### 2021-01-28 Analysis/01_Simulating_spaces_and_stressors.Rmd
Create 01_Simulating_spaces_and_stressors.Rmd

---
title: "Simulating spaces and stressors"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

This is a demo on how to simulate the spaces and apply the different stressors to reflect the mechanisms of interest.

# Simulating spaces

We can use the package `dads` that's in development for simulating the trait spaces.

```{r}
library(devtools)
install_github("TGuillerme/dads")
```

We will build simulate data using a Brownian Motion (the breath of trait values just increases with time, unconstrained), a pure birth tree (no extinction) and will stop the simulation as soon as we reach 100 taxa.
Note that I'll simulate a 2D Brownian Motion trait just for facilitating the visualisation of the stressors, however, for the paper we can stick to one dimension (one trait).

```{r}
## Creating the trait object (2D BM)
my_trait <- make.traits(process = BM.process, n = 2)

## Setting up the birth death parameters (no extinction)
bd_params <- list(speciation = 1)

## Setting up the stopping rules
stop_rule <- list(max.living = 100)

## Simulating the data
simulated_data <- dads(bd.params = bd_params,
                       stop.rule = stop_rule,
                       traits    = my_trait)

## We can then visualise the simulated data (just two show the process)
par(mfrow = c(1, 2))
plot(simulated_data, main = "The simulated BM through time")
plot(simulated_data, main = "The 2D BM", trait = c(1,2))

## The orange data points are node values and will be ignored
## Here we'll select only the tip values from the data
trait_space <- simulated_data$data[rownames(simulated_data$data) %in% simulated_data$tree$tip.label, ]
```

# Simulating the stressors

```{r}

```

### 2021-01-28 Analysis/01_Simulating_spaces_and_stressors.Rmd
Update 01_Simulating_spaces_and_stressors.Rmd
Added random and filtering

@@
 
-We will build simulate data using a Brownian Motion (the breath of trait values just increases with time, unconstrained), a pure birth tree (no extinction) and will stop the simulation as soon as we reach 100 taxa.
+We will build simulate data using a Brownian Motion (the breath of trait values just increases with time, unconstrained), a pure birth tree (no extinction) and will stop the simulation as soon as we reach 200 taxa.
 Note that I'll simulate a 2D Brownian Motion trait just for facilitating the visualisation of the stressors, however, for the paper we can stick to one dimension (one trait).
@@
 ## Setting up the stopping rules
-stop_rule <- list(max.living = 100)
+stop_rule <- list(max.living = 200)
 
@@
 
+The stressors here are the processes we apply to change the normal pattern of the data (the BM motion generated normal distribution of the data) for simulating the different mechanisms.
+To simulate these patterns, we can use the function `dispRity::reduce.space` that has already several algorithms implemented.
+I'm happy to tweak/modify these algorithms (or create new ones!) to get results that better matches our needs.
+
+For the stressors, we can unfortunately only remove species.
+If we take the what-s-done-in-ecology approach this is not a big problem since this is what you'd expect anyways (i.e. your stressors can not "add" species, only make them go extinct).
+So, although we've simulated 200 species, we're gonna resample 100 of them every time.
+
+## Null mechanism
+
+This is the really basic one: just a null model were we randomly sample 100 species out of the 200.
+This can be done using the `stats::sample` function but for making the whole script neater, we can also use the `"random"` algorithm in `dispRity::reduce.space` that just randomly removes n% of species:
+
+```{r}
+## Removing 50% of the elements randomly
+random_reduction <- reduce.space(trait_space,
+                                 type = "random",
+                                 remove = 0.5)
+```
+
+And we can visualise these results compared to the full distribution of 200 species:
+
+```{r}
+## Function for simplifying the plots
+#' @param data the dataset
+#' @param reduction the vector of selected element for the reduction
+#' @param null the vector of selected element for the null
+#' @param col two colours
+#' @param pch the pch (default = 19)
+#' @param trait which trait to plot (default is 1 for plot.1D and c(1,2) for plot.2D)
+#' @param xlab the xlab
+#' @param legend the legend text
+plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch= 19, trait = 1, xlab = "trait values", legend) {
+    ## Plotting the histogram
+    hist(data, main = main, xlab = xlab, col = "white", border = "white")
+    hist(data[null, trait], col = col[1], add = TRUE)
+    hist(data[reduction, trait], col = col[2], add = TRUE)
+    ## Adding the legend
+    if(!missing(legend)) {
+        legend(x = "topleft", col = col, pch = pch, legend = legend)
+    }
+}
+plot.2D <- function(data, reduction, null, main, col = c("blue", "orange"), pch= 19, trait = c(1,2), xlab = "trait values", legend) {
+    ## Plotting the histogram
+    plot(NULL, xlim = range(trait_space[, trait[1]]), ylim = range(trait_space[, trait[2]]), main = main)
+    points(trait_space[null, trait], pch = pch, col = col[1])
+    points(trait_space[reduction, trait], pch = pch, col = col[2])
+    ## Adding the legend
+    if(!missing(legend)) {
+        legend(x = "topleft", col = col, pch = pch, legend = legend)
+    }
+}
+```
+
+```{r}
+## Visualising the results
+par(mfrow = c(1, 2))
+plot.1D(trait_space, reduction = random_reduction, null = rep(TRUE, 200), main = "1D null mechanism", legend = c("Full distribution (200sp)", "Random reduction (100sp)"))
+plot.2D(trait_space, reduction = random_reduction, null = !random_reduction, main = "2D null mechanism")
+```
+
+## Environmental filtering
+
+For the environmental filtering algorithm, we can use the
+We can use the "position" algorithm from the `dispRity` package that shifts that shifts the position of the group @@@
+
+```{r}
+## Removing 50% to shift them
+filtering_reduction <- reduce.space(trait_space,
+                                    type = "position",
+                                    remove = 0.5)
+```
+
+```{r}
+## Visualising the results
+par(mfrow = c(1, 2))
+plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D null mechanism")
+plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D null mechanism")
+```
+
+## Competitive exclusion
+
+Evenness?
+
 ```{r}
@@
 ```
+
+## Facilitation
+
+Adding data in empty spaces
+
+```{r}
+
+```
+
+## Equalizing fitness prosseces
+
+Size
+
+```{r}
+
+```

### 2021-01-29 Analysis/01_Simulating_spaces_and_stressors.Rmd
Update 01_Simulating_spaces_and_stressors.Rmd

@@
 ```{r}
+## Removing 50% to flatten the curve
+competitive_exclusion <- reduce.space(trait_space,
+                                    type = "evenness",
+                                    remove = 0.5)
+```
 
+```{r}
+## Visualising the results
+par(mfrow = c(1, 2))
+plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D null mechanism")
+plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D null mechanism")
+```
+
+## Equalizing fitness processes
+
+Size
+
+```{r}
+## Removing 50% to change the trait space size
+equalizing_fitness <- reduce.space(trait_space,
+                                    type = "size",
+                                    remove = 0.5)
+```
+
+```{r}
+## Visualising the results
+par(mfrow = c(1, 2))
+plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D null mechanism")
+plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D null mechanism")
 ```
@@
 
-Adding data in empty spaces
+Adding data in empty spaces (or clumping things together?)
 
@@
 ```
-
-## Equalizing fitness prosseces
-
-Size
-
-```{r}
-
-```

### 2021-02-18 Analysis/02-Measuring_metric_effect.Rmd
Create 02-Measuring_metric_effect.Rmd

---
title: "Measuring metric effect"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

This is a demo on how to measure the effect of the different stressors from `01_Simulating_spaces_and_stressors.Rmd` on some of our metrics of interest.
This could be done easily using the `dispRity` package.
Here's a simple example.

```{r}
library(dispRity)
library(dads)
library(BAT)
```

## Simulations

Here I'm going to create a wrapping function that simulates the space, apply the stressors and output them into a `"dispRity"` object:

```{r}
## Function for simplifying the plots
#' @param elements the total of elements (before reduction)
#' @param remove the proportion of elements to remove
#' @param stressors the list of stressors as algorithms from dispRity::reduce.space
## Simulating one space 
sim.data <- function(elements, remove, stressors) {
    ## Simulate the data (no stressors)
    evol_data <- dads(stop.rule = list(max.living = elements),
                      bd.params = list(speciation = 1),
                      traits    = make.traits())
    tip_data <- evol_data$data[evol_data$tree$tip.label, , drop = FALSE]
    ## Non evo algorithm (if needed)
    # tip_data <- space.maker(elements = elements, dimensions = 1, distribution = rnorm)
    # rownames(tip_data) <- 1:elements
    ## Apply all the stressors
    stressed_data <- list()
    for(stress in 1:length(stressors)) {
        stressed_data[[stress]] <- reduce.space(tip_data,
                                            remove = remove,
                                            type = stressors[[stress]])
    }
    ## Modify the list to only get the IDs of the selected elements
    names(stressed_data) <- stressors
    stressed_data <- lapply(stressed_data, function(x, data) rownames(data)[which(x)], data = tip_data)

    ## Create the dispRity object
    return(custom.subsets(tip_data, group = stressed_data))
}
```

Here's an example how it works for 10 replicates:

```{r}
## Generating a set of 50 simulations
simulations <- replicate(10, sim.data(elements = 200, remove = 0.5,
                                       stressors = c("random",
                                                     "size",
                                                     "evenness",
                                                     "position",
                                                     "density")),
                          simplify = FALSE)
```

## Metrics

### 2021-02-18 Analysis/01_Simulating_spaces_and_stressors.Rmd
Update 01_Simulating_spaces_and_stressors.Rmd
Wrapped up simulations

--- 2021-01-28 Analysis/01_Simulating_spaces_and_stressors.Rmd
@@
 
-We can use the package `dads` that's in development for simulating the trait spaces.
+We can use the package `dads` that's in development for simulating the trait spaces (note that the package is still pretty unstable but the simple tasks here should not change).
+Later, we can use the `dispRity::reduce.space` package to apply stressors to the data.
+
+```{r, eval = FALSE}
+library(devtools)
+devtools::install_github("TGuillerme/dads")
+devtools::install_github("TGuillerme/dispRity")
+```
 
 ```{r}
-library(devtools)
-install_github("TGuillerme/dads")
+library(dads)     # needs to be version >= 0.1.4
+library(dispRity) # needs to be version >= 1.5.7
 ```
 
-We will build simulate data using a Brownian Motion (the breath of trait values just increases with time, unconstrained), a pure birth tree (no extinction) and will stop the simulation as soon as we reach 100 taxa.
+We will build simulate data using a Brownian Motion (the breath of trait values just increases with time, unconstrained), a pure birth tree (no extinction) and will stop the simulation as soon as we reach 200 taxa.
 Note that I'll simulate a 2D Brownian Motion trait just for facilitating the visualisation of the stressors, however, for the paper we can stick to one dimension (one trait).
@@
 ## We can then visualise the simulated data (just two show the process)
-par(mfrow = c(1, 2))
+par(mfrow = c(2, 1))
 plot(simulated_data, main = "The simulated BM through time")
@@
 
+The stressors here are the processes we apply to change the normal pattern of the data (the BM motion generated normal distribution of the data) for simulating the different mechanisms.
+To simulate these patterns, we can use the function `dispRity::reduce.space` that has already several algorithms implemented.
+I'm happy to tweak/modify these algorithms (or create new ones!) to get results that better matches our needs.
+
+For the stressors, we can unfortunately only remove species.
+If we take the what-s-done-in-ecology approach this is not a big problem since this is what you'd expect anyways (i.e. your stressors can not "add" species, only make them go extinct).
+So, although we've simulated 200 species, we're gonna resample 100 of them every time.
+
+## Null mechanism
+
+This is the really basic one: just a null model were we randomly sample 100 species out of the 200.
+This can be done using the `stats::sample` function but for making the whole script neater, we can also use the `"random"` algorithm in `dispRity::reduce.space` that just randomly removes n% of species:
+
 ```{r}
+## Removing 50% of the elements randomly
+random_reduction <- reduce.space(trait_space,
+                                 type = "random",
+                                 remove = 0.5)
+```
 
+And we can visualise these results compared to the full distribution of 200 species:
+
+```{r}
+## Function for simplifying the plots
+#' @param data the dataset
+#' @param reduction the vector of selected element for the reduction
+#' @param null the vector of selected element for the null
+#' @param col two colours
+#' @param pch the pch (default = 19)
+#' @param trait which trait to plot (default is 1 for plot.1D and c(1,2) for plot.2D)
+#' @param xlab the xlab
+#' @param legend the legend text
+plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch = 19, trait = 1, xlab = "trait values", legend) {
+    ## Get the histograms
+    main_plot <- hist(data, plot = FALSE)
+    plot_1 <- hist(data[null, trait], plot = FALSE, breaks = main_plot$breaks)
+    plot_2 <- hist(data[reduction, trait], plot = FALSE, breaks = main_plot$breaks)
+    plot_height <- max(plot_1$counts, plot_2$counts)
+
+    ## Make the colours as transparent (for overlaps)
+    make.transparent <- function(col, alpha = 0.5) {
+      rgb_vals <- grDevices::col2rgb(col)
+      rgb_args <- list(red   = rgb_vals["red", ],
+                       green = rgb_vals["green", ],
+                       blue  = rgb_vals["blue", ],
+                       maxColorValue = 255,
+                       alpha = alpha * 255)
+      return(do.call(grDevices::rgb, rgb_args))
+    }
+    ## Plot the histograms
+    plot(plot_1, col = make.transparent(col[1]), ylim = c(0, plot_height),
+        main = ifelse(missing(main), "", main),
+        xlab = "Trait values", ylab = "Counts")
+    plot(plot_2, col = make.transparent(col[2]), add = TRUE)
+
+    ## Adding the legend
+    if(!missing(legend)) {
+        legend(x = "topleft", col = col, pch = pch, legend = legend)
+    }
+    ## Add the points
+    points(x = data[null, trait], y = rep(0, sum(null)), col = "black", bg = col[1], pch = 21)
+    points(x = data[reduction, trait], y = rep(-1, sum(reduction)), col = "black", bg = col[2], pch = 21)
+}
+plot.2D <- function(data, reduction, null, main, col = c("blue", "orange"), pch= 19, trait = c(1,2), xlab = "trait values", legend) {
+    ## Plotting the histogram
+    plot(NULL, xlim = range(trait_space[, trait[1]]), ylim = range(trait_space[, trait[2]]), main = main)
+    points(trait_space[null, trait], pch = pch, col = col[1])
+    points(trait_space[reduction, trait], pch = pch, col = col[2])
+    ## Adding the legend
+    if(!missing(legend)) {
+        legend(x = "topleft", col = col, pch = pch, legend = legend)
+    }
+}
 ```
+
+```{r}
+## Visualising the results
+par(mfrow = c(2, 1))
+plot.1D(trait_space, reduction = random_reduction, null = rep(TRUE, 200), main = "1D null mechanism", legend = c("Full distribution (200sp)", "Random reduction (100sp)"))
+plot.2D(trait_space, reduction = random_reduction, null = !random_reduction, main = "2D null mechanism")
+```
+
+## Environmental filtering
+
+For the environmental filtering algorithm, we can use the
+We can use the `"position"` algorithm from the `dispRity::reduce.space` function that shifts the position of the group to one corner of the space so that X% of the data in on side and (1-X)% is on the other side (with side being weirder and weirder as dimensionality increases!):
+
+```{r}
+## Removing 50% to shift them
+filtering_reduction <- reduce.space(trait_space,
+                                    type = "position",
+                                    remove = 0.5)
+```
+
+```{r}
+## Visualising the results
+par(mfrow = c(2, 1))
+plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D null mechanism")
+plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D null mechanism")
+```
+
+## Competitive exclusion
+
+For simulating the competitive exclusion, we could use the `dispRity::reduce.space` `"evenness"` algorithm that basically "flattens" the curve (i.e.
+displacing things from the highest density regions to the sides).
+
+```{r}
+## Removing 50% to flatten the curve
+competitive_exclusion <- reduce.space(trait_space,
+                                    type = "evenness",
+                                    remove = 0.5,
+                                    parameters = list(power = 3))
+```
+
+```{r}
+## Visualising the results
+par(mfrow = c(2, 1))
+plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D null mechanism")
+plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D null mechanism")
+```
+
+## Equalising fitness processes
+
+For simulating equalisation of fitness we can use the `"size"` algorithm from `dispRity::reduce.space`.
+This removes elements around past a distance from the centre of the population (with the distance being determined automatically so that X% are inside the radius and 1-X% are outside):
+
+```{r}
+## Removing 50% to change the trait space size
+equalizing_fitness <- reduce.space(trait_space,
+                                    type = "size",
+                                    remove = 0.5)
+```
+
+```{r}
+## Visualising the results
+par(mfrow = c(2, 1))
+plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D null mechanism")
+plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D null mechanism")
+```
+
+## Facilitation
+
+This one might be a bit trickier (though doable).
+One way to do it (as suggested in our last meeting) could be to simply add a number of data points to places that already have data. 
+However, this will result in non-comparable number of elements in each simulations.
+
+One alternative (at least for now) would be to use the `"density"` algorithm from `dispRity::reduce.space` that will reduce the nearest neighbours distance between elements (basically making clumps of two elements close to each other).
+
+```{r}
+## Removing 50% to change the trait space size
+facilitation <- reduce.space(trait_space,
+                             type = "density",
+                             remove = 0.5)
+```
+
+```{r}
+## Visualising the results
+par(mfrow = c(2, 1))
+plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D null mechanism")
+plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D null mechanism")
+```

### 2021-02-18 Analysis/01_Simulating_spaces_and_stressors.Rmd
updated simulation tests

@@
 date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
 output:
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 6}
 ## Creating the trait object (2D BM)
@@
     points(x = data[null, trait], y = rep(0, sum(null)), col = "black", bg = col[1], pch = 21)
-    points(x = data[reduction, trait], y = rep(-1, sum(reduction)), col = "black", bg = col[2], pch = 21)
+    points(x = data[reduction, trait], y = rep(-plot_height/50, sum(reduction)), col = "black", bg = col[2], pch = 21)
 }
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 6}
 ## Visualising the results
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 6}
 ## Visualising the results
 par(mfrow = c(2, 1))
-plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D null mechanism")
-plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D null mechanism")
+plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
+plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D filtering mechanism")
 ```
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 6}
+
 ## Visualising the results
 par(mfrow = c(2, 1))
-plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D null mechanism")
-plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D null mechanism")
+plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D excluding mechanism")
+plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D excluding mechanism")
 ```
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 6}
+
 ## Visualising the results
 par(mfrow = c(2, 1))
-plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D null mechanism")
-plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D null mechanism")
+plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
+plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")
 ```
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 6}
 ## Visualising the results
 par(mfrow = c(2, 1))
-plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D null mechanism")
-plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D null mechanism")
+plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
+plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D facilitation mechanism")
 ```

### 2021-05-13 Functions/make.simulation.pipeline.R
INIT make.simulation.pipeline

@@ -1,0 +1,27 @@
+#' @title Make the simulation pipeline
+#'
+#' @description Set the variables for running one simulation
+#'
+#' @param sim.data A named list of parameters for the data simulation (default is \code{list{n.traits = 2, speciation = 1, n.taxa = 200}})
+#' @param verbose whether to be verbose (\code{TRUE}, default) or not (\code{FALSE})
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+make.simulation.pipeline <- function(sim.data = list(n.traits   = 2,
+                                                     speciation = 1,
+                                                     n.taxa     = 200),
+                                     verbose  = TRUE) {
+
+
+
+n_traits <- 2
+speciation_rate <- 1
+n_taxa <- 200
+
+    return()
+}

### 2021-05-13 Functions/simulation.pipeline.R
Create simulation.pipeline.R

@@ -1,0 +1,215 @@
+#' @title Simulation pipeline
+#'
+#' @description One simulation pipeline
+#'
+#' @param sim.data A named list of parameters for the data simulation (default is \code{list{n.traits = 2, speciation = 1, n.taxa = 200}})
+#' @param type Which removal type (either \code{"facilitation"}, \code{"equalizing"}, \code{"competition"} or \code{"filtering"})}) to be paired with \code{"random"}.
+#' @param remove How much data to remove (can be a vector of multiple percentages)
+#' @param verbose whether to be verbose (\code{TRUE}, default) or not (\code{FALSE})
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+# sim.data = list(n.traits = 2, speciation = 1, n.taxa = 200)
+# type = "facilitation"
+# remove = c(0.2,.4, 0.6 ,0.8)
+# library(dads)
+# library(dispRity)
+# library(BAT)
+# source("melodic.rao.R")
+
+
+# test <- simulation.pipeline(sim.data = sim.data, type = "facilitation", remove = remove, verbose = TRUE, record.timer = TRUE)
+
+simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.timer = FALSE) {
+
+    if(record.timer) {
+        time_start <- Sys.time()
+    }
+
+    ## Simulating the brownian birth death data
+    simulated_data <- dads::dads(
+            bd.params = list(speciation = sim.data$speciation),
+            stop.rule = list(max.living = sim.data$n.taxa),
+            traits    = dads::make.traits(process = dads::BM.process, n = sim.data$n.traits))
+    trait_space <- simulated_data$data[rownames(simulated_data$data) %in% simulated_data$tree$tip.label, ]
+
+    ## Applying the stressors
+    random_reductions <- sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "random", remove = remove)), trait_space, simplify = FALSE)
+    ## Switch the different types of other reductions
+    target_reductions <- switch(type,
+        "facilitation" = sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "density", remove = remove)), trait_space, simplify = FALSE),
+        "equalizing"   =  sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "size", remove = remove)), trait_space, simplify = FALSE),
[…]

### 2021-05-13 Functions/melodic.rao.R
Create melodic.rao.R

@@ -1,0 +1,30 @@
+##MEan DIssimilarity Components##
+#samp: community matrix; sites in lines, species in columns
+#dis: dissimilarity matrix
+
+melodic.rao <- function(samp,dis,abundance.weighted=TRUE){
+	N<-dim(samp)[1]
+	rao<-richness<-mpd<-simpson<-numeric(N)
+	for (i in 1:N){
+		sppInSample<-names(samp[i,samp[i, ]>0])
+	    #richness[i]<-length(sppInSample)
+		if (length(sppInSample)>1){
+      		sample.dis<-dis[sppInSample,sppInSample]
+			abund.w<-numeric(length(sppInSample))
+			if (abundance.weighted){
+				abund.w <- samp[i , sppInSample] / sum(samp[i , sppInSample])
+			} 	else{
+				abund.w <- rep(1 , length(sppInSample)) / length(sppInSample)
+			}	
+        	sample.weights <- outer(as.numeric( abund.w) , as.numeric( abund.w))
+		  	rao[i] <- sum(sample.weights * sample.dis)
+		  	mpd[i] <- weighted.mean(sample.dis[lower.tri(sample.dis)] , sample.weights[lower.tri(sample.weights)])
+		  	simpson[i]<-sum(2*sample.weights[lower.tri(sample.weights)])
+
+		}	else {
+			rao[i]<- mpd[i]  <- simpson[i]<- 0
+		}
+	}	
+	out<-list(rao=rao, mpd=mpd, simpson=simpson)
+	return(out)	
+}

### 2021-05-13 Functions/make.simulation.pipeline.R
Update make.simulation.pipeline.R

@@ -12,10 +12,15 @@
 #' @author Thomas Guillerme
 #' @export
 
-make.simulation.pipeline <- function(sim.data = list(n.traits   = 2,
+make.simulation.pipeline <- function(type,
+                                     sim.data = list(n.traits   = 2,
                                                      speciation = 1,
                                                      n.taxa     = 200),
-                                     verbose  = TRUE) {
+                                     remove = c(0.2, 0.4, 0.6, 0.8),
+                                     verbose  = TRUE,
+                                     record.timer = TRUE) {
+    return(function() simulation.pipeline(sim.data, type, remove, verbose = verbose, record.timer = record.timer))
+}
 
 
 

### 2021-05-14 Analysis/01_Simulating_spaces_and_stressors.Rmd
Cleaning stressor simulation explanations

@@
 ```{r}
-## Function for simplifying the plots
-#' @param data the dataset
-#' @param reduction the vector of selected element for the reduction
-#' @param null the vector of selected element for the null
-#' @param col two colours
-#' @param pch the pch (default = 19)
-#' @param trait which trait to plot (default is 1 for plot.1D and c(1,2) for plot.2D)
-#' @param xlab the xlab
-#' @param legend the legend text
-plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch = 19, trait = 1, xlab = "trait values", legend) {
-    ## Get the histograms
-    main_plot <- hist(data, plot = FALSE)
-    plot_1 <- hist(data[null, trait], plot = FALSE, breaks = main_plot$breaks)
-    plot_2 <- hist(data[reduction, trait], plot = FALSE, breaks = main_plot$breaks)
-    plot_height <- max(plot_1$counts, plot_2$counts)
-
-    ## Make the colours as transparent (for overlaps)
-    make.transparent <- function(col, alpha = 0.5) {
-      rgb_vals <- grDevices::col2rgb(col)
-      rgb_args <- list(red   = rgb_vals["red", ],
-                       green = rgb_vals["green", ],
-                       blue  = rgb_vals["blue", ],
-                       maxColorValue = 255,
-                       alpha = alpha * 255)
-      return(do.call(grDevices::rgb, rgb_args))
-    }
-    ## Plot the histograms
-    plot(plot_1, col = make.transparent(col[1]), ylim = c(0, plot_height),
-        main = ifelse(missing(main), "", main),
-        xlab = "Trait values", ylab = "Counts")
-    plot(plot_2, col = make.transparent(col[2]), add = TRUE)
-
-    ## Adding the legend
-    if(!missing(legend)) {
-        legend(x = "topleft", col = col, pch = pch, legend = legend)
-    }
-    ## Add the points
-    points(x = data[null, trait], y = rep(0, sum(null)), col = "black", bg = col[1], pch = 21)
-    points(x = data[reduction, trait], y = rep(-plot_height/50, sum(reduction)), col = "black", bg = col[2], pch = 21)
-}
-plot.2D <- function(data, reduction, null, main, col = c("blue", "orange"), pch= 19, trait = c(1,2), xlab = "trait values", legend) {
-    ## Plotting the histogram
-    plot(NULL, xlim = range(trait_space[, trait[1]]), ylim = range(trait_space[, trait[2]]), main = main)
-    points(trait_space[null, trait], pch = pch, col = col[1])
-    points(trait_space[reduction, trait], pch = pch, col = col[2])
-    ## Adding the legend
-    if(!missing(legend)) {
-        legend(x = "topleft", col = col, pch = pch, legend = legend)
-    }
-}
+## Loading the plotting functions
+source("../Functions/plot.fun.R")
 ```

### 2021-05-14 Functions/plot.fun.R

@@ -1,0 +1,50 @@
+## Function for simplifying the plots
+#' @param data the dataset
+#' @param reduction the vector of selected element for the reduction
+#' @param null the vector of selected element for the null
+#' @param col two colours
+#' @param pch the pch (default = 19)
+#' @param trait which trait to plot (default is 1 for plot.1D and c(1,2) for plot.2D)
+#' @param xlab the xlab
+#' @param legend the legend text
+plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch = 19, trait = 1, xlab = "trait values", legend) {
+    ## Get the histograms
+    main_plot <- hist(data, plot = FALSE)
+    plot_1 <- hist(data[null, trait], plot = FALSE, breaks = main_plot$breaks)
+    plot_2 <- hist(data[reduction, trait], plot = FALSE, breaks = main_plot$breaks)
+    plot_height <- max(plot_1$counts, plot_2$counts)
+
+    ## Make the colours as transparent (for overlaps)
+    make.transparent <- function(col, alpha = 0.5) {
+      rgb_vals <- grDevices::col2rgb(col)
+      rgb_args <- list(red   = rgb_vals["red", ],
+                       green = rgb_vals["green", ],
+                       blue  = rgb_vals["blue", ],
+                       maxColorValue = 255,
+                       alpha = alpha * 255)
+      return(do.call(grDevices::rgb, rgb_args))
+    }
+    ## Plot the histograms
+    plot(plot_1, col = make.transparent(col[1]), ylim = c(0, plot_height),
+        main = ifelse(missing(main), "", main),
+        xlab = "Trait values", ylab = "Counts")
+    plot(plot_2, col = make.transparent(col[2]), add = TRUE)
+
+    ## Adding the legend
+    if(!missing(legend)) {
+        legend(x = "topleft", col = col, pch = pch, legend = legend)
+    }
+    ## Add the points
+    points(x = data[null, trait], y = rep(0, sum(null)), col = "black", bg = col[1], pch = 21)
+    points(x = data[reduction, trait], y = rep(-plot_height/50, sum(reduction)), col = "black", bg = col[2], pch = 21)
+}
+plot.2D <- function(data, reduction, null, main, col = c("blue", "orange"), pch= 19, trait = c(1,2), xlab = "trait values", legend) {
+    ## Plotting the histogram
[…]

### 2021-05-14 Analysis/02_Simulation_pipeline.Rmd
Simulation pipeline INIT

---
title: "Simulation pipeline"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

This script is the repeatable script for measuring changes in the different diversity metrics and how they are affected by the different stressors.

## Requirements for the simulations

In order to run the simulations (on your machine, cluster, etc...) you'll need to load the following in your current environment:

```{r, eval = FALSE}
## Getting the non-CRAN (yet) libraries
library(devtools)
devtools::install_github("TGuillerme/dads")
```

```{r}
## Libraries
require(dads)
require(dispRity)
require(BAT)
require(FD)
require(ks)
require(TPD)
```

> Note that the libraries do not need to be loaded (i.e. no need for `library()`) but needs to exist.

You'll also need the following custom functions (located in [here on github](https://github.com/TGuillerme/eco_metrics_simulations/tree/master/Functions))

```{r}
## Loading the functions
source("../Functions/melodic.rao.R")
source("../Functions/optim.replicate.R")
source("../Functions/make.simulation.pipeline.R")
source("../Functions/simulation.pipeline.R")
```

## Setting up the simulations

## Subtitle

```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
```

### 2021-05-14 Functions/make.simulation.pipeline.R
Simulation functions

@@ -21,12 +21,3 @@ make.simulation.pipeline <- function(type,
                                      record.timer = TRUE) {
     return(function() simulation.pipeline(sim.data, type, remove, verbose = verbose, record.timer = record.timer))
 }
-
-
-
-n_traits <- 2
-speciation_rate <- 1
-n_taxa <- 200
-
-    return()
-}

### 2021-05-14 Functions/simulation.pipeline.R

@@ -207,9 +207,11 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
         all_results[[2]] <- NULL
     }
 
+    
+
     if(record.timer) {
-        return(list("results" = results, "trait_space" = trait_space, "reductions" = presences, "timer" = timer))
+        return(list("results" = all_results, "trait_space" = trait_space, "reductions" = presences, "timer" = timer))
     } else {
-        return(list("results" = results, "trait_space" = trait_space, "reductions" = presences))
+        return(list("results" = all_results, "trait_space" = trait_space, "reductions" = presences))
     }
 }

### 2021-05-14 Functions/optim.replicate.R
optimisation function

@@ -1,0 +1,137 @@
+#' @title Optimised replicates
+#'
+#' @description A wrapper for replicating a based on results variance
+#'
+#' @param fun The function to execute. The function should output a single value or a vector or a list of values.
+#' @param diagnose The diagnosis function should intake a vector of values.
+#' @param summarise Optional, if the output of \code{fun()} is complex, a function to summarise that output and pass it to \code{diagnose()}.
+#' @param minimum The minimum of replicates to run (default is 20)
+#' @param maximum Optional, a maximum of iterations (default is Inf).
+#' @param stop.variance The percentage of variance change for stopping (default is 0.05).
+#' @param increment The percentage of increment (if left empty, the value from stop.variance is used).
+#' @param verbose Whether to be verbose (\code{TRUE}) or not (\code{FALSE}; default)
+#' @param parallel Whether to use the \code{\link[future.apply]{future_replicate}} function for parallelisation (\code{TRUE}) or not (\code{FALSE}; default).
+#' @param par.plan If \code{parallel = TRUE}, which plan to use (see \code{\link[future]{plan}}; by default, the plan is "future::multisession").
+#' @param ... Any optional arguments to be passed to the function.
+#' 
+#' @return
+#' A list of:
+#' \itemize{
+#'      \item{results}: a \code{matrix} with the summarised output from \code{fun()} for every set of iterations;
+#'      \item{diagnoses}: a \code{matrix} with changes in variances after each diagnosis;
+#'      \item{outputs}: a \code{list} of the outputs of each iteration of \code{fun()}.
+#' }
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+optim.replicate <- function(fun, diagnose, summarise, minimum = 20, maximum = Inf, stop.variance = 0.05, increment, verbose = FALSE, parallel = FALSE, par.plan = "future::multisession", ...) {
+
+    ## Check increment
+    if(missing(increment)) {
+        increment <- stop.variance
+    }
+
+    ## Check diagnosis
+    if(missing(diagnose)) {
+        stop("Impossible to optimize the replicate loop without a diagnose function.")
+    }
+
+    ## Check summary
+    if(missing(summarise)) {
+        summarise <- function(x) return(x)
+    }
+
+    ## Set the parallel (or not)
[…]

### 2021-05-14 Functions/simulation.pipeline.R
Working pipeline

@@ -6,6 +6,7 @@
 #' @param type Which removal type (either \code{"facilitation"}, \code{"equalizing"}, \code{"competition"} or \code{"filtering"})}) to be paired with \code{"random"}.
 #' @param remove How much data to remove (can be a vector of multiple percentages)
 #' @param verbose whether to be verbose (\code{TRUE}, default) or not (\code{FALSE})
+#' @param record.timer whether to record the timer for each step of the simulations
 #' 
 #' @examples
 #'
@@ -25,28 +26,54 @@
 
 # test <- simulation.pipeline(sim.data = sim.data, type = "facilitation", remove = remove, verbose = TRUE, record.timer = TRUE)
 
+## SOLVED: Errors: Error in hclust(as.dist(thisTree)) : must have n >= 2 objects to cluster
+# set.seed(4)
+test <- simulation.pipeline(sim.data, type = "competition", remove = remove, verbose = TRUE)
 simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.timer = FALSE) {
 
     if(record.timer) {
         time_start <- Sys.time()
     }
 
-    ## Simulating the brownian birth death data
-    simulated_data <- dads::dads(
-            bd.params = list(speciation = sim.data$speciation),
-            stop.rule = list(max.living = sim.data$n.taxa),
-            traits    = dads::make.traits(process = dads::BM.process, n = sim.data$n.traits))
-    trait_space <- simulated_data$data[rownames(simulated_data$data) %in% simulated_data$tree$tip.label, ]
+    if(verbose) {
+        message("Simulating the trait space:", appendLF = FALSE)
+    }
 
-    ## Applying the stressors
-    random_reductions <- sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "random", remove = remove)), trait_space, simplify = FALSE)
-    ## Switch the different types of other reductions
-    target_reductions <- switch(type,
-        "facilitation" = sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "density", remove = remove)), trait_space, simplify = FALSE),
-        "equalizing"   =  sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "size", remove = remove)), trait_space, simplify = FALSE),
[…]

### 2021-05-14 Functions/optim.helpers.R
INIT optim.helpers

@@ -1,0 +1,35 @@
+#' @title The diagnosis function for optim.replicate
+#'
+#' @description 
+#'
+#' @param 
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+optim.diagnose <- function() {
+
+    return()
+}
+
+#' @title The summary function for optim.replicate
+#'
+#' @description 
+#'
+#' @param 
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+#' 
+optim.summarise <- function() {
+
+    return()
+}

### 2021-05-14 Functions/make.simulation.pipeline.R
Update make.simulation.pipeline.R

@@ -2,9 +2,11 @@
 #'
 #' @description Set the variables for running one simulation
 #'
+#' @param type Which removal type (either \code{"facilitation"}, \code{"equalizing"}, \code{"competition"} or \code{"filtering"})}) to be paired with \code{"random"}.
 #' @param sim.data A named list of parameters for the data simulation (default is \code{list{n.traits = 2, speciation = 1, n.taxa = 200}})
+#' @param remove How much data to remove (can be a vector of multiple percentages)
 #' @param verbose whether to be verbose (\code{TRUE}, default) or not (\code{FALSE})
-#' 
+#' @param record.timer whether to record the timer for each step of the simulations#' 
 #' @examples
 #'
 #' @seealso

### 2021-05-14 Analysis/02_Simulation_pipeline.Rmd
Update 02_Simulation_pipeline.Rmd

@@
 source("../Functions/melodic.rao.R")
-source("../Functions/optim.replicate.R")
 source("../Functions/make.simulation.pipeline.R")
@@
 
-## Subtitle
+You then need to set up the simulation parameters using `make.simulation.pipeline()`.
+You can check the manual of that function by opening the function to see the different options but we suggest you use the default to replicate the parameters used in the manuscript.
 
-```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
+By default, the simulation does:
+
+ * Create a tree from a pure birth process with 2 independent birth-death processes to make a 2D trait space. The speciation (birth) rate is set to 1 and the simulation stops when reaching 200 taxa.
+ * Randomly remove 20%, 40%, 60% and 80% of the data to simulate the random (null) stressor.
+ * Remove 20%, 40%, 60% and 80% of the data using the required `type` of stressor (see below and/or `01_Simulating_spaces_and_stressors.Rmd`).
+ * Measure five families of metrics on the randomly reduced and stressor reduced spaces: dendrogram, dissimilarity, convex hull, kernel hypervolume and probability density based metrics.
+ * All the results are then collected into a table with the two stressors as rows and the different 15 metrics as columns.
+
+You then just need to set up which type of stressor you are going to simulate the metrics with:
+
+```{r}
+## Creating the function for simulating facilitation
+simulate.facilitation <- make.simulation.pipeline(type = "facilitaion")
 ```
+
+You can then just run one replicate of the simulation by simply calling that function like (not run here):
+
+```{r, eval = FALSE}
+## Running one replicate
+simulate.facilitation()
+```
+
+## Running the simulations
+
+In order to simulate enough replicates we will replicate the simulation pipeline $n$ times until adding 5% more replicates does not increase the overall variance by more than 5% for any metric.
+In other words, we replicated each simulation $n$ times where $n$ is estimated so that $var(n + 0.05 \times n) < 0.05 \times var(n)$.
+
+For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `simulate.facilitation()` for diagnosis) function
+
+```{r}
+## Loading the optimisation functions
+source("../Functions/optim.replicate.R")
+source("../Functions/optim.helpers.R")
+```
+
+```{r}
+simulate.facilitation <- make.simulation.pipeline(type = "facilitation")
+```

### 2021-05-17 Functions/optim.helpers.R
Updated simulation pipeline

@@ -13,6 +13,8 @@
 
 optim.diagnose <- function() {
 
+    ## Combine all the replicates together
+    #test
     return()
 }
 
@@ -29,7 +31,16 @@ optim.diagnose <- function() {
 #' @author Thomas Guillerme
 #' @export
 #' 
-optim.summarise <- function() {
+optim.summarise <- function(x) {
+    ## Get all the values
+    output <- c(unlist(x$results))
 
-    return()
+    ## Names for one result
+    names <- c(paste(rownames(x$results[[1]])[1], colnames(x$results[[1]]), sep = "_"),
+               paste(rownames(x$results[[1]])[2], colnames(x$results[[1]]), sep = "_"))
+    names <- paste(names, rep(paste0("rm", rep(1:length(x$results))), each = length(names)), sep = "_")
+
+    names(output) <- names
+
+    return(output)
 }

### 2021-05-17 Functions/optim.replicate.R

@@ -1,7 +1,7 @@
 #' @title Optimised replicates
 #'
 #' @description A wrapper for replicating a based on results variance
-#'
+#' 
 #' @param fun The function to execute. The function should output a single value or a vector or a list of values.
 #' @param diagnose The diagnosis function should intake a vector of values.
 #' @param summarise Optional, if the output of \code{fun()} is complex, a function to summarise that output and pass it to \code{diagnose()}.
@@ -12,7 +12,11 @@
 #' @param verbose Whether to be verbose (\code{TRUE}) or not (\code{FALSE}; default)
 #' @param parallel Whether to use the \code{\link[future.apply]{future_replicate}} function for parallelisation (\code{TRUE}) or not (\code{FALSE}; default).
 #' @param par.plan If \code{parallel = TRUE}, which plan to use (see \code{\link[future]{plan}}; by default, the plan is "future::multisession").
+#' @param bkp.path Optional, a \code{"character"} string to be the path to a backup file. See details.
+#' @param bkp.name Optional, if \code{bkp.path} is not missing, a \code{"character"} for the name of the file to save (or to load). See details.
 #' @param ... Any optional arguments to be passed to the function.
+#' 
+#' @details When \code{bkp} is specified, the algorithm first looks if a file with \code{bkp.name} exists (by default, the name is \code{"optim.replicate.bkp"}). If not, the results will be saved after each series of replicates into this file. If it exist, the initial steps are skipped and the algorithm continues from the last backup point. \emph{NOTE} that this option slows down the simulations.
 #' 
 #' @return
 #' A list of:
@@ -28,7 +32,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-optim.replicate <- function(fun, diagnose, summarise, minimum = 20, maximum = Inf, stop.variance = 0.05, increment, verbose = FALSE, parallel = FALSE, par.plan = "future::multisession", ...) {
+optim.replicate <- function(fun, diagnose, summarise, minimum = 20, maximum = Inf, stop.variance = 0.05, increment, verbose = FALSE, parallel = FALSE, par.plan = "future::multisession", bkp.path, bkp.name, ...) {
 
     ## Check increment
     if(missing(increment)) {
@@ -55,37 +59,74 @@ optim.replicate <- function(fun, diagnose, summarise, minimum = 20, maximum = In
[…]

### 2021-05-17 Functions/simulation.pipeline.R

@@ -28,7 +28,7 @@
 
 ## SOLVED: Errors: Error in hclust(as.dist(thisTree)) : must have n >= 2 objects to cluster
 # set.seed(4)
-test <- simulation.pipeline(sim.data, type = "competition", remove = remove, verbose = TRUE)
+# test <- simulation.pipeline(sim.data, type = "facilitation", remove = remove, verbose = TRUE)
 simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.timer = FALSE) {
 
     if(record.timer) {
@@ -36,7 +36,7 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
     }
 
     if(verbose) {
-        message("Simulating the trait space:", appendLF = FALSE)
+        message("\nSimulating the trait space:", appendLF = FALSE)
     }
 
     ## Make empty reductions (all FALSE) for checking if the reduction worked
@@ -139,6 +139,7 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
 
     ## Run the convex.hull analyses
     hypervolume_results <- lapply(presences, fun.kernel, traits = trait_space, verbose)
+    # hypervolume_results <- lapply(presences, fun.tree, tree = tree, verbose) ; warning("DEBUG")
 
     if(record.timer) {
         time_end <- Sys.time()
@@ -154,6 +155,7 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
 
     ## Get the results
     TPD_results <- lapply(presences, fun.proba.den, traits = trait_space, verbose)
+    # TPD_results <- lapply(presences, fun.tree, tree = tree, verbose) ; warning("DEBUG")
 
     if(verbose) {
         message("Done.\n", appendLF = FALSE)

### 2021-05-17 Analysis/02_Simulation_pipeline.Rmd
Update 02_Simulation_pipeline.Rmd

@@
 ## Creating the function for simulating facilitation
-simulate.facilitation <- make.simulation.pipeline(type = "facilitaion")
+simulate.facilitation <- make.simulation.pipeline(type = "facilitation")
 ```
@@
 ```{r}
-simulate.facilitation <- make.simulation.pipeline(type = "facilitation")
+## Running the full optimisation
+results <- optim.replicate(fun = simulate.facilitation(),
+                           diagnose = var,
+                           summarise = optim.summarise,
+                           minimum = 20, maximum = 100,
+                           stop.variance = 0.05,
+                           verbose = TRUE,
+                           bkp.path = "../Data/Processed/",
+                           bkp.name = "sim_facilitation.rda")
 ```

### 2021-05-17 Analysis/02_Simulation_pipeline.Rmd
sim pipeline updated

@@
 ## Running the full optimisation
-results <- optim.replicate(fun = simulate.facilitation(),
+results <- optim.replicate(input.fun = simulate.facilitation,
                            diagnose = var,

### 2021-05-18 Functions/analyse.replicates.R
Fixed maximum

@@ -1,0 +1,19 @@
+#' @title Analyse replicates
+#' 
+#' @description Analyse the replicates results
+#'
+#' @param results the results list or a path to the file
+#' @param what optional, a sub-selection of the results to plot
+#' @param ... any arguments to be passed to \code{lines}.
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+analyse.replicates <- function(results) {
+
+    return()
+}

### 2021-05-18 Functions/plot.results.R
summary and plot prototypes

@@ -1,0 +1,19 @@
+#' @title Plot results
+#'
+#' @description Plotting the results
+#'
+#' @param results the results matrix (typically from summarise.results)
+#' @param what optional, a sub-selection of the results to plot
+#' @param ... any arguments to be passed to \code{lines}.
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+plot.results <- function(results, what, ...) {
+
+    return()
+}

### 2021-05-18 Functions/summarise.results.R

@@ -1,0 +1,20 @@
+#' @title Summarise the results
+#'
+#' @description Summarises the results (from the replications)
+#'
+#' @param results the results or the path to the results
+#' @param what optional, which specific results to summarise
+#' @param scale logical, whether to scale the results relative to the the random reduction (TRUE, default) or not (FALSE).
+#' @param time optional, whether to summarise the running time (TRUE) or the results (FALSE, default)
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+summarise.results <- function(results, what, scale = TRUE, time = FALSE) {
+
+    return()
+}

### 2021-05-18 Functions/analyse.replicates.R
Update analyse.replicates.R

@@ -3,7 +3,9 @@
 #' @description Analyse the replicates results
 #'
 #' @param results the results list or a path to the file
-#' @param what optional, a sub-selection of the results to plot
+#' @param what optional, a sub-selection of the results to plot (either the names of the indices)
+#' @param scale whether to scale the variance for each diagnoses (TRUE) or not (FALSE, default)
+#' @param net.change whether to plot the variance net change (TRUE, default) or not (FALSE).
 #' @param ... any arguments to be passed to \code{lines}.
 #' 
 #' @examples
@@ -13,7 +15,87 @@
 #' @author Thomas Guillerme
 #' @export
 
-analyse.replicates <- function(results) {
+analyse.replicates <- function(results, what, scale = FALSE, net.change = TRUE, ...) {
 
-    return()
+    ## Loading the results from a backup file
+    if(is(results, "character")) {
+        name <- load(results)
+        eval(parse(text = paste0("results <- ", name)))
+    }
+
+    ## Extracting the diagnoses
+    diagnoses <- results$diagnosis
+
+    ## What to plot
+    if(missing(what)) {
+        what <- colnames(results$diagnosis)
+    } else {
+        if(is(what, "numeric")) {
+            what <- colnames(results$diagnosis)[what]
+        }
+    }
+
+    ## Getting the net changes
+    if(net.change) {
+        diagnoses <- abs(diagnoses[-nrow(diagnoses), ]/diagnoses[-1, ]-1)
+        diagnoses[nrow(diagnoses), ] <- 0
+    }
+
+    ## Scaling the values
+    if(scale) {
+        diagnoses <- apply(diagnoses, 2, function(x) x/max(x))
+    }
+
+    ## plot_params
+    plot_params <- list(...)
+    if(is.null(plot_params$xlim)) {
+        plot_params$xlim <- range(as.numeric(rownames(diagnoses)))
+    }
+    if(is.null(plot_params$ylim)) {
+        plot_params$ylim <- range(diagnoses)
+    }
+    if(is.null(plot_params$ylab)) {
+        plot_params$ylab <- "variance"
+        if(net.change) {
+            plot_params$ylab <- paste0("net ", plot_params$ylab, " change")
+        }
+        if(scale) {
+            plot_params$ylab <- paste0("scaled ", plot_params$ylab)
+        }
+    }
+    if(is.null(plot_params$xlab)) {
+        plot_params$xlab <- "iterations"
+    }
[…]

### 2021-05-18 Analysis/02_Simulation_pipeline.Rmd
Update 02_Simulation_pipeline.Rmd

--- 2021-05-17 Analysis/02_Simulation_pipeline.Rmd
@@
 source("../Functions/optim.helpers.R")
+source("../Functions/analyse.replicates.R")
 ```
 
-```{r}
+You can then run this replicate function that will start by running 20 replicates (`mimimum`) and will increase the number of replicates by 5% until reaching 1000 replicates (`maximum`) or until the addition 5% increase does not increase the variance for each metric with each different level of removal by less than 5% (120 parameters!).
+
+There are options to
+
+```{r, eval = FALSE}
 ## Running the full optimisation
-results <- optim.replicate(fun = simulate.facilitation(),
+results <- optim.replicate(input.fun = simulate.facilitation,
                            diagnose = var,
                            summarise = optim.summarise,
-                           minimum = 20, maximum = 100,
+                           minimum = 20, maximum = 1000,
                            stop.variance = 0.05,
@@
 ```
+> Note that you can use the parallel version by adding the argument `parallel = TRUE` that is automatised using the package `future` and `future.apply` which makes the whole process run on all your available cores but is way less verbose.
+Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.
+
+## Checking the results
+
+We can then visualise the "convergence" of the results using the `analyse.replicate` function:
+
+```{r}
+## Check the change in variance for all 120 parameters
+analyse.replicate("../Data/Processed/sim_facilitation.rda")
+```

### 2021-05-18 Analysis/03_Summarise_results.Rmd
Results summary TODO

---
title: "Results"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

This script is the repeatable script for measuring changes in the different diversity metrics and how they are affected by the different stressors. 
In this section we are going to analyse the results from the `optim.replicates` chain.

# TODO
 - [ ] summarise.results: get ready to publish results tables
 - [ ] plot.results: get ready to publish plots (including results per methods and time per method)

### 2021-05-18 Analysis/Simulation.pipeline.R

@@ -1,219 +1,0 @@
-## ------------------------------------------------------------------------
-## 'PAPER TITLE'
-## ------------------------------------------------------------------------
-
-# Authors::
-# Thomas Guillerme (guillert@tcd.ie)
-# Stefano Mammola (stefanomammola@gmail.com)
-# Carlos Carmona (perezcarmonacarlos@gmail.com)
-# Pedro Cardoso (pedro.cardoso@helsinki.fi)
-
-## ------------------------------------------------------------------------
-# 'R script to reproduce the analyses'
-## ------------------------------------------------------------------------
-
-# Analysis performed with R (v. R 4.0.3) and R studio (v. 1.4.1103)
-
-## ------------------------------------------------------------------------
-# Loading R packages ------------------------------------------------------
-## ------------------------------------------------------------------------
-
-library("BAT")
-library("devtools")
-# devtools::install_github("TGuillerme/dads")
-# devtools::install_github("TGuillerme/dispRity")
-library("dads")     # needs to be version >= 0.1.4
-## Loading required package: dispRity
-library("dispRity") # needs to be version >= 1.5.7
-library("doSNOW")
-library("FD")
-library("geometry")
-library("hypervolume")
-library("parallel")
-library("tcltk")
-library("TPD")
-source("melodic.rao.R")
-
-
-## Simulation variables
-## Data simulations
-n_traits <- 2
-speciation_rate <- 1
-n_taxa <- 200
-
-## Simulating the data
-simulated_data <- dads::dads(
-  bd.params = list(speciation = speciation_rate),
-  stop.rule = list(max.living = n_taxa),
-  traits    = dads::make.traits(process = dads::BM.process, n = n_traits))
-trait_space <- simulated_data$data[rownames(simulated_data$data) %in% simulated_data$tree$tip.label, ]
-
-
-
-#TODO:
-
-# 1- Check variance > 0.05 or n_runs < 20
-# 2- Simulate data
-# 3- Run random + one other reduction at 0.2, 0.4, 0.6, 0.8
-# 4- Measure metrics (store as lists)
-# 5- Write outputs as csv (or keep in R)
-
-
-
-
-
-
-# Starting the simulation -------------------------------------------------
[…]

### 2021-05-19 Functions/simulation.pipeline.R
Updated BAT functions (following Stefano's fixes)

@@ -228,14 +228,15 @@ fun.kernel <- function(presence, traits, verbose) {
     }
 
     ## Get the hypervolume (remove messages)
-    suppressMessages(silent <- capture.output(hypervolume <- BAT::kernel.alpha(comm = t(presence), trait = traits, return.hv = TRUE)))
+    suppressMessages(silent <- capture.output(hypervolume <- kernel.build(comm = t(presence), trait = traits)))
+    
+    ## Get the values (remove messages)
+    suppressMessages(silent <- capture.output(richness <- BAT::kernel.alpha(comm=hypervolume )))
+    suppressMessages(silent <- capture.output(dispersion <- BAT::kernel.dispersion(comm = hypervolume)))
+    suppressMessages(silent <- capture.output(regularity <- BAT::kernel.evenness(comm = hypervolume)))
 
     ## Get the values
-    suppressMessages(silent <- capture.output(dispersion <- BAT::kernel.dispersion(comm = hypervolume[[2]])))
-    suppressMessages(silent <- capture.output(regularity <- BAT::kernel.evenness(comm = hypervolume[[2]])))
-
-    ## Get the values
-    return(cbind("hypervolume_richness"   = hypervolume[[1]],
+    return(cbind("hypervolume_richness"   = richness,
           "hypervolume_dispersion" = dispersion,
           "hypervolume_regularity" = regularity))
 }
@@ -246,8 +247,11 @@ fun.convex.hull <- function(presence, traits, verbose) {
         message(".", appendLF = FALSE)
     }
 
-    ## run the hulls (remove messages)
-    suppressMessages(silent <- capture.output(results <- BAT::hull.alpha(comm = t(presence), trait = traits)))
+    ## get the hull
+    suppressMessages(silent <- capture.output(hull <- hull.build(comm = t(presence), trait = traits)))
+    
+    # run the volume
+    suppressMessages(silent <- capture.output(results <- BAT::hull.alpha(hull)))
     results <- t(t(results))
     rownames(results) <- colnames(presence)
     colnames(results) <- "convex.hull"

### 2021-07-02 Analysis/02_Simulation_pipeline.Rmd
Update 02_Simulation_pipeline.Rmd

@@
 ```{r}
-## Creating the function for simulating facilitation
-simulate.facilitation <- make.simulation.pipeline(type = "facilitation")
+## Creating the function for simulating competition
+simulate.competition <- make.simulation.pipeline(type = "competition")
 ```
@@
 ## Running one replicate
-simulate.facilitation()
+simulate.competition()
 ```
@@
 
-For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `simulate.facilitation()` for diagnosis) function
+For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `simulate.competition()` for diagnosis) function
 
@@
 
-```{r, eval = FALSE}
+```{r, eval = TRUE}
 ## Running the full optimisation
-results <- optim.replicate(input.fun = simulate.facilitation,
+results <- optim.replicate(input.fun = simulate.competition,
                            diagnose = var,
@@
                            bkp.path = "../Data/Processed/",
-                           bkp.name = "sim_facilitation.rda")
+                           bkp.name = "sim_competition.rda")
 ```
@@
 
-```{r}
+```{r, eval = FALSE}
+source("../Functions/analyse.replicates.R")
 ## Check the change in variance for all 120 parameters
-analyse.replicate("../Data/Processed/sim_facilitation.rda")
+analyse.replicates("../Data/Processed/sim_filtering.rda") #140
+analyse.replicates("../Data/Processed/sim_equalizing.rda") #155
+analyse.replicates("../Data/Processed/sim_facilitation.rda") #126
+analyse.replicates("../Data/Processed/sim_competition.rda")
 ```

### 2021-07-02 Functions/optim.helpers.R
Update optim.helpers.R

@@ -13,7 +13,7 @@
 
 optim.diagnose <- function() {
 
-    ## Combine all the replicates together
+    ## Combine all the replicates together
     #test
     return()
 }

### 2021-07-08 Analysis/02_Simulation_pipeline.Rmd
results up

@@
 analyse.replicates("../Data/Processed/sim_facilitation.rda") #126
-analyse.replicates("../Data/Processed/sim_competition.rda")
+analyse.replicates("../Data/Processed/sim_competition.rda") #163
 ```

### 2021-07-08 Functions/summarise.results.R
Update summarise.results.R

@@ -3,9 +3,10 @@
 #' @description Summarises the results (from the replications)
 #'
 #' @param results the results or the path to the results
-#' @param what optional, which specific results to summarise
 #' @param scale logical, whether to scale the results relative to the the random reduction (TRUE, default) or not (FALSE).
-#' @param time optional, whether to summarise the running time (TRUE) or the results (FALSE, default)
+#' @param cent.tend function, the central tendency function (default is median)
+#' @param prob the probability quantiles to display (default is c(0.975, 0.75, 0.25, 0.025))
+#' @param output.by whether to return the output sorted by stressor ("stressor"; default) or by metric ("metric").
 #' 
 #' @examples
 #'
@@ -14,7 +15,78 @@
 #' @author Thomas Guillerme
 #' @export
 
-summarise.results <- function(results, what, scale = TRUE, time = FALSE) {
+summarise.results <- function(results, scale = TRUE, cent.tend = median, prob = c(0.975, 0.75, 0.25, 0.025), output.by = "stressor") {
 
-    return()
+    ## Make the results into a list (or not)
+    if(all(names(results) %in% c("results_table", "diagnosis", "output_save", "n_iterations"))) {
+        results <- list(results)
+    }
+
+    ## Extracting the results table
+    results_table <- extract.table(results, scale)
+
+    ## Summarise the results
+    return(lapply(results_table, summarise.table, cent.tend, prob))
+}
+
+
+## Centring on the null
+centre.null.scale <- function(results, scale = FALSE) {
+
+    ## Finding the null results
+    null_results <- grep("random", colnames(results$results_table))
+
+    ## Get the pairs of columns
+    metrics_names <- gsub("stressor_", "", gsub("random_", "", colnames(results$results_table)))
+    pairs <- lapply(as.list(unique(metrics_names)), function(x, names) which(names %in% x), names = metrics_names)
+    
+    ## Centre each pair on the null
+    centre_results <- lapply(pairs, function(pair, matrix) return(matrix[, pair[2]] - matrix[, pair[1]]), matrix = results$results_table)
+    output <- do.call(cbind, centre_results)
+
+    ## Scale each centred results
+    if(scale) {
+        output <- apply(output, 2, function(X) return(X/max(abs(X))))
+    }
+    colnames(output) <- unique(metrics_names)
+    return(output)
+}
+
+## Sort the results per name
+sort.results.names <- function(results) {
[…]

### 2021-07-08 Functions/plot.results.R
Plotting function

@@ -2,9 +2,13 @@
 #'
 #' @description Plotting the results
 #'
-#' @param results the results matrix (typically from summarise.results)
-#' @param what optional, a sub-selection of the results to plot
-#' @param ... any arguments to be passed to \code{lines}.
+#' @param results the different results output from 02_Simulation_pipeline.Rmd
+#' @param metric which metric to display?
+#' @param scale whether to scale all the metric between -1 and 1 (default is TRUE)
+#' @param level.names which level names to use (can be left empty)
+#' @param legend whether to add the legend (TRUE, default) or not (FALSE)
+#' @param legend.pos where to plot the legend (default is "topright")
+#' @param ... any arguments to be passed to \code{boxplot}.
 #' 
 #' @examples
 #'
@@ -12,8 +16,76 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
+plot.results <- function(results, metric, scale = TRUE, level.names, legend = TRUE, legend.pos = "topright", ...) {
 
-plot.results <- function(results, what, ...) {
+    ## Make the results into a list (or not)
+    if(!is.null(names(results)) && all(names(results) %in% c("results_table", "diagnosis", "output_save", "n_iterations"))) {
+        results <- list(results)
+    }
 
-    return()
+    ## Extracting the results table
+    results_table <- extract.table(results, scale)
+    names(results_table) <- names(results)
+
+    ## Plot the metric results
+    plot.one.metric(results_table, metric, scale, level.names, legend, legend.pos, ...)
+}
+
+## Plotting one series of results
+plot.one.metric <- function(results_table, metric, scale, level.names, legend, legend.pos, ...) {
+
+    plot_args <- list(...)
+
+    ## Select the correct metric
+    plot_data <- lapply(results_table, function(x, metric) x[, gsub("_rm.", "", colnames(x)) %in% metric], metric = metric)
+
+    ## Trim lengths
+    plot_data <- lapply(plot_data, function(x, min) x[sample(1:nrow(x), min), ], min = min(unlist(lapply(plot_data, nrow))))
+
+    ## Combine plots
+    plot_data <- do.call(cbind, plot_data)
+
+    ## Get the number of levels
+    levels <- unique(gsub(".*_rm", "", colnames(results_table[[1]])))
+
+    ## Adding the level names
+    if(missing(level.names)) {
[…]

### 2021-07-08 Analysis/03_Summarise_results.Rmd
results up
Update 03_Summarise_results.Rmd

@@
 This script is the repeatable script for measuring changes in the different diversity metrics and how they are affected by the different stressors. 
-In this section we are going to analyse the results from the `optim.replicates` chain.
+In this section we are going to analyse the results from the `optim.replicates` chain
 
-# TODO
- - [ ] summarise.results: get ready to publish results tables
- - [ ] plot.results: get ready to publish plots (including results per methods and time per method)
+```{r}
+## Loading the results
+load("../Data/Processed/sim_equalizing.rda")
+equalizing <- bkp
+load("../Data/Processed/sim_facilitation.rda")
+facilitation <- bkp
+load("../Data/Processed/sim_filtering.rda")
+filtering <- bkp
+load("../Data/Processed/sim_competition.rda")
+competition <- bkp
+```
+
+```{r}
+## Loading the functions
+source("../Functions/summarise.results.R")
+source("../Functions/plot.results.R")
+```
+
+# Results per metrics
+
+Here I'm going to display the results in box plot for each metric in each broad category.
+> We can probably think of a better way to display the results in the future.
+
+For each metric displayed here:
+
+ * The different colours correspond to the different stressors (equalizing, facilitation, filtering and competition)
+ * The different density of each colours correspond to the different levels of each stressor (removing 20, 40, 60 or 80% of the data)
+ * All the metrics values are centred on the null (random) stressor for each individual simulation (i.e.
+1 dataset is simulated, X% of the data is removed, one time randomly, one time biased - following the stressor; the metric is calculated on both the random and the biased removal and the biased removal is then centred on the random one (`biased metric score - random metric score`) for each simulation).
+ * All the centred metric values are then scaled to the absolute maximum change (`centred scores/max(abs(centred scores))`).
+
+```{r}
+## Creating the list of results
+all_results <- list(equalizing   = equalizing,
+                    facilitation = facilitation,
+                    filtering    = filtering,
+                    competition  = competition)
+## The different removal level names
+removals <- c("20%", "40%", "60%", "80%")
+## The colours
+cols <- c("blue", "orange", "red", "green")
+```
+
+## Tree metrics
+
+### Alpha
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "alpha", level.names = removals, col = cols)
+```
+
+### Dispersion
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "dispersion", level.names = removals, col = cols)
+```
+
+### Evenness
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "evenness", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+## Dissimilarity metrics
+
+### Rao
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "FD_Rao", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Divergence (FD)
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "FD_divergence", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Evenness (FD)
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "FD_evenness", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Melodic Rao
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "melodic_Rao", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Melodic MPD
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "melodic_mpd", level.names = removals, col = cols)
+```
+
+## Convex hull metrics
+
+### Convex hull alpha
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "convex.hull", level.names = removals, col = cols)
+```
+
+## Hypervolume metrics
+
+"hypervolume_richness"    
+"hypervolume_dispersion"  
+"hypervolume_regularity"
+
+## Probability density metrics
+
+"TPD_Richness"            
+"TPD_Evenness"            
+"TPD_Divergence"

### 2021-07-08 Analysis/03_Summarise_results.Rmd
Update 03_Summarise_results.Rmd

@@
 
-"hypervolume_richness"    
-"hypervolume_dispersion"  
-"hypervolume_regularity"
+### Richness
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "hypervolume_richness", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Dispersion
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "hypervolume_dispersion", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Regularity
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "hypervolume_regularity", level.names = removals, col = cols)
+```
 

### 2021-07-08 Analysis/03_Summarise_results.Rmd
Update 03_Summarise_results.Rmd

@@
 
-"TPD_Richness"            
-"TPD_Evenness"            
-"TPD_Divergence"
+### Richness
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "TPD_Richness", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Evenness
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "TPD_Evenness", level.names = removals, col = cols, legend.pos = "bottomright")
+```
+
+### Divergence
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "TPD_Divergence", level.names = removals, col = cols, legend.pos = "bottomright")
+```

### 2021-07-08 Functions/simulation.pipeline.R
non UTFs fix

@@ -122,7 +122,7 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
         message("Calculating convex hull:", appendLF = FALSE)
     }
 
-    ## Run the convex.hull analyses
+    ## Run the convex.hull analyses
     convex_hull_results <- lapply(presences, fun.convex.hull, traits = trait_space, verbose)
 
     if(record.timer) {
@@ -137,7 +137,7 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
         message("Calculating kernel density hypervolumes:", appendLF = FALSE)
     }
 
-    ## Run the convex.hull analyses
+    ## Run the convex.hull analyses
     hypervolume_results <- lapply(presences, fun.kernel, traits = trait_space, verbose)
     # hypervolume_results <- lapply(presences, fun.tree, tree = tree, verbose) ; warning("DEBUG")
 
@@ -264,7 +264,7 @@ fun.proba.den <- function(presence, traits, verbose) {
         message(".", appendLF = FALSE)
     }
 
-    ## Get the sds matrix
+    ## Get the sds matrix
     standard_deviations <- matrix(
         data = rep(sqrt(diag(ks::Hpi.diag(traits))), nrow(presence)),
         nrow = nrow(presence),

### 2021-07-08 Functions/plot.results.R
Non centre results (untested)

@@ -5,6 +5,7 @@
 #' @param results the different results output from 02_Simulation_pipeline.Rmd
 #' @param metric which metric to display?
 #' @param scale whether to scale all the metric between -1 and 1 (default is TRUE)
+#' @param centre whether to centre all the metrics on the random (null) changes (TRUE; default). If set to FALSE the results are not centred and the null changes are also displayed.
 #' @param level.names which level names to use (can be left empty)
 #' @param legend whether to add the legend (TRUE, default) or not (FALSE)
 #' @param legend.pos where to plot the legend (default is "topright")
@@ -16,7 +17,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-plot.results <- function(results, metric, scale = TRUE, level.names, legend = TRUE, legend.pos = "topright", ...) {
+plot.results <- function(results, metric, scale = TRUE, centre = TRUE, level.names, legend = TRUE, legend.pos = "topright", ...) {
 
     ## Make the results into a list (or not)
     if(!is.null(names(results)) && all(names(results) %in% c("results_table", "diagnosis", "output_save", "n_iterations"))) {
@@ -24,15 +25,15 @@ plot.results <- function(results, metric, scale = TRUE, level.names, legend = TR
     }
 
     ## Extracting the results table
-    results_table <- extract.table(results, scale)
+    results_table <- extract.table(results, scale, centre)
     names(results_table) <- names(results)
 
     ## Plot the metric results
-    plot.one.metric(results_table, metric, scale, level.names, legend, legend.pos, ...)
+    plot.one.metric(results_table, metric, scale, centre, level.names, legend, legend.pos,...)
 }
 
 ## Plotting one series of results
-plot.one.metric <- function(results_table, metric, scale, level.names, legend, legend.pos, ...) {
+plot.one.metric <- function(results_table, metric, scale, centre, level.names, legend, legend.pos, ...) {
 
     plot_args <- list(...)
 

### 2021-07-08 Functions/summarise.results.R

@@ -15,7 +15,7 @@
 #' @author Thomas Guillerme
 #' @export
 
-summarise.results <- function(results, scale = TRUE, cent.tend = median, prob = c(0.975, 0.75, 0.25, 0.025), output.by = "stressor") {
+summarise.results <- function(results, scale = TRUE, centre = TRUE, cent.tend = median, prob = c(0.975, 0.75, 0.25, 0.025), output.by = "stressor") {
 
     ## Make the results into a list (or not)
     if(all(names(results) %in% c("results_table", "diagnosis", "output_save", "n_iterations"))) {
@@ -23,7 +23,7 @@ summarise.results <- function(results, scale = TRUE, cent.tend = median, prob = 
     }
 
     ## Extracting the results table
-    results_table <- extract.table(results, scale)
+    results_table <- extract.table(results, scale, centre)
 
     ## Summarise the results
     return(lapply(results_table, summarise.table, cent.tend, prob))
@@ -31,7 +31,7 @@ summarise.results <- function(results, scale = TRUE, cent.tend = median, prob = 
 
 
 ## Centring on the null
-centre.null.scale <- function(results, scale = FALSE) {
+centre.null.scale <- function(results, scale = FALSE, centre) {
 
     ## Finding the null results
     null_results <- grep("random", colnames(results$results_table))
@@ -41,36 +41,61 @@ centre.null.scale <- function(results, scale = FALSE) {
     pairs <- lapply(as.list(unique(metrics_names)), function(x, names) which(names %in% x), names = metrics_names)
     
     ## Centre each pair on the null
-    centre_results <- lapply(pairs, function(pair, matrix) return(matrix[, pair[2]] - matrix[, pair[1]]), matrix = results$results_table)
-    output <- do.call(cbind, centre_results)
+    if(centre) {
+        centre_results <- lapply(pairs, function(pair, matrix) return(matrix[, pair[2]] - matrix[, pair[1]]), matrix = results$results_table)
+        output <- do.call(cbind, centre_results)
+    } else {
+        output <- results$results_table
+    }
 
     ## Scale each centred results
     if(scale) {
-        output <- apply(output, 2, function(X) return(X/max(abs(X))))
+        if(centre) {
+            output <- apply(output, 2, function(X) return(X/max(abs(X))))
+        } else {
+            stop("DEBUG: TODO: allow scaling while not centring")
+            #TG: use something with pairs
+        }
     }
-    colnames(output) <- unique(metrics_names)
[…]

### 2021-07-08 Functions/optim.helpers.R
Fixed naming bug

@@ -36,8 +36,11 @@ optim.summarise <- function(x) {
     output <- c(unlist(x$results))
 
     ## Names for one result
-    names <- c(paste(rownames(x$results[[1]])[1], colnames(x$results[[1]]), sep = "_"),
-               paste(rownames(x$results[[1]])[2], colnames(x$results[[1]]), sep = "_"))
+    # names <- c(paste(rownames(x$results[[1]])[1], colnames(x$results[[1]]), sep = "_"),
+    #            paste(rownames(x$results[[1]])[2], colnames(x$results[[1]]), sep = "_"))
+    # names <- paste(names, rep(paste0("rm", rep(1:length(x$results))), each = length(names)), sep = "_")
+    names <- c(sapply(colnames(x$results[[1]]), function(metric_names, x)paste(rownames(x$results[[1]]), metric_names, sep = "_"), x))
+
     names <- paste(names, rep(paste0("rm", rep(1:length(x$results))), each = length(names)), sep = "_")
 
     names(output) <- names

### 2021-07-09 Analysis/03_Summarise_results.Rmd
Updated results

--- 2021-07-08 Analysis/03_Summarise_results.Rmd
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "melodic_Rao", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.results(all_results, metric = "melodic_Rao", level.names = removals, col = cols)
 ```
@@
 
-"hypervolume_richness"    
-"hypervolume_dispersion"  
-"hypervolume_regularity"
+### Richness
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "hypervolume_richness", level.names = removals, col = cols)
+```
+
+### Dispersion
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "hypervolume_dispersion", level.names = removals, col = cols)
+```
+
+### Regularity
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "hypervolume_regularity", level.names = removals, col = cols)
+```
 
@@
 
-"TPD_Richness"            
-"TPD_Evenness"            
-"TPD_Divergence"
+### Richness
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "TPD_Richness", level.names = removals, col = cols)
+```
+
+### Evenness
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "TPD_Evenness", level.names = removals, col = cols)
+```
+
+### Divergence
+
+```{r, fig.height = 4, fig.width = 12}
+plot.results(all_results, metric = "TPD_Divergence", level.names = removals, col = cols)
+```
+
+# My interpretation
+
+ * The fact that most metrics pick up a negative change is interesting: a random decrease in number of elements often results in a smaller decrease in metric score than a non-random decrease in number of elements.
+In other words reducing X% of the elements in a biased way often reduces the score of all the metrics compared to a non-biased way.
+This is promising, if someone uses any metric to measure a decrease in something, it's better that it is due to a non-random cause rather than a random one ;).
+ * The fact that different levels of removal (20%, 40%, etc...) results in different distributions of the metrics is also a good check (the level of the removal seems to often impact either the variance or the mean of the metric score - or both).
+ * The fact that we can roughly see differences between the response of the metric and the different stressors is also encouraging (e.g.
+TPD divergence seems to pick up differences in scenarios pretty neatly!).
+ * And finally, one that seems to be clearly working is the FD_evenness metric which seems to pick up changes in facilitation pretty neatly (if not too many species are removed).
+
+<!-- # Sanity checking
+
+This is what is happening in one simulation (e.g. for alpha)
+ -->
+
+```{r, eval = FALSE, echo = FALSE}
+## Get the result table
+results_table <- facilitation$results_table
+
+## Get the alpha metric
+alpha_results <- results_table[, grep("alpha", colnames(results_table))]
+
+## Get the results of the first simulation
+first_sim_alpha <- alpha_results[1,]
+
+## Get the data removal from the first simulation for 40% removal
+first_sim <- facilitation$output_save[[1]]
+random_sim_40 <- first_sim$reductions[[2]][, "random"]
+stress_sim_40 <- first_sim$reductions[[2]][, "stressor"]
+
+par(mfrow = c(1, 2))
+plot(first_sim$trait_space, col = "grey", pch = 19, main = "stressor")
+points(first_sim$trait_space[stress_sim_40, ], col = "orange", pch = 19)
+plot(first_sim$trait_space, col = "grey", pch = 19, main = "random")
+points(first_sim$trait_space[random_sim_40, ], col = "blue", pch = 19)
+
+## Recalculate the metrics
+tree <- hclust(dist(first_sim$trait_space), method = "average")
+BAT::alpha(t(first_sim$reductions[[2]]), tree)
+#          Richness
+# random   65.29787
+# stressor 46.76265
+first_sim_alpha[3:4] # OK
+```

### 2021-07-12 Analysis/02_Simulation_pipeline.Rmd
Update 02_Simulation_pipeline.Rmd

--- 2021-05-18 Analysis/02_Simulation_pipeline.Rmd
@@
 ```{r}
-## Creating the function for simulating facilitation
-simulate.facilitation <- make.simulation.pipeline(type = "facilitation")
+## Creating the function for simulating equalizing
+simulate.equalizing <- make.simulation.pipeline(type = "equalizing")
 ```
@@
 ## Running one replicate
-simulate.facilitation()
+simulate.equalizing()
 ```
@@
 
-For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `simulate.facilitation()` for diagnosis) function
+For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `simulate.equalizing()` for diagnosis) function
 
@@
 
-```{r, eval = FALSE}
+```{r, eval = TRUE}
 ## Running the full optimisation
-results <- optim.replicate(input.fun = simulate.facilitation,
+results <- optim.replicate(input.fun = simulate.equalizing,
                            diagnose = var,
@@
                            bkp.path = "../Data/Processed/",
-                           bkp.name = "sim_facilitation.rda")
+                           bkp.name = "sim_equalizing.rda")
 ```
@@
 
-```{r}
+```{r, eval = FALSE}
+source("../Functions/analyse.replicates.R")
 ## Check the change in variance for all 120 parameters
-analyse.replicate("../Data/Processed/sim_facilitation.rda")
+analyse.replicates("../Data/Processed/sim_filtering.rda") #140
+analyse.replicates("../Data/Processed/sim_equalizing.rda") #155
+analyse.replicates("../Data/Processed/sim_equalizing.rda") #126
+analyse.replicates("../Data/Processed/sim_equalizing.rda") #163
 ```

### 2021-07-22 Analysis/02_Simulation_pipeline.Rmd
Sim up

@@
 ```{r}
-## Creating the function for simulating equalizing
-simulate.equalizing <- make.simulation.pipeline(type = "equalizing")
+## Creating the function for simulating facilitation
+simulate.facilitation <- make.simulation.pipeline(type = "facilitation")
 ```
@@
 ## Running one replicate
-simulate.equalizing()
+simulate.facilitation()
 ```
@@
 
-For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `simulate.equalizing()` for diagnosis) function
+For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `simulate.facilitation()` for diagnosis) function
 
@@
 ## Running the full optimisation
-results <- optim.replicate(input.fun = simulate.equalizing,
+results <- optim.replicate(input.fun = simulate.facilitation,
                            diagnose = var,
                            summarise = optim.summarise,
-                           minimum = 20, maximum = 1000,
-                           stop.variance = 0.05,
+                           minimum = 20, maximum = 170,
+                           stop.variance = 0.0001,
                            verbose = TRUE,
                            bkp.path = "../Data/Processed/",
-                           bkp.name = "sim_equalizing.rda")
+                           bkp.name = "sim_facilitation.rda")
 ```
@@
 ## Check the change in variance for all 120 parameters
-analyse.replicates("../Data/Processed/sim_filtering.rda") #140
+analyse.replicates("../Data/Processed/sim_filtering.rda") #140 -> #170
 analyse.replicates("../Data/Processed/sim_equalizing.rda") #155
-analyse.replicates("../Data/Processed/sim_equalizing.rda") #126
-analyse.replicates("../Data/Processed/sim_equalizing.rda") #163
+analyse.replicates("../Data/Processed/sim_competition.rda") #126
+analyse.replicates("../Data/Processed/sim_facilitation.rda") #163
 ```

### 2021-07-23 Functions/plot.metric.results.R
change to plot.metric.results (plotting a single metric's results)

@@ -17,7 +17,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-plot.results <- function(results, metric, scale = TRUE, centre = TRUE, level.names, legend = TRUE, legend.pos = "topright", ...) {
+plot.metric.results <- function(results, metric, scale = TRUE, centre = TRUE, level.names, legend = TRUE, legend.pos = "topright", ...) {
 
     ## Make the results into a list (or not)
     if(!is.null(names(results)) && all(names(results) %in% c("results_table", "diagnosis", "output_save", "n_iterations"))) {

### 2021-07-23 Analysis/03_Summarise_results.Rmd
added TODO list + changed metric.results list

@@
 source("../Functions/summarise.results.R")
-source("../Functions/plot.results.R")
+source("../Functions/plot.metric.results.R")
 ```
@@
 
+# TODOs:
+
+  1. reworking the scaling 
+    > Carlos: I suspect the standardization you made removes part of the signal, since you are standardizing within each level of each stressor, right?
+The way I see it, this probably "smoothes out" the differences between levels of the same stressor.
+I would use the maximum absolute change of a method for a stressor ACROSS ALL LEVELS as the value to standardize.
+But maybe I am wrong, or maybe I did not understand it correctly
+    `TODO: work on scale function`
+  2. plotting the metrics per stressors (i.e. all metrics for each stressor).
+    `TODO: re-work the plots (15 rows * 4 columns; each row is a metric and contains the lines for the 4 rarefactions; each column is a stressor)`
+  3. fitting a linear model per metric per stressor to rank models based on their r2
+    > Pedro: Wonder if it makes sense to fit linear (or other) models to each trend (20, 40, 60, 80) and the higher the r2 the higher the clearer or more probable the detection of the trend?
+  4. t-test of difference between each random score and the stressor
+    `TODO 3+4: results table per metric`
+
+  5. run a psych-style pairwise comparison between each scaled metric
+  6. put the results in the context of the expectation table from the paper (Null vs process.docx)
+
 ## Tree metrics
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "alpha", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "alpha", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "dispersion", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "dispersion", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "evenness", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "evenness", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "FD_Rao", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "FD_Rao", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "FD_divergence", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "FD_divergence", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "FD_evenness", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "FD_evenness", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "melodic_Rao", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "melodic_Rao", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "melodic_mpd", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "melodic_mpd", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "convex.hull", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "convex.hull", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "hypervolume_richness", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "hypervolume_richness", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "hypervolume_dispersion", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "hypervolume_dispersion", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "hypervolume_regularity", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "hypervolume_regularity", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "TPD_Richness", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "TPD_Richness", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "TPD_Evenness", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "TPD_Evenness", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "TPD_Divergence", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "TPD_Divergence", level.names = removals, col = cols)
 ```

### 2021-07-23 Analysis/03_Summarise_results.Rmd
Added scaling + t.tests

--- 2021-07-09 Analysis/03_Summarise_results.Rmd
@@
 ## Loading the functions
-source("../Functions/summarise.results.R")
-source("../Functions/plot.results.R")
+source("../Functions/extract.table.R")
+source("../Functions/plot.metric.results.R")
+source("../Functions/paired.tests.R")
 ```
@@
 
+# TODOs:
+
+  * [x]1. reworking the scaling
+  * [ ]2. plotting the metrics per stressors (i.e. all metrics for each stressor).
+    `TODO: re-work the plots (15 rows * 4 columns; each row is a metric and contains the lines for the 4 rarefactions; each column is a stressor)`
+  * [ ]3. fitting a linear model per metric per stressor to rank models based on their r2
+    > Pedro: Wonder if it makes sense to fit linear (or other) models to each trend (20, 40, 60, 80) and the higher the r2 the higher the clearer or more probable the detection of the trend?
+  * [x]4. t-test (paired) of difference between each random score and the stressor
+    `TODO 3+4: results table per metric`
+  * [ ]5. run a psych-style pairwise comparison between each scaled metric
+  * [ ]6. put the results in the context of the expectation table from the paper (Null vs process.docx)
+
+## Scaling and centring the results
+
+> TODO: we centre like this
+> TODO: we scale like this
+
+```{r}
+## Extracting all the scaled results
+scaled_results <- extract.table(all_results, scale.method = "between")
+```
+
+## Measuring the statistics for each metric
+
+> TODO: we measure paired difference between random and stressor
+
+```{r}
+## Measuring the paired t-tests
+metric_null_diff <- lapply(all_results, paired.tests)
+```
+
 ## Tree metrics
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "alpha", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "alpha", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "dispersion", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "dispersion", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "evenness", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "evenness", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "FD_Rao", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "FD_Rao", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "FD_divergence", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "FD_divergence", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "FD_evenness", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "FD_evenness", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "melodic_Rao", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "melodic_Rao", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "melodic_mpd", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "melodic_mpd", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "convex.hull", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "convex.hull", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "hypervolume_richness", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "hypervolume_richness", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "hypervolume_dispersion", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "hypervolume_dispersion", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "hypervolume_regularity", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "hypervolume_regularity", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "TPD_Richness", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "TPD_Richness", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "TPD_Evenness", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "TPD_Evenness", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.results(all_results, metric = "TPD_Divergence", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "TPD_Divergence", scale.method = "within", level.names = removals, col = cols)
 ```

### 2021-07-23 Functions/paired.tests.R
Create paired.tests.R

@@ -1,0 +1,31 @@
+#' @title paired.test
+#'
+#' @description paired t-test for a single stressor
+#'
+#' @param one_stressor the results from one stressor
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+paired.tests <- function(one_stressor) {
+    ## Get the results table
+    results_table <- one_stressor$results_table
+    ## Split the results into two lists (random and stressor)
+    random_list <- unlist(apply(results_table[, seq(from = 1, to = ncol(results_table), by = 2)], 2, list), recursive = FALSE)
+    stressor_list <- unlist(apply(results_table[, seq(from = 2, to = ncol(results_table), by = 2)], 2, list), recursive = FALSE)
+    ## Calculate all the differences
+    test_list <- mapply(t.test, random_list, stressor_list, MoreArgs = list(paired = TRUE), SIMPLIFY = FALSE)
+    ## Extract the parameters (d, t and p)
+    extract.params <- function(one_test) {
+        return(c(d = one_test$estimate[[1]],
+                 t = one_test$statistic[[1]],
+                 p = one_test$p.value[[1]]))
+    }
+    test_table <- t(do.call(cbind, lapply(test_list, extract.params)))
+    rownames(test_table) <- gsub("random_", "", rownames(test_table))
+    return(test_table)
+}

### 2021-07-23 Functions/extract.table.R
moved extract.table as its own function

@@ -1,0 +1,97 @@
+#' @title extract.table
+#'
+#' @description extract the scaled results per metric
+#'
+#' @param results the list of results (per stressor)
+#' @param centre whether to centre the data on the null results (TRUE; default)
+#' @param scale.method optional, which scaling method to use. Either "within" for scaling each metric within each stressor (abs(max(metric$stressor))), "between" for scaling each metric between each stressor (abs(max(unlist(metric)))) or left empty for no scaling.
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+## Extract table
+extract.table <- function(results, centre = TRUE, scale.method) {
+
+
+    # stop("DEBUG extract.table")
+    # results <- all_results
+    
+    ## Centre all the results
+    results <- lapply(results, centre.null, centre = centre)
+
+    ## Scale all the metrics between each stressor
+    if(!missing(scale.method)) {
+        if(scale.method == "between") {
+            ## Merging all the results together to the the max per metric
+            metric_max <- apply(do.call(rbind, results), 2, function(x) return(max(abs(x))))
+            #TG: use absolute values?
+            ## Scaling each result by its maximum
+            results <- lapply(results, function(X, max) t(apply(X, 1, function(x, max) return(x/max), max = metric_max)), max = metric_max)
+        }
+        ## Scale all the metrics within each stressor
+        if(scale.method == "within") {
+            ## Merging all the results together to the the max per metric
+             results <- lapply(results, function(one_results) apply(one_results, 2, function(X) return(X/max(abs(X)))))
+        }
+    }
+
+    ## Sort the results per metric
+    results <- lapply(results, sort.results.names, centre = centre)
+
+    return(results)
+}
+
+## Centring on the null
+centre.null <- function(one_results, centre = TRUE) {
+
+    ## Finding the null results
+    null_results <- grep("random", colnames(one_results$results_table))
+
+    ## Get the pairs of columns
+    metrics_names <- gsub("stressor_", "", gsub("random_", "", colnames(one_results$results_table)))
+    pairs <- lapply(as.list(unique(metrics_names)), function(x, names) which(names %in% x), names = metrics_names)
+    
+    ## Centre each pair on the null
+    if(centre) {
[…]

### 2021-07-23 Functions/summarise.results.R

@@ -3,7 +3,8 @@
 #' @description Summarises the results (from the replications)
 #'
 #' @param results the results or the path to the results
-#' @param scale logical, whether to scale the results relative to the the random reduction (TRUE, default) or not (FALSE).
+#' @param scale.method optional, which scaling method to use. Either "within" for scaling each metric within each stressor (abs(max(metric$stressor))), "between" for scaling each metric between each stressor (abs(max(unlist(metric)))) or left empty for no scaling.
+#' @param centre whether to centre the data on the null results (TRUE; default)
 #' @param cent.tend function, the central tendency function (default is median)
 #' @param prob the probability quantiles to display (default is c(0.975, 0.75, 0.25, 0.025))
 #' @param output.by whether to return the output sorted by stressor ("stressor"; default) or by metric ("metric").
@@ -15,7 +16,7 @@
 #' @author Thomas Guillerme
 #' @export
 
-summarise.results <- function(results, scale = TRUE, centre = TRUE, cent.tend = median, prob = c(0.975, 0.75, 0.25, 0.025), output.by = "stressor") {
+summarise.results <- function(results, scale.method = "between", centre = TRUE, cent.tend = median, prob = c(0.975, 0.75, 0.25, 0.025), output.by = "stressor") {
 
     ## Make the results into a list (or not)
     if(all(names(results) %in% c("results_table", "diagnosis", "output_save", "n_iterations"))) {
@@ -23,82 +24,12 @@ summarise.results <- function(results, scale = TRUE, centre = TRUE, cent.tend = 
     }
 
     ## Extracting the results table
-    results_table <- extract.table(results, scale, centre)
+    results_table <- extract.table(results, centre, scale.method)
 
     ## Summarise the results
     return(lapply(results_table, summarise.table, cent.tend, prob))
 }
 
-
-## Centring on the null
-centre.null.scale <- function(results, scale = FALSE, centre) {
-
-    ## Finding the null results
-    null_results <- grep("random", colnames(results$results_table))
-
-    ## Get the pairs of columns
-    metrics_names <- gsub("stressor_", "", gsub("random_", "", colnames(results$results_table)))
-    pairs <- lapply(as.list(unique(metrics_names)), function(x, names) which(names %in% x), names = metrics_names)
-    
-    ## Centre each pair on the null
-    if(centre) {
[…]

### 2021-07-23 Functions/plot.metric.results.R
Update plot.metric.results.R

@@ -4,7 +4,7 @@
 #'
 #' @param results the different results output from 02_Simulation_pipeline.Rmd
 #' @param metric which metric to display?
-#' @param scale whether to scale all the metric between -1 and 1 (default is TRUE)
+#' @param scale.method optional, which scaling method to use. Either "within" for scaling each metric within each stressor (abs(max(metric$stressor))), "between" for scaling each metric between each stressor (abs(max(unlist(metric)))) or left empty for no scaling.
 #' @param centre whether to centre all the metrics on the random (null) changes (TRUE; default). If set to FALSE the results are not centred and the null changes are also displayed.
 #' @param level.names which level names to use (can be left empty)
 #' @param legend whether to add the legend (TRUE, default) or not (FALSE)
@@ -17,7 +17,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-plot.metric.results <- function(results, metric, scale = TRUE, centre = TRUE, level.names, legend = TRUE, legend.pos = "topright", ...) {
+plot.metric.results <- function(results, metric, scale.method = "between", centre = TRUE, level.names, legend = TRUE, legend.pos = "topright", ...) {
 
     ## Make the results into a list (or not)
     if(!is.null(names(results)) && all(names(results) %in% c("results_table", "diagnosis", "output_save", "n_iterations"))) {
@@ -25,15 +25,15 @@ plot.metric.results <- function(results, metric, scale = TRUE, centre = TRUE, le
     }
 
     ## Extracting the results table
-    results_table <- extract.table(results, scale, centre)
+    results_table <- extract.table(results, centre, scale.method = scale.method)
     names(results_table) <- names(results)
 
     ## Plot the metric results
-    plot.one.metric(results_table, metric, scale, centre, level.names, legend, legend.pos,...)
+    plot.one.metric(results_table, metric, scale.method, centre, level.names, legend, legend.pos,...)
 }
 
 ## Plotting one series of results
-plot.one.metric <- function(results_table, metric, scale, centre, level.names, legend, legend.pos, ...) {
+plot.one.metric <- function(results_table, metric, scale.method, centre, level.names, legend, legend.pos, ...) {
 
     plot_args <- list(...)
 
@@ -60,7 +60,7 @@ plot.one.metric <- function(results_table, metric, scale, centre, level.names, l
     ## Graphics handling
     ## y limit
     if(is.null(plot_args$ylim)) {
-        if(scale) {
[…]

### 2021-07-23 Functions/simulation.pipeline.R
updated metrics names for simplification

@@ -196,7 +196,7 @@ fun.tree <- function(comm, tree, verbose) {
             "tree_dispersion" = BAT::dispersion(t(comm), tree),
             "tree_evenness"   = BAT::evenness(t(comm), tree))
         table <- do.call(cbind, table)
-        colnames(table) <- c("alpha", "dispersion", "evenness")
+        colnames(table) <- c("tree_alpha", "tree_dispersion", "tree_evenness")
         return(table)
 }
 

### 2021-07-23 Functions/split.metric.R
Formalised splitting a table

@@ -1,0 +1,36 @@
+#' @title split.metric
+#'
+#' @description splits a table of metrics (typically from extract.table of results$results_table)
+#'
+#' @param table the table of metrics results
+#' @param stressor whether to split by stressor/random (TRUE) or not (FALSE; default)
+#' @param just.names whether to return just the metric names (TRUE) or not (FALSE; default)
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+split.metric <- function(table, stressor = FALSE, just.names = FALSE) {
+
+    ## Get all the metric names
+    metric_names <- unique(gsub("_rm.", "", colnames(table)))
+
+    ## Split by stressor (if required)
+    if(stressor) {
+        metric_names <- unique(gsub("stressor_", "", gsub("random_", "", metric_names)))
+    }
+
+    ## Return just the names?
+    if(just.names) {
+        return(metric_names)
+    }
+
+    ## Get the list of splits
+    splits <- lapply(as.list(metric_names), function(X, table) grep(X, colnames(table)), table = table)
+
+    ## Splitting the data
+    return(lapply(splits, function(split, table) return(table[, split]), table = table))
+}

### 2021-07-23 Functions/split.metric.R
Update split.metric.R

@@ -32,5 +32,7 @@ split.metric <- function(table, stressor = FALSE, just.names = FALSE) {
     splits <- lapply(as.list(metric_names), function(X, table) grep(X, colnames(table)), table = table)
 
     ## Splitting the data
-    return(lapply(splits, function(split, table) return(table[, split]), table = table))
+    output <- lapply(splits, function(split, table) return(table[, split]), table = table)
+    names(output) <- metric_names
+    return(output)
 }

### 2021-07-23 Functions/fit.model.rm.R
model fitting function

@@ -1,0 +1,48 @@
+#' @title fit.model.rm
+#'
+#' @description Fit a model on the different levels of removal per metric
+#'
+#' @param one_stressor the results from one stressor
+#' @param model.fun the type of model function (kind of only works for lm for now)
+#' @param formula the model (default is metric ~ rm for the metric value function of the removal level)
+#' @param summary.out whether to summarise the output (TRUE; default; returns the intercept, slope, adjust R2 and their p.values) or not (FALSE)
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+fit.model.rm <- function(one_stressor, model.fun = lm, formula = metric ~ rm, summary.out = TRUE) {
+    ## Get the data per metric
+    metric_data <- split.metric(one_stressor)
+
+    ## Table the data
+    table.data <- function(X) {
+        n_rows <- nrow(X)
+        return(data.frame("rm" = rep(1:ncol(X), each = n_rows), "metric" = c(X)))
+    }
+    metric_data <- lapply(metric_data, table.data)
+
+    ## Apply the models
+    all_models <- lapply(metric_data, function(data, fun, formula) fun(formula, data), fun = model.fun, formula = formula)
+    
+    if(!summary.out) {
+        return(all_models)
+    } else {
+        return(t(do.call(cbind, lapply(all_models, summarise.models))))
+    }
+
+    # res <- t(do.call(cbind, lapply(all_models, summarise.models)))
+    # for(i in 1:15) {
+    #     plot(metric_data[[i]])
+    #     abline(a = res[i,1], b = res[i,3])
+    #     Sys.sleep(2)
+    # }
+}
+
+## Extract the summary
+summarise.models <- function(one_model){
+    model <- summary(one_model)
+    return(c(model$coefficients[1,c(1,4)], model$coefficients[2,c(1,4)], "adj.R^2" = model$adj.r.squared))
+}

### 2021-07-23 Analysis/03_Summarise_results.Rmd
added column names

@@
 ```{r}
+library(knitr)
 ## Loading the results
@@
 source("../Functions/paired.tests.R")
+source("../Functions/split.metric.R")
+source("../Functions/fit.model.rm.R")
+source("../Functions/fancy.table.R")
 ```
@@
     `TODO: re-work the plots (15 rows * 4 columns; each row is a metric and contains the lines for the 4 rarefactions; each column is a stressor)`
-  * [ ]3. fitting a linear model per metric per stressor to rank models based on their r2
-    > Pedro: Wonder if it makes sense to fit linear (or other) models to each trend (20, 40, 60, 80) and the higher the r2 the higher the clearer or more probable the detection of the trend?
+  * [x]3. fitting a linear model per metric per stressor to rank models based on their r2
   * [x]4. t-test (paired) of difference between each random score and the stressor
@@
 
-> TODO: we measure paired difference between random and stressor
+> TODO: we measure paired difference between random and stressor on the raw data
+> TODO: we fitted a lm on the scaled data (scaled_metric ~ removal_level)
 
 ```{r}
-## Measuring the paired t-tests
-metric_null_diff <- lapply(all_results, paired.tests)
+## Slope and R2 for each model
+model_fits <- lapply(scaled_results, fit.model.rm)
+## Differences between stressor and random
+null_differences <- lapply(all_results, paired.tests)
+
+## Making the fancy tables
+model_table <- fancy.table(model_fits, columns = list(c(3,4), 5))
+null_table <- fancy.table(null_differences, columns = list(c(1,3), c(4, 6), c(7, 9)))
+```
+
+```{r, echo = FALSE}
+kable(model_table, caption = "Results of the model scaled metric ~ removal level per stressor", align = "c")
+```
+
+```{r, echo = FALSE}
+kable(null_table, caption = "Difference between the null metric and the stressed metric (raw) for each level of removal and each stressor", align = "c")
 ```

### 2021-07-23 Functions/fit.model.rm.R

@@ -44,5 +44,8 @@ fit.model.rm <- function(one_stressor, model.fun = lm, formula = metric ~ rm, su
 ## Extract the summary
 summarise.models <- function(one_model){
     model <- summary(one_model)
-    return(c(model$coefficients[1,c(1,4)], model$coefficients[2,c(1,4)], "adj.R^2" = model$adj.r.squared))
+    results <- c(model$coefficients[1,c(1,4)], model$coefficients[2,c(1,4)], "adj.R^2" = model$adj.r.squared)
+    names(results)[c(1,3)] <- c("Intercept", "Slope")
+    return(results)
+
 }

### 2021-07-23 Functions/paired.tests.R
merged results into the correct format

@@ -27,5 +27,13 @@ paired.tests <- function(one_stressor) {
     }
     test_table <- t(do.call(cbind, lapply(test_list, extract.params)))
     rownames(test_table) <- gsub("random_", "", rownames(test_table))
-    return(test_table)
+    
+    ## Splitting by rm levels
+    test_list <- split.metric(t(test_table))
+    merge.results <- function(one_metric) {
+        out <- c(one_metric)
+        names(out) <- c(t(sapply(paste0(rownames(one_metric), "_rm"), function(X) paste0(X, 1:4))))
+        return(out)
+    }
+    return(t(do.call(cbind, lapply(test_list, merge.results))))
 }

### 2021-07-23 Functions/fancy.table.R
fancy table fun

@@ -1,0 +1,66 @@
+#' @title fancy.table
+#'
+#' @description makes a fancy table per metric and stressor
+#'
+#' @param data the list of results per stressor from which to get the columns
+#' @param columns the list of columns or pair of columns. For pairs of columns, the second column should be the p-value and is replaced by a symbol.
+#' @param symbols a vector of up to 5 symbols per p-value levels (see details)
+#' @param digits the number of digits for rounding (default is 3)
+#' 
+#' @details
+#' By default the symbols are c(" ", ".", "*", "**", "***") for 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+fancy.table <- function(data, columns, symbols = c("", ".", "*", "**", "***"), digits = 3) {
+    ## Making the fancy table
+    return(do.call(cbind, lapply(data, get.columns, columns, symbols, digits)))
+}
+
+## Get the fancy columns for one stressor
+get.columns <- function(one_stressor, columns, symbols, digits) {
+    ## Extracting the columns
+    extracted_data <- lapply(columns, function(col, dat) dat[, col, drop = FALSE], dat = one_stressor)
+    ## Fancied data
+    fancy_data <- lapply(extracted_data, function(col, symbols, digits) apply(col, 1, p.symbol, symbols, digits), symbols = symbols, digits = digits)
+    ## Combine that
+    fancy_data <- as.data.frame(do.call(cbind, fancy_data))
+    colnames(fancy_data) <- colnames(one_stressor)[unlist(lapply(columns, function(x) x[1]))]
+    return(fancy_data)
+}
+
+## Get the symbol for each value
+p.symbol <- function(x, symbols, digits) {
+    ## Arbitrary p_value limits
+    limits <- c(0.1, 0.05, 0.01, 0.001)
+
+    if(length(x) == 1) {
+        ## No p-value symbol needed
+        return(round(x, digits))
+    } else {
+        ## Adding the p symbol
+        if(x[2] > limits[1]) {
+            return(paste0(round(x[1], digits), symbols[1]))
+        } else {
+            if(x[2] > limits[2]) {
+                return(paste0(round(x[1], digits), symbols[2]))
+            } else {
+                if(x[2] > limits[3]) {
[…]

### 2021-07-23 Analysis/03_Summarise_results.Rmd
boring results tables

@@
 ```{r, echo = FALSE}
-kable(null_table, caption = "Difference between the null metric and the stressed metric (raw) for each level of removal and each stressor", align = "c")
+## Even fancier column names
+colnames(null_table) <- gsub(".d_rm", ".rm", colnames(null_table))
+colnames(null_table) <- gsub("equalizing", "equa", colnames(null_table))
+colnames(null_table) <- gsub("facilitation", "faci", colnames(null_table))
+colnames(null_table) <- gsub("filtering", "filt", colnames(null_table))
+colnames(null_table) <- gsub("competition", "comp", colnames(null_table))
+kable(null_table, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor", align = "c")
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "alpha", scale.method = "within", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "tree_alpha", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "dispersion", scale.method = "within", level.names = removals, col = cols)
+plot.metric.results(all_results, metric = "tree_dispersion", scale.method = "within", level.names = removals, col = cols)
 ```
@@
 ```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "evenness", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
+plot.metric.results(all_results, metric = "tree_evenness", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
 ```

### 2021-07-27 Analysis/03_Summarise_results.Rmd
Update 03_Summarise_results.Rmd
Added fancy plot!

@@
 source("../Functions/fancy.table.R")
+source("../Functions/fancy.plot.R")
 ```
@@
   * [x]1. reworking the scaling
-  * [ ]2. plotting the metrics per stressors (i.e. all metrics for each stressor).
+  * [x]2. plotting the metrics per stressors (i.e. all metrics for each stressor).
     `TODO: re-work the plots (15 rows * 4 columns; each row is a metric and contains the lines for the 4 rarefactions; each column is a stressor)`
@@
   * [x]4. t-test (paired) of difference between each random score and the stressor
-    `TODO 3+4: results table per metric`
   * [ ]5. run a psych-style pairwise comparison between each scaled metric
@@
 scaled_results <- extract.table(all_results, scale.method = "between")
+scaled_within <- extract.table(all_results, scale.method = "within")
 ```
@@
 ```
+
+# Plots per stressor
+
+```{r, fig.width = 8, fig.height = 12}
+fancy.plot(scaled_results, col = c("blue", "orange"))
+```
+
+Here is the visualisation of the results per stressor and per metric.
+
+# Plots per metric
 

### 2021-07-27 Functions/fancy.plot.R
Fancy plot fun

@@ -1,0 +1,177 @@
+#' @title Fancy plot
+#'
+#' @description Makes the fancy global plots
+#'
+#' @param results all results
+#' @param probs the quantiles to display (default is c(0.025, 0.25, 0.75, 0.975))
+#' @param cent.tend the central tendency function (default is median)
+#' @param col.metrics the colours for each metric
+#' @param lm optional, lm results to add to the plot
+#' @param null optional, paired t-tests to add to the plot
+#' @param ... plot options
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, ...) {
+
+    ## Get the plot parameters
+    plot_params <- list(...)
+    if(is.null(plot_params$pch)) {
+        plot_params$pch = 19
+    }
+    if(is.null(plot_params$lwd)) {
+        plot_params$lwd = 1
+    }
+
+    ## Setting up the main plotting window
+    par(mfrow = c(1, length(results)+1))
+    ## Plotting each stressor
+    for(one_stressor in 0:length(results)) {
+        
+        if(one_stressor == 0) {
+            ## Making the labels plot
+            par(mar = c(5, 10, 4, 0), bty = "n")
+            empty.plot(results[[1]],
+                       plot.main = "",
+                       add.x = FALSE,
+                       add.y = TRUE,
+                       grid  = FALSE)
+
+        } else {
+            par(mar = c(5, 0, 4, 1))
+            ## Making the empty plot
+            empty.plot(results[[one_stressor]],
+                       plot.main = names(results)[one_stressor],
+                       add.x = TRUE,
+                       add.y = FALSE)
+
+            ## Adding the data
+            add.one.stressor(results[[one_stressor]], plot_params = plot_params, probs = probs, cent.tend = cent.tend, col = col.metrics)
+        }
+    }
+}
+
+## Adding one single line
+#plot(NULL, xlim = c(-1, 1), ylim = c(0, 5))
+#one_line <- one_metric[, 1]
+#plot_params <- list(y = 1, lwd = 1, col = "blue", pch = 19)
+#cent.tend = median
+#probs = c(0.025, 0.25, 0.75, 0.975)
[…]

### 2021-07-27 Functions/fancy.plot.R
added lm + null

@@ -50,17 +50,23 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                        add.y = FALSE)
 
             ## Adding the data
-            add.one.stressor(results[[one_stressor]], plot_params = plot_params, probs = probs, cent.tend = cent.tend, col = col.metrics)
+            if(!missing(lm)) {
+                model_data <- lm[[one_stressor]]
+            } else {
+                model_data <- NULL
+            }
+            if(!missing(null)) {
+                null_data <- null[[one_stressor]]
+            } else {
+                null_data <- NULL
+            }
+            ## Plot the data
+            add.one.stressor(results[[one_stressor]], plot_params = plot_params, probs = probs, cent.tend = cent.tend, col = col.metrics, lm = model_data, null = null_data)
         }
     }
 }
 
 ## Adding one single line
-#plot(NULL, xlim = c(-1, 1), ylim = c(0, 5))
-#one_line <- one_metric[, 1]
-#plot_params <- list(y = 1, lwd = 1, col = "blue", pch = 19)
-#cent.tend = median
-#probs = c(0.025, 0.25, 0.75, 0.975)
 add.one.line <- function(one_line, plot_params, probs, cent.tend) {
 
     ## Get the number of quantiles
@@ -87,13 +93,11 @@ add.one.line <- function(one_line, plot_params, probs, cent.tend) {
     ## Adding the median
     add_point <- plot_params
     add_point$x <- cent.tend(one_line)
-    add_point$pch <- 19
     do.call(points, add_point)
 }
 
 ## Adding the results for one metric
-#one_metric <- one_results_sub[, 1:4]
-add.one.metric <- function(one_metric, plot_params, probs, cent.tend, col) {
+add.one.metric <- function(one_metric, plot_params, probs, cent.tend, col, lm_data = NULL, null_data = NULL) {
     
     ## Selecting the levels
     levels <- ncol(one_metric)
@@ -104,18 +108,48 @@ add.one.metric <- function(one_metric, plot_params, probs, cent.tend, col) {
     ## Get the plotting base
     base <- plot_params$y
 
+    ## Add the linear model (if not null)
+    if(!is.null(lm_data)) {
+        ## Select the line significance
+        lty <- ifelse(all(lm_data[c(2,4)] < 0.05), 1, 3)
+        ## Set the coordinates of the lm
+        y = c((1-0.2),(4+0.2))
[…]

### 2021-07-27 Analysis/03_Summarise_results.Rmd
Added fancy plots

@@
 
-```{r, fig.width = 8, fig.height = 12}
+```{r, fig.width = 8, fig.height = 10, echo = FALSE}
 fancy.plot(scaled_results, col = c("blue", "orange"))
 ```
+Figure X1: Simulation results: the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
+Negative and positive values signify a decrease/increase in the metric score.
+The dots represent the median metric value, the full line their 50% confidence interval (CI) and the dashed line their 95% CI.
+The colours are just here to visually separate the metrics rows but the colour gradient within each row corresponds to a removal of respectively 80%, 60%, 40% and 20% of the data (from top to bottom).
 
-Here is the visualisation of the results per stressor and per metric.
+>TG: Here is the visualisation of the results per stressor and per metric.
+
+```{r, fig.width = 8, fig.height = 10, echo = FALSE}
+fancy.plot(scaled_results, col = c("blue", "orange"), lm = model_fits, null = null_differences)
+```
+Figure X2: See above + Grey lines in the background are a fitted linear model on the scaled metric score function of removal amount and the value displayed is the adjusted R^2 from each of these models.
+Dashed grey lines represent non-significant models (slope or/and intercept).
+The grey line plots represent (CI + median) represent distribution of metrics scores not clearly different from the random metric scores (paired t-test p value > 0.05).
+
+>TG: Here is the visualisation of the results per stressor and per metric.
 
@@
  * And finally, one that seems to be clearly working is the FD_evenness metric which seems to pick up changes in facilitation pretty neatly (if not too many species are removed).
-
-<!-- # Sanity checking
-
-This is what is happening in one simulation (e.g. for alpha)
- -->
-
-```{r, eval = FALSE, echo = FALSE}
-## Get the result table
-results_table <- facilitation$results_table
-
-## Get the alpha metric
-alpha_results <- results_table[, grep("alpha", colnames(results_table))]
-
-## Get the results of the first simulation
-first_sim_alpha <- alpha_results[1,]
-
-## Get the data removal from the first simulation for 40% removal
-first_sim <- facilitation$output_save[[1]]
-random_sim_40 <- first_sim$reductions[[2]][, "random"]
-stress_sim_40 <- first_sim$reductions[[2]][, "stressor"]
-
-par(mfrow = c(1, 2))
-plot(first_sim$trait_space, col = "grey", pch = 19, main = "stressor")
-points(first_sim$trait_space[stress_sim_40, ], col = "orange", pch = 19)
-plot(first_sim$trait_space, col = "grey", pch = 19, main = "random")
-points(first_sim$trait_space[random_sim_40, ], col = "blue", pch = 19)
-
-## Recalculate the metrics
-tree <- hclust(dist(first_sim$trait_space), method = "average")
-BAT::alpha(t(first_sim$reductions[[2]]), tree)
-#          Richness
-# random   65.29787
-# stressor 46.76265
-first_sim_alpha[3:4] # OK
-```

### 2021-07-27 Analysis/03_Summarise_results.Rmd
isolated results per metric

@@
 ```
+
 Figure X1: Simulation results: the y axes represent the different metrics tested (sorted by categories).
@@
 ```
+
 Figure X2: See above + Grey lines in the background are a fitted linear model on the scaled metric score function of removal amount and the value displayed is the adjusted R^2 from each of these models.
@@
 >TG: Here is the visualisation of the results per stressor and per metric.
-
-# Plots per metric
-
-## Tree metrics
-
-### Alpha
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "tree_alpha", scale.method = "within", level.names = removals, col = cols)
-```
-
-### Dispersion
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "tree_dispersion", scale.method = "within", level.names = removals, col = cols)
-```
-
-### Evenness
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "tree_evenness", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
-```
-
-## Dissimilarity metrics
-
-### Rao
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "FD_Rao", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
-```
-
-### Divergence (FD)
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "FD_divergence", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
-```
-
-### Evenness (FD)
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "FD_evenness", scale.method = "within", level.names = removals, col = cols, legend.pos = "bottomright")
-```
-
-### Melodic Rao
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "melodic_Rao", scale.method = "within", level.names = removals, col = cols)
-```
-
-### Melodic MPD
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "melodic_mpd", scale.method = "within", level.names = removals, col = cols)
-```
-
-## Convex hull metrics
-
-### Convex hull alpha
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "convex.hull", scale.method = "within", level.names = removals, col = cols)
-```
-
-## Hypervolume metrics
-
-### Richness
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "hypervolume_richness", scale.method = "within", level.names = removals, col = cols)
-```
-
-### Dispersion
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "hypervolume_dispersion", scale.method = "within", level.names = removals, col = cols)
-```
-
-### Regularity
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "hypervolume_regularity", scale.method = "within", level.names = removals, col = cols)
-```
-
-## Probability density metrics
-
-### Richness
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "TPD_Richness", scale.method = "within", level.names = removals, col = cols)
-```
-
-### Evenness
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "TPD_Evenness", scale.method = "within", level.names = removals, col = cols)
-```
-
-### Divergence
-
-```{r, fig.height = 4, fig.width = 12}
-plot.metric.results(all_results, metric = "TPD_Divergence", scale.method = "within", level.names = removals, col = cols)
-```
 

### 2021-07-27 Analysis/04_results_per_metrics.Rmd
isolated results per metric

---
title: "Results per metrics"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

```{r}
library(knitr)
## Loading the results
load("../Data/Processed/sim_equalizing.rda")
equalizing <- bkp
load("../Data/Processed/sim_facilitation.rda")
facilitation <- bkp
load("../Data/Processed/sim_filtering.rda")
filtering <- bkp
load("../Data/Processed/sim_competition.rda")
competition <- bkp
```

```{r}
## Loading the functions
source("../Functions/extract.table.R")
source("../Functions/plot.metric.results.R")
source("../Functions/paired.tests.R")
source("../Functions/split.metric.R")
source("../Functions/fit.model.rm.R")
source("../Functions/fancy.table.R")
source("../Functions/fancy.plot.R")
```

# Results per metrics

```{r}
## Creating the list of results
all_results <- list(equalizing   = equalizing,
                    facilitation = facilitation,
                    filtering    = filtering,
                    competition  = competition)
## The different removal level names
removals <- c("20%", "40%", "60%", "80%")
## The colours
cols <- c("blue", "orange", "red", "green")
```

# Plots per metric

## Tree metrics

### Alpha

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "tree_alpha", scale.method = "between", level.names = removals, col = cols)
```

### Dispersion

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "tree_dispersion", scale.method = "between", level.names = removals, col = cols)
```

### Evenness

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "tree_evenness", scale.method = "between", level.names = removals, col = cols, legend.pos = "bottomright")
```

## Dissimilarity metrics

### Rao

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "FD_Rao", scale.method = "between", level.names = removals, col = cols, legend.pos = "bottomright")
```

### Divergence (FD)

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "FD_divergence", scale.method = "between", level.names = removals, col = cols, legend.pos = "bottomright")
```

### Evenness (FD)

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "FD_evenness", scale.method = "between", level.names = removals, col = cols, legend.pos = "bottomright")
```

### Melodic Rao

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "melodic_Rao", scale.method = "between", level.names = removals, col = cols)
```

### Melodic MPD

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "melodic_mpd", scale.method = "between", level.names = removals, col = cols)
```

## Convex hull metrics

### Convex hull alpha

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "convex.hull", scale.method = "between", level.names = removals, col = cols)
```

## Hypervolume metrics

### Richness

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "hypervolume_richness", scale.method = "between", level.names = removals, col = cols)
```

### Dispersion

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "hypervolume_dispersion", scale.method = "between", level.names = removals, col = cols)
```

### Regularity

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "hypervolume_regularity", scale.method = "between", level.names = removals, col = cols)
```

## Probability density metrics

### Richness

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "TPD_Richness", scale.method = "between", level.names = removals, col = cols)
```

### Evenness

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "TPD_Evenness", scale.method = "between", level.names = removals, col = cols)
```

### Divergence

```{r, fig.height = 4, fig.width = 12, eval = TRUE, echo = FALSE}
plot.metric.results(all_results, metric = "TPD_Divergence", scale.method = "between", level.names = removals, col = cols)
```

```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
```

### 2021-07-27 Functions/fancy.plot.R
Updated colours of background

@@ -218,9 +218,9 @@ empty.plot <- function(one_results, scaled = TRUE, add.x = FALSE, add.y = FALSE,
         ## Add the separators
         if(scaled) {
             ## Vertical 0 line
-            abline(v = 0, col = "grey")
+            abline(v = 0, col = "lightgrey")
         }
         ## Horizontal metric lines
-        abline(h = 0:length(metric_names)*4+0.5, col = "grey", lty = 2)
+        abline(h = 0:length(metric_names)*4+0.5, col = "lightgrey", lty = 2)
     }
 }

### 2021-08-04 Analysis/03_Summarise_results.Rmd
Update 03_Summarise_results.Rmd

@@
 
-```{r}
+```{r, echo = FALSE}
 library(knitr)
@@
 competition <- bkp
-```
 
-```{r}
 ## Loading the functions
@@
 source("../Functions/fancy.plot.R")
+source("../Functions/fancy.names.R")
 ```
@@
 
-```{r}
+```{r, echo = FALSE}
 ## Creating the list of results
@@
 
-# TODOs:
-
-  * [x]1. reworking the scaling
-  * [x]2. plotting the metrics per stressors (i.e. all metrics for each stressor).
-    `TODO: re-work the plots (15 rows * 4 columns; each row is a metric and contains the lines for the 4 rarefactions; each column is a stressor)`
-  * [x]3. fitting a linear model per metric per stressor to rank models based on their r2
-  * [x]4. t-test (paired) of difference between each random score and the stressor
-  * [ ]5. run a psych-style pairwise comparison between each scaled metric
-  * [ ]6. put the results in the context of the expectation table from the paper (Null vs process.docx)
-
 ## Scaling and centring the results
 
-> TODO: we centre like this
-> TODO: we scale like this
+We centre the metric based on the null results by substracting the score of the metric from the randomly reduced trait space to the stressed trait space (i.e.
+$score_{centred} = score_{stressed} - score_{random}$).
+We then scale the centred metric by dividing it by it's maximum absolute centred value between all stressors (i.e. $score_{scaled} = score_{centred} / \text{max}(|scores_{centred}|))$).
 
-```{r}
+```{r, echo = FALSE}
 ## Extracting all the scaled results
@@
 
-> TODO: we measure paired difference between random and stressor on the raw data
-> TODO: we fitted a lm on the scaled data (scaled_metric ~ removal_level)
+To measure the differences between the random and stressed metric scores, we ran paired t-tests between each pairs of unscaled metric score for the random reduction of the trait space and the stressed reduction of the trait space (`t.test(random_scores, stressor_scores, paired = TRUE)`).
 
-```{r}
+To check the trend in the different level of data removals (20, 40, 60 and 80%) using a linear model with the $scores_{scaled}$ as a response to the different removal levels (`lm(scaled_metric ~ removal_level`).
+
+```{r, echo = FALSE}
 ## Slope and R2 for each model
@@
 
->TG: Here is the visualisation of the results per stressor and per metric.
-
 ```{r, fig.width = 8, fig.height = 10, echo = FALSE}
@@
 
->TG: Here is the visualisation of the results per stressor and per metric.
+> I think that Figure X2 would do a good summary of how the metrics pick up changes. Here's a couple of interpretations of it in the light of my previous work on which metric to choose:
+  * Don't use the TPD richness metric, it doesn't do much.
+  * On the other hand, hypervolume regularity and richness are two good metrics because they only pick up really specific changes (respectively "equalizing" and "facilitation") and nothing else.
 
-# My interpretation
+# Results interpretations in the light of our expectation table
 
- * The fact that most metrics pick up a negative change is interesting: a random decrease in number of elements often results in a smaller decrease in metric score than a non-random decrease in number of elements.
-In other words reducing X% of the elements in a biased way often reduces the score of all the metrics compared to a non-biased way.
-This is promising, if someone uses any metric to measure a decrease in something, it's better that it is due to a non-random cause rather than a random one ;).
- * The fact that different levels of removal (20%, 40%, etc...) results in different distributions of the metrics is also a good check (the level of the removal seems to often impact either the variance or the mean of the metric score - or both).
- * The fact that we can roughly see differences between the response of the metric and the different stressors is also encouraging (e.g.
-TPD divergence seems to pick up differences in scenarios pretty neatly!).
- * And finally, one that seems to be clearly working is the FD_evenness metric which seems to pick up changes in facilitation pretty neatly (if not too many species are removed).
+These were our expectations:
+
+Mechanism | Richness| Dispersion | Regularity
+-------------|-------|-------|-------|
+Equalizing   | Lower | Lower | Higher|
+Facilitation | Higher| Higher| Higher|
+Filtering (exclusion) | Higher| Higher| Higher|
+Competition  | Lower | Lower |Nothing|
+
+And these are our results per metric family:
+
+ * [ok] prediction is correct (lower = decrease; higher = increase)
+ * [-ok] prediction is kinda correct (there is a change compared to the null but not in the right direction)
+ * [NO] to a wrong prediction: no change when change is expected or change when no change is expected
+
+> NOTE: the difference between [ok] and [-ok] is actually not really meaningful here since these graphs shows differences compared to the null results so if both the null and stressor results increase but the stressor increases slowlier than the stressor, the results in the table will decrease.
+Although in general, if this is the case, this is a "bad" results since ideally we would want the null to not change at all (i.e.
+if the metric is just picking up changes in number of elements, this is not really useful).
+
+#### Tree (dissimilarity)
+
+Mechanism    | Richness (alpha)| Dispersion | Regularity
+-------------|------------|-------------|-------|
+Equalizing   | [ok] Lower | [ok] Lower  | [-ok] Higher|
+Facilitation | [NO] Higher| [-ok] Higher| [-ok] Higher|
+Exclusion    | [NO] Higher| [-ok] Higher| [-ok] Higher|
+Competition  | [ok] Lower | [ok] Lower  | [NO] Nothing|
+
+#### FD
+
+Mechanism | Richness (Rao)| Dispersion | Regularity
+-------------|-------|-------|-------|
+Equalizing   | [NO] Lower  | [ok] Lower | [-ok] Higher|
+Facilitation | [NO] Higher | [-ok] Higher| [NO] Higher|
+Exclusion    | [-ok] Higher| [-ok] Higher| [-ok] Higher|
+Competition  | [ok] Lower  | [ok] Lower | [NO] Nothing|
+
+#### hypervolume
+
+Mechanism | Richness (Rao)| Dispersion | Regularity
+-------------|-------|-------|-------|
+Equalizing   | [NO] Lower  | [-ok] Lower | [-ok] Higher|
+Facilitation | [-ok] Higher| [NO] Higher| [NO] Higher|
+Exclusion    | [NO] Higher | [NO] Higher| [NO] Higher|
+Competition  | [ok] Lower  | [NO] Lower | [ok] Nothing|
+
+#### TPD
+
+Mechanism | Richness (Rao)| Dispersion | Regularity
+-------------|-------|-------|-------|
+Equalizing   | [NO] Lower  | [ok] Lower | [-ok] Higher|
+Facilitation | [NO] Higher | [-ok] Higher| [-ok] Higher|
+Exclusion    | [NO] Higher | [-ok] Higher| [-ok] Higher|
+Competition  | [NO] Lower  | [ok] Lower | [NO] Nothing|
+
+#### Other
+
+I actually don't remember of the top of my head what the convex.hull and melodic rao/mpd are representing.
+
+# Metrics correlations
+
+And finally below are all the correlations per stressor between each metric per removal levels (20, 40, 60, 80%).
+
+> TODO: do the PCA plots as suggested by Carlos to see things clearlier.
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+require(psych)
+
+## Split the results per stressor and per metric
+## (to extract the different removal levels)
+splitted_results <- lapply(scaled_results, split.metric)
+
+## Get the right removal level
+get.matrix.level <- function(results, level = 1) {
+    do.call(cbind, lapply(results, function(X, level) return(X[, level, drop = FALSE]), level = level))
+}
+## Wrapper for pairs.panels (with fancy names)
+pairs.panels.wrapper <- function(results, ...) {
+    ## Make fancy names
+    colnames(results) <- fancy.names(colnames(results), sep = ":\n")
+    psych::pairs.panels(results, ...)
+}
+
+```
+
+## Equalizing
+
+### Metrics correlation for equalizing with 20% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 1))
+```
+
+### Metrics correlation for equalizing with 40% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 2))
+```
+
+### Metrics correlation for equalizing with 60% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 3))
+```
+
+### Metrics correlation for equalizing with 80% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 4))
+```
+
+## Facilitation
+
+### Metrics correlation for facilitation with 20% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 1))
+```
+
+### Metrics correlation for facilitation with 40% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 2))
+```
+
+### Metrics correlation for facilitation with 60% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 3))
+```
+
+### Metrics correlation for facilitation with 80% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 4))
+```
+
+## Filtering
+
+### Metrics correlation for filtering with 20% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 1))
+```
+
+### Metrics correlation for filtering with 40% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 2))
+```
+
+### Metrics correlation for filtering with 60% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 3))
+```
+
+### Metrics correlation for filtering with 80% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 4))
+```
+
+## Competition
+
+### Metrics correlation for competition with 20% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 1))
+```
+
+### Metrics correlation for competition with 40% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 2))
+```
+
+### Metrics correlation for competition with 60% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 3))
+```
+
+### Metrics correlation for competition with 80% removal
+
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+## Equalizing level 1
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 4))
+```

### 2021-08-04 Functions/extract.table.R
Functions up

@@ -28,7 +28,6 @@ extract.table <- function(results, centre = TRUE, scale.method) {
         if(scale.method == "between") {
             ## Merging all the results together to the the max per metric
             metric_max <- apply(do.call(rbind, results), 2, function(x) return(max(abs(x))))
-            #TG: use absolute values?
             ## Scaling each result by its maximum
             results <- lapply(results, function(X, max) t(apply(X, 1, function(x, max) return(x/max), max = metric_max)), max = metric_max)
         }

### 2021-08-04 Functions/fancy.plot.R

@@ -211,7 +211,7 @@ empty.plot <- function(one_results, scaled = TRUE, add.x = FALSE, add.y = FALSE,
     ## Add the y labels
     if(add.y) {
         n_levels <- ncol(one_results)/length(metric_names)
-        axis(2, at = 1:length(metric_names)*n_levels-median(1:n_levels)+1, labels = metric_names, las = 2)
+        axis(2, at = 1:length(metric_names)*n_levels-median(1:n_levels)+1, labels = fancy.names(metric_names), las = 2)
     }
 
     if(grid) {

### 2021-08-04 Functions/fancy.names.R
Create fancy.names.R

@@ -1,0 +1,30 @@
+#' @title Fancy names
+#'
+#' @description Makes fancy names out of columns of result tables (for plotting)
+#'
+#' @param names the column names
+#' @param sep how to separate the underscores in names. Default is \n
+#' @param remove.levels whether to remove eventual levels (e.g. _rm1). Default is TRUE.
+#' @param nothing do nothing (just returns the non-fancied names). Default is FALSE.
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+fancy.names <- function(names, sep = ":\n", remove.levels = TRUE, nothing = FALSE) {
+
+    ## Do nothing
+    if(nothing) {
+        return(names)
+    }
+
+    ## Remove levels numbers
+    if(remove.levels) {
+        names <- gsub("_rm.", "", names)
+    }
+    ## Replace underscores
+    return(gsub("_", sep, names))
+}

### 2021-08-09 Analysis/03_Summarise_results.Rmd
Comments on last bit of work
(to be removed eventually)

@@
 ```
+
+# Responses
+
+Carlos:
+-TPD richness potential error: 
+  * I have double checked the scaling and everything seems OK (but I might be mistaken!) The centreing logic happens here: https://github.com/TGuillerme/eco_metrics_simulations/blob/c2a8deb9d673e0456f315d9a9f43285aafdfea93/Functions/extract.table.R#L59 and the scaling logic happens here: https://github.com/TGuillerme/eco_metrics_simulations/blob/c2a8deb9d673e0456f315d9a9f43285aafdfea93/Functions/extract.table.R#L30 (l.30 to 32)).
+  So it might be something with the bandwidth?
+The whole calculations for the TPD metrics happens here (https://github.com/TGuillerme/eco_metrics_simulations/blob/c2a8deb9d673e0456f315d9a9f43285aafdfea93/Functions/simulation.pipeline.R#L262) with no obvious bandwidth part (at least to my eyes).
+So maybe it's just using the defaults?
+
+-A minor comment regarding plots:
+ - [ ] TODO: reorganise plots labels (follow Stefano's pdf)
+ - [ ] TODO: reorganise colours by "Richness / Divergence / Regularity"
+ - [ ] TODO: change Convex.hull = Richness; Rao/mpd = Divergence
+
+-Thomas, I would be happy to make the PCAs if you could send me a file with the correlation matrices
+
+ - [ ] TODO: get the scaled results matrices for PCA
+
+Let's talk about the two points on the expectations and the biological realism!
+
+<!-- Carlos comments:
+
+-Sorry to insist on this, but I really disagree with the expectations table as it is, we really would have to discuss what are our expectations (if any!) for any particular mechanism, and what they are based on.
+This is a big problem for me, since it can give the idea that some particular method (or even all methods) is somehow flawed because it doesnt meet some expectation.
+
+-Biological realism: I think the best at this point would be to use some agent-based model as Pedro suggested.
+Sure, things interact, but I think we can make a model simple enough and change some parameters on it, so that we can isolate more or less different things.
+We recently made one for a paper, and I would be happy to adapt the code and run some simulations (after we decide which things we want to play with): https://onlinelibrary.wiley.com/doi/epdf/10.1111/geb.13203
+
+ -->

### 2021-08-16 Analysis/03_Summarise_results.Rmd
TODO

@@
 scaled_within <- extract.table(all_results, scale.method = "within")
+
+## Re-ordering results for fancyness
+new_order <- c(14, 12, 3, 6,
+               15, 11, 2, 8,
+               7, 4, 5, 13,
+               10, 9, 1)
+new_names <- c("TPD[TPD]\nevenness",
+               "hypervolume[BAT]\nevenness",
+               "dendrogram[BAT]\nevenness",
+               "dissimilarity[FD]\nevenness",
+               "TPD[TPD]\ndivergence",
+               "hypervolume[BAT]\ndispersion",
+               "dendrogram[BAT]\ndispersion",
+               "dissimilarity[melodic]\nMPD",
+               "dissimilarity[melodic]\nRao",
+               "dissimilarity[FD]\nRao",
+               "dissimilarity[FD]\ndivergence",
+               "TPD[TPD]\nalpha",
+               "hypervolume[BAT]\nalpha",
+               "convex hull[BAT]\nalpha",
+               "dendrogram[BAT]\nalpha")
+## TODO: add that to fancy plot as options
 ```
@@
 ```{r, fig.width = 8, fig.height = 10, echo = FALSE}
-fancy.plot(scaled_results, col = c("blue", "orange"))
+fancy.plot(scaled_results, col = c("blue", "orange"), lm = model_fits, null = null_differences)
 ```
 
-Figure X1: Simulation results: the y axes represent the different metrics tested (sorted by categories).
+Figure X2: Simulation results: the y axes represent the different metrics tested (sorted by categories).
 The different columns represent the different stressors.
@@
 The colours are just here to visually separate the metrics rows but the colour gradient within each row corresponds to a removal of respectively 80%, 60%, 40% and 20% of the data (from top to bottom).
-
-```{r, fig.width = 8, fig.height = 10, echo = FALSE}
-fancy.plot(scaled_results, col = c("blue", "orange"), lm = model_fits, null = null_differences)
-```
-
-Figure X2: See above + Grey lines in the background are a fitted linear model on the scaled metric score function of removal amount and the value displayed is the adjusted R^2 from each of these models.
+Grey lines in the background are a fitted linear model on the scaled metric score function of removal amount and the value displayed is the adjusted R^2 from each of these models.
 Dashed grey lines represent non-significant models (slope or/and intercept).
 The grey line plots represent (CI + median) represent distribution of metrics scores not clearly different from the random metric scores (paired t-test p value > 0.05).
-
-> I think that Figure X2 would do a good summary of how the metrics pick up changes. Here's a couple of interpretations of it in the light of my previous work on which metric to choose:
-  * Don't use the TPD richness metric, it doesn't do much.
-  * On the other hand, hypervolume regularity and richness are two good metrics because they only pick up really specific changes (respectively "equalizing" and "facilitation") and nothing else.
 

### 2021-08-18 Functions/extract.table.R
added reorder and rename options

@@ -5,6 +5,8 @@
 #' @param results the list of results (per stressor)
 #' @param centre whether to centre the data on the null results (TRUE; default)
 #' @param scale.method optional, which scaling method to use. Either "within" for scaling each metric within each stressor (abs(max(metric$stressor))), "between" for scaling each metric between each stressor (abs(max(unlist(metric)))) or left empty for no scaling.
+#' @param reorder optional, change the order of the metrics (rows)
+#' @param rename optional, change the names of the metrics
 #' 
 #' @examples
 #'
@@ -14,7 +16,7 @@
 #' @export
 
 ## Extract table
-extract.table <- function(results, centre = TRUE, scale.method) {
+extract.table <- function(results, centre = TRUE, scale.method, reorder, rename) {
 
 
     # stop("DEBUG extract.table")
@@ -39,7 +41,7 @@ extract.table <- function(results, centre = TRUE, scale.method) {
     }
 
     ## Sort the results per metric
-    results <- lapply(results, sort.results.names, centre = centre)
+    results <- lapply(results, sort.results.names, centre = centre, reorder, rename)
 
     return(results)
 }
@@ -69,7 +71,7 @@ centre.null <- function(one_results, centre = TRUE) {
 }
 
 ## Sort the results per name
-sort.results.names <- function(results, centre) {
+sort.results.names <- function(results, centre, reorder, rename) {
 
     ## Find the different removal levels
     levels <- unique(gsub(".*_rm", "", colnames(results)))
@@ -82,6 +84,10 @@ sort.results.names <- function(results, centre) {
         metrics <- unique(gsub("stressor_", "", gsub("random_", "", gsub("_rm.*", "", colnames(results)))))     
     }
 
+    if(!missing(reorder)) {
+        metrics <- metrics[reorder]
+    }
+
     ## Sort the metrics names
     sorted_metrics <- unlist(lapply(as.list(metrics), function(x, levels) paste(x, levels, sep = "_rm"), levels = levels))
 
@@ -90,7 +96,21 @@ sort.results.names <- function(results, centre) {
         sorted_metrics <- c(apply(cbind(paste0("random_", sorted_metrics), paste0("stressor_", sorted_metrics)), 1, c))
     }
 
-    ## Return the sorted column names
-    return(results[, match(sorted_metrics, colnames(results))])
+    ## Sort the column names
+    results <- results[, match(sorted_metrics, colnames(results))]
+    if(!missing(rename)) {
[…]

### 2021-08-18 Functions/extract.table.R
Revert "added reorder and rename options"
This reverts commit 700abdfdfd0ae4a11be40a9f465204c0bf03c9eb.

@@ -5,8 +5,6 @@
 #' @param results the list of results (per stressor)
 #' @param centre whether to centre the data on the null results (TRUE; default)
 #' @param scale.method optional, which scaling method to use. Either "within" for scaling each metric within each stressor (abs(max(metric$stressor))), "between" for scaling each metric between each stressor (abs(max(unlist(metric)))) or left empty for no scaling.
-#' @param reorder optional, change the order of the metrics (rows)
-#' @param rename optional, change the names of the metrics
 #' 
 #' @examples
 #'
@@ -16,7 +14,7 @@
 #' @export
 
 ## Extract table
-extract.table <- function(results, centre = TRUE, scale.method, reorder, rename) {
+extract.table <- function(results, centre = TRUE, scale.method) {
 
 
     # stop("DEBUG extract.table")
@@ -41,7 +39,7 @@ extract.table <- function(results, centre = TRUE, scale.method, reorder, rename)
     }
 
     ## Sort the results per metric
-    results <- lapply(results, sort.results.names, centre = centre, reorder, rename)
+    results <- lapply(results, sort.results.names, centre = centre)
 
     return(results)
 }
@@ -71,7 +69,7 @@ centre.null <- function(one_results, centre = TRUE) {
 }
 
 ## Sort the results per name
-sort.results.names <- function(results, centre, reorder, rename) {
+sort.results.names <- function(results, centre) {
 
     ## Find the different removal levels
     levels <- unique(gsub(".*_rm", "", colnames(results)))
@@ -84,10 +82,6 @@ sort.results.names <- function(results, centre, reorder, rename) {
         metrics <- unique(gsub("stressor_", "", gsub("random_", "", gsub("_rm.*", "", colnames(results)))))     
     }
 
-    if(!missing(reorder)) {
-        metrics <- metrics[reorder]
-    }
-
     ## Sort the metrics names
     sorted_metrics <- unlist(lapply(as.list(metrics), function(x, levels) paste(x, levels, sep = "_rm"), levels = levels))
 
@@ -96,21 +90,7 @@ sort.results.names <- function(results, centre, reorder, rename) {
         sorted_metrics <- c(apply(cbind(paste0("random_", sorted_metrics), paste0("stressor_", sorted_metrics)), 1, c))
     }
 
-    ## Sort the column names
-    results <- results[, match(sorted_metrics, colnames(results))]
-    if(!missing(rename)) {
-        ## TODO: optimise this bit (but just for the challenge)
-        original_names <- metrics
[…]

### 2021-08-18 Functions/fancy.plot.R
Update fancy.plot.R

@@ -8,6 +8,7 @@
 #' @param col.metrics the colours for each metric
 #' @param lm optional, lm results to add to the plot
 #' @param null optional, paired t-tests to add to the plot
+#' @param metric.names optional, a vector of fancy names for the results (rows)
 #' @param ... plot options
 #' 
 #' @examples
@@ -16,7 +17,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, ...) {
+fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, ...) {
 
     ## Get the plot parameters
     plot_params <- list(...)
@@ -39,7 +40,8 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                        plot.main = "",
                        add.x = FALSE,
                        add.y = TRUE,
-                       grid  = FALSE)
+                       grid  = FALSE,
+                       metric.names = metric.names)
 
         } else {
             par(mar = c(5, 0, 4, 1))
@@ -186,7 +188,7 @@ add.one.stressor <- function(one_results, plot_params, probs, cent.tend, col, lm
 }
 
 ## Creating an empty plot
-empty.plot <- function(one_results, scaled = TRUE, add.x = FALSE, add.y = FALSE, grid = TRUE, plot.main) {
+empty.plot <- function(one_results, scaled = TRUE, add.x = FALSE, add.y = FALSE, grid = TRUE, plot.main, metric.names = NULL) {
     
     ## Number of metrics
     y_lim <- c(1,ncol(one_results))
@@ -206,12 +208,16 @@ empty.plot <- function(one_results, scaled = TRUE, add.x = FALSE, add.y = FALSE,
     }
 
     ## Get the metric names
-    metric_names <- split.metric(one_results, just.names = TRUE)
+    if(is.null(metric.names)) {
+        metric_names <- fancy.names(split.metric(one_results, just.names = TRUE))
+    } else {
+        metric_names <- metric.names
+    }
 
     ## Add the y labels
     if(add.y) {
         n_levels <- ncol(one_results)/length(metric_names)
[…]

### 2021-08-18 Analysis/03_Summarise_results.Rmd
Updated results figure and table

--- 2021-08-09 Analysis/03_Summarise_results.Rmd
@@
                     competition  = competition)
+
+## Sort the results per name
+sort.metrics <- function(one_results, order) {
+
+    ## Get the results table
+    all_results <- colnames(one_results$results_table)
+
+    ## Extract the metric names
+    metrics <- unique(gsub("stressor_", "", gsub("random_", "", gsub("_rm.*", "", all_results))))
+
+    ## Reorder these names
+    metrics <- metrics[order]
+
+    ## Get the removal levels
+    levels <- unique(gsub(".*_rm", "", all_results))
+
+    ## Sort the metrics_names
+    sorted_metrics <- unlist(lapply(as.list(metrics), function(x, levels) paste(x, levels, sep = "_rm"), levels = levels))
+    ## Adding the random/stressor bits
+    sorted_metrics <- c(apply(cbind(paste0("random_", sorted_metrics), paste0("stressor_", sorted_metrics)), 1, c))
+
+    ## Match the new column order
+    one_results$results_table <- one_results$results_table[, match(sorted_metrics, all_results)]
+
+    ## Return the updated results
+    return(one_results)
+}
+
+## Re-ordering results for fancyness
+new_order <- c(14, 12, 3, 6,
+               15, 11, 2, 8,
+               7, 4, 5, 13,
+               10, 9, 1)
+all_results <- lapply(all_results, sort.metrics, order = new_order)
+
 ## The different removal level names
@@
 cols <- c("blue", "orange", "red", "green")
+
+## Fancier names
+fancy_names <- c("TPD[TPD]\nregularity",
+                 "hypervolume[BAT]\nregularity",
+                 "dendrogram[BAT]\nregularity",
+                 "dissimilarity[FD]\nregularity",
+                 "TPD[TPD]\ndivergence",
+                 "hypervolume[BAT]\ndivergence",
+                 "dendrogram[BAT]\ndivergence",
+                 "dissimilarity[melodicMPD]\ndivergence(MPD)",
+                 "dissimilarity[melodicRao]\ndivergence(Rao)",
+                 "dissimilarity[FD]\ndivergence(Rao)",
+                 "dissimilarity[FD]\ndivergence",
+                 "TPD[TPD]\nrichness",
+                 "hypervolume[BAT]\nrichness",
+                 "convex hull[BAT]\nrichness",
+                 "dendrogram[BAT]\nrichness")
 ```
@@
 ```{r, echo = FALSE}
+rownames(model_table) <- gsub("\n", "_", fancy_names)
 kable(model_table, caption = "Results of the model scaled metric ~ removal level per stressor", align = "c")
@@
 ```{r, echo = FALSE}
+rownames(null_table) <- fancy_names
 ## Even fancier column names
@@
 ```{r, fig.width = 8, fig.height = 10, echo = FALSE}
-fancy.plot(scaled_results, col = c("blue", "orange"))
+col_vector <- c(rep("blue", 4), # The richness ones
+                rep("orange", 7), # The divergence ones
+                rep("darkgreen", 4)) # The regularity ones
+## The plot
+fancy.plot(scaled_results, col = col_vector, lm = model_fits, null = null_differences, metric.names = fancy_names)
 ```
 
-Figure X1: Simulation results: the y axes represent the different metrics tested (sorted by categories).
+Figure 1: Simulation results: the y axes represent the different metrics tested (sorted by categories).
 The different columns represent the different stressors.

### 2021-08-18 Analysis/03_Summarise_results.Rmd
updated fancy names

@@
 ```{r, echo = FALSE}
-rownames(model_table) <- gsub("\n", "_", fancy_names)
+rownames(model_table) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
 kable(model_table, caption = "Results of the model scaled metric ~ removal level per stressor", align = "c")
@@
 ```{r, echo = FALSE}
-rownames(null_table) <- fancy_names
+rownames(null_table) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
 ## Even fancier column names
@@
 ## Wrapper for pairs.panels (with fancy names)
-pairs.panels.wrapper <- function(results, ...) {
+pairs.panels.wrapper <- function(results, metric.names, ...) {
     ## Make fancy names
+    if(missing(metric.names)) {
         colnames(results) <- fancy.names(colnames(results), sep = ":\n")
+    } else {
+        colnames(results) <- metric.names
+    }
     psych::pairs.panels(results, ...)
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 1))
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 1), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 2))
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 2), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 3))
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 3), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 4))
+pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 4), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 1))
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 1), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 2))
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 2), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 3))
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 3), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 4))
+pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 4), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 1))
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 1), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 2))
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 2), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 3))
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 3), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 4))
+pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 4), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 1))
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 1), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 2))
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 2), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 3))
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 3), metric.names = fancy_names)
 ```
@@
 ## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 4))
+pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 4), metric.names = fancy_names)
 ```

### 2022-10-06 Analysis/01_Simulating_spaces_and_stressors.Rmd
Updated pipeline

@@
 ```{r}
+set.seed(1)
 library(dads)     # needs to be version >= 0.1.4
@@
 ```{r, fig.height = 12, fig.width = 6}
-
 ## Visualising the results
@@
 ```
+
+## Paper figure
+
+```{r}
+plot.text.box <- function(my_text, ...) {
+  plot(1,1, xaxt = "n", yaxt = "n", xlab = "", ylab = "", col = "white", pch = 21)
+  text(c(1,1), my_text, ...)
+}
+
+pdf(file = "../Manuscript/Figures/simulation_protocol.pdf", height = 16, width = 8)
+par(mfrow = c(6, 3), bty = "n")
+
+## First line
+plot.text.box("Tree and trait\nsimulation:", cex = 2)
+plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (1D)", col = c("lightgrey", "blue"), ylab = "Trait 1")
+plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (2D)", col = c("lightgrey", "blue"), trait = c(1,2), xlab = "Trait 1", ylab = "Trait 2")
+
+## Stressors
+plot.text.box("Resampling\nwith stressor:\nnull mechanism", cex = 2)
+plot.1D(trait_space, reduction = random_reduction, null = rep(TRUE, 200), main = "1D null mechanism", legend = c("Full distribution (200sp)", "Resample (100sp)"))
+plot.2D(trait_space, reduction = random_reduction, null = !random_reduction, main = "2D null mechanism")
+
+plot.text.box("Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
+plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D filtering mechanism")
+
+plot.text.box("Reseampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
+plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")
+
+plot.text.box("Reseampling\nwith stressor:\nexcluding\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D excluding mechanism")
+plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D excluding mechanism")
+
+plot.text.box("Reseampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
+plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D facilitation mechanism")
+dev.off()
+```

### 2022-10-06 Analysis/03_Summarise_results.Rmd
Updated pipeline

@@
 fancy.plot(scaled_results, col = col_vector, lm = model_fits, null = null_differences, metric.names = fancy_names)
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor.pdf", height = 9, width = 9)
+fancy.plot(scaled_results, col = col_vector, lm = model_fits, null = null_differences, metric.names = fancy_names)
+dev.off()
+
 ```
@@
 
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = TRUE}
+```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
 ## Equalizing level 1

### 2022-10-06 Functions/plot.fun.R
Updated fun

@@ -24,10 +24,10 @@ plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch 
                        alpha = alpha * 255)
       return(do.call(grDevices::rgb, rgb_args))
     }
-    ## Plot the histograms
+    ## Plot the histograms
     plot(plot_1, col = make.transparent(col[1]), ylim = c(0, plot_height),
         main = ifelse(missing(main), "", main),
-        xlab = "Trait values", ylab = "Counts")
+        xlab = "Trait 1", ylab = "Counts")
     plot(plot_2, col = make.transparent(col[2]), add = TRUE)
 
     ## Adding the legend
@@ -38,9 +38,9 @@ plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch 
     points(x = data[null, trait], y = rep(0, sum(null)), col = "black", bg = col[1], pch = 21)
     points(x = data[reduction, trait], y = rep(-plot_height/50, sum(reduction)), col = "black", bg = col[2], pch = 21)
 }
-plot.2D <- function(data, reduction, null, main, col = c("blue", "orange"), pch= 19, trait = c(1,2), xlab = "trait values", legend) {
+plot.2D <- function(data, reduction, null, main, col = c("blue", "orange"), pch= 19, trait = c(1,2), legend, ...) {
     ## Plotting the histogram
-    plot(NULL, xlim = range(trait_space[, trait[1]]), ylim = range(trait_space[, trait[2]]), main = main)
+    plot(NULL, xlim = range(trait_space[, trait[1]]), ylim = range(trait_space[, trait[2]]), main = main, xlab = "Trait 1", ylab = "Trait 2")
     points(trait_space[null, trait], pch = pch, col = col[1])
     points(trait_space[reduction, trait], pch = pch, col = col[2])
     ## Adding the legend

### 2023-11-21 Analysis/01_Simulating_spaces_and_stressors.Rmd
updated scripts

--- 2021-05-14 Analysis/01_Simulating_spaces_and_stressors.Rmd
@@
 library(devtools)
-devtools::install_github("TGuillerme/dads")
+devtools::install_github("TGuillerme/treats")
 devtools::install_github("TGuillerme/dispRity")
@@
 ```{r}
-library(dads)     # needs to be version >= 0.1.4
+set.seed(1)
+library(treats)     # needs to be version >= 0.1.4
 library(dispRity) # needs to be version >= 1.5.7
@@
 ## Simulating the data
-simulated_data <- dads(bd.params = bd_params,
+simulated_data <- treats(bd.params = bd_params,
                          stop.rule = stop_rule,
@@
 
+## Equalising fitness processes
+
+For simulating equalisation of fitness we can use the `"size"` algorithm from `dispRity::reduce.space`.
+This removes elements around past a distance from the centre of the population (with the distance being determined automatically so that X% are inside the radius and 1-X% are outside):
+
+```{r}
+## Removing 50% to change the trait space size
+equalizing_fitness <- reduce.space(trait_space,
+                                    type = "size",
+                                    remove = 0.5)
+```
+
+```{r, fig.height = 12, fig.width = 6}
+## Visualising the results
+par(mfrow = c(2, 1))
+plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
+plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")
+```
+
 ## Competitive exclusion
@@
 
-## Equalising fitness processes
-
-For simulating equalisation of fitness we can use the `"size"` algorithm from `dispRity::reduce.space`.
-This removes elements around past a distance from the centre of the population (with the distance being determined automatically so that X% are inside the radius and 1-X% are outside):
-
-```{r}
-## Removing 50% to change the trait space size
-equalizing_fitness <- reduce.space(trait_space,
-                                    type = "size",
-                                    remove = 0.5)
-```
-
-```{r, fig.height = 12, fig.width = 6}
-
-## Visualising the results
-par(mfrow = c(2, 1))
-plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
-plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")
-```
-
 ## Facilitation
 
-This one might be a bit trickier (though doable).
-One way to do it (as suggested in our last meeting) could be to simply add a number of data points to places that already have data. 
-However, this will result in non-comparable number of elements in each simulations.
-
-One alternative (at least for now) would be to use the `"density"` algorithm from `dispRity::reduce.space` that will reduce the nearest neighbours distance between elements (basically making clumps of two elements close to each other).
+Using the `"density"` algorithm from `dispRity::reduce.space` that to reduce the nearest neighbours distance between elements (basically making clumps of two elements close to each other).
 
@@
 ```
+
+## Paper figure
+
+```{r}
+plot.text.box <- function(my_text, ...) {
+  plot(1,1, xaxt = "n", yaxt = "n", xlab = "", ylab = "", col = "white", pch = 21)
+  text(c(1,1), my_text, ...)
+}
+
+pdf(file = "../Manuscript/Figures/simulation_protocol.pdf", height = 16, width = 8)
+par(mfrow = c(6, 3), bty = "n")
+
+## First line
+plot.text.box("Tree and trait\nsimulation:", cex = 2)
+plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (1D)", col = c("lightgrey", "blue"), ylab = "Trait 1")
+plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (2D)", col = c("lightgrey", "blue"), trait = c(1,2), xlab = "Trait 1", ylab = "Trait 2")
+
+## Stressors
+plot.text.box("1) Resampling\nwith stressor:\nnull mechanism", cex = 2)
+plot.1D(trait_space, reduction = random_reduction, null = rep(TRUE, 200), main = "1D null mechanism", legend = c("Full distribution (200sp)", "Resample (100sp)"))
+plot.2D(trait_space, reduction = random_reduction, null = !random_reduction, main = "2D null mechanism")
+
+plot.text.box("2) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism", legend = c("Random resampling (100sp)", "Biased resampling (100sp)"))
+plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D filtering mechanism")
+
+plot.text.box("12) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
+plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")
+
+plot.text.box("4) Resampling\nwith stressor:\nexcluding\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D excluding mechanism")
+plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D excluding mechanism")
+
+plot.text.box("5) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
+plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D facilitation mechanism")
+dev.off()
+```

### 2023-11-21 Analysis/02_Simulation_pipeline.Rmd
updated scripts

@@
 library(devtools)
-devtools::install_github("TGuillerme/dads")
+devtools::install_github("TGuillerme/treats")
+install.packages(c("dispRity", "BAT", "FD", "ks", "TPD"))
 ```
@@
 ## Libraries
-require(dads)
+require(treats)
 require(dispRity)

### 2024-01-18 Analysis/03-Empirical_pipeline.rmd
Create 03-Empirical_pipeline.rmd

---
title: "Empirical pipeline"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

This script is the repeatable script for measuring changes in the different diversity metrics and how they are affected by the empirical data.

## Requirements for the simulations

In order to run the simulations (on your machine, cluster, etc...) you'll need to load the following in your current environment:

```{r, eval = FALSE}
## Getting the non-CRAN (yet) libraries
library(devtools)
devtools::install_github("TGuillerme/treats")
install.packages(c("dispRity", "BAT", "FD", "ks", "TPD"))
```

```{r}
## Libraries
require(treats)
require(dispRity)
require(BAT)
require(FD)
require(ks)
require(TPD)
```

> Note that the libraries do not need to be loaded (i.e. no need for `library()`) but needs to exist.

You'll also need the following custom functions (located in [here on github](https://github.com/TGuillerme/eco_metrics_simulations/tree/master/Functions))

```{r}
## Loading the functions
source("../Functions/melodic.rao.R")
source("../Functions/make.simulation.pipeline.R")
source("../Functions/simulation.pipeline.R")
```

## Setting up the empirical data

You can run the pipeline on the empirical data if you have the following elements ready:

 * A trait space (a `"matrix"`) with the rows as species and the columns as dimensions (make sure it has rownames for avoiding any problems down the line).
 * A list of species to remove for each sub-sampling the list should be list of named `logical` vectors (where for example species that should be excluded - e.g. fossils are marked as `FALSE` and the ones to include as `TRUE`).

Here are two dummy examples but you can replace them with your empirical data:

```{r}
## Making a dummy trait space for 200 species (3D)
my_traitspace <- dispRity::space.maker(elements = 200, dimensions = 3,
                                       distribution = rnorm,
                                       elements.names = 1:200)
## This is what it looks like
head(my_traitspace)

## Making a dummy set of sub-sampling
my_subsamples <- list(sample(c(TRUE, FALSE), 200, replace = TRUE),
                      sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.75, 0.25)),
                      sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.90, 0.10)))
## Adding the elements names
my_subsamples <- lapply(my_subsamples, function(x) {names(x) <- 1:200 ; return(x)})

## This is what it looks like
str(my_subsamples)
```

You can then use the function `make.simulation.pipeline` to set up the pipeline by specifying the input data and stressor (the list of species to remove):

```{r}
## Creating the function for running the empirical pipeline
empirical.pipeline <- make.simulation.pipeline(sim.data = my_traitspace,
                                               remove   = my_subsamples)
```

And then you can just run the pipeline by calling this pipeline function:

```{r, eval = FALSE}
## Running the empirical pipeline
empirical.pipeline()
```

## Replicating the empirical pipeline

The pipeline also automatically generates a null model for each level of data removal.
To get enough variance on the null model we will replicate the pipeline $n$ times until adding 5% more replicates does not increase the overall variance by more than 5% for any metric.
In other words, we replicated each simulation $n$ times where $n$ is estimated so that $var(n + 0.05 \times n) < 0.05 \times var(n)$.

For that we will use the custom `optim.replicate()` function which comes with one diagnosis function (checking if the variance for each metric is < 0.05) and one summarising (summarising the output of `empirical.pipeline()` for diagnosis) function

```{r}
## Loading the optimisation functions
source("../Functions/optim.replicate.R")
source("../Functions/optim.helpers.R")
source("../Functions/analyse.replicates.R")
```

You can then run this replicate function that will start by running 20 replicates (`mimimum`) and will increase the number of replicates by 5% until reaching 170 replicates (`maximum`) or until the addition 5% increase does not increase the variance for each metric with each different level of removal by less than 5% (120 parameters!).

There are options to

```{r, eval = TRUE}
## Running the full optimisation
results <- optim.replicate(input.fun = empirical.pipeline,
                           diagnose = var,
                           summarise = optim.summarise,
                           minimum = 20, maximum = 170,
                           stop.variance = 0.0001,
                           verbose = TRUE,
                           bkp.path = "../Data/Processed/",
                           bkp.name = "empirical_data.rda")
```

> Note that you can use the parallel version by adding the argument `parallel = TRUE` that is automatised using the package `future` and `future.apply` which makes the whole process run on all your available cores but is way less verbose. Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.

## Checking the results

We can then visualise the "convergence" of the results using the `analyse.replicate` function:

```{r, eval = FALSE}
source("../Functions/analyse.replicates.R")
## Check the change in variance for all 120 parameters
analyse.replicates("../Data/Processed/sim_filtering.rda") #140 -> #170
analyse.replicates("../Data/Processed/sim_equalizing.rda") #155
analyse.replicates("../Data/Processed/sim_competition.rda") #126
analyse.replicates("../Data/Processed/sim_facilitation.rda") #163
analyse.replicates("../Data/Processed/empirical_data.rda") #163
```

### 2024-01-18 Analysis/04_Summarise_results.Rmd
renumbered markdowns

--- 2022-10-06 Analysis/03_Summarise_results.Rmd
@@
 Dashed grey lines represent non-significant models (slope or/and intercept).
-The grey line plots represent (CI + median) represent distribution of metrics scores not clearly different from the random metric scores (paired t-test p value > 0.05).
+The grey line plots represent distribution of metrics scores not clearly different from the random metric scores (paired t-test p value > 0.05).
 

### 2024-01-18 Functions/make.simulation.pipeline.R
Updated functions for the pipeline

@@ -14,7 +14,7 @@
 #' @author Thomas Guillerme
 #' @export
 
-make.simulation.pipeline <- function(type,
+make.simulation.pipeline <- function(type = NULL,
                                      sim.data = list(n.traits   = 2,
                                                      speciation = 1,
                                                      n.taxa     = 200),

### 2024-01-18 Functions/simulation.pipeline.R

@@ -2,9 +2,9 @@
 #'
 #' @description One simulation pipeline
 #'
-#' @param sim.data A named list of parameters for the data simulation (default is \code{list{n.traits = 2, speciation = 1, n.taxa = 200}})
+#' @param sim.data A named list of parameters for the data simulation (default is \code{list{n.traits = 2, speciation = 1, n.taxa = 200}}) or a traitspace
 #' @param type Which removal type (either \code{"facilitation"}, \code{"equalizing"}, \code{"competition"} or \code{"filtering"})}) to be paired with \code{"random"}.
-#' @param remove How much data to remove (can be a vector of multiple percentages)
+#' @param remove How much data to remove (can be a vector of multiple percentages) or a list of elements to remove.
 #' @param verbose whether to be verbose (\code{TRUE}, default) or not (\code{FALSE})
 #' @param record.timer whether to record the timer for each step of the simulations
 #' 
@@ -18,17 +18,23 @@
 # sim.data = list(n.traits = 2, speciation = 1, n.taxa = 200)
 # type = "facilitation"
 # remove = c(0.2,.4, 0.6 ,0.8)
-# library(dads)
+# library(treats)
 # library(dispRity)
 # library(BAT)
 # source("melodic.rao.R")
 
 
 # test <- simulation.pipeline(sim.data = sim.data, type = "facilitation", remove = remove, verbose = TRUE, record.timer = TRUE)
-
-## SOLVED: Errors: Error in hclust(as.dist(thisTree)) : must have n >= 2 objects to cluster
 # set.seed(4)
 # test <- simulation.pipeline(sim.data, type = "facilitation", remove = remove, verbose = TRUE)
+
+
+## Random 3D dataset with 200 taxa
+# sim.data <- dispRity::space.maker(elements = 200, dimensions = 3, distribution = rnorm)
+# rownames(sim.data) <- 1:200 
+# remove <- list(sample(c(TRUE, FALSE), 200, replace = TRUE),
+#                sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.75, 0.25)),
+#                sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.90, 0.10)))
 simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.timer = FALSE) {
 
     if(record.timer) {
@@ -39,53 +45,90 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
[…]

### 2024-02-29 Analysis/04_Summarise_results.Rmd
updated MS with simulated 2d/3d data
Needs to finish fancy plot for empirical data

@@
 ## Loading the results
-load("../Data/Processed/sim_equalizing.rda")
-equalizing <- bkp
-load("../Data/Processed/sim_facilitation.rda")
-facilitation <- bkp
-load("../Data/Processed/sim_filtering.rda")
-filtering <- bkp
-load("../Data/Processed/sim_competition.rda")
-competition <- bkp
+load("../Data/Processed/sim_equalizing_2d.rda")
+equalizing_2d <- bkp
+load("../Data/Processed/sim_facilitation_2d.rda")
+facilitation_2d <- bkp
+load("../Data/Processed/sim_filtering_2d.rda")
+filtering_2d <- bkp
+load("../Data/Processed/sim_competition_2d.rda")
+competition_2d <- bkp
+
+## Loading the results
+load("../Data/Processed/sim_equalizing_3d.rda")
+equalizing_3d <- bkp
+load("../Data/Processed/sim_facilitation_3d.rda")
+facilitation_3d <- bkp
+load("../Data/Processed/sim_filtering_3d.rda")
+filtering_3d <- bkp
+load("../Data/Processed/sim_competition_3d.rda")
+competition_3d <- bkp
+
+## Loading empirical results
+load("../Data/Processed/empirical_data-hist-present-native.rda")
+empirical_hist <- bkp
+load("../Data/Processed/empirical_data-prehist-present-native.rda")
+empirical_prehist <- bkp
+
+hist_table <- empirical_hist$results_table
+colnames(hist_table) <- gsub("_rm1", "_rm2", colnames(hist_table))
+combined_empirical <- list()
+combined_empirical$results_table <- cbind(empirical_prehist$results_table, hist_table)
 
@@
 ## Creating the list of results
-all_results <- list(equalizing   = equalizing,
-                    facilitation = facilitation,
-                    filtering    = filtering,
-                    competition  = competition)
+all_results_2d <- list(equalizing   = equalizing_2d,
+                       facilitation = facilitation_2d,
+                       filtering    = filtering_2d,
+                       competition  = competition_2d)
+all_results_3d <- list(equalizing   = equalizing_3d,
+                       facilitation = facilitation_3d,
+                       filtering    = filtering_3d,
+                       competition  = competition_3d)
+empirical_results <- list(combined_empirical)
 
@@
                10, 9, 1)
-all_results <- lapply(all_results, sort.metrics, order = new_order)
+all_results_2d <- lapply(all_results_2d, sort.metrics, order = new_order)
+all_results_3d <- lapply(all_results_3d, sort.metrics, order = new_order)
+empirical_results <- lapply(empirical_results, sort.metrics, order = new_order)
 
@@
 ## Extracting all the scaled results
-scaled_results <- extract.table(all_results, scale.method = "between")
-scaled_within <- extract.table(all_results, scale.method = "within")
+scaled_results_2d <- extract.table(all_results_2d, scale.method = "between")
+scaled_within_2d  <- extract.table(all_results_2d, scale.method = "within")
+scaled_results_3d <- extract.table(all_results_3d, scale.method = "between")
+scaled_within_3d  <- extract.table(all_results_3d, scale.method = "within")
+scaled_results_emp<- extract.table(empirical_results, scale.method = "between")
+scaled_within_emp <- extract.table(empirical_results, scale.method = "within")
 ```
@@
 ## Slope and R2 for each model
-model_fits <- lapply(scaled_results, fit.model.rm)
+model_fits_2d  <- lapply(scaled_results_2d, fit.model.rm)
+model_fits_3d  <- lapply(scaled_results_3d, fit.model.rm)
+model_fits_emp <- lapply(scaled_results_emp, fit.model.rm)
+
 ## Differences between stressor and random
-null_differences <- lapply(all_results, paired.tests)
+null_differences_2d  <- lapply(all_results_2d, paired.tests)
+null_differences_3d  <- lapply(all_results_3d, paired.tests)
+null_differences_emp <- lapply(empirical_results, paired.tests)
 
 ## Making the fancy tables
-model_table <- fancy.table(model_fits, columns = list(c(3,4), 5))
-null_table <- fancy.table(null_differences, columns = list(c(1,3), c(4, 6), c(7, 9)))
+model_table_2d  <- fancy.table(model_fits_2d, columns = list(c(3,4), 5))
+model_table_3d  <- fancy.table(model_fits_3d, columns = list(c(3,4), 5))
+model_table_emp <- fancy.table(model_fits_emp, columns = list(c(3,4), 5))
+null_table_2d   <- fancy.table(null_differences_2d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
+null_table_3d   <- fancy.table(null_differences_3d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
+null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), c(4, 6)))
 ```
@@
 ```{r, echo = FALSE}
-rownames(model_table) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-kable(model_table, caption = "Results of the model scaled metric ~ removal level per stressor", align = "c")
+rownames(model_table_2d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
+kable(model_table_2d, caption = "Results of the model scaled metric ~ removal level per stressor (2D)", align = "c")
+```
+```{r, echo = FALSE}
+rownames(model_table_3d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
+kable(model_table_3d, caption = "Results of the model scaled metric ~ removal level per stressor (3D)", align = "c")
+```
+```{r, echo = FALSE}
+rownames(model_table_emp) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
+kable(model_table_emp, caption = "Results of the model scaled metric ~ removal level for the empirical data (3D)", align = "c")
+```
+```{r, echo = FALSE}
+rownames(null_table_2d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
+## Even fancier column names
+colnames(null_table_2d) <- gsub(".d_rm", ".rm", colnames(null_table_2d))
+colnames(null_table_2d) <- gsub("equalizing", "equa", colnames(null_table_2d))
+colnames(null_table_2d) <- gsub("facilitation", "faci", colnames(null_table_2d))
+colnames(null_table_2d) <- gsub("filtering", "filt", colnames(null_table_2d))
+colnames(null_table_2d) <- gsub("competition", "comp", colnames(null_table_2d))
+kable(null_table_2d, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2d)", align = "c")
 ```
@@
 ```{r, echo = FALSE}
-rownames(null_table) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
+rownames(null_table_3d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
 ## Even fancier column names
-colnames(null_table) <- gsub(".d_rm", ".rm", colnames(null_table))
-colnames(null_table) <- gsub("equalizing", "equa", colnames(null_table))
-colnames(null_table) <- gsub("facilitation", "faci", colnames(null_table))
-colnames(null_table) <- gsub("filtering", "filt", colnames(null_table))
-colnames(null_table) <- gsub("competition", "comp", colnames(null_table))
-kable(null_table, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor", align = "c")
+colnames(null_table_3d) <- gsub(".d_rm", ".rm", colnames(null_table_3d))
+colnames(null_table_3d) <- gsub("equalizing", "equa", colnames(null_table_3d))
+colnames(null_table_3d) <- gsub("facilitation", "faci", colnames(null_table_3d))
+colnames(null_table_3d) <- gsub("filtering", "filt", colnames(null_table_3d))
+colnames(null_table_3d) <- gsub("competition", "comp", colnames(null_table_3d))
+kable(null_table_3d, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (3d)", align = "c")
+```
+
+```{r, echo = FALSE}
+rownames(null_table_emp) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
+## Even fancier column names
+colnames(null_table_emp) <- gsub(".d_rm", ".rm", colnames(null_table_emp))
+colnames(null_table_emp) <- gsub("equalizing", "equa", colnames(null_table_emp))
+colnames(null_table_emp) <- gsub("facilitation", "faci", colnames(null_table_emp))
+colnames(null_table_emp) <- gsub("filtering", "filt", colnames(null_table_emp))
+colnames(null_table_emp) <- gsub("competition", "comp", colnames(null_table_emp))
+kable(null_table_emp, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal for the empirical data (pre-history and history)", align = "c")
 ```
@@
 ## The plot
-fancy.plot(scaled_results, col = col_vector, lm = model_fits, null = null_differences, metric.names = fancy_names)
+fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names)
 
 ## Pdf version
-pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor.pdf", height = 9, width = 9)
-fancy.plot(scaled_results, col = col_vector, lm = model_fits, null = null_differences, metric.names = fancy_names)
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_2d.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names)
 dev.off()
-
 ```
@@
 The grey line plots represent distribution of metrics scores not clearly different from the random metric scores (paired t-test p value > 0.05).
+
+```{r}
+## The plot
+fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names)
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_3d.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names)
+dev.off()
+```
+
+```{r}
+## The plot
+fancy.plot(scaled_results_emp, col = col_vector, lm = model_fits_emp, null = null_differences_emp, metric.names = fancy_names)
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_emp, col = col_vector, lm = model_fits_emp, null = null_differences_emp, metric.names = fancy_names)
+dev.off()
+```
 

### 2024-02-29 Functions/paired.tests.R

@@ -32,7 +32,7 @@ paired.tests <- function(one_stressor) {
     test_list <- split.metric(t(test_table))
     merge.results <- function(one_metric) {
         out <- c(one_metric)
-        names(out) <- c(t(sapply(paste0(rownames(one_metric), "_rm"), function(X) paste0(X, 1:4))))
+        names(out) <- c(t(sapply(paste0(rownames(one_metric), "_rm"), function(X) paste0(X, 1:ncol(one_metric)))))
         return(out)
     }
     return(t(do.call(cbind, lapply(test_list, merge.results))))

### 2024-03-01 Functions/fancy.plot.R
Updated pipeline for empirical data

@@ -28,43 +28,67 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
         plot_params$lwd = 1
     }
 
-    ## Setting up the main plotting window
-    par(mfrow = c(1, length(results)+1))
-    ## Plotting each stressor
-    for(one_stressor in 0:length(results)) {
-        
-        if(one_stressor == 0) {
-            ## Making the labels plot
-            par(mar = c(5, 10, 4, 0), bty = "n")
-            empty.plot(results[[1]],
-                       plot.main = "",
-                       add.x = FALSE,
-                       add.y = TRUE,
-                       grid  = FALSE,
-                       metric.names = metric.names)
+    if(length(results) > 1) {
+        ## Setting up the main plotting window
+        par(mfrow = c(1, length(results)+1))
+        ## Plotting each stressor
+        for(one_stressor in 0:length(results)) {
+            
+            if(one_stressor == 0) {
+                ## Making the labels plot
+                par(mar = c(5, 10, 4, 0), bty = "n")
+                empty.plot(results[[1]],
+                           plot.main = "",
+                           add.x = FALSE,
+                           add.y = TRUE,
+                           grid  = FALSE,
+                           metric.names = metric.names)
 
-        } else {
-            par(mar = c(5, 0, 4, 1))
-            ## Making the empty plot
-            empty.plot(results[[one_stressor]],
-                       plot.main = names(results)[one_stressor],
-                       add.x = TRUE,
-                       add.y = FALSE)
+            } else {
+                par(mar = c(5, 0, 4, 1))
+                ## Making the empty plot
+                empty.plot(results[[one_stressor]],
+                           plot.main = names(results)[one_stressor],
+                           add.x = TRUE,
+                           add.y = FALSE)
 
-            ## Adding the data
-            if(!missing(lm)) {
-                model_data <- lm[[one_stressor]]
-            } else {
-                model_data <- NULL
+                ## Adding the data
+                if(!missing(lm)) {
+                    model_data <- lm[[one_stressor]]
+                } else {
+                    model_data <- NULL
+                }
+                if(!missing(null)) {
[…]

### 2024-03-04 Analysis/04_Summarise_results.Rmd
Update 04_Summarise_results.Rmd

@@
                        competition  = competition_3d)
-empirical_results <- list(combined_empirical)
+empirical_results <- list(empirical = combined_empirical)
 
@@
 ## Pdf version
-pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_emp, col = col_vector, lm = model_fits_emp, null = null_differences_emp, metric.names = fancy_names)
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 9, width = 4)
+fancy.plot(scaled_results_emp, col = col_vector, lm = model_fits_emp, null = null_differences_emp, metric.names = rev(fancy_names))
 dev.off()

### 2024-03-15 Functions/dbFD.R
up vert.txt fix

@@ -1,0 +1,740 @@
+`dbFD_novert` <- function(x, a , w, w.abun = TRUE, stand.x = TRUE, ord = c("podani", "metric"), asym.bin = NULL, corr = c("sqrt", "cailliez", "lingoes", "none"),  calc.FRic = TRUE, m = "max", stand.FRic = FALSE, scale.RaoQ = FALSE, calc.FGR = FALSE, clust.type = "ward", km.inf.gr = 2, km.sup.gr = nrow(x) - 1, km.iter = 100, km.crit = c("calinski", "ssi"),  calc.CWM = TRUE, CWM.type = c("dom", "all"), calc.FDiv = TRUE, dist.bin = 2, print.pco = FALSE, messages = TRUE){
+
+# get tolerance value specifying small values that should be 0
+tol <- .Machine$double.eps
+
+# match and check arguments
+corr <- match.arg(corr)
+ord <- match.arg(ord)
+CWM.type <- match.arg(CWM.type)
+km.crit <- match.arg(km.crit)
+if (!is.logical(messages) ) stop("'messages' must be TRUE or FALSE.", "\n")
+if (!is.logical(stand.FRic) ) stop("'stand.FRic' must be TRUE or FALSE.","\n")
+if (!is.logical(stand.x) ) stop("'stand.x' must be TRUE or FALSE.","\n")
+if (!is.logical(w.abun) ) stop("'w.abun' must be TRUE or FALSE.","\n")
+if (!is.logical(calc.FRic) ) stop("'calc.FRic' must be TRUE or FALSE.","\n")
+if (!is.logical(calc.FDiv) ) stop("'calc.FDiv' must be TRUE or FALSE.","\n")
+if (!is.logical(calc.FGR) ) stop("'calc.FGR' musts be TRUE or FALSE.","\n")
+if (!is.logical(calc.CWM) ) stop("'calc.CWM' must be TRUE or FALSE.","\n")
+if (!is.logical(scale.RaoQ) ) stop("'scale.RaoQ' must be TRUE or FALSE.","\n")
+if (!is.logical(print.pco) ) stop("'print.pco' must be TRUE or FALSE.","\n")
+
+# check coherence of number of species in 'x' and 'a'
+if (is.matrix(x) | is.data.frame(x) )
+{
+    is.dist.x <- FALSE
+    s.x <- dim(x)[1]
+    t.x <- dim(x)[2]
[…]

### 2024-03-15 Functions/simulation.pipeline.R

@@ -252,7 +252,7 @@ fun.dissimilarity <- function(presence, traits, verbose) {
     ## Only keep columns data have data in common
     keep_rows <- rowSums(presence) > 0
     ## Get the distance based metrics
-    distance_FD <- FD::dbFD(x = traits[keep_rows, ], a = t(presence[keep_rows, ]), print.pco = FALSE, calc.FGR = FALSE, messages = FALSE)
+    distance_FD <- dbFD_novert(x = traits[keep_rows, ], a = t(presence[keep_rows, ]), print.pco = FALSE, calc.FGR = FALSE, messages = TRUE)
 
     ## Get the melodic rao
     melo_rao <- melodic.rao(samp = t(presence[keep_rows, ]), dis = as.matrix(FD::gowdis(traits[keep_rows, ])))

### 2024-03-18 Functions/simulation.pipeline.R
Update simulation.pipeline.R

@@ -314,9 +314,14 @@ fun.proba.den <- function(presence, traits, verbose) {
         byrow = TRUE,
         dimnames = list(rownames(presence), paste0("sdA", 1:ncol(traits))))
 
-    ## Measure probability density metrics 
+    ## Measure probability density metrics
+    if(ncol(traits) <= 4) {
+        traits_4dmax <- traits
+    } else {
+        traits_4dmax <- traits[, c(1:4)]
+    }
     suppressMessages(TPD_species <- TPD::TPDsMean(species = rownames(presence),
-                                 means = traits,
+                                 means = traits_4dmax,
                                  sds = standard_deviations,
                                  alpha = 0.99))
     TPD_cases   <- TPD::TPDc(TPDs = TPD_species, sampUnit = t(presence))

### 2024-03-19 Functions/fancy.plot.R
remove R2

@@ -149,7 +149,7 @@ add.one.metric <- function(one_metric, plot_params, probs, cent.tend, col, lm_da
         lines(x, y, lty = lty, col = "darkgrey")
 
         ## Add the R^2
-        text(paste(names(lm_data[5]), round(lm_data[5], 3), sep = "\n") , y = base-0.4, x = 0.6, cex = 1)
+        # text(paste(names(lm_data[5]), round(lm_data[5], 3), sep = "\n") , y = base-0.4, x = 0.6, cex = 1)
     }
 
     ## Plot each level

### 2024-03-19 Analysis/01_Simulating_spaces_and_stressors.Rmd
removed legend

@@
 plot.text.box("1) Resampling\nwith stressor:\nnull mechanism", cex = 2)
-plot.1D(trait_space, reduction = random_reduction, null = rep(TRUE, 200), main = "1D null mechanism", legend = c("Full distribution (200sp)", "Resample (100sp)"))
+plot.1D(trait_space, reduction = random_reduction, null = !random_reduction, main = "1D null mechanism")
 plot.2D(trait_space, reduction = random_reduction, null = !random_reduction, main = "2D null mechanism")
@@
 plot.text.box("2) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
-plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism", legend = c("Random resampling (100sp)", "Biased resampling (100sp)"))
+plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
 plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D filtering mechanism")

### 2024-03-21 Analysis/01_Simulating_spaces_and_stressors.Rmd
Update 01_Simulating_spaces_and_stressors.Rmd

@@
 
-plot.text.box("2) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
+plot.text.box("2) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
+plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")
+
+plot.text.box("3) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
+plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D facilitation mechanism")
+
+plot.text.box("4) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
@@
 
-plot.text.box("12) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
-plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
-plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")
+plot.text.box("5) Resampling\nwith stressor:\ncompetition\nmechanism", cex = 2)
+plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D competition mechanism")
+plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D competition mechanism")
 
-plot.text.box("4) Resampling\nwith stressor:\nexcluding\nmechanism", cex = 2)
-plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D excluding mechanism")
-plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D excluding mechanism")
-
-plot.text.box("5) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
-plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
-plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D facilitation mechanism")
 dev.off()

### 2024-04-11 Analysis/02_Simulation_pipeline.Rmd
added non verbose version of FD:dbFD
 (for cluster runs)

@@
 source("../Functions/simulation.pipeline.R")
+source("../Functions/dbFD.R")
 ```

### 2024-04-11 Functions/simulation.pipeline.R
bandwidth estimated separately
Following Tom Matthews suggestion

@@ -270,9 +270,12 @@ fun.kernel <- function(presence, traits, verbose) {
         message(".", appendLF = FALSE)
     }
 
+    ## estimate the bandwidth (now formally done on the whole traitspace in nD)
+    suppressMessages(silent <- capture.output(estimated_bw <- hypervolume::estimate_bandwidth(traits, method = "silverman")))
+
     ## Get the hypervolume (remove messages)
-    suppressMessages(silent <- capture.output(hypervolume <- BAT::kernel.build(comm = t(presence), trait = traits)))
-    
+    suppressMessages(silent <- capture.output(hypervolume <- BAT::kernel.build(comm = t(presence), trait = traits, kde.bandwidth = estimated_bw)))
+
     ## Get the values (remove messages)
     suppressMessages(silent <- capture.output(richness <- BAT::kernel.alpha(comm=hypervolume )))
     suppressMessages(silent <- capture.output(dispersion <- BAT::kernel.dispersion(comm = hypervolume)))

### 2024-04-11 Functions/simulation.pipeline.R
added 4D fix by Maria

@@ -310,19 +310,20 @@ fun.proba.den <- function(presence, traits, verbose) {
         message(".", appendLF = FALSE)
     }
 
-    ## Get the sds matrix
-    standard_deviations <- matrix(
-        data = rep(sqrt(diag(ks::Hpi.diag(traits))), nrow(presence)),
-        nrow = nrow(presence),
-        byrow = TRUE,
-        dimnames = list(rownames(presence), paste0("sdA", 1:ncol(traits))))
-
     ## Measure probability density metrics
     if(ncol(traits) <= 4) {
         traits_4dmax <- traits
     } else {
         traits_4dmax <- traits[, c(1:4)]
     }
+
+    ## Get the sds matrix
+    standard_deviations <- matrix(
+        data = rep(sqrt(diag(ks::Hpi.diag(traits_4dmax))), nrow(presence)), # Hpo.diag was originally: traits
+        nrow = nrow(presence),
+        byrow = TRUE,
+        dimnames = list(rownames(presence), paste0("sdA", 1:ncol(traits_4dmax)))) # ncol argument was origninally: traits
+
     suppressMessages(TPD_species <- TPD::TPDsMean(species = rownames(presence),
                                  means = traits_4dmax,
                                  sds = standard_deviations,

### 2024-04-11 Analysis/3-eco-dis-case-study-empirical-pipeline.R
Create 3-eco-dis-case-study-empirical-pipeline.R

@@ -1,0 +1,346 @@
+
+
+
+
+############################################################################################################################
+############################################ ECODISSIMILARITY CASE STUDY: HAWAII ###########################################
+############################################################################################################################
+
+setwd("C:/Users/mwj207/OneDrive - University of Birmingham/hawaii/eco-dissimilarity/github")
+
+library(devtools)
+require(treats)
+require(dispRity)
+require(BAT)
+require(FD)
+require(ks)
+require(TPD)
+library(tidyverse)
+
+## Loading the functions
+source("./Functions/dbFD.R") # I added this in order to import the dbFD function
+source("./Functions/melodic.rao.R")
+source("./Functions/make.simulation.pipeline.R")
+source("./Functions/simulation.pipeline.R")
+
+## Loading the optimisation functions
+source("./Functions/optim.replicate.R")
+source("./Functions/optim.helpers.R")
+source("./Functions/analyse.replicates.R")
+
+
+
+###### SLIGHTLY ALTERED FUNCTIONS ##########################################################################################
+
+fun.proba.den.2 = function(presence, traits, verbose) {
+  if(verbose) {
+    message(".", appendLF = FALSE)
+  }
+  
+  if(ncol(traits) <= 4) {
+    traits_4dmax <- traits
+  } else {
+    traits_4dmax <- traits[, c(1:4)] # moved this from under "Measure probability density metrics
+  }
+  
+  ## Get the sds matrix
+  standard_deviations <- matrix(
+    data = rep(sqrt(diag(ks::Hpi.diag(traits_4dmax))), nrow(presence)), # Hpo.diag was originally: traits
+    nrow = nrow(presence),
+    byrow = TRUE,
+    dimnames = list(rownames(presence), paste0("sdA", 1:ncol(traits_4dmax)))) # ncol argument was origninally: traits
+  
+  ## Measure probability density metrics
+
+  suppressMessages(TPD_species <- TPD::TPDsMean(species = rownames(presence),
+                                                means = traits_4dmax,
+                                                sds = standard_deviations,
+                                                alpha = 0.99))
+  TPD_cases   <- TPD::TPDc(TPDs = TPD_species, sampUnit = t(presence))
+  output <- suppressMessages(do.call(cbind, TPD::REND(TPDc = TPD_cases)$communities))
[…]

### 2024-04-12 Functions/simulation.pipeline.R
added n_taxa internal variable

@@ -108,14 +108,15 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
     } else {
         ## Get trait_space
         trait_space <- sim.data
+        n_taxa <- nrow(trait_space) # number of taxa automatically detected (Maria's suggestion)
         if(is.null(rownames(trait_space))) {
-            rownames(trait_space) <- 1:200
+            rownames(trait_space) <- 1:n_taxa
         }
         ## Create the presences list
         make.presences <- function(one_remove) {
             ## Create the random variable (null)
             random <- rep(FALSE, length(one_remove))
-            random[sample(1:200, size = sum(one_remove))] <- TRUE
+            random[sample(1:n_taxa, size = sum(one_remove))] <- TRUE
             ## Get the dimnames
             if(is.null(names(one_remove))) {
                 rownames <- 1:length(one_remove)

### 2024-04-19 Functions/simulation.pipeline.R
Added safety checks to filtering

@@ -15,8 +15,8 @@
 #' @author Thomas Guillerme
 #' @export
 
-# sim.data = list(n.traits = 2, speciation = 1, n.taxa = 200)
-# type = "facilitation"
+# sim.data = list(n.traits = 8, speciation = 1, n.taxa = 200)
+# type = "filtering"
 # remove = c(0.2,.4, 0.6 ,0.8)
 # library(treats)
 # library(dispRity)
@@ -93,6 +93,20 @@ simulation.pipeline <- function(sim.data, type, remove, verbose = FALSE, record.
                 "competition"  =  sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "evenness", remove = remove, parameters = list(power = 3))), trait_space, simplify = FALSE),
                 "filtering"    =  sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "position", remove = remove)), trait_space, simplify = FALSE)
                 )
+            ## Redo if filtering didn't work
+            if(type == "filtering") {
+                ## checking if it roughly worked
+                test_targets <- as.list(unlist(lapply(target_reductions, sum))/sim.data$n.taxa)
+                check.targets <- function(one_test, one_remove, tolerance = 0.1) {
+                    one_test >= (one_remove-tolerance) && one_test <= (one_remove+tolerance)
+                }
+                while(!all(mapply(check.targets, test_targets, as.list(remove)))) {
+                    ## Re run the reduction until it works
+                    target_reductions <- sapply(remove, function(remove, trait_space) return(dispRity::reduce.space(trait_space, type = "density", remove = remove)), trait_space, simplify = FALSE)
+                    test_targets <- as.list(unlist(lapply(target_reductions, sum))/sim.data$n.taxa)
+                    if(verbose) message(".", appendLF = FALSE)
+                }
+            }
         }
 
         ## Wrapping up the data

### 2024-04-19 Analysis/04_Summarise_results.Rmd
up vert.txt fix
Update 04_Summarise_results.Rmd

@@
 In this section we are going to analyse the results from the `optim.replicates` chain
+
+TODO:
+ - [ ] remove R2 on plots
+ - [ ] finalise empirical plot
+ - [ ] run simulations for 4D and 8D
 
@@
 load("../Data/Processed/empirical_data-hist-present-native.rda")
+#empirical_data-hist-extant
+
 empirical_hist <- bkp
 load("../Data/Processed/empirical_data-prehist-present-native.rda")
+#empirical_data-prehist-hist-prehist-extant
 empirical_prehist <- bkp

### 2024-04-22 Analysis/01_Simulating_spaces_and_stressors.Rmd
added letters

@@
 ## First line
-plot.text.box("Tree and trait\nsimulation:", cex = 2)
+plot.text.box("A) Tree and trait\nsimulation:", cex = 2)
 plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (1D)", col = c("lightgrey", "blue"), ylab = "Trait 1")
@@
 ## Stressors
-plot.text.box("1) Resampling\nwith stressor:\nnull mechanism", cex = 2)
+plot.text.box("B) Resampling\nwith stressor:\nnull mechanism", cex = 2)
 plot.1D(trait_space, reduction = random_reduction, null = !random_reduction, main = "1D null mechanism")
@@
 
-plot.text.box("2) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
+plot.text.box("C) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
@@
 
-plot.text.box("3) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
+plot.text.box("D) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
@@
 
-plot.text.box("4) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
+plot.text.box("E) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
@@
 
-plot.text.box("5) Resampling\nwith stressor:\ncompetition\nmechanism", cex = 2)
+plot.text.box("F) Resampling\nwith stressor:\ncompetition\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D competition mechanism")

### 2024-04-22 Analysis/04_Summarise_results.Rmd
piped data input + added time per metric plot

--- 2024-03-04 Analysis/04_Summarise_results.Rmd
@@
 library(knitr)
-## Loading the results
-load("../Data/Processed/sim_equalizing_2d.rda")
-equalizing_2d <- bkp
-load("../Data/Processed/sim_facilitation_2d.rda")
-facilitation_2d <- bkp
-load("../Data/Processed/sim_filtering_2d.rda")
-filtering_2d <- bkp
-load("../Data/Processed/sim_competition_2d.rda")
-competition_2d <- bkp
-
-## Loading the results
-load("../Data/Processed/sim_equalizing_3d.rda")
-equalizing_3d <- bkp
-load("../Data/Processed/sim_facilitation_3d.rda")
-facilitation_3d <- bkp
-load("../Data/Processed/sim_filtering_3d.rda")
-filtering_3d <- bkp
-load("../Data/Processed/sim_competition_3d.rda")
-competition_3d <- bkp
-
-## Loading empirical results
-load("../Data/Processed/empirical_data-hist-present-native.rda")
-empirical_hist <- bkp
-load("../Data/Processed/empirical_data-prehist-present-native.rda")
-empirical_prehist <- bkp
-
-hist_table <- empirical_hist$results_table
-colnames(hist_table) <- gsub("_rm1", "_rm2", colnames(hist_table))
-combined_empirical <- list()
-combined_empirical$results_table <- cbind(empirical_prehist$results_table, hist_table)
-
 ## Loading the functions
@@
 source("../Functions/fancy.names.R")
+```
+
+```{r, echo = FALSE}
+## Loading all results per dimensions
+
+## Simulated results:
+read.results <- function(nd, stressors = c("equalizing", "facilitation", "filtering", "competition")) {
+  list_load <- list()
+  for(one_stressor in stressors) {
+    load(paste0("../Data/Processed/sim_", one_stressor, "_", nd, ".rda"))
+    list_load[[length(list_load)+1]] <- bkp
+    names(list_load)[length(list_load)] <- one_stressor
+  }
+  return(list_load)
+}
+
+## 2d
+all_results_2d <- read.results("2d")
+
+## 3d
+all_results_3d <- read.results("3d")
+## placeholders
+all_results_4d <- read.results("3d")
+all_results_5d <- read.results("3d")
+all_results_8d <- read.results("3d")
+
+## 4d
+# all_results_4d <- read.results("4d")
+
+# 5d
+# all_results_5d <- read.results("5d")
+
+## 8d
+# all_results_8d <- read.results("8d")
+
+## Empirical
+
+## Loading empirical results
+load("../Data/Processed/empirical_data-hist-extant.rda")
+#empirical_data-hist-extant
+empirical_hist_extant <- bkp
+
+load("../Data/Processed/empirical_data-prehist-hist-prehist-extant.rda")
+#empirical_data-prehist-hist-prehist-extant
+empirical_prehist_hist_prehist_extant <- bkp
+
+hist_table <- empirical_hist$results_table
+colnames(hist_table) <- gsub("_rm1", "_rm2", colnames(hist_table))
+combined_empirical <- list()
+combined_empirical$results_table <- cbind(empirical_prehist$results_table, hist_table)
+empirical_results <- list(empirical = combined_empirical)
 ```
@@
 ```{r, echo = FALSE}
-## Creating the list of results
-all_results_2d <- list(equalizing   = equalizing_2d,
-                       facilitation = facilitation_2d,
-                       filtering    = filtering_2d,
-                       competition  = competition_2d)
-all_results_3d <- list(equalizing   = equalizing_3d,
-                       facilitation = facilitation_3d,
-                       filtering    = filtering_3d,
-                       competition  = competition_3d)
-empirical_results <- list(empirical = combined_empirical)
-
 ## Sort the results per name
@@
 dev.off()
+```
+
+## Get the computation time per metric type
+
+```{r}
+#@ combine logical, whether to combine the results per stressor (TRUE) or not (FALSE)
+get.time.per.type <- function(all_results, combine = TRUE) {
+  ## Get the timers per stressor
+  get.time.per.stressors <- function(one_stressor_results) {
+    return(do.call(rbind, lapply(one_stressor_results$output_save, function(x) return(log(as.numeric(as.difftime(x$timer, units = "secs")))))))
+  }
+
+  ## Get all the timers
+  metrics_times <- lapply(all_results, get.time.per.stressors)
+
+  ## Combine it
+  if(combine) {
+    metric_times <- do.call(rbind, metrics_times)
+  }
+}
+
+## times per dimensions
+times_2d <- apply(get.time.per.type(all_results_2d), 2, quantile, probs = c(0.025, 0.5, 0.975))
+times_3d <- apply(get.time.per.type(all_results_3d), 2, quantile, probs = c(0.025, 0.5, 0.975))
+times_4d <- apply(get.time.per.type(all_results_4d), 2, quantile, probs = c(0.025, 0.5, 0.975))
+times_5d <- apply(get.time.per.type(all_results_5d), 2, quantile, probs = c(0.025, 0.5, 0.975))
+times_8d <- apply(get.time.per.type(all_results_8d), 2, quantile, probs = c(0.025, 0.5, 0.975))
+
+times_tree <- cbind(times_2d[,2], times_3d[,2], times_4d[,2], times_5d[,2], times_8d[,2])
+times_dissimilarity <- cbind(times_2d[,3], times_3d[,3], times_4d[,3], times_5d[,3], times_8d[,3])
+times_hull <- cbind(times_2d[,4], times_3d[,4], times_4d[,4], times_5d[,4], times_8d[,4])
+times_hypervolume <- cbind(times_2d[,5], times_3d[,5], times_4d[,5], times_5d[,5], times_8d[,5])
+times_probabilities <-  cbind(times_2d[,6], times_3d[,6], times_4d[,6], times_5d[,6], times_8d[,6])
+
+## Plot times per dimensions
+ylim <- range(c(times_tree, times_dissimilarity, times_hull, times_hypervolume, times_probabilities))
+## Plot colours
+metric_cols <- c("#d55e00", "#cc79a7", "#0072b2", "#f0e442", "#009e73")
+
+## Empty plot
+plot(NULL, ylim = range(pretty(ylim)), xlim = c(2, 8), ylab = "Time (log(s))", xlab = "dimensions")
+
+## Medians
+points(x = c(2,3,4,5,8)+0.00, c(times_tree[2,]), pch = 19, col = metric_cols[1])
+points(x = c(2,3,4,5,8)+0.04, c(times_dissimilarity[2,]), pch = 19, col = metric_cols[2])
+points(x = c(2,3,4,5,8)+0.08, c(times_hull[2,]), pch = 19, col = metric_cols[3])
+points(x = c(2,3,4,5,8)+0.12, c(times_hypervolume[2,]), pch = 19, col = metric_cols[4])
+points(x = c(2,3,4,5,8)+0.16, c(times_probabilities[2,]), pch = 19, col = metric_cols[5])
+## Linkings
+lines(x = c(2,3,4,5,8)+0.00, c(times_tree[2,]), pch = 19, col = metric_cols[1])
+lines(x = c(2,3,4,5,8)+0.04, c(times_dissimilarity[2,]), pch = 19, col = metric_cols[2])
+lines(x = c(2,3,4,5,8)+0.08, c(times_hull[2,]), pch = 19, col = metric_cols[3])
+lines(x = c(2,3,4,5,8)+0.12, c(times_hypervolume[2,]), pch = 19, col = metric_cols[4])
+lines(x = c(2,3,4,5,8)+0.16, c(times_probabilities[2,]), pch = 19, col = metric_cols[5])
+## 95% CI
+dimensions <- c(2,3,4,5,8)
+for(one_ci in 1:5) {
+  lines(y = c(times_tree[c(1,3), one_ci]), x = rep(dimensions[one_ci], 2)+0.00, col = metric_cols[1])
+}
+for(one_ci in 1:5) {
+  lines(y = c(times_dissimilarity[c(1,3), one_ci]), x = rep(dimensions[one_ci], 2)+0.04, col = metric_cols[2])
+}
+for(one_ci in 1:5) {
+  lines(y = c(times_hull[c(1,3), one_ci]), x = rep(dimensions[one_ci], 2)+0.08, col = metric_cols[3])
+}
+for(one_ci in 1:5) {
+  lines(y = c(times_hypervolume[c(1,3), one_ci]), x = rep(dimensions[one_ci], 2)+0.12, col = metric_cols[4])
+}
+for(one_ci in 1:5) {
+  lines(y = c(times_probabilities[c(1,3), one_ci]), x = rep(dimensions[one_ci], 2)+0.16, col = metric_cols[5])
+}
+## legend
+legend("topleft", col = metric_cols, pch = 19, legend = c("tree", "dissimilarity", "convex hull", "hypervolume", "probability density"))
 ```

### 2024-04-22 Analysis/02_Simulation_pipeline.Rmd
Update 02_Simulation_pipeline.Rmd

@@
 
+<!-- 
 ## Checking the results
@@
 analyse.replicates("../Data/Processed/sim_facilitation.rda") #163
-```
+``` -->

### 2024-04-22 Analysis/script_empirical.R
Create script_empirical.R

@@ -1,0 +1,45 @@
+## Libraries
+require(treats)
+require(dispRity)
+require(BAT)
+require(FD)
+require(ks)
+require(TPD)
+
+## Loading the functions
+source("../Functions/melodic.rao.R")
+source("../Functions/make.simulation.pipeline.R")
+source("../Functions/simulation.pipeline.R")
+source("../Functions/optim.replicate.R")
+source("../Functions/optim.helpers.R")
+source("../Functions/analyse.replicates.R")
+source("../Functions/dbFD.R")
+
+## global trait space 
+empirical_traitspace <- read.csv('../Data/Raw/trait-space-5d-prehistoric-community.csv', row.names = 1) 
+
+## Species lists
+prehistoric_species <- rownames(read.csv("../Data/Raw/prehistoric-species-list.csv"))
+historic_species <- rownames(read.csv('../Data/Raw/historic-species-list-minus-a-flammeus.csv', row.names = 1))
+extant_species <- rownames(read.csv('../Data/Raw/present-species-list-minus-a-flammeus.csv', row.names = 1))
+
+## Traitspace groups
+extant_group <- rownames(empirical_traitspace) %in% extant_species ## 37 species present
+historic_group <- rownames(empirical_traitspace) %in% historic_species ## 63 species present
+prehistoric_group <- !(rownames(empirical_traitspace) %in% c(extant_species, historic_species)) ## 55 species present
+names(extant_group) <- names(historic_group) <- names(prehistoric_group) <- rownames(empirical_traitspace)
+
+groups <- list("extant" = extant_group, "historic" = historic_group, "prehistoric" = prehistoric_group)
+
+## Set simulation type
+simulate <- make.simulation.pipeline(sim.data = empirical_traitspace, remove = groups)
+
+## Running the full optimisation
+results <- optim.replicate(input.fun = simulate,
+                           diagnose = var,
+                           summarise = optim.summarise,
+                           minimum = 20, maximum = 170,
+                           stop.variance = 0.0001,
+                           verbose = TRUE,
+                           bkp.path = "../Data/Processed/",
+                           bkp.name = "empirical_data.rda")

### 2024-04-23 Analysis/04_Summarise_results.Rmd
Update 04_Summarise_results.Rmd

@@
 fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names)
+dev.off()
+```
+
+```{r}
+## The plot
+fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names)
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_4d.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names)
+dev.off()
+```
+
+```{r}
+## The plot
+fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names)
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_5d.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names)
+dev.off()
+```
+
+```{r}
+## The plot
+fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names)
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_8d.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names)
 dev.off()

### 2024-04-24 Analysis/04_Summarise_results.Rmd
Update 04_Summarise_results.Rmd

@@
 
-## 2d
 all_results_2d <- read.results("2d")
-
-## 3d
 all_results_3d <- read.results("3d")
-## placeholders
-all_results_4d <- read.results("3d")
-all_results_5d <- read.results("3d")
-all_results_8d <- read.results("3d")
-
-## 4d
-# all_results_4d <- read.results("4d")
-
-# 5d
-# all_results_5d <- read.results("5d")
-
-## 8d
-# all_results_8d <- read.results("8d")
+all_results_4d <- read.results("4d")
+all_results_5d <- read.results("5d")
+all_results_8d <- read.results("8d")
 
 ## Empirical
-
-## Loading empirical results
-load("../Data/Processed/empirical_data-hist-extant.rda")
-#empirical_data-hist-extant
-empirical_hist_extant <- bkp
-
-load("../Data/Processed/empirical_data-prehist-hist-prehist-extant.rda")
-#empirical_data-prehist-hist-prehist-extant
-empirical_prehist_hist_prehist_extant <- bkp
-
-hist_table <- empirical_hist$results_table
-colnames(hist_table) <- gsub("_rm1", "_rm2", colnames(hist_table))
-combined_empirical <- list()
-combined_empirical$results_table <- cbind(empirical_prehist$results_table, hist_table)
-empirical_results <- list(empirical = combined_empirical)
+load("../Data/Processed/empirical_data.rda")
+empirical_results <- list("empirical" = bkp)
 ```
@@
                15, 11, 2, 8,
-               7, 4, 5, 13,
+               7, 4, #5,
+               13,
                10, 9, 1)
@@
 all_results_3d <- lapply(all_results_3d, sort.metrics, order = new_order)
+all_results_4d <- lapply(all_results_4d, sort.metrics, order = new_order)
+all_results_5d <- lapply(all_results_5d, sort.metrics, order = new_order)
+all_results_8d <- lapply(all_results_8d, sort.metrics, order = new_order)
 empirical_results <- lapply(empirical_results, sort.metrics, order = new_order)
@@
                  "dissimilarity[FD]\ndivergence(Rao)",
-                 "dissimilarity[FD]\ndivergence",
+                 #"dissimilarity[FD]\ndivergence",
                  "TPD[TPD]\nrichness",
@@
 scaled_within_3d  <- extract.table(all_results_3d, scale.method = "within")
+scaled_results_4d <- extract.table(all_results_4d, scale.method = "between")
+scaled_within_4d  <- extract.table(all_results_4d, scale.method = "within")
+scaled_results_5d <- extract.table(all_results_5d, scale.method = "between")
+scaled_within_5d  <- extract.table(all_results_5d, scale.method = "within")
+scaled_results_8d <- extract.table(all_results_8d, scale.method = "between")
+scaled_within_8d  <- extract.table(all_results_8d, scale.method = "within")
 scaled_results_emp<- extract.table(empirical_results, scale.method = "between")
@@
 model_fits_3d  <- lapply(scaled_results_3d, fit.model.rm)
+model_fits_4d  <- lapply(scaled_results_4d, fit.model.rm)
+model_fits_5d  <- lapply(scaled_results_5d, fit.model.rm)
+model_fits_8d  <- lapply(scaled_results_8d, fit.model.rm)
 model_fits_emp <- lapply(scaled_results_emp, fit.model.rm)
@@
 null_differences_3d  <- lapply(all_results_3d, paired.tests)
+null_differences_4d  <- lapply(all_results_4d, paired.tests)
+null_differences_5d  <- lapply(all_results_5d, paired.tests)
+null_differences_8d  <- lapply(all_results_8d, paired.tests)
 null_differences_emp <- lapply(empirical_results, paired.tests)
@@
 model_table_3d  <- fancy.table(model_fits_3d, columns = list(c(3,4), 5))
+model_table_4d  <- fancy.table(model_fits_4d, columns = list(c(3,4), 5))
+model_table_5d  <- fancy.table(model_fits_5d, columns = list(c(3,4), 5))
+model_table_8d  <- fancy.table(model_fits_8d, columns = list(c(3,4), 5))
 model_table_emp <- fancy.table(model_fits_emp, columns = list(c(3,4), 5))
@@
 null_table_3d   <- fancy.table(null_differences_3d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
+null_table_4d   <- fancy.table(null_differences_4d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
+null_table_5d   <- fancy.table(null_differences_5d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
+null_table_8d   <- fancy.table(null_differences_8d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
 null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), c(4, 6)))
@@
 legend("topleft", col = metric_cols, pch = 19, legend = c("tree", "dissimilarity", "convex hull", "hypervolume", "probability density"))
+
+## Ti me
+axis(4, at = c(1, 4.1, 7, 8.189), labels = c("1 sec", "1 min", "30 min", "1 hour"), las = 2)
+
 ```

### 2024-04-26 Analysis/04_Summarise_results.Rmd
Update 04_Summarise_results.Rmd

@@
 empirical_results <- list("empirical" = bkp)
+#rm1 = "extant", rm2 = "historic", rm3 = "prehistoric"
+
+#TODO: reorder so to have rm2, rm3, rm1
+
 ```
@@
 ## The plot
-fancy.plot(scaled_results_emp, col = col_vector, lm = model_fits_emp, null = null_differences_emp, metric.names = fancy_names)
+fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 9, width = 4)
-fancy.plot(scaled_results_emp, col = col_vector, lm = model_fits_emp, null = null_differences_emp, metric.names = rev(fancy_names))
+fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = rev(fancy_names), , pch = c(19,15,17))
 dev.off()
 ```
+
+Figure 3: Empirical data.
+Plain circles represents the metric for historic data (53 % - 63/118 species), lightly fainted squares for prehistoric data (46% - 55/118 species), and fainted triangles for extant data (31% - 37/118 species).
+<!-- TODO: this caption is for the empirical data re-arranged as rm2, rm3, rm1 -->
 
@@
 
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
+```{r, eval = FALSE, echo = FALSE, message = FALSE}
 require(psych)

### 2024-04-26 Functions/fancy.plot.R
Update fancy.plot.R

@@ -9,6 +9,7 @@
 #' @param lm optional, lm results to add to the plot
 #' @param null optional, paired t-tests to add to the plot
 #' @param metric.names optional, a vector of fancy names for the results (rows)
+#' @param empirical optional, whether to plot empirical data or not
 #' @param ... plot options
 #' 
 #' @examples
@@ -17,7 +18,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, ...) {
+fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, empirical = FALSE, ...) {
 
     ## Get the plot parameters
     plot_params <- list(...)
@@ -155,12 +156,16 @@ add.one.metric <- function(one_metric, plot_params, probs, cent.tend, col, lm_da
     ## Plot each level
     for(one_level in 1:levels) {
         ## Setting the plot params
-        plot_params$col <- colours[one_level]
-        plot_params$y   <- rev(((base-levels)+1):base)[one_level]
+        plotting_params <- plot_params
+        plotting_params$col <- colours[one_level]
+        plotting_params$y   <- rev(((base-levels)+1):base)[one_level]
+        if(length(plot_params$pch == levels)) {
+            plotting_params$pch <- plot_params$pch[one_level]
+        }
         ## Setting the pch param
         if(!is.null(null_data)) {
             if(null_data[3*one_level] > 0.05) {
-                plot_params$col <- "grey"
+                plotting_params$col <- "grey"
             }
             # plot_params$pch <- 1
             # if(null_data[3*one_level] < 0.05) {
@@ -170,7 +175,7 @@ add.one.metric <- function(one_metric, plot_params, probs, cent.tend, col, lm_da
             #     }
             # }
         }
-        add.one.line(one_metric[, one_level], plot_params = plot_params, probs = probs, cent.tend = cent.tend)
+        add.one.line(one_metric[, one_level], plot_params = plotting_params, probs = probs, cent.tend = cent.tend)
     }
 }
 

### 2024-04-29 Analysis/04_Summarise_results.Rmd
Update 04_Summarise_results.Rmd

@@
 #rm1 = "extant", rm2 = "historic", rm3 = "prehistoric"
-
-#TODO: reorder so to have rm2, rm3, rm1
-
+extants <- grep("rm1", colnames(empirical_results$empirical$results_table))
+historics <- grep("rm2", colnames(empirical_results$empirical$results_table))
+prehistorics <- grep("rm3", colnames(empirical_results$empirical$results_table))
+## Reorder by percentage removed (historics > prehistorics > extants)
+empirical_results$empirical$results_table <- empirical_results$empirical$results_table[, c(historics, prehistorics, extants)]
 ```
@@
 ## The plot
-fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names)
+fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_2d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names)
+fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names)
+fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names, pch = rep(19,4))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_3d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names)
+fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names, pch = rep(19,4))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names)
+fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names, pch = rep(19,4))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_4d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names)
+fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names, pch = rep(19,4))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names)
+fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_5d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names)
+fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names)
+fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_8d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names)
+fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4))
 dev.off()
@@
 ## Pdf version
-pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 9, width = 4)
-fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = rev(fancy_names), , pch = c(19,15,17))
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 9, width = 4.5)
+fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = rev(fancy_names), pch = c(19,15,17))
 dev.off()
@@
 Plain circles represents the metric for historic data (53 % - 63/118 species), lightly fainted squares for prehistoric data (46% - 55/118 species), and fainted triangles for extant data (31% - 37/118 species).
-<!-- TODO: this caption is for the empirical data re-arranged as rm2, rm3, rm1 -->
 
@@
 
+pdf(file = "../Manuscript/Figures/time_per_metric.pdf", height = 6, width = 7)
+par(mar =c(5, 4, 4, 4) + 0.1)
 ## Empty plot
 plot(NULL, ylim = range(pretty(ylim)), xlim = c(2, 8), ylab = "Time (log(s))", xlab = "dimensions")
+abline(h = c(1, 4.1, 7, 8.189), col = "grey")
 
@@
 
-## Ti me
+## Time
 axis(4, at = c(1, 4.1, 7, 8.189), labels = c("1 sec", "1 min", "30 min", "1 hour"), las = 2)
+dev.off()
 

### 2024-05-01 Analysis/04_Summarise_results.Rmd
Update 04_Summarise_results.Rmd

@@
 historics <- grep("rm2", colnames(empirical_results$empirical$results_table))
-prehistorics <- grep("rm3", colnames(empirical_results$empirical$results_table))
+# prehistorics <- grep("rm3", colnames(empirical_results$empirical$results_table))
 ## Reorder by percentage removed (historics > prehistorics > extants)
-empirical_results$empirical$results_table <- empirical_results$empirical$results_table[, c(historics, prehistorics, extants)]
+empirical_results$empirical$results_table <- empirical_results$empirical$results_table[, c(historics, extants)]
 ```
@@
 ## Fancier names
-fancy_names <- c("TPD[TPD]\nregularity",
-                 "hypervolume[BAT]\nregularity",
-                 "dendrogram[BAT]\nregularity",
-                 "dissimilarity[FD]\nregularity",
-                 "TPD[TPD]\ndivergence",
-                 "hypervolume[BAT]\ndivergence",
-                 "dendrogram[BAT]\ndivergence",
-                 "dissimilarity[melodicMPD]\ndivergence(MPD)",
-                 "dissimilarity[melodicRao]\ndivergence(Rao)",
-                 "dissimilarity[FD]\ndivergence(Rao)",
+fancy_names <- rev(
+               c("Functional evenness\nprobability density\nTPD::REND",
+                 "Functional evenness\nhypervolume\nBAT::kernel.evenness",
+                 "Functional evenness\ndendrogram\nBAT::evenness",
+                 "Functional evenness\ndistances\nFD::dbFD()$FEve",
+                 "Divergence\nprobability density\nTPD::REND",             
+                 "Functional dispersion\nhypervolume\nBAT::kernel.dispersion",
+                 "Functional dispersion\ndendrogram\nBAT::dispersion",
+                 "Melodic Mean Pairwise Distance\ndistances\nmelodic.rao",
+                 "Melodic Rao\ndistances\nmelodic.rao",
+                 "Rao's quadratic entropy\ndistances\nFD::dbFD()$RaoQ",
                  #"dissimilarity[FD]\ndivergence",
-                 "TPD[TPD]\nrichness",
-                 "hypervolume[BAT]\nrichness",
-                 "convex hull[BAT]\nrichness",
-                 "dendrogram[BAT]\nrichness")
+                 "Functional richness\nprobability density\nTPD::REND",
+                 "Functional richness\nhypervolume\nBAT::kernel.alpha",
+                 "Functional richness\nconvex hull\nBAT::hull.alpha",
+                 "Alpha diversity\ndendrogram\nBAT::alpha"))
 ```
@@
 col_vector <- c(rep("blue", 4), # The richness ones
-                rep("orange", 7), # The divergence ones
+                rep("orange", 6), # The divergence ones
                 rep("darkgreen", 4)) # The regularity ones
@@
 ## Pdf version
-pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 9, width = 4.5)
-fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = rev(fancy_names), pch = c(19,15,17))
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 11, width = 4.5)
+fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17))
 dev.off()
@@
 ## legend
-legend("topleft", col = metric_cols, pch = 19, legend = c("tree", "dissimilarity", "convex hull", "hypervolume", "probability density"))
+legend("topleft", col = metric_cols, pch = 19, legend = c("dendrogram", "dissimilarity", "convex hull", "hypervolume", "probability density"))
 

### 2024-05-09 Analysis/script_empirical.R
Update script_empirical.R

@@ -19,17 +19,14 @@ source("../Functions/dbFD.R")
 empirical_traitspace <- read.csv('../Data/Raw/trait-space-5d-prehistoric-community.csv', row.names = 1) 
 
 ## Species lists
-prehistoric_species <- rownames(read.csv("../Data/Raw/prehistoric-species-list.csv"))
+# prehistoric_species <- rownames(read.csv("../Data/Raw/prehistoric-species-list.csv"))
 historic_species <- rownames(read.csv('../Data/Raw/historic-species-list-minus-a-flammeus.csv', row.names = 1))
 extant_species <- rownames(read.csv('../Data/Raw/present-species-list-minus-a-flammeus.csv', row.names = 1))
 
 ## Traitspace groups
 extant_group <- rownames(empirical_traitspace) %in% extant_species ## 37 species present
 historic_group <- rownames(empirical_traitspace) %in% historic_species ## 63 species present
-prehistoric_group <- !(rownames(empirical_traitspace) %in% c(extant_species, historic_species)) ## 55 species present
-names(extant_group) <- names(historic_group) <- names(prehistoric_group) <- rownames(empirical_traitspace)
-
-groups <- list("extant" = extant_group, "historic" = historic_group, "prehistoric" = prehistoric_group)
+groups <- list("extant" = extant_group, "historic" = historic_group)
 
 ## Set simulation type
 simulate <- make.simulation.pipeline(sim.data = empirical_traitspace, remove = groups)

### 2024-05-09 Analysis/testing_empirical_data.Rmd
Testing metrics for empirical data

---
title: "Testing empirical trait space increase"
output:
  html_document: default
  pdf_document: default
---

# Loading the traitspace and the groups

```{r}
library(dispRity)
library(knitr)
## global trait space 
empirical_traitspace <- read.csv('../Data/Raw/trait-space-5d-prehistoric-community.csv', row.names = 1)

## Species lists
# prehistoric_species <- rownames(read.csv("../Data/Raw/prehistoric-species-list.csv"))
historic_species <- rownames(read.csv('../Data/Raw/historic-species-list-minus-a-flammeus.csv', row.names = 1))
extant_species <- rownames(read.csv('../Data/Raw/present-species-list-minus-a-flammeus.csv', row.names = 1))

## Traitspace groups
extant_group <- rownames(empirical_traitspace) %in% extant_species ## 37 species present
historic_group <- rownames(empirical_traitspace) %in% historic_species ## 63 species present
groups <- list("extant" = extant_group, "historic" = historic_group)
```

# Creating the groups and visualising the space

```{r}
## Creating the subsets
custom_groups <- lapply(groups, which)
custom_groups$all <- 1:nrow(empirical_traitspace)
grouped_space <- custom.subsets(empirical_traitspace, group = custom_groups)
```

```{r, fig.height = 8, fig.width = 8}
## Visualising the traitspace (you can also just use: plot(grouped_space)!)
plot(empirical_traitspace[, c(1,2)], pch = 19, col = "darkgreen", xlim = c(-4,6.5), ylim = c(-6.5, 4), xlab = "PC1 (79.11%)", ylab = "PC2 (7.81%)", cex = 1.5)
points(empirical_traitspace[custom_groups$historic, c(1,2)], pch = 19, col = "lightblue", cex = 1.25)
points(empirical_traitspace[custom_groups$extant, c(1,2)], pch = 19, col = "orange", cex = 0.9)
legend("bottomright", legend = c("all (prehist)", "historic", "extant"), pch = 19, col = c("darkgreen", "lightblue", "orange"))
```

```{r, fig.height = 8, fig.width = 8}
## Visualising the space in n_dimensionss
col_groups <- rep("darkgreen", nrow(empirical_traitspace))
col_groups[custom_groups$historic] <- "lightblue"
col_groups[custom_groups$extant] <- "orange"
plot(empirical_traitspace, col = col_groups, pch = 19)
```

# Measuring disparity

## Functional divergence (Villeger 2009)

The rarefaction here is for checking any potential sample size effect:

```{r}
summary(dispRity(boot.matrix(grouped_space, rarefaction = 37), metric = func.div))
```

So here we have `prehistoric < historic < extant` i.e. and increase in functional divergence through time.

## pseudo half hyperperimeter

We can then check using the metric `c(sum, quantiles)` which is the sum of the spread of the 95% quantile i.e. the sum of the spread on each dimensions without potential outliers.
The two times the sum of lengths is the perimeter of a parallelepiped so here we have the "pseudo semi hyperperimeter" or something like that?

```{r}
summary(dispRity(boot.matrix(grouped_space, rarefaction = 37), metric = c(sum, quantiles)))
```

Using this we have `prehistoric > historic > extant`, i.e. a decrease in trait space occupancy through time.

# Testing the metrics

This basically requires us to actually test what we mean by changes in traitspace occupancy through time.
For example, when we mean functional divergence, what do we expect this to capture in terms of size, position, density, etc. of the traitspace?
Let's test both in terms of traitspace size, just for the example:

## Functional divergence

```{r}
## Testing the functional divergence
func_div_test <- test.metric(grouped_space, metric = func.div, shifts = c("random", "size"), replicates = 30)
```

The results for the functional divergence tests:
```{r, echo = FALSE}
kable(summary(func_div_test))
```

The visualisation of the results:
```{r, echo = FALSE, fig.height = 12, fig.width = 6}
plot(func_div_test)
```

If we just focus on changes of the traitspace size, we see that the relationship between this metric and the changes in traitspace size are not linear which might explain the difficulties in interpretations!
First, we see that the random removal is actually good: the metric does not change on average when removing random number of elements (the flat slope on the first plot).

Then, when looking at size increase (orange slope), we see an actual gradual decrease of the metric with the amount of data. The more data, the smaller the functional divergence (this is what we actually have in our empirical example!).
When looking at hollowness (blue slope - the whole left in the traitspace when removing data), the relationship with amounts of data and functional divergence is quiet funky and non linear!

## pseudo half hyperperimeter

```{r}
## Testing the sum of quantiles
sum_quant_test <- test.metric(grouped_space, metric = c(sum, quantiles),  shifts = c("random", "size"), replicates = 30)
```

The results for the sum of the quantiles tests:
```{r, echo = FALSE}
kable(summary(sum_quant_test))
```

The visualisation of the results:
```{r, echo = FALSE, fig.height = 12, fig.width = 6}
plot(sum_quant_test)
```

For the sum of quantiles, we have something worse going on here.
When looking at just the random changes, there is an increase of metric through time, so basically it's gonna be hard to say if this metric picks up meaningful changes in traitspace or random ones.

## median distances from the centroids

A third alternative could be to use the mediand distances from each groups' centroid: this one seem more intuitive, at least for the size increase.

```{r}
plot(test.metric(grouped_space, metric = c(median, centroids),  shifts = c("random", "size"), replicates = 10))
```

### 2024-05-14 Manuscript/main.tex
Manuscript INIT

\documentclass[12pt,letterpaper]{article}

%Packages
% \usepackage{textcomp}
% \usepackage{latexsym}
% \usepackage{url}
% \usepackage{amssymb}
% \usepackage{amsmath}
% \usepackage{mathtools}
% \usepackage{bm}
% \usepackage{array}
% \usepackage[version=3]{mhchem}
% \usepackage{ifthen}
% \usepackage{amsthm}
% \usepackage{amstext}
% \usepackage{enumerate}
% \usepackage{dcolumn}

\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{pdflscape}
\usepackage{mathtools}
\usepackage[osf]{mathpazo}
\usepackage{fullpage}
\usepackage{float}
\usepackage{xr} %linking to supplementaries
\externaldocument{supplementaries}

\pagenumbering{arabic}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}
%Running head
\begin{flushright}
Version dated: \today
\end{flushright}

\bigskip
\bigskip
\begin{center}

\noindent{\Large \bf The what, how, and why of trait-based analyses in ecology}
\bigskip

\noindent {\normalsize \sc
Thomas Guillerme$^{1,*}$, 
Pedro Cardoso$^{2,3}$,
Carlos P. Carmona$^{4}$,
Maria Wagner J\o rgensen$^{5}$,
Stefano Mammola$^{3,6,7}$,
Thomas J. Matthews$^{5,8}$}\\
\noindent {\small \it 
$^1$School of Biosciences, The University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$CE3C—Centre for Ecology, Evolution and Environmental Changes, CHANGE – Global Change and Sustainability Institute, Faculty of Sciences, University of Lisbon, Lisbon, Portugal\\
$^3$Laboratory for Integrative Biodiversity Research (LIBRe), Finnish Museum of Natural History (Luomus), University of Helsinki, Helsinki, Finland\\
$^4$Institute of Ecology and Earth Sciences, University of Tartu, Tartu, Estonia\\
$^5$School of Geography, Earth and Environmental Sciences and Birmingham Institute of Forest Research, University of Birmingham, Birmingham B15 2TT, UK\\
$^6$Molecular Ecology Group (MEG), Water Research Institute, National Research Council (CNR-IRSA), Verbania Pallanza, Italy\\
$^7$National Biodiversity Future Center, Piazza Marina 61, 90133 Palermo, Italy\\
$^8$Centre for Ecology, Evolution and Environmental Changes/Azorean Biodiversity Group / CHANGE – Global Change and Sustainability Institute and Universidade dos Açores – Faculty of Agricultural Sciences and Environment; PT-9700-042, Angra do Hero\'ismo, A\c ores, Portugal.\\}

\end{center}
\bigskip
\noindent{*\bf Corresponding author.} \textit{t.guillerme@sheffield.ac.uk}\\ 
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\noindent (Keywords: )\\

\section{Abstract}
Objective: Showing which patterns are recovered by different functional diversity metrics under different frameworks, and how different metrics recover different signals from different processes linked to common mechanisms in ecology.

Approach: We simulate some one dimensional datasets using neutral evolutionary models (pure birth tree with brownian motion traits) then perturb the resulting dataset with some common ecological stressor mechanisms (environmental filtering, competitive exclusion, equalising fitness and facilitation).
We then measure different space occupancy metrics on the perturbed and unperturbed simulations, and check how the different metrics capture the resultant changes.

\section{Acknowledgements}
Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.\\
Cite grants

% \bibliographystyle{Science}
\bibliography{references}

\end{document}

### 2024-05-14 Manuscript/main.tex
transfered text tables and figures

@@
 
+\section{Introduction}
+
+In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses (@MammolaFuncEcol2021, @Palacio2022).
+Although numerous definitions exist (@Dawson), we here consider a trait to be any characteristic - e.g.
+, morphological, anatomical, ecological, physiological, behavioural, phenological—measured on individual organisms at any level, from genes to whole organisms.
+A trait-based approach has two advantages for answering core questions in ecology and evolution.
+First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
+Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g., unrelated species pools) and time (e.g., anatomical traits are comparable between palaeontology and ecology, which is not always the case with species) (@luza2023going).
+Already used routinely in palaeontology (@Raup1961, @Gould1991, @Foote1995, @GuillermeBiolLet2020), this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology (@McGill), biogeography (@Violle), conservation biology (@Chichorro2022), micro- (@Chapin), macro-evolution (@Wiens), and applied fields (e.g.
+agronomy @Martin2015).
+This is because trait-based approaches closely align with the general analytical framework proposed by @AnandCoenoses1994 to answer three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism) (Box 1).
+
+\break
+
+\hrule
+\section*{Box}
+@AnandCoenoses1994 conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism).
+This framework can be effectively applied to trait-based analyses:
+
+\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., “how?” and “why?” as defined below).
+Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; @ViolleOikos2007); ii) arranging these traits into some kind of trait space (@GuillermeBiolLet2020, @MammolaFuncEcol2021); and iii) using one or more unidimensional statistics to summarise properties of the trait space.
+For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g.
+ community weighted mean).
+For multidimensional spaces, these are usually named disparity metrics or indices (@GuillermeBiolLet2020) or functional diversity metrics (@MammolaFuncEcol2021), but all attempt to capture some pattern of interest in the trait space.
+
+\textit{How?} Process description can be seen as linking the pattern of interest (“What?”) to some dynamic element.
+This can be a punctual change.
+For example, change of the pattern under a certain condition, such as how traits X differ between two habitats (@Martinez).
+But it can also be one or more continuous or ordinal changes, such as how the pattern X changes in space, time and/or along ecological gradients (@Belmaker, @Jaryna, @Lamanna, @bjorkman, @McLean).
+The distinction here might seem trivial but it is significant: usually, the process designates the change of the pattern, not the change of the traits.
+Although the traits are what is really changing, researchers will usually analyse some emergent property of the trait aggregation as described by the change of statistics between two or more conditions.
+
+\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
+In most cases, this is what researchers are actually trying to understand.
+For example, one might be interested in understanding the effect of climate change on some trait (e.g. Does temperature affect leaf surface area?).
+In fact, most researchers work on understanding the causal link between variables (i.e. “why” are “what” and “how” linked).
+This can be very useful, for example, to provide predictions about the past or the future.
+Ultimately, studying the mechanisms (“why””) is often the reason why researchers get funded (or not).
+
+Note that the distinction between what, how and why is not categorical in its nature and is often nuanced, with patterns and process or process and mechanisms sometimes being used interchangeably to describe the same questions
+
+\hrule
+
+While the first part of this framework (“what?”) has been studied extensively in the last decade (@CITECITECITE), this was often undertaken outside of the “what, how, and why” framework.
+Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
+The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
+Using an inappropriate metric for describing a pattern can lead to biased conclusions.
+For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - “why?”) on an island compared to another island with only one of the species (the process - “how?”).
+In such a scenario, measuring the occupied area of each population on each island (a pattern - “what?”) will not be the most appropriate way to understand the potential competition between these populations.
+In this simplistic example, the functional overlap between populations (“what and how”) might be more appropriate.
+See for example Carvalho \& Cardoso (@ref) for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.
+
+Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure 1).
+We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands as patterns and processes capturing the effects of species extinction on species trait distributions.
+We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
+We suggest caution when summarising patterns in observed data and propose tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
+
+\section{Methods}
+\subsection{Simulating trait space patterns}
+
+First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.).
+This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (sensu @BausmanPhilSci2018).
+We refer to this as the “non-stressed trait space”.
+
+\subsubsection{Applying stressors to the trait space}
+
+We then applied five different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
+In removing the data, we applied these five specific algorithms (Figure 1; Table 1; all algorithms, except “Evenness”, were previously described in @GuillermeEcoEvo2020):
+
+\begin{itemize}
+\item Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data. This approximates our \textbf{null mechanism}.
+
+\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius)  $\rho$ of the centre of the trait-space.
+This approximates our \textbf{equalising mechanism}.
+
+\item Increasing density: by removing the required amount of pairs of points with a pair distance of at least $D$ (i.e.
+removing the n pairs of points that are at least $D$ distance away from each other).
+This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
+
+\item Shifting space: by removing the required amount of data from a distance (radius) $\rho$ of the element with the maximum value.
+This approximates our \textbf{filtering mechanism}.
+
+\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e.
+an estimated number of categories summarising the distribution, like in a histogram).
+Each bandwidth has an observed probability of sampling of n (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and p is a factor increasing the flattening/steepening (here we used $p=3$).
+This approximates our \textbf{competition mechanism}.
+\end{itemize}
+
+For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, Guillerme 2018, @GuillermeEcoEvo2020).
+These steps resulted in 4000 simulated trait spaces (200 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
+See Table 1 for a biological description of these stressors and Figure 1 for a visual description of them.
+
+Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
+For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
+However, if the data are distributed uniformly, this does not happen.
+Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
+However, this is not tackled here for simplicity.
+
+We repeated each simulation pipeline (generate a traitspace and apply the stressor) for uncorrelated 4 dimensional traits (figure 2) and uncorrelated 2, 3, 5 and 8 dimensional traits (supplementary materials).
+We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions) but also to avoid dealing with the curse of dimensionality.
+This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
+For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
+However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
+This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g.
+for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
+Note also that in our simulations, all dimensions have the same properties (i.e.
+same variance and distribution).
+This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.
+
+\begin{table}
+\center
+\scriptsize
+\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
+\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
+\hline
+  Null mechanism & Random removal (“random”) & No overall change in species traits or community structure except for the reduction in the number of elements.\\
+  Equalising fitness & Size change (“size”) & When species with trait combinations that are more extreme (i.e.
+species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
+\\
+  Facilitation & Density change (“density”) & When species with closely shared trait combinations (i.e.
+located close together in the trait space) are more likely to survive the stressor.
+In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
+  Environmental filtering & Position change (“position”) & When an increase in species trait similarity happens through some strong abiotic constraints (@Cornwell2006).
+For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e.
+some regions of the trait space become unsuitable).\\
+  Competitive exclusion & Evenness change (“evenness”) & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; @Hardin 1960).
+At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; @MacArthur1967)\\
+
+    \end{tabular}
+    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it.
+We also provide a more detailed description of the stressor.
+Note that in biological data, we don’t expect any of the mechanisms to be applied to the data alone (e.g.
+equalising fitness and facilitation can both act on the distribution of species traits).
+Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges of a trait distribution or/and through environmental filtering).
+These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
+}
+    \label{Tab:mechanisms}
+\end{table}
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
+\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
+A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting in a one dimensional trait space or two dimensional space.
+We then applied different stressors to the resulting trait space using the following:
+B) null mechanism: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
+C) equalising mechanism: removing species on the edge of the distribution;
+D) facilitation mechanism: removing species to reduce the distance between pairs of species (increasing density);
+E) filtering mechanism: removing species on one extreme of the distribution;
+F) competition mechanism: proportionally removing species on the centre of the distribution (“flattening the curve”)..}}
+\label{Fig:simulations}
+\end{figure}
+\bigskip
+
+\subsubsection{Metrics for measuring trait space occupancy dissimilarity/disparity:}
+
+We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (@Mammola2021):
+1) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in @GuillermeEcoEvo2020);
+2) Divergence, encompassing metrics reflecting the average difference among observations (6 metrics); and
+3) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in @GuillermeEcoEvo2020?).
+Note that we focused here on three packages fully devoted to functional diversity analyses (BAT, FD, and TPD) in the R statistical environments (see @MammolaFuncEcol2021).
+
+\begin{table}
+\center
+\scriptsize
+\begin{tabular}{p{0.1\linewidth}|p{0.15\linewidth}|p{0.15\linewidth}|p{0.2\linewidth}|p{0.15\linewidth}|p{0.1\linewidth}}
+
+\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
+\hline
+ & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & @Carmona2019\\
+ & \cr 
+ & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & @MammolaCardoso2020\\
+ & \cr 
+ & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & @Cardoso2015\\
+ & \cr 
+ & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & @Villeger2008\\
+\hline
+Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & Carmona2019\\
+ & \cr 
+ & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & @MammolaCardoso2020\\
+ & \cr 
+ & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & @Cardoso2015\\
+ & \cr  
+ & Distances & Melodic Mean Pairwise Distance & @@@NEEDS DESCRIPTION  & melodic.rao & @Botta-Dukat2005\\
+ & \cr 
+ & Distances & Melodic Rao & @@@NEEDS DESCRIPTION & melodic.rao & @Botta-Dukat2005\\
+ & \cr 
+ & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & Botta-Dukat2005\\
+\hline
+Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & @Carmona2019\\
+ & \cr 
+ & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & @MammolaCardoso2020\\
+ & \cr
+ & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & @Cornwell2006\\
+ & \cr
+ & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & @PetcheyGaston2002; @Cardoso2015\\
+\hline
+
+\end{tabular}
+\caption{\scriptsize{\textbf{Functional diversity metrics tested in our simulations.}}
+}
+\label{Tab:metrics}
+\end{table}
+
+\subsubsection{Scaling the results to be proportional to the random removals (null)}
+
+To ease interpretation of the results, we centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
+Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
+This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null, and 1 and –1 the maximum positive and negative deviations relative to the null.
+In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
+The results of these tests are displayed in table @@@ and in figure 2 (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
+
+\subsubsection{Measuring the effect of the strength of the stressor}
+
+Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score ~ amount of data removed.
+For each of these models, we reported the model fit (adjusted R2) and whether the slope was clearly distinguishable from 0 (p-values). See figure 2.
+
+\subsection{Empirical data}
+
+We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers @Walther2022
+A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE), and extant species list of the native community for each island was taken from @matthews along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain)
+We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
+We extracted traits for extant species from the AVONET database (@tobias)
+We used 9 morphological traits associated with dietary/foraging preference and dispersal ability @tobias @sheard to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather @tobias); and tail length
+Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.
+
+The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years
+Fifty-five species are known to have gone extinct prior to 1500 CE, and 26 after 1500 CE
+For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'')
+We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a traitspace reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a traitspace reduction of 69\%).
+
+To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets
+We selected the first 5 axes to represent the trait-spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
+We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).
+
+\section{Results}
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d.pdf}
+\caption{\scriptsize{\textbf{Simulation results:} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
+Negative and positive values signify a decrease/increase in the metric score.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value > 0.05).
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
+Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+Similar figures are available in the supplementary materials for 2, 3, 5 and 8 dimensions.}}
+\label{Fig:simulation_results}
+\end{figure}
+\bigskip
+
+The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.
+It ranged from metrics capturing no clear pattern (e.g.
+the Functional richness based on the hypervolume method) to metrics clearly capturing one specific pattern (e.g.
+Rao’s quadratic entropy based on pairwise distances capturing clear changes in the equalising mechanism but less clear changes for the other mechanisms).
+Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
+Functional richness using the convex hull method for equalizing) or non-linear ones (e.g. Functional richness using the probability density method for facilitation).
+This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
+Some cases resulted in an increase in metric score (positive values)!
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
+\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (blue = regularity, yellow = divergence, green = richness).
+Plain circles represent the metrics for the historic dataset (46\% removal), and fainted squares for the extant dataset (69\% removal).
+}}
+\label{Fig:simulation_results}
+\end{figure}
+\bigskip
+
+For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern.
+For example, all richness metrics (green; Figure 3) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
+However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g.
+Function dispersion based on dendrograms or hypervolumes).
+Finally, some metrics pick up a decrease for both datasets (e.g.
+Functional evenness based on hypervolumes).
+Note that this apparent global increase in most metric values may appear counterintuitive given that 46\% and 69\% of species went extinct over the two time periods.
+This is due to the scaling of the metrics compared to random equivalent extinctions.
+That is, Figure 3 is not displaying absolute changes in traitspace occupancy but rather changes in traitspace occupancy relative to a crude neutral model where all species are equally likely to go extinct.
+However, this is still an intriguing finding, as it indicates that, at least for certain metrics, the loss of Hawaiian bird FD is less than expected than if extinctions were random with respect to traits.
+This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct, and thus that FD loss is larger than through random extinctions (e.g., @Ferran2021 @Matthews2022).
+This could partly be due to the proportion of passerine extinctions in Hawaii.
+Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in FD).
+However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
+It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines – indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology @Walther2022 – but simply that this morphological variation is less than that observed across all birds.
+But these results – given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines –  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
+relative to random extinctions).
+@@@**add sentence about the large number of rail extinctions**.
+In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage), and different results may have been found if focusing exclusively on terrestrial birds.
+
+More broadly, as can be seen from the Hawaiian case study, patterns are likely to be much less clear with empirical data due to multiple stressors acting simultaneously.
+For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
+In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds @Samuel2011, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but rather genetic susceptibility (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range @book.
+
+\section{Discussion}
+
+We tested 14 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
+Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
+Therefore, the choice of traits space occupancy metric is essential to accurately describing a pattern of interest.
+
+Our results may be influenced by our choice of space occupancy metrics – specifically, we focused on metrics available in three statistical packages in R, the most common statistical language currently used in ecology (@cite), but many other ones could have been used.
+Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
+Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
+The simulated traitspaces in 2, 3, 4, 5 or 8 dimensions had the same variance on all dimensions (i.e.
+we used the same trait simulation process for all dimensions) whereas in empirical traitspaces, commonly generated using ordination techniques (e.g.
+PCA or PCO), the dimensions have by definition a decreasing variance.
+
+Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (@@@ examples).
+In a higher number of dimensions (usually >10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
+This sometimes makes metric interpretation more chaotic (@curse\_of\_dimensionality) in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
+In fact, the relationship between observations has an important effect on how our results scale with higher dimensionality (e.g.
+the volume of an evenly occupied space scales differently in high dimensions compared to the same space but with normally distributed observations).
+Therefore we advise workers to test assumptions specifically about their traitspace and “play” around with it before choosing their functional diversity/disparity framework and metric of interest.
+
+Another cautionary note could be made about our choice of scaling the metrics relative to a random traitspace reduction (with the same percentage of reduction).
+This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern.
+For example, if removing 80\% of the edges of a distribution (equalising mechanism) was distinguishable or not from removing 80\% of a distribution randomly (null mechanism).
+Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).
+In other words, our statistical question was “Does metric X distinguish between removing N \% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?”.
+This definition of null hypothesis is also appropriate for the empirical data if the question is the same.
+However, it is very likely that workers will ask a more exciting or intriguing question based on these data.
+For example, an interesting one could be “Are the extinctions uniform across all birds in Hawaii?".
+This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
+Here instead we asked the more colloquial question of  “Does metric X distinguish between removing N \% of data in a biassed way (the observed extinctions) and removing N \% of data randomly (the null)?” which we will argue is only a prerequisite question for answering the more interesting one highlighted above: identifying the metric(s) X that can capture the mechanism of interest.
+
+Following our results and these thoughts, we would like to encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
+We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g.
+using simulations in @treats or the @moms interactive package).
+Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
+For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
+facilitation + filtering; @ref), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; @ref)
+
+\section{Conclusion}
+Different ecological and evolutionary processes or mechanisms do not always lead to different patterns.
+Different observed patterns can originate from the same processes and mechanisms.
+Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of traitspace occupancy metric (disparity, dissimilarity or functional diversity metrics).
+Different metrics are designed to capture different aspects (@GuillermeEcoevo2020, @MammolaFuncEcol2021) but also perform better or worse at their designed task depending on the mechanism or process of interest.
+In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or based on a previous inspiring publication.
+
+\section{Repeatability and reproducibility}
+The results figures and tables are entirely reproducible via \url{https://github.com/TGuillerme/eco_metrics_simulations}.
+
 \section{Acknowledgements}

### 2024-05-14 Manuscript/main.tex
references input

@@
 \documentclass[12pt,letterpaper]{article}
+\usepackage{natbib}
 
@@
 
-In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses (@MammolaFuncEcol2021, @Palacio2022).
-Although numerous definitions exist (@Dawson), we here consider a trait to be any characteristic - e.g.
+In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}).
+Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any characteristic - e.g.
 , morphological, anatomical, ecological, physiological, behavioural, phenological—measured on individual organisms at any level, from genes to whole organisms.
@@
 First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
-Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g., unrelated species pools) and time (e.g., anatomical traits are comparable between palaeontology and ecology, which is not always the case with species) (@luza2023going).
-Already used routinely in palaeontology (@Raup1961, @Gould1991, @Foote1995, @GuillermeBiolLet2020), this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology (@McGill), biogeography (@Violle), conservation biology (@Chichorro2022), micro- (@Chapin), macro-evolution (@Wiens), and applied fields (e.g.
-agronomy @Martin2015).
-This is because trait-based approaches closely align with the general analytical framework proposed by @AnandCoenoses1994 to answer three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism) (Box 1).
+Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g.
+unrelated species pools) and time (e.g.
+anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
+Already used routinely in palaeontology \cite{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \cite{mcgill2006rebuilding}, biogeography \cite{violle2014emergence}, conservation biology \cite{chichorro2022trait}, micro- \cite{chapin1993evolution}, macro-evolution \cite{wiens2023trait} %TG: check to find another cite
+, and applied fields (e.g. agronomy \citealt{martin2015plant}).
+This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process), and \textit{``why?''} (understanding the mechanism) (Box 1).
 
-\break
+\bigskip
+\bigskip
+\hrule
+\hrule
 
-\hrule
 \section*{Box}
-@AnandCoenoses1994 conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism).
+\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process), and \textit{``why?''} (understanding the mechanism)
 This framework can be effectively applied to trait-based analyses:
 
-\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., “how?” and “why?” as defined below).
-Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; @ViolleOikos2007); ii) arranging these traits into some kind of trait space (@GuillermeBiolLet2020, @MammolaFuncEcol2021); and iii) using one or more unidimensional statistics to summarise properties of the trait space.
-For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g.
- community weighted mean).
-For multidimensional spaces, these are usually named disparity metrics or indices (@GuillermeBiolLet2020) or functional diversity metrics (@MammolaFuncEcol2021), but all attempt to capture some pattern of interest in the trait space.
+\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
+Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional statistics to summarise properties of the trait space.
+For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
+For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.
 
-\textit{How?} Process description can be seen as linking the pattern of interest (“What?”) to some dynamic element.
+\textit{How?} Process description can be seen as linking the pattern of interest (``What?'') to some dynamic element.
 This can be a punctual change.
-For example, change of the pattern under a certain condition, such as how traits X differ between two habitats (@Martinez).
-But it can also be one or more continuous or ordinal changes, such as how the pattern X changes in space, time and/or along ecological gradients (@Belmaker, @Jaryna, @Lamanna, @bjorkman, @McLean).
+For example, change of the pattern under a certain condition, such as how traits X differ between two habitats \citep{martinez2021habitat}.
+But it can also be one or more continuous or ordinal changes, such as how the pattern X changes in space, time and/or along ecological gradients \citep{belmaker2013spatial, jarzyna2018taxonomic,lamanna2014functional,bjorkman2018plant,mclean2021trait}.
 The distinction here might seem trivial but it is significant: usually, the process designates the change of the pattern, not the change of the traits.
@@
 In most cases, this is what researchers are actually trying to understand.
-For example, one might be interested in understanding the effect of climate change on some trait (e.g. Does temperature affect leaf surface area?).
-In fact, most researchers work on understanding the causal link between variables (i.e. “why” are “what” and “how” linked).
+For example, one might be interested in understanding the effect of climate change on some trait (e.g. Does temperature affect leaf surface area? @@@). %TG: to check
+In fact, most researchers work on understanding the causal link between variables (i.e. ``why'' are ``what'' and ``how'' linked).
 This can be very useful, for example, to provide predictions about the past or the future.
-Ultimately, studying the mechanisms (“why””) is often the reason why researchers get funded (or not).
+Ultimately, studying the mechanisms (``why'') is often the reason why researchers get funded (or not).
 
@@
 
+\bigskip
 \hrule
+\hrule
+\bigskip
 
-While the first part of this framework (“what?”) has been studied extensively in the last decade (@CITECITECITE), this was often undertaken outside of the “what, how, and why” framework.
+While the first part of this framework (``what?'') has been studied extensively in the last decade (@CITECITECITE), %TG: TODO
+ this was often undertaken outside of the ``what, how, and why'' framework.
 Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
@@
 Using an inappropriate metric for describing a pattern can lead to biased conclusions.
-For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - “why?”) on an island compared to another island with only one of the species (the process - “how?”).
-In such a scenario, measuring the occupied area of each population on each island (a pattern - “what?”) will not be the most appropriate way to understand the potential competition between these populations.
-In this simplistic example, the functional overlap between populations (“what and how”) might be more appropriate.
-See for example Carvalho \& Cardoso (@ref) for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.
+For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on an island compared to another island with only one of the species (the process - ``how?'').
+In such a scenario, measuring the occupied area of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
+In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
+See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.
 
-Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure 1).
+Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations}).
 We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands as patterns and processes capturing the effects of species extinction on species trait distributions.
@@
 
-First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.).
-This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (sensu @BausmanPhilSci2018).
-We refer to this as the “non-stressed trait space”.
+First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.) in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
+This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
+We refer to this as the ``non-stressed trait space''.
 
@@
 We then applied five different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
-In removing the data, we applied these five specific algorithms (Figure 1; Table 1; all algorithms, except “Evenness”, were previously described in @GuillermeEcoEvo2020):
+In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):
 
@@
 
-For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, Guillerme 2018, @GuillermeEcoEvo2020).
+For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
 These steps resulted in 4000 simulated trait spaces (200 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
-See Table 1 for a biological description of these stressors and Figure 1 for a visual description of them.
+See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.
 
@@
 
-We repeated each simulation pipeline (generate a traitspace and apply the stressor) for uncorrelated 4 dimensional traits (figure 2) and uncorrelated 2, 3, 5 and 8 dimensional traits (supplementary materials).
-We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions) but also to avoid dealing with the curse of dimensionality.
+We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2, 3, 5 and 8 dimensional traits (supplementary materials).
+We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
 This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
@@
 \hline
-  Null mechanism & Random removal (“random”) & No overall change in species traits or community structure except for the reduction in the number of elements.\\
-  Equalising fitness & Size change (“size”) & When species with trait combinations that are more extreme (i.e.
+  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of elements.\\
+  Equalising fitness & Size change (``size'') & When species with trait combinations that are more extreme (i.e.
 species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
 \\
-  Facilitation & Density change (“density”) & When species with closely shared trait combinations (i.e.
+  Facilitation & Density change (``density'') & When species with closely shared trait combinations (i.e.
 located close together in the trait space) are more likely to survive the stressor.
 In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
-  Environmental filtering & Position change (“position”) & When an increase in species trait similarity happens through some strong abiotic constraints (@Cornwell2006).
+  Environmental filtering & Position change (``position'') & When an increase in species trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}.
 For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e.
 some regions of the trait space become unsuitable).\\
-  Competitive exclusion & Evenness change (“evenness”) & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; @Hardin 1960).
-At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; @MacArthur1967)\\
+  Competitive exclusion & Evenness change (``evenness'') & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; \citealt{hardin1960competitive}).
+At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\
 
@@
 E) filtering mechanism: removing species on one extreme of the distribution;
-F) competition mechanism: proportionally removing species on the centre of the distribution (“flattening the curve”)..}}
+F) competition mechanism: proportionally removing species on the centre of the distribution (``flattening the curve'')..}}
 \label{Fig:simulations}
@@
 
-We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (@Mammola2021):
-1) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in @GuillermeEcoEvo2020);
+We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
+1) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting});
 2) Divergence, encompassing metrics reflecting the average difference among observations (6 metrics); and
 3) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in @GuillermeEcoEvo2020?).
-Note that we focused here on three packages fully devoted to functional diversity analyses (BAT, FD, and TPD) in the R statistical environments (see @MammolaFuncEcol2021).
+Note that we focused here on three packages fully devoted to functional diversity analyses (BAT, FD, and TPD) in the R statistical environments (see \citealt{mammola2021concepts}).
 
@@
 \hline
- & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & @Carmona2019\\
+ & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & \cite{carmona2019trait}\\
  & \cr 
- & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & @MammolaCardoso2020\\
+ & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
  & \cr 
- & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & @Cardoso2015\\
+ & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat}\\
  & \cr 
- & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & @Villeger2008\\
+ & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
 \hline
-Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & Carmona2019\\
+Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
  & \cr 
- & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & @MammolaCardoso2020\\
+ & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
  & \cr 
- & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & @Cardoso2015\\
+ & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
  & \cr  
- & Distances & Melodic Mean Pairwise Distance & @@@NEEDS DESCRIPTION  & melodic.rao & @Botta-Dukat2005\\
+ & Distances & Melodic Mean Pairwise Distance & @@@NEEDS DESCRIPTION  & melodic.rao & \cite{botta2005rao}\\
  & \cr 
@@
  & \cr 
- & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & Botta-Dukat2005\\
+ & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
 \hline
-Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & @Carmona2019\\
+Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
  & \cr 
- & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & @MammolaCardoso2020\\
+ & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
  & \cr
- & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & @Cornwell2006\\
+ & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
  & \cr
- & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & @PetcheyGaston2002; @Cardoso2015\\
+ & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
 \hline
@@
 In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
-The results of these tests are displayed in table @@@ and in figure 2 (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
+The results of these tests are displayed in table @@@ %TG: TODO: add table
+ and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
 
@@
 
-We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers @Walther2022
-A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE), and extant species list of the native community for each island was taken from @matthews along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain)
+We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
+A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE), and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
 We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
-We extracted traits for extant species from the AVONET database (@tobias)
-We used 9 morphological traits associated with dietary/foraging preference and dispersal ability @tobias @sheard to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather @tobias); and tail length
+We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}
+We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather; \citealt{tobias2022avonet}); and tail length
 Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.
 
-The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years
-Fifty-five species are known to have gone extinct prior to 1500 CE, and 26 after 1500 CE
-For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'')
-We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a traitspace reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a traitspace reduction of 69\%).
+The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
+Fifty-five species are known to have gone extinct prior to 1500 CE, and 26 after 1500 CE.
+For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
+We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a trait space reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a trait space reduction of 69\%).
 
-To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets
+To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets.
 We selected the first 5 axes to represent the trait-spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
@@
 }}
-\label{Fig:simulation_results}
+\label{Fig:empirical_results}
 \end{figure}
@@
 This is due to the scaling of the metrics compared to random equivalent extinctions.
-That is, Figure 3 is not displaying absolute changes in traitspace occupancy but rather changes in traitspace occupancy relative to a crude neutral model where all species are equally likely to go extinct.
+That is, Figure 3 is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.
+%TG: Move below to discussion?
 However, this is still an intriguing finding, as it indicates that, at least for certain metrics, the loss of Hawaiian bird FD is less than expected than if extinctions were random with respect to traits.
-This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct, and thus that FD loss is larger than through random extinctions (e.g., @Ferran2021 @Matthews2022).
+This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct, and thus that FD loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
 This could partly be due to the proportion of passerine extinctions in Hawaii.
@@
 However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
-It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines – indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology @Walther2022 – but simply that this morphological variation is less than that observed across all birds.
+It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines – indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} – but simply that this morphological variation is less than that observed across all birds.
 But these results – given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines –  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
 relative to random extinctions).
-@@@**add sentence about the large number of rail extinctions**.
+% @@@**add sentence about the large number of rail extinctions**.
 In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage), and different results may have been found if focusing exclusively on terrestrial birds.
@@
 More broadly, as can be seen from the Hawaiian case study, patterns are likely to be much less clear with empirical data due to multiple stressors acting simultaneously.
+%TG: Move above to discussion?
 For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
-In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds @Samuel2011, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but rather genetic susceptibility (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range @book.
+In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \cite{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but rather genetic susceptibility (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
 
@@
 
-Our results may be influenced by our choice of space occupancy metrics – specifically, we focused on metrics available in three statistical packages in R, the most common statistical language currently used in ecology (@cite), but many other ones could have been used.
+Our results may be influenced by our choice of space occupancy metrics – specifically, we focused on metrics available in three statistical packages in R, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many other ones could have been used.
 Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
 Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
-The simulated traitspaces in 2, 3, 4, 5 or 8 dimensions had the same variance on all dimensions (i.e.
-we used the same trait simulation process for all dimensions) whereas in empirical traitspaces, commonly generated using ordination techniques (e.g.
+The simulated trait spaces in 2, 3, 4, 5 or 8 dimensions had the same variance on all dimensions (i.e.
+we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g.
 PCA or PCO), the dimensions have by definition a decreasing variance.
@@
 the volume of an evenly occupied space scales differently in high dimensions compared to the same space but with normally distributed observations).
-Therefore we advise workers to test assumptions specifically about their traitspace and “play” around with it before choosing their functional diversity/disparity framework and metric of interest.
+Therefore we advise workers to test assumptions specifically about their trait space and ``play'' around with it before choosing their functional diversity/disparity framework and metric of interest.
 
-Another cautionary note could be made about our choice of scaling the metrics relative to a random traitspace reduction (with the same percentage of reduction).
-This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern.
+Another cautionary note could be made about our choice of scaling the metrics relative to a random trait space reduction (with the same percentage of reduction).
+This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern \cite{bausman2018modeling}.
 For example, if removing 80\% of the edges of a distribution (equalising mechanism) was distinguishable or not from removing 80\% of a distribution randomly (null mechanism).
 Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).
-In other words, our statistical question was “Does metric X distinguish between removing N \% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?”.
+In other words, our statistical question was ``Does metric X distinguish between removing N \% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?''.
 This definition of null hypothesis is also appropriate for the empirical data if the question is the same.
 However, it is very likely that workers will ask a more exciting or intriguing question based on these data.
-For example, an interesting one could be “Are the extinctions uniform across all birds in Hawaii?".
+For example, an interesting one could be ``Are the extinctions uniform across all birds in Hawaii?".
 This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
-Here instead we asked the more colloquial question of  “Does metric X distinguish between removing N \% of data in a biassed way (the observed extinctions) and removing N \% of data randomly (the null)?” which we will argue is only a prerequisite question for answering the more interesting one highlighted above: identifying the metric(s) X that can capture the mechanism of interest.
+Here instead we asked the more colloquial question of  ``Does metric X distinguish between removing N \% of data in a biassed way (the observed extinctions) and removing N \% of data randomly (the null)?'' which we will argue is only a prerequisite question for answering the more interesting one highlighted above: identifying the metric(s) X that can capture the mechanism of interest.
 
@@
 We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g.
-using simulations in @treats or the @moms interactive package).
+using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package) \citealt{guillerme2020shifting}.
 Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
 For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
-facilitation + filtering; @ref), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; @ref)
+facilitation + filtering; @ref), %TG: TODOL find a reference
+ or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
 
@@
 Different observed patterns can originate from the same processes and mechanisms.
-Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of traitspace occupancy metric (disparity, dissimilarity or functional diversity metrics).
-Different metrics are designed to capture different aspects (@GuillermeEcoevo2020, @MammolaFuncEcol2021) but also perform better or worse at their designed task depending on the mechanism or process of interest.
+Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (disparity, dissimilarity or functional diversity metrics).
+Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the mechanism or process of interest.
 In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or based on a previous inspiring publication.
@@
 
-% \bibliographystyle{Science}
+\bibliographystyle{sysbio}
 \bibliography{references}

### 2024-05-15 Analysis/04_Summarise_results.Rmd
Removed raos and reorder metrics

@@
 ## Re-ordering results for fancyness
-new_order <- c(14, 12, 3, 6,
-               15, 11, 2, 8,
-               7, 4, #5,
-               13,
-               10, 9, 1)
+new_order <- c(# Richness
+               13, 10, 9, 1,
+               # Divergence
+               15, 11, 2, #8,
+               #7,
+               4, #5,
+               # Regularity
+               14, 12, 3, 6)
+
 all_results_2d <- lapply(all_results_2d, sort.metrics, order = new_order)
@@
 fancy_names <- rev(
-               c("Functional evenness\nprobability density\nTPD::REND",
-                 "Functional evenness\nhypervolume\nBAT::kernel.evenness",
-                 "Functional evenness\ndendrogram\nBAT::evenness",
-                 "Functional evenness\ndistances\nFD::dbFD()$FEve",
+               c(
+                 "Functional richness\nprobability density\nTPD::REND",
+                 "Functional richness\nhypervolume\nBAT::kernel.alpha",
+                 "Functional richness\nconvex hull\nBAT::hull.alpha",
+                 "Alpha diversity\ndendrogram\nBAT::alpha",
+
                  "Divergence\nprobability density\nTPD::REND",             
@@
                  "Functional dispersion\ndendrogram\nBAT::dispersion",
-                 "Melodic Mean Pairwise Distance\ndistances\nmelodic.rao",
-                 "Melodic Rao\ndistances\nmelodic.rao",
+                 #"Melodic Mean Pairwise Distance\ndistances\nmelodic.rao",
+                 #"Melodic Rao\ndistances\nmelodic.rao",
                  "Rao's quadratic entropy\ndistances\nFD::dbFD()$RaoQ",
                  #"dissimilarity[FD]\ndivergence",
-                 "Functional richness\nprobability density\nTPD::REND",
-                 "Functional richness\nhypervolume\nBAT::kernel.alpha",
-                 "Functional richness\nconvex hull\nBAT::hull.alpha",
-                 "Alpha diversity\ndendrogram\nBAT::alpha"))
+
+                 "Functional evenness\nprobability density\nTPD::REND",
+                 "Functional evenness\nhypervolume\nBAT::kernel.evenness",
+                 "Functional evenness\ndendrogram\nBAT::evenness",
+                 "Functional evenness\ndistances\nFD::dbFD()$FEve"
+               ))
 ```
@@
 ```{r, fig.width = 8, fig.height = 10, echo = FALSE}
-col_vector <- c(rep("blue", 4), # The richness ones
-                rep("orange", 6), # The divergence ones
-                rep("darkgreen", 4)) # The regularity ones
+col_vector <- c(rep("darkgreen", 4), # The regularity ones
+                rep("orange", 4), # The divergence ones
+                rep("blue", 4)) # The richness ones
 ## The plot

### 2024-05-15 Manuscript/main.tex
Update main.tex

@@
 
-\noindent{\Large \bf The what, how, and why of trait-based analyses in ecology}
+\noindent{\Large \bf The what, how and why of trait-based analyses in ecology}
 \bigskip
@@
 \section{Abstract}
-Objective: Showing which patterns are recovered by different functional diversity metrics under different frameworks, and how different metrics recover different signals from different processes linked to common mechanisms in ecology.
+Objective: Showing which patterns are recovered by different functional diversity metrics under different frameworks and how different metrics recover different signals from different processes linked to common mechanisms in ecology.
 
 Approach: We simulate some one dimensional datasets using neutral evolutionary models (pure birth tree with brownian motion traits) then perturb the resulting dataset with some common ecological stressor mechanisms (environmental filtering, competitive exclusion, equalising fitness and facilitation).
-We then measure different space occupancy metrics on the perturbed and unperturbed simulations, and check how the different metrics capture the resultant changes.
+We then measure different space occupancy metrics on the perturbed and unperturbed simulations and check how the different metrics capture the resultant changes.
 
@@
 Already used routinely in palaeontology \cite{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \cite{mcgill2006rebuilding}, biogeography \cite{violle2014emergence}, conservation biology \cite{chichorro2022trait}, micro- \cite{chapin1993evolution}, macro-evolution \cite{wiens2023trait} %TG: check to find another cite
-, and applied fields (e.g. agronomy \citealt{martin2015plant}).
-This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process), and \textit{``why?''} (understanding the mechanism) (Box 1).
+ and applied fields (e.g. agronomy \citealt{martin2015plant}).
+This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (Box 1).
 
@@
 \section*{Box}
-\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process), and \textit{``why?''} (understanding the mechanism)
+\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism)
 This framework can be effectively applied to trait-based analyses:
@@
 In most cases, this is what researchers are actually trying to understand.
-For example, one might be interested in understanding the effect of climate change on some trait (e.g. Does temperature affect leaf surface area? @@@). %TG: to check
+For example, one might be interested in understanding the effect of climate change on some trait \citep{boonman2022trait}.
 In fact, most researchers work on understanding the causal link between variables (i.e. ``why'' are ``what'' and ``how'' linked).
@@
 While the first part of this framework (``what?'') has been studied extensively in the last decade (@CITECITECITE), %TG: TODO
- this was often undertaken outside of the ``what, how, and why'' framework.
+ this was often undertaken outside of the ``what, how and why'' framework.
 Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
@@
 Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations}).
-We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands as patterns and processes capturing the effects of species extinction on species trait distributions.
+We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands (process) resulting in a trait space modified by extinctions (the pattern) capturing the effects of species extinction on species trait distributions (the mechanism).
 We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
@@
 Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
-However, this is not tackled here for simplicity.
+However, this is not tackled here for both simplicity and to understand how they work in isolation.
 
-We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2, 3, 5 and 8 dimensional traits (supplementary materials).
+We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2 and 8 dimensional traits (supplementary materials).
 We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
@@
 for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
-Note also that in our simulations, all dimensions have the same properties (i.e.
-same variance and distribution).
+Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
 This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.
@@
 equalising fitness and facilitation can both act on the distribution of species traits).
-Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges of a trait distribution or/and through environmental filtering).
+Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution).
 These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
@@
 We then applied different stressors to the resulting trait space using the following:
-B) null mechanism: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
-C) equalising mechanism: removing species on the edge of the distribution;
-D) facilitation mechanism: removing species to reduce the distance between pairs of species (increasing density);
-E) filtering mechanism: removing species on one extreme of the distribution;
-F) competition mechanism: proportionally removing species on the centre of the distribution (``flattening the curve'')..}}
+B) \textbf{null mechanism}: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
+C) \textbf{equalising mechanism}: removing species on the edge of the distribution;
+D) \textbf{facilitation mechanism}: removing species to reduce the distance between pairs of species (increasing local density; decreasing evenness);
+E) \textbf{filtering mechanism}: removing species on one extreme of the distribution;
+F) \textbf{competition mechanism}: proportionally removing species on the centre of the distribution (``flattening the curve'').}}
 \label{Fig:simulations}
@@
 We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
-1) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting});
-2) Divergence, encompassing metrics reflecting the average difference among observations (6 metrics); and
-3) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in @GuillermeEcoEvo2020?).
-Note that we focused here on three packages fully devoted to functional diversity analyses (BAT, FD, and TPD) in the R statistical environments (see \citealt{mammola2021concepts}).
+1) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});
+2) Divergence, encompassing metrics reflecting the average difference among observations (4 metrics); and
+3) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting}).
+Note that we focused here on three packages fully devoted to functional diversity analyses (\texttt{BAT} \citealt{cardoso2015bat} \texttt{FD} \citealt{Laliberte2010FD} and \texttt{TPD} \citealt{carmona2019trait}) in the \texttt{R} statistical environments (see \citealt{mammola2021concepts}).
 
@@
  & \cr 
- & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat}\\
+ & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
  & \cr 
@@
  & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
- & \cr  
- & Distances & Melodic Mean Pairwise Distance & @@@NEEDS DESCRIPTION  & melodic.rao & \cite{botta2005rao}\\
- & \cr 
- & Distances & Melodic Rao & @@@NEEDS DESCRIPTION & melodic.rao & @Botta-Dukat2005\\
  & \cr  
@@
 Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
-This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null, and 1 and –1 the maximum positive and negative deviations relative to the null.
+This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
 In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
@@
 We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
-A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE), and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
+A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
 We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
@@
 The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
-Fifty-five species are known to have gone extinct prior to 1500 CE, and 26 after 1500 CE.
+Fifty-five species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
 For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
@@
 Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
-Similar figures are available in the supplementary materials for 2, 3, 5 and 8 dimensions.}}
+Similar figures are available in the supplementary materials for 2 and 8 dimensions.}}
 \label{Fig:simulation_results}
@@
 \caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (blue = regularity, yellow = divergence, green = richness).
-Plain circles represent the metrics for the historic dataset (46\% removal), and fainted squares for the extant dataset (69\% removal).
+Plain circles represent the metrics for the historic dataset (46\% removal) and fainted squares for the extant dataset (69\% removal).
 }}
@@
 However, this is still an intriguing finding, as it indicates that, at least for certain metrics, the loss of Hawaiian bird FD is less than expected than if extinctions were random with respect to traits.
-This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct, and thus that FD loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
+This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that FD loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
 This could partly be due to the proportion of passerine extinctions in Hawaii.
@@
 % @@@**add sentence about the large number of rail extinctions**.
-In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage), and different results may have been found if focusing exclusively on terrestrial birds.
+In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
 

### 2024-05-15 Manuscript/supplementaries.tex
Create supplementaries.tex

--- 2024-05-14 Manuscript/main.tex
@@
 
-\noindent (Keywords: )\\
+\section{Supplementary materials}
 
-\section{Abstract}
-Objective: Showing which patterns are recovered by different functional diversity metrics under different frameworks, and how different metrics recover different signals from different processes linked to common mechanisms in ecology.
-
-Approach: We simulate some one dimensional datasets using neutral evolutionary models (pure birth tree with brownian motion traits) then perturb the resulting dataset with some common ecological stressor mechanisms (environmental filtering, competitive exclusion, equalising fitness and facilitation).
-We then measure different space occupancy metrics on the perturbed and unperturbed simulations, and check how the different metrics capture the resultant changes.
-
-\section{Introduction}
-
-In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses (@MammolaFuncEcol2021, @Palacio2022).
-Although numerous definitions exist (@Dawson), we here consider a trait to be any characteristic - e.g.
-, morphological, anatomical, ecological, physiological, behavioural, phenological—measured on individual organisms at any level, from genes to whole organisms.
-A trait-based approach has two advantages for answering core questions in ecology and evolution.
-First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
-Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g., unrelated species pools) and time (e.g., anatomical traits are comparable between palaeontology and ecology, which is not always the case with species) (@luza2023going).
-Already used routinely in palaeontology (@Raup1961, @Gould1991, @Foote1995, @GuillermeBiolLet2020), this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology (@McGill), biogeography (@Violle), conservation biology (@Chichorro2022), micro- (@Chapin), macro-evolution (@Wiens), and applied fields (e.g.
-agronomy @Martin2015).
-This is because trait-based approaches closely align with the general analytical framework proposed by @AnandCoenoses1994 to answer three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism) (Box 1).
-
-\break
-
-\hrule
-\section*{Box}
-@AnandCoenoses1994 conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism).
-This framework can be effectively applied to trait-based analyses:
-
-\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., “how?” and “why?” as defined below).
-Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; @ViolleOikos2007); ii) arranging these traits into some kind of trait space (@GuillermeBiolLet2020, @MammolaFuncEcol2021); and iii) using one or more unidimensional statistics to summarise properties of the trait space.
-For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g.
- community weighted mean).
-For multidimensional spaces, these are usually named disparity metrics or indices (@GuillermeBiolLet2020) or functional diversity metrics (@MammolaFuncEcol2021), but all attempt to capture some pattern of interest in the trait space.
-
-\textit{How?} Process description can be seen as linking the pattern of interest (“What?”) to some dynamic element.
-This can be a punctual change.
-For example, change of the pattern under a certain condition, such as how traits X differ between two habitats (@Martinez).
-But it can also be one or more continuous or ordinal changes, such as how the pattern X changes in space, time and/or along ecological gradients (@Belmaker, @Jaryna, @Lamanna, @bjorkman, @McLean).
-The distinction here might seem trivial but it is significant: usually, the process designates the change of the pattern, not the change of the traits.
-Although the traits are what is really changing, researchers will usually analyse some emergent property of the trait aggregation as described by the change of statistics between two or more conditions.
-
-\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
-In most cases, this is what researchers are actually trying to understand.
-For example, one might be interested in understanding the effect of climate change on some trait (e.g. Does temperature affect leaf surface area?).
-In fact, most researchers work on understanding the causal link between variables (i.e. “why” are “what” and “how” linked).
-This can be very useful, for example, to provide predictions about the past or the future.
-Ultimately, studying the mechanisms (“why””) is often the reason why researchers get funded (or not).
-
-Note that the distinction between what, how and why is not categorical in its nature and is often nuanced, with patterns and process or process and mechanisms sometimes being used interchangeably to describe the same questions
-
-\hrule
-
-While the first part of this framework (“what?”) has been studied extensively in the last decade (@CITECITECITE), this was often undertaken outside of the “what, how, and why” framework.
-Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
-The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
-Using an inappropriate metric for describing a pattern can lead to biased conclusions.
-For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - “why?”) on an island compared to another island with only one of the species (the process - “how?”).
-In such a scenario, measuring the occupied area of each population on each island (a pattern - “what?”) will not be the most appropriate way to understand the potential competition between these populations.
-In this simplistic example, the functional overlap between populations (“what and how”) might be more appropriate.
-See for example Carvalho \& Cardoso (@ref) for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.
-
-Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure 1).
-We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands as patterns and processes capturing the effects of species extinction on species trait distributions.
-We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
-We suggest caution when summarising patterns in observed data and propose tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
-
-\section{Methods}
-\subsection{Simulating trait space patterns}
-
-First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.).
-This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (sensu @BausmanPhilSci2018).
-We refer to this as the “non-stressed trait space”.
-
-\subsubsection{Applying stressors to the trait space}
-
-We then applied five different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
-In removing the data, we applied these five specific algorithms (Figure 1; Table 1; all algorithms, except “Evenness”, were previously described in @GuillermeEcoEvo2020):
-
-\begin{itemize}
-\item Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data. This approximates our \textbf{null mechanism}.
-
-\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius)  $\rho$ of the centre of the trait-space.
-This approximates our \textbf{equalising mechanism}.
-
-\item Increasing density: by removing the required amount of pairs of points with a pair distance of at least $D$ (i.e.
-removing the n pairs of points that are at least $D$ distance away from each other).
-This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
-
-\item Shifting space: by removing the required amount of data from a distance (radius) $\rho$ of the element with the maximum value.
-This approximates our \textbf{filtering mechanism}.
-
-\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e.
-an estimated number of categories summarising the distribution, like in a histogram).
-Each bandwidth has an observed probability of sampling of n (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and p is a factor increasing the flattening/steepening (here we used $p=3$).
-This approximates our \textbf{competition mechanism}.
-\end{itemize}
-
-For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, Guillerme 2018, @GuillermeEcoEvo2020).
-These steps resulted in 4000 simulated trait spaces (200 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
-See Table 1 for a biological description of these stressors and Figure 1 for a visual description of them.
-
-Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
-For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
-However, if the data are distributed uniformly, this does not happen.
-Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
-However, this is not tackled here for simplicity.
-
-We repeated each simulation pipeline (generate a traitspace and apply the stressor) for uncorrelated 4 dimensional traits (figure 2) and uncorrelated 2, 3, 5 and 8 dimensional traits (supplementary materials).
-We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions) but also to avoid dealing with the curse of dimensionality.
-This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
-For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
-However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
-This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g.
-for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
-Note also that in our simulations, all dimensions have the same properties (i.e.
-same variance and distribution).
-This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.
-
-\begin{table}
-\center
-\scriptsize
-\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
-\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
-\hline
-  Null mechanism & Random removal (“random”) & No overall change in species traits or community structure except for the reduction in the number of elements.\\
-  Equalising fitness & Size change (“size”) & When species with trait combinations that are more extreme (i.e.
-species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
-\\
-  Facilitation & Density change (“density”) & When species with closely shared trait combinations (i.e.
-located close together in the trait space) are more likely to survive the stressor.
-In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
-  Environmental filtering & Position change (“position”) & When an increase in species trait similarity happens through some strong abiotic constraints (@Cornwell2006).
-For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e.
-some regions of the trait space become unsuitable).\\
-  Competitive exclusion & Evenness change (“evenness”) & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; @Hardin 1960).
-At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; @MacArthur1967)\\
-
-    \end{tabular}
-    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it.
-We also provide a more detailed description of the stressor.
-Note that in biological data, we don’t expect any of the mechanisms to be applied to the data alone (e.g.
-equalising fitness and facilitation can both act on the distribution of species traits).
-Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges of a trait distribution or/and through environmental filtering).
-These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
-}
-    \label{Tab:mechanisms}
-\end{table}
+\subsection{Supplementary results}
 
@@
 \centering
-   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
-\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
-A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting in a one dimensional trait space or two dimensional space.
-We then applied different stressors to the resulting trait space using the following:
-B) null mechanism: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
-C) equalising mechanism: removing species on the edge of the distribution;
-D) facilitation mechanism: removing species to reduce the distance between pairs of species (increasing density);
-E) filtering mechanism: removing species on one extreme of the distribution;
-F) competition mechanism: proportionally removing species on the centre of the distribution (“flattening the curve”)..}}
-\label{Fig:simulations}
-\end{figure}
-\bigskip
-
-\subsubsection{Metrics for measuring trait space occupancy dissimilarity/disparity:}
-
-We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (@Mammola2021):
-1) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in @GuillermeEcoEvo2020);
-2) Divergence, encompassing metrics reflecting the average difference among observations (6 metrics); and
-3) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in @GuillermeEcoEvo2020?).
-Note that we focused here on three packages fully devoted to functional diversity analyses (BAT, FD, and TPD) in the R statistical environments (see @MammolaFuncEcol2021).
-
-\begin{table}
-\center
-\scriptsize
-\begin{tabular}{p{0.1\linewidth}|p{0.15\linewidth}|p{0.15\linewidth}|p{0.2\linewidth}|p{0.15\linewidth}|p{0.1\linewidth}}
-
-\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
-\hline
- & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & @Carmona2019\\
- & \cr 
- & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & @MammolaCardoso2020\\
- & \cr 
- & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & @Cardoso2015\\
- & \cr 
- & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & @Villeger2008\\
-\hline
-Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & Carmona2019\\
- & \cr 
- & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & @MammolaCardoso2020\\
- & \cr 
- & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & @Cardoso2015\\
- & \cr  
- & Distances & Melodic Mean Pairwise Distance & @@@NEEDS DESCRIPTION  & melodic.rao & @Botta-Dukat2005\\
- & \cr 
- & Distances & Melodic Rao & @@@NEEDS DESCRIPTION & melodic.rao & @Botta-Dukat2005\\
- & \cr 
- & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & Botta-Dukat2005\\
-\hline
-Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & @Carmona2019\\
- & \cr 
- & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & @MammolaCardoso2020\\
- & \cr
- & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & @Cornwell2006\\
- & \cr
- & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & @PetcheyGaston2002; @Cardoso2015\\
-\hline
-
-\end{tabular}
-\caption{\scriptsize{\textbf{Functional diversity metrics tested in our simulations.}}
-}
-\label{Tab:metrics}
-\end{table}
-
-\subsubsection{Scaling the results to be proportional to the random removals (null)}
-
-To ease interpretation of the results, we centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
-Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
-This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null, and 1 and –1 the maximum positive and negative deviations relative to the null.
-In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
-The results of these tests are displayed in table @@@ and in figure 2 (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
-
-\subsubsection{Measuring the effect of the strength of the stressor}
-
-Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score ~ amount of data removed.
-For each of these models, we reported the model fit (adjusted R2) and whether the slope was clearly distinguishable from 0 (p-values). See figure 2.
-
-\subsection{Empirical data}
-
-We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers @Walther2022
-A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE), and extant species list of the native community for each island was taken from @matthews along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain)
-We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
-We extracted traits for extant species from the AVONET database (@tobias)
-We used 9 morphological traits associated with dietary/foraging preference and dispersal ability @tobias @sheard to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather @tobias); and tail length
-Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.
-
-The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years
-Fifty-five species are known to have gone extinct prior to 1500 CE, and 26 after 1500 CE
-For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'')
-We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a traitspace reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a traitspace reduction of 69\%).
-
-To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets
-We selected the first 5 axes to represent the trait-spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
-We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).
-
-\section{Results}
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d.pdf}
-\caption{\scriptsize{\textbf{Simulation results:} the y axes represent the different metrics tested (sorted by categories).
+   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_2d.pdf}
+\caption{\scriptsize{\textbf{Simulation results for 2 dimensions:} the y axes represent the different metrics tested (sorted by categories).
 The different columns represent the different stressors.
@@
 Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
-Similar figures are available in the supplementary materials for 2, 3, 5 and 8 dimensions.}}
-\label{Fig:simulation_results}
-\end{figure}
-\bigskip
-
-The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.
-It ranged from metrics capturing no clear pattern (e.g.
-the Functional richness based on the hypervolume method) to metrics clearly capturing one specific pattern (e.g.
-Rao’s quadratic entropy based on pairwise distances capturing clear changes in the equalising mechanism but less clear changes for the other mechanisms).
-Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
-Functional richness using the convex hull method for equalizing) or non-linear ones (e.g. Functional richness using the probability density method for facilitation).
-This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
-Some cases resulted in an increase in metric score (positive values)!
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
-\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (blue = regularity, yellow = divergence, green = richness).
-Plain circles represent the metrics for the historic dataset (46\% removal), and fainted squares for the extant dataset (69\% removal).
 }}
@@
 
-For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern.
-For example, all richness metrics (green; Figure 3) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
-However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g.
-Function dispersion based on dendrograms or hypervolumes).
-Finally, some metrics pick up a decrease for both datasets (e.g.
-Functional evenness based on hypervolumes).
-Note that this apparent global increase in most metric values may appear counterintuitive given that 46\% and 69\% of species went extinct over the two time periods.
-This is due to the scaling of the metrics compared to random equivalent extinctions.
-That is, Figure 3 is not displaying absolute changes in traitspace occupancy but rather changes in traitspace occupancy relative to a crude neutral model where all species are equally likely to go extinct.
-However, this is still an intriguing finding, as it indicates that, at least for certain metrics, the loss of Hawaiian bird FD is less than expected than if extinctions were random with respect to traits.
-This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct, and thus that FD loss is larger than through random extinctions (e.g., @Ferran2021 @Matthews2022).
-This could partly be due to the proportion of passerine extinctions in Hawaii.
-Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in FD).
-However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
-It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines – indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology @Walther2022 – but simply that this morphological variation is less than that observed across all birds.
-But these results – given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines –  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
-relative to random extinctions).
-@@@**add sentence about the large number of rail extinctions**.
-In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage), and different results may have been found if focusing exclusively on terrestrial birds.
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_8d.pdf}
+\caption{\scriptsize{\textbf{Simulation results for 8 dimensions:} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
+Negative and positive values signify a decrease/increase in the metric score.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value > 0.05).
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
+Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+}}
+\label{Fig:simulation_results}
+\end{figure}
+\bigskip
 
-More broadly, as can be seen from the Hawaiian case study, patterns are likely to be much less clear with empirical data due to multiple stressors acting simultaneously.
-For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
-In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds @Samuel2011, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but rather genetic susceptibility (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range @book.
-
-\section{Discussion}
-
-We tested 14 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
-Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
-Therefore, the choice of traits space occupancy metric is essential to accurately describing a pattern of interest.
-
-Our results may be influenced by our choice of space occupancy metrics – specifically, we focused on metrics available in three statistical packages in R, the most common statistical language currently used in ecology (@cite), but many other ones could have been used.
-Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
-Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
-The simulated traitspaces in 2, 3, 4, 5 or 8 dimensions had the same variance on all dimensions (i.e.
-we used the same trait simulation process for all dimensions) whereas in empirical traitspaces, commonly generated using ordination techniques (e.g.
-PCA or PCO), the dimensions have by definition a decreasing variance.
-
-Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (@@@ examples).
-In a higher number of dimensions (usually >10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
-This sometimes makes metric interpretation more chaotic (@curse\_of\_dimensionality) in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
-In fact, the relationship between observations has an important effect on how our results scale with higher dimensionality (e.g.
-the volume of an evenly occupied space scales differently in high dimensions compared to the same space but with normally distributed observations).
-Therefore we advise workers to test assumptions specifically about their traitspace and “play” around with it before choosing their functional diversity/disparity framework and metric of interest.
-
-Another cautionary note could be made about our choice of scaling the metrics relative to a random traitspace reduction (with the same percentage of reduction).
-This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern.
-For example, if removing 80\% of the edges of a distribution (equalising mechanism) was distinguishable or not from removing 80\% of a distribution randomly (null mechanism).
-Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).
-In other words, our statistical question was “Does metric X distinguish between removing N \% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?”.
-This definition of null hypothesis is also appropriate for the empirical data if the question is the same.
-However, it is very likely that workers will ask a more exciting or intriguing question based on these data.
-For example, an interesting one could be “Are the extinctions uniform across all birds in Hawaii?".
-This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
-Here instead we asked the more colloquial question of  “Does metric X distinguish between removing N \% of data in a biassed way (the observed extinctions) and removing N \% of data randomly (the null)?” which we will argue is only a prerequisite question for answering the more interesting one highlighted above: identifying the metric(s) X that can capture the mechanism of interest.
-
-Following our results and these thoughts, we would like to encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
-We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g.
-using simulations in @treats or the @moms interactive package).
-Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
-For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
-facilitation + filtering; @ref), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; @ref)
-
-\section{Conclusion}
-Different ecological and evolutionary processes or mechanisms do not always lead to different patterns.
-Different observed patterns can originate from the same processes and mechanisms.
-Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of traitspace occupancy metric (disparity, dissimilarity or functional diversity metrics).
-Different metrics are designed to capture different aspects (@GuillermeEcoevo2020, @MammolaFuncEcol2021) but also perform better or worse at their designed task depending on the mechanism or process of interest.
-In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or based on a previous inspiring publication.
-
-\section{Repeatability and reproducibility}
-The results figures and tables are entirely reproducible via \url{https://github.com/TGuillerme/eco_metrics_simulations}.
-
-\section{Acknowledgements}
-Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.\\
-Cite grants
-
-% \bibliographystyle{Science}
+\bibliographystyle{sysbio}
 \bibliography{references}

### 2024-05-15 Manuscript/main.tex
Update main.tex

@@
 For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
-We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a trait space reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a trait space reduction of 69\%).
+We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a trait space richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a trait space richness reduction of 69\%).
 

### 2024-05-17 Manuscript/main.tex
Update main.tex

@@
 
-In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}).
-Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any characteristic - e.g.
-, morphological, anatomical, ecological, physiological, behavioural, phenological—measured on individual organisms at any level, from genes to whole organisms.
+In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
+Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g.
+morphological, anatomical, ecological, physiological, behavioural, phenological, etc.) measured on individual organisms at any level, from genes to whole organisms.
 A trait-based approach has two advantages for answering core questions in ecology and evolution.
@@
 anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
-Already used routinely in palaeontology \cite{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \cite{mcgill2006rebuilding}, biogeography \cite{violle2014emergence}, conservation biology \cite{chichorro2022trait}, micro- \cite{chapin1993evolution}, macro-evolution \cite{wiens2023trait} %TG: check to find another cite
- and applied fields (e.g. agronomy \citealt{martin2015plant}).
-This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (Box 1).
+Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g.
+agronomy \citealt{martin2015plant}).
+This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).
 
@@
 
-\section*{Box}
-\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism)
+\section*{Box 1}
+\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
 This framework can be effectively applied to trait-based analyses:
@@
 
-Note that the distinction between what, how and why is not categorical in its nature and is often nuanced, with patterns and process or process and mechanisms sometimes being used interchangeably to describe the same questions
+Note that the distinction between what, how and why is not categorical in its nature and is often nuanced, with patterns and process or process and mechanisms sometimes being used interchangeably to describe the same questions.
 
@@
 
-While the first part of this framework (``what?'') has been studied extensively in the last decade (@CITECITECITE), %TG: TODO
- this was often undertaken outside of the ``what, how and why'' framework.
+While the first part of this framework (``what?'') has been studied extensively in the last decade this was often undertaken outside of the ``what, how and why'' framework.
 Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
@@
 Using an inappropriate metric for describing a pattern can lead to biased conclusions.
-For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on an island compared to another island with only one of the species (the process - ``how?'').
-In such a scenario, measuring the occupied area of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
+For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home two both populations and another one home to only one of the population (the process - ``how?'').
+In such a scenario, measuring the occupied area (i.e.
+$km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
 In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
@@
 
-We then applied five different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
+We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
 In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):
@@
 \item Increasing density: by removing the required amount of pairs of points with a pair distance of at least $D$ (i.e.
-removing the n pairs of points that are at least $D$ distance away from each other).
+removing the $n$ pairs of points that are at least $D$ distance away from each other).
 This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
@@
 an estimated number of categories summarising the distribution, like in a histogram).
-Each bandwidth has an observed probability of sampling of n (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and p is a factor increasing the flattening/steepening (here we used $p=3$).
+Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and $p$ is a factor increasing the flattening/steepening (here we used $p=3$).
 This approximates our \textbf{competition mechanism}.
@@
 For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
-These steps resulted in 4000 simulated trait spaces (200 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
+These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
+We used 171 replicates per simulation because that was empirically the smallest number of simulation required to reach a variance between simulation lower than 1\% across all metrics (i.e.
+any additional simulation added less that 1\% extra variance).
 See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.
@@
 \caption{\scriptsize{Illustration of the simulation protocol used in this paper.
-A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting in a one dimensional trait space or two dimensional space.
+A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting a trait space of 2, 4, or 8 dimensions.
 We then applied different stressors to the resulting trait space using the following:
@@
 We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
-1) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});
-2) Divergence, encompassing metrics reflecting the average difference among observations (4 metrics); and
+1) Richness, encompassing metrics reflecting the sum of differences among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});
+2) Divergence, encompassing metrics reflecting the average differences among observations (4 metrics); and
 3) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting}).
 Note that we focused here on three packages fully devoted to functional diversity analyses (\texttt{BAT} \citealt{cardoso2015bat} \texttt{FD} \citealt{Laliberte2010FD} and \texttt{TPD} \citealt{carmona2019trait}) in the \texttt{R} statistical environments (see \citealt{mammola2021concepts}).
+
+%  STOPPED HERE
 
@@
 \hline
- & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & \cite{carmona2019trait}\\
+ & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
+the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
  & \cr 
@@
  & \cr  
- & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
+ & Distances & Rao's quadratic entropy & The mean Euclidean distance between every two random pairs of species & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
 \hline
@@
 
-To ease interpretation of the results, we centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
+To ease interpretation of the results, w e centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
 Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
@@
 In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
-The results of these tests are displayed in table @@@ %TG: TODO: add table
- and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
+The results of these tests are displayed in supplementary tables 1 to 3 and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
 
@@
 A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
-We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
+We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss} data.
 We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}
@@
 It ranged from metrics capturing no clear pattern (e.g.
-the Functional richness based on the hypervolume method) to metrics clearly capturing one specific pattern (e.g.
-Rao’s quadratic entropy based on pairwise distances capturing clear changes in the equalising mechanism but less clear changes for the other mechanisms).
+the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g.
+Divergence based on probability density for the equalizing mechanism for a decrease in space occupancy).
 Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
-Functional richness using the convex hull method for equalizing) or non-linear ones (e.g. Functional richness using the probability density method for facilitation).
+the functional evenness based on hypervolume for and increase) or non-linear ones (e.g.
+Alpha diversity based on dendrogram for the equalizing mechanism).
 This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
-Some cases resulted in an increase in metric score (positive values)!
+Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results})!
 
@@
 That is, Figure 3 is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.
-%TG: Move below to discussion?
-However, this is still an intriguing finding, as it indicates that, at least for certain metrics, the loss of Hawaiian bird FD is less than expected than if extinctions were random with respect to traits.
-This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that FD loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
-This could partly be due to the proportion of passerine extinctions in Hawaii.
-Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in FD).
-However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
-It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines – indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} – but simply that this morphological variation is less than that observed across all birds.
-But these results – given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines –  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
-relative to random extinctions).
-% @@@**add sentence about the large number of rail extinctions**.
-In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
 
 More broadly, as can be seen from the Hawaiian case study, patterns are likely to be much less clear with empirical data due to multiple stressors acting simultaneously.
-%TG: Move above to discussion?
 For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
@@
 
-Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (@@@ examples).
+Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (e.g.
+in \citealt{van2023should}).
 In a higher number of dimensions (usually >10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
-This sometimes makes metric interpretation more chaotic (@curse\_of\_dimensionality) in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
+This sometimes makes metric interpretation more chaotic \cite{bellman1957dynamic} in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
 In fact, the relationship between observations has an important effect on how our results scale with higher dimensionality (e.g.
@@
 
+The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to traits (Figure \ref{Fig:empirical_results}).
+This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
+This could partly be due to the proportion of passerine extinctions in Hawaii.
+Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
+However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
+It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
+But these results, given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines,  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
+relative to random extinctions).
+In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
+
 Following our results and these thoughts, we would like to encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
@@
 For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
-facilitation + filtering; @ref), %TG: TODOL find a reference
- or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
+facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
 
@@
 Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the mechanism or process of interest.
-In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or based on a previous inspiring publication.
+Using space occupancy metrics for describing functional diversity is tricky.
+This is because the metrics have properties that can be counterintuitive based on the data at hand.
+And also because functions of ecosystems or organisms and niches within it are hard to capture/understand/define.
+This means that measuring functional diversity is based on both the definition of the metric and the function of interest and thus cannot be served by a one-fits-all metric.
+In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.
 
@@
 Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.\\
-Cite grants
+TG was funded by the UKRI-NERC grant NE/X016781/1.
 

### 2024-05-17 Manuscript/main.tex
Update main.tex

@@
 A trait-based approach has two advantages for answering core questions in ecology and evolution.
-First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
+First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than organisms themselves.
 Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g.
@@
 In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
-See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.
+See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.
 
@@
 
-\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius)  $\rho$ of the centre of the trait-space.
+\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius)  $\rho$ of the centre of the trait space.
 This approximates our \textbf{equalising mechanism}.
@@
 We also provide a more detailed description of the stressor.
-Note that in biological data, we don’t expect any of the mechanisms to be applied to the data alone (e.g.
+Note that in biological data, we don't expect any of the mechanisms to be applied to the data alone (e.g.
 equalising fitness and facilitation can both act on the distribution of species traits).
@@
 Note that we focused here on three packages fully devoted to functional diversity analyses (\texttt{BAT} \citealt{cardoso2015bat} \texttt{FD} \citealt{Laliberte2010FD} and \texttt{TPD} \citealt{carmona2019trait}) in the \texttt{R} statistical environments (see \citealt{mammola2021concepts}).
-
-%  STOPPED HERE
 
@@
 
-To ease interpretation of the results, w e centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
+To ease interpretation of the results, we centred the results for the 12 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
 Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
 This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
-In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
-The results of these tests are displayed in supplementary tables 1 to 3 and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
+In parallel, we also compared the distribution of the 171 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
+The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
 
@@
 Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score ~ amount of data removed.
-For each of these models, we reported the model fit (adjusted R2) and whether the slope was clearly distinguishable from 0 (p-values). See figure 2.
+For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables 1 to 3 and whether the slope was clearly distinguishable from 0 (Figure \ref{Fig:simulation_results}).
 
@@
 
-We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
-A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
-We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss} data.
-We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}
-We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather; \citealt{tobias2022avonet}); and tail length
+We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
+A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
+We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
+We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}.
+We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
 Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.
@@
 The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
-Fifty-five species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
+55 species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
 For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
@@
 
-To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets.
-We selected the first 5 axes to represent the trait-spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
+To build the community trait space, we undertook a PCA including all 118 species, which we then subset to calculate the trait space for the historic and extant datasets.
+We selected the first 5 axes to represent the trait spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
 We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).
+
+%  STOPPED HERE
 
@@
 
-We tested 14 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
+We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
 Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).

### 2024-05-20 Analysis/02_Simulation_pipeline.Rmd
Update 02_Simulation_pipeline.Rmd

@@
                            minimum = 20, maximum = 170,
-                           stop.variance = 0.0001,
+                           stop.variance = 0.01,
                            verbose = TRUE,

### 2024-05-20 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 
+\begin{landscape}
+\begin{table}
+\scriptsize
+\caption{Results of the model scaled metric $\sim$ removal level per stressor (2D)}
+\centering
+\begin{tabular}[t]{l|c|c|c|c|c|c|c|c}
+\hline
+  & equalizing slope & equalizing adj.$R^{2}$ & facilitation slope & facilitation adj.$R^{2}$ & filtering slope & filtering adj.$R^{2}$ & competition slope & competition adj.$R^{2}$\\
+\hline
+Functional evenness\\distances\\ \texttt{FD::dbFD()\$FEve} & 0.017*** & 0.02 & -0.015** & 0.012 & 0.026*** & 0.057 & -0.001 & -0.001\\
+\hline
+Functional evenness\\dendrogram\\ \texttt{BAT::evenness} & 0.014** & 0.011 & -0.016** & 0.014 & 0.024*** & 0.038 & -0.009* & 0.007\\
+\hline
+Functional evenness\\hypervolume\\ \texttt{BAT::kernel.evenness} & 0.007 & 0.002 & -0.006 & 0 & 0.03*** & 0.069 & 0.004 & 0\\
+\hline
+Functional evenness\\probability density\\ \texttt{TPD::REND} & 0.002 & -0.001 & -0.055*** & 0.07 & 0.001 & -0.001 & -0.02*** & 0.048\\
+\hline
+Rao's quadratic entropy\\distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.185*** & 0.578 & 0.022** & 0.009 & 0.102*** & 0.117 & 0.022** & 0.013\\
+\hline
+Functional dispersion\\dendrogram\\ \texttt{BAT::dispersion} & 0.08*** & 0.396 & 0.009. & 0.003 & 0.054*** & 0.195 & 0.008* & 0.005\\
+\hline
+Functional dispersion\\hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.018** & 0.009 & -0.023*** & 0.04 & 0.014* & 0.004 & -0.022*** & 0.048\\
+\hline
+Divergence\\probability density\\ \texttt{TPD::REND} & 0.066*** & 0.436 & -0.028*** & 0.046 & -0.009. & 0.004 & -0.022*** & 0.038\\
+\hline
+Alpha diversity\\dendrogram\\ \texttt{BAT::alpha} & 0.227*** & 0.572 & 0.044*** & 0.037 & 0.091*** & 0.155 & 0.002 & -0.001\\
+\hline
+Functional richness\\convex hull\\ \texttt{BAT::hull.alpha} & -0.09*** & 0.293 & -0.002 & -0.001 & -0.027*** & 0.043 & -0.011* & 0.008\\
+\hline
+Functional richness\\hypervolume\\ \texttt{BAT::kernel.alpha} & 0.018*** & 0.049 & 0.046*** & 0.179 & 0.004 & 0 & -0.012*** & 0.02\\
+\hline
+Functional richness\\probability density\\ \texttt{TPD::REND} & 0.09*** & 0.356 & 0.14*** & 0.28 & 0.022*** & 0.037 & -0.009* & 0.005\\
+\hline
+\end{tabular}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\scriptsize
+\caption{Results of the model scaled metric $\sim$ removal level per stressor (4D)}
+\centering
+\begin{tabular}[t]{l|c|c|c|c|c|c|c|c}
+\hline
+  & equalizing slope & equalizing adj.$R^{2}$ & facilitation slope & facilitation adj.$R^{2}$ & filtering slope & filtering adj.$R^{2}$ & competition slope & competition adj.$R^{2}$\\
+\hline
+Functional evenness\\distances\\ \texttt{FD::dbFD()\$FEve} & 0.054*** & 0.143 & -0.009 & 0.002 & 0.032*** & 0.076 & 0.002 & 0\\
+\hline
+Functional evenness\\dendrogram\\ \texttt{BAT::evenness} & 0.013* & 0.007 & -0.032*** & 0.092 & 0.015** & 0.012 & -0.002 & 0\\
+\hline
+Functional evenness\\hypervolume\\ \texttt{BAT::kernel.evenness} & 0.049*** & 0.122 & 0.008 & 0.002 & 0.061*** & 0.194 & 0.018*** & 0.038\\
+\hline
+Functional evenness\\probability\\ \texttt{TPD::REND} & 0.03*** & 0.135 & -0.01 & 0.002 & 0.024*** & 0.09 & 0.001 & -0.001\\
+\hline
+Rao's quadratic entropy\\distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.164*** & 0.539 & -0.02** & 0.012 & 0.039*** & 0.023 & -0.002 & -0.001\\
+\hline
+Functional dispersion\\dendrogram\\ \texttt{BAT::dispersion} & 0.108*** & 0.535 & -0.009** & 0.009 & 0.067*** & 0.258 & 0.012*** & 0.025\\
+\hline
+Functional dispersion\\hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.064*** & 0.145 & -0.004 & 0.002 & 0.045*** & 0.05 & 0.005* & 0.004\\
+\hline
+Divergence\\probability density\\ \texttt{TPD::REND} & 0.11*** & 0.52 & -0.03*** & 0.071 & 0.034*** & 0.065 & 0.004 & 0.001\\
+\hline
+Alpha diversity\\dendrogram\\ \texttt{BAT::alpha} & 0.153*** & 0.393 & 0.059*** & 0.083 & 0.014* & 0.005 & -0.035*** & 0.048\\
+\hline
+Functional richness\\convex hull\\ \texttt{BAT::hull.alpha} & -0.118*** & 0.517 & 0.013*** & 0.018 & -0.025*** & 0.06 & -0.002 & 0\\
+\hline
+Functional richness\\hypervolume\\ \texttt{BAT::kernel.alpha} & -0.008** & 0.009 & -0.015*** & 0.016 & -0.006 & 0.002 & -0.001 & -0.001\\
+\hline
+Functional richness\\probability density\\ \texttt{TPD::REND} & 0.016*** & 0.037 & 0.196*** & 0.535 & 0.014*** & 0.02 & -0.001 & -0.001\\
+\hline
+\end{tabular}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\scriptsize
+\caption{Results of the model scaled metric $\sim$ removal level per stressor (8D)}
+\centering
+\begin{tabular}[t]{l|c|c|c|c|c|c|c|c}
+\hline
+  & equalizing slope & equalizing adj.$R^{2}$ & facilitation slope & facilitation adj.$R^{2}$ & filtering slope & filtering adj.$R^{2}$ & competition slope & competition adj.$R^{2}$\\
+\hline
+Functional evenness\\distances\\ \texttt{FD::dbFD()\$FEve} & 0.039*** & 0.071 & 0.013** & 0.009 & 0.026*** & 0.051 & 0 & -0.001\\
+\hline
+Functional evenness\\dendrogram\\ \texttt{BAT::evenness} & -0.023*** & 0.041 & -0.038*** & 0.122 & -0.024*** & 0.102 & -0.008*** & 0.056\\
+\hline
+Functional evenness\\hypervolume\\ \texttt{BAT::kernel.evenness} & -0.055*** & 0.159 & -0.051*** & 0.121 & -0.034*** & 0.093 & -0.007*** & 0.015\\
+\hline
+Functional evenness\\probability\\ \texttt{TPD::REND} & -0.032*** & 0.134 & -0.118*** & 0.345 & -0.059*** & 0.185 & -0.008*** & 0.03\\
+\hline
+Rao's quadratic entropy\\distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.088*** & 0.157 & -0.02** & 0.01 & 0.013 & 0.002 & -0.012. & 0.004\\
+\hline
+Functional dispersion\\dendrogram\\ \texttt{BAT::dispersion} & 0.05*** & 0.166 & -0.022*** & 0.055 & 0.01. & 0.003 & -0.005 & 0.002\\
+\hline
+Functional dispersion\\hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.035*** & 0.035 & -0.004 & 0.001 & 0.028*** & 0.029 & -0.007** & 0.009\\
+\hline
+Divergence\\probability density\\ \texttt{TPD::REND} & 0.037*** & 0.121 & -0.029*** & 0.101 & -0.005 & 0 & -0.017*** & 0.041\\
+\hline
+Alpha diversity\\dendrogram\\ \texttt{BAT::alpha} & 0.03*** & 0.019 & 0.08*** & 0.131 & 0.041*** & 0.028 & -0.041*** & 0.05\\
+\hline
+Functional richness\\convex hull\\ \texttt{BAT::hull.alpha} & -0.077*** & 0.265 & 0.031*** & 0.09 & -0.009* & 0.006 & 0.002 & -0.001\\
+\hline
+Functional richness\\hypervolume\\ \texttt{BAT::kernel.alpha} & -0.019** & 0.013 & -0.023*** & 0.017 & -0.02*** & 0.018 & 0.002 & -0.001\\
+\hline
+Functional richness\\probability density\\ \texttt{TPD::REND} & 0.014** & 0.012 & 0.175*** & 0.574 & 0.08*** & 0.112 & -0.002 & -0.001\\
+\hline
+\end{tabular}
+\end{table}
+\end{landscape}
+
 \bibliographystyle{sysbio}

### 2024-05-20 Manuscript/main.tex
MS update done

@@
 
-%  STOPPED HERE
-
 \section{Results}
@@
 The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
-The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
-The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value > 0.05).
+The colours are here to visually separate the metrics by categories (green = richness, yellow = divergence, blue = regularity); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test $p$ value $> 0.05$).
 Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
-Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+Dashed thin grey lines represent non-significant models ($p$ value of slope or/and intercept $> 0.05$).
 Similar figures are available in the supplementary materials for 2 and 8 dimensions.}}
@@
 Divergence based on probability density for the equalizing mechanism for a decrease in space occupancy).
-Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
-the functional evenness based on hypervolume for and increase) or non-linear ones (e.g.
+Some metrics even captured an opposite pattern to what was expected: for all metrics, we naively expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
+the functional evenness based on hypervolume for the equalizing mechanism) or non-linear ones (e.g.
 Alpha diversity based on dendrogram for the equalizing mechanism).
@@
    \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
-\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (blue = regularity, yellow = divergence, green = richness).
+\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (green = richness, yellow = divergence, blue = regularity).
 Plain circles represent the metrics for the historic dataset (46\% removal) and fainted squares for the extant dataset (69\% removal).
@@
 For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern.
-For example, all richness metrics (green; Figure 3) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
+For example, all richness metrics (green; Figure \ref{Fig:empirical_results}) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
 However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g.
-Function dispersion based on dendrograms or hypervolumes).
-Finally, some metrics pick up a decrease for both datasets (e.g.
-Functional evenness based on hypervolumes).
-Note that this apparent global increase in most metric values may appear counterintuitive given that 46\% and 69\% of species went extinct over the two time periods.
+functional dispersion based on dendrograms or hypervolumes).
+Finally, some metrics pick up a decrease for both datasets (e.g. functional evenness based on hypervolumes - although the 95\% confidence intervals overlapped with 0).
+Note that this apparent global increase in most metric values may appear counter intuitive given that 46\% and 69\% of species went extinct over the two time periods.
 This is due to the scaling of the metrics compared to random equivalent extinctions.
-That is, Figure 3 is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.
+That is, Figure \ref{Fig:empirical_results} is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.
 
-More broadly, as can be seen from the Hawaiian case study, patterns are likely to be much less clear with empirical data due to multiple stressors acting simultaneously.
+More broadly, as can be seen from this empirical case, patterns are likely to be much less clear than with simulated data due to multiple stressors acting simultaneously.
 For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
@@
 
-Our results may be influenced by our choice of space occupancy metrics – specifically, we focused on metrics available in three statistical packages in R, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many other ones could have been used.
+Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many other ones could have been used (e.g.
+\citealt{guillerme2020shifting}).
 Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
 Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
-The simulated trait spaces in 2, 3, 4, 5 or 8 dimensions had the same variance on all dimensions (i.e.
+The simulated trait spaces in 2, 4 or 8 dimensions had the same variance on all dimensions (i.e.
 we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g.
@@
 Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).
-In other words, our statistical question was ``Does metric X distinguish between removing N \% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?''.
+In other words, our statistical question was ``Does metric X distinguish between removing N\% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?''.
 This definition of null hypothesis is also appropriate for the empirical data if the question is the same.
@@
 This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
-Here instead we asked the more colloquial question of  ``Does metric X distinguish between removing N \% of data in a biassed way (the observed extinctions) and removing N \% of data randomly (the null)?'' which we will argue is only a prerequisite question for answering the more interesting one highlighted above: identifying the metric(s) X that can capture the mechanism of interest.
+%@@@ TODO: add supplementary
 
 The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to traits (Figure \ref{Fig:empirical_results}).
-This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
+This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g.
+\citealt{sayol2021loss,Matthews2022}).
 This could partly be due to the proportion of passerine extinctions in Hawaii.
@@
 We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g.
-using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package) \citealt{guillerme2020shifting}.
+using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package \citealt{guillerme2020shifting}).
 Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
@@
 \section{Conclusion}
-Different ecological and evolutionary processes or mechanisms do not always lead to different patterns.
-Different observed patterns can originate from the same processes and mechanisms.
-Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (disparity, dissimilarity or functional diversity metrics).
-Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the mechanism or process of interest.
+Different ecological and evolutionary processes or mechanisms do not always results in different patterns.
+Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (\textit{aka} disparity, dissimilarity or functional diversity metrics).
+Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the data, the mechanism or/and process of interest.
 Using space occupancy metrics for describing functional diversity is tricky.
-This is because the metrics have properties that can be counterintuitive based on the data at hand.
+This is because the metrics have properties that can be counter intuitive based on the data at hand.
 And also because functions of ecosystems or organisms and niches within it are hard to capture/understand/define.

### 2024-05-24 Manuscript/main.tex
Added Tom's comments

@@
 Pedro Cardoso$^{2,3}$,
-Carlos P. Carmona$^{4}$,
-Maria Wagner J\o rgensen$^{5}$,
-Stefano Mammola$^{3,6,7}$,
-Thomas J. Matthews$^{5,8}$}\\
+% Carlos P. Carmona$^{4}$,
+Maria Wagner J\o rgensen$^{4}$,
+Stefano Mammola$^{3,5,6}$,
+Thomas J. Matthews$^{4,7}$}\\
 \noindent {\small \it 
@@
 $^3$Laboratory for Integrative Biodiversity Research (LIBRe), Finnish Museum of Natural History (Luomus), University of Helsinki, Helsinki, Finland\\
-$^4$Institute of Ecology and Earth Sciences, University of Tartu, Tartu, Estonia\\
-$^5$School of Geography, Earth and Environmental Sciences and Birmingham Institute of Forest Research, University of Birmingham, Birmingham B15 2TT, UK\\
-$^6$Molecular Ecology Group (MEG), Water Research Institute, National Research Council (CNR-IRSA), Verbania Pallanza, Italy\\
-$^7$National Biodiversity Future Center, Piazza Marina 61, 90133 Palermo, Italy\\
-$^8$Centre for Ecology, Evolution and Environmental Changes/Azorean Biodiversity Group / CHANGE – Global Change and Sustainability Institute and Universidade dos Açores – Faculty of Agricultural Sciences and Environment; PT-9700-042, Angra do Hero\'ismo, A\c ores, Portugal.\\}
+% $^4$Institute of Ecology and Earth Sciences, University of Tartu, Tartu, Estonia\\
+$^4$School of Geography, Earth and Environmental Sciences and Birmingham Institute of Forest Research, University of Birmingham, Birmingham B15 2TT, UK\\
+$^5$Molecular Ecology Group (MEG), Water Research Institute, National Research Council (CNR-IRSA), Verbania Pallanza, Italy\\
+$^6$National Biodiversity Future Center, Piazza Marina 61, 90133 Palermo, Italy\\
+$^7$Centre for Ecology, Evolution and Environmental Changes/Azorean Biodiversity Group / CHANGE – Global Change and Sustainability Institute and Universidade dos Açores – Faculty of Agricultural Sciences and Environment; PT-9700-042, Angra do Hero\'ismo, A\c ores, Portugal.\\}
 
@@
 Using an inappropriate metric for describing a pattern can lead to biased conclusions.
-For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home two both populations and another one home to only one of the population (the process - ``how?'').
+For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?'').
 In such a scenario, measuring the occupied area (i.e.
@@
 These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
-We used 171 replicates per simulation because that was empirically the smallest number of simulation required to reach a variance between simulation lower than 1\% across all metrics (i.e.
+We used 171 replicates per simulation because that was empirically the smallest number of simulations required to reach a variance between simulations lower than 1\% across all metrics (i.e.
 any additional simulation added less that 1\% extra variance).
@@
 
-Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score ~ amount of data removed.
+Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score $\sim$ amount of data removed.
 For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables 1 to 3 and whether the slope was clearly distinguishable from 0 (Figure \ref{Fig:simulation_results}).
@@
 For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
-We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a trait space richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a trait space richness reduction of 69\%).
+We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).
 
@@
 For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
-In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \cite{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but rather genetic susceptibility (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
+In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \citep{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but is related to genetics (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
 
@@
 in \citealt{van2023should}).
-In a higher number of dimensions (usually >10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
+In a higher number of dimensions (usually $>$10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
 This sometimes makes metric interpretation more chaotic \cite{bellman1957dynamic} in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
@@
 Another cautionary note could be made about our choice of scaling the metrics relative to a random trait space reduction (with the same percentage of reduction).
-This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern \cite{bausman2018modeling}.
+This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern \citep{bausman2018modeling}.
 For example, if removing 80\% of the edges of a distribution (equalising mechanism) was distinguishable or not from removing 80\% of a distribution randomly (null mechanism).
@@
 %@@@ TODO: add supplementary
+
+% TODO: also add the section suggested by Tom:
+% For the discussion with Pedro (you haven't attached the picture - but I think I know what you mean): yes, there is a weird behaviour with 60% vs.
+80% throughout most of the tests, I've tried to look into it whether it was a bug or not but could not find any bugged bits.
+I never thought of it in terms of null but yeah you folks reasoning actually makes a lot of sense.
+If you remove 100% of the data in a random way and then 100% in a biased way and then correct the results by the null (random), you should end up with no decrease at all.
+So I assume if you add data a little by little, the biased removal is at first not that different from the null one (e.g.
+same results wit 99% removed, 95, 90, etc...) and maybe only up until a certain threshold (maybe around 60% then?) do you see a clear effect of the bias (i.e.
+removing the data in a biased way is clearly different than randomly).
+This is actually totally worth adding somewhere in the manuscript!
 
@@
 relative to random extinctions).
-In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
+In addition, it is worth stressing that these analyses included marine species (which constitute less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
 
@@
 TG was funded by the UKRI-NERC grant NE/X016781/1.
+SM was supported by NBFC, funded by the Italian Ministry of University and Research, PNRR, Missione 4, Componente 2, ``Dalla ricerca all'impresa'', Investimento 1.4, Project CN00000033.
 

### 2024-05-31 Manuscript/main.tex
Update main.tex MWJ acknowledgements
Addedd ecol lett bits

@@
 \bigskip
-\bigskip
 \begin{center}
@@
 \bigskip
+
+\noindent RH: The what how and why of trait-based analyses
 
@@
 \noindent{*\bf Corresponding author.} \textit{t.guillerme@sheffield.ac.uk}\\
-\vspace{1in}
 
-%Line numbering
-\modulolinenumbers[1]
-\linenumbers
+\noindent (Keywords: Functional diversity, dissimilarity, disparity, patterns, processes, mechanisms)\\
+
+\begin{itemize}
+   \item Article type: Letter
+   \item Number of words in the abstract: @@@
+   \item Number of words in the main text: @@@
+   \item Number of words in the text box: @@@
+   \item Number of references: @@@
+   \item 3 Figures, 2 tables 1 text box @@@
+\end{itemize}
+
+\subsection{Statement of authorship}
+
+TG, PC, MWJ, SM and TJM designed the study, analysed the data and wrote the manuscript. TG wrote the code to analyse the data.
+
+\subsection{Data accessibility statement}
+
+All code and data required to reproduce this analysis is available on @@@
 
@@
 
-\noindent (Keywords: )\\
+\newpage
+%Line numbering
+\modulolinenumbers[1]
+\linenumbers
 
@@
 We then measure different space occupancy metrics on the perturbed and unperturbed simulations and check how the different metrics capture the resultant changes.
+
+\newpage
 
@@
 \section{Repeatability and reproducibility}
-The results figures and tables are entirely reproducible via \url{https://github.com/TGuillerme/eco_metrics_simulations}.
+The simulations, results figures and tables are entirely reproducible via \url{https://github.com/TGuillerme/eco_metrics_simulations}.
 
 \section{Acknowledgements}
-Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.\\
+Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.
 TG was funded by the UKRI-NERC grant NE/X016781/1.
 SM was supported by NBFC, funded by the Italian Ministry of University and Research, PNRR, Missione 4, Componente 2, ``Dalla ricerca all'impresa'', Investimento 1.4, Project CN00000033.
+MWJ was supported by NERC CENTA2 grant NE/S007350/1 and University of Birmingham.
 

### 2024-05-31 Manuscript/main.tex
Abstract 1st draft

@@
 \section{Abstract}
-Objective: Showing which patterns are recovered by different functional diversity metrics under different frameworks and how different metrics recover different signals from different processes linked to common mechanisms in ecology.
 
-Approach: We simulate some one dimensional datasets using neutral evolutionary models (pure birth tree with brownian motion traits) then perturb the resulting dataset with some common ecological stressor mechanisms (environmental filtering, competitive exclusion, equalising fitness and facilitation).
-We then measure different space occupancy metrics on the perturbed and unperturbed simulations and check how the different metrics capture the resultant changes.
+Functional diversity is now a commonly used statistic to describe populations and communities in ecology beyond specific richness.
+It is a very efficient tool allowing to summarise a complex space occupancy pattern (what is changing?) that can be used to understand changes in communities and/or populations (the process; how is it changing?) in response to some stressors (the mechanism; why is it changing?).
+However, although researchers have the choice between many different metrics (what) to study their questions of interest (how and why), it is often not directly clear which metric is more readily appropriate for which question.
+Here we show which patterns are recovered by different functional diversity metrics under different frameworks and how different metrics recover different signals from different processes linked to common mechanisms in ecology.
+Using both simulated data with different common ecological mechanisms approximations (environmental filtering, competitive exclusion, equalising fitness and facilitation) and an empirical dataset affected by more complex and nuanced mechanisms; we test the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
+We show that different metrics perform better for different tasks and discuss the importance of not using a one-fit-all metric but rather think and test whether a particular one will be good at capturing a pattern of interest.
 

### 2024-05-31 Manuscript/CoverLetter.tex
Create CoverLetter.tex

\documentclass[11pt]{letter}
\usepackage[a4paper,left=2.5cm, right=2.5cm, top=1cm, bottom=1cm]{geometry}
\usepackage{hyperref}
\usepackage[osf]{mathpazo}
\signature{Thomas Guillerme}
\address{School of Biosciences\\University of Sheffield\\Sheffield, S10 2TN\\United Kingdom\\t.guillerme@sheffield.ac.uk}
\longindentation=0pt
\begin{document}

\begin{letter}{}
\opening{Dear Editors,}

Since the seminal publication by Petchey and Gaston in Ecology Letters in 2002 on functional diversity (``Functional diversity (FD), species richness and community composition''; cumulating $>$ 2000 citations), functional diversity is now routinely used as a statistic to describe complex multidimensional datasets.
This has been driven by a really motivated community of ecologists proposing many different ways to measure functional diversity depending on different datasets or different questions.
More than 20 years later however, despite the enthusiasm of the ecology and evolution community there is no consensual way on which metric to use and in which scenarios they perform better.
This has led to the positive and rich paradigm of a diverse number of metrics routinely used for different tasks.

However, this richness of ways to measure functional diversity can make it daunting and or confusing for researchers that want to add functional diversity to their research methodology toolkit.
We argue that this confusions can stem from both an epistemological misunderstanding as well as practical hurdles.
The second is nicely alleviated by a dynamic community of researchers proposing open source and easy to use implementations to measure functional diveristy.
But the first hurdle remains.
Indeed, although it is often rather clear how functional diversity in general can capture changes communities (investigating processes; answering the question ``how are communities changing?'') and relate them to general understanding in ecology (understanding mechanisms; ``why are communities changing?''); surprisingly little work has been published in the last few years on how the metrics used to measure functional diversity can accurately measure complex multidimensional patterns (capturing patterns; ``what is changing?'').

Here we argue that a better understanding of the ``capturability'' of a pattern should be taken more into account for functional diversity to move forward as a great tool in ecology and evolution.
Through simulated and empirical data we show that the choice of the functional diversity metric (what) can lead to often very different interpretations of the same processes and mechanisms (how and why).
For example, in our empirical data looking at endemic birds extinctions in Hawaii we show that different functional diversity metrics can lead to widely different interpretations of the actual biological mechanisms changing Hawaiian bird communities.
I.e. with the same dataset and a drastic rate of extinction for endemic birds, different metrics can either show a clear decrease in functional diversity (expected) but also a clear increase (unexpected).
We have written this manuscript with an emphasis on the didactic side of the question and we hope it will be of great interest and usefulness to the many readers of Ecology Letters using functional diversity in their research.

I look forward to hearing from you soon.

\closing{Yours sincerely,\\On behalf of my co-authors}

\end{letter}
\end{document}

### 2024-05-31 Manuscript/CoverLetter.tex
README INIT

@@
 
-Here we argue that a better understanding of the ``capturability'' of a pattern should be taken more into account for functional diversity to move forward as a great tool in ecology and evolution.
+Here, in \textit{``The what, how and why of trait-based analyses in ecology''} we argue that a better understanding of the ``capturability'' of a pattern should be taken more into account for functional diversity to move forward as a great tool in ecology and evolution.
 Through simulated and empirical data we show that the choice of the functional diversity metric (what) can lead to often very different interpretations of the same processes and mechanisms (how and why).

### 2024-06-01 Manuscript/main.tex
main_PC.tex

--- 2024-05-31 Manuscript/main.tex
@@
 
-\noindent RH: The what how and why of trait-based analyses
+\noindent RH: The what, how and why of trait-based analyses
 
@@
 \section{Abstract}
-Objective: Showing which patterns are recovered by different functional diversity metrics under different frameworks and how different metrics recover different signals from different processes linked to common mechanisms in ecology.
 
-Approach: We simulate some one dimensional datasets using neutral evolutionary models (pure birth tree with brownian motion traits) then perturb the resulting dataset with some common ecological stressor mechanisms (environmental filtering, competitive exclusion, equalising fitness and facilitation).
-We then measure different space occupancy metrics on the perturbed and unperturbed simulations and check how the different metrics capture the resultant changes.
+Functional diversity is commonly used to describe populations and communities in ecology beyond taxonomic diversity.
+It allows summarising complex space  or time-occupancy patterns (what is changing?) that can be used to understand changes in communities and/or populations (the process; how is it changing?) in response to some stressors (the mechanism; why is it changing?).
+Although researchers have the choice between many different metrics (what) to study their questions of interest (how and why), it is often not directly clear which metric is more readily appropriate for which question.
+Here we show which patterns are recovered by multiple functional diversity metrics under different frameworks and how different metrics recover contrasting signals from different processes linked to common mechanisms in ecology.
+Using both simulated data with common ecological mechanism approximations (environmental filtering, competitive exclusion, equalising fitness and facilitation) and an empirical dataset affected by more complex and nuanced mechanisms, we test the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
+We show that metrics perform differently for many tasks and discuss the importance of not using a one-fit-all metric but rather think and test whether a particular one will be good at capturing a pattern of interest.
 
@@
 
-In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
+In the last two decades, there has been a progressive expansion in ecology and evolution studies from taxonomic-oriented approaches, with species as the focal point, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
 Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g.
@@
 A trait-based approach has two advantages for answering core questions in ecology and evolution.
-First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than organisms themselves.
+First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than the organisms themselves.
 Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g.
@@
 agronomy \citealt{martin2015plant}).
-This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).
+This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).
 
@@
 \section*{Box 1}
-\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
+\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
 This framework can be effectively applied to trait-based analyses:
@@
 \textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
-Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional statistics to summarise properties of the trait space.
-For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
+Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space.
+For unidimensional spaces (i.e.
+distributions), these statistics can be as simple as the mean and the standard deviation (e.g.
+community weighted mean).
 For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.
@@
 
-\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
-In most cases, this is what researchers are actually trying to understand.
+\textit{Why?} Mechanism description is then at the core of answering the biological question at hand. In most cases, this is what researchers are actually trying to understand.
 For example, one might be interested in understanding the effect of climate change on some trait \citep{boonman2022trait}.
-In fact, most researchers work on understanding the causal link between variables (i.e. ``why'' are ``what'' and ``how'' linked).
+In fact, most researchers work on understanding the causal link between variables (i.e.
+``why'' are ``what'' and ``how'' linked).
 This can be very useful, for example, to provide predictions about the past or the future.
@@
 
-While the first part of this framework (``what?'') has been studied extensively in the last decade this was often undertaken outside of the ``what, how and why'' framework.
+While the first part of this framework (``what?'') has been studied extensively in the last decade, this was often undertaken outside of the ``what, how and why'' framework.
 Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
@@
 We also provide a more detailed description of the stressor.
-Note that in biological data, we don't expect any of the mechanisms to be applied to the data alone (e.g.
+Note that in biological data, we don't expect any of the mechanisms to act on communities alone (e.g.
 equalising fitness and facilitation can both act on the distribution of species traits).
@@
 \hline
- & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
-the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
+Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
  & \cr 
- & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
+ & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
  & \cr
- & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
+ & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
  & \cr
- & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
+ & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
 \hline
@@
 \hline
-Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
+ & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
+the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
  & \cr 
- & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
+ & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
  & \cr 
- & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
+ & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
  & \cr 
- & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
+ & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
 \hline
@@
 55 species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
-For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
+For our analyses, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
 We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).
@@
 Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
-Therefore, the choice of traits space occupancy metric is essential to accurately describing a pattern of interest.
+Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
 
-Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many other ones could have been used (e.g.
+Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many others could have been used (e.g.
 \citealt{guillerme2020shifting}).
@@
 
-Following our results and these thoughts, we would like to encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
-We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g.
+Following our results and these thoughts, we encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
+We encourage workers to test their metrics using simplified example datasets to see if they have the ability to capture at least crude changes as exemplified in this work (e.g.
 using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package \citealt{guillerme2020shifting}).

### 2024-06-03 Manuscript/main.tex
Update main.tex
Updates
-Counts
-Abstract
-Caveat from Tom

@@
    \item Article type: Letter
-   \item Number of words in the abstract: @@@
-   \item Number of words in the main text: @@@
-   \item Number of words in the text box: @@@
-   \item Number of references: @@@
-   \item 3 Figures, 2 tables 1 text box @@@
+   \item Number of words in the abstract: 189
+   \item Number of words in the main text: 3949
+   \item Number of words in the text box: 419
+   \item Number of references: 54
+   \item 3 Figures, 2 tables and 1 text box
 \end{itemize}
@@
 
-Functional diversity is commonly used to describe populations and communities in ecology beyond taxonomic diversity.
-It allows summarising complex space  or time-occupancy patterns (what is changing?) that can be used to understand changes in communities and/or populations (the process; how is it changing?) in response to some stressors (the mechanism; why is it changing?).
-Although researchers have the choice between many different metrics (what) to study their questions of interest (how and why), it is often not directly clear which metric is more readily appropriate for which question.
-Here we show which patterns are recovered by multiple functional diversity metrics under different frameworks and how different metrics recover contrasting signals from different processes linked to common mechanisms in ecology.
-Using both simulated data with common ecological mechanism approximations (environmental filtering, competitive exclusion, equalising fitness and facilitation) and an empirical dataset affected by more complex and nuanced mechanisms, we test the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
-We show that metrics perform differently for many tasks and discuss the importance of not using a one-fit-all metric but rather think and test whether a particular one will be good at capturing a pattern of interest.
+Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
+Indeed, functional diversity metrics allow researchers to summarize complex occupancy patterns in space and/or time (what is changing?) that lead to changes in communities and/or populations (the process; how is it changing?) in response to some stressors (the mechanism; why is it changing?).
+However, as the diversity of functional diversity metrics and methods increases, it is often not directly clear which metric is more readily appropriate for which question.
+We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms (environmental filtering, competitive exclusion, equalizing fitness, and facilitation) in community ecology.
+Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
+We show that different metrics perform better for different tasks, emphasizing the importance of not using a one-size-fits-all metric.
+Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.
 
@@
 Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space.
-For unidimensional spaces (i.e.
-distributions), these statistics can be as simple as the mean and the standard deviation (e.g.
-community weighted mean).
+For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
 For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.
@@
 
-\textit{Why?} Mechanism description is then at the core of answering the biological question at hand. In most cases, this is what researchers are actually trying to understand.
+\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
+In most cases, this is what researchers are actually trying to understand.
 For example, one might be interested in understanding the effect of climate change on some trait \citep{boonman2022trait}.
-In fact, most researchers work on understanding the causal link between variables (i.e.
-``why'' are ``what'' and ``how'' linked).
+In fact, most researchers work on understanding the causal link between variables (i.e. ``why'' are ``what'' and ``how'' linked).
 This can be very useful, for example, to provide predictions about the past or the future.
@@
 \begin{itemize}
-\item Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data. This approximates our \textbf{null mechanism}.
+\item Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data.
+This approximates our \textbf{null mechanism}.
 
@@
 %@@@ TODO: add supplementary
-
-% TODO: also add the section suggested by Tom:
-% For the discussion with Pedro (you haven't attached the picture - but I think I know what you mean): yes, there is a weird behaviour with 60% vs.
-80% throughout most of the tests, I've tried to look into it whether it was a bug or not but could not find any bugged bits.
-I never thought of it in terms of null but yeah you folks reasoning actually makes a lot of sense.
-If you remove 100% of the data in a random way and then 100% in a biased way and then correct the results by the null (random), you should end up with no decrease at all.
-So I assume if you add data a little by little, the biased removal is at first not that different from the null one (e.g.
-same results wit 99% removed, 95, 90, etc...) and maybe only up until a certain threshold (maybe around 60% then?) do you see a clear effect of the bias (i.e.
-removing the data in a biased way is clearly different than randomly).
-This is actually totally worth adding somewhere in the manuscript!
+Furthermore, the scaling of the metrics paired with the levels of removal (80\%, 60\%, 40\% and 20\%) creates an expected artifact similarity when comparing the random removals against the other ones: when less data is available overall, the metrics are more likely to be similar (e.g.
+if removing 99\% of the data, we expect the random and removal metric score to be nearly identical to the non-random ones).
 

### 2024-06-03 Analysis/01_Simulating_spaces_and_stressors.Rmd
Cleaned up analysis folder

@@
 
-We can use the package `dads` that's in development for simulating the trait spaces (note that the package is still pretty unstable but the simple tasks here should not change).
+We can use the package `treats` that's in development for simulating the trait spaces (note that the package is still pretty unstable but the simple tasks here should not change).
 Later, we can use the `dispRity::reduce.space` package to apply stressors to the data.
@@
 
-## Paper figure
+## Paper figure 1
 
-```{r}
+```{r, eval = FALSE}
 plot.text.box <- function(my_text, ...) {

### 2024-06-03 Analysis/02_Simulation_pipeline.Rmd
Cleaned up analysis folder

@@
 source("../Functions/simulation.pipeline.R")
+source("../Functions/analyse.replicates.R")
 source("../Functions/dbFD.R")
@@
 
-There are options to
-
 ```{r, eval = TRUE}
@@
 ```
+
 > Note that you can use the parallel version by adding the argument `parallel = TRUE` that is automatised using the package `future` and `future.apply` which makes the whole process run on all your available cores but is way less verbose.
 Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.
+
+You can run this pipeline on a cluster using/modifying the `cluster_script_example.Rscript` file in `Analysis/`.
 

### 2024-06-03 Analysis/03_Empirical_pipeline.Rmd
Cleaned up analysis folder

--- 2024-01-18 Analysis/03-Empirical_pipeline.rmd
@@
 source("../Functions/simulation.pipeline.R")
+source("../Functions/analyse.replicates.R")
+source("../Functions/dbFD.R")
 ```
@@
 
-You can run the pipeline on the empirical data if you have the following elements ready:
+To run the simulations on an empirical data set you'll need to:
 
- * A trait space (a `"matrix"`) with the rows as species and the columns as dimensions (make sure it has rownames for avoiding any problems down the line).
- * A list of species to remove for each sub-sampling the list should be list of named `logical` vectors (where for example species that should be excluded - e.g.
-fossils are marked as `FALSE` and the ones to include as `TRUE`).
+ 1. load an empirical traitspace
+ 2. load a list of species to remove
+ 3. generate the pipeline as for the simulated data
 
-Here are two dummy examples but you can replace them with your empirical data:
+Here's the pipeline applied to the Hawaiian birds dataset/
 
 ```{r}
-## Making a dummy trait space for 200 species (3D)
-my_traitspace <- dispRity::space.maker(elements = 200, dimensions = 3,
-                                       distribution = rnorm,
-                                       elements.names = 1:200)
-## This is what it looks like
-head(my_traitspace)
+## global trait space 
+empirical_traitspace <- read.csv('../Data/Raw/trait-space-5d-prehistoric-community.csv', row.names = 1)
 
-## Making a dummy set of sub-sampling
-my_subsamples <- list(sample(c(TRUE, FALSE), 200, replace = TRUE),
-                      sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.75, 0.25)),
-                      sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.90, 0.10)))
-## Adding the elements names
-my_subsamples <- lapply(my_subsamples, function(x) {names(x) <- 1:200 ; return(x)})
+## Species lists
+historic_species <- rownames(read.csv('../Data/Raw/historic-species-list-minus-a-flammeus.csv', row.names = 1))
+extant_species <- rownames(read.csv('../Data/Raw/present-species-list-minus-a-flammeus.csv', row.names = 1))
 
-## This is what it looks like
-str(my_subsamples)
+## Traitspace groups
+extant_group <- rownames(empirical_traitspace) %in% extant_species ## 37 species present
+historic_group <- rownames(empirical_traitspace) %in% historic_species ## 63 species present
+groups <- list("extant" = extant_group, "historic" = historic_group)
 ```
@@
 ## Creating the function for running the empirical pipeline
-empirical.pipeline <- make.simulation.pipeline(sim.data = my_traitspace,
-                                               remove   = my_subsamples)
+empirical.pipeline <- make.simulation.pipeline(sim.data = empirical_traitspace,
+                                               remove = groups)
 ```
@@
 Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.
-
-## Checking the results
-
-We can then visualise the "convergence" of the results using the `analyse.replicate` function:
-
-```{r, eval = FALSE}
-source("../Functions/analyse.replicates.R")
-## Check the change in variance for all 120 parameters
-analyse.replicates("../Data/Processed/sim_filtering.rda") #140 -> #170
-analyse.replicates("../Data/Processed/sim_equalizing.rda") #155
-analyse.replicates("../Data/Processed/sim_competition.rda") #126
-analyse.replicates("../Data/Processed/sim_facilitation.rda") #163
-analyse.replicates("../Data/Processed/empirical_data.rda") #163
-```

### 2024-06-03 Analysis/04_Summarise_results.Rmd
Cleaned up analysis folder

@@
 ```
-
-# Results interpretations in the light of our expectation table
-
-These were our expectations:
-
-Mechanism | Richness| Dispersion | Regularity
--------------|-------|-------|-------|
-Equalizing   | Lower | Lower | Higher|
-Facilitation | Higher| Higher| Higher|
-Filtering (exclusion) | Higher| Higher| Higher|
-Competition  | Lower | Lower |Nothing|
-
-And these are our results per metric family:
-
- * [ok] prediction is correct (lower = decrease; higher = increase)
- * [-ok] prediction is kinda correct (there is a change compared to the null but not in the right direction)
- * [NO] to a wrong prediction: no change when change is expected or change when no change is expected
-
-> NOTE: the difference between [ok] and [-ok] is actually not really meaningful here since these graphs shows differences compared to the null results so if both the null and stressor results increase but the stressor increases slowlier than the stressor, the results in the table will decrease.
-Although in general, if this is the case, this is a "bad" results since ideally we would want the null to not change at all (i.e.
-if the metric is just picking up changes in number of elements, this is not really useful).
-
-#### Tree (dissimilarity)
-
-Mechanism    | Richness (alpha)| Dispersion | Regularity
--------------|------------|-------------|-------|
-Equalizing   | [ok] Lower | [ok] Lower  | [-ok] Higher|
-Facilitation | [NO] Higher| [-ok] Higher| [-ok] Higher|
-Exclusion    | [NO] Higher| [-ok] Higher| [-ok] Higher|
-Competition  | [ok] Lower | [ok] Lower  | [NO] Nothing|
-
-#### FD
-
-Mechanism | Richness (Rao)| Dispersion | Regularity
--------------|-------|-------|-------|
-Equalizing   | [NO] Lower  | [ok] Lower | [-ok] Higher|
-Facilitation | [NO] Higher | [-ok] Higher| [NO] Higher|
-Exclusion    | [-ok] Higher| [-ok] Higher| [-ok] Higher|
-Competition  | [ok] Lower  | [ok] Lower | [NO] Nothing|
-
-#### hypervolume
-
-Mechanism | Richness (Rao)| Dispersion | Regularity
--------------|-------|-------|-------|
-Equalizing   | [NO] Lower  | [-ok] Lower | [-ok] Higher|
-Facilitation | [-ok] Higher| [NO] Higher| [NO] Higher|
-Exclusion    | [NO] Higher | [NO] Higher| [NO] Higher|
-Competition  | [ok] Lower  | [NO] Lower | [ok] Nothing|
-
-#### TPD
-
-Mechanism | Richness (Rao)| Dispersion | Regularity
--------------|-------|-------|-------|
-Equalizing   | [NO] Lower  | [ok] Lower | [-ok] Higher|
-Facilitation | [NO] Higher | [-ok] Higher| [-ok] Higher|
-Exclusion    | [NO] Higher | [-ok] Higher| [-ok] Higher|
-Competition  | [NO] Lower  | [ok] Lower | [NO] Nothing|
-
-#### Other
-
-I actually don't remember of the top of my head what the convex.hull and melodic rao/mpd are representing.
-
-# Metrics correlations
-
-And finally below are all the correlations per stressor between each metric per removal levels (20, 40, 60, 80%).
-
-> TODO: do the PCA plots as suggested by Carlos to see things clearlier.
-
-```{r, eval = FALSE, echo = FALSE, message = FALSE}
-require(psych)
-
-## Split the results per stressor and per metric
-## (to extract the different removal levels)
-splitted_results <- lapply(scaled_results, split.metric)
-
-## Get the right removal level
-get.matrix.level <- function(results, level = 1) {
-    do.call(cbind, lapply(results, function(X, level) return(X[, level, drop = FALSE]), level = level))
-}
-## Wrapper for pairs.panels (with fancy names)
-pairs.panels.wrapper <- function(results, metric.names, ...) {
-    ## Make fancy names
-    if(missing(metric.names)) {
-        colnames(results) <- fancy.names(colnames(results), sep = ":\n")
-    } else {
-        colnames(results) <- metric.names
-    }
-    psych::pairs.panels(results, ...)
-}
-
-```
-
-## Equalizing
-
-### Metrics correlation for equalizing with 20% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 1), metric.names = fancy_names)
-```
-
-### Metrics correlation for equalizing with 40% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 2), metric.names = fancy_names)
-```
-
-### Metrics correlation for equalizing with 60% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 3), metric.names = fancy_names)
-```
-
-### Metrics correlation for equalizing with 80% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$equalizing, level = 4), metric.names = fancy_names)
-```
-
-## Facilitation
-
-### Metrics correlation for facilitation with 20% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 1), metric.names = fancy_names)
-```
-
-### Metrics correlation for facilitation with 40% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 2), metric.names = fancy_names)
-```
-
-### Metrics correlation for facilitation with 60% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 3), metric.names = fancy_names)
-```
-
-### Metrics correlation for facilitation with 80% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$facilitation, level = 4), metric.names = fancy_names)
-```
-
-## Filtering
-
-### Metrics correlation for filtering with 20% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 1), metric.names = fancy_names)
-```
-
-### Metrics correlation for filtering with 40% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 2), metric.names = fancy_names)
-```
-
-### Metrics correlation for filtering with 60% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 3), metric.names = fancy_names)
-```
-
-### Metrics correlation for filtering with 80% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$filtering, level = 4), metric.names = fancy_names)
-```
-
-## Competition
-
-### Metrics correlation for competition with 20% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 1), metric.names = fancy_names)
-```
-
-### Metrics correlation for competition with 40% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 2), metric.names = fancy_names)
-```
-
-### Metrics correlation for competition with 60% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 3), metric.names = fancy_names)
-```
-
-### Metrics correlation for competition with 80% removal
-
-```{r, fig.height = 18, fig.width = 18, echo = FALSE, eval = FALSE}
-## Equalizing level 1
-pairs.panels.wrapper(get.matrix.level(splitted_results$competition, level = 4), metric.names = fancy_names)
-```
-
-# Responses
-
-Carlos:
--TPD richness potential error: 
-  * I have double checked the scaling and everything seems OK (but I might be mistaken!) The centreing logic happens here: https://github.com/TGuillerme/eco_metrics_simulations/blob/c2a8deb9d673e0456f315d9a9f43285aafdfea93/Functions/extract.table.R#L59 and the scaling logic happens here: https://github.com/TGuillerme/eco_metrics_simulations/blob/c2a8deb9d673e0456f315d9a9f43285aafdfea93/Functions/extract.table.R#L30 (l.30 to 32)).
-  So it might be something with the bandwidth?
-The whole calculations for the TPD metrics happens here (https://github.com/TGuillerme/eco_metrics_simulations/blob/c2a8deb9d673e0456f315d9a9f43285aafdfea93/Functions/simulation.pipeline.R#L262) with no obvious bandwidth part (at least to my eyes).
-So maybe it's just using the defaults?
-
--A minor comment regarding plots:
- - [ ] TODO: reorganise plots labels (follow Stefano's pdf)
- - [ ] TODO: reorganise colours by "Richness / Divergence / Regularity"
- - [ ] TODO: change Convex.hull = Richness; Rao/mpd = Divergence
-
--Thomas, I would be happy to make the PCAs if you could send me a file with the correlation matrices
-
- - [ ] TODO: get the scaled results matrices for PCA
-
-Let's talk about the two points on the expectations and the biological realism!
-
-<!-- Carlos comments:
-
--Sorry to insist on this, but I really disagree with the expectations table as it is, we really would have to discuss what are our expectations (if any!) for any particular mechanism, and what they are based on.
-This is a big problem for me, since it can give the idea that some particular method (or even all methods) is somehow flawed because it doesnt meet some expectation.
-
--Biological realism: I think the best at this point would be to use some agent-based model as Pedro suggested.
-Sure, things interact, but I think we can make a model simple enough and change some parameters on it, so that we can isolate more or less different things.
-We recently made one for a paper, and I would be happy to adapt the code and run some simulations (after we decide which things we want to play with): https://onlinelibrary.wiley.com/doi/epdf/10.1111/geb.13203
-
- -->

### 2024-06-03 Analysis/3-eco-dis-case-study-empirical-pipeline.R

@@ -1,346 +1,0 @@
-
-
-
-
-############################################################################################################################
-############################################ ECODISSIMILARITY CASE STUDY: HAWAII ###########################################
-############################################################################################################################
-
-setwd("C:/Users/mwj207/OneDrive - University of Birmingham/hawaii/eco-dissimilarity/github")
-
-library(devtools)
-require(treats)
-require(dispRity)
-require(BAT)
-require(FD)
-require(ks)
-require(TPD)
-library(tidyverse)
-
-## Loading the functions
-source("./Functions/dbFD.R") # I added this in order to import the dbFD function
-source("./Functions/melodic.rao.R")
-source("./Functions/make.simulation.pipeline.R")
-source("./Functions/simulation.pipeline.R")
-
-## Loading the optimisation functions
-source("./Functions/optim.replicate.R")
-source("./Functions/optim.helpers.R")
-source("./Functions/analyse.replicates.R")
-
-
-
-###### SLIGHTLY ALTERED FUNCTIONS ##########################################################################################
-
-fun.proba.den.2 = function(presence, traits, verbose) {
-  if(verbose) {
-    message(".", appendLF = FALSE)
-  }
-  
-  if(ncol(traits) <= 4) {
-    traits_4dmax <- traits
-  } else {
-    traits_4dmax <- traits[, c(1:4)] # moved this from under "Measure probability density metrics
-  }
-  
-  ## Get the sds matrix
-  standard_deviations <- matrix(
-    data = rep(sqrt(diag(ks::Hpi.diag(traits_4dmax))), nrow(presence)), # Hpo.diag was originally: traits
-    nrow = nrow(presence),
-    byrow = TRUE,
-    dimnames = list(rownames(presence), paste0("sdA", 1:ncol(traits_4dmax)))) # ncol argument was origninally: traits
-  
-  ## Measure probability density metrics
-
-  suppressMessages(TPD_species <- TPD::TPDsMean(species = rownames(presence),
-                                                means = traits_4dmax,
-                                                sds = standard_deviations,
-                                                alpha = 0.99))
-  TPD_cases   <- TPD::TPDc(TPDs = TPD_species, sampUnit = t(presence))
-  output <- suppressMessages(do.call(cbind, TPD::REND(TPDc = TPD_cases)$communities))
[…]

### 2024-06-03 Analysis/script_empirical.R

@@ -1,42 +1,0 @@
-## Libraries
-require(treats)
-require(dispRity)
-require(BAT)
-require(FD)
-require(ks)
-require(TPD)
-
-## Loading the functions
-source("../Functions/melodic.rao.R")
-source("../Functions/make.simulation.pipeline.R")
-source("../Functions/simulation.pipeline.R")
-source("../Functions/optim.replicate.R")
-source("../Functions/optim.helpers.R")
-source("../Functions/analyse.replicates.R")
-source("../Functions/dbFD.R")
-
-## global trait space 
-empirical_traitspace <- read.csv('../Data/Raw/trait-space-5d-prehistoric-community.csv', row.names = 1) 
-
-## Species lists
-# prehistoric_species <- rownames(read.csv("../Data/Raw/prehistoric-species-list.csv"))
-historic_species <- rownames(read.csv('../Data/Raw/historic-species-list-minus-a-flammeus.csv', row.names = 1))
-extant_species <- rownames(read.csv('../Data/Raw/present-species-list-minus-a-flammeus.csv', row.names = 1))
-
-## Traitspace groups
-extant_group <- rownames(empirical_traitspace) %in% extant_species ## 37 species present
-historic_group <- rownames(empirical_traitspace) %in% historic_species ## 63 species present
-groups <- list("extant" = extant_group, "historic" = historic_group)
-
-## Set simulation type
-simulate <- make.simulation.pipeline(sim.data = empirical_traitspace, remove = groups)
-
-## Running the full optimisation
-results <- optim.replicate(input.fun = simulate,
-                           diagnose = var,
-                           summarise = optim.summarise,
-                           minimum = 20, maximum = 170,
-                           stop.variance = 0.0001,
-                           verbose = TRUE,
-                           bkp.path = "../Data/Processed/",
-                           bkp.name = "empirical_data.rda")

### 2024-06-03 Functions/fit.model.rm.R
Update fit.model.rm.R

@@ -25,7 +25,14 @@ fit.model.rm <- function(one_stressor, model.fun = lm, formula = metric ~ rm, su
     metric_data <- lapply(metric_data, table.data)
 
     ## Apply the models
-    all_models <- lapply(metric_data, function(data, fun, formula) fun(formula, data), fun = model.fun, formula = formula)
+    apply.model <- function(data, fun, formula) {
+        if(any(is.na(data))) {
+            return(NA)
+        } else {
+            return(fun(formula, data))
+        }
+    }
+    all_models <- lapply(metric_data, apply.model, fun = model.fun, formula = formula)
     
     if(!summary.out) {
         return(all_models)
@@ -43,9 +50,11 @@ fit.model.rm <- function(one_stressor, model.fun = lm, formula = metric ~ rm, su
 
 ## Extract the summary
 summarise.models <- function(one_model){
+    # if(is.na(one_model)) {
+    #     return(NA)
+    # }
     model <- summary(one_model)
     results <- c(model$coefficients[1,c(1,4)], model$coefficients[2,c(1,4)], "adj.R^2" = model$adj.r.squared)
     names(results)[c(1,3)] <- c("Intercept", "Slope")
     return(results)
-
 }

### 2024-06-03 Simulation_FD-2.R
Cleaned home/

@@ -1,599 +1,0 @@
-## ------------------------------------------------------------------------
-## 'PAPER TITLE'
-## ------------------------------------------------------------------------
-
-# Authors::
-# Thomas Guillerme (guillert@tcd.ie)
-# Stefano Mammola (stefanomammola@gmail.com)
-# Carlos Carmona (perezcarmonacarlos@gmail.com)
-# Pedro Cardoso (pedro.cardoso@helsinki.fi)
-
-## ------------------------------------------------------------------------
-# 'R script to reproduce the analyses'
-## ------------------------------------------------------------------------
-
-# Analysis performed with R (v. R 4.0.3) and R studio (v. 1.4.1103)
-
-## ------------------------------------------------------------------------
-# Loading R packages ------------------------------------------------------
-## ------------------------------------------------------------------------
-
-library("BAT")
-library("devtools")
-# devtools::install_github("TGuillerme/dads")
-# devtools::install_github("TGuillerme/dispRity")
-library("dads")     # needs to be version >= 0.1.4
-## Loading required package: dispRity
-library("dispRity") # needs to be version >= 1.5.7
-library("doSNOW")
-library("FD")
-library("geometry")
-library("hypervolume")
-library("parallel")
-library("tcltk")
-library("TPD")
-source("melodic.rao.R")
-
-## ------------------------------------------------------------------------
-# Custom R functions ------------------------------------------------------
-## ------------------------------------------------------------------------
-
-## Function for simplifying the plots (1D)
-#' @param data the dataset
-#' @param reduction the vector of selected element for the reduction
-#' @param null the vector of selected element for the null
-#' @param col two colours
-#' @param pch the pch (default = 19)
-#' @param trait which trait to plot (default is 1 for plot.1D and c(1,2) for plot.2D)
-#' @param xlab the xlab
-#' @param legend the legend text
-plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch = 19, trait = 1, xlab = "trait values", legend) {
-  ## Get the histograms
-  main_plot <- hist(data, plot = FALSE)
-  plot_1 <- hist(data[null, trait], plot = FALSE, breaks = main_plot$breaks)
-  plot_2 <- hist(data[reduction, trait], plot = FALSE, breaks = main_plot$breaks)
[…]

### 2024-06-03 Simulation_FD.R

@@ -1,587 +1,0 @@
-## ------------------------------------------------------------------------
-## 'PAPER TITLE'
-## ------------------------------------------------------------------------
-
-# Authors::
-# Thomas Guillerme (guillert@tcd.ie)
-# Stefano Mammola (stefanomammola@gmail.com)
-# Carlos Carmona (perezcarmonacarlos@gmail.com)
-# Pedro Cardoso (pedro.cardoso@helsinki.fi)
-
-## ------------------------------------------------------------------------
-# 'R script to reproduce the analyses'
-## ------------------------------------------------------------------------
-
-# Analysis performed with R (v. R 4.0.3) and R studio (v. 1.4.1103)
-
-## ------------------------------------------------------------------------
-# Loading R packages ------------------------------------------------------
-## ------------------------------------------------------------------------
-
-library("BAT")
-library("devtools")
-# devtools::install_github("TGuillerme/dads")
-# devtools::install_github("TGuillerme/dispRity")
-library("dads")     # needs to be version >= 0.1.4
-## Loading required package: dispRity
-library("dispRity") # needs to be version >= 1.5.7
-library("FD")
-library("geometry")
-library("hypervolume")
-library("TPD")
-
-## ------------------------------------------------------------------------
-# Custom R functions ------------------------------------------------------
-## ------------------------------------------------------------------------
-
-## Function for simplifying the plots (1D)
-#' @param data the dataset
-#' @param reduction the vector of selected element for the reduction
-#' @param null the vector of selected element for the null
-#' @param col two colours
-#' @param pch the pch (default = 19)
-#' @param trait which trait to plot (default is 1 for plot.1D and c(1,2) for plot.2D)
-#' @param xlab the xlab
-#' @param legend the legend text
-plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch = 19, trait = 1, xlab = "trait values", legend) {
-  ## Get the histograms
-  main_plot <- hist(data, plot = FALSE)
-  plot_1 <- hist(data[null, trait], plot = FALSE, breaks = main_plot$breaks)
-  plot_2 <- hist(data[reduction, trait], plot = FALSE, breaks = main_plot$breaks)
-  plot_height <- max(plot_1$counts, plot_2$counts)
-  
-  ## Make the colours as transparent (for overlaps)
-  make.transparent <- function(col, alpha = 0.5) {
-    rgb_vals <- grDevices::col2rgb(col)
[…]

### 2024-06-03 Simulation_FD_mod.R

@@ -1,397 +1,0 @@
-## ------------------------------------------------------------------------
-## 'PAPER TITLE'
-## ------------------------------------------------------------------------
-
-# Authors::
-# Thomas Guillerme (guillert@tcd.ie)
-# Stefano Mammola (stefanomammola@gmail.com)
-# Carlos Carmona (perezcarmonacarlos@gmail.com)
-# Pedro Cardoso (pedro.cardoso@helsinki.fi)
-
-## ------------------------------------------------------------------------
-# 'R script to reproduce the analyses'
-## ------------------------------------------------------------------------
-
-# Analysis performed with R (v. R 4.0.3) and R studio (v. 1.4.1103)
-
-## ------------------------------------------------------------------------
-# Loading R packages ------------------------------------------------------
-## ------------------------------------------------------------------------
-
-library("BAT")
-library("devtools")
-# devtools::install_github("TGuillerme/dads")
-# devtools::install_github("TGuillerme/dispRity")
-library("dads")     # needs to be version >= 0.1.4
-## Loading required package: dispRity
-library("dispRity") # needs to be version >= 1.5.7
-library("FD")
-library("geometry")
-library("hypervolume")
-library("TPD")
-
-## ------------------------------------------------------------------------
-# Custom R functions ------------------------------------------------------
-## ------------------------------------------------------------------------
-
-## Function for simplifying the plots (1D)
-#' @param data the dataset
-#' @param reduction the vector of selected element for the reduction
-#' @param null the vector of selected element for the null
-#' @param col two colours
-#' @param pch the pch (default = 19)
-#' @param trait which trait to plot (default is 1 for plot.1D and c(1,2) for plot.2D)
-#' @param xlab the xlab
-#' @param legend the legend text
-plot.1D <- function(data, reduction, null, main, col = c("blue", "orange"), pch = 19, trait = 1, xlab = "trait values", legend) {
-  ## Get the histograms
-  main_plot <- hist(data, plot = FALSE)
-  plot_1 <- hist(data[null, trait], plot = FALSE, breaks = main_plot$breaks)
-  plot_2 <- hist(data[reduction, trait], plot = FALSE, breaks = main_plot$breaks)
-  plot_height <- max(plot_1$counts, plot_2$counts)
-  
-  ## Make the colours as transparent (for overlaps)
-  make.transparent <- function(col, alpha = 0.5) {
-    rgb_vals <- grDevices::col2rgb(col)
[…]

### 2024-06-03 melodic.rao.R

@@ -1,30 +1,0 @@
-##MEan DIssimilarity Components##
-#samp: community matrix; sites in lines, species in columns
-#dis: dissimilarity matrix
-
-melodic.rao <- function(samp,dis,abundance.weighted=TRUE){
-	N<-dim(samp)[1]
-	rao<-richness<-mpd<-simpson<-numeric(N)
-	for (i in 1:N){
-		sppInSample<-names(samp[i,samp[i, ]>0])
-	    #richness[i]<-length(sppInSample)
-		if (length(sppInSample)>1){
-      		sample.dis<-dis[sppInSample,sppInSample]
-			abund.w<-numeric(length(sppInSample))
-			if (abundance.weighted){
-				abund.w <- samp[i , sppInSample] / sum(samp[i , sppInSample])
-			} 	else{
-				abund.w <- rep(1 , length(sppInSample)) / length(sppInSample)
-			}	
-        	sample.weights <- outer(as.numeric( abund.w) , as.numeric( abund.w))
-		  	rao[i] <- sum(sample.weights * sample.dis)
-		  	mpd[i] <- weighted.mean(sample.dis[lower.tri(sample.dis)] , sample.weights[lower.tri(sample.weights)])
-		  	simpson[i]<-sum(2*sample.weights[lower.tri(sample.weights)])
-
-		}	else {
-			rao[i]<- mpd[i]  <- simpson[i]<- 0
-		}
-	}	
-	out<-list(rao=rao, mpd=mpd, simpson=simpson)
-	return(out)	
-}

### 2024-06-03 Analysis/Choosing_metrics_vignette.Rmd
Create Choosing_metrics_vignette.Rmd

---
title: "Choosing a space occupancy metric"
author: "Thomas Guillerme"
bibliography: ../Manuscript/references.bib
output:
  html_document: default
  pdf_document: default
---

```{r}
library(dispRity)
library(knitr)
```

This script illustrates a potential pipeline of how to choose a correct space occupancy metric (_aka_ functional diversity, dissimilarity, disparity metric) to answer a biological question.
Our question here could be whether the functional diversity of Hawaiian endemic birds decreased more in the last 500 years than ever before?
I.e. is the decrease in trait space occupancy greater in the present than in 1500 compared to all species that where recorded.

> Note that the exercise in this vignette can be done using the [`moms`](https://tguillerme.shinyapps.io/moms/) graphical user interface (with the associate manual [here](https://raw.githack.com/TGuillerme/moms/master/inst/moms_vignette.html)).

# The data

Here we will use the empirical data from @guillerme2024. The trait data is compiled from @tobias2022avonet, @pigot2020macroevolutionary and @sheard2020ecological and the extinction data is compiled from @matthews2023global.
It consists of a trait space of 118 species and 5 dimensions recorded in the last 125,000 years.
From this trait space we have 2 subsets that corresponds to the endemic species present in Hawaii in 1500CE (the historic group of 63 species) and the species present today (37 species).
We will group all that in a `dispRity` object to facilitate the pipeline.

```{r}
## global trait space 
empirical_traitspace <- read.csv('../Data/Raw/trait-space-5d-prehistoric-community.csv', row.names = 1)

## Species lists
historic_species <- rownames(read.csv('../Data/Raw/historic-species-list-minus-a-flammeus.csv', row.names = 1))
extant_species <- rownames(read.csv('../Data/Raw/present-species-list-minus-a-flammeus.csv', row.names = 1))

## Traitspace groups
extant_group <- rownames(empirical_traitspace) %in% extant_species ## 37 species present
historic_group <- rownames(empirical_traitspace) %in% historic_species ## 63 species present
groups <- list("extant" = extant_group, "historic" = historic_group)

## Creating a dispRity object containing all the groups
custom_groups <- lapply(groups, which)
custom_groups$all <- 1:nrow(empirical_traitspace)
grouped_space <- custom.subsets(empirical_traitspace, group = custom_groups)
```

## Visualising the data

```{r, fig.height = 8, fig.width = 8}
## Visualising the group
plot(grouped_space)
```

Note here that the groups are subsets of each other so the species that are in the group `"extant"` (red) are also in the group `"historic"` (green) and `"all"` (blue).
From visualising the data we can first see that both PC are a fair representation of the data (around 85% of the data).
In other words, some things might happen on the other dimensions here (e.g. all the elements of one group might be weirdly distributed on some other PC axes) but that can only affect 15% of the data.

# Choosing a space occupancy metric

Once we have set up the data as a `dispRity` object, we can choose a space occupancy metric to reflect changes in the dataset.
We can do that using already implemented metrics in `dispRity` (see `?dispRity.metric`) or by making or own metrics (see details below).
The objective here will be to have a metric that reflects changes of interest.
Again, the question being: "the functional diversity of Hawaiian endemic birds decreased more in the last 500 years than ever before?"; we can posit that changes in functional diversity should be approximated by changes in trait space size.
The bigger the trait space, the more functional diversity.

> Note that this assumption should not be generalised. It is just what we posit for this dataset and this specific question.

## A simple case: the convex hull volume

We can start with a simple and rather intuitive measurement of trait space size: the volume of the convex hull of each group, that is volume of the smallest non-regular polygon that contains all points of each groups.

```{r}
## Measuring the convhull volume
convhull_volume <- dispRity(grouped_space,
                            metric = convhull.volume)
```

We can then summarise this in a table:

```{r, eval = FALSE}
## Printing the table
summary(convhull_volume)
```

```{r, echo = FALSE}
## Printing the table (for markdown fancy print)
kable(summary(convhull_volume))
```

This seems satisfying with or prior knowledge on species extinctions in Hawaii: there has been a huge reduction in functional diversity since the event of globalisation (1500CE onwards): we have a reduction of `r round(((summary(convhull_volume)$obs[3] - summary(convhull_volume)$obs[2])/summary(convhull_volume)$obs[3])*100, digits = 2)`% in the last 12500 years (something like loosing 0.0002 % of the trait space per year).
Then, we have a reduction `r round(((summary(convhull_volume)$obs[3] - summary(convhull_volume)$obs[1])/summary(convhull_volume)$obs[3])*100, digits = 2)`% in the last 500 years (something like loosing 0.14 % of the trait space per year).

## Checking sample size

However, we could note here that this change in volume is also closely linked to the change in the number of species, at least qualitatively (loosing `r round(((summary(convhull_volume)$n[3] - summary(convhull_volume)$n[2])/summary(convhull_volume)$n[3])*100, digits = 2)`% of species for the historic data and `r round(((summary(convhull_volume)$n[3] - summary(convhull_volume)$n[1])/summary(convhull_volume)$n[3])*100, digits = 2)`% for the extant data).
This might not be a problem _per se_ but in our case we are interested in the loss of function diversity not just the number of species.
So we have to first check whether the changes of functional diversity are potentially de-coupled with changes in number of species.
One way to do that is by rarefying the data, i.e. resampling each subset so they contain the same number of species.
We can do that easily in `dispRity` using the `boot.matrix` function:

```{r}
## Measuring the convhull volume on the rarefied subsets (to 37 elements)
rarefied_conv_vol <- dispRity(boot.matrix(grouped_space, rarefaction = 37),
                              metric = convhull.volume)
```

```{r, eval = FALSE}
## Printing the table
summary(rarefied_conv_vol)
```

```{r, echo = FALSE}
## Printing the table (for markdown fancy print)
kable(summary(rarefied_conv_vol))
```

Or plotting it:

```{r, fig.height = 8, fig.width = 4}
par(mfrow = c(1, 2))
plot(rarefied_conv_vol, main = "raw data")
plot(rarefied_conv_vol, rarefaction = 37, main = "rarefied data")
```

Note that now the differences are not that clear.
This might suggest that there is something with specific outliers in our dataset or that the metric is actually more capturing changes in the number of elements in the trait space rather than directly the trait space occupancy (_sensu_ functional diversity).

## Checking the ability of the convex hull volume to capture changes in size

To check that, we can actually run some simulations to see whether the metric can actual capture changes in trait space size!
We can do that using the `dispRity::test.metric` function from that uses the same pipeline used in @guillerme2024 and @guillerme2020shifting.

```{r}
## Testing the convex hull volume metric
convvol_test <- test.metric(grouped_space, metric = convhull.volume, shifts = c("random", "size"), replicates = 10)
```

The results for the functional divergence tests:

```{r, eval = FALSE}
summary(convvol_test)
```

```{r, echo = FALSE}
kable(summary(convvol_test))
```

The visualisation of the results:

```{r, fig.height = 12, fig.width = 6}
plot(convvol_test)
```

Here we can clearly see that no matter which trait reduction we applied, we would always see a change of the convex hull volume linked to the change in the number of elements: we can see that the slope are very similar whether we randomly remove elements (black), remove elements from elements on the edge of the distribution (orange) or remove elements from the inside of the distribution (blue).

This should lead us to conclude that, even though the metric results make sense, they only do here because they are linked to the changes in the number of elements and we can never be sure whether it captures changes in trait space size (i.e. or proxy for functional diversity) or the number of elements (species richness).

> Note that this conclusion here is not a generalisation about this trait space or this metric. It is only valid for this specific trait space, this specific metric and this specific interpretation of function diversity.

# Testing different metrics

Unfortunately there is no one-size fit all recipe on how to find the correct metric for the dataset and question of interest.
However, using tools like [`moms`](https://tguillerme.shinyapps.io/moms/) or `dispRity` we can test several ones and see which one fits our needs the best.
Here is a series of them:

## The average spanning tree length

```{r, fig.height = 12, fig.width = 6}
plot(test.metric(grouped_space, metric = c(mean, span.tree.length), shifts = c("random", "size"), replicates = 10))
```

This metric is also not suitable as it seems inversely correlated to the sample size. Furthermore, it also seem to not pick changes in size (orange) but rather in hollowness (blue).

## The ratio of deviation from the centroid (functional diversity _sensu_ @villeger2008new)

```{r, fig.height = 12, fig.width = 6}
plot(test.metric(grouped_space, metric = func.div, shifts = c("random", "size"), replicates = 10))
```

This metric, is more interesting as it is not directly related to the sample size and therefore more clearly decoupled from species richness. Changes in trait space sizes are linearly and inversely related to the amount of data considered (i.e. the more volume, the lower the functional diversity) which is not a problem _per se_ but can be counterintuitive (hence the usefulness of testing it before interpreting it). However it is also interest to not the non-linear relationship with hollowness (blue) that can follow the same trend as the trait space size (blue) so it might be hard to disentangle either.

## The mean distance to the centroid

```{r, fig.height = 12, fig.width = 6}
plot(test.metric(grouped_space, metric = c(mean, centroids), shifts = c("random", "size"), replicates = 10))
```

Unsurprising, this metric seem to give a similar pattern as the deviation from the centroid. However, we can note here that the metric increases with size (orange) which is a good behaviour in our case. Furthermore, the hollowness (blue) seems not to be following the changes in size making it potentially easier to distinguish between both changes in the trait space.

# Results

```{r}
## Measuring the average distance from centroid on the rarefied subsets (to 37 elements)
rarefied_mean_cen <- dispRity(boot.matrix(grouped_space, rarefaction = 37),
                              metric =  c(mean, centroids))
```

```{r, eval = FALSE}
## Printing the table
summary(rarefied_mean_cen)
```

```{r, echo = FALSE}
## Printing the table (for markdown fancy print)
kable(summary(rarefied_mean_cen))
```

```{r, fig.height = 4, fig.width = 8}
par(mfrow = c(1, 2))
plot(rarefied_mean_cen, main = "raw data")
plot(rarefied_mean_cen, rarefaction = 37, main = "rarefied data")
```

Using this metric, we see a more complex picture of the changes in functional diversity (defined here as the mean distance to centroid) where it seems to not decrease through time compared to using the convex hull volume.
This might be because of many reasons but looking at the 2D representation of the trait space at the start of this vignette, this might be because the biggest traitspace loss might actually have occurred for the historic species (the blue dots in the 2D trait space plot).

# References

### 2024-06-04 Analysis/Choosing_metrics_vignette.Rmd
Update Choosing_metrics_vignette.Rmd

@@
 
-```{r, fig.height = 8, fig.width = 4}
+```{r, fig.height = 4, fig.width = 8}
 par(mfrow = c(1, 2))

### 2024-06-04 Manuscript/CoverLetter.tex
Update CoverLetter.tex

@@
 
-Since the seminal publication by Petchey and Gaston in Ecology Letters in 2002 on functional diversity (``Functional diversity (FD), species richness and community composition''; cumulating $>$ 2000 citations), functional diversity is now routinely used as a statistic to describe complex multidimensional datasets.
+Since the seminal publication by Petchey and Gaston in Ecology Letters in 2002 \textit{``Functional diversity (FD), species richness and community composition''} (cumulating $>$ 2000 citations), functional diversity is now routinely used as a statistic to describe complex multidimensional datasets.
 This has been driven by a really motivated community of ecologists proposing many different ways to measure functional diversity depending on different datasets or different questions.
@@
 However, this richness of ways to measure functional diversity can make it daunting and or confusing for researchers that want to add functional diversity to their research methodology toolkit.
-We argue that this confusions can stem from both an epistemological misunderstanding as well as practical hurdles.
-The second is nicely alleviated by a dynamic community of researchers proposing open source and easy to use implementations to measure functional diveristy.
-But the first hurdle remains.
+We argue that this confusions can stem from both (1) an epistemological misunderstanding as well as (2) practical hurdles.
+The latter is nicely alleviated by a dynamic community of researchers proposing open source and easy to use implementations to measure functional diveristy.
+But the former misunderstanding remains.
 Indeed, although it is often rather clear how functional diversity in general can capture changes communities (investigating processes; answering the question ``how are communities changing?'') and relate them to general understanding in ecology (understanding mechanisms; ``why are communities changing?''); surprisingly little work has been published in the last few years on how the metrics used to measure functional diversity can accurately measure complex multidimensional patterns (capturing patterns; ``what is changing?'').

### 2024-06-05 Analysis/Choosing_metrics_vignette.Rmd
Choosing metrics vignette: Maria's edits

@@
 This script illustrates a potential pipeline of how to choose a correct space occupancy metric (_aka_ functional diversity, dissimilarity, disparity metric) to answer a biological question.
-Our question here could be whether the functional diversity of Hawaiian endemic birds decreased more in the last 500 years than ever before?
-I.e. is the decrease in trait space occupancy greater in the present than in 1500 compared to all species that where recorded.
+Our question here could be whether the functional diversity of Hawaiian native birds decreased more in the last 500 years than ever before?
+I.e. does the loss of species in the last 500 years constitute a greater decrease in trait space occupancy compared to the loss of all species recorded between 125,000 years ago and 500 years ago.
 
@@
 Here we will use the empirical data from @guillerme2024.
-The trait data is compiled from @tobias2022avonet, @pigot2020macroevolutionary and @sheard2020ecological and the extinction data is compiled from @matthews2023global.
-It consists of a trait space of 118 species and 5 dimensions recorded in the last 125,000 years.
+The trait data of extant species is compiled from @tobias2022avonet  while extinct trait data is from @sayol (in review).
+Species lists were compiled from @matthews2023global with the addition of two marine species.
+It consists of a trait space of 118 species in 5 dimensions recorded in the last 125,000 years.
 From this trait space we have 2 subsets that corresponds to the endemic species present in Hawaii in 1500CE (the historic group of 63 species) and the species present today (37 species).
@@
 Once we have set up the data as a `dispRity` object, we can choose a space occupancy metric to reflect changes in the dataset.
-We can do that using already implemented metrics in `dispRity` (see `?dispRity.metric`) or by making or own metrics (see details below).
+We can do that using already implemented metrics in `dispRity` (see `?dispRity.metric`) or by making our own metrics (see details below).
 The objective here will be to have a metric that reflects changes of interest.
@@
 the more volume, the lower the functional diversity) which is not a problem _per se_ but can be counterintuitive (hence the usefulness of testing it before interpreting it).
-However it is also interest to not the non-linear relationship with hollowness (blue) that can follow the same trend as the trait space size (blue) so it might be hard to disentangle either.
+However it is also interest to note the non-linear relationship with hollowness (blue) that can follow the same trend as the trait space size (blue) so it might be hard to disentangle either.
 

### 2024-06-05 Manuscript/main.tex
submission BS added
paper up

@@
 Thomas Guillerme$^{1,*}$, 
-Pedro Cardoso$^{2,3}$,
+Pedro Cardoso$^{2,3}$, %pedro.cardoso@helsinki.fi
 % Carlos P. Carmona$^{4}$,
-Maria Wagner J\o rgensen$^{4}$,
-Stefano Mammola$^{3,5,6}$,
-Thomas J. Matthews$^{4,7}$}\\
+Maria Wagner J\o rgensen$^{4}$, %mwj207@student.bham.ac.uk
+Stefano Mammola$^{3,5,6}$, %stefanomammola@gmail.com
+Thomas J. Matthews$^{4,7}$}\\ % txm676@gmail.com
 \noindent {\small \it 
@@
    \item Number of words in the abstract: 189
-   \item Number of words in the main text: 3949
+   \item Number of words in the main text: 3960
    \item Number of words in the text box: 419
@@
 The simulations, results figures and tables are entirely reproducible via \url{https://github.com/TGuillerme/eco_metrics_simulations}.
+An additional vignette on how to choose metrics is available at \url{https://github.com/TGuillerme/eco_metrics_simulations/blob/master/Analysis/Choosing_metrics_vignette.Rmd}.
 

### 2024-06-05 Manuscript/supplementaries.tex
paper up

@@
 The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
-The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value > 0.05).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
 Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.

### 2024-06-05 Manuscript/main.tex
Update main.tex

@@
 
-All code and data required to reproduce this analysis is available on @@@
+All code and data required to reproduce this analysis is available on GitHub: \url{https://github.com/TGuillerme/eco_metrics_simulations/}.
 
@@
 This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
-%@@@ TODO: add supplementary
 Furthermore, the scaling of the metrics paired with the levels of removal (80\%, 60\%, 40\% and 20\%) creates an expected artifact similarity when comparing the random removals against the other ones: when less data is available overall, the metrics are more likely to be similar (e.g.

### 2024-09-23 Manuscript/main_ecography_R1.tex
R1 version INIT

--- 2024-06-05 Manuscript/main.tex
@@
 
-\noindent {\normalsize \sc
-Thomas Guillerme$^{1,*}$, 
-Pedro Cardoso$^{2,3}$, %pedro.cardoso@helsinki.fi
-% Carlos P. Carmona$^{4}$,
-Maria Wagner J\o rgensen$^{4}$, %mwj207@student.bham.ac.uk
-Stefano Mammola$^{3,5,6}$, %stefanomammola@gmail.com
-Thomas J. Matthews$^{4,7}$}\\ % txm676@gmail.com
-\noindent {\small \it 
-$^1$School of Biosciences, The University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$CE3C—Centre for Ecology, Evolution and Environmental Changes, CHANGE – Global Change and Sustainability Institute, Faculty of Sciences, University of Lisbon, Lisbon, Portugal\\
-$^3$Laboratory for Integrative Biodiversity Research (LIBRe), Finnish Museum of Natural History (Luomus), University of Helsinki, Helsinki, Finland\\
-% $^4$Institute of Ecology and Earth Sciences, University of Tartu, Tartu, Estonia\\
-$^4$School of Geography, Earth and Environmental Sciences and Birmingham Institute of Forest Research, University of Birmingham, Birmingham B15 2TT, UK\\
-$^5$Molecular Ecology Group (MEG), Water Research Institute, National Research Council (CNR-IRSA), Verbania Pallanza, Italy\\
-$^6$National Biodiversity Future Center, Piazza Marina 61, 90133 Palermo, Italy\\
-$^7$Centre for Ecology, Evolution and Environmental Changes/Azorean Biodiversity Group / CHANGE – Global Change and Sustainability Institute and Universidade dos Açores – Faculty of Agricultural Sciences and Environment; PT-9700-042, Angra do Hero\'ismo, A\c ores, Portugal.\\}
+% \noindent {\normalsize \sc
+% Thomas Guillerme$^{1,*}$, 
+% Pedro Cardoso$^{2,3}$, %pedro.cardoso@helsinki.fi
+% % Carlos P. Carmona$^{4}$,
+% Maria Wagner J\o rgensen$^{4}$, %mwj207@student.bham.ac.uk
+% Stefano Mammola$^{3,5,6}$, %stefanomammola@gmail.com
+% Thomas J. Matthews$^{4,7}$}\\ % txm676@gmail.com
+% \noindent {\small \it 
+% $^1$School of Biosciences, The University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
+% $^2$CE3C—Centre for Ecology, Evolution and Environmental Changes, CHANGE – Global Change and Sustainability Institute, Faculty of Sciences, University of Lisbon, Lisbon, Portugal\\
+% $^3$Laboratory for Integrative Biodiversity Research (LIBRe), Finnish Museum of Natural History (Luomus), University of Helsinki, Helsinki, Finland\\
+% % $^4$Institute of Ecology and Earth Sciences, University of Tartu, Tartu, Estonia\\
+% $^4$School of Geography, Earth and Environmental Sciences and Birmingham Institute of Forest Research, University of Birmingham, Birmingham B15 2TT, UK\\
+% $^5$Molecular Ecology Group (MEG), Water Research Institute, National Research Council (CNR-IRSA), Verbania Pallanza, Italy\\
+% $^6$National Biodiversity Future Center, Piazza Marina 61, 90133 Palermo, Italy\\
+% $^7$Centre for Ecology, Evolution and Environmental Changes/Azorean Biodiversity Group / CHANGE – Global Change and Sustainability Institute and Universidade dos Açores – Faculty of Agricultural Sciences and Environment; PT-9700-042, Angra do Hero\'ismo, A\c ores, Portugal.\\}
 
 \end{center}
-\bigskip
-\noindent{*\bf Corresponding author.} \textit{t.guillerme@sheffield.ac.uk}\\
+% \bigskip
+% \noindent{*\bf Corresponding author.} \textit{t.guillerme@sheffield.ac.uk}\\
 
-\noindent (Keywords: Functional diversity, dissimilarity, disparity, patterns, processes, mechanisms)\\
+% \noindent (Keywords: Functional diversity, dissimilarity, disparity, patterns, processes, mechanisms)\\
 
-\begin{itemize}
-   \item Article type: Letter
-   \item Number of words in the abstract: 189
-   \item Number of words in the main text: 3960
-   \item Number of words in the text box: 419
-   \item Number of references: 54
-   \item 3 Figures, 2 tables and 1 text box
-\end{itemize}
+% \begin{itemize}
+%    \item Article type: Perspective
+%    \item Number of words in the abstract: 189
+%    \item Number of words in the main text: 3960
+%    \item Number of words in the text box: 419
+%    \item Number of references: 54
+%    \item 3 Figures, 2 tables and 1 text box
+% \end{itemize}
 
-\subsection{Statement of authorship}
+% \subsection{Statement of authorship}
 
-TG, PC, MWJ, SM and TJM designed the study, analysed the data and wrote the manuscript. TG wrote the code to analyse the data.
+% TG, PC, MWJ, SM and TJM designed the study, analysed the data and wrote the manuscript. TG wrote the code to analyse the data.
 
-\subsection{Data accessibility statement}
+% \subsection{Data accessibility statement}
 
-All code and data required to reproduce this analysis is available on GitHub: \url{https://github.com/TGuillerme/eco_metrics_simulations/}.
+% All code and data required to reproduce this analysis is available on GitHub: \url{https://github.com/TGuillerme/eco_metrics_simulations/}.
 
-%---------------------------------------------
-%
-%       ABSTRACT
-%
-%---------------------------------------------
+% %---------------------------------------------
+% %
+% %       ABSTRACT
+% %
+% %---------------------------------------------
 
@@
 Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.
+
+\noindent (Keywords: Functional diversity, dissimilarity, disparity, patterns, processes, mechanisms)\\
 
@@
 
-\section{Acknowledgements}
-Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.
-TG was funded by the UKRI-NERC grant NE/X016781/1.
-SM was supported by NBFC, funded by the Italian Ministry of University and Research, PNRR, Missione 4, Componente 2, ``Dalla ricerca all'impresa'', Investimento 1.4, Project CN00000033.
-MWJ was supported by NERC CENTA2 grant NE/S007350/1 and University of Birmingham.
+% \section{Acknowledgements}
+% Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.
+% TG was funded by the UKRI-NERC grant NE/X016781/1.
+% SM was supported by NBFC, funded by the Italian Ministry of University and Research, PNRR, Missione 4, Componente 2, ``Dalla ricerca all'impresa'', Investimento 1.4, Project CN00000033.
+% MWJ was supported by NERC CENTA2 grant NE/S007350/1 and University of Birmingham.
 

### 2024-09-26 Manuscript/main_ecography_R1.tex
Updated bits of the abstract

@@
 Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
-Indeed, functional diversity metrics allow researchers to summarize complex occupancy patterns in space and/or time (what is changing?) that lead to changes in communities and/or populations (the process; how is it changing?) in response to some stressors (the mechanism; why is it changing?).
+Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to some stressors.
+In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism. 
 However, as the diversity of functional diversity metrics and methods increases, it is often not directly clear which metric is more readily appropriate for which question.
-We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms (environmental filtering, competitive exclusion, equalizing fitness, and facilitation) in community ecology.
+We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.
 Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
-We show that different metrics perform better for different tasks, emphasizing the importance of not using a one-size-fits-all metric.
+We show that different metrics perform better for different tasks, emphasising the importance of not using a one-size-fits-all metric.
 Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.
@@
 the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g.
-Divergence based on probability density for the equalizing mechanism for a decrease in space occupancy).
+Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).
 Some metrics even captured an opposite pattern to what was expected: for all metrics, we naively expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
-the functional evenness based on hypervolume for the equalizing mechanism) or non-linear ones (e.g.
-Alpha diversity based on dendrogram for the equalizing mechanism).
+the functional evenness based on hypervolume for the equalising mechanism) or non-linear ones (e.g.
+Alpha diversity based on dendrogram for the equalising mechanism).
 This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.

### 2024-09-26 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 
-In the last two decades, there has been a progressive expansion in ecology and evolution studies from taxonomic-oriented approaches, with species as the focal point, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
+In the last two decades, there has been a progressive expansion in ecology and evolution studies from taxonomic-oriented approaches, with species as the focal point, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{violle2007let, mammola2021concepts, palacio2022protocol}.
 Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g.
@@
 First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than the organisms themselves.
-Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g.
+Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualisation of general principles broadly valid in space (e.g.
 unrelated species pools) and time (e.g.
 anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
-Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g.
+Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g.
 agronomy \citealt{martin2015plant}).
@@
 
-\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
+\textit{Why?} Mechanism description is then about linking the pattern and the process to some biological properties (sometimes referred to as ``rules'') and is at the core of answering the biological question at hand.
 In most cases, this is what researchers are actually trying to understand.
@@
 In such a scenario, measuring the occupied area (i.e.
-$km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
+$km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.
 In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
@@
 
-Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations}).
-We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands (process) resulting in a trait space modified by extinctions (the pattern) capturing the effects of species extinction on species trait distributions (the mechanism).
+Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and \ref{Tab:mechanisms}).
+To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands (process) resulting in a trait space modified by extinctions (the pattern) capturing the effects of species extinction on species trait distributions (the mechanism).
 We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
@@
 
-First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.) in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
+First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
+a pure birth speciation model) until reaching 200 elements in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
+These elements can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. 
 This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
@@
 
-\item Increasing density: by removing the required amount of pairs of points with a pair distance of at least $D$ (i.e.
-removing the $n$ pairs of points that are at least $D$ distance away from each other).
+\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$ (i.e.
+removing the $n$ pairs of points that are at least $D$ distance away from each other to remove the desired number of elements).
 This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
@@
 an estimated number of categories summarising the distribution, like in a histogram).
-Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and $p$ is a factor increasing the flattening/steepening (here we used $p=3$).
+Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$. 
+For bandwidths with a low density (low values of $n$) we used $m>0$ to increase their resampling probability, and the opposite for bandwidths with a high density (high values of $n$; using $m<0$ to decrease sampling property).
+This effectively flattened the normal distribution of the data by some factor $p$ (here we used $p=3$).
 This approximates our \textbf{competition mechanism}.
@@
 \caption{\scriptsize{Illustration of the simulation protocol used in this paper.
-A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting a trait space of 2, 4, or 8 dimensions.
-We then applied different stressors to the resulting trait space using the following:
-B) \textbf{null mechanism}: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
-C) \textbf{equalising mechanism}: removing species on the edge of the distribution;
-D) \textbf{facilitation mechanism}: removing species to reduce the distance between pairs of species (increasing local density; decreasing evenness);
-E) \textbf{filtering mechanism}: removing species on one extreme of the distribution;
-F) \textbf{competition mechanism}: proportionally removing species on the centre of the distribution (``flattening the curve'').}}
+A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 elements resulting a trait space of 2, 4, or 8 dimensions.
+We then applied different stressors to the resulting trait space.
+We represent the elements removed by the stressors in blue and the ones kept in the trait space in orange.
+The first column illustrate the distribution of the elements in one dimension with the overlap between the removed (blue) and kept (orange) elements in brown (note we did not measure any metric on 1D spaces).
+The second column illustrate the same distributions in 2 dimensions.
+We did not illustrate the distributions in 4 or 8 dimensions here.
+We used the following stressors:
+B) \textbf{null mechanism}: randomly removing half of the elements;
+C) \textbf{equalising mechanism}: removing elements on the edge of the distribution;
+D) \textbf{facilitation mechanism}: removing elements to reduce the distance between pairs of elements (increasing local density; decreasing evenness);
+E) \textbf{filtering mechanism}: removing elements on one extreme of the distribution;
+F) \textbf{competition mechanism}: proportionally removing elements on the centre of the distribution (``flattening the curve'').}}
 \label{Fig:simulations}
@@
 
-We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
+We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts} using 12 metrics (Tab. \ref{Tab:metrics}):
 1) Richness, encompassing metrics reflecting the sum of differences among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});

### 2024-09-26 Manuscript/main_ecography_R1.tex
added box idea

@@
 
-\section*{Box 1}
+\section*{Box 1: what, how and why = pattern, process and mechanism}
 \cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
@@
 The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test $p$ value $> 0.05$).
-Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model ``centred and scaled metric score $\sim$ amount of data removed''.
 Dashed thin grey lines represent non-significant models ($p$ value of slope or/and intercept $> 0.05$).
+The exact model results are available in supplementary table @@@.
 Similar figures are available in the supplementary materials for 2 and 8 dimensions.}}
@@
 Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).
-Some metrics even captured an opposite pattern to what was expected: for all metrics, we naively expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
+Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an increase of trait space occupancy (e.g.
 the functional evenness based on hypervolume for the equalising mechanism) or non-linear ones (e.g.
@@
 This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
-Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results})!
+Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
 
@@
 
+\subsections{Results}
+
 We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
@@
 Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
+
+\subsubsection{Empirical results}
+
+The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to traits (Figure \ref{Fig:empirical_results}).
+This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g.
+\citealt{sayol2021loss,Matthews2022}).
+This could partly be due to the proportion of passerine extinctions in Hawaii.
+Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
+However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
+It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
+But these results, given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines,  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
+relative to random extinctions).
+In addition, it is worth stressing that these analyses included marine species (which constitute less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
+
+More broadly, as can be seen from this empirical case, patterns are likely to be much less clear than with simulated data due to multiple stressors acting simultaneously.
+For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
+In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \citep{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but is related to genetics (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
+
+\subsection{Caveats}
 
@@
 the volume of an evenly occupied space scales differently in high dimensions compared to the same space but with normally distributed observations).
-Therefore we advise workers to test assumptions specifically about their trait space and ``play'' around with it before choosing their functional diversity/disparity framework and metric of interest.
 
@@
 
-The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to traits (Figure \ref{Fig:empirical_results}).
-This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g.
-\citealt{sayol2021loss,Matthews2022}).
-This could partly be due to the proportion of passerine extinctions in Hawaii.
-Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
-However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
-It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
-But these results, given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines,  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
-relative to random extinctions).
-In addition, it is worth stressing that these analyses included marine species (which constitute less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
+\bigskip
+\bigskip
+\hrule
+\hrule
+
+\section*{Box 2: so which metric do \textit{I} choose?}
+\susbsection{So which metric to choose?}
+
+TL;DR: it depends!
 
@@
 facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
+
+% Maybe use the empirical example here.
+
+1 - Which questions to ask yourself?
+
+- Why? Think of the broad biological mechanism of interest. For example: ``we are interested in inter specific competition''.
+- How?
+With the data you have at hand, identify the process you could analyse.
+For example ``we have data collected on two islands, one control with only one species and another one with two species''.
+- What? Then select your pattern and test whether it could answer your question.
+
+2 - Choose your broad type of metric: what pattern do you want to capture?
+
+-Is one group bigger than an other? -> Richness
+-Is one group more dispersed that the other? -> Divergence
+-Is one group more evenly dispersed in the space than another? -> Regularity 
+-Is it something else -> There are many many more subtleties to these broad distinctions in the literature!
+
+3 -  Choose your broad method: what method to use to capture the pattern
+
+Do you have a lot of data or not?
+Do you have a lot of computational time available or not?
+What type of data do you use (categorical, continuous, transformed)?
+
+See Figure 1 in \cite{mammola2021concepts}.
+
+4 - Optional: check whether your method works!
+
+Depending on some properties of your space, you can use simulations to see if it works!
+
+% Therefore we advise workers to test assumptions specifically about their trait space and ``play'' around with it before choosing their functional diversity/disparity framework and metric of interest.
+
+\bigskip
+\bigskip
+\hrule
+\hrule
 

### 2024-09-27 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 a pure birth speciation model) until reaching 200 elements in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
+This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
 These elements can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. 
@@
 This definition of null hypothesis is also appropriate for the empirical data if the question is the same.
-However, it is very likely that workers will ask a more exciting or intriguing question based on these data.
+However, it is very likely that researchers will ask a more exciting or intriguing question based on these data.
 For example, an interesting one could be ``Are the extinctions uniform across all birds in Hawaii?".

### 2024-09-27 Manuscript/supplementaries.tex
added glossary

@@
 \section{Supplementary materials}
+
+\subsection{Glossary}
+
+\begin{table}[ht]
+\caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2020shifting} and \cite{mammola2021concepts}.}
+\centering
+\begin{tabular}{p{4cm}p{5cm}p{5cm}p{3cm}}
+Mathematics & Ecology & Macroevolution & This paper \\
+\hline
+Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, traitspace, etc. & trait space \\
+\hline
+Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs etc. & observations \\
+\hline
+Columns ($d$) & Traits, Ordination scores, distances, etc. & Traits, ordination scores, distances, etc. & dimensions \\
+\hline
+Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional trait; e.g. geographical location & uni/multidimensional trait; e.g. landmark coordinates & trait \\
+\hline
+Rows subset ($m \times d$; $m \leq n$) & Treatments, phylogenetic group (clade), etc. & Clades, geological stratum, etc. & group \\
+\hline
+Statistic (i.e. a measure)  & Dissimilarity index or metric, hypervolume, functional diversity, etc. & Disparity metric or index & space occupancy measure \\
+\hline
+\end{tabular}
+\end{table}
 

### 2024-09-27 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
-a pure birth speciation model) until reaching 200 elements in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
+a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
 This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
-These elements can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. 
+These observations can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. %@@@ link to glossary
 This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
@@
 
-We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
+We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 observations).
 In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):
@@
 \item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$ (i.e.
-removing the $n$ pairs of points that are at least $D$ distance away from each other to remove the desired number of elements).
+removing the $n$ pairs of points that are at least $D$ distance away from each other to remove the desired number of observations).
 This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
@@
 an estimated number of categories summarising the distribution, like in a histogram).
-Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$. 
+Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of observations in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$. 
 For bandwidths with a low density (low values of $n$) we used $m>0$ to increase their resampling probability, and the opposite for bandwidths with a high density (high values of $n$; using $m<0$ to decrease sampling property).
@@
 This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g.
-for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
+for two spaces with 200 observations and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
 Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
@@
 \hline
-  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of elements.\\
+  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of observations.\\
   Equalising fitness & Size change (``size'') & When species with trait combinations that are more extreme (i.e.
@@
 \caption{\scriptsize{Illustration of the simulation protocol used in this paper.
-A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 elements resulting a trait space of 2, 4, or 8 dimensions.
+A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 observations resulting a trait space of 2, 4, or 8 dimensions.
 We then applied different stressors to the resulting trait space.
-We represent the elements removed by the stressors in blue and the ones kept in the trait space in orange.
-The first column illustrate the distribution of the elements in one dimension with the overlap between the removed (blue) and kept (orange) elements in brown (note we did not measure any metric on 1D spaces).
+We represent the observations removed by the stressors in blue and the ones kept in the trait space in orange.
+The first column illustrate the distribution of the observations in one dimension with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces).
 The second column illustrate the same distributions in 2 dimensions.
@@
 We used the following stressors:
-B) \textbf{null mechanism}: randomly removing half of the elements;
-C) \textbf{equalising mechanism}: removing elements on the edge of the distribution;
-D) \textbf{facilitation mechanism}: removing elements to reduce the distance between pairs of elements (increasing local density; decreasing evenness);
-E) \textbf{filtering mechanism}: removing elements on one extreme of the distribution;
-F) \textbf{competition mechanism}: proportionally removing elements on the centre of the distribution (``flattening the curve'').}}
+B) \textbf{null mechanism}: randomly removing a proportion of the observations;
+C) \textbf{equalising mechanism}: removing observations on the edge of the distribution;
+D) \textbf{facilitation mechanism}: removing observations to reduce the distance between pairs of observations (increasing local density; decreasing evenness);
+E) \textbf{observations mechanism}: removing observations on one extreme of the distribution;
+F) \textbf{competition mechanism}: proportionally removing observations on the centre of the distribution (``flattening the curve'').}}
 \label{Fig:simulations}
@@
 \citealt{guillerme2020shifting}).
-Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
+Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of observations in trait spaces in a non-random manner.
 Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.

### 2024-09-27 Manuscript/supplementaries.tex
Update supplementaries.tex

--- 2024-06-05 Manuscript/supplementaries.tex
@@
 \section{Supplementary materials}
+
+\subsection{Glossary}
+
+\begin{table}[ht]
+\caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2020shifting} and \cite{mammola2021concepts}.}
+\centering
+\begin{tabular}{p{4cm}p{5cm}p{5cm}p{3cm}}
+This paper & Ecology & Macroevolution & Definition \\
+\hline
+Trait space & Functional space, morphospace , etc. & Morphospace, traitspace, etc. & Matrix ($n \times d$) with a structural relation between rows and columns  \\
+\hline
+Observations & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs etc. & Rows ($n$) \\
+\hline
+Dimensions & Traits, Ordination scores, distances, etc. & Traits, ordination scores, distances, etc. & Columns ($d$) \\
+\hline
+Traits & uni/multidimensional trait; e.g. geographical location & uni/multidimensional trait; e.g. landmark coordinates & Columns subset ($b \times n$; $b \leq d$) \\
+\hline
+Group & Treatments, phylogenetic group (clade), etc. & Clades, geological stratum, etc. & Rows subset ($m \times d$; $m \leq n$) \\
+\hline
+Metric  & Dissimilarity index or metric, hypervolume, functional diversity, etc. & Disparity metric or index & Statistic (i.e. a measure) \\
+\hline
+Stressor & Introduction of invasive species, change of landscape use, climate change, polution, experimental design, etc.
+& Mass extinction, Tectonic change, climate change, etc.
+& An algorithm selecting a proportion of the observations.
+\end{tabular}
+\end{table}
 

### 2024-10-01 Manuscript/supplementaries.tex
Update supplementaries.tex

--- 2024-06-05 Manuscript/supplementaries.tex
@@
 \section{Supplementary materials}
+
+\subsection{Glossary}
+
+\begin{table}[ht]
+\caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
+\centering
+\begin{tabular}{p{2.5cm}p{4cm}p{5cm}p{5cm}}
+Term in this paper & definition & in ecology & in macroevolution \\
+\hline
+Trait space & Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, traitspace, etc.  \\
+\hline
+Observations & Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs etc. \\
+\hline
+Dimensions & Columns ($d$) & Traits, Ordination scores, distances, etc. & Traits, ordination scores, distances, etc. \\
+\hline
+Traits & Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional trait; e.g. geographical location & uni/multidimensional trait; e.g. landmark coordinates \\
+\hline
+Group & Rows subset ($m \times d$; $m \leq n$) & Treatments, phylogenetic group (clade), etc. & Clades, geological stratum, etc. \\
+\hline
+Metric & Statistic (i.e. a measure) & Dissimilarity index or metric, hypervolume, functional diversity, etc. & Disparity metric or index \\
+\hline
+Stressors & An algorithm removing a proportion of the observations & Introduction of invasive species, change of landscape use, climate change, pollution, experimental design, etc.
+& Mass extinction, Tectonic change, climate change, etc.
+\\
+\hline
+Stressors' strength & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction,  etc. \\
+\end{tabular}
+\end{table}
 

### 2024-10-01 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2 and 8 dimensional traits (supplementary materials).
-We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
+We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
 This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
@@
  & \cr 
- & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
+ & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where dimensions are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
  & \cr 
@@
 We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
-We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}.
-We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
+We extracted dimensions for extant species from the AVONET database \citep{tobias2022avonet}.
+We used 9 morphological dimensions associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
 Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.
@@
 For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
-In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \citep{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but is related to genetics (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
+In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \citep{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the dimensions we have considered here), but is related to genetics (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
 
@@
 
-The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to traits (Figure \ref{Fig:empirical_results}).
+The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to dimensions (Figure \ref{Fig:empirical_results}).
 This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g.

### 2024-10-01 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 
-We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 observations).
-In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):
+We then applied five different stressors to the trait space with different intensities.
+The five different stressors are effectively five specific algorithms removing either 20\%, 40\%, 60\% or 80\% of the data (the intensities of the stressors, resulting in trait spaces with, respectively, 160, 120, 80 and 40 observations).
+We used the following specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``increasing evenness'', were previously described in \cite{guillerme2020shifting}):
 
@@
 \item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius) $\rho$ of the centre of the trait space.
+$\rho$ is estimated for each trait space to match the required intensity.
+For example, for one trait space, the algorithm estimated $\rho_{20}$ to be the minimum radius excluding 20\% of the observations, $\rho_{40}$ the one removing 40\%, etc.
+(with $\rho_{20} > \rho_{40} > \rho_{60} > \rho_{80}$).
 This approximates our \textbf{equalising mechanism}.
 
-\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$ (i.e.
-removing the $n$ pairs of points that are at least $D$ distance away from each other to remove the desired number of observations).
+\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
+$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
+For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc.
+(with $D_{20} > D_{40} > D_{60} > D_{80}$).
+Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
 This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
 
-\item Shifting space: by removing the required amount of data from a distance (radius) $\rho$ of the element with the maximum value.
+\item Shifting space: by removing the required amount of data from a distance (radius) $\rho'$ of the observation with the maximum numerical value on all dimensions.
+This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
+In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
 This approximates our \textbf{filtering mechanism}.
 
-\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e.
+\item Increasing evenness: by resampling the proportion of data (i.e.
+20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities per bandwidth (i.e.
 an estimated number of categories summarising the distribution, like in a histogram).
@@
 
-For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
-These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
+For the decreasing size, increasing density and shifting space stressors, we estimated the parameters $\rho$, $D$ and $\rho'$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
+Note that the algorithm used here to simulate the stressors depend, to some extent, to the distribution of the data for each trait space.
+The algorithms used can lead to similar stressors effects depending on the simulated data distribution.
+For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their name suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit no identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors.)
+These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 intensities of data removal).
 We used 171 replicates per simulation because that was empirically the smallest number of simulations required to reach a variance between simulations lower than 1\% across all metrics (i.e.
@@
 We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
-This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
+This curse changes the properties of space occupancy in a non-linear way depending on the observation's distribution and the number of dimensions.
 For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
-However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
+However, the rate at which it approaches zero is not linear and depends on the distribution of each observation on every dimension.
 This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g.
@@
 
-\subsections{Results}
+\subsection{Results}
 
@@
 \citealt{guillerme2020shifting}).
-Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of observations in trait spaces in a non-random manner.
+Also, our results are likely affected by the use of simplified stressors designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of observations in trait spaces in a non-random manner.
 Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
@@
 \section*{Box 2: so which metric do \textit{I} choose?}
-\susbsection{So which metric to choose?}
+\subsection{So which metric to choose?}
 

### 2024-10-01 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 
-\item Increasing evenness: by resampling the proportion of data (i.e.
-20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities per bandwidth (i.e.
-an estimated number of categories summarising the distribution, like in a histogram).
-Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of observations in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$. 
-For bandwidths with a low density (low values of $n$) we used $m>0$ to increase their resampling probability, and the opposite for bandwidths with a high density (high values of $n$; using $m<0$ to decrease sampling property).
-This effectively flattened the normal distribution of the data by some factor $p$ (here we used $p=3$).
-This approximates our \textbf{competition mechanism}.
+\item Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
+In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
+This is done by selecting of $B$ discrete categories to summarise the distribution - i.e.
+bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0()} function in \texttt{R};][]{silverman1986density}.
+Each of these discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$.
+Where $p$ is factor increasing the scaling power of the algorithm (here we used $p=3$).
+In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
+This approximates our \textbf{competition mechanism}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
 \end{itemize}

### 2024-10-07 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 intensities of data removal).
-We used 171 replicates per simulation because that was empirically the smallest number of simulations required to reach a variance between simulations lower than 1\% across all metrics (i.e.
-any additional simulation added less that 1\% extra variance).
+We used 171 replicates because that was empirically the smallest number of replicates required to reach a variance between replicates lower than 1\% across all metrics (i.e.
+any additional replicates beyond 171 added less that 1\% extra variance).
 See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.
@@
 
-We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2 and 8 dimensional traits (supplementary materials).
+We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}) for 2 and 8 dimensional traits with not correlation between each dimensions in each simulation pipeline (supplementary materials).
 We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
@@
 This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g.
-for two spaces with 200 observations and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
+for two spaces with $200$ observations and $10$ dimensions generated in the exact same way, one might have a hypervolume of nearly $0$ and the other one of $10^5$).
 Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
@@
 \hline
-  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of observations.\\
-  Equalising fitness & Size change (``size'') & When species with trait combinations that are more extreme (i.e.
-species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
+  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations.\\
+  Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e.
+observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
 \\
-  Facilitation & Density change (``density'') & When species with closely shared trait combinations (i.e.
+  Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e.
 located close together in the trait space) are more likely to survive the stressor.
-In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
-  Environmental filtering & Position change (``position'') & When an increase in species trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}.
+In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g.
+a community of species) with shared traits rather than the opposite.\\
+  Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}.
 For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e.
 some regions of the trait space become unsuitable).\\
-  Competitive exclusion & Evenness change (``evenness'') & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; \citealt{hardin1960competitive}).
+  Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}).
 At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\
@@
 To ease interpretation of the results, we centred the results for the 12 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
-Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
+Effectively, for each metric and each single replicate, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each replicate.
 This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.

### 2024-10-07 Manuscript/main_ecography_R1.tex
up

@@
 We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}) for 2 and 8 dimensional traits with not correlation between each dimensions in each simulation pipeline (supplementary materials).
-We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
+We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait} - note that int the supplementary figures, we measured the \texttt{TPD} metrics on only the four first dimensions) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
 This curse changes the properties of space occupancy in a non-linear way depending on the observation's distribution and the number of dimensions.
@@
 
-We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts} using 12 metrics (Tab. \ref{Tab:metrics}):
+We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (trait space occupancy metrics; \cite{mammola2021concepts}) using 12 metrics (Tab.
+\ref{Tab:metrics}):
 1) Richness, encompassing metrics reflecting the sum of differences among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});
@@
 
-Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score $\sim$ amount of data removed.
-For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables 1 to 3 and whether the slope was clearly distinguishable from 0 (Figure \ref{Fig:simulation_results}).
+Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score as a function of the amount of data removed.
+For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables $1$ to $3$ and whether the slope was clearly distinguishable from $0$ (Figure \ref{Fig:simulation_results}).
 
@@
 We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
-A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
+A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at $1500$ CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
 We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
 We extracted dimensions for extant species from the AVONET database \citep{tobias2022avonet}.
-We used 9 morphological dimensions associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
-Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.
+We used $9$ morphological dimensions associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
+Traits were log transformed and standardised to a mean of $0$ and standard deviation of $1$, prior to analyses.
 
-The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
-55 species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
-For our analyses, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
-We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).
+The full trait dataset comprised $118$ bird species, all native species known to have existed on these islands over the last $125,000$ years.
+$55$ species are known to have gone extinct prior to $1500$ CE and $26$ after $1500$ CE.
+For our analyses, we focused on two time periods: (i) the avifauna present in $1500$ CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
+We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-$1500$ CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).
 
-To build the community trait space, we undertook a PCA including all 118 species, which we then subset to calculate the trait space for the historic and extant datasets.
-We selected the first 5 axes to represent the trait spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
-We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).
+To build the community trait space, we undertook a PCA including all $118$ species, which we then subset to calculate the trait space for the historic and extant datasets.
+We selected the first $5$ axes to represent the trait spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
+We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing $55$ and $81$ species for each stressor and scaling the results proportionally to this null mechanism (as described above).
 
@@
 The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test $p$ value $> 0.05$).
-Grey lines in the background across the distributions of different removal amounts represent the fitted linear model ``centred and scaled metric score $\sim$ amount of data removed''.
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model ``centred and scaled metric score function of the amount of data removed''.
 Dashed thin grey lines represent non-significant models ($p$ value of slope or/and intercept $> 0.05$).
@@
 Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).
-Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an increase of trait space occupancy (e.g.
-the functional evenness based on hypervolume for the equalising mechanism) or non-linear ones (e.g.
+Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted into a score increase (e.g.
+the functional evenness based on hypervolume for the equalising mechanism) or non-linear responses (e.g.
 Alpha diversity based on dendrogram for the equalising mechanism).
-This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
 Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
@@
 We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
-Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
+Our results show that different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
 Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
@@
 This could partly be due to the proportion of passerine extinctions in Hawaii.
-Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
-However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
+Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines (approximatively 58\%), the majority of global known anthropogenic extinctions (approx.
+75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
+However, in Hawaii, where the same proportion of known species were passerines (approx. 58\%), the opposite is true: the majority of extinctions (approx. 60\%) have been of passerine species.
 It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
@@
 
-Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (e.g.
-in \citealt{van2023should}).
+Furthermore, we only presented results based on a relatively small number of dimensions (up to 8).
+Although this number of dimensions is in the range of many studies in ecology (e.g.
+6 dimensions in \citealt{healy2019animal}), it is not uncommon to use a much greater number of dimensions, especially in palaeontology (e.g.
+more than 200 dimensions in \citealt{van2023should}).
 In a higher number of dimensions (usually $>$10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.

### 2024-10-08 Analysis/01_Simulating_spaces_and_stressors.Rmd
Update 01_Simulating_spaces_and_stressors.Rmd

@@
 plot.text.box("A) Tree and trait\nsimulation:", cex = 2)
-plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (1D)", col = c("lightgrey", "blue"), ylab = "Trait 1")
+plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (1D)", col = c(tips = "blue", "nodes" = "lightgrey"), ylab = "Trait 1")
 plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (2D)", col = c("lightgrey", "blue"), trait = c(1,2), xlab = "Trait 1", ylab = "Trait 2")

### 2024-10-08 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 observations resulting a trait space of 2, 4, or 8 dimensions.
+The first plot on row A illustrates one such simulation in one dimensions through time (with time on the horizontal axis and trait value for one dimension on the vertical axis).
+The second plot on the same row shows the same results but with 2 dimensions of the trait represented here.
 We then applied different stressors to the resulting trait space.
 We represent the observations removed by the stressors in blue and the ones kept in the trait space in orange.
-The first column illustrate the distribution of the observations in one dimension with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces).
-The second column illustrate the same distributions in 2 dimensions.
+The first column (histograms) illustrates the distribution of observations in one dimension of the 2D, 4D or 8D traits with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces).
+The second column illustrate the trait in 2 dimensions.
+For traits 
 We did not illustrate the distributions in 4 or 8 dimensions here.

### 2024-10-10 Analysis/01_Simulating_spaces_and_stressors.Rmd
Update 01_Simulating_spaces_and_stressors.Rmd

@@
 
-plot.text.box("C) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
+plot.text.box("C) Resampling\nwith stressor:\napproximating\nequalizing\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
@@
 
-plot.text.box("D) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
+plot.text.box("D) Resampling\nwith stressor:\napproximating\nfacilitating\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
@@
 
-plot.text.box("E) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
+plot.text.box("E) Resampling\nwith stressor:\napproximating\nfiltering\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
@@
 
-plot.text.box("F) Resampling\nwith stressor:\ncompetition\nmechanism", cex = 2)
+plot.text.box("F) Resampling\nwith stressor:\napproximating\ncompetition\nmechanism", cex = 2)
 plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D competition mechanism")

### 2024-10-10 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 The second column illustrate the trait in 2 dimensions.
-For traits 
-We did not illustrate the distributions in 4 or 8 dimensions here.
+We did not illustrate the data beyond 2 dimensions for traits in 4 or 8 dimensions.
 We used the following stressors:
 B) \textbf{null mechanism}: randomly removing a proportion of the observations;
-C) \textbf{equalising mechanism}: removing observations on the edge of the distribution;
-D) \textbf{facilitation mechanism}: removing observations to reduce the distance between pairs of observations (increasing local density; decreasing evenness);
-E) \textbf{observations mechanism}: removing observations on one extreme of the distribution;
-F) \textbf{competition mechanism}: proportionally removing observations on the centre of the distribution (``flattening the curve'').}}
+C) \textbf{approximating equalising mechanism}: removing observations on the edge of the distribution;
+D) \textbf{approximating facilitation mechanism}: removing observations to reduce the distance between pairs of observations (increasing local density; decreasing evenness);
+E) \textbf{approximating filtering mechanism}: removing observations on one extreme of the distribution;
+F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution (``flattening the curve'').}}
 \label{Fig:simulations}

### 2024-10-14 Analysis/04_Summarise_results.Rmd
updated analyses to include raw results

@@
 scaled_results_2d <- extract.table(all_results_2d, scale.method = "between")
-scaled_within_2d  <- extract.table(all_results_2d, scale.method = "within")
 scaled_results_3d <- extract.table(all_results_3d, scale.method = "between")
-scaled_within_3d  <- extract.table(all_results_3d, scale.method = "within")
 scaled_results_4d <- extract.table(all_results_4d, scale.method = "between")
-scaled_within_4d  <- extract.table(all_results_4d, scale.method = "within")
 scaled_results_5d <- extract.table(all_results_5d, scale.method = "between")
-scaled_within_5d  <- extract.table(all_results_5d, scale.method = "within")
 scaled_results_8d <- extract.table(all_results_8d, scale.method = "between")
-scaled_within_8d  <- extract.table(all_results_8d, scale.method = "within")
 scaled_results_emp<- extract.table(empirical_results, scale.method = "between")
-scaled_within_emp <- extract.table(empirical_results, scale.method = "within")
+
+## Extracting all the results but not centred (still scaled)
+raw_results_2d <- extract.table(all_results_2d, scale.method = "between", centre = FALSE)
+raw_results_3d <- extract.table(all_results_3d, scale.method = "between", centre = FALSE)
+raw_results_4d <- extract.table(all_results_4d, scale.method = "between", centre = FALSE)
+raw_results_5d <- extract.table(all_results_5d, scale.method = "between", centre = FALSE)
+raw_results_8d <- extract.table(all_results_8d, scale.method = "between", centre = FALSE)
+raw_results_emp<- extract.table(empirical_results, scale.method = "between", centre = FALSE)
 ```
@@
 ## The plot
-fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4), xlab = "scaled metric change")
 
@@
 ## Pdf version
-pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 11, width = 4.5)
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 11, width = 3.5)
 fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17))
@@
 Plain circles represents the metric for historic data (53 % - 63/118 species), lightly fainted squares for prehistoric data (46% - 55/118 species), and fainted triangles for extant data (31% - 37/118 species).
+
+## Raw results per metric
+
+```{r}
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_2d.pdf", height = 11, width = 8.5)
+fancy.plot(raw_results_2d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_4d.pdf", height = 11, width = 8.5)
+fancy.plot(raw_results_4d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_8d.pdf", height = 11, width = 8.5)
+fancy.plot(raw_results_8d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_emp.pdf", height = 11, width = 5.5)
+fancy.plot(raw_results_emp, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = c(19,15,17), xlab = "scaled metric", xlim = c(0,1))
+dev.off()
+```
 
@@
 points(x = c(2,3,4,5,8)+0.12, c(times_hypervolume[2,]), pch = 19, col = metric_cols[4])
-points(x = c(2,3,4,5,8)+0.16, c(times_probabilities[2,]), pch = 19, col = metric_cols[5])
+points(x = c(2,3,4)+0.16, c(times_probabilities[2,c(1:3)]), pch = 19, col = metric_cols[5])
 ## Linkings
@@
 lines(x = c(2,3,4,5,8)+0.12, c(times_hypervolume[2,]), pch = 19, col = metric_cols[4])
-lines(x = c(2,3,4,5,8)+0.16, c(times_probabilities[2,]), pch = 19, col = metric_cols[5])
+lines(x = c(2,3,4)+0.16, c(times_probabilities[2,c(1:3)]), pch = 19, col = metric_cols[5])
 ## 95% CI
@@
 }
-for(one_ci in 1:5) {
+for(one_ci in 1:3) {
   lines(y = c(times_probabilities[c(1,3), one_ci]), x = rep(dimensions[one_ci], 2)+0.16, col = metric_cols[5])

### 2024-10-14 Functions/extract.table.R

@@ -16,12 +16,23 @@
 ## Extract table
 extract.table <- function(results, centre = TRUE, scale.method) {
 
-
     # stop("DEBUG extract.table")
     # results <- all_results
     
     ## Centre all the results
     results <- lapply(results, centre.null, centre = centre)
+
+    if(!centre) {
+        ## Extract and merge the nulls
+        nulls <- do.call(rbind, lapply(results, `[[`, "null"))
+        colnames(nulls) <- gsub("random_", "", colnames(nulls)) 
+
+        ## Extract the stressors
+        results <- lapply(results, `[[`, "stressor")
+
+        ## Combine into a list of 5
+        results$nulls <- nulls
+    }
 
     ## Scale all the metrics between each stressor
     if(!missing(scale.method)) {
@@ -39,7 +50,7 @@ extract.table <- function(results, centre = TRUE, scale.method) {
     }
 
     ## Sort the results per metric
-    results <- lapply(results, sort.results.names, centre = centre)
+    results <- lapply(results, sort.results.names, centre)
 
     return(results)
 }
@@ -59,9 +70,12 @@ centre.null <- function(one_results, centre = TRUE) {
         centre_results <- lapply(pairs, function(pair, matrix) return(matrix[, pair[2]] - matrix[, pair[1]]), matrix = one_results$results_table)
         output <- do.call(cbind, centre_results)
     } else {
-        output <- one_results$results_table
+        ## Separate the nulls from the results
+        nulls <- one_results$results_table[, null_results]
+        stressors <- one_results$results_table[, -null_results]
+        colnames(stressors) <- unique(metrics_names)
+        output <- list("stressor" = stressors, "null" = nulls)
     }
-
     if(centre) {
         colnames(output) <- unique(metrics_names)
     } 
@@ -69,28 +83,28 @@ centre.null <- function(one_results, centre = TRUE) {
 }
 
 ## Sort the results per name
-sort.results.names <- function(results, centre) {
+sort.results.names <- function(one_results, centre) {
 
     ## Find the different removal levels
-    levels <- unique(gsub(".*_rm", "", colnames(results)))
+    levels <- unique(gsub(".*_rm", "", colnames(one_results)))
      
     if(centre) {   
         ## Find the metrics names    
-        metrics <- unique(gsub("_rm.*", "", colnames(results)))
[…]

### 2024-10-14 Functions/fancy.plot.R

@@ -18,7 +18,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, empirical = FALSE, ...) {
+fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, empirical = FALSE, xlab, xlim = c(-1, 0), ...) {
 
     ## Get the plot parameters
     plot_params <- list(...)
@@ -43,6 +43,8 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                            add.x = FALSE,
                            add.y = TRUE,
                            grid  = FALSE,
+                           xlab  = xlab,
+                           xlim  = xlim,
                            metric.names = metric.names)
 
             } else {
@@ -51,7 +53,9 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                 empty.plot(results[[one_stressor]],
                            plot.main = names(results)[one_stressor],
                            add.x = TRUE,
-                           add.y = FALSE)
+                           add.y = FALSE,
+                           xlim  = xlim,
+                           xlab  = xlab)
 
                 ## Adding the data
                 if(!missing(lm)) {
@@ -75,6 +79,8 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                    plot.main = names(results)[1],
                    add.x = TRUE,
                    add.y = TRUE,
+                   xlab  = xlab,
+                   xlim  = xlim,
                    metric.names = metric.names)
 
         ## Adding the data
@@ -217,7 +223,7 @@ add.one.stressor <- function(one_results, plot_params, probs, cent.tend, col, lm
 }
 
 ## Creating an empty plot
-empty.plot <- function(one_results, scaled = TRUE, add.x = FALSE, add.y = FALSE, grid = TRUE,  plot.main, metric.names = NULL) {
+empty.plot <- function(one_results, scaled = TRUE, add.x = FALSE, add.y = FALSE, grid = TRUE,  plot.main, metric.names = NULL, xlab, xlim) {
     
[…]

### 2024-10-14 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 This approximates our \textbf{null mechanism}.
+This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly.
 
@@
 E) \textbf{approximating filtering mechanism}: removing observations on one extreme of the distribution;
-F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution (``flattening the curve'').}}
+F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution.}}
 \label{Fig:simulations}
@@
 
-\subsubsection{Scaling the results to be proportional to the random removals (null)}
+\subsubsection{Comparing and scaling the results to the random removals (null stressor)}
 
-To ease interpretation of the results, we centred the results for the 12 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
-Effectively, for each metric and each single replicate, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each replicate.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
+$metric_stressor_20\% - metric_null_20\%$).
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).
+In other words, for each metric and each single replicate, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each replicate.
 This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
@@
 The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
-Negative and positive values signify a decrease/increase in the metric score.
+Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
 The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
@@
 Grey lines in the background across the distributions of different removal amounts represent the fitted linear model ``centred and scaled metric score function of the amount of data removed''.
-Dashed thin grey lines represent non-significant models ($p$ value of slope or/and intercept $> 0.05$).
-The exact model results are available in supplementary table @@@.
+Dashed thin grey lines represent non-significant linear models (slope's $p$ value $> 0.05$).
+The model results are available in supplementary table 3.
 Similar figures are available in the supplementary materials for 2 and 8 dimensions.}}
@@
 Plain circles represent the metrics for the historic dataset (46\% removal) and fainted squares for the extant dataset (69\% removal).
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
 }}

### 2024-10-14 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 \hline
-Stressors' strength & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction,  etc. \\
+Stressors' intensity & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction,  etc. \\
 \end{tabular}
@@
 }}
-\label{Fig:simulation_results}
+\label{Fig:simulation_results_2d}
 \end{figure}
@@
 Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
 }}
-\label{Fig:simulation_results}
+\label{Fig:simulation_results_8d}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_2d.pdf}
+\caption{\scriptsize{\textbf{Raw metric scores for 2 dimensions:} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the scaled metric scores.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+}}
+\label{Fig:raw_results_2d}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_4d.pdf}
+\caption{\scriptsize{\textbf{Raw metric scores for 4 dimensions:} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the scaled metric scores.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+}}
+\label{Fig:raw_results_4d}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_2d.pdf}
+\caption{\scriptsize{\textbf{Raw metric scores for 8 dimensions:} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the scaled metric scores.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+}
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
+}
+\label{Fig:raw_results_8d}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_emp.pdf}
+\caption{\scriptsize{\textbf{Raw metric scores for the empirical data:} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the scaled metric scores.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
+}}
+\label{Fig:raw_results_emp}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/time_per_metric}
+\caption{\scriptsize{\textbf{Computational time per metric:} log time (seconds) distribution per metric.
+The dots correspond to the median computational time for one metric for $n$ elements (40, 80, 120, 160) for $D$ dimensions (2, 3, 4, 5, 8).
+And colour vertical lines represent the 95\% distribution range for each measurement.
+Note that for the probability density metrics (\texttt{TPD} package), the number of dimensions is restricted to 4.
+The horizontal grey lines are indicating the log seconds time equivalent in other units (seconds, minutes, hours) for facilitating interpretation.
+}}
+\label{Fig:metrics_time}
 \end{figure}

### 2024-10-22 Analysis/03_Empirical_pipeline.Rmd
Update 03_Empirical_pipeline.Rmd

@@
 Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.
+
+## Empirical example for box 2
+
+```{r}
+library(dispRity)
+## Plotting the trait space
+data <- custom.subsets(empirical_traitspace, group = lapply(groups, which))
+pdf(file = "../Manuscript/Figures/empirical_example.pdf", height = 4, width = 4)
+plot(data, legend = FALSE, col = c("blue", "orange"))
+legend("topright", pch = 19, col = c("blue", "orange"), legend = c("Current species", "Pre 1500 species"))
+dev.off()
+
+## Measuring disparity
+library(xtable)
+xtable(summary(dispRity(data, metric = convhull.volume)))
+xtable(summary(dispRity(data, metric = displacements)))
+```
+
+## Empirical number of dimensions selection
+
+```{r}
+library(dispRity)
+pca <- readRDS("../Data/Processed/pca-object-historic-community.RDS")
+data <- custom.subsets(pca$x, group = list(extant = extant_species, pre_1500 = historic_species))
+pdf(file = "../Manuscript/Figures/empirical_data_dimensions.pdf", height = 8, width = 4)
+plot(select.axes(data))
+dev.off()
+```

### 2024-10-22 Functions/fancy.plot.R
Update fancy.plot.R

@@ -18,7 +18,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, empirical = FALSE, xlab, xlim = c(-1, 0), ...) {
+fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, empirical = FALSE, xlab, xlim = c(-1, 1), ...) {
 
     ## Get the plot parameters
     plot_params <- list(...)

### 2024-10-22 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and \ref{Tab:mechanisms}).
-To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands (process) resulting in a trait space modified by extinctions (the pattern) capturing the effects of species extinction on species trait distributions (the mechanism).
-We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
-We suggest caution when summarising patterns in observed data and propose tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
+To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.
+We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.
+We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
 
@@
 \centering
-   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
+   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol_explained.pdf}
 \caption{\scriptsize{Illustration of the simulation protocol used in this paper.
@@
 E) \textbf{approximating filtering mechanism}: removing observations on one extreme of the distribution;
-F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution.}}
+F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution.
+G) we presented the results of the simualtions as the scaled metric scores for the approximated mechanisms (C to F) relative to the null mechanism.
+The dot represents the median metric score and the thick and dashed lines respectively the 50\% and 95\% confidence intervals.}}
 \label{Fig:simulations}
@@
 
-\subsubsection{Metrics for measuring trait space occupancy dissimilarity/disparity:}
+\subsubsection{Metrics for measuring trait space occupancy (\textit{aka} function diversity, dissimilarity, disparity):}
 
@@
 
-To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
-$metric_stressor_20\% - metric_null_20\%$).
-We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).
-In other words, for each metric and each single replicate, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each replicate.
-This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
-In parallel, we also compared the distribution of the 171 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between $0$ and $1$.
+We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.
+$metric_{stressor20\%} - metric_{null20\%}$).
+This resulted in each metric being scaled between $–1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
+approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \ref{Fig:simulations}, \ref{Fig:simulation_results} and \ref{Fig:empirical_results}).
+We interpret this last scenario as the metric score performing poorly, i.e. it is unable to clearly distinguish between a random removal of elements and a non random one.
+To assess these poor performances, we compared the distributions of the $171$ scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.
 The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
@@
 
-Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score as a function of the amount of data removed.
+Finally, we measured the effect of the different strengths of the stressors (i.e.
+removing 20\%, 40\%, 60\% and 80\% the data) by fitting a linear regression using the scaled metric difference to null score as a function of the amount of data removed.
 For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables $1$ to $3$ and whether the slope was clearly distinguishable from $0$ (Figure \ref{Fig:simulation_results}).
+This allowed us to see if there was any predictable linearity in the metric scores.
+In other words testing whether the metric scores relative to null removals where linearly scaling with the amount of data removed.
 
@@
 
+Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
+For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
+facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
+
 \bigskip
@@
 \section*{Box 2: so which metric do \textit{I} choose?}
-\subsection{So which metric to choose?}
 
@@
 
-Following our results and these thoughts, we encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
-We encourage workers to test their metrics using simplified example datasets to see if they have the ability to capture at least crude changes as exemplified in this work (e.g.
-using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package \citealt{guillerme2020shifting}).
-Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
-For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
-facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
+Here is a proposed step by step protocol that could help one decide what to measure by clearly specifying their mechanism, process and pattern of interest prior to any functional diversity or disparity measurements.
 
-% Maybe use the empirical example here.
+\textit{We have a data of bird traits for native species in Hawaii through time before and after 1500.
+We want to know how anthropogenic stresses (hunting, invasive species, land use change) have affected the trait diversity in native Hawaiian birds.}
 
-1 - Which questions to ask yourself?
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/empirical_example.pdf}
+\caption{Example of a traitspace based on the empirical data described above.}
+\label{Fig:trait_space_example}
+\end{figure}
+\bigskip
 
-- Why? Think of the broad biological mechanism of interest. For example: ``we are interested in inter specific competition''.
-- How?
-With the data you have at hand, identify the process you could analyse.
-For example ``we have data collected on two islands, one control with only one species and another one with two species''.
-- What? Then select your pattern and test whether it could answer your question.
+\subsubsection{Frame the question:}
 
-2 - Choose your broad type of metric: what pattern do you want to capture?
+\begin{itemize}
 
--Is one group bigger than an other? -> Richness
--Is one group more dispersed that the other? -> Divergence
--Is one group more evenly dispersed in the space than another? -> Regularity 
--Is it something else -> There are many many more subtleties to these broad distinctions in the literature!
+    \item \textbf{Why?} Think of the broad biological mechanism of interest.
+In our example, this could be asking why (due to what mechanism) has the community changed: \textit{the trait diversity has changed due to a filtering mechanism where anthropogenic stressors affect specific species with specific traits (e.g.
+through hunting, predation on eggs by mammals, etc...).}
+    \item \textbf{How?} With the data we have at hand, identify the process we could analyse.
+In our example, this could be native bird species traits through time: \textit{how did filtering affect the native species after 1500?}
 
-3 -  Choose your broad method: what method to use to capture the pattern
+\end{itemize}
 
+Although asking \textbf{why} and \textbf{how} is a relatively standard step in the literature, we suggest that proposing them in this framework more easily leads to the next, often less asked question of what pattern to capture:
+
+\begin{itemize}
+    \item \textbf{What?} Choose the appropriate pattern that would capture changes through time in relation to a filtering mechanism.
+In our example, this could be a continuous metric that measures either changes in the trait space position or/and changes in the trait space side.
+The idea being that filtering would remove either the edges or a specific portion of the trait space without affecting to many other parts of the trait space.
+\end{itemize}
+
+\subsubsection{Find the type of metrics of interest:}
+
+Once the one has clearly defined the pattern they want to capture (\textbf{what}), we believe it gets easier to pick up the type of metric that would be the most adequate.
+For example, we can follow the suggestions from \cite{mammola2021concepts} (Table \ref{Tab:box2}).
+
+\begin{table}
+\center
+\begin{tabular}{c | c | c}
+\textbf{Metric type}  & \textbf{Question} & \textbf{Are we interested?}\\
+\hline
+Size & Is one group bigger than another one? & Yes \\
+Divergence & Is one group more spread than another one? & Not really \\
+Regularity & Is one group more evenly disperse in space? & Not really \\
+Something else  & Does the position of a group changes in space? & Yes \\
+\end{tabular}
+\caption{\textbf{Example of questions we can ask with the different metric types.}The categories are arbitrary and are taken from \cite{mammola2021concepts}}
+\label{Tab:box2}
+\end{table}
+
+\subsubsection{Get the metric most appropriate to your data:}
+
+So to describe the pattern for our specific example question, we will be interested in size and position of the trait space.
+There is a lot of different metrics that capture changes in size and in position of the trait space and their choice should depend on the following questions:
 Do you have a lot of data or not?
@@
 What type of data do you use (categorical, continuous, transformed)?
+Are you familiar with the metric (is there a software implementation you can use)?
+Of course, these are example questions and are not necessary the only ones you should ask.
+\cite{mammola2021concepts} proposes a very handy flow chart.
 
-See Figure 1 in \cite{mammola2021concepts}.
+Here following our experience and the nature of the data, we will measure the absolute size of the trait space using the convex hull size and the potential displacement of the trait space using the displacement metric \citep{guillerme2020shifting}.
+Results are displayed in tables \ref{Tab:results_convhull} and \ref{Tab:displacement}.
 
-4 - Optional: check whether your method works!
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrr}
+  \hline
+ group & n & convex hull volume \\ 
+  \hline
+ current &  37 & 25.44 \\ 
+ pre 1500 &  63 & 60.84 \\ 
+   \hline
+\end{tabular}
+\caption{Trait space size measured as the convex hull volume for both groups.}
+\label{Tab:results_convhull}
+\end{table}
 
-Depending on some properties of your space, you can use simulations to see if it works!
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrr}
+  \hline
+ group & n & median displacement & 2.5\% & 25\% & 75\% & 97.5\% \\ 
+  \hline
+ current &  37 & 1.04 & 0.95 & 0.99 & 1.05 & 1.10 \\ 
+ pre 1500 &  63 & 1.12 & 0.85 & 0.94 & 1.19 & 1.34 \\ 
+   \hline
+\end{tabular}
+\caption{Trait space position distribution measured as the displacement metric: the percentages are the confidence intervals of the distribution.}
+\label{Tab:displacement}
+\end{table}
 
-% Therefore we advise workers to test assumptions specifically about their trait space and ``play'' around with it before choosing their functional diversity/disparity framework and metric of interest.
+Note that it is also possible to test whether the metric is actually capturing the broad mechanism of interest using evolutionary simulations (e.g.
+by simulating evolutionary scenarios that would include species traits that would go through a filtering mechanisms - \citealt{guillerme2024treats}) or, like in this manuscript, using stressors (e.g.
+by using the \texttt{moms} interactive package \citealt{guillerme2020shifting}).
 

### 2024-10-22 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 \centering
-   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_emp.pdf}
+   \includegraphics[width=0.6\textwidth]{Figures/raw_results_per_metric_per_stressor_emp.pdf}
 \caption{\scriptsize{\textbf{Raw metric scores for the empirical data:} the y axes represent the different metrics tested (sorted by categories).
@@
 \label{Fig:raw_results_emp}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics{Figures/empirical_data_dimensions.pdf}
+\caption{\scriptsize{\textbf{Selection of the number of dimensions for the empirical data:} each histograms corresponds to the variance per dimension (dark grey) and the cumulative variance per dimension (light grey) for each group in the empirical data: the 37 currently endemic Hawaiian species (extant species) the 63 species present before 1500 (pre-1500 group).
+The plot shows that although 4 dimensions are enough to describe at least 95\% of the variance for the extant group, 5 dimensions are required for the pre-1500 group (following the method proposed in \cite{guillerme2023innovation}).
+}}
+\label{Fig:empirical_dimensions}
 \end{figure}

### 2024-10-23 Manuscript/title_page.tex
Create title_page.tex

--- 2024-06-05 Manuscript/main.tex
@@
 
-\noindent (Keywords: Functional diversity, dissimilarity, disparity, patterns, processes, mechanisms)\\
-
-\begin{itemize}
-   \item Article type: Letter
-   \item Number of words in the abstract: 189
-   \item Number of words in the main text: 3960
-   \item Number of words in the text box: 419
-   \item Number of references: 54
-   \item 3 Figures, 2 tables and 1 text box
-\end{itemize}
-
-\subsection{Statement of authorship}
-
-TG, PC, MWJ, SM and TJM designed the study, analysed the data and wrote the manuscript. TG wrote the code to analyse the data.
-
-\subsection{Data accessibility statement}
-
-All code and data required to reproduce this analysis is available on @@@
-
-%---------------------------------------------
-%
-%       ABSTRACT
-%
-%---------------------------------------------
-
-\newpage
-%Line numbering
-\modulolinenumbers[1]
-\linenumbers
-
-\section{Abstract}
-
-Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
-Indeed, functional diversity metrics allow researchers to summarize complex occupancy patterns in space and/or time (what is changing?) that lead to changes in communities and/or populations (the process; how is it changing?) in response to some stressors (the mechanism; why is it changing?).
-However, as the diversity of functional diversity metrics and methods increases, it is often not directly clear which metric is more readily appropriate for which question.
-We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms (environmental filtering, competitive exclusion, equalizing fitness, and facilitation) in community ecology.
-Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
-We show that different metrics perform better for different tasks, emphasizing the importance of not using a one-size-fits-all metric.
-Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.
-
-\newpage
-
-\section{Introduction}
-
-In the last two decades, there has been a progressive expansion in ecology and evolution studies from taxonomic-oriented approaches, with species as the focal point, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
-Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g.
-morphological, anatomical, ecological, physiological, behavioural, phenological, etc.) measured on individual organisms at any level, from genes to whole organisms.
-A trait-based approach has two advantages for answering core questions in ecology and evolution.
-First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than the organisms themselves.
-Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g.
-unrelated species pools) and time (e.g.
-anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
-Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g.
-agronomy \citealt{martin2015plant}).
-This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).
-
-\bigskip
-\bigskip
-\hrule
-\hrule
-
-\section*{Box 1}
-\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
-This framework can be effectively applied to trait-based analyses:
-
-\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
-Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space.
-For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
-For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.
-
-\textit{How?} Process description can be seen as linking the pattern of interest (``What?'') to some dynamic element.
-This can be a punctual change.
-For example, change of the pattern under a certain condition, such as how traits X differ between two habitats \citep{martinez2021habitat}.
-But it can also be one or more continuous or ordinal changes, such as how the pattern X changes in space, time and/or along ecological gradients \citep{belmaker2013spatial, jarzyna2018taxonomic,lamanna2014functional,bjorkman2018plant,mclean2021trait}.
-The distinction here might seem trivial but it is significant: usually, the process designates the change of the pattern, not the change of the traits.
-Although the traits are what is really changing, researchers will usually analyse some emergent property of the trait aggregation as described by the change of statistics between two or more conditions.
-
-\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
-In most cases, this is what researchers are actually trying to understand.
-For example, one might be interested in understanding the effect of climate change on some trait \citep{boonman2022trait}.
-In fact, most researchers work on understanding the causal link between variables (i.e. ``why'' are ``what'' and ``how'' linked).
-This can be very useful, for example, to provide predictions about the past or the future.
-Ultimately, studying the mechanisms (``why'') is often the reason why researchers get funded (or not).
-
-Note that the distinction between what, how and why is not categorical in its nature and is often nuanced, with patterns and process or process and mechanisms sometimes being used interchangeably to describe the same questions.
-
-\bigskip
-\hrule
-\hrule
-\bigskip
-
-While the first part of this framework (``what?'') has been studied extensively in the last decade, this was often undertaken outside of the ``what, how and why'' framework.
-Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
-The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
-Using an inappropriate metric for describing a pattern can lead to biased conclusions.
-For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?'').
-In such a scenario, measuring the occupied area (i.e.
-$km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
-In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
-See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.
-
-Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations}).
-We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands (process) resulting in a trait space modified by extinctions (the pattern) capturing the effects of species extinction on species trait distributions (the mechanism).
-We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
-We suggest caution when summarising patterns in observed data and propose tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
-
-\section{Methods}
-\subsection{Simulating trait space patterns}
-
-First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.) in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
-This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
-We refer to this as the ``non-stressed trait space''.
-
-\subsubsection{Applying stressors to the trait space}
-
-We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
-In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):
-
-\begin{itemize}
-\item Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data.
-This approximates our \textbf{null mechanism}.
-
-\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius)  $\rho$ of the centre of the trait space.
-This approximates our \textbf{equalising mechanism}.
-
-\item Increasing density: by removing the required amount of pairs of points with a pair distance of at least $D$ (i.e.
-removing the $n$ pairs of points that are at least $D$ distance away from each other).
-This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
-
-\item Shifting space: by removing the required amount of data from a distance (radius) $\rho$ of the element with the maximum value.
-This approximates our \textbf{filtering mechanism}.
-
-\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e.
-an estimated number of categories summarising the distribution, like in a histogram).
-Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and $p$ is a factor increasing the flattening/steepening (here we used $p=3$).
-This approximates our \textbf{competition mechanism}.
-\end{itemize}
-
-For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
-These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
-We used 171 replicates per simulation because that was empirically the smallest number of simulations required to reach a variance between simulations lower than 1\% across all metrics (i.e.
-any additional simulation added less that 1\% extra variance).
-See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.
-
-Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
-For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
-However, if the data are distributed uniformly, this does not happen.
-Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
-However, this is not tackled here for both simplicity and to understand how they work in isolation.
-
-We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2 and 8 dimensional traits (supplementary materials).
-We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
-This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
-For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
-However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
-This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g.
-for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
-Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
-This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.
-
-\begin{table}
-\center
-\scriptsize
-\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
-\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
-\hline
-  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of elements.\\
-  Equalising fitness & Size change (``size'') & When species with trait combinations that are more extreme (i.e.
-species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
-\\
-  Facilitation & Density change (``density'') & When species with closely shared trait combinations (i.e.
-located close together in the trait space) are more likely to survive the stressor.
-In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
-  Environmental filtering & Position change (``position'') & When an increase in species trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}.
-For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e.
-some regions of the trait space become unsuitable).\\
-  Competitive exclusion & Evenness change (``evenness'') & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; \citealt{hardin1960competitive}).
-At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\
-
-    \end{tabular}
-    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it.
-We also provide a more detailed description of the stressor.
-Note that in biological data, we don't expect any of the mechanisms to act on communities alone (e.g.
-equalising fitness and facilitation can both act on the distribution of species traits).
-Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution).
-These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
-}
-    \label{Tab:mechanisms}
-\end{table}
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
-\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
-A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting a trait space of 2, 4, or 8 dimensions.
-We then applied different stressors to the resulting trait space using the following:
-B) \textbf{null mechanism}: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
-C) \textbf{equalising mechanism}: removing species on the edge of the distribution;
-D) \textbf{facilitation mechanism}: removing species to reduce the distance between pairs of species (increasing local density; decreasing evenness);
-E) \textbf{filtering mechanism}: removing species on one extreme of the distribution;
-F) \textbf{competition mechanism}: proportionally removing species on the centre of the distribution (``flattening the curve'').}}
-\label{Fig:simulations}
-\end{figure}
-\bigskip
-
-\subsubsection{Metrics for measuring trait space occupancy dissimilarity/disparity:}
-
-We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
-1) Richness, encompassing metrics reflecting the sum of differences among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});
-2) Divergence, encompassing metrics reflecting the average differences among observations (4 metrics); and
-3) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting}).
-Note that we focused here on three packages fully devoted to functional diversity analyses (\texttt{BAT} \citealt{cardoso2015bat} \texttt{FD} \citealt{Laliberte2010FD} and \texttt{TPD} \citealt{carmona2019trait}) in the \texttt{R} statistical environments (see \citealt{mammola2021concepts}).
-
-\begin{table}
-\center
-\scriptsize
-\begin{tabular}{p{0.1\linewidth}|p{0.15\linewidth}|p{0.15\linewidth}|p{0.2\linewidth}|p{0.15\linewidth}|p{0.1\linewidth}}
-
-\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
-\hline
-Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
- & \cr 
- & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
- & \cr
- & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
- & \cr
- & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
-\hline
-Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
- & \cr 
- & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
- & \cr 
- & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
- & \cr  
- & Distances & Rao's quadratic entropy & The mean Euclidean distance between every two random pairs of species & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
-\hline
- & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
-the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
- & \cr 
- & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
- & \cr 
- & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
- & \cr 
- & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
-\hline
-
-\end{tabular}
-\caption{\scriptsize{\textbf{Functional diversity metrics tested in our simulations.}}
-}
-\label{Tab:metrics}
-\end{table}
-
-\subsubsection{Scaling the results to be proportional to the random removals (null)}
-
-To ease interpretation of the results, we centred the results for the 12 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
-Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
-This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
-In parallel, we also compared the distribution of the 171 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
-The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).
-
-\subsubsection{Measuring the effect of the strength of the stressor}
-
-Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score $\sim$ amount of data removed.
-For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables 1 to 3 and whether the slope was clearly distinguishable from 0 (Figure \ref{Fig:simulation_results}).
-
-\subsection{Empirical data}
-
-We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
-A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
-We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
-We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}.
-We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
-Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.
-
-The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
-55 species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
-For our analyses, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
-We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).
-
-To build the community trait space, we undertook a PCA including all 118 species, which we then subset to calculate the trait space for the historic and extant datasets.
-We selected the first 5 axes to represent the trait spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
-We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).
-
-\section{Results}
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d.pdf}
-\caption{\scriptsize{\textbf{Simulation results:} the y axes represent the different metrics tested (sorted by categories).
-The different columns represent the different stressors.
-The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
-Negative and positive values signify a decrease/increase in the metric score.
-The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
-The colours are here to visually separate the metrics by categories (green = richness, yellow = divergence, blue = regularity); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
-The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test $p$ value $> 0.05$).
-Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
-Dashed thin grey lines represent non-significant models ($p$ value of slope or/and intercept $> 0.05$).
-Similar figures are available in the supplementary materials for 2 and 8 dimensions.}}
-\label{Fig:simulation_results}
-\end{figure}
-\bigskip
-
-The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.
-It ranged from metrics capturing no clear pattern (e.g.
-the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g.
-Divergence based on probability density for the equalizing mechanism for a decrease in space occupancy).
-Some metrics even captured an opposite pattern to what was expected: for all metrics, we naively expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
-the functional evenness based on hypervolume for the equalizing mechanism) or non-linear ones (e.g.
-Alpha diversity based on dendrogram for the equalizing mechanism).
-This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
-Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results})!
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
-\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (green = richness, yellow = divergence, blue = regularity).
-Plain circles represent the metrics for the historic dataset (46\% removal) and fainted squares for the extant dataset (69\% removal).
-}}
-\label{Fig:empirical_results}
-\end{figure}
-\bigskip
-
-For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern.
-For example, all richness metrics (green; Figure \ref{Fig:empirical_results}) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
-However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g.
-functional dispersion based on dendrograms or hypervolumes).
-Finally, some metrics pick up a decrease for both datasets (e.g. functional evenness based on hypervolumes - although the 95\% confidence intervals overlapped with 0).
-Note that this apparent global increase in most metric values may appear counter intuitive given that 46\% and 69\% of species went extinct over the two time periods.
-This is due to the scaling of the metrics compared to random equivalent extinctions.
-That is, Figure \ref{Fig:empirical_results} is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.
-
-More broadly, as can be seen from this empirical case, patterns are likely to be much less clear than with simulated data due to multiple stressors acting simultaneously.
-For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
-In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \citep{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but is related to genetics (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
-
-\section{Discussion}
-
-We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
-Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
-Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
-
-Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many others could have been used (e.g.
-\citealt{guillerme2020shifting}).
-Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
-Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
-The simulated trait spaces in 2, 4 or 8 dimensions had the same variance on all dimensions (i.e.
-we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g.
-PCA or PCO), the dimensions have by definition a decreasing variance.
-
-Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (e.g.
-in \citealt{van2023should}).
-In a higher number of dimensions (usually $>$10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
-This sometimes makes metric interpretation more chaotic \cite{bellman1957dynamic} in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
-In fact, the relationship between observations has an important effect on how our results scale with higher dimensionality (e.g.
-the volume of an evenly occupied space scales differently in high dimensions compared to the same space but with normally distributed observations).
-Therefore we advise workers to test assumptions specifically about their trait space and ``play'' around with it before choosing their functional diversity/disparity framework and metric of interest.
-
-Another cautionary note could be made about our choice of scaling the metrics relative to a random trait space reduction (with the same percentage of reduction).
-This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern \citep{bausman2018modeling}.
-For example, if removing 80\% of the edges of a distribution (equalising mechanism) was distinguishable or not from removing 80\% of a distribution randomly (null mechanism).
-Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).
-In other words, our statistical question was ``Does metric X distinguish between removing N\% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?''.
-This definition of null hypothesis is also appropriate for the empirical data if the question is the same.
-However, it is very likely that workers will ask a more exciting or intriguing question based on these data.
-For example, an interesting one could be ``Are the extinctions uniform across all birds in Hawaii?".
-This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
-%@@@ TODO: add supplementary
-Furthermore, the scaling of the metrics paired with the levels of removal (80\%, 60\%, 40\% and 20\%) creates an expected artifact similarity when comparing the random removals against the other ones: when less data is available overall, the metrics are more likely to be similar (e.g.
-if removing 99\% of the data, we expect the random and removal metric score to be nearly identical to the non-random ones).
-
-The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to traits (Figure \ref{Fig:empirical_results}).
-This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g.
-\citealt{sayol2021loss,Matthews2022}).
-This could partly be due to the proportion of passerine extinctions in Hawaii.
-Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
-However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
-It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
-But these results, given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines,  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
-relative to random extinctions).
-In addition, it is worth stressing that these analyses included marine species (which constitute less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.
-
-Following our results and these thoughts, we encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
-We encourage workers to test their metrics using simplified example datasets to see if they have the ability to capture at least crude changes as exemplified in this work (e.g.
-using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package \citealt{guillerme2020shifting}).
-Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
-For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
-facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
-
-\section{Conclusion}
-Different ecological and evolutionary processes or mechanisms do not always results in different patterns.
-Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (\textit{aka} disparity, dissimilarity or functional diversity metrics).
-Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the data, the mechanism or/and process of interest.
-Using space occupancy metrics for describing functional diversity is tricky.
-This is because the metrics have properties that can be counter intuitive based on the data at hand.
-And also because functions of ecosystems or organisms and niches within it are hard to capture/understand/define.
-This means that measuring functional diversity is based on both the definition of the metric and the function of interest and thus cannot be served by a one-fits-all metric.
-In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.
-
-\section{Repeatability and reproducibility}
-The simulations, results figures and tables are entirely reproducible via \url{https://github.com/TGuillerme/eco_metrics_simulations}.
-An additional vignette on how to choose metrics is available at \url{https://github.com/TGuillerme/eco_metrics_simulations/blob/master/Analysis/Choosing_metrics_vignette.Rmd}.
-
 \section{Acknowledgements}
@@
 
-\bibliographystyle{sysbio}
-\bibliography{references}
+\subsection{Conflict of interest}
+We declare no conflict of interest.
+
+\subsection{Authors contributions}
+TG, PC, MWJ, SM and TJM designed the study, analysed the data and wrote the manuscript. TG wrote the code to analyse the data.
+
+\subsection{Data accessibility statement}
+The simulations, results figures and tables are entirely reproducible via \url{https://github.com/TGuillerme/eco_metrics_simulations}.
+An additional vignette on how to choose metrics is available at \url{https://github.com/TGuillerme/eco_metrics_simulations/blob/master/Analysis/Choosing_metrics_vignette.Rmd}.
 

### 2024-10-30 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
-We show that different metrics perform better for different tasks, emphasising the importance of not using a one-size-fits-all metric.
+We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).
+For example the competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
+This emphasises the importance of not using a one-size-fits-all metric.
 Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.
@@
 Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and \ref{Tab:mechanisms}).
+Our goal is to the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
+We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).
 To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.
 We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.
-We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
 
@@
 \scriptsize
-\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
-\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
+\begin{tabular}{p{0.2\linewidth}|p{0.2\linewidth}|p{0.4\linewidth}|p{0.2\linewidth}}
+\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description} & \textbf{Type of metric expected to recover the mechanism}\\
 \hline
-  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations.\\
+  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations.
+& Metrics sensitive to the amount of data \\
   Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e.
 observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
-\\
+& Metrics capturing changes in richness (but not divergence and regularity) \\
   Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e.
@@
 In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g.
-a community of species) with shared traits rather than the opposite.\\
+a community of species) with shared traits rather than the opposite.
+& Metrics capturing changes in regularity (but not richness and regularity) \\
   Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}.
 For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e.
-some regions of the trait space become unsuitable).\\
+some regions of the trait space become unsuitable).
+& Metrics capturing changes in position (and richness) \\\\
   Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}).
-At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\
+At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}).
+& Metrics capturing changes in divergence (but not richness and regularity) \\\\
 
@@
 equalising fitness and facilitation can both act on the distribution of species traits).
-Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution).
+Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution), thus, the type of metric expected to recover the mechanism can be variable.
 These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
@@
 \hline
- & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
+Regularity & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
 the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
@@
 
+To understand the ability of each metric to capture a mechanism of interest, we compared their scores relatively to the same metric scores of a null mechanism.
 To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between $0$ and $1$.
@@
 Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
+We found that different metrics can capture changes in pattern relative to the null mechanisms in different ways depending on the approximated mechanism of interest.
+For example, for the approximated equalising mechanism, most metrics capture a pattern clearly smaller than the one from the null mechanism (i.e.
+randomly removing elements has more effects on the metric than removing them following the mechanism).
+However, some metrics (e.g. the functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
+This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.
+
+For example, say one is interested in looking at changes in functional diversity using a divergence metric using the probability density method (e.g. \texttt{TPD::REND}).
+They can then apply this metric to their trait space and get two relative metric scores of, says, 0.7 and 0.65 (pattern) for two levels of extinction (process).
+The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.
+In our simulations, if the mechanism approximated was the equalising one, the pattern can clearly be interpreted as a change in functional diversity, but if the mechanism is a facilitation one, the pattern can be meaningless (i.e.
+not clearly distinguishable from a null mechanism - Fig.
+\ref{Fig:simulation_results} shows this example by contrasting row 5, columns 1 and 2).
+In this made up example, the procedure is done in reverse order where the the pattern is specified first and then the process and the mechanism (``What process and mechanism are needed to interpret pattern X?'').
+In Box 2 we suggest an approach that specifies the process and the mechanism first (``how do I interpret pattern X under a specific process and mechanism''; see Box 2).
+
+Regarding the different approximated mechanisms, it is interesting to note that our different approximations lead to different general behaviours of the metrics: for example the range of variance of metrics behaviour is greater for the equalising mechanism and smaller for the competition mechanism (albeit nearly always distinguishable from the null mechanism).
+This differences in metrics variance per mechanisms suggest that some mechanisms leads to more subttle changes or/and that our approximation of these mechanisms can be variable (see caveats below).
 
@@
 In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.
+We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
 

### 2024-10-30 Analysis/04_Summarise_results.Rmd
calculate ses

@@
 model_table_emp <- fancy.table(model_fits_emp, columns = list(c(3,4), 5))
-null_table_2d   <- fancy.table(null_differences_2d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_3d   <- fancy.table(null_differences_3d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_4d   <- fancy.table(null_differences_4d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_5d   <- fancy.table(null_differences_5d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_8d   <- fancy.table(null_differences_8d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), c(4, 6)))
+null_table_2d   <- fancy.table(null_differences_2d, columns = list(c(1,3), c(1,3)+4, c(1,3)+8, c(1,3)+12))
+null_table_3d   <- fancy.table(null_differences_3d, columns = list(c(1,3), c(1,3)+4, c(1,3)+8, c(1,3)+12))
+null_table_4d   <- fancy.table(null_differences_4d, columns = list(c(1,3), c(1,3)+4, c(1,3)+8, c(1,3)+12))
+null_table_5d   <- fancy.table(null_differences_5d, columns = list(c(1,3), c(1,3)+4, c(1,3)+8, c(1,3)+12))
+null_table_8d   <- fancy.table(null_differences_8d, columns = list(c(1,3), c(1,3)+4, c(1,3)+8, c(1,3)+12))
+null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), c(1,3)+4))
 ```
@@
 rownames(model_table_emp) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-kable(model_table_emp, caption = "Results of the model scaled metric ~ removal level for the empirical data (3D)", align = "c")
+kable(model_table_emp, caption = "Results of the model scaled metric ~ removal level for the empirical data (5D)", align = "c")
 ```

### 2024-10-30 Functions/fancy.table.R

@@ -27,9 +27,12 @@ get.columns <- function(one_stressor, columns, symbols, digits) {
     extracted_data <- lapply(columns, function(col, dat) dat[, col, drop = FALSE], dat = one_stressor)
     ## Fancied data
     fancy_data <- lapply(extracted_data, function(col, symbols, digits) apply(col, 1, p.symbol, symbols, digits), symbols = symbols, digits = digits)
+
+    fancy_data1 <- lapply(extracted_data, function(col, symbols, digits))
+
     ## Combine that
     fancy_data <- as.data.frame(do.call(cbind, fancy_data))
-    colnames(fancy_data) <- colnames(one_stressor)[unlist(lapply(columns, function(x) x[1]))]
+    colnames(fancy_data) <- colnames(one_stressor)[unlist(lapply(columns, function(x) x[w]))]
     return(fancy_data)
 }
 
@@ -38,7 +41,7 @@ p.symbol <- function(x, symbols, digits) {
     ## Arbitrary p_value limits
     limits <- c(0.1, 0.05, 0.01, 0.001)
 
-    if(length(x) == 1) {
+    if(missing(symbols)) {
         ## No p-value symbol needed
         return(round(x, digits))
     } else {

### 2024-10-30 Functions/paired.tests.R

@@ -19,6 +19,7 @@ paired.tests <- function(one_stressor) {
     stressor_list <- unlist(apply(results_table[, seq(from = 2, to = ncol(results_table), by = 2)], 2, list), recursive = FALSE)
     ## Calculate all the differences
     test_list <- mapply(t.test, random_list, stressor_list, MoreArgs = list(paired = TRUE), SIMPLIFY = FALSE)
+    ses_list <- mapply(function(x,y) return(effectsize::hedges_g(x, y)[[1]]), random_list, stressor_list)
     ## Extract the parameters (d, t and p)
     extract.params <- function(one_test) {
         return(c(d = one_test$estimate[[1]],
@@ -27,6 +28,7 @@ paired.tests <- function(one_stressor) {
     }
     test_table <- t(do.call(cbind, lapply(test_list, extract.params)))
     rownames(test_table) <- gsub("random_", "", rownames(test_table))
+    test_table <- cbind(test_table, ses = ses_list)
     
     ## Splitting by rm levels
     test_list <- split.metric(t(test_table))

### 2024-10-30 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
   Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e.
-observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient.
+observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient \citep{chesson2000mechanisms,barot2004mechanisms}.
 & Metrics capturing changes in richness (but not divergence and regularity) \\
   Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e.
-located close together in the trait space) are more likely to survive the stressor.
+located close together in the trait space) are more likely to survive the stressor \citep{bruno2003inclusion, danet2024species}.
 In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g.
@@
 
-For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern.
+For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure \ref{Fig:empirical_results}, Supplementary table 5).
 For example, all richness metrics (green; Figure \ref{Fig:empirical_results}) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
@@
 functional dispersion based on dendrograms or hypervolumes).
-Finally, some metrics pick up a decrease for both datasets (e.g. functional evenness based on hypervolumes - although the 95\% confidence intervals overlapped with 0).
-Note that this apparent global increase in most metric values may appear counter intuitive given that 46\% and 69\% of species went extinct over the two time periods.
+Finally, some metrics pick up a decrease for both datasets (e.g. functional evenness based on hypervolumes - although the $95$\% confidence intervals overlapped with $0$).
+Note that this apparent global increase in most metric values may appear counter intuitive given that $46$\% and $69$\% of species went extinct over the two time periods.
 This is due to the scaling of the metrics compared to random equivalent extinctions.

### 2024-11-04 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 Dashed thin grey lines represent non-significant linear models (slope's $p$ value $> 0.05$).
-The model results are available in supplementary table 3.
+The model results (including $R^{2}$ and t-statistics) are available in supplementary table 3.
 Similar figures are available in the supplementary materials for 2 and 8 dimensions.}}
@@
 \caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (green = richness, yellow = divergence, blue = regularity).
-Plain circles represent the metrics for the historic dataset (46\% removal) and fainted squares for the extant dataset (69\% removal).
+Plain circles represent the metrics for the species present on Hawaii before 1500 (historic data; 46\% removal) and fainted squares for the species currently present on Hawaii (extant data; 69\% removal).
 \textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
+The model results (including $R^{2}$ and t-statistics) are available in supplementary table 4.
 }}
@@
 
-The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to dimensions (Figure \ref{Fig:empirical_results}).
+The empirical results are also intriguing in that the loss of Hawaiian bird's functional diversity since 1500 is not greater than expected by chance (at least for certain metrics; Figure \ref{Fig:empirical_results}) because one would expect that extinctions since 1500 target specific areas of functional diversity space (i.e.
+birds with specific trait combinations).
 This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g.
 \citealt{sayol2021loss,Matthews2022}).
+% TODO:
+% This illustrates the importance of setting the question + metric
+% For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance for the species present before 1500 and larger than expected by change for the current species.
+% This could lead to the interpretation that the trait space was not changed under any specific mechanism before 1500 (i.e.
+null mechanism) but that since then, specific mechanisms have been reducing specific areas of the trait space (e.g.
+filtering mechanism).
+% However, when measuring divergence using the Rao's quadratic entropy based on distance matrices, we did not find that.
 This could partly be due to the proportion of passerine extinctions in Hawaii.
@@
 This choice allowed us to simulate a null model designed to test whether the mechanisms of interest were causing the observed change in pattern \citep{bausman2018modeling}.
-For example, if removing 80\% of the edges of a distribution (equalising mechanism) was distinguishable or not from removing 80\% of a distribution randomly (null mechanism).
+For example, if removing 80\% of the edges of a distribution was distinguishable or not from removing 80\% of a distribution randomly (e.g.
+facilitation mechanism measured as divergence using the probability density method - not distinguishable from null when removing 80\% or 60\% of the data; Figure \ref{Fig:simulations}).
 Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).

### 2024-11-06 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 \citealt{sayol2021loss,Matthews2022}).
-% TODO:
-% This illustrates the importance of setting the question + metric
-% For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance for the species present before 1500 and larger than expected by change for the current species.
-% This could lead to the interpretation that the trait space was not changed under any specific mechanism before 1500 (i.e.
+This illustrates the importance of positing the mechanism and process of interest as well as the pattern to capture it before analysing the results.
+For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance for the species present before 1500 and larger than expected by change for the current species.
+This could lead to the interpretation that the trait space was not changed under any specific mechanism before 1500 (i.e.
 null mechanism) but that since then, specific mechanisms have been reducing specific areas of the trait space (e.g.
 filtering mechanism).
-% However, when measuring divergence using the Rao's quadratic entropy based on distance matrices, we did not find that.
-This could partly be due to the proportion of passerine extinctions in Hawaii.
-Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines (approximatively 58\%), the majority of global known anthropogenic extinctions (approx.
+However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before 1500 leading to very different interpretation of the results.
+
+Our results also indicate that although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines (approximatively 58\%), the majority of global known anthropogenic extinctions (approx.
 75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).

### 2024-11-06 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 To build the community trait space, we undertook a PCA including all $118$ species, which we then subset to calculate the trait space for the historic and extant datasets.
-We selected the first $5$ axes to represent the trait spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
+We selected the first $5$ axes to represent the trait spaces that explained at least 95\% of the variance in each specific group (historic and extant species - see supplementary figure 7; following recommendations of \citealt{pigot2020macroevolutionary} and dimensionality selection method from \citealt{guillerme2023innovation}).
 We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing $55$ and $81$ species for each stressor and scaling the results proportionally to this null mechanism (as described above).

### 2024-11-07 Analysis/01_Simulating_spaces_and_stressors.Rmd
added precision on mechanism expectations

--- 2024-10-08 Analysis/01_Simulating_spaces_and_stressors.Rmd
@@
 
-```{r, eval = FALSE}
+```{r, eval = TRUE}
 plot.text.box <- function(my_text, ...) {
@@
 ## Stressors
-plot.text.box("B) Resampling\nwith stressor:\nnull mechanism", cex = 2)
+plot.text.box("B) Null mechanism:\nno specific\nmetric change", cex = 2)
 plot.1D(trait_space, reduction = random_reduction, null = !random_reduction, main = "1D null mechanism")
@@
 
-plot.text.box("C) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
+plot.text.box("C) Approximating\nequalizing\nmechanism:\nexpected changes\nin richness", cex = 2)
 plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
@@
 
-plot.text.box("D) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
+plot.text.box("D) Approximating\nfacilitating\nmechanism:\nexpected changes\nin regularity", cex = 2)
 plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
@@
 
-plot.text.box("E) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
+plot.text.box("E) Approximating\nfiltering\nmechanism:\nexpected changes\nin position", cex = 2)
 plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
@@
 
-plot.text.box("F) Resampling\nwith stressor:\ncompetition\nmechanism", cex = 2)
+plot.text.box("F) Approximating\ncompetition\nmechanism:\nexpected changes\nin divergence", cex = 2)
 plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D competition mechanism")

### 2024-11-07 Manuscript/main_ecography_R1.tex
added precision on mechanism expectations

@@
 
-\section*{Box 1: what, how and why = pattern, process and mechanism}
+\section*{Box 1: what, how and why: pattern, process and mechanism}
 \cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
@@
 This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly.
+We chose this algorithm to approximate the absence of any specific stressor (i.e.
+the removal of data only changes the number of observations, not the properties of the trait space).
 
@@
 This approximates our \textbf{equalising mechanism}. 
+We chose this algorithm to approximate equalisation assuming that a stressor could increase the probability of extinction for observations with more rare trait combinations (i.e.
+observations on the edges of the trait space).
 
@@
 This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
+We chose this algorithm to approximate facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
+I.e. pairs of observations that share a close trait combinations are more likely to remain (i.e. observations adjacent to each other in the trait space).
 
@@
 This approximates our \textbf{filtering mechanism}.
+We chose this algorithm to approximate a flitering assuming that a stressor that could increase the probability of extinction for observations further from some trait combination optimum.
+This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).
 
@@
 This approximates our \textbf{competition mechanism}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
+We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
+This could approximate the fact that observation that have similar trait combinations are more likely to go extinct, due to competition (e.g. using the same resources). 
 \end{itemize}
@@
 However, this is not tackled here for both simplicity and to understand how they work in isolation. 
+Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \ref{Tab:mechanisms}) should be taken as a cautious guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.
 
@@
 & Metrics sensitive to the amount of data \\
+
   Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e.
@@
 & Metrics capturing changes in richness (but not divergence and regularity) \\
+
   Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e.
@@
 & Metrics capturing changes in regularity (but not richness and regularity) \\
+
   Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}.
@@
 & Metrics capturing changes in position (and richness) \\\\
+
   Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}).
@@
 F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution.
-G) we presented the results of the simualtions as the scaled metric scores for the approximated mechanisms (C to F) relative to the null mechanism.
+The expected changes indicate the main type of metric affect by the stressor (following the aspects in \citealt{mammola2021concepts}).
+G) we presented the results of the simulations as the scaled metric scores for the approximated mechanisms (C to F) relative to the null mechanism.
 The dot represents the median metric score and the thick and dashed lines respectively the 50\% and 95\% confidence intervals.}}

### 2024-11-07 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 
-The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.
+The ability of different metrics to capture the different patterns (and thus approximate the mechanisms) was highly variable.
 It ranged from metrics capturing no clear pattern (e.g.
@@
 Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
+Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulation show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).
 

### 2024-11-25 Manuscript/main_ecography_R1.tex
edits from Tom and Maria
Up to Comparing and scaling the results to the random removals (null stressor)

@@
 We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).
-For example the competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
+For example competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
 This emphasises the importance of not using a one-size-fits-all metric.
@@
 
-Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and \ref{Tab:mechanisms}).
-Our goal is to the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
-We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).
-To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.
-We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.
+Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and Table \ref{Tab:mechanisms}).
+Our goal is to assess the relative performance of different metrics in capturing the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
+We expect that our ability to capture a pattern does not only depend on the choice of metric (what) but also on the process and mechanism at hand (how and why).
+To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data.
+We show that, with a fixed process and mechanisms approximation, the choice of statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data, with the exact same data being sometimes interpreted in opposite ways.
 
@@
 This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
-These observations can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. %@@@ link to glossary
+Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive).
+These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
+For example, for trait defined as ``leaf insertion angle'', this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).
+These observations on which the traits are measured can represent any biological focal entities, e.g.
+tips in a phylogenetic tree, individuals, species, populations, OTUs, etc.
+(see Glossary table in supplementary materials).
 This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
@@
 We chose this algorithm to approximate the absence of any specific stressor (i.e.
-the removal of data only changes the number of observations, not the properties of the trait space).
+the removal of data only changes the number of observations, not the properties of the trait space in a systematic way).
 
@@
 This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
-We chose this algorithm to approximate facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
-I.e. pairs of observations that share a close trait combinations are more likely to remain (i.e. observations adjacent to each other in the trait space).
+We chose this algorithm to approximate the facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
+I.e. pairs of observations that share similar trait combinations are more likely to remain (i.e. observations adjacent to each other in the trait space).
 
@@
 This approximates our \textbf{filtering mechanism}.
-We chose this algorithm to approximate a flitering assuming that a stressor that could increase the probability of extinction for observations further from some trait combination optimum.
+We chose this algorithm to approximate a flitering mechanism assuming a stressor that could increase the probability of extinction for observations further from some trait combination optimum.
 This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).
@@
 In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
-This is done by selecting of $B$ discrete categories to summarise the distribution - i.e.
+This is done by selecting $B$ discrete categories to summarise the distribution - i.e.
 bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0()} function in \texttt{R};][]{silverman1986density}.
-Each of these discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$.
-Where $p$ is factor increasing the scaling power of the algorithm (here we used $p=3$).
+Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$.
+Where $p$ is a factor increasing the scaling power of the algorithm (here we used $p=3$).
 In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
-This approximates our \textbf{competition mechanism}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
+This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
 We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
-This could approximate the fact that observation that have similar trait combinations are more likely to go extinct, due to competition (e.g. using the same resources). 
+This could approximate the fact that observations that have similar trait combinations are more likely to go extinct due to competition (e.g. using the same resources). 
 \end{itemize}
@@
 For the decreasing size, increasing density and shifting space stressors, we estimated the parameters $\rho$, $D$ and $\rho'$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
-Note that the algorithm used here to simulate the stressors depend, to some extent, to the distribution of the data for each trait space.
-The algorithms used can lead to similar stressors effects depending on the simulated data distribution.
-For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their name suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit no identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors.)
+Note that the algorithm used here to simulate the stressors depends, to some extent, on the distribution of the data for each trait space.
+The algorithms used can lead to similar stressor effects depending on the simulated data distribution.
+For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors.)
 These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 intensities of data removal).
@@
 
-We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}) for 2 and 8 dimensional traits with not correlation between each dimensions in each simulation pipeline (supplementary materials).
-We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait} - note that int the supplementary figures, we measured the \texttt{TPD} metrics on only the four first dimensions) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
-This curse changes the properties of space occupancy in a non-linear way depending on the observation's distribution and the number of dimensions.
+We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}), and for 2 and 8 dimensional (see supplementary materials).
+In each pipeline, the dimensions where simulated as uncorrelated.
+We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait} - note that in the supplementary figures, we measured the \texttt{TPD} metrics on only the four first dimensions) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
+This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.
 For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
@@
 \hline
-  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations.
+  Null mechanism & Random removal (``random'') & No overall systematic change in observations' traits or community structure except for the reduction in the number of observations.
 & Metrics sensitive to the amount of data \\
@@
 observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient \citep{chesson2000mechanisms,barot2004mechanisms}.
-& Metrics capturing changes in richness (but not divergence and regularity) \\
+& Metrics mainly capturing changes in richness (but less in divergence and not in regularity) \\
 
-  Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e.
+  Facilitation & Density change (``density'') & When observations with similar trait combinations (i.e.
 located close together in the trait space) are more likely to survive the stressor \citep{bruno2003inclusion, danet2024species}.
@@
 a community of species) with shared traits rather than the opposite.
-& Metrics capturing changes in regularity (but not richness and regularity) \\
+& Metrics capturing changes in regularity (but not richness and divergence) \\
 
@@
   Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}).
-At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}).
+At the extreme, it implies that only dissimilar observations will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}).
 & Metrics capturing changes in divergence (but not richness and regularity) \\\\
@@
 \caption{\scriptsize{Illustration of the simulation protocol used in this paper.
-A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 observations resulting a trait space of 2, 4, or 8 dimensions.
-The first plot on row A illustrates one such simulation in one dimensions through time (with time on the horizontal axis and trait value for one dimension on the vertical axis).
-The second plot on the same row shows the same results but with 2 dimensions of the trait represented here.
+A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 observations resulting in a trait space of 2, 4, or 8 dimensions.
+The first plot on row A illustrates one such simulation in one dimension through time (with time on the horizontal axis and trait value for one dimension on the vertical axis).
+The second plot on the same row shows the same results but with a two dimensional trait represented here.
 We then applied different stressors to the resulting trait space.
@@
 The first column (histograms) illustrates the distribution of observations in one dimension of the 2D, 4D or 8D traits with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces).
-The second column illustrate the trait in 2 dimensions.
+The second column illustrates the trait in 2 dimensions.
 We did not illustrate the data beyond 2 dimensions for traits in 4 or 8 dimensions.
@@
 D) \textbf{approximating facilitation mechanism}: removing observations to reduce the distance between pairs of observations (increasing local density; decreasing evenness);
-E) \textbf{approximating filtering mechanism}: removing observations on one extreme of the distribution;
-F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution.
+E) \textbf{approximating filtering mechanism}: removing observations from one extreme of the distribution;
+F) \textbf{approximating competition mechanism}: proportionally removing observations from the centre of the distribution.
 The expected changes indicate the main type of metric affect by the stressor (following the aspects in \citealt{mammola2021concepts}).
@@
 
-\subsubsection{Metrics for measuring trait space occupancy (\textit{aka} function diversity, dissimilarity, disparity):}
+\subsubsection{Metrics for measuring trait space occupancy (\textit{aka} functional diversity, dissimilarity, disparity):}
 

### 2024-11-25 Manuscript/main_ecography_R1.tex
edits up to box 2

@@
 
-To understand the ability of each metric to capture a mechanism of interest, we compared their scores relatively to the same metric scores of a null mechanism.
-To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between $0$ and $1$.
+To understand the ability of each metric to capture a mechanism of interest, we compared their scores relative to the same metric scores of a null mechanism.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metric by their absolute greatest value to scale them between $0$ and $1$.
 We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.
@@
 Finally, we measured the effect of the different strengths of the stressors (i.e.
-removing 20\%, 40\%, 60\% and 80\% the data) by fitting a linear regression using the scaled metric difference to null score as a function of the amount of data removed.
-For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables $1$ to $3$ and whether the slope was clearly distinguishable from $0$ (Figure \ref{Fig:simulation_results}).
+removing 20\%, 40\%, 60\% and 80\% the data) by fitting a linear regression using the scaled metric difference to the null score as a function of the amount of data removed.
+For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables $1$ to $3$ and whether the slope was clearly distinguishable from $0$ (Figure \ref{Fig:simulation_results}; \textit{sensu} \cite{dushoff2019can}).
 This allowed us to see if there was any predictable linearity in the metric scores.
-In other words testing whether the metric scores relative to null removals where linearly scaling with the amount of data removed.
+In other words testing whether the metric scores relative to null removals were linearly scaling with the amount of data removed.
 
@@
 We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
-A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at $1500$ CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
+A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at $1500$ CE) and extant species list of the native community for each island were taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
 We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
-We extracted dimensions for extant species from the AVONET database \citep{tobias2022avonet}.
-We used $9$ morphological dimensions associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
+We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}.
+We used $9$ morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
 Traits were log transformed and standardised to a mean of $0$ and standard deviation of $1$, prior to analyses.
@@
 $55$ species are known to have gone extinct prior to $1500$ CE and $26$ after $1500$ CE.
+Note that we used the extinct species data as presented in \cite{matthews2023global}, and thus we do not include de Hawaiian species declared extinct by the IUCN in $2024$.
 For our analyses, we focused on two time periods: (i) the avifauna present in $1500$ CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
@@
 To build the community trait space, we undertook a PCA including all $118$ species, which we then subset to calculate the trait space for the historic and extant datasets.
-We selected the first $5$ axes to represent the trait spaces that explained at least 95\% of the variance in each specific group (historic and extant species - see supplementary figure 7; following recommendations of \citealt{pigot2020macroevolutionary} and dimensionality selection method from \citealt{guillerme2023innovation}).
+We selected the first $5$ axes to represent the trait spaces that explained at least 95\% of the variance in each specific group (historic and extant species - see supplementary figure 7; following the recommendations of \citealt{pigot2020macroevolutionary} and the dimensionality selection method from \citealt{guillerme2023innovation}).
 We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing $55$ and $81$ species for each stressor and scaling the results proportionally to this null mechanism (as described above).
@@
 The colours are here to visually separate the metrics by categories (green = richness, yellow = divergence, blue = regularity); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
-The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test $p$ value $> 0.05$).
+The grey dots and corresponding CI lines represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test $p$ value $> 0.05$).
 Grey lines in the background across the distributions of different removal amounts represent the fitted linear model ``centred and scaled metric score function of the amount of data removed''.
@@
 Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).
-Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted into a score increase (e.g.
-the functional evenness based on hypervolume for the equalising mechanism) or non-linear responses (e.g.
-Alpha diversity based on dendrogram for the equalising mechanism).
+Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted in a score increase (e.g.
+the Functional evenness based on a hypervolume for the equalising mechanism) or non-linear responses (e.g.
+Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores).
 Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
-Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulation show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).
+Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulations show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).
 
@@
 
-More broadly, as can be seen from this empirical case, patterns are likely to be much less clear than with simulated data due to multiple stressors acting simultaneously.
-For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
-In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \citep{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the dimensions we have considered here), but is related to genetics (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.
-
 \section{Discussion}
 
-\subsection{Results}
+\subsection{Capturing approximated mechanisms with different metrics}
 
@@
 randomly removing elements has more effects on the metric than removing them following the mechanism).
-However, some metrics (e.g. the functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
+However, some metrics (e.g. the Functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
 This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.
@@
 
-Regarding the different approximated mechanisms, it is interesting to note that our different approximations lead to different general behaviours of the metrics: for example the range of variance of metrics behaviour is greater for the equalising mechanism and smaller for the competition mechanism (albeit nearly always distinguishable from the null mechanism).
-This differences in metrics variance per mechanisms suggest that some mechanisms leads to more subttle changes or/and that our approximation of these mechanisms can be variable (see caveats below).
+Regarding the different approximated mechanisms, it is interesting to note that our different approximations lead to different general behaviours of the metrics: for example the range of variance in the metric values is greater for the equalising mechanism and smaller for the competition mechanism (albeit nearly always distinguishable from the null mechanism).
+These differences in metric variance across mechanisms suggest that some mechanisms lead to more subtle changes or/and that our approximation of these mechanisms can be variable (see caveats below).
 
@@
 
-The empirical results are also intriguing in that the loss of Hawaiian bird's functional diversity since 1500 is not greater than expected by chance (at least for certain metrics; Figure \ref{Fig:empirical_results}) because one would expect that extinctions since 1500 target specific areas of functional diversity space (i.e.
+The empirical results are also intriguing in that the loss of Hawaiian bird's functional diversity due to extinction is not greater than expected by chance (at least for certain metrics; Figure \ref{Fig:empirical_results}) because one would expect that extinctions target specific areas of functional diversity space (i.e.
 birds with specific trait combinations).
@@
 This illustrates the importance of positing the mechanism and process of interest as well as the pattern to capture it before analysing the results.
-For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance for the species present before 1500 and larger than expected by change for the current species.
-This could lead to the interpretation that the trait space was not changed under any specific mechanism before 1500 (i.e.
+For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance when focusing on all extinctions and larger than expected by chance when considereing only pre-$1500$ extinctions.
+% This could lead to the interpretation that the trait space was not changed under any specific mechanism before $1500$ (i.e.
 null mechanism) but that since then, specific mechanisms have been reducing specific areas of the trait space (e.g.
 filtering mechanism).
-However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before 1500 leading to very different interpretation of the results.
+% However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before $1500$ leading to very different interpretation of the results.
+%TM: can we make this interpretation? as this isn't the two time periods, i.e., there isn't a before and after 1500 comparison. It is all extinctions, and then just pre-1500 extinctions. 
+%TM: And also see my previous comment - is not saying the opposite? That there was some mechanism pre 1500? But not when considering all extinctions?
 
-Our results also indicate that although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines (approximatively 58\%), the majority of global known anthropogenic extinctions (approx.
-75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
-However, in Hawaii, where the same proportion of known species were passerines (approx. 58\%), the opposite is true: the majority of extinctions (approx. 60\%) have been of passerine species.
+Our results indicating that the loss of functional diversity (based on several metrics) was not more than expected by chance when focusing on all extinctions could be due to the fact that although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines (approximatively $58$\%), the majority of global known anthropogenic extinctions (approx.
+$75$\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
+However, in Hawaii, where the same proportion of known species were passerines (approx. $58$\%), the opposite is true: the majority of extinctions (approx. $60$\%) have been of passerine species.
 It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
-But these results, given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines,  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
+Given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines, our results indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
 relative to random extinctions).
@@
 However, it is very likely that researchers will ask a more exciting or intriguing question based on these data.
-For example, an interesting one could be ``Are the extinctions uniform across all birds in Hawaii?".
+For example, an interesting one could be ``Are the extinctions spread uniformly in the trait space across all birds in Hawaii?".
 This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
@@
 
-Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
-For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
+Importantly, it should always be borne in mind that mechanisms ought of course to be more complex in real world scenarios.
+For example, evolutionary mechanisms can vary through time or across clades (and so patterns are often the results of multiple processes); and ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
 facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})

### 2024-11-25 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 
-\textit{We have a data of bird traits for native species in Hawaii through time before and after 1500.
-We want to know how anthropogenic stresses (hunting, invasive species, land use change) have affected the trait diversity in native Hawaiian birds.}
+\textit{We have data of bird traits for native species in Hawaii through time (e.g. before and after human colonisation of the archipelago).
+We want to know how anthropogenic stresses (e.g. hunting, invasive species, land use change) have affected the trait diversity of native Hawaiian birds.}
 
@@
 
+%TG: TODO: check the pre-1500 story from email from Tom and Maria
+
 \begin{table}[ht]
@@
 This is because the metrics have properties that can be counter intuitive based on the data at hand.
-And also because functions of ecosystems or organisms and niches within it are hard to capture/understand/define.
+And also because functions of ecosystems, or the organisms and niches within them, are hard to capture/understand/define.
 This means that measuring functional diversity is based on both the definition of the metric and the function of interest and thus cannot be served by a one-fits-all metric.
-In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.
+In other words, it is important to choose a disparity/dissimilarity/functional diversity metric based on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.
 We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.

### 2024-11-27 Analysis/03_Empirical_pipeline.Rmd
Update 03_Empirical_pipeline.Rmd

--- 2024-06-03 Analysis/03_Empirical_pipeline.Rmd
@@
 Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.
+
+## Empirical example for box 2
+
+```{r}
+library(dispRity)
+## Plotting the trait space
+data <- custom.subsets(empirical_traitspace, group = lapply(groups, which))
+pdf(file = "../Manuscript/Figures/empirical_example.pdf", height = 4, width = 4)
+plot(data, legend = FALSE, col = c("blue", "orange"))
+legend("topright", pch = 19, col = c("blue", "orange"), legend = c("Current species", "Species in 1500"))
+dev.off()
+
+## Measuring disparity
+library(xtable)
+xtable(summary(dispRity(data, metric = convhull.volume)))
+xtable(summary(dispRity(data, metric = displacements)))
+```
+
+## Empirical number of dimensions selection
+
+```{r}
+library(dispRity)
+pca <- readRDS("../Data/Processed/pca-object-historic-community.RDS")
+data <- custom.subsets(pca$x, group = list(extant = extant_species, at_1500 = historic_species))
+pdf(file = "../Manuscript/Figures/empirical_data_dimensions.pdf", height = 8, width = 4)
+plot(select.axes(data))
+dev.off()
+```

### 2024-11-27 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at $1500$ CE) and extant species list of the native community for each island were taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
-We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
+We also added two marine species to the prehistoric species list (\textit{Haliaeetus albicilla} and \textit{Pterodroma hypoleuca}); we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
 We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}.
@@
 %TM: And also see my previous comment - is not saying the opposite? That there was some mechanism pre 1500? But not when considering all extinctions?
+%TODO@@@
 
@@
  current &  37 & 25.44 \\ 
- pre 1500 &  63 & 60.84 \\ 
+ at 1500 &  63 & 60.84 \\ 
    \hline
@@
  current &  37 & 1.04 & 0.95 & 0.99 & 1.05 & 1.10 \\ 
- pre 1500 &  63 & 1.12 & 0.85 & 0.94 & 1.19 & 1.34 \\ 
+ at 1500 &  63 & 1.12 & 0.85 & 0.94 & 1.19 & 1.34 \\ 
    \hline

### 2024-11-28 Manuscript/Tables/results_differences2d_1.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & EQ.d\_1 & EQ.ses\_1 & EQ.d\_2 & EQ.ses\_2 & EQ.d\_3 & EQ.ses\_3 & EQ.d\_4 & EQ.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 66.11*** & 4.1 & 68.86*** & 3.87 & 60.36*** & 3 & 43.83*** & 2.01 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 43.52*** & 4.02 & 41.95*** & 3.83 & 34.67*** & 2.89 & 23.62*** & 1.85 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 36.44*** & 4.88 & 41.74*** & 5.17 & 39.04*** & 4.21 & 30.16*** & 2.99 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 31.3*** & 5.32 & 37.36*** & 4.98 & 33.84*** & 3.53 & 23.2*** & 2.13 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.11*** & 3.48 & 0.05*** & 1.64 & 0.01*** & 0.44 & -0.01*** & -0.25 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 2.06*** & 4.98 & 1.78*** & 4.39 & 1.38*** & 3.23 & 0.92*** & 2.06 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.24*** & 2.01 & 0.17*** & 1.51 & 0.11*** & 1.03 & 0.06*** & 0.59 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 2.81*** & 20.32 & 2.07*** & 18.96 & 1.43*** & 15.61 & 0.82*** & 11.23 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.09*** & 4.35 & 0.04*** & 2.12 & 0.01*** & 0.7 & -0.01*** & -0.49 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.33*** & -5.46 & -0.24*** & -4.94 & -0.16*** & -3.81 & -0.09*** & -2.12 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0 & -0.01 & 0* & -0.18 & 0*** & -0.27 & 0*** & -0.23 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & -0.02*** & -0.69 & -0.03*** & -1.32 & -0.03*** & -1.58 & -0.03*** & -1.54 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D). EQ = equalising; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences2d_2.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & FA.d\_1 & FA.ses\_1 & FA.d\_2 & FA.ses\_2 & FA.d\_3 & FA.ses\_3 & FA.d\_4 & FA.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 34.1*** & 1.51 & 32.08*** & 1.19 & 32.28*** & 1.14 & 28.07*** & 0.98 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 21.27*** & 1.29 & 18.72*** & 1.03 & 17.59*** & 0.96 & 14.25*** & 0.77 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 21.81*** & 1.94 & 21.56*** & 1.56 & 22.13*** & 1.53 & 19.58*** & 1.33 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 22.76*** & 2.2 & 26.13*** & 1.63 & 29.21*** & 2.18 & 24.35*** & 1.85 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0 & -0.03 & 0 & 0.1 & -0.01** & -0.16 & -0.01*** & -0.2 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.76*** & 1.12 & 0.6*** & 0.89 & 0.55*** & 0.82 & 0.46*** & 0.69 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.08*** & 0.77 & 0.06*** & 0.59 & 0.05*** & 0.47 & 0.03*** & 0.33 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.88*** & 3.28 & 0.7*** & 3.57 & 0.6*** & 4.16 & 0.44*** & 4.18 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.02*** & 0.61 & 0.01*** & 0.53 & 0.01*** & 0.31 & 0 & -0.09 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.09*** & -1.01 & -0.06*** & -0.84 & -0.05*** & -0.81 & -0.04*** & -0.73 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0** & 0.35 & 0 & 0.09 & 0*** & -0.24 & 0*** & -0.45 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.25*** & 5.31 & 0.15*** & 3.43 & 0.06*** & 2.12 & -0.01*** & -0.45 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D). FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences2d_3.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & FI.d\_1 & FI.ses\_1 & FI.d\_2 & FI.ses\_2 & FI.d\_3 & FI.ses\_3 & FI.d\_4 & FI.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 44.79*** & 2.45 & 42.48*** & 1.98 & 36.94*** & 1.58 & 25.79*** & 1.08 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 31.28*** & 2.26 & 27.45*** & 1.85 & 22.78*** & 1.45 & 14.93*** & 0.95 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 28.59*** & 3.14 & 28.96*** & 2.67 & 26.06*** & 2.22 & 18.98*** & 1.58 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 21.29*** & 3.08 & 24.42*** & 2.89 & 22.75*** & 2.08 & 15.6*** & 1.39 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.04*** & 0.95 & 0.01 & 0.14 & -0.01** & -0.26 & -0.02*** & -0.41 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 1.27*** & 2.4 & 1.03*** & 1.88 & 0.81*** & 1.48 & 0.53*** & 0.98 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.14*** & 1.34 & 0.09*** & 0.92 & 0.06*** & 0.62 & 0.03*** & 0.33 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 1.2*** & 4.88 & 0.99*** & 5.02 & 0.78*** & 4.56 & 0.49*** & 4.4 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.06*** & 2.38 & 0.04*** & 1.71 & 0.03*** & 1.1 & 0.01*** & 0.46 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.05*** & -0.71 & -0.02*** & -0.4 & -0.01*** & -0.23 & 0 & -0.07 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0* & 0.25 & 0. & 0.17 & 0 & 0.11 & 0 & -0.03 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.02*** & 0.46 & 0.01*** & 0.34 & 0 & -0.03 & 0*** & -0.2 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D). FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences2d_4.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & CO.d\_1 & CO.ses\_1 & CO.d\_2 & CO.ses\_2 & CO.d\_3 & CO.ses\_3 & CO.d\_4 & CO.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 17.17*** & 0.79 & 18.21*** & 0.76 & 16.88*** & 0.68 & 12.58*** & 0.48 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 13.24*** & 0.89 & 14.64*** & 0.96 & 12.41*** & 0.81 & 9.06*** & 0.57 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 11.24*** & 1.1 & 12.26*** & 1.05 & 11.73*** & 0.95 & 8.88*** & 0.67 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 7.71*** & 1.17 & 11*** & 1.24 & 11.32*** & 1.14 & 8.68*** & 0.74 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.04*** & 0.87 & 0.04*** & 1.01 & 0.03*** & 0.72 & 0.02*** & 0.49 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.51*** & 0.9 & 0.51*** & 0.93 & 0.41*** & 0.77 & 0.31*** & 0.56 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.06*** & 0.58 & 0.05*** & 0.52 & 0.04*** & 0.4 & 0.03*** & 0.26 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.71*** & 3.18 & 0.67*** & 4.05 & 0.53*** & 4.31 & 0.36*** & 3.85 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.04*** & 1.3 & 0.04*** & 1.73 & 0.03*** & 1.25 & 0.02*** & 0.83 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.02** & -0.29 & -0.01. & -0.13 & 0 & -0.08 & 0 & -0.01 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0** & 0.32 & 0*** & 0.35 & 0*** & 0.3 & 0*** & 0.19 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.03*** & 0.76 & 0.03*** & 0.99 & 0.01*** & 0.62 & 0.01*** & 0.32 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences4d_1.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & EQ.d\_1 & EQ.ses\_1 & EQ.d\_2 & EQ.ses\_2 & EQ.d\_3 & EQ.ses\_3 & EQ.d\_4 & EQ.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 1024.68*** & 2.29 & 1305.5*** & 1.94 & 1221.81*** & 1.37 & 794.58*** & 0.74 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 2277.38*** & 2.33 & 2433.1*** & 2.04 & 2101.05*** & 1.48 & 1349.76*** & 0.84 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 449.97*** & 3.43 & 720.41*** & 3.62 & 780.96*** & 2.77 & 628.64*** & 1.8 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 42.72*** & 4.14 & 51.54*** & 3.7 & 46.87*** & 2.64 & 30.12*** & 1.5 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.13*** & 3.66 & 0.07*** & 1.91 & 0.03*** & 0.92 & 0.01*** & 0.16 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 2*** & 3.53 & 1.65*** & 2.84 & 1.25*** & 2.06 & 0.78*** & 1.23 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.16*** & 1.76 & 0.11*** & 1.25 & 0.08*** & 0.88 & 0.04*** & 0.52 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 4.19*** & 15.4 & 2.92*** & 15.04 & 1.97*** & 11.6 & 1.1*** & 8.48 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.01*** & 1.2 & 0*** & 0.37 & 0*** & -0.21 & 0*** & -0.55 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.23*** & -5.95 & -0.14*** & -5.16 & -0.08*** & -4.15 & -0.04*** & -2.33 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.56 & 0*** & -0.4 & 0*** & -0.49 & 0** & -0.12 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & -0.01** & -0.33 & -0.01*** & -0.68 & -0.01*** & -0.62 & -0.01*** & -0.54 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D). EQ = equalising; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences4d_2.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & FA.d\_1 & FA.ses\_1 & FA.d\_2 & FA.ses\_2 & FA.d\_3 & FA.ses\_3 & FA.d\_4 & FA.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 517.78*** & 0.82 & 783.51*** & 0.96 & 877.91*** & 0.9 & 724.23*** & 0.67 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 940.19*** & 0.71 & 1134.97*** & 0.72 & 1169.29*** & 0.68 & 945.03*** & 0.53 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 235.02*** & 1.11 & 401.13*** & 1.49 & 477.06*** & 1.45 & 439.74*** & 1.23 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 29.27*** & 0.74 & 51.57*** & 1.85 & 57.78*** & 2.4 & 43.32*** & 2.14 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0 & -0.03 & 0 & 0.05 & 0.01*** & 0.21 & 0*** & 0.1 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.36*** & 0.47 & 0.4*** & 0.53 & 0.39*** & 0.52 & 0.31*** & 0.43 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.03*** & 0.4 & 0.03*** & 0.37 & 0.03*** & 0.33 & 0.02*** & 0.24 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.53*** & 1.38 & 0.62*** & 2.38 & 0.6*** & 3.1 & 0.46*** & 3.63 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.03*** & 2.88 & 0.02*** & 2.69 & 0.02*** & 2.31 & 0.01*** & 0.93 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.03*** & -0.75 & -0.03*** & -1 & -0.03*** & -1.14 & -0.02*** & -0.92 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0. & -0.21 & 0*** & -0.75 & 0*** & -0.52 & 0*** & -0.4 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.22*** & 4.28 & 0.11*** & 3.5 & 0.03*** & 1.38 & -0.01*** & -0.55 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D). FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences4d_3.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & FI.d\_1 & FI.ses\_1 & FI.d\_2 & FI.ses\_2 & FI.d\_3 & FI.ses\_3 & FI.d\_4 & FI.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 667.99*** & 1.31 & 871.01*** & 1.14 & 808.97*** & 0.83 & 549.73*** & 0.48 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 1625.13*** & 1.34 & 1706.46*** & 1.14 & 1401.89*** & 0.83 & 908.74*** & 0.49 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 398.17*** & 2.59 & 587.5*** & 2.44 & 586.06*** & 1.89 & 437.87*** & 1.12 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 33.97*** & 3.24 & 40.35*** & 3.04 & 36.06*** & 1.83 & 23.54*** & 1.03 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.05*** & 1.29 & 0.04*** & 1.08 & 0.02*** & 0.67 & 0.01*** & 0.31 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 1.16*** & 1.84 & 0.91*** & 1.4 & 0.67*** & 1.01 & 0.42*** & 0.62 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.1*** & 1.15 & 0.07*** & 0.83 & 0.04*** & 0.55 & 0.02*** & 0.32 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 1.83*** & 4.43 & 1.42*** & 4.4 & 1*** & 3.98 & 0.59*** & 3.23 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.01*** & 1 & 0.01*** & 1.18 & 0.01*** & 0.85 & 0*** & 0.34 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.04*** & -1.03 & -0.02*** & -0.87 & -0.01*** & -0.49 & -0.01*** & -0.25 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0* & -0.26 & 0 & -0.14 & 0 & 0.1 & 0 & 0.07 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.02*** & 0.88 & 0.01*** & 0.54 & 0* & 0.18 & 0 & 0.09 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D). FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences4d_4.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & CO.d\_1 & CO.ses\_1 & CO.d\_2 & CO.ses\_2 & CO.d\_3 & CO.ses\_3 & CO.d\_4 & CO.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 160.54*** & 0.28 & 239.24*** & 0.28 & 239.86*** & 0.23 & 181.09*** & 0.15 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 436.95*** & 0.3 & 491.69*** & 0.28 & 445.45*** & 0.24 & 307.31*** & 0.15 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 115.67*** & 0.61 & 153.13*** & 0.48 & 159.12*** & 0.42 & 123.17*** & 0.27 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 6.44*** & 0.69 & 8.26*** & 0.61 & 8.72*** & 0.52 & 6.84*** & 0.36 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.04*** & 0.93 & 0.03*** & 0.91 & 0.02*** & 0.7 & 0.01*** & 0.44 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.28*** & 0.41 & 0.24*** & 0.34 & 0.19*** & 0.28 & 0.13*** & 0.18 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.03*** & 0.35 & 0.02*** & 0.26 & 0.02*** & 0.23 & 0.01*** & 0.14 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.68*** & 2.23 & 0.59*** & 2.72 & 0.44*** & 2.8 & 0.28*** & 2.85 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0*** & 0.39 & 0.01*** & 0.61 & 0.01*** & 0.61 & 0*** & 0.39 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.01*** & -0.43 & -0.01*** & -0.38 & -0.01*** & -0.3 & 0*** & -0.22 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0 & -0.08 & 0 & -0.02 & 0 & -0.02 & 0 & -0.02 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.01*** & 0.48 & 0.01*** & 0.5 & 0.01*** & 0.48 & 0*** & 0.21 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences8d_1.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & EQ.d\_1 & EQ.ses\_1 & EQ.d\_2 & EQ.ses\_2 & EQ.d\_3 & EQ.ses\_3 & EQ.d\_4 & EQ.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 761.11*** & 1.34 & 962.64*** & 1.11 & 873.68*** & 0.81 & 544.78*** & 0.43 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 1031482.51*** & 0.39 & 1415980.4*** & 0.33 & 1316657.33*** & 0.23 & 798612.09*** & 0.12 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 7348.88*** & 2.06 & 29098.27*** & 1.93 & 48535.88*** & 1.7 & 47957.63*** & 1.11 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 52.07*** & 3.25 & 62.28*** & 3 & 55*** & 2.41 & 33.6*** & 1.35 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.08*** & 2.07 & 0.05*** & 1.37 & 0.03*** & 0.72 & 0.01*** & 0.29 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 2*** & 2.46 & 1.49*** & 1.83 & 1.07*** & 1.27 & 0.62*** & 0.73 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.12*** & 1.6 & 0.08*** & 1.16 & 0.06*** & 0.85 & 0.03*** & 0.5 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 6.09*** & 12.85 & 4.25*** & 12.1 & 2.83*** & 10.04 & 1.56*** & 7.82 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0*** & 0.47 & 0*** & 0.43 & 0*** & 0.22 & 0 & -0.01 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.07*** & -4.87 & -0.03*** & -3.81 & -0.02*** & -2.76 & -0.01*** & -1.71 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.72 & 0*** & -0.42 & 0*** & -0.29 & 0*** & -0.25 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.02*** & 0.86 & 0.01*** & 0.47 & 0*** & 0.25 & 0 & 0.05 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D). EQ = equalising; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences8d_2.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & FA.d\_1 & FA.ses\_1 & FA.d\_2 & FA.ses\_2 & FA.d\_3 & FA.ses\_3 & FA.d\_4 & FA.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 514.85*** & 1.18 & 555.45*** & 0.78 & 600.74*** & 0.78 & 443.13*** & 0.49 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 877212.48*** & 0.53 & 1040106.12*** & 0.36 & 1253555.6*** & 0.35 & 961196.13*** & 0.22 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 5448.5*** & 0.89 & 18489.32*** & 0.91 & 35772.76*** & 1.48 & 39487.59*** & 1.06 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 63.57*** & 1.5 & 83.62*** & 1.59 & 97.82*** & 4.02 & 75.03*** & 2.69 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & -0.02*** & -0.54 & -0.01*** & -0.4 & -0.01*** & -0.22 & 0*** & -0.11 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.13*** & 0.17 & 0.14*** & 0.17 & 0.14*** & 0.18 & 0.13*** & 0.16 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.02*** & 0.39 & 0.01*** & 0.24 & 0.01*** & 0.22 & 0.01*** & 0.16 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.49*** & 0.97 & 0.44*** & 1.4 & 0.43*** & 1.91 & 0.38*** & 2.35 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.02*** & 3.95 & 0.02*** & 2.45 & 0.01*** & 1.65 & 0.01*** & 0.72 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & 0* & -0.25 & -0.01*** & -0.72 & 0*** & -0.77 & 0*** & -0.69 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -1.09 & 0*** & -1.26 & 0*** & -1.19 & 0*** & -0.67 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.25*** & 7.86 & 0.13*** & 4.81 & 0.05*** & 2.64 & 0.01*** & 0.64 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D). FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences8d_3.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & FI.d\_1 & FI.ses\_1 & FI.d\_2 & FI.ses\_2 & FI.d\_3 & FI.ses\_3 & FI.d\_4 & FI.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 512.45*** & 0.98 & 601.87*** & 0.74 & 538.56*** & 0.54 & 373.2*** & 0.32 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 759454.01*** & 0.51 & 1036742.44*** & 0.41 & 984207.67*** & 0.3 & 724993.83*** & 0.18 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 6282.8*** & 1.77 & 22928.75*** & 1.72 & 33761.59*** & 1.38 & 35235.64*** & 0.95 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 54.91*** & 3.53 & 71.72*** & 3.02 & 67.56*** & 2.39 & 46.86*** & 1.62 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.02*** & 0.45 & 0.01* & 0.15 & 0.01* & 0.15 & 0** & 0.12 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.66*** & 0.77 & 0.45*** & 0.56 & 0.33*** & 0.43 & 0.22*** & 0.29 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.06*** & 0.8 & 0.03*** & 0.5 & 0.02*** & 0.32 & 0.01*** & 0.19 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 1.73*** & 1.99 & 1.19*** & 1.89 & 0.86*** & 2.12 & 0.57*** & 2.4 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.02*** & 1.8 & 0.01*** & 1.45 & 0.01*** & 1.29 & 0*** & 0.52 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.01*** & -0.79 & 0*** & -0.57 & 0*** & -0.49 & 0*** & -0.32 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.96 & 0*** & -0.95 & 0*** & -0.73 & 0*** & -0.5 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.13*** & 1.73 & 0.07*** & 1.59 & 0.03*** & 1.41 & 0.01*** & 0.52 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D). FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differences8d_4.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & CO.d\_1 & CO.ses\_1 & CO.d\_2 & CO.ses\_2 & CO.d\_3 & CO.ses\_3 & CO.d\_4 & CO.ses\_4 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 79.64*** & 0.14 & 130.26*** & 0.16 & 105.45*** & 0.1 & 92.81*** & 0.08 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 67526.58*** & 0.04 & 173367.22*** & 0.06 & 216655.44*** & 0.06 & 150636.07*** & 0.04 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 1312.02*** & 0.3 & 6855.69*** & 0.47 & 8664.07*** & 0.32 & 8093.46*** & 0.2 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 3.79*** & 0.29 & 7.28*** & 0.44 & 6.82*** & 0.32 & 5.38*** & 0.23 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.02*** & 0.54 & 0.02*** & 0.5 & 0.01*** & 0.4 & 0.01*** & 0.3 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.21*** & 0.27 & 0.21*** & 0.28 & 0.16*** & 0.21 & 0.1*** & 0.13 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.01*** & 0.18 & 0.01*** & 0.2 & 0.01*** & 0.14 & 0.01*** & 0.11 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.6*** & 1.4 & 0.61*** & 2.07 & 0.47*** & 2.15 & 0.32*** & 2.17 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0 & 0.11 & 0** & 0.23 & 0*** & 0.37 & 0*** & 0.27 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & 0** & -0.29 & 0*** & -0.54 & 0*** & -0.34 & 0*** & -0.21 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0 & -0.03 & 0 & -0.1 & 0 & 0.05 & 0 & -0.1 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.01* & 0.24 & 0.01*** & 0.31 & 0*** & 0.25 & 0 & 0.07 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%). Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_differencesempirical.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:22:09 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllll}
  \hline
 & empirical.d\_1 & empirical.ses\_1 & empirical.d\_2 & empirical.ses\_2 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & -7.28*** & -1.53 & -4.8*** & -1.08 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & -44.19*** & -1.74 & -9.72*** & -0.37 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & -17.07*** & -2.29 & -3.07*** & -0.55 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & -6.04*** & -2.38 & -4.32*** & -1.78 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.02*** & 1.45 & -0.03*** & -1.63 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.11*** & 1.28 & -0.05*** & -0.44 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.03*** & 1.64 & -0.03*** & -1.19 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & -0.41*** & -1.91 & -0.16*** & -0.52 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0* & 0.24 & 0.01*** & 2.35 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & 0.01*** & 0.97 & 0.02*** & 1.06 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.72 & 0*** & 0.69 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & -0.03*** & -2.23 & -0.02*** & -0.7 \\ 
   \hline
\end{tabular}
\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each
stressor (empirical); d = difference, s = standardised effect size. Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_model2d.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:16:05 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & equalizing.Slope & equalizing.adj.R\verb|^|2 & facilitation.Slope & facilitation.adj.R\verb|^|2 & filtering.Slope & filtering.adj.R\verb|^|2 & competition.Slope & competition.adj.R\verb|^|2 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 0.017*** & 0.02 & -0.015** & 0.012 & 0.026*** & 0.057 & -0.001 & -0.001 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 0.014** & 0.011 & -0.016** & 0.014 & 0.024*** & 0.038 & -0.009* & 0.007 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 0.007 & 0.002 & -0.006 & 0 & 0.03*** & 0.069 & 0.004 & 0 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 0.002 & -0.001 & -0.055*** & 0.07 & 0.001 & -0.001 & -0.02*** & 0.048 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.185*** & 0.578 & 0.022** & 0.009 & 0.102*** & 0.117 & 0.022** & 0.013 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.08*** & 0.396 & 0.009. & 0.003 & 0.054*** & 0.195 & 0.008* & 0.005 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.018** & 0.009 & -0.023*** & 0.04 & 0.014* & 0.004 & -0.022*** & 0.048 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.066*** & 0.436 & -0.028*** & 0.046 & -0.009. & 0.004 & -0.022*** & 0.038 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.227*** & 0.572 & 0.044*** & 0.037 & 0.091*** & 0.155 & 0.002 & -0.001 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.09*** & 0.293 & -0.002 & -0.001 & -0.027*** & 0.043 & -0.011* & 0.008 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0.018*** & 0.049 & 0.046*** & 0.179 & 0.004 & 0 & -0.012*** & 0.02 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.09*** & 0.356 & 0.14*** & 0.28 & 0.022*** & 0.037 & -0.009* & 0.005 \\ 
   \hline
\end{tabular}
\caption{Results of the model scaled metric ~ removal level per stressor (2D)} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_model4d.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:16:05 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & equalizing.Slope & equalizing.adj.R\verb|^|2 & facilitation.Slope & facilitation.adj.R\verb|^|2 & filtering.Slope & filtering.adj.R\verb|^|2 & competition.Slope & competition.adj.R\verb|^|2 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 0.054*** & 0.143 & -0.009 & 0.002 & 0.032*** & 0.076 & 0.002 & 0 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 0.013* & 0.007 & -0.032*** & 0.092 & 0.015** & 0.012 & -0.002 & 0 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 0.049*** & 0.122 & 0.008 & 0.002 & 0.061*** & 0.194 & 0.018*** & 0.038 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & 0.03*** & 0.135 & -0.01 & 0.002 & 0.024*** & 0.09 & 0.001 & -0.001 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.164*** & 0.539 & -0.02** & 0.012 & 0.039*** & 0.023 & -0.002 & -0.001 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.108*** & 0.535 & -0.009** & 0.009 & 0.067*** & 0.258 & 0.012*** & 0.025 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.064*** & 0.145 & -0.004 & 0.002 & 0.045*** & 0.05 & 0.005* & 0.004 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.11*** & 0.52 & -0.03*** & 0.071 & 0.034*** & 0.065 & 0.004 & 0.001 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.153*** & 0.393 & 0.059*** & 0.083 & 0.014* & 0.005 & -0.035*** & 0.048 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.118*** & 0.517 & 0.013*** & 0.018 & -0.025*** & 0.06 & -0.002 & 0 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & -0.008** & 0.009 & -0.015*** & 0.016 & -0.006 & 0.002 & -0.001 & -0.001 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.016*** & 0.037 & 0.196*** & 0.535 & 0.014*** & 0.02 & -0.001 & -0.001 \\ 
   \hline
\end{tabular}
\caption{Results of the model scaled metric ~ removal level per stressor (4D)} 
\end{table}

### 2024-11-28 Manuscript/Tables/results_model8d.tex
Added all results tables

% latex table generated in R 4.4.2 by xtable 1.8-4 package
% Thu Nov 28 15:16:05 2024
\begin{table}[ht]
\scriptsize
\centering
\begin{tabular}{rllllllll}
  \hline
 & equalizing.Slope & equalizing.adj.R\verb|^|2 & facilitation.Slope & facilitation.adj.R\verb|^|2 & filtering.Slope & filtering.adj.R\verb|^|2 & competition.Slope & competition.adj.R\verb|^|2 \\ 
  \hline
Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 0.039*** & 0.071 & 0.013** & 0.009 & 0.026*** & 0.051 & 0 & -0.001 \\ 
  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & -0.023*** & 0.041 & -0.038*** & 0.122 & -0.024*** & 0.102 & -0.008*** & 0.056 \\ 
  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & -0.055*** & 0.159 & -0.051*** & 0.121 & -0.034*** & 0.093 & -0.007*** & 0.015 \\ 
  Functional evenness\\probability density\\\texttt{TPD::REND} & -0.032*** & 0.134 & -0.118*** & 0.345 & -0.059*** & 0.185 & -0.008*** & 0.03 \\ 
  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.088*** & 0.157 & -0.02** & 0.01 & 0.013 & 0.002 & -0.012. & 0.004 \\ 
  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.05*** & 0.166 & -0.022*** & 0.055 & 0.01. & 0.003 & -0.005 & 0.002 \\ 
  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.035*** & 0.035 & -0.004 & 0.001 & 0.028*** & 0.029 & -0.007** & 0.009 \\ 
  Divergence\\probability density\\\texttt{TPD::REND} & 0.037*** & 0.121 & -0.029*** & 0.101 & -0.005 & 0 & -0.017*** & 0.041 \\ 
  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.03*** & 0.019 & 0.08*** & 0.131 & 0.041*** & 0.028 & -0.041*** & 0.05 \\ 
  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.077*** & 0.265 & 0.031*** & 0.09 & -0.009* & 0.006 & 0.002 & -0.001 \\ 
  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & -0.019** & 0.013 & -0.023*** & 0.017 & -0.02*** & 0.018 & 0.002 & -0.001 \\ 
  Functional richness\\probability density\\\texttt{TPD::REND} & 0.014** & 0.012 & 0.175*** & 0.574 & 0.08*** & 0.112 & -0.002 & -0.001 \\ 
   \hline
\end{tabular}
\caption{Results of the model scaled metric ~ removal level per stressor (8D)} 
\end{table}

### 2024-11-28 Manuscript/supplementaries.tex
added empirical table
Added all results tables

@@
 \end{table}
+
+\newpage
 
@@
 
+\newpage
+
+\subsection{Model results}
+
 \begin{landscape}
-\begin{table}
-\scriptsize
-\caption{Results of the model scaled metric $\sim$ removal level per stressor (2D)}
-\centering
-\begin{tabular}[t]{l|c|c|c|c|c|c|c|c}
-\hline
-  & equalizing slope & equalizing adj.$R^{2}$ & facilitation slope & facilitation adj.$R^{2}$ & filtering slope & filtering adj.$R^{2}$ & competition slope & competition adj.$R^{2}$\\
-\hline
-Functional evenness\\distances\\ \texttt{FD::dbFD()\$FEve} & 0.017*** & 0.02 & -0.015** & 0.012 & 0.026*** & 0.057 & -0.001 & -0.001\\
-\hline
-Functional evenness\\dendrogram\\ \texttt{BAT::evenness} & 0.014** & 0.011 & -0.016** & 0.014 & 0.024*** & 0.038 & -0.009* & 0.007\\
-\hline
-Functional evenness\\hypervolume\\ \texttt{BAT::kernel.evenness} & 0.007 & 0.002 & -0.006 & 0 & 0.03*** & 0.069 & 0.004 & 0\\
-\hline
-Functional evenness\\probability density\\ \texttt{TPD::REND} & 0.002 & -0.001 & -0.055*** & 0.07 & 0.001 & -0.001 & -0.02*** & 0.048\\
-\hline
-Rao's quadratic entropy\\distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.185*** & 0.578 & 0.022** & 0.009 & 0.102*** & 0.117 & 0.022** & 0.013\\
-\hline
-Functional dispersion\\dendrogram\\ \texttt{BAT::dispersion} & 0.08*** & 0.396 & 0.009. & 0.003 & 0.054*** & 0.195 & 0.008* & 0.005\\
-\hline
-Functional dispersion\\hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.018** & 0.009 & -0.023*** & 0.04 & 0.014* & 0.004 & -0.022*** & 0.048\\
-\hline
-Divergence\\probability density\\ \texttt{TPD::REND} & 0.066*** & 0.436 & -0.028*** & 0.046 & -0.009. & 0.004 & -0.022*** & 0.038\\
-\hline
-Alpha diversity\\dendrogram\\ \texttt{BAT::alpha} & 0.227*** & 0.572 & 0.044*** & 0.037 & 0.091*** & 0.155 & 0.002 & -0.001\\
-\hline
-Functional richness\\convex hull\\ \texttt{BAT::hull.alpha} & -0.09*** & 0.293 & -0.002 & -0.001 & -0.027*** & 0.043 & -0.011* & 0.008\\
-\hline
-Functional richness\\hypervolume\\ \texttt{BAT::kernel.alpha} & 0.018*** & 0.049 & 0.046*** & 0.179 & 0.004 & 0 & -0.012*** & 0.02\\
-\hline
-Functional richness\\probability density\\ \texttt{TPD::REND} & 0.09*** & 0.356 & 0.14*** & 0.28 & 0.022*** & 0.037 & -0.009* & 0.005\\
-\hline
-\end{tabular}
-\end{table}
+\input{Tables/results_model2d.tex}
 \end{landscape}
@@
 \begin{landscape}
-\begin{table}
-\scriptsize
-\caption{Results of the model scaled metric $\sim$ removal level per stressor (4D)}
-\centering
-\begin{tabular}[t]{l|c|c|c|c|c|c|c|c}
-\hline
-  & equalizing slope & equalizing adj.$R^{2}$ & facilitation slope & facilitation adj.$R^{2}$ & filtering slope & filtering adj.$R^{2}$ & competition slope & competition adj.$R^{2}$\\
-\hline
-Functional evenness\\distances\\ \texttt{FD::dbFD()\$FEve} & 0.054*** & 0.143 & -0.009 & 0.002 & 0.032*** & 0.076 & 0.002 & 0\\
-\hline
-Functional evenness\\dendrogram\\ \texttt{BAT::evenness} & 0.013* & 0.007 & -0.032*** & 0.092 & 0.015** & 0.012 & -0.002 & 0\\
-\hline
-Functional evenness\\hypervolume\\ \texttt{BAT::kernel.evenness} & 0.049*** & 0.122 & 0.008 & 0.002 & 0.061*** & 0.194 & 0.018*** & 0.038\\
-\hline
-Functional evenness\\probability\\ \texttt{TPD::REND} & 0.03*** & 0.135 & -0.01 & 0.002 & 0.024*** & 0.09 & 0.001 & -0.001\\
-\hline
-Rao's quadratic entropy\\distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.164*** & 0.539 & -0.02** & 0.012 & 0.039*** & 0.023 & -0.002 & -0.001\\
-\hline
-Functional dispersion\\dendrogram\\ \texttt{BAT::dispersion} & 0.108*** & 0.535 & -0.009** & 0.009 & 0.067*** & 0.258 & 0.012*** & 0.025\\
-\hline
-Functional dispersion\\hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.064*** & 0.145 & -0.004 & 0.002 & 0.045*** & 0.05 & 0.005* & 0.004\\
-\hline
-Divergence\\probability density\\ \texttt{TPD::REND} & 0.11*** & 0.52 & -0.03*** & 0.071 & 0.034*** & 0.065 & 0.004 & 0.001\\
-\hline
-Alpha diversity\\dendrogram\\ \texttt{BAT::alpha} & 0.153*** & 0.393 & 0.059*** & 0.083 & 0.014* & 0.005 & -0.035*** & 0.048\\
-\hline
-Functional richness\\convex hull\\ \texttt{BAT::hull.alpha} & -0.118*** & 0.517 & 0.013*** & 0.018 & -0.025*** & 0.06 & -0.002 & 0\\
-\hline
-Functional richness\\hypervolume\\ \texttt{BAT::kernel.alpha} & -0.008** & 0.009 & -0.015*** & 0.016 & -0.006 & 0.002 & -0.001 & -0.001\\
-\hline
-Functional richness\\probability density\\ \texttt{TPD::REND} & 0.016*** & 0.037 & 0.196*** & 0.535 & 0.014*** & 0.02 & -0.001 & -0.001\\
-\hline
-\end{tabular}
-\end{table}
+\input{Tables/results_model4d.tex}
 \end{landscape}
@@
 \begin{landscape}
-\begin{table}
-\scriptsize
-\caption{Results of the model scaled metric $\sim$ removal level per stressor (8D)}
-\centering
-\begin{tabular}[t]{l|c|c|c|c|c|c|c|c}
-\hline
-  & equalizing slope & equalizing adj.$R^{2}$ & facilitation slope & facilitation adj.$R^{2}$ & filtering slope & filtering adj.$R^{2}$ & competition slope & competition adj.$R^{2}$\\
-\hline
-Functional evenness\\distances\\ \texttt{FD::dbFD()\$FEve} & 0.039*** & 0.071 & 0.013** & 0.009 & 0.026*** & 0.051 & 0 & -0.001\\
-\hline
-Functional evenness\\dendrogram\\ \texttt{BAT::evenness} & -0.023*** & 0.041 & -0.038*** & 0.122 & -0.024*** & 0.102 & -0.008*** & 0.056\\
-\hline
-Functional evenness\\hypervolume\\ \texttt{BAT::kernel.evenness} & -0.055*** & 0.159 & -0.051*** & 0.121 & -0.034*** & 0.093 & -0.007*** & 0.015\\
-\hline
-Functional evenness\\probability\\ \texttt{TPD::REND} & -0.032*** & 0.134 & -0.118*** & 0.345 & -0.059*** & 0.185 & -0.008*** & 0.03\\
-\hline
-Rao's quadratic entropy\\distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.088*** & 0.157 & -0.02** & 0.01 & 0.013 & 0.002 & -0.012. & 0.004\\
-\hline
-Functional dispersion\\dendrogram\\ \texttt{BAT::dispersion} & 0.05*** & 0.166 & -0.022*** & 0.055 & 0.01. & 0.003 & -0.005 & 0.002\\
-\hline
-Functional dispersion\\hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.035*** & 0.035 & -0.004 & 0.001 & 0.028*** & 0.029 & -0.007** & 0.009\\
-\hline
-Divergence\\probability density\\ \texttt{TPD::REND} & 0.037*** & 0.121 & -0.029*** & 0.101 & -0.005 & 0 & -0.017*** & 0.041\\
-\hline
-Alpha diversity\\dendrogram\\ \texttt{BAT::alpha} & 0.03*** & 0.019 & 0.08*** & 0.131 & 0.041*** & 0.028 & -0.041*** & 0.05\\
-\hline
-Functional richness\\convex hull\\ \texttt{BAT::hull.alpha} & -0.077*** & 0.265 & 0.031*** & 0.09 & -0.009* & 0.006 & 0.002 & -0.001\\
-\hline
-Functional richness\\hypervolume\\ \texttt{BAT::kernel.alpha} & -0.019** & 0.013 & -0.023*** & 0.017 & -0.02*** & 0.018 & 0.002 & -0.001\\
-\hline
-Functional richness\\probability density\\ \texttt{TPD::REND} & 0.014** & 0.012 & 0.175*** & 0.574 & 0.08*** & 0.112 & -0.002 & -0.001\\
-\hline
-\end{tabular}
-\end{table}
+\input{Tables/results_model8d.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_empirical.tex}
+\end{landscape}
+
+\newpage
+
+\subsection{Average differences between null and focal stressors}
+
+\begin{landscape}
+\input{Tables/results_differences2d_1.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences2d_2.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences2d_3.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences2d_4.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences4d_1.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences4d_2.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences4d_3.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences4d_4.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences8d_1.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences8d_2.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences8d_3.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differences8d_4.tex}
+\end{landscape}
+
+\begin{landscape}
+\input{Tables/results_differencesempirical.tex}
 \end{landscape}

### 2024-11-28 Analysis/04_Summarise_results.Rmd
added SES

--- 2024-10-14 Analysis/04_Summarise_results.Rmd
@@
 library(knitr)
+library(xtable)
 ## Loading the functions
@@
 
-## Making the fancy tables
+## Making the fancy tables for the models
 model_table_2d  <- fancy.table(model_fits_2d, columns = list(c(3,4), 5))
@@
 model_table_emp <- fancy.table(model_fits_emp, columns = list(c(3,4), 5))
-null_table_2d   <- fancy.table(null_differences_2d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_3d   <- fancy.table(null_differences_3d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_4d   <- fancy.table(null_differences_4d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_5d   <- fancy.table(null_differences_5d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_8d   <- fancy.table(null_differences_8d, columns = list(c(1,3), c(4, 6), c(7, 9), c(10, 12)))
-null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), c(4, 6)))
+## For the differences
+columns_diff_ses <- list(c(1,3), 4)
+columns_diff_ses <- c(columns_diff_ses,
+                      lapply(columns_diff_ses, `+`, 4),
+                      lapply(columns_diff_ses, `+`, 8),
+                      lapply(columns_diff_ses, `+`, 12))
+null_table_2d   <- fancy.table(null_differences_2d, columns = columns_diff_ses, digits = 2)
+null_table_3d   <- fancy.table(null_differences_3d, columns = columns_diff_ses, digits = 2)
+null_table_4d   <- fancy.table(null_differences_4d, columns = columns_diff_ses, digits = 2)
+null_table_5d   <- fancy.table(null_differences_5d, columns = columns_diff_ses, digits = 2)
+null_table_8d   <- fancy.table(null_differences_8d, columns = columns_diff_ses, digits = 2)
+null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), 4, c(1,3)+4, 4+4), digits = 2)
 ```
 
+#### Printing the latex tables for the models
+
 ```{r, echo = FALSE}
-rownames(model_table_2d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-kable(model_table_2d, caption = "Results of the model scaled metric ~ removal level per stressor (2D)", align = "c")
-```
-```{r, echo = FALSE}
-rownames(model_table_3d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-kable(model_table_3d, caption = "Results of the model scaled metric ~ removal level per stressor (3D)", align = "c")
-```
-```{r, echo = FALSE}
-rownames(model_table_emp) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-kable(model_table_emp, caption = "Results of the model scaled metric ~ removal level for the empirical data (3D)", align = "c")
-```
-```{r, echo = FALSE}
-rownames(null_table_2d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-## Even fancier column names
-colnames(null_table_2d) <- gsub(".d_rm", ".rm", colnames(null_table_2d))
-colnames(null_table_2d) <- gsub("equalizing", "equa", colnames(null_table_2d))
-colnames(null_table_2d) <- gsub("facilitation", "faci", colnames(null_table_2d))
-colnames(null_table_2d) <- gsub("filtering", "filt", colnames(null_table_2d))
-colnames(null_table_2d) <- gsub("competition", "comp", colnames(null_table_2d))
-kable(null_table_2d, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2d)", align = "c")
+# rownames(model_table_2d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
+change.names <- function(table, fancy_names, output = "xtable", colnames = FALSE) {
+  if(output == "xtable") {
+    ## Format the row names for xtable
+    ## split should be \\ and the third element should be \texttt{}
+    row_names <- sapply(fancy_names, strsplit, split = "\\n")
+    add.ttt <- function(x) {
+      x[[3]] <- paste0("\\texttt{", x[[3]], "}")
+      return(x)
+    }
+    row_names <- lapply(row_names, add.ttt)
+    row_names <- lapply(row_names, function(x) paste(x, collapse = "\\\\"))
+  }
+  if(colnames) {
+    colnames(table) <- gsub("equalizing", "EQ", colnames(table))
+    colnames(table) <- gsub("facilitation", "FA", colnames(table))
+    colnames(table) <- gsub("filtering", "FI", colnames(table))
+    colnames(table) <- gsub("competition", "CO", colnames(table))
+    colnames(table) <- gsub("rm", "", colnames(table))
+  }
+  rownames(table) <- row_names
+  return(table)
+}
+# kable(model_table_2d, caption = "Results of the model scaled metric ~ removal level per stressor (2D)", align = "c")
+print(xtable(change.names(model_table_2d, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (2D)"), file = "../Manuscript/Tables/results_model2d.tex")
+
+print(xtable(change.names(model_table_4d, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (4D)"), file = "../Manuscript/Tables/results_model4d.tex")
+
+print(xtable(change.names(model_table_8d, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (8D)"), file = "../Manuscript/Tables/results_model8d.tex")
+
+print(xtable(change.names(model_table_emp, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (empirical data)"), file = "../Manuscript/Tables/results_empirical.tex")
 ```
 
+#### Printing the latex tables for the differences
+
 ```{r, echo = FALSE}
-rownames(null_table_3d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-## Even fancier column names
-colnames(null_table_3d) <- gsub(".d_rm", ".rm", colnames(null_table_3d))
-colnames(null_table_3d) <- gsub("equalizing", "equa", colnames(null_table_3d))
-colnames(null_table_3d) <- gsub("facilitation", "faci", colnames(null_table_3d))
-colnames(null_table_3d) <- gsub("filtering", "filt", colnames(null_table_3d))
-colnames(null_table_3d) <- gsub("competition", "comp", colnames(null_table_3d))
-kable(null_table_3d, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (3d)", align = "c")
+print(xtable(change.names(null_table_2d[, 1:8], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
+EQ = equalising; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences2d_1.tex")
+
+print(xtable(change.names(null_table_2d[, 9:16], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
+FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences2d_2.tex")
+
+print(xtable(change.names(null_table_2d[, 17:24], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
+FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences2d_3.tex")
+
+print(xtable(change.names(null_table_2d[, 25:32], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences2d_4.tex")
+
+print(xtable(change.names(null_table_4d[, 1:8], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
+EQ = equalising; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences4d_1.tex")
+
+print(xtable(change.names(null_table_4d[, 9:16], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
+FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences4d_2.tex")
+
+print(xtable(change.names(null_table_4d[, 17:24], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
+FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences4d_3.tex")
+
+print(xtable(change.names(null_table_4d[, 25:32], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences4d_4.tex")
+
+print(xtable(change.names(null_table_8d[, 1:8], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
+EQ = equalising; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences8d_1.tex")
+
+print(xtable(change.names(null_table_8d[, 9:16], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
+FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences8d_2.tex")
+
+print(xtable(change.names(null_table_8d[, 17:24], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
+FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences8d_3.tex")
+
+print(xtable(change.names(null_table_8d[, 25:32], fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differences8d_4.tex")
+
+print(xtable(change.names(null_table_emp, fancy_names, colnames = TRUE),
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each
+stressor (empirical); d = difference, s = standardised effect size. Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+  file = "../Manuscript/Tables/results_differencesempirical.tex")
 ```
 
-```{r, echo = FALSE}
-rownames(null_table_emp) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
-## Even fancier column names
-colnames(null_table_emp) <- gsub(".d_rm", ".rm", colnames(null_table_emp))
-colnames(null_table_emp) <- gsub("equalizing", "equa", colnames(null_table_emp))
-colnames(null_table_emp) <- gsub("facilitation", "faci", colnames(null_table_emp))
-colnames(null_table_emp) <- gsub("filtering", "filt", colnames(null_table_emp))
-colnames(null_table_emp) <- gsub("competition", "comp", colnames(null_table_emp))
-kable(null_table_emp, caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal for the empirical data (pre-history and history)", align = "c")
-```
+<!-- Table cleaning (in shell)
+
+## Cleaning (in shell)
+## search replace
+## $\backslash$ by \
+## \{ by {
+## \} by }
+## \centering by \scriptsize\n\centering
+
+ -->
 

### 2024-11-28 Functions/fancy.table.R

@@ -16,7 +16,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-fancy.table <- function(data, columns, symbols = c("", ".", "*", "**", "***"), digits = 3) {
+fancy.table <- function(data, columns, symbols = c("", ".", "*", "**", "***"), digits = 2) {
     ## Making the fancy table
     return(do.call(cbind, lapply(data, get.columns, columns, symbols, digits)))
 }
@@ -28,11 +28,9 @@ get.columns <- function(one_stressor, columns, symbols, digits) {
     ## Fancied data
     fancy_data <- lapply(extracted_data, function(col, symbols, digits) apply(col, 1, p.symbol, symbols, digits), symbols = symbols, digits = digits)
 
-    fancy_data1 <- lapply(extracted_data, function(col, symbols, digits))
-
     ## Combine that
     fancy_data <- as.data.frame(do.call(cbind, fancy_data))
-    colnames(fancy_data) <- colnames(one_stressor)[unlist(lapply(columns, function(x) x[w]))]
+    colnames(fancy_data) <- colnames(one_stressor)[unlist(lapply(columns, function(x) x[1]))]
     return(fancy_data)
 }
 
@@ -41,7 +39,7 @@ p.symbol <- function(x, symbols, digits) {
     ## Arbitrary p_value limits
     limits <- c(0.1, 0.05, 0.01, 0.001)
 
-    if(missing(symbols)) {
+    if(length(x) == 1) {
         ## No p-value symbol needed
         return(round(x, digits))
     } else {

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
Create ecography_reponse_letter.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Response to reviewers},
  pdfauthor={Thomas Guillerme},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Response to reviewers}
\author{Thomas Guillerme}

\begin{document}
\maketitle

Dear Editor,

We are grateful for the detailed and useful comments from all three reviewers
We have taken all their suggestions on board, integrated changes associated with the suggestion into the manuscript and responded to them in detail below
For clarity we have displayed the \textcolor{blue}{reviewers' comments in blue} and our own responses in black
We fully describe each change below.

Best regards,

Thomas Guillerme, on behalf of all co-authors.

\textcolor{blue}{\textbf{Reviewer: 1}}

\textcolor{blue}{Thank you for the opportunity to review this manuscript, which highlights the critical importance of selecting the appropriate metric for functional diversity when analyzing biological data
The authors used simulated and empirical data to explore the extent to which different functional diversity metrics effectively capture the ecological mechanisms that underlie the relationship between functional diversity patterns and ecological processes
By examining four distinct ecological mechanisms and assessing the ability of various diversity metrics to detect their effect on functional diversity compared to a null model, the authors demonstrated that the same dataset could yield different interpretations depending on the chosen metric
This variability underscores the risk of misinterpreting results and creating confusion in the explanation of key ecological mechanisms.}

\textcolor{blue}{The study is valuable and it has the potential to raise awareness of a significant methodological caveat in functional ecology that, if overlooked, could lead to misconceptions or a loss of knowledge
With the increasing availability of large public datasets on functional traits and the widespread use of functional analyses, it is critical to improve clarity around commonly used methods
However, I believe the manuscript requires revision before it can be recommended for publication
I have several points of critique and corresponding suggestions for improving the next version of this paper
In particular, more clarity is needed in several sections of the manuscript, and the results and discussion should be more effectively linked to the introduction and the study's aims
Additionally, parts of the results and discussion should be reframed to better convey the key message of the study and enhance the overall significance of the paper
I believe this approach would help make the paper less descriptive and provide readers with a clearer framework for selecting the appropriate functional metrics when conducting a functional analysis.}

\textcolor{blue}{Below, I have itemized my major comments and provided more detailed comments organized by the different sections of the manuscript.}

\textcolor{blue}{\textbf{Item 1.}}
\textcolor{blue}{You present four different ecological mechanisms that could drive the reduction observed in your data, and you assess the effectiveness of various diversity metrics in detecting this effect
However, the methodology for testing these relationships is not clearly presented
Specifically, it is unclear how these mechanisms were tested in relation to the null stressor model
The purpose of using a null model is to establish a reference point against which alternative hypotheses can be compared (in your case the four ecological mechanisms), allowing for the identification of whether the focal mechanisms significantly explain the observed patterns in the data (see Gotelli and Graves 1996 ``Null Models in Ecology'').}

We have now specified more clearly throughout the manuscript the role of our null mechanism in the experimental design
Namely:

\textit{This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly.}
Lines @@@

\textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.~\(metric\_stressor\_20 \- metric\_null\_20\)).}
\textit{We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).}
Lines @@@

% The effect of the focal mechanisms is well explained, but I believe a more explicit explanation of how these mechanisms were tested is necessary
% I recommend adding clear statements at the end of the introduction that outline how you tested the different mechanisms against the null model, along with the associated hypotheses and predictions (perhaps in table 1)
% While you briefly mention this in the methods section (e.g., lines 198-201), it would be beneficial to include this information earlier to provide readers with a clearer framework for understanding your approach.

% We have now clarified our goal and expectations at the end of the introduction:
%  \textit{Our goal is to assess the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
%  \textit{We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
%  Lines @@@.

% We have clarified how and why we compare the metrics to the null mechanisms:
%  \textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relatively to the same metric scores of a null mechanism.}
%  \textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between \(0\) and \(1\).}
%  \textit{We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.~\(metric\_{stressor20
% %} \- metric\_{null20
% %}\)).}
%  \textit{This resulted in each metric being scaled between \(–1\) and \(1\), where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.~approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near \(0\) meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \textbackslash ref\{Fig:simulations\}, \textbackslash ref\{Fig:simulation\_results\} and \textbackslash ref\{Fig:empirical\_results\}).}
%  \textit{We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.}
%  Lines @@@

% Furthermore, we have added an expected results column to Table 1 as suggested.

% Additionally, the use of both simulated and real data in your study should be better explained, with more details provided on the value and rationale behind this choice
% This would help to further justify your methodology and enhance the overall clarity and impact of your manuscript.

% We have rewritten the last paragraph of the introduction to include the following:
%  \textit{Beyond simulations, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.}
%  \textit{We show that, with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.}
%  \textit{We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.}
%  Lines @@@

% \textbf{Item 2.}
%  While I appreciate the detailed explanation of the methodology used to produce the simulated space patterns and the application of the stressors, this section appears disproportionate compared to the rest of the methods, especially concerning the testing of focal metrics
% I recommend enhancing the clarity of Sections 3.1.3 and 3.1.4
% Together with providing more explicit statements regarding your hypotheses and predictions (see item 1), it would help readers navigate the different aspects of the analyses
% For instance, clarifying how the performance of each metric in capturing various focal mechanisms was evaluated, and how different results support or challenge this evaluation, would be beneficial.

% We have now rewritten sections 3.1.3 and 3.14 to be clearer:
%  \textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values, normalizing them to a range between \(0\) and \(1\).}
%  \textit{We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.~\(metric\_{stressor20
% %} \- metric\_{null20
% %}\)).}
%  \textit{This resulted in each metric being scaled between \(–1\) and \(1\) where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.~approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near \(0\) meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \textbackslash ref\{Fig:simulations\}, \textbackslash ref\{Fig:simulation\_results\} and \textbackslash ref\{Fig:empirical\_results\}).}
%  \textit{We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.}
%  \textit{To assess these poor performances, we compared the distributions of the \(171\) scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.}
%  \textit{The results of these tests are displayed in figure \textbackslash ref\{Fig:simulation\_results\} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).}
%  Lines @@@

% \textit{Finally, we measured the effect of the different strengths of the stressors (i.e.~removing 20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% and 80\textbackslash\% the data) by fitting a linear regression using the scaled metric difference to null score as a function of the amount of data removed.}
%  \textit{For each of these models, we reported the variance explained by the model (adjusted \(R^2\)) in the supplementary tables \(1\) to \(3\) and whether the slope was clearly distinguishable from \(0\) (Figure \textbackslash ref\{Fig:simulation\_results\}).}
%  \textit{This allowed us to see if there was any predictable linearity in the metric scores.}
%  \textit{In other words testing whether the metric scores relative to null removals where linearly scaling with the amount of data removed.}
%  Lines @@@

% We have also added a panel to figure 1 helping the reader to easily interpret our results in figures 2 and 3.

% \textbf{Item 3.}
%  In my opinion the primary issue in the manuscript lies in how the results are communicated
% As mentioned in the abstract and introduction, your objective is to evaluate the effectiveness of various functional diversity metrics in capturing patterns and signals associated with ten distinct processes linked to common community assembly mechanisms
% From what I understand, your aim is to offer a framework that assists ecologists in selecting the most appropriate diversity metric based on the relevant mechanism and observed patterns
% However, I believe this crucial message is not clearly conveyed
% Instead, the main takeaway from your results and their interpretation seems to be the high variability observed in the performance of the studied metrics.

% I recommend revising the results and discussion sections to address this issue more effectively
% Specifically, it would be helpful to provide a more detailed account of your main findings in the results section, particularly how different metrics performed in relation to the significance of your results, especially regarding the impact of data removal
% Additionally, a more thorough interpretation of these results is needed in the discussion (what really means that a specific metric deviate from the null model, and what the effect of data removal from 20\% to 80\% really implies)
% In the current manuscript, the discussion begins with an early focus on the limitations and caveats of your methods
% While acknowledging these aspects is important, it would be more effective to simplify this discussion and move it to a later section
% Instead, I would emphasize the interpretation of the performance of the various metrics to better communicate your main findings.

% We have now entirely reorganised the discussion section with clearer subsections:

% \begin{itemize} \tightlist \item   Results (including a subsection on empirical results) discussing in   more details the result
% Lines @@@
%  \item   Caveats: discussing clearly the caveats of our approach
% Lines @@@
%  \item   Box 2: ``so which metric do \textbackslash textit\{I\} choose?''   proposing a didactic approach and reflections to hopefully easily help   workers to choose their own metric
% Lines @@@ \end{itemize}

% \textbf{Item 4.}
%  Finally, the discussion section needs additional work to better address the key aspects of ``what,'' ``how,'' and ``why'' in functional analyses
% The introduction emphasizes applying these fundamental questions of ecological analysis to the study of functional diversity metrics
% However, this focus seems to be lost in the discussion, which lacks a strong connection to the introduction
% Given that this is the core of your study (as reflected in the title), it should be the primary focus of the discussion section.
%  I recommend revising the initial part of the introduction to better align it with your main question and hypotheses and to clearly convey the core message of your study
% For example, a more detailed interpretation of how your results on different metrics relate to this aspect of your framework would be helpful
% As I mentioned in my previous item, acknowledging the limitations of your methods is important, but it should be addressed in a separate paragraph following the interpretation of your results, rather than being the primary focus of the discussion.

% We have now linked the framework of ``what'', ``how'' and ``why'' more clearly in the discussion:
%  \textit{For example, say one is interested in looking at changes in functional diversity using a divergence metric calculated with the probability density method (e.g.~\textbackslash texttt\{TPD::REND\}).}
%  \textit{They can apply this metric to their trait space and get two relative metric scores of, says, 0.7 and 0.65 (pattern) for two levels of extinction (process).}
%  \textit{The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.}
%  \textit{In our simulations, if the mechanism approximated was the equalising one, the pattern can clearly be interpreted as a change in functional diversity, but if the mechanism is a facilitation one, the pattern can be meaningless (i.e.~not clearly distinguishable from a null mechanism - Fig
% \textbackslash ref\{Fig:simulation\_results\} shows this example by contrasting row 5, columns 1 and 2).}
%  \textit{In this made up example, the procedure is done in reverse order where the the pattern is specified first and then the process and the mechanism (``What process and mechanism are needed to interpret pattern X?'\,').}
%  \textit{In Box 2 we suggest an approach that specifies the process and the mechanism first (``how do I interpret pattern X under a specific process and mechanism'\,'; see Box 2).}
%  Lines @@@

% As well as in the new Box 2
% Lines @@@

% \textbf{Item 5.}
%  Ensure consistency in the use of terms such as ``elements of the model'' versus ``tips'' or ``species'', and adhere to a uniform citation format (e.g., using or omitting a comma before the year).

% We have now standardised terminology and provide a glossary in the supplementary materials (based on Mammola 2021).

% Overall, addressing these areas will significantly enhance the paper's impact, readability, and contribution to the field of functional ecology.

% Here are more detailed comments on the different sections of the paper.

% \textbf{Abstract}.
%  Line 3-6: Please rephrase to simplify and avoid brackets.
%  We changed the sentence to:
%  \textit{Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to various stressors
% In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism.}
%  Lines @@@

% Line 10-11: Please rephrase to avoid brackets.
%  We changed the sentence to:
%  \textit{We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.}
%  Lines @@@

% Line 14-15: If space requirements permit, please include more detailed information about your results
% The current description is too vague and does not effectively convey your main findings.
%  We've expanded a little on the results
% One point of keeping the vagueness in the abstract though was to not give the impression that we provide any prescriptive results but rather a broad understanding that the choice of metric is important.
%  \textit{We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).}
%  \textit{For example, the competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.}

% \textbf{Introduction.}
%  Line 25-28
% It appears to me that you are using the definition of a functional trait from Violle et al.~(2007) in Oikos
% Please include a citation for this reference if it is the case.
%  We've added the reference as suggested.

% Line 37
% I would avoid using the term ``new trait-focused ecology.'' While it's true that a more ``explicit'' approach to functional ecology has emerged over the past few decades, the foundation of functional analyses is rooted in the field of ecology itself
% For example, as you mentioned, Darwin's finches represent an early functional analysis centered on beak traits
% Although the specific term ``functional traits'' was less commonly used in the past, broader functional analyses were not uncommon, particularly in plant ecology (see, for example, Calow, P
% 1987
% ``Towards a Definition of Functional Ecology.'' Functional Ecology; and Duarte et al., 1995
% ``Comparative Functional Plant Ecology: Rationale and Potentials.'' Trends in Ecology \& Evolution).
%  This is a really good point
% We've removed the mention of ``new''.

% Line 73-74
% Please clarify how this is related to the previous question ``what'' and ``how''.
%  We've expanded the sentence to:
%  \textit{Mechanism description is then about linking the pattern and the process to some biological properties (sometimes referred to as ``rules'\,') and is at the core of answering the biological question at hand.}
%  Lines @@@

% Line 93-96
% Please explain why the chosen metric is not appropriate for the proposed example.
%  We've expanded the sentence to:
%  \textit{In such a scenario, measuring the occupied area (i.e.~\(km^{2}\)) of each population on each island (a pattern - ``what?'\,') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.}
%  Lines @@@

% Line 100
% I think it would be beneficial to more clearly address your research question here
% Is your goal simply to describe the patterns of functional diversity across different processes, or are you aiming to test how well different metrics capture the effects of various mechanisms? If it's the latter, how did you test this, and what are your expectations in relation to the null model?
%  We have now clarified what our goal and expectations are in the last paragraph of the introduction:
%  \textit{Our goal is to assess the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
%  \textit{We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
%  Lines @@@

% Line 102
% Please refer to Table 1 together with figure 1, to provide more details on the ecological mechanisms.
%  We now refer to both the table and the figure.

% Line 103-106
% Please provide more information on why you chose to use a combination of empirical and simulated data
% From my understanding, your intention was to apply your framework within a more realistic context
% If that is the case, please clarify this point to strengthen your rationale.
%  We changed the sentence to:
%  \textit{Beyond simulations, we also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands (process) resulting in a trait space modified by extinctions (the pattern) capturing the effects of species extinction on species trait distributions (the mechanism).}
%  Lines @@@

% Line 106-111
% This content could be more suitable at the start of the discussion section
% Here you could maybe focus on outlining your hypotheses and predictions rather than presenting your conclusions
% By doing so, you establish a clearer context for interpreting your findings and guide the reader through the rationale behind your study.
%  We have now moved this sentence to the end of the conclusion (Lines @@@) to end the introduction on a clear description of our goals and expectations:

% \textit{Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \textbackslash ref\{Fig:simulations\} and \textbackslash ref\{Tab:mechanisms\}).}
%  \textit{Our goal is to the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
%  \textit{We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
%  \textit{To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.}
%  \textit{We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.}

% Lines @@@

% \textbf{Methods}.
%  Line 115
% Please provide more details of the elements of your model, as not all readers may be familiar with the Brownian Motion trait model.
%  We clarified the sentence as follows:
%  \textit{First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.~a pure birth speciation model) until reaching 200 elements in \textbackslash texttt\{R\} \textbackslash citep\{rcore\} using \textbackslash texttt\{treats\} \textbackslash citep\{guillerme2024treats\}.}
%  Lines @@@

% Line 131
% Please explain what the distance D is.
%  We clarified this point as follows:
%  \textit{Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least \(D\) (i.e.~removing the \(n\) pairs of points that are at least \(D\) distance away from each other to remove the desired number of elements).}
%  Lines @@@

% Line 139-144
% I suggest breaking this section into two sentences and improve readability
% The current structure is challenging to follow.
%  We split this sentence in three for clarity:
%  \textit{Each bandwidth has an observed probability of sampling of \(n\) (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of \(n 
% times m^{p}\)
% For bandwidths with a low density (low values of \(n\)) we used \(m\>0\) to increase their resampling probability, and the opposite for bandwidths with a high density (high values of \(n\); using \(m\<0\) to decrease sampling property)
% This effectively flattened the normal distribution of the data by some factor \(p\) (here we used \(p=3\)).}
%  Lines @@@

% Table 1
% Please provide references for the Equalizing and Facilitation mechanisms to offer clearer context for the focal ecological mechanisms studied.
%  We now added references to \textbackslash citep\{chesson2000mechanisms,barot2004mechanisms\} for the equalising mechanism and to \textbackslash citep\{bruno2003inclusion, danet2024species\} for the facilitation one.
%  Lines @@@

% Line 183
% Please provide the total number of the different metrics you used and refer to the table below (table 2).
%  We've now added a reference to table 2 and summarised the total amount of metrics.
%  Lines @@@

% Figure 1
% I believe adding a color legend to the figure would enhance its clarity.
%  We have now reworked figure 1 to be more didactic (especially for the results interpretation)
% We have not added the color legend to the plot itself (although we have added orange arrows to make it clearer) but in the caption to not overload an already busy figure.

% Figure 1
% Please include information in the figure caption explaining what the two types of graphs represent.
%  We added the following clarifications to the figure caption:
%  \textit{We then applied different stressors to the resulting trait space
% We represent the elements removed by the stressors in blue and the ones kept in the trait space in orange
% The first column illustrate the distribution of the elements in one dimension with the overlap between the removed (blue) and kept (orange) elements in brown (note we did not measure any metric on 1D spaces)
% The second column illustrate the same distributions in 2 dimensions
% We did not illustrate the distributions in 4 or 8 dimensions here
% We used the following stressors:}
%  Lines @@@

% Line 194-195
% Please improve clarity.
%  @@@TODO: rewrite following other reviewers comments (stats for results)

% Line 201-203
% Please simplify and improve clarity.
%  @@@TODO: rewrite following other reviewers comments (stats for results)

% Line 207-211
% Please provide a clearer explanation of the effect size and statistical significance of these model outcomes, similar to the explanation provided in lines 198-201.
%  @@@TODO: rewrite following other reviewers comments (stats for results)
%  @@@TODO: calculate effect sizes
%  We now added all the statistics results in the supplementary results from table 6 to 18.

% Line 221
% Please specify the added species.
%  @@@TODO: name of the two added marine species

% Table 2
% The aspect for the third row is missing.
%  We are not sure about what the reviewer meant here
% The tables are all formatted in LaTeX and we prefer to leave it to the production team to set it up as they like should the manuscript be accepted.

% \textbf{Results}.
%  Line 248-251
% Please provide more detail regarding these results (e.g., what the scaled metric scores are, the effect size, and the significance of data reduction caused by the stressor)
% Additionally, refer to the relevant section of Figure 2 (to explain how the observed trends in the figure support this assertion), and to the tables in supplementary material.
%  @@@TODO: add stats for results (following other reviewers comments)

% Line 252
% I would avoid the term ``naively''.
%  We removed this term.

% Line 254
% Please ensure that the use of italics follows the guidelines from the journal.
%  We did not find any specific guideline regarding emphasising words with italics but removed it nonetheless.

% Line 254-256
% Apply the same changes mentioned in the comment below to lines 248-251.
%  @@@TODO: add stats for results (following other reviewers comments)

% Line 260
% Please avoid using the exclamation mark.
%  We changed the punctuation to a dot.

% Line 262-269
% Please provide more detail regarding these results
% Additionally, refer to the table in supplementary material.
%  @@@TODO: add stats for results (following other reviewers comments)
%  We also now refer to the supplementary results:
%  \textit{For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure \textbackslash ref\{Fig:empirical\_results\}, Supplementary table 5).}
%  Lines @@@

% Line 275-283
% This should be in the discussion section.
%  We moved this paragraph to the discussion on lines @@@.

% Figure 2
% Maybe I misunderstood, but I think that you are missing the adjusted R square from the figure.
%  We removed mention of the R2 (from a previous cluttered version of the figure) and now refer to it in the supplementaries in the figure caption.

% Figure 3
% Please provide more details in the figure caption.
%  Please include the results from the t-test in the supplementary material
% While the significance of the test is demonstrated through the gray plots in Figures 2 and 3, the relative values are also necessary.
%  The tables in the supplementary materials now contain the R2 values for both the Figures 2 and 3
% We have now clarified that in the figure captions:
%  \textit{The model results (including \(R^{2}\) and t-statistics) are available in supplementary table 4.}
%  Lines @@@ and Lines @@@

% \textbf{Discussion}.
%  Line 285-291
% This section effectively summarizes your main goal and key message
% However, a more detailed interpretation of your main findings is needed
% I believe that simply stating that ``different metrics capture different patterns (what), leading to inferences about different processes (how) and mechanisms (why)'' is insufficient
% I recommend focusing more on the performance of the various metrics to offer readers a clearer framework for selecting the appropriate metric for each mechanism
% Please refer to items 3 and 4 for further details.
%  Line 291-302
% I appreciate the detailed discussion of the potential limitations of your methodology
% However, in my opinion, this should follow a more in-depth interpretation of your results and their relation to the ``what'', ``how'', and ``why'' questions
% Consider placing the discussion of caveats and limitations in a separate paragraph, where you address the constraints of your framework.
%  Line 303-315
% Based on your preference, I recommend moving this section further into the discussion
% This will allow you to address the importance of selecting the appropriate metric in relation to the number of dimensions used for the trait space.
%  We have now re-orginased the discussion and clearly separated each subsection
% The first one now discusses the results (simulated and empirical), the second one focuses on the caveats of our approach
% And the third section is now a standalone box (box 2) discussing how we believe the what, how and why questions can help workers choose their functional diversity metric of interest based on their own data and question
% Hopefully these clear distinctions will make it easier to follow (or skip) for the reader.

% Line 313-316
% Please rephrase to improve readability.
%  We have now removed this sentence.

% Line 316-337
% I think this section should be simplified
% The discussion focuses too heavily on the limitations of the approach, making it overly methodological
% A greater emphasis on the ecological implications of the results would better convey your main message
% I recommend streamlining this section and, if preferred, moving part of it to the supplementary material, where you can address the implications of scaling your results with the null model.
%  We have now moved this section into the named ``caveats'' section
% Hopefully this will make it easier to skip for readers no interested in the methodological aspects of our work
% However, if the reviewer still thinks this section is too focused, we're happy to reduce it in a future version.

% Line 321-324
% When you refer to your findings, please provide a more explicit reference to the values you found and the relative graph to better guide the reader through your results and your interpretation.
%  We have now specified which part of figure 2 we were using to illustrate this point:
%  \textit{(e.g.~facilitation mechanism measured as divergence using the probability density method - not distinguishable from null when removing 80\textbackslash\% or 60\textbackslash\% of the data; Figure \textbackslash ref\{Fig:simulations\})}
%  Lines @@@

% Line 327
% Please replace `workers' with `ecologists,' as you are referring to an example from an ecological study.
%  We changed ``workers'' to ``researchers'' because not only ecologists are likely to work on such data/questions (e.g.~evolutionary biologists, palaeobiologists, etc.)

% Line 339-340
% Please simplify and increase readability.
%  We reworded this sentence as follows:
%  \textit{The empirical results are also intriguing in that the loss of Hawaiian bird's functional diversity since 1500 is not greater than expected by chance (at least for certain metrics; Figure \textbackslash ref\{Fig:empirical\_results\}) because one would expect that extinctions since 1500 target specific areas of functional diversity space (i.e.~birds with specific trait combinations).}
%  Lines @@@

% Line 340-361
% I believe this section should be simplified and partially reframed
% While it's clear that you applied your framework to a real study case using empirical data, this section currently diverges too much from the main focus of your study
% Since your primary goal is to test functional diversity metrics, the main discussion should center on the performance of these metrics, their differences from the simulated data, and their relevance to the ``why'', ``what'', and ``how'' framework
% After addressing these points, you can briefly discuss the extinction dynamics of avian species in Hawaii.
%  We have now expanded this section to be clearer:
%  \textit{This illustrates the importance of positing the mechanism and process of interest as well as the pattern to capture it before analysing the results.}
%  \textit{For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance for the species present before 1500 and larger than expected by change for the current species.}
%  \textit{This could lead to the interpretation that the trait space was not changed under any specific mechanism before 1500 (i.e.~null mechanism) but that since then, specific mechanisms have been reducing specific areas of the trait space (e.g.~filtering mechanism).}
%  \textit{However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before 1500 leading to very different interpretation of the results.}
%  Lines @@@

% \textbf{Conclusion}.
%  Similarly to my comments on the discussion, I suggest strengthening the connection between your conclusion and the main ``what, how, and why'' framework
% This would enhance the overall flow of the paper and help the reader more clearly connect the main question and introduction to the final conclusions and discussion.
%  We have now added the box 2 that would hopefully help with strengthening the connection between our results and the need to set up the ``what how and why'' framework for functional diversity analyses.

% \textbf{Supplementary material}.
%  Like for figure 2 in the main manuscript, I think that you are missing the adjusted R square from the figure.
%  We have now specified in the caption that the adjusted R squares are available in the supplementary tables to not clutter the figure.

% I enjoyed reviewing and with some time in revisions, this paper and the messaging will be stronger and more applied
% I appreciate the framework and application this paper will support.

% \textbf{Reviewer: 2}

% Upon reading the title and introduction, I was intrigued and had high expectations for the proposed ``what, how, why'' framework of trait-based analysis
% However, after thoroughly reviewing the manuscript, it became clear that the study does not fully deliver on its initial promise of providing a guideline for applying this framework to ecological data analysis
% Instead, it focuses more on evaluating the varying responses of different functional diversity indices when specific data points are selectively removed from a dataset
% The primary outcome is that different functional diversity indices exhibit distinct and often unpredictable responses to data removal
% However, this observation, while valid, does not seem to offer significant new insights into ecological processes
% In essence, the study highlights the dependence of analytical outcomes on the choice of index---a finding that, while true, feels somewhat self-evident.

% It is possible that these concerns arise due to insufficient descriptions of the methodology and inconsistencies in terminology (specific points are detailed below)
% As the manuscript stands, I am unable to recommend it for publication
% That said, I have outlined several constructive suggestions that I hope will assist the authors in revising and refining the paper.

% \textbf{General comments}

% \textbf{Main model idea}
%  The paper falls short in demonstrating how the ``what, how, why'' framework can be applied to ecological data
% This issue may stem from examining two distinct layers: the simple removal of points from datasets and the subsequent interpretation of these removals in terms of ecological processes, which are not explicitly simulated in the model
% The distinction between these layers is confusing and is primarily addressed in Table 1, which does not effectively clarify these points
% To improve clarity, I recommend a more systematic presentation of these issues in the main text: This would involve, describing the relevant ecological processes (e.g., facilitation, filtering, competition), explaining how these processes influence species-associated trait distributions, and then outlining how these influences are represented in the five trait-removal processes.

% We have now improved both the mechanisms approximation description (see comment below) as well as added more justification on why we believe our algorithms approximate the focal mechanisms (see response to the next comment).
%  Furthermore, we have also added some general predictions as to which type of metric would capture which type of changes in the trait space (Table 1).
%  We hope that these three points improve the clarity of our objectives and justifications.

% \textbf{Model description:}
%  The section on simulating trait-space patterns needs to be significantly expanded to include more technical details regarding how the simulations were conducted
% Key information, such as the model parameters, is entirely absent
% A more thorough explanation of the model setup, parameters, and procedures is required to ensure that the simulations can be reproduced without needing to consult external publications (such as those by Guillerme and Mammola)
% To address this, I suggest adding a detailed explanation of the model in either the main Methods section or in a supplementary appendix, to ensure that all necessary parameters and computational steps are clearly specified.

% We have now developed the explanation of the algorithms description as follows:
%  \textit{\textbackslash begin\{itemize\}}
%  \textit{\textbackslash item Random removal: by randomly removing 20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\% of the data.}
%  \textit{This approximates our \textbackslash textbf\{null mechanism\}
% This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly
% We chose this algorithm to approximate the absence of any specific stressor (i.e.~the removal of data only changes the number of observations, not the properties of the trait space).}

% \textit{\textbackslash item Decreasing size: by removing the required amount (20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\%) of data away from a distance (radius) \(
% rho\) of the centre of the trait space.}
%  \textit{\(
% rho\) is estimated for each trait space to match the required intensity
% For example, for one trait space, the algorithm estimated \(
% rho\_{20}\) to be the minimum radius excluding 20\textbackslash\% of the observations, \(
% rho\_{40}\) the one removing 40\textbackslash\%, etc
% (with \(
% rho\_{20} \> 
% rho\_{40} \> 
% rho\_{60} \> 
% rho\_{80}\)).}
%  \textit{This approximates our \textbackslash textbf\{equalising mechanism\}.}
%  \textit{We chose this algorithm to approximate equalisation assuming that a stressor could increase the probability of extinction for observations with more rare trait combinations (i.e.~observations on the edges of the trait space).}

% \textit{\textbackslash item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least \(D\).}
%  \textit{\(D\) is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least \(n\) pairs of points corresponding to the intensity of the stressor.}
%  \textit{For example \(D\_{20}\) is the distance that excludes any \(n\_{20}\) pairs of points that are a distance of at least \(D\_{20}\) from each other, \(D\_{40}\) is the distance that excludes \(n\_{40}\) pairs, etc
% (with \(D\_{20} \> D\_{40} \> D\_{60} \> D\_{80}\)).}
%  \textit{Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.}
%  \textit{This approximates our \textbackslash textbf\{facilitation mechanism\} (i.e.~the points left are only ones that are close to at least another point in space).}
%  \textit{We chose this algorithm to approximate facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.}
%  \textit{I.e
% pairs of observations that share a close trait combinations are more likely to remain (i.e.~observations adjacent to each other in the trait space).}

% \textit{\textbackslash item Shifting space: by removing the required amount of data from a distance (radius) \(
% rho'\) of the observation with the maximum numerical value on all dimensions.}
%  \textit{This is similar to the decreasing size algorithm but instead of choosing \(
% rho\) to be the radius from the centre of the trait space, \(
% rho'\) is a radius from the ``top right corner'\,' of the trait space.}
%  \textit{In other words the centre of the radius \(
% rho'\) is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with \(
% rho'\_{20} \> 
% rho'\_{40} \> 
% rho'\_{60} \> 
% rho'\_{80}\)).}
%  \textit{This approximates our \textbackslash textbf\{filtering mechanism\}.}
%  \textit{We chose this algorithm to approximate a flitering assuming that a stressor that could increase the probability of extinction for observations further from some trait combination optimum.}
%  \textit{This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).}

% \textit{\textbackslash item Increasing evenness: by resampling the proportion of data (i.e.~20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\%) but with skewed resampling probabilities.}
%  \textit{In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.}
%  \textit{This is done by selecting of \(B\) discrete categories to summarise the distribution - i.e.~bandwidths - using Silverman's ``rule of thumb'\,' \textbackslash citep{[}\textbackslash texttt\{bw.nrd0()\} function in \textbackslash texttt\{R\};{]}{[}{]}\{silverman1986density\}.}
%  \textit{Each of these discrete category \(B\) has an observed probability of sampling of \(b\) (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove \(i\)) of \(i 
% times (1-b)^{p}\).}
%  \textit{Where \(p\) is factor increasing the scaling power of the algorithm (here we used \(p=3\)).}
%  \textit{In other words, observations in categories with a low density (\(b\<0.5\)) were more likely to be resampled and the ones in categories with high density (\(b\>0.5\)) were less likely to be resampled.}
%  \textit{This approximates our \textbackslash textbf\{competition mechanism\}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.}
%  \textit{We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.}
%  \textit{This could approximate the fact that observation that have similar trait combinations are more likely to go extinct, due to competition (e.g.~using the same resources).}
%  \textit{\textbackslash end\{itemize\}}
%  Lines @@@

% \textbf{Consistent terminology:}
%  The paper's clarity is hindered by inconsistent, imprecise, and shifting terminology, making it unnecessarily difficult to follow
% For example, the entities generated by the model are referred to interchangeably as ``elements,'' ``traits,'' ``tips,'' or ``species,'' with no clear rationale for the variation
% Similar, for many other terms that are often used inconsistently or fuzzy (see detailed comments below)
% This lack of consistency creates confusion and often forced me to repeatedly revisit sections of the text to discern the intended meaning
% To improve readability, I recommend establishing a consistent terminology early in the paper and adhering to it throughout
% This would significantly reduce ambiguity and enhance the reader's ability to follow the methodology and conclusions.
%  We have now standardised the terminology following \textbackslash cite\{guillerme2020shifting\} and \textbackslash cite\{mammola2021concepts\}
% In brief, elements, tips or species are now referred to as ``observations''; and traits, dimensions and columns as ``dimensions'' when referring to our simulation approach
% We also added a Glossary to the supplementary materials @@@ for clarity.

% \textbf{Detailed comments:}

% L114: Really ``A single Brownian motion trait''? This implies that only a single trait dimension is generated, whereas from the later text it seems that higher-dimensional traits are simulated.
%  By ``trait'' here we refer to the single process (Brownian Motion) which can be simulated in any number of dimensions (i.e.~generating a time dependent element value using 2, 4 or 8 dimension coordinates to describe it)
% In this context we differentiate ``dimensions'' which can be interpreted as a column in the data matrix, and the ``trait'' which is the definition of the measured (or simulated) feature
% For example, the trait ``geographical location'' can be described using two dimensions (latitude and longitude) or can be described as two traits (``latitude coordinate'' and ``longitude coordinate'')
% In this manuscript we preferred using the first definition
% We've clarified this as follows:
%  \textit{First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.~a pure birth speciation model) until reaching 200 elements in \textbackslash texttt\{R\} \textbackslash citep\{rcore\} using \textbackslash texttt\{treats\} \textbackslash citep\{guillerme2024treats\}
% This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensional space.}
%  Lines@@@

% L115: what is a ``tip''?
%  By tip we meant a node with no descendants in an evolutionary tree
% We have now standardised the terminology (see response to general comment above).

% L115, ``200 elements'': Why are so few data points used? A dataset of only 200 elements (reduced to 40 after data removal) seems exceptionally sparse, especially in a 4- or 8-dimensional space, as noted later in the text
% What is the motivation behind studying such a sparse space? One of the main advantages of using simple models like a branching process is that they allow for the generation of larger datasets, which would provide more robust statistics and better analytical power.
%  @@@TODO: Rerun the simulations with 400, 800 and 1000 elements.
%  But also highlight that 40 species is already quite big
% We can also link that to the use of FD in palaeontology (disparity) where 40 is also already quite big!

% LL 120-124, ``four different stressors\ldots{} five specific algorithms'': Inconsistency in the used terminology
% In this section, ``stressors'' appear to refer to the amount of data being removed (ranging from 20\% to 80\%), but later, the term ``stressors'' is used to describe the five data-removal processes---referred to as ``specific algorithms'' here.
%  We have now standardised the terminology throughout the manuscript (and included it in the glossary table in the supplementary materials): ``algorithm'' now designate the code algorithm, ``stressors'' (and ``stressors' intensity'') now designates how we modify the data (using the algorithms mentioned above) and ``mechanisms'' now designates the biological mechanisms we are approximating using a specific stressor (that is based on a specific algorithm)
% We've namely changed this section for clarity:
%  \textit{We then applied five different stressors to the trait space with different intensities.}
%  \textit{The five different stressors are effectively five specific algorithms removing either 20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\% of the data (the intensities of the stressors, resulting in trait spaces with, respectively, 160, 120, 80 and 40 observations).}
%  \textit{We used the following specific algorithms (Figure \textbackslash ref\{Fig:simulations\}; Table \textbackslash ref\{Tab:mechanisms\}; all algorithms, except ``increasing evenness'\,', were previously described in \textbackslash cite\{guillerme2020shifting\}):}
%  Lines @@@

% L130, ``increasing density'': What is the motivation for this term? While I understand that this removal scheme may result in an overall increase in average density, the removal itself is not directly based on density but rather on pairwise distance in trait space
% For instance, the equalizing scheme would also lead to an increase in density.
%  We used the term ``increasing density'' here to differentiate from ``decreasing size''
% The reviewer is totally correct in their understanding that, especially for normally distributed data, changes in density in the data actually occur to some extent with all the algorithms used here
% The stressor names and their algorithms are based on Guillerme et al 2020 and their implementation in the dispRity package: in their description they don't assume any specific distribution of the data a priori
% In fact, when using a uniform distribution, the decreasing size stressor does not change the density and the increasing density algorithm does not decrease the size of the trait space.
%  We've now added the following note to include the reviewer's comment:
%  \textit{Note that the algorithm used here to simulate the stressors depend, to some extent, to the distribution of the data for each trait space.}
%  \textit{The algorithms used can lead to similar stressors effects depending on the simulated data distribution.}
%  \textit{For example, ``increasing density'\,' and ``decreasing size'\,' stressors will lead to just what their name suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'\,' and just overall decrease in size for the ``decreasing size'\,' stressors) but will have more or less similar effects (albeit no identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'\,' and ``decreasing size'\,' stressors.)}
%  Lines @@@

% L134, ``shifting space'': Similar to the previous point, this term is somewhat misleading
% This removal scheme does not explicitly shift the distribution; rather, it cuts the distribution at one edge, causing the mean to shift
% However, this effect of shifting the mean is not unique to this scheme---other removal schemes will generally result in a shift of the mean as well.
%  This is also a fair point
% See comment above.

% L135, ``with the maximal value'': What does ``maximal value'' mean for points in a higher-dimensional space? Maybe this refer to the location in trait space with the highest density of trait elements?
%  We have now specified what we meant by maximal value:
%  \textit{Shifting space: by removing the required amount of data from a distance (radius) \(
% rho'\) of the observation with the maximum numerical value on all dimensions.}
%  \textit{This is similar to the decreasing size algorithm but instead of choosing \(
% rho\) to be the radius from the centre of the trait space, \(
% rho'\) is a radius from the ``top right corner'\,' of the trait space.}
%  \textit{In other words the centre of the radius \(
% rho'\) is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with \(
% rho'\_{20} \> 
% rho'\_{40} \> 
% rho'\_{60} \> 
% rho'\_{80}\)).}
%  Lines @@@

% LL138- 145: I found this section on the competition method difficult to follow:
%  - The term ``bandwidth'' is unclear
% What does ``bandwidth'' refer to in this context? Does it imply that resampling was done using a frequency-based method, or is it referring to a specific range or interval of trait values in the distribution? This ambiguity ties back to my earlier request for more technical details regarding the simulation process.
%  - I am also unclear on the meaning of ``n''
% Could you clarify what ``n'' represents here? For example, what is meant by the ``probability of sampling of n'' and the reference to ``high/low values of n''? Without this information, it is difficult to grasp how the competition process is functioning.
%  - The term ``flattening the curve'' needs more context
% What curve is being flattened, and how does this relate to the competition method?
%  We have now completely rewritten the description of this stressor by defining more clearly what the bandwidth is, renaming the variables to be easier to understand, removed mention of flattening the curve and rewrote the equations to be specific to the increase in evenness (previous equations were generalised to both increase or decrease the eveness depending on the variable \textit{i} (formerly \textit{n}) being now always a proportion of removal (e.g.~0.2, 0.4, \ldots{} - former description worked with \textit{n} being also a proportion to add e.g.~1.2, 1.4, etc\ldots):
%  \textit{Increasing evenness: by resampling the proportion of data (i.e.~20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\%) but with skewed resampling probabilities.}
%  \textit{In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.}
%  \textit{This is done by selecting of \(B\) discrete categories to summarise the distribution - i.e.~bandwidths - using Silverman's ``rule of thumb'\,' \textbackslash citep{[}\textbackslash texttt\{bw.nrd0()\} function in \textbackslash texttt\{R\};{]}{[}{]}\{silverman1986density\}.}
%  \textit{Each of these discrete category \(B\) has an observed probability of sampling of \(b\) (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove \(i\)) of \(i 
% times (1-b)^{p}\).}
%  \textit{Where \(p\) is factor increasing the scaling power of the algorithm (here we used \(p=3\)).}
%  \textit{In other words, observations in categories with a low density (\(b\<0.5\)) were more likely to be resampled and the ones in categories with high density (\(b\>0.5\)) were less likely to be resampled.}
%  \textit{This approximates our \textbackslash textbf\{competition mechanism\}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.}
%  Lines @@@

% Figure 1: The caption is confusing and needs significant improvement
% Below are some specific points that require attention:
%  - The caption completely omits any description of what is shown in left and right columns (i.e., 1d and 2d case).
%  We have now clarified what each column contains in the figure caption.
%  - It is unclear whether this figure (both the 1d and 2d cases) is based on the actual simulations or is merely a schematic sketch
% For instance, in the right column, it seems unlikely that the figure actually depicts 200 points.
%  The figures do depict 200 points (200 blue in row A and 100 blue + 100 orange in the other columns)
% We have now specified this in the caption.
%  - The caption mentions a ``trait space of 2, 4, or 8 dimensions,'' yet a 1d trait space is shown
% If a 1d trait space is an illustrative example, this should be specified in the caption.
%  We have now specified that we are showing 1 or 2 dimensions regardless of the number of dimensions per traits (i.e.~8 dimensional traits are not fully represented here but what's happening on their first and second dimensions is displayed here)
% We think one confusion on our part is the distinction between a dimension (a column in a matrix) and a trait (a measurable feature for an observation)
% In our case we are always only looking at one trait per trait space
% Although these traits can have 2, 4 or 8 dimensions
% We have now clarified this in the glossary.
%  - The pattern shown in the subfigure labeled ``2d facilitating mechanism'' appears questionable
% If point-pairs from a 2d distribution are removed based on a threshold distance, the resulting pattern would not typically resemble what is depicted here.
%  The illustration of the 2D facilitating mechanism directly comes from the output of the dispRity::reduce.space function where the algorithm is implemented
% Here it selected every \textit{n} pairs of observations that are the closest to each other so that 50\% of the data is selected
% Here \textit{n} is 50 so that 100 observations are selected
% The algorithm is recursive so that once a pair of observations is selected, it cannot be selected again.
%  - Terminology: The use of the term ``species'' in the caption is confusing, as the methods seem to generate traits rather than species.
%  We have now removed any mention of species.
%  - The phrase ``removing half of the species, resulting in two groups of tips'' is unclear
% How are these two groups defined? Presumably, one group consists of the removed traits and the other of non-removed traits
% Additionally, it is unclear which color (blue or orange) corresponds to which group.
%  We have now specified this in the new caption.
%  - I am skeptical about the use of ecological terminology such as ``competition'' and ``facilitation'' in this context
% What is actually being shown are data distributions resulting from different point-removal algorithms, not ecological interactions per se
% While these processes can be discussed and associated with ecological mechanisms in the text or discussion (if well justified), using such terms directly in the figure caption is highly misleading.
%  @@@TODO: Links back to main comment from reviewer 2: Go back to the previous version of table 1 that gives more description of the models
% One thing here is probably to justify more why we think that, for example, ``removing points from the side of a traitspace = filtering''.
%  We now changed the labels in the figure and the caption to highlight that we are indeed approximating such mechanisms, not simulating them per se as the reviewers pointed out.

% LL 150-152: The description is unclear and appears to use the term ``simulation'' inconsistently
% In the phrase ``replicates per simulations,'' the term ``simulation'' seems to have a different meaning than in ``smallest number of simulations.'' Additionally, the first part of the sentence suggests that variance decreases with an increasing number of simulations, whereas the second part implies that adding more simulations leads to increased variance.
%  There was a confusion indeed
% We now refer to ``the simulation pipeline'' for the whole set of replicates per stressors, and ``replicates'' for each individual run (one trait space + one stressor)
% We changed this sentence to:
%  \textit{We used 171 replicates because that was empirically the smallest number of replicates required to reach a variance between replicates lower than 1\textbackslash\% across all metrics (i.e.~any additional replicates beyond 171 added less that 1\textbackslash\% extra variance).}
%  Lines @@@

% L 166: what does ``uncorrelated'' mean here? Does this mean that the branching process was independently repeated four times, once for each trait dimension, or does it mean that it generates four trait dimensions simultaneously, with these dimensions being uncorrelated?
%  We meant that the dimensions for each multidimensional traits are uncorrelated (i.e.~the trait is a uncorrelated Brownian motion process in n dimensions)
% We have now clarified this point:
%  \textit{We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \textbackslash ref\{Fig:simulation\_results\}) for 2 and 8 dimensional traits with not correlation between each dimensions in each simulation pipeline (supplementary materials).}
%  Lines @@@

% LL 170-178: why then did the authors not simply generate more data points (see my comment above)?
%  @@@TODO: Rerun the simulations with 400, 800 and 1000 elements.

% Table 1: The integration of this table into the main text could be improved
% Specifically, the abrupt mention of species (which are not part of the model) is confusing.
%  We have now improved the context of the tables with predictions and removed most of the mentions of ``species'' now replaced by the standardised term of ``observations'' (unless they were specific examples).

% L 178 and several instances later ``o'' should read ``0'' (zero)
%  These integers were compiled using the latex mathpazo font
% We have now clarified them by changing into a default serif font.

% L184, I suggest to mention here, that these functional metrics are denoted as trait space occupancy later on.
%  Great suggestions
% We added the following:
%  \textit{We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (trait space occupancy metrics; \textbackslash cite\{mammola2021concepts\}) using 12 metrics (Tab
% \textbackslash ref\{Tab:metrics\}):}
%  Lines@@@

% L 208, ``\textasciitilde{}'' symbol: This usage of mathematical notation (instead of plain words) in a sentence (here, but also later in the text) is difficult to understand.
%  We removed tilde symbols throughout the manuscript.

% Table 2: Column two (``Methods'') does not provide any information beyond what is explained in ``Description'' and can be removed.
%  We decided to keep the ``methods'' column to make it easier for the reader to refer to the terms used in Mammola et al.~2021 at a glance.

% LL 252-260 ``opposite pattern to what was expected'' and ``naïve expectation'': I do not agree
% It is well known that removing data points can increase the evenness of a distribution, so it is not surprising that removal procedures can result in a reduction of the metric scores
% The term ``space occupancy'' could be contributing to this misconception.
%  We rewritten this section to be in line with the reviewer's comment:
%  \textit{Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted into a score increase (e.g.~the functional evenness based on hypervolume for the equalising mechanism) or non-linear responses (e.g.~Alpha diversity based on dendrogram for the equalising mechanism).}
%  Lines @@@
%  Additionally, there appears to be a logical inconsistency
% Figure 2 illustrates the difference between two functional diversity metrics, rather than the metrics themselves
% Therefore, changes in this difference do not provide information about the individual increases or decreases of the metrics.
%  We have now added supplementary figures showing the scaled metric score changes (to keep the values manageable between 0 and 1) but not also scaled to the null mechanism changes (to display just the raw metric changes).

% L 287: Why would this be ``unexpected''?
%  We removed the word ``unexpected'' that was maybe ironic (and irrelevant regardless).

% LL 287-289: I do not see that the ``what, how, why''-framework has really been used in this study
% Please specify, for example, which mechanisms have correctly been explained in the analysis?
%  We have now reworked this section with the inclusion of a Box explaining in more details how we believe that using the ``what, how, why'' framework helps choosing the correct metric for a specific dataset and question.

% \textbf{Reviewer: 3}

% The question of whether modern methods are effective in extracting functional diversity patterns and relating them to ecological drivers is fundamentally important
% The authors conducted a consistent and accurate set of simulations to test how different methods and metrics measure functional diversity under various ecological processes, considering both simulated and real examples
% The approach used is reasonable and appropriate
% However, as it stands, I find it challenging to identify the specific hypotheses that the study aims to address, particularly regarding how different metrics or methodologies should approximate the different mechanisms tested
% Additionally, I found the discussion of the results to be too broad and lacking clear justification for the findings
% Consequently, the manuscript, as it reads, does not fully convey the novelty and the scientific gap filled by this work
% Therefore, I believe that a revision would be beneficial to improve the clarity of the manuscript's message.

% \textbf{Specific comments:}

% I think that, while the idea of the ``what?'', ``how?'', and ``why?'' questions is intriguing, the Introduction does not sufficiently articulate the specific aims addressed in the manuscript
% Specifically, through simulations, the authors approximate different ecological mechanisms that produce patterns in the functional space (as partially shown in Figure 1)
% These patterns and mechanisms are addressed in section 3.1.1 of the Methods, yet it is not clear how the different functional diversity metrics (i.e., richness, divergence, and evenness) should behave under these spatial reorganizations, particularly when compared to a null mechanism
% For instance, under the Equalizing mechanism, I would hypothesize that functional richness and divergence would decrease compared to the Null mechanism, while species distribution would become more even, given that species removed from the extremes of the space would pack towards the centre
% The signal might be expected to be stronger with increasing species removal, as species packing increases
% An accurate algorithm or method should be capable of detecting such a signal
% However, for mechanisms such as competition, the reasoning is not as straightforward (e.g., I am not convinced that richness metrics would decrease compared to the null mechanism in this case), particularly concerning expectations for divergence and evenness, as well as the relationship of these metrics to increasing species loss
% I believe that formulating clear ``hypotheses'' on species patterns in the space in relation to the various mechanisms tested may help readers better navigate through the metrics analysed
% This part could also be enhanced by introducing a new figure or expanding Figure 1 or Table 1 (which I found very well done).

% We have now improved the description of the algorithms (lines @@@) as well as added our predictions to Table 1 and improved Figure 1 to be a better tool to understand our simulation protocol and how to read the results figures.
%  Furthermore we agree with the reviewer that the effect of different mechanisms can be sometimes more intuitive than others (e.g.~in their example and counterexample of equalizing vs.~competition).

% We've expanded on the following paragraph:
%  \textit{Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.}
%  \textit{For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.}
%  \textit{However, if the data are distributed uniformly, this does not happen.}
%  \textit{Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.}
%  \textit{However, this is not tackled here for both simplicity and to understand how they work in isolation.}
%  \textit{Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \textbackslash ref\{Tab:mechanisms\}) should be taken as a cautious guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.}
%  Lines @@@

% Lines 165-170 and 303-307: As mentioned, TPD methods are implemented for a maximum of four dimensions
% I am curious about how the metric for the eight-dimensional space was computed
% Furthermore, although eight-dimensional spaces may be considered ``relatively small'' in morphological studies, ecological studies addressing both plants (Laughlin 2014, Díaz et al.~2016, Weigelt et al.~2021) and animals (Healy et al.~2019, Mouillot et al.~2021, Cox et al.~2022, Beccari et al.~2024) generally use far fewer than eight dimensions
% Accordingly, I wonder if this was the reason behind selecting the four-dimensional space as the main analysis in Figure 2, or was it to allow comparison with the dimensionality of the empirical trait space? I believe this selection requires justification in the text.

% We add the following clarification in the methods for the 8 dimensional spaces:
%  \textit{We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \textbackslash texttt\{TPD::TPDsMean\} is only implemented for up to 4 dimensions; \textbackslash citealt\{carmona2019trait\} - note that int the supplementary figures, we measured the \textbackslash texttt\{TPD\} metrics on only the four first dimensions)}
%  Lines @@@
%  And:
%  \textit{Although this number of dimensions is in the range of many studies in ecology (e.g.~6 dimensions in \textbackslash citealt\{healy2019animal\}), it is not uncommon to use a much greater number of dimensions, especially in palaeontology (e.g.~more than 200 dimensions in \textbackslash citealt\{van2023should\}).}
%  Lines @@@

% We also specified the that we used only the four first dimensions for the TPD metrics in the supplementary results

% The use of bird traits is an interesting aspect that fits well with the scope of the manuscript
% However, for the same reason mentioned in the previous comment and as stated by the authors in lines 165-181, I am puzzled by the choice of five dimensions to portray the AVONET variation
% Morphological traits are generally highly correlated, resulting in a few axes explaining the highest proportion of trait variation and a space highly clustered in the centre with few species occupying its outer part (Pigot et al.~2020, Germain et al.~2023, Toussaint et al.~2024)
% How is the selection of five dimensions justified for the analysis? Are these properties of the dataset considered important for the example?
%  We chose to use five dimensions to describe the Hawaiian birds trait space to make sure that each focal group has at least 95\% of their variance described in the trait space (following Guillerme et al.~2023 protocol).
%  Furthermore, the choice of 5 dimensions is congruent with the suggestions from Pigot et al 2020 (``\textit{We have demonstrated that a minimum of four independent morphological trait axes are required to predict variation in avian trophic niches, calling into question the validity of trait-based macroecological analyses that assess functional diversity on the basis of fewer morphological trait dimensions}.'').
%  We have now specified this in the main text:
%  \textit{We selected the first \(5\) axes to represent the trait spaces that explained at least 95\textbackslash\% of the variance in each specific group (historic and extant species - see supplementary figure 7; following recommendations of \textbackslash citealt\{pigot2020macroevolutionary\} and dimensionality selection method from \textbackslash citealt\{guillerme2023innovation\}).}
%  Lines @@@

% I found the Results section very broad and believe it would benefit from specifying further details about discrepancies and similarities found among metrics
% Referring to my first comment, the authors mention expected patterns (like a decrease in trait space occupancy) that are not previously introduced in the text
% Similarly, I am not convinced by the expectation that all metrics (i.e., richness, evenness, and divergence) should generally increase compared to a null mechanism
% For instance, if the competition mechanism is removing species from the centre of the distribution (as specified in Figure 1), wouldn't it produce a decrease in functional richness but an increase in its divergence?

% We have now shorten the results to be more descriptive:

% \textit{The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.}
%  \textit{It ranged from metrics capturing no clear pattern (e.g.~the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g.~Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).}
%  \textit{Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted into a score increase (e.g.~the functional evenness based on hypervolume for the equalising mechanism) or non-linear responses (e.g.~Alpha diversity based on dendrogram for the equalising mechanism).}
%  \textit{Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \textbackslash ref\{Fig:simulation\_results\}).}
%  \textit{Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulation show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).}
%  Lines @@@

% However, we have now greatly expanded our discussion of these results in the discussions section under two subheadings (results, Lines @@@; and caveats Lines @@@)
% We felt there was a dilemma in 1) expanding the results at the cost of the discussion (i.e.~longer results section and shorter discussion one) which would have highlighted specific results but maybe downplayed some caveats; or 2) expanding the discussion at the cost of the results section which we felt more highlighted what we believe is the main takeaway from the results (``\textit{The ability of different metrics to capture the different patterns (and thus approximate the mechanisms) was highly variable.''}) and expanding where this variability comes from in the discussion (difficulties of interpreting multidimensional spaces + caveats stemming from our methodological choices).

% Lines 332-337: The rationale behind the choice of scaling for the null mechanism is unclear to me
% While I agree with the authors that this transformation may help in data visualization and interpretation, if scaling introduces such an artifact, why rely so heavily on these comparisons? Wouldn't it be more straightforward to simply compare metrics across each other without scaling?

% We have now expanded the section in the methods discussing the scaling:
%  \textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relatively to the same metric scores of a null mechanism.}
%  \textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between \(0\) and \(1\).}
%  \textit{We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.~\(metric\_{stressor20
% %} \- metric\_{null20
% %}\)).}
%  \textit{This resulted in each metric being scaled between \(–1\) and \(1\) where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.~approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near \(0\) meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \textbackslash ref\{Fig:simulations\}, \textbackslash ref\{Fig:simulation\_results\} and \textbackslash ref\{Fig:empirical\_results\}).}
%  \textit{We interpret this last scenario as the metric score performing poorly, i.e.~it is unable to clearly distinguish between a random removal of elements and a non random one.}
%  Lines @@@

% As well as updated figure 1 to hopefully improve the readability of the scaled results
% We believe that they are more interpretable than just displaying the raw results (now also available in the supplementaries), especially since it highlights how some metrics cannot distinguish between the null and the focal mechanism (e.g.~functional evenness using the dendrogram method for the competition) but can still distinguish from the null even when the range of change seems rather small (e.g.~functional evenness using the hypervolume method for competition).

% While I agree with the conclusion that the metric of choice should align with the research question or the specific pattern analyzed, I feel the Discussion is missing some mechanistic links
% The take-home message that different metrics are suitable for different ecological processes should be paired with an explanation or at least a suggestion of which metric may be more appropriate for which process
% For instance, the highest dispersion observed in the empirical dataset for RaoQ compared to other metrics may relate to the lower accuracy of the metric
% Without further details on the mechanisms of the metrics analyzed, I fear the reader may lose sight of the importance of metric choice and the implications of their different behaviors in addressing ecological processes.

% Although we did wish to expand the discussion with more clear messages of which metrics to use in which circumstances, we do not believe that the findings of our analysis justify prescriptive choice of specific metrics for specific scenarios
% As highlighted by the reviewer, although some metrics seem to work as better tools to capture changes in some specific scenarios, these are highly dependent on the type of data used here
% In the example of using the RaoQ having a lower accuracy than other methods, this might be true for this example dataset but we cannot posit that it would be the case for other datasets.

% To convey this message more clearly, we have now entirely restructured the discussion by separating it into sub-sections (simulated results, empirical results and caveats) as well as added a new text box (box 2) summarising how we believe that thinking more carefully about the pattern, process and mechanism allows for a better choice of functional diversity metric.

\end{document}

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
update response

@@
 \textit{This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly.}
-Lines @@@
+Lines 149-151.
 
-\textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.~\(metric\_stressor\_20 \- metric\_null\_20\)).}
-\textit{We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).}
-Lines @@@
+\textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
+$metric_{stressor20\%} - metric_{null20\%}$).
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).}
+Lines 274-279.
 
-% The effect of the focal mechanisms is well explained, but I believe a more explicit explanation of how these mechanisms were tested is necessary
-% I recommend adding clear statements at the end of the introduction that outline how you tested the different mechanisms against the null model, along with the associated hypotheses and predictions (perhaps in table 1)
-% While you briefly mention this in the methods section (e.g., lines 198-201), it would be beneficial to include this information earlier to provide readers with a clearer framework for understanding your approach.
+\textcolor{blue}{The effect of the focal mechanisms is well explained, but I believe a more explicit explanation of how these mechanisms were tested is necessary.
+I recommend adding clear statements at the end of the introduction that outline how you tested the different mechanisms against the null model, along with the associated hypotheses and predictions (perhaps in table 1)
+While you briefly mention this in the methods section (e.g., lines 198-201), it would be beneficial to include this information earlier to provide readers with a clearer framework for understanding your approach.}
 
-% We have now clarified our goal and expectations at the end of the introduction:
-%  \textit{Our goal is to assess the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
-%  \textit{We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
-%  Lines @@@.
+We have now clarified our goal and expectations at the end of the introduction:
 
-% We have clarified how and why we compare the metrics to the null mechanisms:
-%  \textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relatively to the same metric scores of a null mechanism.}
-%  \textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between \(0\) and \(1\).}
-%  \textit{We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.~\(metric\_{stressor20
-% %} \- metric\_{null20
-% %}\)).}
-%  \textit{This resulted in each metric being scaled between \(–1\) and \(1\), where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.~approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near \(0\) meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \textbackslash ref\{Fig:simulations\}, \textbackslash ref\{Fig:simulation\_results\} and \textbackslash ref\{Fig:empirical\_results\}).}
-%  \textit{We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.}
-%  Lines @@@
+\textit{Our goal is to assess the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
+We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
+Lines 109-114.
 
-% Furthermore, we have added an expected results column to Table 1 as suggested.
+We have clarified how and why we compare the metrics to the null mechanisms:
 
-% Additionally, the use of both simulated and real data in your study should be better explained, with more details provided on the value and rationale behind this choice
-% This would help to further justify your methodology and enhance the overall clarity and impact of your manuscript.
+\textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relative to the same metric scores of a null mechanism.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
+$metric_{stressor20\%} - metric_{null20\%}$).
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).
+This resulted in each metric being scaled between $–1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
+approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs 1, 2 and 3).
+We interpret this last scenario as the metric score performing poorly, i.e. it is unable to clearly distinguish between a random removal of elements and a non random one.}
+Lines 273-287.
+
+Furthermore, we have added an expected results column to Table 1 as suggested.
+
+\textcolor{blue}{Additionally, the use of both simulated and real data in your study should be better explained, with more details provided on the value and rationale behind this choice.
+This would help to further justify your methodology and enhance the overall clarity and impact of your manuscript.}
 

### 2024-11-28 Manuscript/main_ecography_R1.tex
update response

@@
 To understand the ability of each metric to capture a mechanism of interest, we compared their scores relative to the same metric scores of a null mechanism.
-To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metric by their absolute greatest value to scale them between $0$ and $1$.
-We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
 $metric_{stressor20\%} - metric_{null20\%}$).
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).
 This resulted in each metric being scaled between $–1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
Update ecography_reponse_letter.tex

@@
 
-% We have rewritten the last paragraph of the introduction to include the following:
-%  \textit{Beyond simulations, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.}
-%  \textit{We show that, with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.}
-%  \textit{We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.}
-%  Lines @@@
+We have rewritten the last paragraph of the introduction to include the following:
+
+\textit{To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data.
+We show that, with a fixed process and mechanisms approximation, the choice of statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data, with the exact same data being sometimes interpreted in opposite ways.}
+Lines 114-122.
+
+\textit{We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.}
+Lines 570-573.
 

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
up

@@
 We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).}
-Lines 274-279.
+Lines 273-279.
 
@@
 $metric_{stressor20\%} - metric_{null20\%}$).
-We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).
-This resulted in each metric being scaled between $–1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.)
+This resulted in each metric being scaled between $-1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
 approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs 1, 2 and 3).
-We interpret this last scenario as the metric score performing poorly, i.e. it is unable to clearly distinguish between a random removal of elements and a non random one.}
+We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.}
 Lines 273-287.
@@
 
-% \textbf{Item 2.}
-%  While I appreciate the detailed explanation of the methodology used to produce the simulated space patterns and the application of the stressors, this section appears disproportionate compared to the rest of the methods, especially concerning the testing of focal metrics
-% I recommend enhancing the clarity of Sections 3.1.3 and 3.1.4
-% Together with providing more explicit statements regarding your hypotheses and predictions (see item 1), it would help readers navigate the different aspects of the analyses
-% For instance, clarifying how the performance of each metric in capturing various focal mechanisms was evaluated, and how different results support or challenge this evaluation, would be beneficial.
+\textcolor{blue}{\textbf{Item 2.}}
 
-% We have now rewritten sections 3.1.3 and 3.14 to be clearer:
-%  \textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values, normalizing them to a range between \(0\) and \(1\).}
-%  \textit{We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.~\(metric\_{stressor20
-% %} \- metric\_{null20
-% %}\)).}
-%  \textit{This resulted in each metric being scaled between \(–1\) and \(1\) where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.~approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near \(0\) meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \textbackslash ref\{Fig:simulations\}, \textbackslash ref\{Fig:simulation\_results\} and \textbackslash ref\{Fig:empirical\_results\}).}
-%  \textit{We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.}
-%  \textit{To assess these poor performances, we compared the distributions of the \(171\) scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.}
-%  \textit{The results of these tests are displayed in figure \textbackslash ref\{Fig:simulation\_results\} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).}
-%  Lines @@@
+\textcolor{blue}{While I appreciate the detailed explanation of the methodology used to produce the simulated space patterns and the application of the stressors, this section appears disproportionate compared to the rest of the methods, especially concerning the testing of focal metrics
+I recommend enhancing the clarity of Sections 3.1.3 and 3.1.4
+Together with providing more explicit statements regarding your hypotheses and predictions (see item 1), it would help readers navigate the different aspects of the analyses
+For instance, clarifying how the performance of each metric in capturing various focal mechanisms was evaluated, and how different results support or challenge this evaluation, would be beneficial.}
+
+We have now rewritten sections 3.1.3 and 3.14 to be clearer:
+
+\textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relative to the same metric scores of a null mechanism.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
+$metric_{stressor20\%} - metric_{null20\%}$).
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.)
+This resulted in each metric being scaled between $-1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
+approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \ref{Fig:simulations}, \ref{Fig:simulation_results} and \ref{Fig:empirical_results}).
+We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.
+To assess these poor performances, we compared the distributions of the $171$ scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.
+The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).}
+Lines 273-291.
 

### 2024-11-28 Manuscript/main_ecography_R1.tex
up

@@
 $metric_{stressor20\%} - metric_{null20\%}$).
-We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.).
-This resulted in each metric being scaled between $–1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.)
+This resulted in each metric being scaled between $-1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
 approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \ref{Fig:simulations}, \ref{Fig:simulation_results} and \ref{Fig:empirical_results}).
-We interpret this last scenario as the metric score performing poorly, i.e. it is unable to clearly distinguish between a random removal of elements and a non random one.
+We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.
 To assess these poor performances, we compared the distributions of the $171$ scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
reviewer 1 item 2 done

@@
 This resulted in each metric being scaled between $-1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
-approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \ref{Fig:simulations}, \ref{Fig:simulation_results} and \ref{Fig:empirical_results}).
+approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs 1, 2 and 3).
 We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.
 To assess these poor performances, we compared the distributions of the $171$ scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.
-The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).}
+The results of these tests are displayed in figure 2 (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).}
 Lines 273-291.
 
-% \textit{Finally, we measured the effect of the different strengths of the stressors (i.e.~removing 20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% and 80\textbackslash\% the data) by fitting a linear regression using the scaled metric difference to null score as a function of the amount of data removed.}
-%  \textit{For each of these models, we reported the variance explained by the model (adjusted \(R^2\)) in the supplementary tables \(1\) to \(3\) and whether the slope was clearly distinguishable from \(0\) (Figure \textbackslash ref\{Fig:simulation\_results\}).}
-%  \textit{This allowed us to see if there was any predictable linearity in the metric scores.}
-%  \textit{In other words testing whether the metric scores relative to null removals where linearly scaling with the amount of data removed.}
-%  Lines @@@
+\textit{Finally, we measured the effect of the different strengths of the stressors (i.e.
+removing 20\%, 40\%, 60\% and 80\% the data) by fitting a linear regression using the scaled metric difference to the null score as a function of the amount of data removed
+For each of these models, we reported the variance explained by the model (adjusted $R^2$) in the supplementary tables $2$ to $5$ and whether the slope was clearly distinguishable from $0$ (Figure 2b; \textit{sensu} Dushoff et al.
+2019).
+This allowed us to see if there was any predictable linearity in the metric scores.
+In other words testing whether the metric scores relative to null removals where linearly scaling with the amount of data removed.}
+Lines 292-299.
 
-% We have also added a panel to figure 1 helping the reader to easily interpret our results in figures 2 and 3.
+We have also added a panel to figure 1 helping the reader to easily interpret our results in figures 2 and 3.
 

### 2024-11-28 Manuscript/main_ecography_R1.tex
reviewer 1 item 2 done

@@
 Finally, we measured the effect of the different strengths of the stressors (i.e.
-removing 20\%, 40\%, 60\% and 80\% the data) by fitting a linear regression using the scaled metric difference to the null score as a function of the amount of data removed.
-For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables $1$ to $3$ and whether the slope was clearly distinguishable from $0$ (Figure \ref{Fig:simulation_results}; \textit{sensu} \cite{dushoff2019can}).
+removing 20\%, 40\%, 60\% and 80\% the data) by fitting a linear regression using the scaled metric difference to the null score as a function of the amount of data removed
+For each of these models, we reported the variance explained by the model (adjusted $R^2$) in the supplementary tables $2$ to $5$ and whether the slope was clearly distinguishable from $0$ (Figure \ref{Fig:simulation_results}; \textit{sensu} \citealt{dushoff2019can}).
 This allowed us to see if there was any predictable linearity in the metric scores.
-In other words testing whether the metric scores relative to null removals were linearly scaling with the amount of data removed.
+In other words testing whether the metric scores relative to null removals where linearly scaling with the amount of data removed.
 

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
Item 3

@@
 
-% \textbf{Item 3.}
-%  In my opinion the primary issue in the manuscript lies in how the results are communicated
-% As mentioned in the abstract and introduction, your objective is to evaluate the effectiveness of various functional diversity metrics in capturing patterns and signals associated with ten distinct processes linked to common community assembly mechanisms
-% From what I understand, your aim is to offer a framework that assists ecologists in selecting the most appropriate diversity metric based on the relevant mechanism and observed patterns
-% However, I believe this crucial message is not clearly conveyed
-% Instead, the main takeaway from your results and their interpretation seems to be the high variability observed in the performance of the studied metrics.
+\textcolor{blue}{\textbf{Item 3.}}
+\textcolor{blue}{In my opinion the primary issue in the manuscript lies in how the results are communicated.
+As mentioned in the abstract and introduction, your objective is to evaluate the effectiveness of various functional diversity metrics in capturing patterns and signals associated with ten distinct processes linked to common community assembly mechanisms.
+From what I understand, your aim is to offer a framework that assists ecologists in selecting the most appropriate diversity metric based on the relevant mechanism and observed patterns.
+However, I believe this crucial message is not clearly conveyed.
+Instead, the main takeaway from your results and their interpretation seems to be the high variability observed in the performance of the studied metrics.}
 
-% I recommend revising the results and discussion sections to address this issue more effectively
-% Specifically, it would be helpful to provide a more detailed account of your main findings in the results section, particularly how different metrics performed in relation to the significance of your results, especially regarding the impact of data removal
-% Additionally, a more thorough interpretation of these results is needed in the discussion (what really means that a specific metric deviate from the null model, and what the effect of data removal from 20\% to 80\% really implies)
-% In the current manuscript, the discussion begins with an early focus on the limitations and caveats of your methods
-% While acknowledging these aspects is important, it would be more effective to simplify this discussion and move it to a later section
-% Instead, I would emphasize the interpretation of the performance of the various metrics to better communicate your main findings.
+\textcolor{blue}{I recommend revising the results and discussion sections to address this issue more effectively
+Specifically, it would be helpful to provide a more detailed account of your main findings in the results section, particularly how different metrics performed in relation to the significance of your results, especially regarding the impact of data removal
+Additionally, a more thorough interpretation of these results is needed in the discussion (what really means that a specific metric deviate from the null model, and what the effect of data removal from 20\% to 80\% really implies)
+In the current manuscript, the discussion begins with an early focus on the limitations and caveats of your methods
+While acknowledging these aspects is important, it would be more effective to simplify this discussion and move it to a later section
+Instead, I would emphasize the interpretation of the performance of the various metrics to better communicate your main findings.}
 
-% We have now entirely reorganised the discussion section with clearer subsections:
+We have now entirely reorganised the discussion section with clearer subsections:
 
-% \begin{itemize} \tightlist \item   Results (including a subsection on empirical results) discussing in   more details the result
-% Lines @@@
-%  \item   Caveats: discussing clearly the caveats of our approach
-% Lines @@@
-%  \item   Box 2: ``so which metric do \textbackslash textit\{I\} choose?''   proposing a didactic approach and reflections to hopefully easily help   workers to choose their own metric
-% Lines @@@ \end{itemize}
+\begin{itemize} 
+  \item \textbf{Capturing approximated mechanisms with different metrics} discussing the simulated result in more details. Lines 369-405
+  \item \textbf{Empirical results} focusing on the empirical results. Lines 407-448
+  \item \textbf{Caveats} discussing clearly the caveats of our approach. Lines 450-505.
+  \item  \textbf{Box 2: ``so which metric do \textit\{I\} choose?''} proposing a didactic approach and reflections to hopefully easily help workers to choose their own metric. Lines 506-553.
+  \item \textbf{Conclusion} to briefly wrap up our findings. Lines 555-572.
+\end{itemize}
 

### 2024-11-28 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

@@
 
-For example, say one is interested in looking at changes in functional diversity using a divergence metric using the probability density method (e.g. \texttt{TPD::REND}).
-They can then apply this metric to their trait space and get two relative metric scores of, says, 0.7 and 0.65 (pattern) for two levels of extinction (process).
+For example, say one is interested in looking at changes in functional diversity using a divergence metric calculated with the probability density method (e.g. \texttt{TPD::REND}).
+They can apply this metric to their trait space and get two relative metric scores of, say, $0.7$ and $0.65$ (pattern) for two levels of extinction (process).
 The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
Update ecography_reponse_letter.tex

@@
   \item \textbf{Caveats} discussing clearly the caveats of our approach. Lines 450-505.
-  \item  \textbf{Box 2: ``so which metric do \textit\{I\} choose?''} proposing a didactic approach and reflections to hopefully easily help workers to choose their own metric. Lines 506-553.
+  \item  \textbf{Box 2: ``so which metric do \textit{I} choose?''} proposing a didactic approach and reflections to hopefully easily help workers to choose their own metric. Lines 506-553.
   \item \textbf{Conclusion} to briefly wrap up our findings. Lines 555-572.
@@
 
-% \textbf{Item 4.}
-%  Finally, the discussion section needs additional work to better address the key aspects of ``what,'' ``how,'' and ``why'' in functional analyses
-% The introduction emphasizes applying these fundamental questions of ecological analysis to the study of functional diversity metrics
-% However, this focus seems to be lost in the discussion, which lacks a strong connection to the introduction
-% Given that this is the core of your study (as reflected in the title), it should be the primary focus of the discussion section.
-%  I recommend revising the initial part of the introduction to better align it with your main question and hypotheses and to clearly convey the core message of your study
-% For example, a more detailed interpretation of how your results on different metrics relate to this aspect of your framework would be helpful
-% As I mentioned in my previous item, acknowledging the limitations of your methods is important, but it should be addressed in a separate paragraph following the interpretation of your results, rather than being the primary focus of the discussion.
+\textcolor{blue}{\textbf{Item 4.}}
 
-% We have now linked the framework of ``what'', ``how'' and ``why'' more clearly in the discussion:
-%  \textit{For example, say one is interested in looking at changes in functional diversity using a divergence metric calculated with the probability density method (e.g.~\textbackslash texttt\{TPD::REND\}).}
-%  \textit{They can apply this metric to their trait space and get two relative metric scores of, says, 0.7 and 0.65 (pattern) for two levels of extinction (process).}
-%  \textit{The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.}
-%  \textit{In our simulations, if the mechanism approximated was the equalising one, the pattern can clearly be interpreted as a change in functional diversity, but if the mechanism is a facilitation one, the pattern can be meaningless (i.e.~not clearly distinguishable from a null mechanism - Fig
-% \textbackslash ref\{Fig:simulation\_results\} shows this example by contrasting row 5, columns 1 and 2).}
-%  \textit{In this made up example, the procedure is done in reverse order where the the pattern is specified first and then the process and the mechanism (``What process and mechanism are needed to interpret pattern X?'\,').}
-%  \textit{In Box 2 we suggest an approach that specifies the process and the mechanism first (``how do I interpret pattern X under a specific process and mechanism'\,'; see Box 2).}
-%  Lines @@@
+\textcolor{blue}{Finally, the discussion section needs additional work to better address the key aspects of ``what,'' ``how,'' and ``why'' in functional analyses.
+The introduction emphasizes applying these fundamental questions of ecological analysis to the study of functional diversity metrics.
+However, this focus seems to be lost in the discussion, which lacks a strong connection to the introduction.
+Given that this is the core of your study (as reflected in the title), it should be the primary focus of the discussion section.
+I recommend revising the initial part of the introduction to better align it with your main question and hypotheses and to clearly convey the core message of your study.
+For example, a more detailed interpretation of how your results on different metrics relate to this aspect of your framework would be helpful.
+As I mentioned in my previous item, acknowledging the limitations of your methods is important, but it should be addressed in a separate paragraph following the interpretation of your results, rather than being the primary focus of the discussion.}
 
-% As well as in the new Box 2
-% Lines @@@
+We have now linked the framework of ``what'', ``how'' and ``why'' more clearly in the discussion:
 
-% \textbf{Item 5.}
-%  Ensure consistency in the use of terms such as ``elements of the model'' versus ``tips'' or ``species'', and adhere to a uniform citation format (e.g., using or omitting a comma before the year).
+\textit{For example, say one is interested in looking at changes in functional diversity using a divergence metric calculated with the probability density method (e.g. \texttt{TPD::REND}).
+They can apply this metric to their trait space and get two relative metric scores of, say, $0.7$ and $0.65$ (pattern) for two levels of extinction (process).
+The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.
+In our simulations, if the mechanism approximated was the equalising one, the pattern can clearly be interpreted as a change in functional diversity, but if the mechanism is a facilitation one, the pattern can be meaningless (i.e.
+not clearly distinguishable from a null mechanism - Fig.
+2 shows this example by contrasting row 5, columns 1 and 2).
+In this made up example, the procedure is done in reverse order where the the pattern is specified first and then the process and the mechanism (``What process and mechanism are needed to interpret pattern X?'').
+In Box 2 we suggest an approach that specifies the process and the mechanism first (``how do I interpret pattern X under a specific process and mechanism''; see Box 2).}
+Lines 385-398
 
-% We have now standardised terminology and provide a glossary in the supplementary materials (based on Mammola 2021).
+As well as in the new Box 2 lines 506-553.
 
-% Overall, addressing these areas will significantly enhance the paper's impact, readability, and contribution to the field of functional ecology.
+\textcolor{blue}{\textbf{Item 5.}}
 
-% Here are more detailed comments on the different sections of the paper.
+\textcolor{blue}{Ensure consistency in the use of terms such as ``elements of the model'' versus ``tips'' or ``species'', and adhere to a uniform citation format (e.g., using or omitting a comma before the year).}
 
-% \textbf{Abstract}.
-%  Line 3-6: Please rephrase to simplify and avoid brackets.
+We have now standardised terminology and provide a glossary in the supplementary materials (table 1; based on Mammola 2021).
+
+\textcolor{blue}{Overall, addressing these areas will significantly enhance the paper's impact, readability, and contribution to the field of functional ecology.}
+
+\textcolor{blue}{Here are more detailed comments on the different sections of the paper.}
+
+\textcolor{blue}{\textbf{Abstract}.}
+
+\textcolor{blue}{Line 3-6: Please rephrase to simplify and avoid brackets.}
 %  We changed the sentence to:
@@
 
-% Line 10-11: Please rephrase to avoid brackets.
+\textcolor{blue}{Line 10-11: Please rephrase to avoid brackets.}
 %  We changed the sentence to:
@@
 
-% Line 14-15: If space requirements permit, please include more detailed information about your results
-% The current description is too vague and does not effectively convey your main findings.
+\textcolor{blue}{Line 14-15: If space requirements permit, please include more detailed information about your results.
+The current description is too vague and does not effectively convey your main findings.}
 %  We've expanded a little on the results
@@
 
-% \textbf{Introduction.}
-%  Line 25-28
-% It appears to me that you are using the definition of a functional trait from Violle et al.~(2007) in Oikos
-% Please include a citation for this reference if it is the case.
+\textcolor{blue}{\textbf{Introduction.}}
+
+\textcolor{blue}{Line 25-28.
+It appears to me that you are using the definition of a functional trait from Violle et al.(2007) in Oikos.
+Please include a citation for this reference if it is the case.}
 %  We've added the reference as suggested.
 
-% Line 37
-% I would avoid using the term ``new trait-focused ecology.'' While it's true that a more ``explicit'' approach to functional ecology has emerged over the past few decades, the foundation of functional analyses is rooted in the field of ecology itself
-% For example, as you mentioned, Darwin's finches represent an early functional analysis centered on beak traits
-% Although the specific term ``functional traits'' was less commonly used in the past, broader functional analyses were not uncommon, particularly in plant ecology (see, for example, Calow, P
-% 1987
-% ``Towards a Definition of Functional Ecology.'' Functional Ecology; and Duarte et al., 1995
-% ``Comparative Functional Plant Ecology: Rationale and Potentials.'' Trends in Ecology \& Evolution).
+\textcolor{blue}{Line 37.
+I would avoid using the term ``new trait-focused ecology.'' While it's true that a more ``explicit'' approach to functional ecology has emerged over the past few decades, the foundation of functional analyses is rooted in the field of ecology itself.
+For example, as you mentioned, Darwin's finches represent an early functional analysis centered on beak traits.
+Although the specific term ``functional traits'' was less commonly used in the past, broader functional analyses were not uncommon, particularly in plant ecology (see, for example, Calow, P 1987 ``Towards a Definition of Functional Ecology.'' Functional Ecology; and Duarte et al., 1995 ``Comparative Functional Plant Ecology: Rationale and Potentials.'' Trends in Ecology \& Evolution).}
 %  This is a really good point
@@
 
-% Line 73-74
-% Please clarify how this is related to the previous question ``what'' and ``how''.
+\textcolor{blue}{Line 73-74. Please clarify how this is related to the previous question ``what'' and ``how''.}
 %  We've expanded the sentence to:
@@
 
-% Line 93-96
-% Please explain why the chosen metric is not appropriate for the proposed example.
+\textcolor{blue}{Line 93-96. Please explain why the chosen metric is not appropriate for the proposed example.}
 %  We've expanded the sentence to:
@@
 
-% Line 100
-% I think it would be beneficial to more clearly address your research question here
-% Is your goal simply to describe the patterns of functional diversity across different processes, or are you aiming to test how well different metrics capture the effects of various mechanisms?
-If it's the latter, how did you test this, and what are your expectations in relation to the null model?
+\textcolor{blue}{Line 100.
+I think it would be beneficial to more clearly address your research question here.
+Is your goal simply to describe the patterns of functional diversity across different processes, or are you aiming to test how well different metrics capture the effects of various mechanisms?
+If it's the latter, how did you test this, and what are your expectations in relation to the null model?}
 %  We have now clarified what our goal and expectations are in the last paragraph of the introduction:
@@
 
-% Line 102
-% Please refer to Table 1 together with figure 1, to provide more details on the ecological mechanisms.
+\textcolor{blue}{Line 102. Please refer to Table 1 together with figure 1, to provide more details on the ecological mechanisms.}
 %  We now refer to both the table and the figure.
 
-% Line 103-106
-% Please provide more information on why you chose to use a combination of empirical and simulated data
-% From my understanding, your intention was to apply your framework within a more realistic context
-% If that is the case, please clarify this point to strengthen your rationale.
+\textcolor{blue}{Line 103-106.Please provide more information on why you chose to use a combination of empirical and simulated data.
+From my understanding, your intention was to apply your framework within a more realistic context.
+If that is the case, please clarify this point to strengthen your rationale.}
 %  We changed the sentence to:
@@
 
-% Line 106-111
-% This content could be more suitable at the start of the discussion section
-% Here you could maybe focus on outlining your hypotheses and predictions rather than presenting your conclusions
-% By doing so, you establish a clearer context for interpreting your findings and guide the reader through the rationale behind your study.
+\textcolor{blue}{Line 106-111.
+This content could be more suitable at the start of the discussion section.
+Here you could maybe focus on outlining your hypotheses and predictions rather than presenting your conclusions.
+By doing so, you establish a clearer context for interpreting your findings and guide the reader through the rationale behind your study.}
 %  We have now moved this sentence to the end of the conclusion (Lines @@@) to end the introduction on a clear description of our goals and expectations:
@@
 % \textit{Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \textbackslash ref\{Fig:simulations\} and \textbackslash ref\{Tab:mechanisms\}).}
-%  \textit{Our goal is to the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
+%  \textit{Our goal is to test the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
 %  \textit{We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
@@
 
-% \textbf{Methods}.
-%  Line 115
-% Please provide more details of the elements of your model, as not all readers may be familiar with the Brownian Motion trait model.
+\textcolor{blue}{\textbf{Methods}.}
+
+\textcolor{blue}{Line 115. Please provide more details of the elements of your model, as not all readers may be familiar with the Brownian Motion trait model.}
 %  We clarified the sentence as follows:
@@
 
-% Line 131
-% Please explain what the distance D is.
+\textcolor{blue}{Line 131. Please explain what the distance D is.}
 %  We clarified this point as follows:
@@
 
-% Line 139-144
-% I suggest breaking this section into two sentences and improve readability
-% The current structure is challenging to follow.
+\textcolor{blue}{Line 139-144. I suggest breaking this section into two sentences and improve readability.
+The current structure is challenging to follow.}
 %  We split this sentence in three for clarity:
@@
 
-% Table 1
-% Please provide references for the Equalizing and Facilitation mechanisms to offer clearer context for the focal ecological mechanisms studied.
+\textcolor{blue}{Table 1. Please provide references for the Equalizing and Facilitation mechanisms to offer clearer context for the focal ecological mechanisms studied.}
 %  We now added references to \textbackslash citep\{chesson2000mechanisms,barot2004mechanisms\} for the equalising mechanism and to \textbackslash citep\{bruno2003inclusion, danet2024species\} for the facilitation one.
@@
 
-% Line 183
-% Please provide the total number of the different metrics you used and refer to the table below (table 2).
+\textcolor{blue}{Line 183. Please provide the total number of the different metrics you used and refer to the table below (table 2).}
 %  We've now added a reference to table 2 and summarised the total amount of metrics.
@@
 
-% Figure 1
-% I believe adding a color legend to the figure would enhance its clarity.
+\textcolor{blue}{Figure 1. I believe adding a color legend to the figure would enhance its clarity.}
 %  We have now reworked figure 1 to be more didactic (especially for the results interpretation)
@@
 
-% Figure 1
-% Please include information in the figure caption explaining what the two types of graphs represent.
+\textcolor{blue}{Figure 1. Please include information in the figure caption explaining what the two types of graphs represent.}
 %  We added the following clarifications to the figure caption:
@@
 
-% Line 194-195
-% Please improve clarity.
+\textcolor{blue}{Line 194-195. Please improve clarity.}
 %  @@@TODO: rewrite following other reviewers comments (stats for results)
 
-% Line 201-203
-% Please simplify and improve clarity.
+\textcolor{blue}{Line 201-203. Please simplify and improve clarity.}
 %  @@@TODO: rewrite following other reviewers comments (stats for results)
 
-% Line 207-211
-% Please provide a clearer explanation of the effect size and statistical significance of these model outcomes, similar to the explanation provided in lines 198-201.
+\textcolor{blue}{Line 207-211. Please provide a clearer explanation of the effect size and statistical significance of these model outcomes, similar to the explanation provided in lines 198-201.}
 %  @@@TODO: rewrite following other reviewers comments (stats for results)
@@
 
-% Line 221
-% Please specify the added species.
+\textcolor{blue}{Line 221. Please specify the added species.}
 %  @@@TODO: name of the two added marine species
 
-% Table 2
-% The aspect for the third row is missing.
+\textcolor{blue}{Table 2. The aspect for the third row is missing.}
 %  We are not sure about what the reviewer meant here
@@
 
-% \textbf{Results}.
-%  Line 248-251
-% Please provide more detail regarding these results (e.g., what the scaled metric scores are, the effect size, and the significance of data reduction caused by the stressor)
-% Additionally, refer to the relevant section of Figure 2 (to explain how the observed trends in the figure support this assertion), and to the tables in supplementary material.
+\textcolor{blue}{\textbf{Results}.}
+
+\textcolor{blue}{Line 248-251.
+Please provide more detail regarding these results (e.g., what the scaled metric scores are, the effect size, and the significance of data reduction caused by the stressor).
+Additionally, refer to the relevant section of Figure 2 (to explain how the observed trends in the figure support this assertion), and to the tables in supplementary material.}
 %  @@@TODO: add stats for results (following other reviewers comments)
 
-% Line 252
-% I would avoid the term ``naively''.
+\textcolor{blue}{Line 252. I would avoid the term ``naively''.}
 %  We removed this term.
 
-% Line 254
-% Please ensure that the use of italics follows the guidelines from the journal.
+\textcolor{blue}{Line 254. Please ensure that the use of italics follows the guidelines from the journal.}
 %  We did not find any specific guideline regarding emphasising words with italics but removed it nonetheless.
 
-% Line 254-256
-% Apply the same changes mentioned in the comment below to lines 248-251.
+\textcolor{blue}{Line 254-256. Apply the same changes mentioned in the comment below to lines 248-251.}
 %  @@@TODO: add stats for results (following other reviewers comments)
 
-% Line 260
-% Please avoid using the exclamation mark.
+\textcolor{blue}{Line 260. Please avoid using the exclamation mark.}
 %  We changed the punctuation to a dot.
 
-% Line 262-269
-% Please provide more detail regarding these results
-% Additionally, refer to the table in supplementary material.
+\textcolor{blue}{Line 262-269. Please provide more detail regarding these results. Additionally, refer to the table in supplementary material.}
 %  @@@TODO: add stats for results (following other reviewers comments)
@@
 
-% Line 275-283
-% This should be in the discussion section.
+\textcolor{blue}{Line 275-283. This should be in the discussion section.}
 %  We moved this paragraph to the discussion on lines @@@.
 
-% Figure 2
-% Maybe I misunderstood, but I think that you are missing the adjusted R square from the figure.
+\textcolor{blue}{Figure 2. Maybe I misunderstood, but I think that you are missing the adjusted R square from the figure.}
+
 %  We removed mention of the R2 (from a previous cluttered version of the figure) and now refer to it in the supplementaries in the figure caption.
 
-% Figure 3
-% Please provide more details in the figure caption.
-%  Please include the results from the t-test in the supplementary material
-% While the significance of the test is demonstrated through the gray plots in Figures 2 and 3, the relative values are also necessary.
+\textcolor{blue}{Figure 3. Please provide more details in the figure caption. Please include the results from the t-test in the supplementary material.
+While the significance of the test is demonstrated through the gray plots in Figures 2 and 3, the relative values are also necessary.}
 %  The tables in the supplementary materials now contain the R2 values for both the Figures 2 and 3
@@
 
-% \textbf{Discussion}.
-%  Line 285-291
-% This section effectively summarizes your main goal and key message
-% However, a more detailed interpretation of your main findings is needed
-% I believe that simply stating that ``different metrics capture different patterns (what), leading to inferences about different processes (how) and mechanisms (why)'' is insufficient
-% I recommend focusing more on the performance of the various metrics to offer readers a clearer framework for selecting the appropriate metric for each mechanism
-% Please refer to items 3 and 4 for further details.
-%  Line 291-302
-% I appreciate the detailed discussion of the potential limitations of your methodology
-% However, in my opinion, this should follow a more in-depth interpretation of your results and their relation to the ``what'', ``how'', and ``why'' questions
-% Consider placing the discussion of caveats and limitations in a separate paragraph, where you address the constraints of your framework.
-%  Line 303-315
-% Based on your preference, I recommend moving this section further into the discussion
-% This will allow you to address the importance of selecting the appropriate metric in relation to the number of dimensions used for the trait space.
+\textcolor{blue}{\textbf{Discussion}.}
+\textcolor{blue}{Line 285-291. This section effectively summarizes your main goal and key message.
+However, a more detailed interpretation of your main findings is needed.
+I believe that simply stating that ``different metrics capture different patterns (what), leading to inferences about different processes (how) and mechanisms (why)'' is insufficient.
+I recommend focusing more on the performance of the various metrics to offer readers a clearer framework for selecting the appropriate metric for each mechanism.
+Please refer to items 3 and 4 for further details.}
+
+\textcolor{blue}{Line 291-302.
+I appreciate the detailed discussion of the potential limitations of your methodology.
+However, in my opinion, this should follow a more in-depth interpretation of your results and their relation to the ``what'', ``how'', and ``why'' questions.
+Consider placing the discussion of caveats and limitations in a separate paragraph, where you address the constraints of your framework.}
+
+\textcolor{blue}{Line 303-315.
+Based on your preference, I recommend moving this section further into the discussion.
+This will allow you to address the importance of selecting the appropriate metric in relation to the number of dimensions used for the trait space.}
 % We have now re-orginased the discussion and clearly separated each subsection
-% The first one now discusses the results (simulated and empirical), the second one focuses on the caveats of our approach
+% The first one now discusses the results (simulated and empirical), the second one focuses on the caveats of our approach.
 % And the third section is now a standalone box (box 2) discussing how we believe the what, how and why questions can help workers choose their functional diversity metric of interest based on their own data and question
@@
 
-% Line 313-316
-% Please rephrase to improve readability.
+\textcolor{blue}{Line 313-316. Please rephrase to improve readability.}
 %  We have now removed this sentence.
 
-% Line 316-337
-% I think this section should be simplified
-% The discussion focuses too heavily on the limitations of the approach, making it overly methodological
-% A greater emphasis on the ecological implications of the results would better convey your main message
-% I recommend streamlining this section and, if preferred, moving part of it to the supplementary material, where you can address the implications of scaling your results with the null model.
+\textcolor{blue}{Line 316-337. I think this section should be simplified.
+The discussion focuses too heavily on the limitations of the approach, making it overly methodological.
+A greater emphasis on the ecological implications of the results would better convey your main message.
+I recommend streamlining this section and, if preferred, moving part of it to the supplementary material, where you can address the implications of scaling your results with the null model.}
 %  We have now moved this section into the named ``caveats'' section
@@
 
-% Line 321-324
-% When you refer to your findings, please provide a more explicit reference to the values you found and the relative graph to better guide the reader through your results and your interpretation.
+\textcolor{blue}{Line 321-324.
+When you refer to your findings, please provide a more explicit reference to the values you found and the relative graph to better guide the reader through your results and your interpretation.}
 %  We have now specified which part of figure 2 we were using to illustrate this point:
@@
 
-% Line 327
-% Please replace `workers' with `ecologists,' as you are referring to an example from an ecological study.
+\textcolor{blue}{Line 327. Please replace `workers' with `ecologists,' as you are referring to an example from an ecological study.}
 %  We changed ``workers'' to ``researchers'' because not only ecologists are likely to work on such data/questions (e.g.~evolutionary biologists, palaeobiologists, etc.)
 
-% Line 339-340
-% Please simplify and increase readability.
+\textcolor{blue}{Line 339-340. Please simplify and increase readability.}
 %  We reworded this sentence as follows:
@@
 
-% Line 340-361
-% I believe this section should be simplified and partially reframed
-% While it's clear that you applied your framework to a real study case using empirical data, this section currently diverges too much from the main focus of your study
-% Since your primary goal is to test functional diversity metrics, the main discussion should center on the performance of these metrics, their differences from the simulated data, and their relevance to the ``why'', ``what'', and ``how'' framework
-% After addressing these points, you can briefly discuss the extinction dynamics of avian species in Hawaii.
+\textcolor{blue}{Line 340-361. I believe this section should be simplified and partially reframed.
+While it's clear that you applied your framework to a real study case using empirical data, this section currently diverges too much from the main focus of your study.
+Since your primary goal is to test functional diversity metrics, the main discussion should center on the performance of these metrics, their differences from the simulated data, and their relevance to the ``why'', ``what'', and ``how'' framework.
+After addressing these points, you can briefly discuss the extinction dynamics of avian species in Hawaii.}
 %  We have now expanded this section to be clearer:
@@
 
-% \textbf{Conclusion}.
-%  Similarly to my comments on the discussion, I suggest strengthening the connection between your conclusion and the main ``what, how, and why'' framework
-% This would enhance the overall flow of the paper and help the reader more clearly connect the main question and introduction to the final conclusions and discussion.
+\textcolor{blue}{\textbf{Conclusion}.}
+
+\textcolor{blue}{Similarly to my comments on the discussion, I suggest strengthening the connection between your conclusion and the main ``what, how, and why'' framewor.
+This would enhance the overall flow of the paper and help the reader more clearly connect the main question and introduction to the final conclusions and discussion.}
 %  We have now added the box 2 that would hopefully help with strengthening the connection between our results and the need to set up the ``what how and why'' framework for functional diversity analyses.
 
-% \textbf{Supplementary material}.
-%  Like for figure 2 in the main manuscript, I think that you are missing the adjusted R square from the figure.
+\textcolor{blue}{\textbf{Supplementary material}.}
+
+\textcolor{blue}{Like for figure 2 in the main manuscript, I think that you are missing the adjusted R square from the figure.}
 %  We have now specified in the caption that the adjusted R squares are available in the supplementary tables to not clutter the figure.
 
-% I enjoyed reviewing and with some time in revisions, this paper and the messaging will be stronger and more applied
-% I appreciate the framework and application this paper will support.
+\textcolor{blue}{I enjoyed reviewing and with some time in revisions, this paper and the messaging will be stronger and more applied.
+I appreciate the framework and application this paper will support.}
 

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
Update ecography_reponse_letter.tex

@@
 
-% \textbf{Reviewer: 2}
+\textcolor{blue}{\textbf{Reviewer: 2}}
 
-% Upon reading the title and introduction, I was intrigued and had high expectations for the proposed ``what, how, why'' framework of trait-based analysis
-% However, after thoroughly reviewing the manuscript, it became clear that the study does not fully deliver on its initial promise of providing a guideline for applying this framework to ecological data analysis
-% Instead, it focuses more on evaluating the varying responses of different functional diversity indices when specific data points are selectively removed from a dataset
-% The primary outcome is that different functional diversity indices exhibit distinct and often unpredictable responses to data removal
-% However, this observation, while valid, does not seem to offer significant new insights into ecological processes
-% In essence, the study highlights the dependence of analytical outcomes on the choice of index---a finding that, while true, feels somewhat self-evident.
+\textcolor{blue}{Upon reading the title and introduction, I was intrigued and had high expectations for the proposed ``what, how, why'' framework of trait-based analysis.
+However, after thoroughly reviewing the manuscript, it became clear that the study does not fully deliver on its initial promise of providing a guideline for applying this framework to ecological data analysis.
+Instead, it focuses more on evaluating the varying responses of different functional diversity indices when specific data points are selectively removed from a dataset.
+The primary outcome is that different functional diversity indices exhibit distinct and often unpredictable responses to data removal.
+However, this observation, while valid, does not seem to offer significant new insights into ecological processes.
+In essence, the study highlights the dependence of analytical outcomes on the choice of index - a finding that, while true, feels somewhat self-evident.}
 
-% It is possible that these concerns arise due to insufficient descriptions of the methodology and inconsistencies in terminology (specific points are detailed below)
-% As the manuscript stands, I am unable to recommend it for publication
-% That said, I have outlined several constructive suggestions that I hope will assist the authors in revising and refining the paper.
+\textcolor{blue}{It is possible that these concerns arise due to insufficient descriptions of the methodology and inconsistencies in terminology (specific points are detailed below).
+As the manuscript stands, I am unable to recommend it for publication.
+That said, I have outlined several constructive suggestions that I hope will assist the authors in revising and refining the paper.}
 
-% \textbf{General comments}
+\textcolor{blue}{\textbf{General comments}}
 
-% \textbf{Main model idea}
-%  The paper falls short in demonstrating how the ``what, how, why'' framework can be applied to ecological data
-% This issue may stem from examining two distinct layers: the simple removal of points from datasets and the subsequent interpretation of these removals in terms of ecological processes, which are not explicitly simulated in the model
-% The distinction between these layers is confusing and is primarily addressed in Table 1, which does not effectively clarify these points
-% To improve clarity, I recommend a more systematic presentation of these issues in the main text: This would involve, describing the relevant ecological processes (e.g., facilitation, filtering, competition), explaining how these processes influence species-associated trait distributions, and then outlining how these influences are represented in the five trait-removal processes.
+\textcolor{blue}{\textbf{Main model idea}}
+
+\textcolor{blue}{The paper falls short in demonstrating how the ``what, how, why'' framework can be applied to ecological data.
+This issue may stem from examining two distinct layers: the simple removal of points from datasets and the subsequent interpretation of these removals in terms of ecological processes, which are not explicitly simulated in the model.
+The distinction between these layers is confusing and is primarily addressed in Table 1, which does not effectively clarify these points.
+To improve clarity, I recommend a more systematic presentation of these issues in the main text: This would involve, describing the relevant ecological processes (e.g., facilitation, filtering, competition), explaining how these processes influence species-associated trait distributions, and then outlining how these influences are represented in the five trait-removal processes.}
 
@@
 
-% \textbf{Model description:}
-%  The section on simulating trait-space patterns needs to be significantly expanded to include more technical details regarding how the simulations were conducted
-% Key information, such as the model parameters, is entirely absent
-% A more thorough explanation of the model setup, parameters, and procedures is required to ensure that the simulations can be reproduced without needing to consult external publications (such as those by Guillerme and Mammola)
-% To address this, I suggest adding a detailed explanation of the model in either the main Methods section or in a supplementary appendix, to ensure that all necessary parameters and computational steps are clearly specified.
+\textcolor{blue}{\textbf{Model description:}}
+The section on simulating trait-space patterns needs to be significantly expanded to include more technical details regarding how the simulations were conducted.
+Key information, such as the model parameters, is entirely absent.
+A more thorough explanation of the model setup, parameters, and procedures is required to ensure that the simulations can be reproduced without needing to consult external publications (such as those by Guillerme and Mammola).
+To address this, I suggest adding a detailed explanation of the model in either the main Methods section or in a supplementary appendix, to ensure that all necessary parameters and computational steps are clearly specified.}
 
@@
 
-% \textbf{Consistent terminology:}
-%  The paper's clarity is hindered by inconsistent, imprecise, and shifting terminology, making it unnecessarily difficult to follow
-% For example, the entities generated by the model are referred to interchangeably as ``elements,'' ``traits,'' ``tips,'' or ``species,'' with no clear rationale for the variation
-% Similar, for many other terms that are often used inconsistently or fuzzy (see detailed comments below)
-% This lack of consistency creates confusion and often forced me to repeatedly revisit sections of the text to discern the intended meaning
-% To improve readability, I recommend establishing a consistent terminology early in the paper and adhering to it throughout
-% This would significantly reduce ambiguity and enhance the reader's ability to follow the methodology and conclusions.
+\textcolor{blue}{\textbf{Consistent terminology:}}
+
+\textcolor{blue}{The paper's clarity is hindered by inconsistent, imprecise, and shifting terminology, making it unnecessarily difficult to follow.
+For example, the entities generated by the model are referred to interchangeably as ``elements,'' ``traits,'' ``tips,'' or ``species,'' with no clear rationale for the variation.
+Similar, for many other terms that are often used inconsistently or fuzzy (see detailed comments below).
+This lack of consistency creates confusion and often forced me to repeatedly revisit sections of the text to discern the intended meaning.
+To improve readability, I recommend establishing a consistent terminology early in the paper and adhering to it throughout.
+This would significantly reduce ambiguity and enhance the reader's ability to follow the methodology and conclusions.}
+
 % We have now standardised the terminology following \textbackslash cite\{guillerme2020shifting\} and \textbackslash cite\{mammola2021concepts\}
@@
 
-% \textbf{Detailed comments:}
+\textcolor{blue}{\textbf{Detailed comments:}}
 
-% L114: Really ``A single Brownian motion trait''? This implies that only a single trait dimension is generated, whereas from the later text it seems that higher-dimensional traits are simulated.
+\textcolor{blue}{L114: Really ``A single Brownian motion trait''?
+This implies that only a single trait dimension is generated, whereas from the later text it seems that higher-dimensional traits are simulated.}
 %  By ``trait'' here we refer to the single process (Brownian Motion) which can be simulated in any number of dimensions (i.e.~generating a time dependent element value using 2, 4 or 8 dimension coordinates to describe it)
@@
 
-% L115: what is a ``tip''?
+\textcolor{blue}{L115: what is a ``tip''?}
 %  By tip we meant a node with no descendants in an evolutionary tree
@@
 
-% L115, ``200 elements'': Why are so few data points used?
-A dataset of only 200 elements (reduced to 40 after data removal) seems exceptionally sparse, especially in a 4- or 8-dimensional space, as noted later in the text
-% What is the motivation behind studying such a sparse space?
-One of the main advantages of using simple models like a branching process is that they allow for the generation of larger datasets, which would provide more robust statistics and better analytical power.
+\textcolor{blue}{L115, ``200 elements'': Why are so few data points used?
+A dataset of only 200 elements (reduced to 40 after data removal) seems exceptionally sparse, especially in a 4- or 8-dimensional space, as noted later in the text.
+What is the motivation behind studying such a sparse space?
+One of the main advantages of using simple models like a branching process is that they allow for the generation of larger datasets, which would provide more robust statistics and better analytical power.}
 %  @@@TODO: Rerun the simulations with 400, 800 and 1000 elements.
@@
 % We can also link that to the use of FD in palaeontology (disparity) where 40 is also already quite big!
+% Tom's comment: *Reviewer 2's point about why we use only 200 data points - it's a while since I read our paper, but I guess we did that to make it more realistic in terms of the sample sizes people use.
+In island studies, 200 is pretty large, and I have many references (including my own ...) where people calculate FD using 1-5 species.
+So again we could make the argument that it represents a trade-off, between realistic sample sizes that people analyse vs large enough that we can really test the mechanisms
 
-% LL 120-124, ``four different stressors\ldots{} five specific algorithms'': Inconsistency in the used terminology
-% In this section, ``stressors'' appear to refer to the amount of data being removed (ranging from 20\% to 80\%), but later, the term ``stressors'' is used to describe the five data-removal processes---referred to as ``specific algorithms'' here.
+\textcolor{blue}{LL 120-124, ``four different stressors, five specific algorithms'': Inconsistency in the used terminology.
+In this section, ``stressors'' appear to refer to the amount of data being removed (ranging from 20\% to 80\%), but later, the term ``stressors'' is used to describe the five data-removal processes---referred to as ``specific algorithms'' here.}
 % We have now standardised the terminology throughout the manuscript (and included it in the glossary table in the supplementary materials): ``algorithm'' now designate the code algorithm, ``stressors'' (and ``stressors' intensity'') now designates how we modify the data (using the algorithms mentioned above) and ``mechanisms'' now designates the biological mechanisms we are approximating using a specific stressor (that is based on a specific algorithm)
@@
 
-% L130, ``increasing density'': What is the motivation for this term?
-While I understand that this removal scheme may result in an overall increase in average density, the removal itself is not directly based on density but rather on pairwise distance in trait space
-% For instance, the equalizing scheme would also lead to an increase in density.
+\textcolor{blue}{L130, ``increasing density'': What is the motivation for this term?
+While I understand that this removal scheme may result in an overall increase in average density, the removal itself is not directly based on density but rather on pairwise distance in trait space.
+For instance, the equalizing scheme would also lead to an increase in density.}
 %  We used the term ``increasing density'' here to differentiate from ``decreasing size''
@@
 
-% L134, ``shifting space'': Similar to the previous point, this term is somewhat misleading
-% This removal scheme does not explicitly shift the distribution; rather, it cuts the distribution at one edge, causing the mean to shift
-% However, this effect of shifting the mean is not unique to this scheme---other removal schemes will generally result in a shift of the mean as well.
+\textcolor{blue}{L134, ``shifting space'': Similar to the previous point, this term is somewhat misleading.
+This removal scheme does not explicitly shift the distribution; rather, it cuts the distribution at one edge, causing the mean to shift.
+However, this effect of shifting the mean is not unique to this scheme - other removal schemes will generally result in a shift of the mean as well.}
 %  This is also a fair point
@@
 
-% L135, ``with the maximal value'': What does ``maximal value'' mean for points in a higher-dimensional space?
-Maybe this refer to the location in trait space with the highest density of trait elements?
+\textcolor{blue}{L135, ``with the maximal value'': What does ``maximal value'' mean for points in a higher-dimensional space?
+Maybe this refer to the location in trait space with the highest density of trait elements?}
 %  We have now specified what we meant by maximal value:
@@
 
-% LL138- 145: I found this section on the competition method difficult to follow:
-%  - The term ``bandwidth'' is unclear
-% What does ``bandwidth'' refer to in this context?
+\textcolor{blue}{LL138- 145: I found this section on the competition method difficult to follow:}
+
+\textcolor{blue}{- The term ``bandwidth'' is unclear.
+What does ``bandwidth'' refer to in this context?
 Does it imply that resampling was done using a frequency-based method, or is it referring to a specific range or interval of trait values in the distribution?
-This ambiguity ties back to my earlier request for more technical details regarding the simulation process.
-%  - I am also unclear on the meaning of ``n''
-% Could you clarify what ``n'' represents here?
+This ambiguity ties back to my earlier request for more technical details regarding the simulation process.}
+
+\textcolor{blue}{- I am also unclear on the meaning of ``n''.
+Could you clarify what ``n'' represents here?
 For example, what is meant by the ``probability of sampling of n'' and the reference to ``high/low values of n''?
-Without this information, it is difficult to grasp how the competition process is functioning.
-%  - The term ``flattening the curve'' needs more context
-% What curve is being flattened, and how does this relate to the competition method?
+Without this information, it is difficult to grasp how the competition process is functioning.}
+
+\textcolor{blue}{- The term ``flattening the curve'' needs more context.
+What curve is being flattened, and how does this relate to the competition method?}
+
 %  We have now completely rewritten the description of this stressor by defining more clearly what the bandwidth is, renaming the variables to be easier to understand, removed mention of flattening the curve and rewrote the equations to be specific to the increase in evenness (previous equations were generalised to both increase or decrease the eveness depending on the variable \textit{i} (formerly \textit{n}) being now always a proportion of removal (e.g.~0.2, 0.4, \ldots{} - former description worked with \textit{n} being also a proportion to add e.g.~1.2, 1.4, etc\ldots):
@@
 
-% Figure 1: The caption is confusing and needs significant improvement
-% Below are some specific points that require attention:
-%  - The caption completely omits any description of what is shown in left and right columns (i.e., 1d and 2d case).
+\textcolor{blue}{Figure 1: The caption is confusing and needs significant improvement.
+Below are some specific points that require attention:}
+
+\textcolor{blue}{- The caption completely omits any description of what is shown in left and right columns (i.e., 1d and 2d case).}
 %  We have now clarified what each column contains in the figure caption.
-%  - It is unclear whether this figure (both the 1d and 2d cases) is based on the actual simulations or is merely a schematic sketch
-% For instance, in the right column, it seems unlikely that the figure actually depicts 200 points.
-%  The figures do depict 200 points (200 blue in row A and 100 blue + 100 orange in the other columns)
+\textcolor{blue}{- It is unclear whether this figure (both the 1d and 2d cases) is based on the actual simulations or is merely a schematic sketch.
+For instance, in the right column, it seems unlikely that the figure actually depicts 200 points.}
+
+% The figures does depict 200 points (200 blue in row A and 100 blue + 100 orange in the other columns).
 % We have now specified this in the caption.
-%  - The caption mentions a ``trait space of 2, 4, or 8 dimensions,'' yet a 1d trait space is shown
-% If a 1d trait space is an illustrative example, this should be specified in the caption.
+\textcolor{blue}{- The caption mentions a ``trait space of 2, 4, or 8 dimensions,'' yet a 1d trait space is shown.
+If a 1d trait space is an illustrative example, this should be specified in the caption.}
 %  We have now specified that we are showing 1 or 2 dimensions regardless of the number of dimensions per traits (i.e.~8 dimensional traits are not fully represented here but what's happening on their first and second dimensions is displayed here)
@@
 % We have now clarified this in the glossary.
-%  - The pattern shown in the subfigure labeled ``2d facilitating mechanism'' appears questionable
-% If point-pairs from a 2d distribution are removed based on a threshold distance, the resulting pattern would not typically resemble what is depicted here.
+\textcolor{blue}{- The pattern shown in the subfigure labeled ``2d facilitating mechanism'' appears questionable.
+If point-pairs from a 2d distribution are removed based on a threshold distance, the resulting pattern would not typically resemble what is depicted here.}
 %  The illustration of the 2D facilitating mechanism directly comes from the output of the dispRity::reduce.space function where the algorithm is implemented
@@
 % The algorithm is recursive so that once a pair of observations is selected, it cannot be selected again.
-%  - Terminology: The use of the term ``species'' in the caption is confusing, as the methods seem to generate traits rather than species.
+\textcolor{blue}{- Terminology: The use of the term ``species'' in the caption is confusing, as the methods seem to generate traits rather than species.}
 %  We have now removed any mention of species.
-%  - The phrase ``removing half of the species, resulting in two groups of tips'' is unclear
-% How are these two groups defined? Presumably, one group consists of the removed traits and the other of non-removed traits
-% Additionally, it is unclear which color (blue or orange) corresponds to which group.
+\textcolor{blue}{- The phrase ``removing half of the species, resulting in two groups of tips'' is unclear.
+How are these two groups defined? Presumably, one group consists of the removed traits and the other of non-removed traits.
+Additionally, it is unclear which color (blue or orange) corresponds to which group.}
 %  We have now specified this in the new caption.
-%  - I am skeptical about the use of ecological terminology such as ``competition'' and ``facilitation'' in this context
-% What is actually being shown are data distributions resulting from different point-removal algorithms, not ecological interactions per se
-% While these processes can be discussed and associated with ecological mechanisms in the text or discussion (if well justified), using such terms directly in the figure caption is highly misleading.
+\textcolor{blue}{- I am skeptical about the use of ecological terminology such as ``competition'' and ``facilitation'' in this context.
+What is actually being shown are data distributions resulting from different point-removal algorithms, not ecological interactions per se.
+While these processes can be discussed and associated with ecological mechanisms in the text or discussion (if well justified), using such terms directly in the figure caption is highly misleading.}
 %  @@@TODO: Links back to main comment from reviewer 2: Go back to the previous version of table 1 that gives more description of the models
@@
 
-% LL 150-152: The description is unclear and appears to use the term ``simulation'' inconsistently
-% In the phrase ``replicates per simulations,'' the term ``simulation'' seems to have a different meaning than in ``smallest number of simulations.'' Additionally, the first part of the sentence suggests that variance decreases with an increasing number of simulations, whereas the second part implies that adding more simulations leads to increased variance.
+\textcolor{blue}{LL 150-152: The description is unclear and appears to use the term ``simulation'' inconsistently.
+In the phrase ``replicates per simulations,'' the term ``simulation'' seems to have a different meaning than in ``smallest number of simulations.'' Additionally, the first part of the sentence suggests that variance decreases with an increasing number of simulations, whereas the second part implies that adding more simulations leads to increased variance.}
 %  There was a confusion indeed
@@
 
-% L 166: what does ``uncorrelated'' mean here?
-Does this mean that the branching process was independently repeated four times, once for each trait dimension, or does it mean that it generates four trait dimensions simultaneously, with these dimensions being uncorrelated?
+\textcolor{blue}{L 166: what does ``uncorrelated'' mean here?
+Does this mean that the branching process was independently repeated four times, once for each trait dimension, or does it mean that it generates four trait dimensions simultaneously, with these dimensions being uncorrelated?}
 %  We meant that the dimensions for each multidimensional traits are uncorrelated (i.e.~the trait is a uncorrelated Brownian motion process in n dimensions)
@@
 
-% LL 170-178: why then did the authors not simply generate more data points (see my comment above)?
+\textcolor{blue}{LL 170-178: why then did the authors not simply generate more data points (see my comment above)?}
 %  @@@TODO: Rerun the simulations with 400, 800 and 1000 elements.
 
-% Table 1: The integration of this table into the main text could be improved
-% Specifically, the abrupt mention of species (which are not part of the model) is confusing.
+\textcolor{blue}{Table 1: The integration of this table into the main text could be improved.
+Specifically, the abrupt mention of species (which are not part of the model) is confusing.}
 %  We have now improved the context of the tables with predictions and removed most of the mentions of ``species'' now replaced by the standardised term of ``observations'' (unless they were specific examples).
 
-% L 178 and several instances later ``o'' should read ``0'' (zero)
+\textcolor{blue}{L 178 and several instances later ``o'' should read ``0'' (zero).}
 %  These integers were compiled using the latex mathpazo font
@@
 
-% L184, I suggest to mention here, that these functional metrics are denoted as trait space occupancy later on.
+\textcolor{blue}{L184, I suggest to mention here, that these functional metrics are denoted as trait space occupancy later on.}
 %  Great suggestions
@@
 
-% L 208, ``\textasciitilde{}'' symbol: This usage of mathematical notation (instead of plain words) in a sentence (here, but also later in the text) is difficult to understand.
+\textcolor{blue}{L 208, ``\textasciitilde{}'' symbol: This usage of mathematical notation (instead of plain words) in a sentence (here, but also later in the text) is difficult to understand.}
 %  We removed tilde symbols throughout the manuscript.
 
-% Table 2: Column two (``Methods'') does not provide any information beyond what is explained in ``Description'' and can be removed.
+\textcolor{blue}{Table 2: Column two (``Methods'') does not provide any information beyond what is explained in ``Description'' and can be removed.}
 %  We decided to keep the ``methods'' column to make it easier for the reader to refer to the terms used in Mammola et al.~2021 at a glance.
 
-% LL 252-260 ``opposite pattern to what was expected'' and ``naïve expectation'': I do not agree
-% It is well known that removing data points can increase the evenness of a distribution, so it is not surprising that removal procedures can result in a reduction of the metric scores
-% The term ``space occupancy'' could be contributing to this misconception.
+\textcolor{blue}{LL 252-260 ``opposite pattern to what was expected'' and ``naïve expectation'': I do not agree.
+It is well known that removing data points can increase the evenness of a distribution, so it is not surprising that removal procedures can result in a reduction of the metric scores.
+The term ``space occupancy'' could be contributing to this misconception.}
 %  We rewritten this section to be in line with the reviewer's comment:
@@
 
-% L 287: Why would this be ``unexpected''?
+\textcolor{blue}{L 287: Why would this be ``unexpected''?}
 %  We removed the word ``unexpected'' that was maybe ironic (and irrelevant regardless).
 
-% LL 287-289: I do not see that the ``what, how, why''-framework has really been used in this study
-% Please specify, for example, which mechanisms have correctly been explained in the analysis?
+\textcolor{blue}{LL 287-289: I do not see that the ``what, how, why''-framework has really been used in this study.
+Please specify, for example, which mechanisms have correctly been explained in the analysis?}
 %  We have now reworked this section with the inclusion of a Box explaining in more details how we believe that using the ``what, how, why'' framework helps choosing the correct metric for a specific dataset and question.

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
Update ecography_reponse_letter.tex

@@
 
-% \textbf{Reviewer: 3}
+\textcolor{blue}{\textbf{Reviewer: 3}}
 
-% The question of whether modern methods are effective in extracting functional diversity patterns and relating them to ecological drivers is fundamentally important
-% The authors conducted a consistent and accurate set of simulations to test how different methods and metrics measure functional diversity under various ecological processes, considering both simulated and real examples
-% The approach used is reasonable and appropriate
-% However, as it stands, I find it challenging to identify the specific hypotheses that the study aims to address, particularly regarding how different metrics or methodologies should approximate the different mechanisms tested
-% Additionally, I found the discussion of the results to be too broad and lacking clear justification for the findings
-% Consequently, the manuscript, as it reads, does not fully convey the novelty and the scientific gap filled by this work
-% Therefore, I believe that a revision would be beneficial to improve the clarity of the manuscript's message.
+\textcolor{blue}{The question of whether modern methods are effective in extracting functional diversity patterns and relating them to ecological drivers is fundamentally important.
+The authors conducted a consistent and accurate set of simulations to test how different methods and metrics measure functional diversity under various ecological processes, considering both simulated and real examples.
+The approach used is reasonable and appropriate.
+However, as it stands, I find it challenging to identify the specific hypotheses that the study aims to address, particularly regarding how different metrics or methodologies should approximate the different mechanisms tested.
+Additionally, I found the discussion of the results to be too broad and lacking clear justification for the findings.
+Consequently, the manuscript, as it reads, does not fully convey the novelty and the scientific gap filled by this work.
+Therefore, I believe that a revision would be beneficial to improve the clarity of the manuscript's message.}
 
-% \textbf{Specific comments:}
+\textcolor{blue}{\textbf{Specific comments:}}
 
-% I think that, while the idea of the ``what?'', ``how?'', and ``why?'' questions is intriguing, the Introduction does not sufficiently articulate the specific aims addressed in the manuscript
-% Specifically, through simulations, the authors approximate different ecological mechanisms that produce patterns in the functional space (as partially shown in Figure 1)
-% These patterns and mechanisms are addressed in section 3.1.1 of the Methods, yet it is not clear how the different functional diversity metrics (i.e., richness, divergence, and evenness) should behave under these spatial reorganizations, particularly when compared to a null mechanism
-% For instance, under the Equalizing mechanism, I would hypothesize that functional richness and divergence would decrease compared to the Null mechanism, while species distribution would become more even, given that species removed from the extremes of the space would pack towards the centre
-% The signal might be expected to be stronger with increasing species removal, as species packing increases
-% An accurate algorithm or method should be capable of detecting such a signal
-% However, for mechanisms such as competition, the reasoning is not as straightforward (e.g., I am not convinced that richness metrics would decrease compared to the null mechanism in this case), particularly concerning expectations for divergence and evenness, as well as the relationship of these metrics to increasing species loss
-% I believe that formulating clear ``hypotheses'' on species patterns in the space in relation to the various mechanisms tested may help readers better navigate through the metrics analysed
-% This part could also be enhanced by introducing a new figure or expanding Figure 1 or Table 1 (which I found very well done).
+\textcolor{blue}{I think that, while the idea of the ``what?'', ``how?'', and ``why?'' questions is intriguing, the Introduction does not sufficiently articulate the specific aims addressed in the manuscript.
+Specifically, through simulations, the authors approximate different ecological mechanisms that produce patterns in the functional space (as partially shown in Figure 1).
+These patterns and mechanisms are addressed in section 3.1.1 of the Methods, yet it is not clear how the different functional diversity metrics (i.e., richness, divergence, and evenness) should behave under these spatial reorganizations, particularly when compared to a null mechanism.
+For instance, under the Equalizing mechanism, I would hypothesize that functional richness and divergence would decrease compared to the Null mechanism, while species distribution would become more even, given that species removed from the extremes of the space would pack towards the centre.
+The signal might be expected to be stronger with increasing species removal, as species packing increases.
+An accurate algorithm or method should be capable of detecting such a signal.
+However, for mechanisms such as competition, the reasoning is not as straightforward (e.g., I am not convinced that richness metrics would decrease compared to the null mechanism in this case), particularly concerning expectations for divergence and evenness, as well as the relationship of these metrics to increasing species loss.
+I believe that formulating clear ``hypotheses'' on species patterns in the space in relation to the various mechanisms tested may help readers better navigate through the metrics analysed.
+This part could also be enhanced by introducing a new figure or expanding Figure 1 or Table 1 (which I found very well done).}
 
@@
 
-% Lines 165-170 and 303-307: As mentioned, TPD methods are implemented for a maximum of four dimensions
-% I am curious about how the metric for the eight-dimensional space was computed
-% Furthermore, although eight-dimensional spaces may be considered ``relatively small'' in morphological studies, ecological studies addressing both plants (Laughlin 2014, Díaz et al.~2016, Weigelt et al.~2021) and animals (Healy et al.~2019, Mouillot et al.~2021, Cox et al.~2022, Beccari et al.~2024) generally use far fewer than eight dimensions
-% Accordingly, I wonder if this was the reason behind selecting the four-dimensional space as the main analysis in Figure 2, or was it to allow comparison with the dimensionality of the empirical trait space?
-I believe this selection requires justification in the text.
+\textcolor{blue}{Lines 165-170 and 303-307: As mentioned, TPD methods are implemented for a maximum of four dimensions.
+I am curious about how the metric for the eight-dimensional space was computed.
+Furthermore, although eight-dimensional spaces may be considered ``relatively small'' in morphological studies, ecological studies addressing both plants (Laughlin 2014, Díaz et al.2016, Weigelt et al.2021) and animals (Healy et al.2019, Mouillot et al.2021, Cox et al.2022, Beccari et al.2024) generally use far fewer than eight dimensions.
+Accordingly, I wonder if this was the reason behind selecting the four-dimensional space as the main analysis in Figure 2, or was it to allow comparison with the dimensionality of the empirical trait space?
+I believe this selection requires justification in the text.}
 
@@
 
-% The use of bird traits is an interesting aspect that fits well with the scope of the manuscript
-% However, for the same reason mentioned in the previous comment and as stated by the authors in lines 165-181, I am puzzled by the choice of five dimensions to portray the AVONET variation
-% Morphological traits are generally highly correlated, resulting in a few axes explaining the highest proportion of trait variation and a space highly clustered in the centre with few species occupying its outer part (Pigot et al.~2020, Germain et al.~2023, Toussaint et al.~2024)
-% How is the selection of five dimensions justified for the analysis? Are these properties of the dataset considered important for the example?
+\textcolor{blue}{The use of bird traits is an interesting aspect that fits well with the scope of the manuscript.
+However, for the same reason mentioned in the previous comment and as stated by the authors in lines 165-181, I am puzzled by the choice of five dimensions to portray the AVONET variation.
+Morphological traits are generally highly correlated, resulting in a few axes explaining the highest proportion of trait variation and a space highly clustered in the centre with few species occupying its outer part (Pigot et al.2020, Germain et al.2023, Toussaint et al.2024).
+How is the selection of five dimensions justified for the analysis? Are these properties of the dataset considered important for the example?}
 %  We chose to use five dimensions to describe the Hawaiian birds trait space to make sure that each focal group has at least 95\% of their variance described in the trait space (following Guillerme et al.~2023 protocol).
@@
 
-% I found the Results section very broad and believe it would benefit from specifying further details about discrepancies and similarities found among metrics
-% Referring to my first comment, the authors mention expected patterns (like a decrease in trait space occupancy) that are not previously introduced in the text
-% Similarly, I am not convinced by the expectation that all metrics (i.e., richness, evenness, and divergence) should generally increase compared to a null mechanism
-% For instance, if the competition mechanism is removing species from the centre of the distribution (as specified in Figure 1), wouldn't it produce a decrease in functional richness but an increase in its divergence?
+\textcolor{blue}{I found the Results section very broad and believe it would benefit from specifying further details about discrepancies and similarities found among metrics.
+Referring to my first comment, the authors mention expected patterns (like a decrease in trait space occupancy) that are not previously introduced in the text.
+Similarly, I am not convinced by the expectation that all metrics (i.e., richness, evenness, and divergence) should generally increase compared to a null mechanism.
+For instance, if the competition mechanism is removing species from the centre of the distribution (as specified in Figure 1), wouldn't it produce a decrease in functional richness but an increase in its divergence?}
 
@@
 
-% Lines 332-337: The rationale behind the choice of scaling for the null mechanism is unclear to me
-% While I agree with the authors that this transformation may help in data visualization and interpretation, if scaling introduces such an artifact, why rely so heavily on these comparisons?
-Wouldn't it be more straightforward to simply compare metrics across each other without scaling?
+\textcolor{blue}{Lines 332-337: The rationale behind the choice of scaling for the null mechanism is unclear to me.
+While I agree with the authors that this transformation may help in data visualization and interpretation, if scaling introduces such an artifact, why rely so heavily on these comparisons?
+Wouldn't it be more straightforward to simply compare metrics across each other without scaling?}
 
@@
 
-% While I agree with the conclusion that the metric of choice should align with the research question or the specific pattern analyzed, I feel the Discussion is missing some mechanistic links
-% The take-home message that different metrics are suitable for different ecological processes should be paired with an explanation or at least a suggestion of which metric may be more appropriate for which process
-% For instance, the highest dispersion observed in the empirical dataset for RaoQ compared to other metrics may relate to the lower accuracy of the metric
-% Without further details on the mechanisms of the metrics analyzed, I fear the reader may lose sight of the importance of metric choice and the implications of their different behaviors in addressing ecological processes.
+\textcolor{blue}{While I agree with the conclusion that the metric of choice should align with the research question or the specific pattern analyzed, I feel the Discussion is missing some mechanistic links.
+The take-home message that different metrics are suitable for different ecological processes should be paired with an explanation or at least a suggestion of which metric may be more appropriate for which process.
+For instance, the highest dispersion observed in the empirical dataset for RaoQ compared to other metrics may relate to the lower accuracy of the metric.
+Without further details on the mechanisms of the metrics analyzed, I fear the reader may lose sight of the importance of metric choice and the implications of their different behaviors in addressing ecological processes.}
 

### 2024-11-28 Manuscript/ecography_reponse_letter.tex
updated responses to reviewer 1

@@
 \textcolor{blue}{Line 3-6: Please rephrase to simplify and avoid brackets.}
-%  We changed the sentence to:
-%  \textit{Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to various stressors
-% In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism.}
-%  Lines @@@
+
+We changed the sentence to:
+\textit{Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to various stressors
+In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism.}
+Lines 3-7.
 
 \textcolor{blue}{Line 10-11: Please rephrase to avoid brackets.}
-%  We changed the sentence to:
-%  \textit{We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.}
-%  Lines @@@
+
+We changed the sentence to:
+\textit{We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.}
+Lines 9-12.
 
@@
 The current description is too vague and does not effectively convey your main findings.}
-%  We've expanded a little on the results
-% One point of keeping the vagueness in the abstract though was to not give the impression that we provide any prescriptive results but rather a broad understanding that the choice of metric is important.
-%  \textit{We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).}
-%  \textit{For example, the competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.}
+
+We've expanded a little on the results.
+One point of keeping the vagueness in the abstract though was to not give the impression that we provide any prescriptive results but rather a broad understanding that the choice of metric is important.
+
+\textit{We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).
+For example, the competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.}
+Lines 15-18.
 
@@
 Please include a citation for this reference if it is the case.}
-%  We've added the reference as suggested.
+
+We've added the reference as suggested.
 
@@
 Although the specific term ``functional traits'' was less commonly used in the past, broader functional analyses were not uncommon, particularly in plant ecology (see, for example, Calow, P 1987 ``Towards a Definition of Functional Ecology.'' Functional Ecology; and Duarte et al., 1995 ``Comparative Functional Plant Ecology: Rationale and Potentials.'' Trends in Ecology \& Evolution).}
-%  This is a really good point
-% We've removed the mention of ``new''.
+
+This is a really good point
+We've removed the mention of ``new''.
 
 \textcolor{blue}{Line 73-74. Please clarify how this is related to the previous question ``what'' and ``how''.}
-%  We've expanded the sentence to:
-%  \textit{Mechanism description is then about linking the pattern and the process to some biological properties (sometimes referred to as ``rules'\,') and is at the core of answering the biological question at hand.}
-%  Lines @@@
+
+We've expanded the sentence to:
+\textit{Mechanism description is then about linking the pattern and the process to some biological properties (sometimes referred to as ``rules'') and is at the core of answering the biological question at hand.}
+Lines 77-79.
 
 \textcolor{blue}{Line 93-96. Please explain why the chosen metric is not appropriate for the proposed example.}
-%  We've expanded the sentence to:
-%  \textit{In such a scenario, measuring the occupied area (i.e.~\(km^{2}\)) of each population on each island (a pattern - ``what?'\,') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.}
-%  Lines @@@
+
+We've expanded the sentence to:
+\textit{In such a scenario, measuring the occupied area (i.e.
+$km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.}
+Lines 99-103.
 
@@
 If it's the latter, how did you test this, and what are your expectations in relation to the null model?}
-%  We have now clarified what our goal and expectations are in the last paragraph of the introduction:
-%  \textit{Our goal is to assess the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
-%  \textit{We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
-%  Lines @@@
+
+We have now clarified what our goal and expectations are in the last paragraph of the introduction:
+\textit{Our goal is to assess the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
+We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
+Lines 109-114.
 
 \textcolor{blue}{Line 102. Please refer to Table 1 together with figure 1, to provide more details on the ecological mechanisms.}
-%  We now refer to both the table and the figure.
+
+We now refer to both the table and the figure.
 
@@
 If that is the case, please clarify this point to strengthen your rationale.}
-%  We changed the sentence to:
-%  \textit{Beyond simulations, we also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands (process) resulting in a trait space modified by extinctions (the pattern) capturing the effects of species extinction on species trait distributions (the mechanism).}
-%  Lines @@@
+
+We changed the sentence to:
+\textit{To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data..}
+Lines 114-119.
 
@@
 By doing so, you establish a clearer context for interpreting your findings and guide the reader through the rationale behind your study.}
-%  We have now moved this sentence to the end of the conclusion (Lines @@@) to end the introduction on a clear description of our goals and expectations:
 
-% \textit{Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \textbackslash ref\{Fig:simulations\} and \textbackslash ref\{Tab:mechanisms\}).}
-%  \textit{Our goal is to test the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).}
-%  \textit{We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).}
-%  \textit{To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.}
-%  \textit{We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.}
+We have now moved this sentence to the end of the conclusion (Lines 569-572) to end the introduction on a clear description of our goals and expectations:
 
-% Lines @@@
+\textit{Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure 1 and Table 1).
+Our goal is to test the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
+We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).
+To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data.
+We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.}
+Lines 107-122.
 
@@
 \textcolor{blue}{Line 115. Please provide more details of the elements of your model, as not all readers may be familiar with the Brownian Motion trait model.}
-%  We clarified the sentence as follows:
-%  \textit{First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.~a pure birth speciation model) until reaching 200 elements in \textbackslash texttt\{R\} \textbackslash citep\{rcore\} using \textbackslash texttt\{treats\} \textbackslash citep\{guillerme2024treats\}.}
-%  Lines @@@
+
+We clarified the sentence as follows:
+\textit{First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
+a pure birth speciation model) until reaching 200 observations in \texttt{R} (R Core Team, 2024) using \texttt{treats} (Guillerme, 2024).}
+Lines 125-127.
 
 \textcolor{blue}{Line 131. Please explain what the distance D is.}
-%  We clarified this point as follows:
-%  \textit{Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least \(D\) (i.e.~removing the \(n\) pairs of points that are at least \(D\) distance away from each other to remove the desired number of elements).}
-%  Lines @@@
+
+We clarified this point as follows:
+\textit{Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
+$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.}
+Lines 164-167.
 

### 2024-11-28 Manuscript/main_ecography_R1.tex
updated responses to reviewer 1

@@
 Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
-Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to some stressors.
+Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to various stressors.
 In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism.
@@
 Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and Table \ref{Tab:mechanisms}).
-Our goal is to assess the relative performance of different metrics in capturing the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
-We expect that our ability to capture a pattern does not only depend on the choice of metric (what) but also on the process and mechanism at hand (how and why).
+Our goal is to test the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
+We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).
 To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data.
-We show that, with a fixed process and mechanisms approximation, the choice of statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data, with the exact same data being sometimes interpreted in opposite ways.
+We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.
 

### 2024-11-29 Analysis/04_Summarise_results.Rmd
fixed p-value selection in fancy plots
bug came from added ses column

@@
 
-```{r, echo = FALSE}
+```{r, echo = FALSE, eval = FALSE}
 # rownames(model_table_2d) <- gsub("\\]", "_", gsub("\\[", "_", gsub("\n", "", fancy_names)))
@@
 
-```{r, echo = FALSE}
+```{r, echo = FALSE, eval = FALSE}
 print(xtable(change.names(null_table_2d[, 1:8], fancy_names, colnames = TRUE),

### 2024-11-29 Functions/fancy.plot.R

@@ -34,7 +34,7 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
         par(mfrow = c(1, length(results)+1))
         ## Plotting each stressor
         for(one_stressor in 0:length(results)) {
-            
+
             if(one_stressor == 0) {
                 ## Making the labels plot
                 par(mar = c(5, 10, 4, 0), bty = "n")
@@ -46,7 +46,6 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                            xlab  = xlab,
                            xlim  = xlim,
                            metric.names = metric.names)
-
             } else {
                 par(mar = c(5, 0, 4, 1))
                 ## Making the empty plot
@@ -170,16 +169,9 @@ add.one.metric <- function(one_metric, plot_params, probs, cent.tend, col, lm_da
         }
         ## Setting the pch param
         if(!is.null(null_data)) {
-            if(null_data[3*one_level] > 0.05) {
+            if(null_data[3+((one_level*4)-4)] > 0.05) {
                 plotting_params$col <- "grey"
             }
-            # plot_params$pch <- 1
-            # if(null_data[3*one_level] < 0.05) {
-            #     plot_params$pch <- 20
-            #     if(null_data[3*one_level] < 0.001) {
-            #         plot_params$pch <- 19
-            #     }
-            # }
         }
         add.one.line(one_metric[, one_level], plot_params = plotting_params, probs = probs, cent.tend = cent.tend)
     }

### 2024-11-29 Manuscript/ecography_reponse_letter.tex
Reviewer 1 mostly done

@@
 The current structure is challenging to follow.}
-%  We split this sentence in three for clarity:
-%  \textit{Each bandwidth has an observed probability of sampling of \(n\) (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of \(n 
-% times m^{p}\)
-% For bandwidths with a low density (low values of \(n\)) we used \(m\>0\) to increase their resampling probability, and the opposite for bandwidths with a high density (high values of \(n\); using \(m\<0\) to decrease sampling property)
-% This effectively flattened the normal distribution of the data by some factor \(p\) (here we used \(p=3\)).}
-%  Lines @@@
+
+We re-wrote this section to be hopefully clearer:
+\textit{In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
+This is done by selecting $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' (texttt{bw.nrd0()} function in \texttt{R};Silverman, 1986).
+Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$.
+Where $p$ is a factor increasing the scaling power of the algorithm (here we used $p=3$).
+In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.}
+Lines 192-203.
 
 \textcolor{blue}{Table 1. Please provide references for the Equalizing and Facilitation mechanisms to offer clearer context for the focal ecological mechanisms studied.}
-%  We now added references to \textbackslash citep\{chesson2000mechanisms,barot2004mechanisms\} for the equalising mechanism and to \textbackslash citep\{bruno2003inclusion, danet2024species\} for the facilitation one.
-%  Lines @@@
+
+We now added references to Chesson 2000 and Barot 2004 for the equalising mechanism and to Bruno et al. 2003 and Danet et al. 2024 for the facilitation one in Table 1.
 
 \textcolor{blue}{Line 183. Please provide the total number of the different metrics you used and refer to the table below (table 2).}
-%  We've now added a reference to table 2 and summarised the total amount of metrics.
-%  Lines @@@
+
+We've now added a reference to table 2 and summarised the total amount of metrics.
+Lines 264.
 
 \textcolor{blue}{Figure 1. I believe adding a color legend to the figure would enhance its clarity.}
-%  We have now reworked figure 1 to be more didactic (especially for the results interpretation)
-% We have not added the color legend to the plot itself (although we have added orange arrows to make it clearer) but in the caption to not overload an already busy figure.
+
+We have now reworked figure 1 to be more didactic (especially for the results interpretation)
+We have not added the color legend to the plot itself (although we have added orange arrows to make it clearer) but in the caption to not overload an already busy figure.
 
 \textcolor{blue}{Figure 1. Please include information in the figure caption explaining what the two types of graphs represent.}
-%  We added the following clarifications to the figure caption:
-%  \textit{We then applied different stressors to the resulting trait space
-% We represent the elements removed by the stressors in blue and the ones kept in the trait space in orange
-% The first column illustrate the distribution of the elements in one dimension with the overlap between the removed (blue) and kept (orange) elements in brown (note we did not measure any metric on 1D spaces)
-% The second column illustrate the same distributions in 2 dimensions
-% We did not illustrate the distributions in 4 or 8 dimensions here
-% We used the following stressors:}
-%  Lines @@@
+
+We added the following clarifications to the figure caption:
+\textit{We then applied different stressors to the resulting trait space
+We represent the elements removed by the stressors in blue and the ones kept in the trait space in orange.
+The first column illustrate the distribution of the elements in one dimension with the overlap between the removed (blue) and kept (orange) elements in brown (note we did not measure any metric on 1D spaces)
+The second column illustrate the same distributions in 2 dimensions.
+We did not illustrate the distributions in 4 or 8 dimensions here.}
+Figure 1.
 
 \textcolor{blue}{Line 194-195. Please improve clarity.}
+
 %  @@@TODO: rewrite following other reviewers comments (stats for results)
@@
 \textcolor{blue}{Line 201-203. Please simplify and improve clarity.}
+
 %  @@@TODO: rewrite following other reviewers comments (stats for results)
@@
 \textcolor{blue}{Line 207-211. Please provide a clearer explanation of the effect size and statistical significance of these model outcomes, similar to the explanation provided in lines 198-201.}
+
 %  @@@TODO: rewrite following other reviewers comments (stats for results)
 %  @@@TODO: calculate effect sizes
-%  We now added all the statistics results in the supplementary results from table 6 to 18.
+We now added all the statistics results in the supplementary results from table 6 to 18.
 
 \textcolor{blue}{Line 221. Please specify the added species.}
+
 %  @@@TODO: name of the two added marine species
@@
 \textcolor{blue}{Table 2. The aspect for the third row is missing.}
-%  We are not sure about what the reviewer meant here
-% The tables are all formatted in LaTeX and we prefer to leave it to the production team to set it up as they like should the manuscript be accepted.
+
+We are not sure about what the reviewer meant here.
+The tables are all formatted in LaTeX and we prefer to leave it to the production team to set it up as they like should the manuscript be accepted.
 
@@
 Additionally, refer to the relevant section of Figure 2 (to explain how the observed trends in the figure support this assertion), and to the tables in supplementary material.}
+
 %  @@@TODO: add stats for results (following other reviewers comments)
@@
 \textcolor{blue}{Line 252. I would avoid the term ``naively''.}
-%  We removed this term.
+
+We removed this term.
 
 \textcolor{blue}{Line 254. Please ensure that the use of italics follows the guidelines from the journal.}
-%  We did not find any specific guideline regarding emphasising words with italics but removed it nonetheless.
+
+We did not find any specific guideline regarding emphasising words with italics but removed it nonetheless.
 
@@
 \textcolor{blue}{Line 260. Please avoid using the exclamation mark.}
-%  We changed the punctuation to a dot.
+
+We changed the punctuation to a dot.
 
 \textcolor{blue}{Line 262-269. Please provide more detail regarding these results. Additionally, refer to the table in supplementary material.}
+
 %  @@@TODO: add stats for results (following other reviewers comments)
-%  We also now refer to the supplementary results:
-%  \textit{For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure \textbackslash ref\{Fig:empirical\_results\}, Supplementary table 5).}
-%  Lines @@@
+
+We also now refer to the supplementary results:
+\textit{For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure 3, Supplementary table 5).}
+Lines 354-355.
 
 \textcolor{blue}{Line 275-283. This should be in the discussion section.}
-%  We moved this paragraph to the discussion on lines @@@.
+
+We moved this paragraph to the discussion on lines 440-448.
 
@@
 
-%  We removed mention of the R2 (from a previous cluttered version of the figure) and now refer to it in the supplementaries in the figure caption.
+We removed mention of the R2 (from a previous cluttered version of the figure) and now refer to it in the supplementaries in the figure caption.
 
@@
 While the significance of the test is demonstrated through the gray plots in Figures 2 and 3, the relative values are also necessary.}
-%  The tables in the supplementary materials now contain the R2 values for both the Figures 2 and 3
-% We have now clarified that in the figure captions:
-%  \textit{The model results (including \(R^{2}\) and t-statistics) are available in supplementary table 4.}
-%  Lines @@@ and Lines @@@
+
+The tables in the supplementary materials now contain the R2 values for both the Figures 2 and 3.
+We have now clarified that in the figure captions:
+\textit{The model results (including \$R^{2}$ and t-statistics) are available in supplementary table 3.}
 
@@
 This will allow you to address the importance of selecting the appropriate metric in relation to the number of dimensions used for the trait space.}
-% We have now re-orginased the discussion and clearly separated each subsection
-% The first one now discusses the results (simulated and empirical), the second one focuses on the caveats of our approach.
-% And the third section is now a standalone box (box 2) discussing how we believe the what, how and why questions can help workers choose their functional diversity metric of interest based on their own data and question
-% Hopefully these clear distinctions will make it easier to follow (or skip) for the reader.
+
+We have now re-orginased the discussion and clearly separated each subsection.
+The first one now discusses the results (simulated and empirical), the second one focuses on the caveats of our approach.
+And the third section is now a standalone box (box 2) discussing how we believe the what, how and why questions can help workers choose their functional diversity metric of interest based on their own data and question.
+Hopefully these clear distinctions will make it easier to follow (or skip) for the reader.
 
 \textcolor{blue}{Line 313-316. Please rephrase to improve readability.}
-%  We have now removed this sentence.
+
+We have now removed this sentence.
 
@@
 I recommend streamlining this section and, if preferred, moving part of it to the supplementary material, where you can address the implications of scaling your results with the null model.}
-%  We have now moved this section into the named ``caveats'' section
-% Hopefully this will make it easier to skip for readers no interested in the methodological aspects of our work
-% However, if the reviewer still thinks this section is too focused, we're happy to reduce it in a future version.
+
+We have now moved this section into the named ``caveats'' section.
+Hopefully this will make it easier to skip for readers no interested in the methodological aspects of our work.
+However, if the reviewer still thinks this section is too focused, we're happy to reduce it in a future version.
 
@@
 When you refer to your findings, please provide a more explicit reference to the values you found and the relative graph to better guide the reader through your results and your interpretation.}
-%  We have now specified which part of figure 2 we were using to illustrate this point:
-%  \textit{(e.g.~facilitation mechanism measured as divergence using the probability density method - not distinguishable from null when removing 80\textbackslash\% or 60\textbackslash\% of the data; Figure \textbackslash ref\{Fig:simulations\})}
-%  Lines @@@
+
+We have now specified which part of figure 2 we were using to illustrate this point:
+\textit{(e.g. facilitation mechanism measured as divergence using the probability density method - not distinguishable from null when removing 80\% or 60\% of the data; Figure 2)}. Lines 480-482.
 
 \textcolor{blue}{Line 327. Please replace `workers' with `ecologists,' as you are referring to an example from an ecological study.}
-%  We changed ``workers'' to ``researchers'' because not only ecologists are likely to work on such data/questions (e.g.~evolutionary biologists, palaeobiologists, etc.)
+
+We changed ``workers'' to ``researchers'' because not only ecologists are likely to work on such data/questions (e.g. evolutionary biologists, palaeobiologists, etc.)
 
 \textcolor{blue}{Line 339-340. Please simplify and increase readability.}
-%  We reworded this sentence as follows:
-%  \textit{The empirical results are also intriguing in that the loss of Hawaiian bird's functional diversity since 1500 is not greater than expected by chance (at least for certain metrics; Figure \textbackslash ref\{Fig:empirical\_results\}) because one would expect that extinctions since 1500 target specific areas of functional diversity space (i.e.~birds with specific trait combinations).}
-%  Lines @@@
+
+We reworded this sentence as follows:
+\textit{The empirical results are also intriguing in that the loss of Hawaiian bird's functional diversity due to extinction is not greater than expected by chance (at least for certain metrics; Figure 3) because one would expect that extinctions target specific areas of functional diversity space (i.e.
+birds with specific trait combinations).}
+Lines 407-410.
 
@@
 After addressing these points, you can briefly discuss the extinction dynamics of avian species in Hawaii.}
-%  We have now expanded this section to be clearer:
-%  \textit{This illustrates the importance of positing the mechanism and process of interest as well as the pattern to capture it before analysing the results.}
-%  \textit{For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance for the species present before 1500 and larger than expected by change for the current species.}
-%  \textit{This could lead to the interpretation that the trait space was not changed under any specific mechanism before 1500 (i.e.~null mechanism) but that since then, specific mechanisms have been reducing specific areas of the trait space (e.g.~filtering mechanism).}
-%  \textit{However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before 1500 leading to very different interpretation of the results.}
-%  Lines @@@
+
+We have now expanded this section to be clearer:
+\textit{This illustrates the importance of positing the mechanism and process of interest as well as the pattern to capture it before analysing the results.
+For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance when focusing on all extinctions and larger than expected by chance when considering only pre-$1500$ extinctions.
+This could lead to the interpretation that the trait space was not changed under any specific mechanism up to $1500$ (i.e.
+null mechanism) but that since then, specific mechanisms have been reducing specific areas of the trait space (e.g.
+filtering mechanism).
+However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before $1500$ leading to very different interpretation of the results.}
+Lines 413-424.
 
@@
 This would enhance the overall flow of the paper and help the reader more clearly connect the main question and introduction to the final conclusions and discussion.}
-%  We have now added the box 2 that would hopefully help with strengthening the connection between our results and the need to set up the ``what how and why'' framework for functional diversity analyses.
+
+We have now added the box 2 that would hopefully help with strengthening the connection between our results and the need to set up the ``what how and why'' framework for functional diversity analyses.
 
@@
 \textcolor{blue}{Like for figure 2 in the main manuscript, I think that you are missing the adjusted R square from the figure.}
-%  We have now specified in the caption that the adjusted R squares are available in the supplementary tables to not clutter the figure.
+
+We have now specified in the caption that the adjusted R squares are available in the supplementary tables to not clutter the figure.
 

### 2024-11-29 Manuscript/main_ecography_R1.tex
Reviewer 1 mostly done

@@
 The second plot on the same row shows the same results but with a two dimensional trait represented here.
-We then applied different stressors to the resulting trait space.
-We represent the observations removed by the stressors in blue and the ones kept in the trait space in orange.
-The first column (histograms) illustrates the distribution of observations in one dimension of the 2D, 4D or 8D traits with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces).
-The second column illustrates the trait in 2 dimensions.
-We did not illustrate the data beyond 2 dimensions for traits in 4 or 8 dimensions.
+We then applied different stressors to the resulting trait space
+We represent the elements removed by the stressors in blue and the ones kept in the trait space in orange.
+The first column illustrate the distribution of the elements in one dimension with the overlap between the removed (blue) and kept (orange) elements in brown (note we did not measure any metric on 1D spaces)
+The second column illustrate the same distributions in 2 dimensions.
+We did not illustrate the distributions in 4 or 8 dimensions here.
 We used the following stressors:
@@
 This illustrates the importance of positing the mechanism and process of interest as well as the pattern to capture it before analysing the results.
-For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance when focusing on all extinctions and larger than expected by chance when considereing only pre-$1500$ extinctions.
-% This could lead to the interpretation that the trait space was not changed under any specific mechanism before $1500$ (i.e.
+For example, for the divergence metrics using probability density, hypervolume or dendrogram methods, we recovered changes in trait space that are smaller than expected by chance when focusing on all extinctions and larger than expected by chance when considering only pre-$1500$ extinctions.
+This could lead to the interpretation that the trait space was not changed under any specific mechanism up to $1500$ (i.e.
 null mechanism) but that since then, specific mechanisms have been reducing specific areas of the trait space (e.g.
 filtering mechanism).
-% However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before $1500$ leading to very different interpretation of the results.
-%TM: can we make this interpretation? as this isn't the two time periods, i.e., there isn't a before and after 1500 comparison. It is all extinctions, and then just pre-1500 extinctions. 
-%TM: And also see my previous comment - is not saying the opposite? That there was some mechanism pre 1500? But not when considering all extinctions?
-%TODO@@@
+However, when using a different metric (say the Rao's quadratic entropy based on distance matrices), the results would indicate that a reduction of specific areas of the trait space could have also occured before $1500$ leading to very different interpretation of the results.
 
@@
 For example, if removing 80\% of the edges of a distribution was distinguishable or not from removing 80\% of a distribution randomly (e.g.
-facilitation mechanism measured as divergence using the probability density method - not distinguishable from null when removing 80\% or 60\% of the data; Figure \ref{Fig:simulations}).
+facilitation mechanism measured as divergence using the probability density method - not distinguishable from null when removing 80\% or 60\% of the data; Figure \ref{Fig:simulation_results}).
 Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).

### 2024-11-29 Manuscript/ecography_reponse_letter.tex
dent in reviewer 2 comments

@@
 We have now clarified that in the figure captions:
-\textit{The model results (including \$R^{2}$ and t-statistics) are available in supplementary table 3.}
+\textit{The model results (including $R^{2}$ and t-statistics) are available in supplementary table 3.}
 
@@
 
-% We have now improved both the mechanisms approximation description (see comment below) as well as added more justification on why we believe our algorithms approximate the focal mechanisms (see response to the next comment).
-%  Furthermore, we have also added some general predictions as to which type of metric would capture which type of changes in the trait space (Table 1).
-%  We hope that these three points improve the clarity of our objectives and justifications.
+We have now improved both the mechanisms approximation description (see comment below) as well as added more justification on why we believe our algorithms approximate the focal mechanisms (see response to the next comment).
+Furthermore, we have also added some general predictions as to which type of metric would capture which type of changes in the trait space (Table 1).
+We hope that these three points improve the clarity of our objectives and justifications.
 
 \textcolor{blue}{\textbf{Model description:}}
-The section on simulating trait-space patterns needs to be significantly expanded to include more technical details regarding how the simulations were conducted.
+
+\textcolor{blue}{The section on simulating trait-space patterns needs to be significantly expanded to include more technical details regarding how the simulations were conducted.
 Key information, such as the model parameters, is entirely absent.
@@
 
-% We have now developed the explanation of the algorithms description as follows:
-%  \textit{\textbackslash begin\{itemize\}}
-%  \textit{\textbackslash item Random removal: by randomly removing 20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\% of the data.}
-%  \textit{This approximates our \textbackslash textbf\{null mechanism\}
-% This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly
-% We chose this algorithm to approximate the absence of any specific stressor (i.e.~the removal of data only changes the number of observations, not the properties of the trait space).}
+We have now developed the explanation of the algorithms description as follows:
 
-% \textit{\textbackslash item Decreasing size: by removing the required amount (20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\%) of data away from a distance (radius) \(
-% rho\) of the centre of the trait space.}
-%  \textit{\(
-% rho\) is estimated for each trait space to match the required intensity
-% For example, for one trait space, the algorithm estimated \(
-% rho\_{20}\) to be the minimum radius excluding 20\textbackslash\% of the observations, \(
-% rho\_{40}\) the one removing 40\textbackslash\%, etc
-% (with \(
-% rho\_{20} \> 
-% rho\_{40} \> 
-% rho\_{60} \> 
-% rho\_{80}\)).}
-%  \textit{This approximates our \textbackslash textbf\{equalising mechanism\}.}
-%  \textit{We chose this algorithm to approximate equalisation assuming that a stressor could increase the probability of extinction for observations with more rare trait combinations (i.e.~observations on the edges of the trait space).}
+\begin{itemize}
+\item \textit{Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data.
+This approximates our \textbf{null mechanism}.
+This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly.
+We chose this algorithm to approximate the absence of any specific stressor (i.e.
+the removal of data only changes the number of observations, not the properties of the trait space in a systematic way).}
 
-% \textit{\textbackslash item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least \(D\).}
-%  \textit{\(D\) is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least \(n\) pairs of points corresponding to the intensity of the stressor.}
-%  \textit{For example \(D\_{20}\) is the distance that excludes any \(n\_{20}\) pairs of points that are a distance of at least \(D\_{20}\) from each other, \(D\_{40}\) is the distance that excludes \(n\_{40}\) pairs, etc
-% (with \(D\_{20} \> D\_{40} \> D\_{60} \> D\_{80}\)).}
-%  \textit{Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.}
-%  \textit{This approximates our \textbackslash textbf\{facilitation mechanism\} (i.e.~the points left are only ones that are close to at least another point in space).}
-%  \textit{We chose this algorithm to approximate facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.}
-%  \textit{I.e
-% pairs of observations that share a close trait combinations are more likely to remain (i.e.~observations adjacent to each other in the trait space).}
+\item \textit{Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius) $\rho$ of the centre of the trait space.
+$\rho$ is estimated for each trait space to match the required intensity.
+For example, for one trait space, the algorithm estimated $\rho_{20}$ to be the minimum radius excluding 20\% of the observations, $\rho_{40}$ the one removing 40\%, etc.
+(with $\rho_{20} > \rho_{40} > \rho_{60} > \rho_{80}$).
+This approximates our \textbf{equalising mechanism}. 
+We chose this algorithm to approximate equalisation assuming that a stressor could increase the probability of extinction for observations with more rare trait combinations (i.e.
+observations on the edges of the trait space).}
 
-% \textit{\textbackslash item Shifting space: by removing the required amount of data from a distance (radius) \(
-% rho'\) of the observation with the maximum numerical value on all dimensions.}
-%  \textit{This is similar to the decreasing size algorithm but instead of choosing \(
-% rho\) to be the radius from the centre of the trait space, \(
-% rho'\) is a radius from the ``top right corner'\,' of the trait space.}
-%  \textit{In other words the centre of the radius \(
-% rho'\) is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with \(
-% rho'\_{20} \> 
-% rho'\_{40} \> 
-% rho'\_{60} \> 
-% rho'\_{80}\)).}
-%  \textit{This approximates our \textbackslash textbf\{filtering mechanism\}.}
-%  \textit{We chose this algorithm to approximate a flitering assuming that a stressor that could increase the probability of extinction for observations further from some trait combination optimum.}
-%  \textit{This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).}
+\item \textit{Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
+$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
+For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc.
+(with $D_{20} > D_{40} > D_{60} > D_{80}$).
+Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
+This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
+We chose this algorithm to approximate the facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
+I.e. pairs of observations that share similar trait combinations are more likely to remain (i.e. observations adjacent to each other in the trait space).}
 
-% \textit{\textbackslash item Increasing evenness: by resampling the proportion of data (i.e.~20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\%) but with skewed resampling probabilities.}
-%  \textit{In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.}
-%  \textit{This is done by selecting of \(B\) discrete categories to summarise the distribution - i.e.~bandwidths - using Silverman's ``rule of thumb'\,' \textbackslash citep{[}\textbackslash texttt\{bw.nrd0()\} function in \textbackslash texttt\{R\};{]}{[}{]}\{silverman1986density\}.}
-%  \textit{Each of these discrete category \(B\) has an observed probability of sampling of \(b\) (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove \(i\)) of \(i
-% times (1-b)^{p}\).}
-%  \textit{Where \(p\) is factor increasing the scaling power of the algorithm (here we used \(p=3\)).}
-%  \textit{In other words, observations in categories with a low density (\(b\<0.5\)) were more likely to be resampled and the ones in categories with high density (\(b\>0.5\)) were less likely to be resampled.}
-%  \textit{This approximates our \textbackslash textbf\{competition mechanism\}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.}
-%  \textit{We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.}
-%  \textit{This could approximate the fact that observation that have similar trait combinations are more likely to go extinct, due to competition (e.g.~using the same resources).}
-%  \textit{\textbackslash end\{itemize\}}
-%  Lines @@@
+\item \textit{Shifting space: by removing the required amount of data from a distance (radius) $\rho'$ of the observation with the maximum numerical value on all dimensions.
+This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
+In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
+This approximates our \textbf{filtering mechanism}.
+We chose this algorithm to approximate a flitering mechanism assuming a stressor that could increase the probability of extinction for observations further from some trait combination optimum.
+This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).}
+
+\item \textit{Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
+In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
+This is done by selecting $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' (\texttt{bw.nrd0()} function in \texttt{R}; Silverman 1986).
+Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$.
+Where $p$ is a factor increasing the scaling power of the algorithm (here we used $p=3$).
+In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
+This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
+We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
+This could approximate the fact that observations that have similar trait combinations are more likely to go extinct due to competition (e.g. using the same resources).}
+\end{itemize}
+Lines 148-209
 
@@
 
-% We have now standardised the terminology following \textbackslash cite\{guillerme2020shifting\} and \textbackslash cite\{mammola2021concepts\}
-% In brief, elements, tips or species are now referred to as ``observations''; and traits, dimensions and columns as ``dimensions'' when referring to our simulation approach
-% We also added a Glossary to the supplementary materials @@@ for clarity.
+We have now standardised the terminology following Guillerme et al. 2020 and Mammola et al 2021.
+In brief, elements, tips or species are now referred to as ``observations''; and traits, dimensions and columns as ``dimensions'' when referring to our simulation approach
+We also added a Glossary to the supplementary materials table 1 for clarity.
 
@@
 This implies that only a single trait dimension is generated, whereas from the later text it seems that higher-dimensional traits are simulated.}
-%  By ``trait'' here we refer to the single process (Brownian Motion) which can be simulated in any number of dimensions (i.e.~generating a time dependent element value using 2, 4 or 8 dimension coordinates to describe it)
-% In this context we differentiate ``dimensions'' which can be interpreted as a column in the data matrix, and the ``trait'' which is the definition of the measured (or simulated) feature
-% For example, the trait ``geographical location'' can be described using two dimensions (latitude and longitude) or can be described as two traits (``latitude coordinate'' and ``longitude coordinate'')
-% In this manuscript we preferred using the first definition
-% We've clarified this as follows:
-%  \textit{First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.~a pure birth speciation model) until reaching 200 elements in \textbackslash texttt\{R\} \textbackslash citep\{rcore\} using \textbackslash texttt\{treats\} \textbackslash citep\{guillerme2024treats\}
-% This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensional space.}
-%  Lines@@@
+
+By ``trait'' here we refer to the single process (Brownian Motion) which can be simulated in any number of dimensions (i.e.
+generating a time dependent element value using 2, 4 or 8 dimension coordinates to describe it).
+In this context we differentiate ``dimensions'' which can be interpreted as a column in the data matrix, and the ``trait'' which is the definition of the measured (or simulated) feature.
+For example, the trait ``geographical location'' can be described using two dimensions (latitude and longitude) or can be described as two traits (``latitude coordinate'' and ``longitude coordinate'').
+In this manuscript we preferred using the first definition.
+
+We've clarified this as follows:
+\textit{First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
+a pure birth speciation model) until reaching 200 observations in \texttt{R} (R Core Team, 2024) using \texttt{treats} (Guillerme, 2024).
+This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensional space.}
+Lines 125-129.
 
 \textcolor{blue}{L115: what is a ``tip''?}
-%  By tip we meant a node with no descendants in an evolutionary tree
-% We have now standardised the terminology (see response to general comment above).
+
+By tip we meant a node with no descendants in an evolutionary tree.
+We have now standardised the terminology (see response to general comment above) and changed it to observation.
 
@@
 One of the main advantages of using simple models like a branching process is that they allow for the generation of larger datasets, which would provide more robust statistics and better analytical power.}
+
 %  @@@TODO: Rerun the simulations with 400, 800 and 1000 elements.
@@
 \textcolor{blue}{LL 120-124, ``four different stressors, five specific algorithms'': Inconsistency in the used terminology.
-In this section, ``stressors'' appear to refer to the amount of data being removed (ranging from 20\% to 80\%), but later, the term ``stressors'' is used to describe the five data-removal processes---referred to as ``specific algorithms'' here.}
-% We have now standardised the terminology throughout the manuscript (and included it in the glossary table in the supplementary materials): ``algorithm'' now designate the code algorithm, ``stressors'' (and ``stressors' intensity'') now designates how we modify the data (using the algorithms mentioned above) and ``mechanisms'' now designates the biological mechanisms we are approximating using a specific stressor (that is based on a specific algorithm)
-% We've namely changed this section for clarity:
-%  \textit{We then applied five different stressors to the trait space with different intensities.}
-%  \textit{The five different stressors are effectively five specific algorithms removing either 20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\% of the data (the intensities of the stressors, resulting in trait spaces with, respectively, 160, 120, 80 and 40 observations).}
-%  \textit{We used the following specific algorithms (Figure \textbackslash ref\{Fig:simulations\}; Table \textbackslash ref\{Tab:mechanisms\}; all algorithms, except ``increasing evenness'\,', were previously described in \textbackslash cite\{guillerme2020shifting\}):}
-%  Lines @@@
+In this section, ``stressors'' appear to refer to the amount of data being removed (ranging from 20\% to 80\%), but later, the term ``stressors'' is used to describe the five data-removal processes - referred to as ``specific algorithms'' here.}
+
+We have now standardised the terminology throughout the manuscript (and included it in the glossary table in the supplementary materials): ``algorithm'' now designate the code algorithm, ``stressors'' (and ``stressors' intensity'') now designates how we modify the data (using the algorithms mentioned above) and ``mechanisms'' now designates the biological mechanisms we are approximating using a specific stressor (that is based on a specific algorithm).
+
+We've namely changed this section for clarity:
+\textit{We then applied five different stressors to the trait space with different intensities.
+The five different stressors are effectively five specific algorithms removing either 20\%, 40\%, 60\% or 80\% of the data (the intensities of the stressors, resulting in trait spaces with, respectively, 160, 120, 80 and 40 observations).
+We used the following specific algorithms (Figure 1; Table 1; all algorithms, except ``increasing evenness'', were previously described in Guillerme et al. 2020):}
+Lines 142-147.
 
@@
 For instance, the equalizing scheme would also lead to an increase in density.}
-%  We used the term ``increasing density'' here to differentiate from ``decreasing size''
-% The reviewer is totally correct in their understanding that, especially for normally distributed data, changes in density in the data actually occur to some extent with all the algorithms used here
-% The stressor names and their algorithms are based on Guillerme et al 2020 and their implementation in the dispRity package: in their description they don't assume any specific distribution of the data a priori
-% In fact, when using a uniform distribution, the decreasing size stressor does not change the density and the increasing density algorithm does not decrease the size of the trait space.
-%  We've now added the following note to include the reviewer's comment:
-%  \textit{Note that the algorithm used here to simulate the stressors depend, to some extent, to the distribution of the data for each trait space.}
-%  \textit{The algorithms used can lead to similar stressors effects depending on the simulated data distribution.}
-%  \textit{For example, ``increasing density'\,' and ``decreasing size'\,' stressors will lead to just what their name suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'\,' and just overall decrease in size for the ``decreasing size'\,' stressors) but will have more or less similar effects (albeit no identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'\,' and ``decreasing size'\,' stressors.)}
-%  Lines @@@
+
+We used the term ``increasing density'' here to differentiate from ``decreasing size''.
+The reviewer is totally correct in their understanding that, especially for normally distributed data, changes in density in the data actually occur to some extent with all the algorithms used here.
+The stressor names and their algorithms are based on Guillerme et al 2020 and their implementation in the \texttt{dispRity} package: in their description they don't assume any specific distribution of the data \textit{a priori}.
+In fact, when using a uniform distribution, the decreasing size stressor does not change the density and the increasing density algorithm does not decrease the size of the trait space.
+
+We've now added the following note to include the reviewer's comment:
+\textit{Note that the algorithm used here to simulate the stressors depends, to some extent, on the distribution of the data for each trait space.
+The algorithms used can lead to similar stressor effects depending on the simulated data distribution.
+For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors.)}
+Lines 213-221.
 
@@
 However, this effect of shifting the mean is not unique to this scheme - other removal schemes will generally result in a shift of the mean as well.}
-%  This is also a fair point
-% See comment above.
+
+This is also a very fair point.
+See our response to the comment above.
 
@@
 Maybe this refer to the location in trait space with the highest density of trait elements?}
-%  We have now specified what we meant by maximal value:
-%  \textit{Shifting space: by removing the required amount of data from a distance (radius) \(
-% rho'\) of the observation with the maximum numerical value on all dimensions.}
-%  \textit{This is similar to the decreasing size algorithm but instead of choosing \(
-% rho\) to be the radius from the centre of the trait space, \(
-% rho'\) is a radius from the ``top right corner'\,' of the trait space.}
-%  \textit{In other words the centre of the radius \(
-% rho'\) is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with \(
-% rho'\_{20} \> 
-% rho'\_{40} \> 
-% rho'\_{60} \> 
-% rho'\_{80}\)).}
-%  Lines @@@
+
+We have now specified what we meant by maximal value:
+\textit{Shifting space: by removing the required amount of data from a distance (radius) $\rho'$ of the observation with the maximum numerical value on all dimensions.
+This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
+In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).}
+Lines 168-184.
 
@@
 
-%  We have now completely rewritten the description of this stressor by defining more clearly what the bandwidth is, renaming the variables to be easier to understand, removed mention of flattening the curve and rewrote the equations to be specific to the increase in evenness (previous equations were generalised to both increase or decrease the eveness depending on the variable \textit{i} (formerly \textit{n}) being now always a proportion of removal (e.g.~0.2, 0.4, \ldots{} - former description worked with \textit{n} being also a proportion to add e.g.~1.2, 1.4, etc\ldots):
-%  \textit{Increasing evenness: by resampling the proportion of data (i.e.~20\textbackslash\%, 40\textbackslash\%, 60\textbackslash\% or 80\textbackslash\%) but with skewed resampling probabilities.}
-%  \textit{In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.}
-%  \textit{This is done by selecting of \(B\) discrete categories to summarise the distribution - i.e.~bandwidths - using Silverman's ``rule of thumb'\,' \textbackslash citep{[}\textbackslash texttt\{bw.nrd0()\} function in \textbackslash texttt\{R\};{]}{[}{]}\{silverman1986density\}.}
-%  \textit{Each of these discrete category \(B\) has an observed probability of sampling of \(b\) (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove \(i\)) of \(i
-% times (1-b)^{p}\).}
-%  \textit{Where \(p\) is factor increasing the scaling power of the algorithm (here we used \(p=3\)).}
-%  \textit{In other words, observations in categories with a low density (\(b\<0.5\)) were more likely to be resampled and the ones in categories with high density (\(b\>0.5\)) were less likely to be resampled.}
-%  \textit{This approximates our \textbackslash textbf\{competition mechanism\}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.}
-%  Lines @@@
+We have now completely rewritten the description of this stressor by defining more clearly what the bandwidth is, renaming the variables to be easier to understand, removed mention of flattening the curve and rewrote the equations to be specific to the increase in evenness (previous equations were generalised to both increase or decrease the eveness depending on the variable \textit{i} (formerly \textit{n}) being now always a proportion of removal (e.g.
+0.2, 0.4, ...
+- former description worked with \textit{n} being also a proportion to add e.g.
+1.2, 1.4, etc.):
+
+\textit{Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
+In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
+This is done by selecting $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' (\texttt{bw.nrd0()} function in \texttt{R}; Silverman, 1986).
+Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$.
+Where $p$ is a factor increasing the scaling power of the algorithm (here we used $p=3$).
+In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
+This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space..}
+Lines 191-205.
 
@@
 \textcolor{blue}{- The caption completely omits any description of what is shown in left and right columns (i.e., 1d and 2d case).}
-%  We have now clarified what each column contains in the figure caption.
+
+We have now clarified what each column contains in the figure caption.
+
 \textcolor{blue}{- It is unclear whether this figure (both the 1d and 2d cases) is based on the actual simulations or is merely a schematic sketch.
@@
 
-% The figures does depict 200 points (200 blue in row A and 100 blue + 100 orange in the other columns).
-% We have now specified this in the caption.
+The figures does depict 200 points (200 blue in row A and 100 blue + 100 orange in the other columns; see specific script: \url{https://github.com/TGuillerme/eco_metrics_simulations/blob/a094183ab65543adc805db0b0c4f24314c9c7e99/Analysis/01_Simulating_spaces_and_stressors.Rmd#L40}).
+We have now specified this in the caption.
+
 \textcolor{blue}{- The caption mentions a ``trait space of 2, 4, or 8 dimensions,'' yet a 1d trait space is shown.
 If a 1d trait space is an illustrative example, this should be specified in the caption.}
-%  We have now specified that we are showing 1 or 2 dimensions regardless of the number of dimensions per traits (i.e.~8 dimensional traits are not fully represented here but what's happening on their first and second dimensions is displayed here)
-% We think one confusion on our part is the distinction between a dimension (a column in a matrix) and a trait (a measurable feature for an observation)
-% In our case we are always only looking at one trait per trait space
-% Although these traits can have 2, 4 or 8 dimensions
-% We have now clarified this in the glossary.
+
+We have now specified that we are showing 1 or 2 dimensions regardless of the number of dimensions per traits (i.e.
+8 dimensional traits are not fully represented here but what's happening on their first and second dimensions is displayed here).
+We think one confusion on our part is the distinction between a dimension (a column in a matrix) and a trait (a measurable feature for an observation).
+In our case we are always only looking at one trait per trait space.
+Although these traits can have 2, 4 or 8 dimensions.
+We have now clarified this in the glossary (supplementary table 1).
+
 \textcolor{blue}{- The pattern shown in the subfigure labeled ``2d facilitating mechanism'' appears questionable.
 If point-pairs from a 2d distribution are removed based on a threshold distance, the resulting pattern would not typically resemble what is depicted here.}
-%  The illustration of the 2D facilitating mechanism directly comes from the output of the dispRity::reduce.space function where the algorithm is implemented
-% Here it selected every \textit{n} pairs of observations that are the closest to each other so that 50\% of the data is selected
-% Here \textit{n} is 50 so that 100 observations are selected
-% The algorithm is recursive so that once a pair of observations is selected, it cannot be selected again.
+
+The illustration of the 2D facilitating mechanism directly comes from the output of the \texttt{dispRity::reduce.space} function where the algorithm is implemented (implementation: \url{https://github.com/TGuillerme/dispRity/blob/master/R/reduce.space.R}; test: \url{https://github.com/TGuillerme/dispRity/blob/master/tests/testthat/test-reduce.space.R}.
+Here it selected every \textit{n} pairs of observations that are the closest to each other so that 50\% of the data is selected.
+Here \textit{n} is 50 so that 100 observations are selected.
+The algorithm is recursive so that once a pair of observations is selected, it cannot be selected again.
+
 \textcolor{blue}{- Terminology: The use of the term ``species'' in the caption is confusing, as the methods seem to generate traits rather than species.}
-%  We have now removed any mention of species.
+
+We have now removed any mention of ``species'' and replaced it with ``observations''.
+
 \textcolor{blue}{- The phrase ``removing half of the species, resulting in two groups of tips'' is unclear.
@@
 Additionally, it is unclear which color (blue or orange) corresponds to which group.}
-%  We have now specified this in the new caption.
+
+We have now specified this in the new caption.
+
 \textcolor{blue}{- I am skeptical about the use of ecological terminology such as ``competition'' and ``facilitation'' in this context.
@@
 While these processes can be discussed and associated with ecological mechanisms in the text or discussion (if well justified), using such terms directly in the figure caption is highly misleading.}
+
 %  @@@TODO: Links back to main comment from reviewer 2: Go back to the previous version of table 1 that gives more description of the models

### 2024-11-29 Manuscript/main_ecography_R1.tex
dent in reviewer 2 comments

@@
 The algorithms used can lead to similar stressor effects depending on the simulated data distribution.
-For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors.)
+For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors).
 These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 intensities of data removal).

### 2024-11-29 Manuscript/ecography_reponse_letter.tex
updated reviewer 2 responses

@@
 In the phrase ``replicates per simulations,'' the term ``simulation'' seems to have a different meaning than in ``smallest number of simulations.'' Additionally, the first part of the sentence suggests that variance decreases with an increasing number of simulations, whereas the second part implies that adding more simulations leads to increased variance.}
-%  There was a confusion indeed
-% We now refer to ``the simulation pipeline'' for the whole set of replicates per stressors, and ``replicates'' for each individual run (one trait space + one stressor)
-% We changed this sentence to:
-%  \textit{We used 171 replicates because that was empirically the smallest number of replicates required to reach a variance between replicates lower than 1\textbackslash\% across all metrics (i.e.~any additional replicates beyond 171 added less that 1\textbackslash\% extra variance).}
-%  Lines @@@
+
+There was a confusion indeed.
+We now refer to ``the simulation pipeline'' for the whole set of replicates per stressors, and ``replicates'' for each individual run (one trait space + one stressor).
+We changed this sentence to:
+\textit{We used 171 replicates because that was empirically the smallest number of replicates required to reach a variance between replicates lower than 1\% across all metrics (i.e.
+any additional replicates beyond 171 added less that 1\% extra variance).}
+Lines 223-225
 
@@
 Does this mean that the branching process was independently repeated four times, once for each trait dimension, or does it mean that it generates four trait dimensions simultaneously, with these dimensions being uncorrelated?}
-%  We meant that the dimensions for each multidimensional traits are uncorrelated (i.e.~the trait is a uncorrelated Brownian motion process in n dimensions)
-% We have now clarified this point:
-%  \textit{We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \textbackslash ref\{Fig:simulation\_results\}) for 2 and 8 dimensional traits with not correlation between each dimensions in each simulation pipeline (supplementary materials).}
-%  Lines @@@
+
+We meant that the dimensions for each multidimensional traits are uncorrelated (i.e. the trait is a uncorrelated Brownian motion process in $n$ dimensions).
+We have now clarified this point:
+\textit{We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure 2) for 2 and 8 dimensional traits with not correlation between each dimensions in each simulation pipeline (supplementary materials figures 1 and 2).}
+Lines 241-243.
 
@@
 Specifically, the abrupt mention of species (which are not part of the model) is confusing.}
-%  We have now improved the context of the tables with predictions and removed most of the mentions of ``species'' now replaced by the standardised term of ``observations'' (unless they were specific examples).
+
+We have now improved the context of the tables with predictions and removed most of the mentions of ``species'' now replaced by the standardised term of ``observations'' (unless they were specific examples).
 
 \textcolor{blue}{L 178 and several instances later ``o'' should read ``0'' (zero).}
-%  These integers were compiled using the latex mathpazo font
-% We have now clarified them by changing into a default serif font.
+
+These integers were compiled using the latex mathpazo font.
+We have now clarified them by changing into a default serif font.
 
 \textcolor{blue}{L184, I suggest to mention here, that these functional metrics are denoted as trait space occupancy later on.}
-%  Great suggestions
-% We added the following:
-%  \textit{We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (trait space occupancy metrics; \textbackslash cite\{mammola2021concepts\}) using 12 metrics (Tab
-% \textbackslash ref\{Tab:metrics\}):}
-%  Lines@@@
+
+Great suggestions!
+We added the following:
+\textit{We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (trait space occupancy metrics; Mammola et al.
+2021) using 12 metrics (Table 2):}
+Lines 262-264.
 
 \textcolor{blue}{L 208, ``\textasciitilde{}'' symbol: This usage of mathematical notation (instead of plain words) in a sentence (here, but also later in the text) is difficult to understand.}
-%  We removed tilde symbols throughout the manuscript.
+
+We removed tilde symbols throughout the manuscript.
 
 \textcolor{blue}{Table 2: Column two (``Methods'') does not provide any information beyond what is explained in ``Description'' and can be removed.}
-%  We decided to keep the ``methods'' column to make it easier for the reader to refer to the terms used in Mammola et al.~2021 at a glance.
+
+We decided to keep the ``methods'' column to make it easier for the reader to refer to the terms used in Mammola et al. 2021 at a glance.
 
@@
 The term ``space occupancy'' could be contributing to this misconception.}
-%  We rewritten this section to be in line with the reviewer's comment:
-%  \textit{Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted into a score increase (e.g.~the functional evenness based on hypervolume for the equalising mechanism) or non-linear responses (e.g.~Alpha diversity based on dendrogram for the equalising mechanism).}
-%  Lines @@@
-%  Additionally, there appears to be a logical inconsistency
+
+We re-wrote this section to be in line with the reviewer's comment:
+\textit{Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted in a score increase (e.g.
+the Functional evenness based on a hypervolume for the equalising mechanism) or non-linear responses (e.g.
+Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores).}
+Lines 344-348.
+
+Additionally, there appears to be a logical inconsistency.
 % Figure 2 illustrates the difference between two functional diversity metrics, rather than the metrics themselves

### 2024-11-29 Manuscript/main_ecography_R1.tex
updated reviewer 2 responses

@@
 
-We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}), and for 2 and 8 dimensional (see supplementary materials).
+We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}), and for 2 and 8 dimensional (see supplementary materials figures 1 and 2).
 In each pipeline, the dimensions where simulated as uncorrelated.
@@
 
-We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (trait space occupancy metrics; \cite{mammola2021concepts}) using 12 metrics (Tab.
-\ref{Tab:metrics}):
+We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (trait space occupancy metrics; \cite{mammola2021concepts}) using 12 metrics (Table \ref{Tab:metrics}):
 1) Richness, encompassing metrics reflecting the sum of differences among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});

### 2024-11-29 Manuscript/ecography_reponse_letter.tex
reviewer 2 done

@@
 
-Additionally, there appears to be a logical inconsistency.
-% Figure 2 illustrates the difference between two functional diversity metrics, rather than the metrics themselves
-% Therefore, changes in this difference do not provide information about the individual increases or decreases of the metrics.
-%  We have now added supplementary figures showing the scaled metric score changes (to keep the values manageable between 0 and 1) but not also scaled to the null mechanism changes (to display just the raw metric changes).
+\textcolor{blue}{Additionally, there appears to be a logical inconsistency.
+Figure 2 illustrates the difference between two functional diversity metrics, rather than the metrics themselves.
+Therefore, changes in this difference do not provide information about the individual increases or decreases of the metrics.}
+
+We have now added supplementary figures 3, 4 and 5 showing the scaled metric score changes (to keep the values manageable between 0 and 1) but not also scaled to the null mechanism changes (to display just the raw metric changes).
 
 \textcolor{blue}{L 287: Why would this be ``unexpected''?}
-%  We removed the word ``unexpected'' that was maybe ironic (and irrelevant regardless).
+
+We removed the word ``unexpected'' that was maybe ironic (and irrelevant regardless).
 
@@
 Please specify, for example, which mechanisms have correctly been explained in the analysis?}
-%  We have now reworked this section with the inclusion of a Box explaining in more details how we believe that using the ``what, how, why'' framework helps choosing the correct metric for a specific dataset and question.
+
+We have now reworked this section with the inclusion of a Box explaining in more details how we believe that using the ``what, how, why'' framework helps choosing the correct metric for a specific dataset and question.
 

### 2024-11-29 Manuscript/ecography_reponse_letter.tex
reviewer 3 done

@@
 
-% We have now improved the description of the algorithms (lines @@@) as well as added our predictions to Table 1 and improved Figure 1 to be a better tool to understand our simulation protocol and how to read the results figures.
-%  Furthermore we agree with the reviewer that the effect of different mechanisms can be sometimes more intuitive than others (e.g.~in their example and counterexample of equalizing vs.~competition).
+We have now improved the description of the algorithms (lines 148-209) as well as added our predictions to Table 1 and improved Figure 1 to be a better tool to understand our simulation protocol and how to read the results figures.
+Furthermore we agree with the reviewer that the effect of different mechanisms can be sometimes more intuitive than others (e.g. in their example and counterexample of equalizing vs. competition).
 
-% We've expanded on the following paragraph:
-%  \textit{Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.}
-%  \textit{For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.}
-%  \textit{However, if the data are distributed uniformly, this does not happen.}
-%  \textit{Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.}
-%  \textit{However, this is not tackled here for both simplicity and to understand how they work in isolation.}
-%  \textit{Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \textbackslash ref\{Tab:mechanisms\}) should be taken as a cautious guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.}
-%  Lines @@@
+We've expanded on the following paragraph:
+\textit{Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
+For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
+However, if the data are distributed uniformly, this does not happen.
+Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
+However, this is not tackled here for both simplicity and to understand how they work in isolation. 
+Therefore the prediction of how each different mechanism approximation would affect the metrics (Table 1) should be taken as a cautious guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.}
+Lines 228-240.
 
@@
 
-% We add the following clarification in the methods for the 8 dimensional spaces:
-%  \textit{We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \textbackslash texttt\{TPD::TPDsMean\} is only implemented for up to 4 dimensions; \textbackslash citealt\{carmona2019trait\} - note that int the supplementary figures, we measured the \textbackslash texttt\{TPD\} metrics on only the four first dimensions)}
-%  Lines @@@
-%  And:
-%  \textit{Although this number of dimensions is in the range of many studies in ecology (e.g.~6 dimensions in \textbackslash citealt\{healy2019animal\}), it is not uncommon to use a much greater number of dimensions, especially in palaeontology (e.g.~more than 200 dimensions in \textbackslash citealt\{van2023should\}).}
-%  Lines @@@
+We add the following clarification in the methods for the 8 dimensional spaces:
+\textit{We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; Carmona et al.
+2019 - note that in the supplementary figures, we measured the \texttt{TPD} metrics on only the four first dimensions)}.
+Lines 244-247.
 
-% We also specified the that we used only the four first dimensions for the TPD metrics in the supplementary results
+And:
+\textit{Although this number of dimensions is in the range of many studies in ecology (e.g.
+6 dimensions in Healy et al.
+2019), it is not uncommon to use a much greater number of dimensions, especially in palaeontology (e.g.
+more than 200 dimensions in Van Den Ende et al.
+2023).}
+Lines 469-472.
+
+We also specified the that we used only the four first dimensions for the TPD metrics in the supplementary results.
 
@@
 How is the selection of five dimensions justified for the analysis? Are these properties of the dataset considered important for the example?}
-%  We chose to use five dimensions to describe the Hawaiian birds trait space to make sure that each focal group has at least 95\% of their variance described in the trait space (following Guillerme et al.~2023 protocol).
-%  Furthermore, the choice of 5 dimensions is congruent with the suggestions from Pigot et al 2020 (``\textit{We have demonstrated that a minimum of four independent morphological trait axes are required to predict variation in avian trophic niches, calling into question the validity of trait-based macroecological analyses that assess functional diversity on the basis of fewer morphological trait dimensions}.'').
-%  We have now specified this in the main text:
-%  \textit{We selected the first \(5\) axes to represent the trait spaces that explained at least 95\textbackslash\% of the variance in each specific group (historic and extant species - see supplementary figure 7; following recommendations of \textbackslash citealt\{pigot2020macroevolutionary\} and dimensionality selection method from \textbackslash citealt\{guillerme2023innovation\}).}
-%  Lines @@@
+
+We chose to use five dimensions to describe the Hawaiian birds trait space to make sure that each focal group has at least 95\% of their variance described in the trait space (following Guillerme et al.~2023 protocol).
+Furthermore, the choice of 5 dimensions is congruent with the suggestions from Pigot et al 2020 (``\textit{We have demonstrated that a minimum of four independent morphological trait axes are required to predict variation in avian trophic niches, calling into question the validity of trait-based macroecological analyses that assess functional diversity on the basis of fewer morphological trait dimensions}.'').
+
+We have now specified this in the main text:
+\textit{We selected the first $5$ axes to represent the trait spaces that explained at least 95\% of the variance in each specific group (historic and extant species - see supplementary figure 7; following the recommendations of Pigot et al.
+2020 and the dimensionality selection method from Guillerme et al.
+2023).
+}
+Lines 331-335.
 
@@
 
-% We have now shorten the results to be more descriptive:
+We have now shorten the results to be more descriptive:
 
-% \textit{The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.}
-%  \textit{It ranged from metrics capturing no clear pattern (e.g.~the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g.~Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).}
-%  \textit{Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted into a score increase (e.g.~the functional evenness based on hypervolume for the equalising mechanism) or non-linear responses (e.g.~Alpha diversity based on dendrogram for the equalising mechanism).}
-%  \textit{Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \textbackslash ref\{Fig:simulation\_results\}).}
-%  \textit{Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulation show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).}
-%  Lines @@@
+\textit{The ability of different metrics to capture the different patterns (and thus approximate the mechanisms) was highly variable.
+It ranged from metrics capturing no clear pattern (e.g.
+the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g.
+Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).
+Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted in a score increase (e.g.
+the Functional evenness based on a hypervolume for the equalising mechanism) or non-linear responses (e.g.
+Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores).
+Some cases resulted in an increase in metric score relative to random removals (positive values in Figure 2).
+Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulations show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).}
+Lines 340-353.
 
-% However, we have now greatly expanded our discussion of these results in the discussions section under two subheadings (results, Lines @@@; and caveats Lines @@@)
-% We felt there was a dilemma in 1) expanding the results at the cost of the discussion (i.e.~longer results section and shorter discussion one) which would have highlighted specific results but maybe downplayed some caveats; or 2) expanding the discussion at the cost of the results section which we felt more highlighted what we believe is the main takeaway from the results (``\textit{The ability of different metrics to capture the different patterns (and thus approximate the mechanisms) was highly variable.''}) and expanding where this variability comes from in the discussion (difficulties of interpreting multidimensional spaces + caveats stemming from our methodological choices).
+However, we have now greatly expanded our discussion of these results in the discussions section under two subheadings (``\textbf{Capturing approximated mechanisms with different metrics}'', lines 370-454; and ``\textbf{Caveats}'' lines 456-511).
+We felt there was a dilemma in 1) expanding the results at the cost of the discussion (i.e.~ longer results section and shorter discussion one) which would have highlighted specific results but maybe downplayed some caveats; or 2) expanding the discussion at the cost of the results section which we felt more highlighted what we believe is the main takeaway from the results (``\textit{The ability of different metrics to capture the different patterns (and thus approximate the mechanisms) was highly variable.''}) and expanding where this variability comes from in the discussion (difficulties of interpreting multidimensional spaces + caveats stemming from our methodological choices).
 
@@
 
-% We have now expanded the section in the methods discussing the scaling:
-%  \textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relatively to the same metric scores of a null mechanism.}
-%  \textit{To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between \(0\) and \(1\).}
-%  \textit{We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g.~\(metric\_{stressor20
-% %} \- metric\_{null20
-% %}\)).}
-%  \textit{This resulted in each metric being scaled between \(–1\) and \(1\) where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.~approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near \(0\) meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \textbackslash ref\{Fig:simulations\}, \textbackslash ref\{Fig:simulation\_results\} and \textbackslash ref\{Fig:empirical\_results\}).}
-%  \textit{We interpret this last scenario as the metric score performing poorly, i.e.~it is unable to clearly distinguish between a random removal of elements and a non random one.}
-%  Lines @@@
+We have now expanded the section in the methods discussing the scaling:
+ \textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relative to the same metric scores of a null mechanism.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
+$metric_{stressor20\%} - metric_{null20\%}$).
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.)
+This resulted in each metric being scaled between $-1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
+approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs 1, 2 and 3).
+We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.}
+Lines 273-287.
 
-% As well as updated figure 1 to hopefully improve the readability of the scaled results
-% We believe that they are more interpretable than just displaying the raw results (now also available in the supplementaries), especially since it highlights how some metrics cannot distinguish between the null and the focal mechanism (e.g.~functional evenness using the dendrogram method for the competition) but can still distinguish from the null even when the range of change seems rather small (e.g.~functional evenness using the hypervolume method for competition).
+As well as updated figure 1 to hopefully improve the readability of the scaled results.
+We believe that they are more interpretable than just displaying the raw results (now also available in the supplementaries), especially since it highlights how some metrics cannot distinguish between the null and the focal mechanism (e.g.
+functional evenness using the dendrogram method for the competition) but can still distinguish from the null even when the range of change seems rather small (e.g.
+functional evenness using the hypervolume method for competition).
 
@@
 
-% Although we did wish to expand the discussion with more clear messages of which metrics to use in which circumstances, we do not believe that the findings of our analysis justify prescriptive choice of specific metrics for specific scenarios
-% As highlighted by the reviewer, although some metrics seem to work as better tools to capture changes in some specific scenarios, these are highly dependent on the type of data used here
-% In the example of using the RaoQ having a lower accuracy than other methods, this might be true for this example dataset but we cannot posit that it would be the case for other datasets.
+Although we did wish to expand the discussion with more clear messages of which metrics to use in which circumstances, we do not believe that the findings of our analysis justify prescriptive choice of specific metrics for specific scenarios.
+As highlighted by the reviewer, although some metrics seem to work as better tools to capture changes in some specific scenarios, these are highly dependent on the type of data used here.
+In the example of using the RaoQ having a lower accuracy than other methods, this might be true for this example dataset but we cannot posit that it would be the case for other datasets.
 
-% To convey this message more clearly, we have now entirely restructured the discussion by separating it into sub-sections (simulated results, empirical results and caveats) as well as added a new text box (box 2) summarising how we believe that thinking more carefully about the pattern, process and mechanism allows for a better choice of functional diversity metric.
+To convey this message more clearly, we have now entirely restructured the discussion by separating it into sub-sections (simulated results, empirical results and caveats) as well as added a new text box (box 2) summarising how we believe that thinking more carefully about the pattern, process and mechanism allows for a better choice of functional diversity metric.
 

### 2024-11-29 Analysis/04_Summarise_results.Rmd
added 500 and 1000 elements

@@
 all_results_8d <- read.results("8d")
+
+all_results_2d500 <- read.results("2d500")
+all_results_4d500 <- read.results("4d500")
+all_results_2d1000 <- read.results("2d1000")
+all_results_4d1000 <- read.results("4d1000")
 
@@
 
+all_results_2d500 <- lapply(all_results_2d500, sort.metrics, order = new_order)
+all_results_4d500 <- lapply(all_results_4d500, sort.metrics, order = new_order)
+all_results_2d1000 <- lapply(all_results_2d1000, sort.metrics, order = new_order)
+all_results_4d1000 <- lapply(all_results_4d1000, sort.metrics, order = new_order)
+
 ## The different removal level names
@@
 
+scaled_results_2d500 <- extract.table(all_results_2d500, scale.method = "between")
+scaled_results_4d500 <- extract.table(all_results_4d500, scale.method = "between")
+scaled_results_2d1000 <- extract.table(all_results_2d1000, scale.method = "between")
+scaled_results_4d1000 <- extract.table(all_results_4d1000, scale.method = "between")
+
 ## Extracting all the results but not centred (still scaled)
@@
 raw_results_emp<- extract.table(empirical_results, scale.method = "between", centre = FALSE)
+
+raw_results_2d500 <- extract.table(all_results_2d500, scale.method = "between", centre = FALSE)
+raw_results_4d500 <- extract.table(all_results_4d500, scale.method = "between", centre = FALSE)
+raw_results_2d1000 <- extract.table(all_results_2d1000, scale.method = "between", centre = FALSE)
+raw_results_4d1000 <- extract.table(all_results_4d1000, scale.method = "between", centre = FALSE)
+
 ```
@@
 null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), 4, c(1,3)+4, 4+4), digits = 2)
+
+model_fits_2d500  <- lapply(scaled_results_2d500, fit.model.rm)
+null_differences_2d500  <- lapply(all_results_2d500, paired.tests)
+model_table_2d500  <- fancy.table(model_fits_2d500, columns = list(c(3,4), 5))
+null_table_2d500   <- fancy.table(null_differences_2d500, columns = columns_diff_ses, digits = 2)
+model_fits_4d500  <- lapply(scaled_results_4d500, fit.model.rm)
+null_differences_4d500  <- lapply(all_results_4d500, paired.tests)
+model_table_4d500  <- fancy.table(model_fits_4d500, columns = list(c(3,4), 5))
+null_table_4d500   <- fancy.table(null_differences_4d500, columns = columns_diff_ses, digits = 2)
+
+model_fits_2d1000  <- lapply(scaled_results_2d1000, fit.model.rm)
+null_differences_2d1000  <- lapply(all_results_2d1000, paired.tests)
+model_table_2d1000  <- fancy.table(model_fits_2d1000, columns = list(c(3,4), 5))
+null_table_2d1000   <- fancy.table(null_differences_2d1000, columns = columns_diff_ses, digits = 2)
+model_fits_4d1000  <- lapply(scaled_results_4d1000, fit.model.rm)
+null_differences_4d1000  <- lapply(all_results_4d1000, paired.tests)
+model_table_4d1000  <- fancy.table(model_fits_4d1000, columns = list(c(3,4), 5))
+null_table_4d1000   <- fancy.table(null_differences_4d1000, columns = columns_diff_ses, digits = 2)
+
 ```
@@
 
+```{r}
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_2d500.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_2d500, col = col_vector, lm = model_fits_2d500, null = null_differences_2d500, metric.names = fancy_names, pch = rep(19,4))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_4d500.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_4d500, col = col_vector, lm = model_fits_4d500, null = null_differences_4d500, metric.names = fancy_names, pch = rep(19,4))
+dev.off()
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_2d1000.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_2d1000, col = col_vector, lm = model_fits_2d1000, null = null_differences_2d1000, metric.names = fancy_names, pch = rep(19,4))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_4d1000.pdf", height = 9, width = 9)
+fancy.plot(scaled_results_4d1000, col = col_vector, lm = model_fits_4d1000, null = null_differences_4d1000, metric.names = fancy_names, pch = rep(19,4))
+dev.off()
+```
+
 ## Raw results per metric

### 2024-11-29 Manuscript/supplementaries.tex
added 500 and 1000 elements

@@
 \bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_2d500.pdf}
+\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 500 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
+Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
+Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
+}}
+\label{Fig:simulation_results_2d500}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_2d1000.pdf}
+\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 1000 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
+Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
+Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
+}}
+\label{Fig:simulation_results_2d1000}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d500.pdf}
+\caption{\scriptsize{\textbf{Simulation results for 4 dimensions and 500 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
+Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
+Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
+}}
+\label{Fig:simulation_results_4d500}
+\end{figure}
+\bigskip
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d1000.pdf}
+\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 1000 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
+The different columns represent the different stressors.
+The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
+Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
+The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
+The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
+The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
+Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
+Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
+\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
+}}
+\label{Fig:simulation_results_4d1000}
+\end{figure}
+\bigskip
+
+\subsection{Raw results}
 

### 2024-11-30 Manuscript/ecography_reponse_letter.tex
Update ecography_reponse_letter.tex

@@
 \textcolor{blue}{Line 194-195. Please improve clarity.}
-
-%  @@@TODO: rewrite following other reviewers comments (stats for results)
-
 \textcolor{blue}{Line 201-203. Please simplify and improve clarity.}
-
-%  @@@TODO: rewrite following other reviewers comments (stats for results)
-
 \textcolor{blue}{Line 207-211. Please provide a clearer explanation of the effect size and statistical significance of these model outcomes, similar to the explanation provided in lines 198-201.}
 
-%  @@@TODO: rewrite following other reviewers comments (stats for results)
-%  @@@TODO: calculate effect sizes
-We now added all the statistics results in the supplementary results from table 6 to 18.
+We rewrote this entire section to hopefully improve clarity:
+
+\textit{To understand the ability of each metric to capture a mechanism of interest, we compared their scores relative to the same metric scores of a null mechanism.
+To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first calculated the difference between the stressor of interest and the null stressor (e.g.
+$metric_{stressor20\%} - metric_{null20\%}$).
+We then scaled these differences to be relative to the highest relative differences between all the simulations for that one metric and stressor (across the 4 levels of removal - 20\%, 40\%, etc.)
+This resulted in each metric being scaled between $-1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e.
+approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs 1, 2 and 3).
+We interpret this last scenario as the metric score being unable to clearly distinguish between a random removal of elements and a non random one.
+To assess these poor performances, we compared the distributions of the $171$ scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.
+The results of these tests are displayed in figure 2 (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).}
+Lines 273-290.
+
+Furtheremore, we have also added all the statistics results in the supplementary results from table 6 to 18 (including effect sizes).
 
@@
 
-%  @@@TODO: name of the two added marine species
+We now changed the sentence to:
+\textit{We also added two marine species to the prehistoric species list (}Haliaeetus albicilla\textit{ and }Pterodroma hypoleuca\textit{);}. Lines 309-310.
 
@@
 
-%  @@@TODO: Rerun the simulations with 400, 800 and 1000 elements.
-%  But also highlight that 40 species is already quite big
-% We can also link that to the use of FD in palaeontology (disparity) where 40 is also already quite big!
-% Tom's comment: *Reviewer 2's point about why we use only 200 data points - it's a while since I read our paper, but I guess we did that to make it more realistic in terms of the sample sizes people use.
-In island studies, 200 is pretty large, and I have many references (including my own ...) where people calculate FD using 1-5 species.
-So again we could make the argument that it represents a trade-off, between realistic sample sizes that people analyse vs large enough that we can really test the mechanisms
+We re-ran the simulations with 500 and 1000 observations for 2D and 4D traits (now in the supplementary figures 3 and 4).
+We did not however changed the main results in the manuscript because we believe that datasets with a low number of observations are actually not that uncommon across evolutionary biology (e.g.
+in palaeontology) and ecology.
+For example, in island studies, 200 observations is a pretty large number (e.g. the empirical dataset in this paper).
+We believe that 200 to 40 observations remains a fair trade-off between realistic sample sizes that researchers often analyse (e.g.
+Matthews et al.
+2023) \textit{versus} much larger ones that are sometimes used (e.g.
+Guillerme et al.
+2023).
 
@@
 \textcolor{blue}{LL 170-178: why then did the authors not simply generate more data points (see my comment above)?}
-%  @@@TODO: Rerun the simulations with 400, 800 and 1000 elements.
+
+We have now added the results for 500 and 1000 observations in the supplementary figures 4 and 5.
 

### 2024-11-30 Manuscript/ecography_reponse_letter.tex
final comment

@@
 
-%  @@@TODO: add stats for results (following other reviewers comments)
+We have now added all the statistics in the supplementary materials table 2 to 5 (for the model results) and 6 to 18 (for the metrics differences to the nulls results).
 
@@
 \textcolor{blue}{Line 254-256. Apply the same changes mentioned in the comment below to lines 248-251.}
-%  @@@TODO: add stats for results (following other reviewers comments)
+
+We now refer to the added statistics in the supplementary materials table 2 to 5 (for the model results) and 6 to 18 (for the metrics differences to the nulls results).
 
@@
 \textcolor{blue}{Line 262-269. Please provide more detail regarding these results. Additionally, refer to the table in supplementary material.}
-
-%  @@@TODO: add stats for results (following other reviewers comments)
 
@@
 
-%  @@@TODO: Links back to main comment from reviewer 2: Go back to the previous version of table 1 that gives more description of the models
-% One thing here is probably to justify more why we think that, for example, ``removing points from the side of a traitspace = filtering''.
-%  We now changed the labels in the figure and the caption to highlight that we are indeed approximating such mechanisms, not simulating them per se as the reviewers pointed out.
+We have now improved both the mechanisms approximation description (see comment below) as well as added more justification on why we believe our algorithms approximate the focal mechanisms (see response to the next comment).
+Furthermore, we have also added some general predictions as to which type of metric would capture which type of changes in the trait space (Table 1).
+We hope that these three points improve the clarity of our objectives and justifications.
+We now changed the labels in the figure and the caption to highlight that we are indeed approximating such mechanisms, not simulating them \textit{per se} as the reviewers pointed out.
 

### 2024-11-30 Manuscript/main_ecography_R1.tex
final comment

@@
 tips in a phylogenetic tree, individuals, species, populations, OTUs, etc.
-(see Glossary table in supplementary materials).
+(see Glossary in the supplementary materials, table 1).
 This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
@@
 the Functional evenness based on a hypervolume for the equalising mechanism) or non-linear responses (e.g.
-Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores).
+Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores; see supplementary tables 2 to 5 and 6 to 17).
 Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
@@
 
-For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure \ref{Fig:empirical_results}, Supplementary table 5).
+For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure \ref{Fig:empirical_results}, Supplementary table 5 and 18).
 For example, all richness metrics (green; Figure \ref{Fig:empirical_results}) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.

### 2024-11-30 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 
-\noindent {\normalsize \sc
-Thomas Guillerme$^{1,*}$, 
-Pedro Cardoso$^{2,3}$,
-% Carlos P. Carmona$^{4}$,
-Maria Wagner J\o rgensen$^{4}$,
-Stefano Mammola$^{3,5,6}$,
-Thomas J. Matthews$^{4,7}$}\\
-\noindent {\small \it 
-$^1$School of Biosciences, The University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$CE3C—Centre for Ecology, Evolution and Environmental Changes, CHANGE – Global Change and Sustainability Institute, Faculty of Sciences, University of Lisbon, Lisbon, Portugal\\
-$^3$Laboratory for Integrative Biodiversity Research (LIBRe), Finnish Museum of Natural History (Luomus), University of Helsinki, Helsinki, Finland\\
-% $^4$Institute of Ecology and Earth Sciences, University of Tartu, Tartu, Estonia\\
-$^4$School of Geography, Earth and Environmental Sciences and Birmingham Institute of Forest Research, University of Birmingham, Birmingham B15 2TT, UK\\
-$^5$Molecular Ecology Group (MEG), Water Research Institute, National Research Council (CNR-IRSA), Verbania Pallanza, Italy\\
-$^6$National Biodiversity Future Center, Piazza Marina 61, 90133 Palermo, Italy\\
-$^7$Centre for Ecology, Evolution and Environmental Changes/Azorean Biodiversity Group / CHANGE – Global Change and Sustainability Institute and Universidade dos Açores – Faculty of Agricultural Sciences and Environment; PT-9700-042, Angra do Hero\'ismo, A\c ores, Portugal.\\}
+% \noindent {\normalsize \sc
+% Thomas Guillerme$^{1,*}$, 
+% Pedro Cardoso$^{2,3}$,
+% Maria Wagner J\o rgensen$^{4}$,
+% Stefano Mammola$^{3,5,6}$,
+% Thomas J. Matthews$^{4,7}$}\\
+% \noindent {\small \it 
+% $^1$School of Biosciences, The University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
+% $^2$CE3C—Centre for Ecology, Evolution and Environmental Changes, CHANGE – Global Change and Sustainability Institute, Faculty of Sciences, University of Lisbon, Lisbon, Portugal\\
+% $^3$Laboratory for Integrative Biodiversity Research (LIBRe), Finnish Museum of Natural History (Luomus), University of Helsinki, Helsinki, Finland\\
+% $^4$School of Geography, Earth and Environmental Sciences and Birmingham Institute of Forest Research, University of Birmingham, Birmingham B15 2TT, UK\\
+% $^5$Molecular Ecology Group (MEG), Water Research Institute, National Research Council (CNR-IRSA), Verbania Pallanza, Italy\\
+% $^6$National Biodiversity Future Center, Piazza Marina 61, 90133 Palermo, Italy\\
+% $^7$Centre for Ecology, Evolution and Environmental Changes/Azorean Biodiversity Group / CHANGE – Global Change and Sustainability Institute and Universidade dos Açores – Faculty of Agricultural Sciences and Environment; PT-9700-042, Angra do Hero\'ismo, A\c ores, Portugal.\\}
 
@@
 \bigskip
-\noindent{*\bf Corresponding author.} \textit{t.guillerme@sheffield.ac.uk}\\ 
+% \noindent{*\bf Corresponding author.} \textit{t.guillerme@sheffield.ac.uk}\\ 
 \vspace{1in}

### 2025-01-02 Manuscript/ecography_reponse_letter_2.tex
R2 INIT

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Response to reviewers},
  pdfauthor={Thomas Guillerme},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Response to reviewers}
\author{Thomas Guillerme}

\begin{document}
\maketitle

Dear Editor,

We are grateful for the detailed and useful comments from all two reviewers
We have taken all their suggestions on board, integrated changes associated with the suggestion into the manuscript and responded to them in detail below
For clarity we have displayed the \textcolor{blue}{reviewers' comments in blue} and our own responses in black
We fully describe each change below.
We have also attached a tracked change version of the manuscript to hopefully improve clarifying our changes.

Best regards,

Thomas Guillerme, on behalf of all co-authors.

\textcolor{blue}{Recommendation by the Subject Editor (Dr. Barry Brook):}

% \textcolor{blue}{While Reviewer 1 is largely satisfied with the improvements and only requests minor revisions (e.g., clarifying a few sentences, adding references, and fixing small typographical issues), Reviewer 2 identifies more substantive points that need further attention—particularly around your approach to comparing metrics across different dimensionalities, the conceptual definition of single vs. multiple traits, and the impact of your second scaling step on the observed patterns. Clarifying why and how you simulate trait dimensions and potentially standardising them at four dimensions for TPD comparisons, seem to be important for interpretability.}

% \textcolor{blue}{In your revision,

% - please focus on resolving these methodological and conceptual concerns to improve transparency and strengthen the presentation of your core findings.
% - You will also need to address the referees’ points regarding:
%   - the consistency of dimension usage,
%   - the rationale for simulating “multidimensional” traits,
%   - and the potential biases introduced by your second scaling step.

% Additionally, please follow Reviewer 1’s minor editorial suggestions to refine readability. }

\textcolor{blue}{\textbf{Reviewer: 1}}

\textcolor{blue}{Thank you for dedicating your time and attention to addressing the comments. Your manuscript has significantly improved in readability, content, discussion, and interpretations. The addition of the text box at the end is particularly appreciated, as it effectively highlights the key message of the manuscript. At this stage, I only recommend minor revisions, and after those are addressed, I recommend the manuscript for publication. Below I provided my minor comments and edits to the manuscript.}

\textcolor{blue}{Line 89-90: I think that some references are necessary here.}

\textcolor{blue}{Line 175-177: This sentence is not very clear. I recommend rephrasing.}

\textcolor{blue}{Line 183: Remove “-“.}

\textcolor{blue}{Line 196: I don’t think you need the brackets when citing the function. }

\textcolor{blue}{Line 201: Please explain why you chose p = 3.}

\textcolor{blue}{Line 322: Correct “de Hawaiian species”.}

\textcolor{blue}{Line 339: I would recommend placing figure 2 after the first results paragraph.}

\textcolor{blue}{Figure 3: I would recommend placing figure 3 in the results section. }

\textcolor{blue}{Line 388: Remove “say”.}

\textcolor{blue}{Line 426-432: Please rephrase and simplify.}

\textcolor{blue}{Line 542: Please specify what suggestions you are referring to.}

\textcolor{blue}{\textbf{Reviewer: 3}}

\textcolor{blue}{I greatly appreciate the effort the authors have put into revising the manuscript. The substantial reformulation has significantly improved the understanding of processes, mechanisms, and expectations. I particularly value the expanded explanation of how the stressors are applied. However, I have a few concerns that I believe should be addressed to further enhance the clarity and methodological rigor of the manuscript. First, the comparison of metrics across different dimensionalities raises questions, particularly when metrics from lower-dimensional spaces (e.g., the TPD package) are compared to those measured in higher-dimensional spaces. Given the manuscript's aim to compare metrics measured in a standardized way, ensuring consistency across dimensions is fundamental. Additionally, this version of the manuscript has highlighted concerns regarding the simulation of traits—specifically, whether the approach involves a single trait with multiple dimensions or multiple independent traits. This distinction has important implications for the assumptions of independence and variance, and subsequently, for the simulated patterns within the trait space. Below, I provide specific comments and suggestions to address these points. I will recommend a further revision to clarify the selection and implementation of these approaches, as this will enhance the transparency and methodological rigor of the manuscript.}

\textcolor{blue}{Specific comments: }

\textcolor{blue}{Lines 128-135: The capacity of the algorithm to simulate the stressor strongly depends on the distribution of the data and the assumption of independence in the simulated dimensions (as should be the case in the simulation example). Given this, it is unclear why a single trait is simulated as multidimensional. This approach feels unnecessarily complex, potentially confounding the reader and introducing covariation that should be tested. To clarify, a trait is typically defined as a single measurable characteristic of an organism, particularly when considering the definition referenced in Violle et al. (2007). The concept of multidimensionality in trait spaces usually applies when multiple traits are combined into a matrix, with each column representing a distinct trait or a principal component summarizing trait variation. However, it is not common practice to describe a single trait using multiple variables (i.e., columns or dimensions). Regarding the statement in Lines 132–135:
“These aspects are often expressed in one dimension but can in fact be described in any number of dimensions. For example, for trait defined as “leaf insertion angle”, this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).”
While this explanation clarifies the authors’ definition of a “multidimensional trait,” it appears to complicate the methodology. If traits are often expressed in a single dimension (e.g., angle in degrees), I do not see the necessity to simulate them as “a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.” (lines 128-129 If I understood the text correctly, this approach results in a trait space built around a single trait defined by multiple dimensions. Instead, it would be more intuitive and broadly applicable to simulate 2, 4, or 8 separate traits. These traits, together, would define a 2-, 4-, or 8-dimensional trait space. Since these traits would be uncorrelated, they would correspond directly to the same number of principal components, aligning with the definition of dimensions provided in Supplementary Table 1.
Moreover, simulating a single multidimensional trait may inadvertently introduce covariation across dimensions. For example, using “leaf insertion angle” as an illustration, the trigonometric relationships of the triangle’s sides and angles inherently interconnect mathematically (e.g., through sine, cosine, and tangent). This interconnection could introduce correlation in the variance, even under Brownian assumptions. While this correlation may not be inherently incorrect or distinct from real ecological data, the assumption of independence in the simulation should be explicitly tested. Correlated dimensions might create clustering in the trait space, potentially biasing the application of stressors in the simulation context. 
Finally, simulating independent traits using Brownian motion without fixing the variance could also address the issue of homogeneity of variance across dimensions (Lines 463–464). Homogeneity of variance may introduce similar bias in the simulated mechanisms and it is far from approximating a set of independent trait or dimensions. I belive that simulating independent trait may mitigate that concern.}

\textcolor{blue}{Lines 191 – 209: This paragraph includes parts that require clarification. It begins by stating that the 'algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations', which corresponds to the equation ix(1-b)p.  However, if this algorithm is intended to approximate a competition mechanism ‘where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.’ and ‘We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination’ shouldn’t the pattern be the reverse? Furthermore, the formula, by subtracting 1 from the probability, seems to skew sampling towards less populated regions of the space, which, if I understand correctly, does not align with the concept of competitive exclusion.}

\textcolor{blue}{Lines 280-289: It seems that I may need further clarification regarding the second step of the scaling. If the data are only divided by the maximum value, the results would generally have a maximum of 1 in cases of consistently negative or positive metrics across all levels of species exclusion (e.g., -2/-2 = 1 and +2/+2 = 1). Similarly, when some metrics have negative values and others are positive (across different levels of removal), the range would not necessarily be constrained between -1 and 1 unless all values fall within the range [0, 1] (e.g., functional richness might not range between -1 and 1). Furthermore, I am concerned that this second scaling could reduce or bias the patterns of the metrics, potentially affecting the conclusions of the manuscript. For instance, in Supplementary Figure 8 (simulated data for 4 dimensions), taking FD:FEve for facilitation as a reference, subtracting the MetricStressor from the MetricNull results in a relatively linear pattern, with a general decrease in evenness as species removal increases (if I am seeing well from the figure), which aligns with expectations under this mechanism. However, in Figure 2, the same metric exhibits a more drastic decrease in stressor evenness at 60\% and an increase at 80\%, challenging to reconcile with the expected mechanism, given that the data were simulated to approximate a facilitation mechanism. A similar trend can be observed across other examples (e.g., Supplementary Figures with raw data), which may explain the contrasting patterns between historical and extant empirical data, with some metrics showing fewer extreme differences than observed. I suggest further exploring the impact of this second scaling on the final results.}

\textcolor{blue}{Figure 1: To enhance readability, I suggest including a legend for the colours (orange and blue) directly in the figure, not just in the caption. I appreciate the improvements made to expand the figure. However, if metrics C to F are scaled for the null mechanisms, I assume that in the first two panels of section G, the term 'scaled' could be removed, as they likely represent the actual metric used to compute the scaled metric in panel 3. If the metrics are further scaled, this should be clarified in the text. For completeness and readability, I also suggest explicitly adding the details of the second scaling performed relative to the four levels of removal.}

\textcolor{blue}{Lines 330 – 335/ Figure 4: I appreciate the clarification regarding the TPD metric. However, I regret to say that comparing a metric built on four dimensions with one working on five dimensions does not seem appropriate, particularly given the aims of the manuscript. Considering the amount of variation explained by the first four dimensions (Supplementary Figure 11) and the manuscript's focus on exploring metric performance with a real dataset rather than depicting bird diversity after extinction, standardizing the dimensionality across all metrics is necessary.  This approach would enable a fair comparison, even if it requires a slight compromise on the 95\% variation threshold, targeting approximately 90\% instead (as observed in the histogram for four dimensions, which aligns with the minimum number of dimensions recommended by Pigot et al. 2020). This adjustment is particularly important, as metric performance may vary significantly across different dimensionalities (e.g., metrics evaluated across 2, 4, or 8 dimensions). Standardizing to four dimensions would also ensure consistency with the dimensionality used in the simulated dataset and maintain uniformity throughout the analyses. Regarding the 8-dimensional space, since it is not central to the analysis, I suggest removing the TPD approach for this case to avoid introducing unnecessary variability, while clearly justifying this decision as a limit of the package in the text.
Furthermore, upon focusing on Supplementary Figure 11, I have a concern. Based on the description in the main text, I understand that the analyses are performed starting from a single trait space (comprising 118 points), where species are removed either randomly or based on actual extinction events, and metrics are then recalculated. This approach seems appropriate for evaluating how species are distributed and how metric performance changes under different scenarios. However, the figure comparing extant and pre-1500 species leads me to wonder whether the trait spaces were recalculated based on the subset of species following extinctions. Am I correct in assuming that the analyses are based on a single trait space constructed with all 118 points?}

\textcolor{blue}{Discussion: While I appreciate the revisions and the inclusion of the Caveats section, I feel that Box 2 might not add substantial value, as it primarily refers to previously published materials rather than the manuscript's main findings. Expanding the Results/Discussion section instead could be more beneficial. For instance, the simulations were performed across 2, 4, and 8 dimensions, and the metrics and their relationships seem to vary across these spaces with different dimensionalities. Addressing this aspect in greater detail would add significant value to the manuscript and better align with its stated aims.}

\textcolor{blue}{Supplementary Table 1: This table is a valuable addition to the manuscript. However, geographical location or landmark coordinates should not be considered a trait of an observation, particularly when applying the definition of a trait as used throughout the text (e.g., Violle et al. 2007 and others). Additionally, I suggest revising the terminology in the same column to avoid defining a 'trait' as a 'trait'; using terms like 'characteristic' or 'aspect of observations' might align better with the main text's definitions. Finally, I recommend considering the inclusion of this table in the main text, as it enhances the reader's understanding of the concepts presented.}

\textcolor{blue}{Supplementary Figures 7–10: I appreciate the addition of the new figures showing the raw metrics. However, given the range of the values (particularly for functional richness) and the legend indicating that the metrics are scaled, it is unclear to me what the values are referencing. I suggest not scaling the metrics in this case, as they should naturally correspond to the same units as the null values, ensuring clarity and consistency. This adjustment would enhance the interpretability of the data.}

\textcolor{blue}{Supplementary Tables 6–18: I suggest including the measured metrics in addition to the differences currently reported. Additionally, I noticed that some mechanisms (e.g., Tables 8, 10, 12, 14, and 16) show differences that warrant double-checking. For example, if the reported values are raw, the differences for functional divergence with TPD should be constrained between 0 and 1, as the metric itself is bounded within this range. However, some values in the tables exceed this limit. Finally, could you clarify how the SES is calculated in these cases?}

\textcolor{blue}{Lastly, I recommend a thorough review of the text, as I noticed several instances of missing punctuation throughout.}

\end{document}

### 2025-01-02 Manuscript/ecography_reponse_letter_2.tex
Update ecography_reponse_letter_2.tex

@@
 This approach feels unnecessarily complex, potentially confounding the reader and introducing covariation that should be tested.
-To clarify, a trait is typically defined as a single measurable characteristic of an organism, particularly when considering the definition referenced in Violle et al.
-(2007).
+To clarify, a trait is typically defined as a single measurable characteristic of an organism, particularly when considering the definition referenced in Violle et al. (2007).
 The concept of multidimensionality in trait spaces usually applies when multiple traits are combined into a matrix, with each column representing a distinct trait or a principal component summarizing trait variation.
 However, it is not common practice to describe a single trait using multiple variables (i.e., columns or dimensions).
+
 Regarding the statement in Lines 132–135:
-“These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
-For example, for trait defined as “leaf insertion angle”, this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).”
+\textit{“These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
+For example, for trait defined as “leaf insertion angle”, this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).”}
 While this explanation clarifies the authors’ definition of a “multidimensional trait,” it appears to complicate the methodology.
 If traits are often expressed in a single dimension (e.g., angle in degrees), I do not see the necessity to simulate them as “a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.” (lines 128-129 If I understood the text correctly, this approach results in a trait space built around a single trait defined by multiple dimensions.
+
 Instead, it would be more intuitive and broadly applicable to simulate 2, 4, or 8 separate traits.
@@
 
+% TODO:  Change the trait definition from single to multiple.
+And say it's because in treats "traits" are treated as processes in n-dimensions but that we agree that this is adds extra complication here.
+
 \textcolor{blue}{Lines 191 – 209: This paragraph includes parts that require clarification.
@@
 Furthermore, the formula, by subtracting 1 from the probability, seems to skew sampling towards less populated regions of the space, which, if I understand correctly, does not align with the concept of competitive exclusion.}
+
+% TODO: double check the definition and make sure it aligns with the correct one. Say we fucked up explanation because of the negative effect (i.e. removing X is the same as keeping not X).
 
@@
 
+% TODO: re double checking how the scaling works and what it does to the data!
+
 \textcolor{blue}{Figure 1: To enhance readability, I suggest including a legend for the colours (orange and blue) directly in the figure, not just in the caption.
@@
 For completeness and readability, I also suggest explicitly adding the details of the second scaling performed relative to the four levels of removal.}
+
+% TODO: add colour legend to the plot
+% TODO: distinguish between the two scalings in the figure
+% TODO: expand scaling explanation in the caption
 
@@
 
+% TODO: clarify the artificial extinctions in the birds dataset (yes, we always start from 118 points).
+% TODO: Remove TPD metric altogether when it's not calculable?
+% TODO: double check if that's what the reviewer means.
+
 \textcolor{blue}{Discussion: While I appreciate the revisions and the inclusion of the Caveats section, I feel that Box 2 might not add substantial value, as it primarily refers to previously published materials rather than the manuscript's main findings.
@@
 Addressing this aspect in greater detail would add significant value to the manuscript and better align with its stated aims.}
+
+% TODO: say we're keeping Box 2 because.
+% TODO: but also develop points they want.
 
@@
 
+% TODO: add the table to the main text
+% TODO: say we keep our definition of a trait as "measurable characteristic" (but that we use the more common defintion for the simulations)
+
 \textcolor{blue}{Supplementary Figures 7–10: I appreciate the addition of the new figures showing the raw metrics.
@@
 This adjustment would enhance the interpretability of the data.}
+
+% TODO: redo the supplementary plots but with different scale bars for every plot (and don't do the scaling)
 
@@
 
+% TODO: double check what's in the tables
+% TODO: in caption, explain how SES is calculated (reference the R package)
+
 \textcolor{blue}{Lastly, I recommend a thorough review of the text, as I noticed several instances of missing punctuation throughout.}

### 2025-01-06 Manuscript/ecography_reponse_letter_2.tex
up reviewer 1

@@
 \textcolor{blue}{Line 89-90: I think that some references are necessary here.}
+We changed this sentence and merged it with the next one after discovering some previously missed literature:
+
+\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g.
+\cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines @@@.
 
@@
 
+We rephrased the sentence to:
+
+\textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e.
+observations adjacent to each other in the trait space are more likely compared to common ones).} Lines @@@
+
 \textcolor{blue}{Line 183: Remove “-“.}
+
+We replaced the dash by a comma.
 
@@
 
+We removed the brackets.
+
 \textcolor{blue}{Line 201: Please explain why you chose p = 3.}
+
+We've now explained why we chose p=3:
+
+\textit{here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D} Lines @@@.
 
@@
 
+We fixed the type.
+
 \textcolor{blue}{Line 339: I would recommend placing figure 2 after the first results paragraph.}
+\textcolor{blue}{Figure 3: I would recommend placing figure 3 in the results section. }
 
-\textcolor{blue}{Figure 3: I would recommend placing figure 3 in the results section. }
+The placing of figures is done automatically here by the compiler.
+But we'll make sure to forward this suggestion to the copy editors after the manuscript is accepted.
 
@@
 
+We remove the word ``say''.
+
 \textcolor{blue}{Line 426-432: Please rephrase and simplify.}
 
+We've now simplified the sentence to:
+
+\textit{Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
+This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
+These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines @@@
+
 \textcolor{blue}{Line 542: Please specify what suggestions you are referring to.}
+
+We've now specified the following:
+
+\textit{For example, we can follow the suggestions from \cite{mammola2021concepts} highlighting which kind of methodological pipeline best supports which type of metric of interest (Table 3).} Lines @@@
 
@@
 The concept of multidimensionality in trait spaces usually applies when multiple traits are combined into a matrix, with each column representing a distinct trait or a principal component summarizing trait variation.
-However, it is not common practice to describe a single trait using multiple variables (i.e., columns or dimensions).
+However, it is not common practice to describe a single trait using multiple variables (i.e., columns or dimensions).}
 
-Regarding the statement in Lines 132–135:
+\textcolor{blue}{Regarding the statement in Lines 132–135:
 \textit{“These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
@@
 While this explanation clarifies the authors’ definition of a “multidimensional trait,” it appears to complicate the methodology.
-If traits are often expressed in a single dimension (e.g., angle in degrees), I do not see the necessity to simulate them as “a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.” (lines 128-129 If I understood the text correctly, this approach results in a trait space built around a single trait defined by multiple dimensions.
+If traits are often expressed in a single dimension (e.g., angle in degrees), I do not see the necessity to simulate them as “a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.” (lines 128-129 If I understood the text correctly, this approach results in a trait space built around a single trait defined by multiple dimensions.}
 
-Instead, it would be more intuitive and broadly applicable to simulate 2, 4, or 8 separate traits.
+\textcolor{blue}{Instead, it would be more intuitive and broadly applicable to simulate 2, 4, or 8 separate traits.
 These traits, together, would define a 2-, 4-, or 8-dimensional trait space.

### 2025-01-06 Manuscript/main_ecography_R2.tex
R2 INIT
up reviewer 1

--- 2024-11-30 Manuscript/main_ecography_R1.tex
@@
 
-While the first part of this framework (``what?'') has been studied extensively in the last decade, this was often undertaken outside of the ``what, how and why'' framework.
-Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
+Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g.
+\cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.
 The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
@@
 We chose this algorithm to approximate the facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
-I.e. pairs of observations that share similar trait combinations are more likely to remain (i.e. observations adjacent to each other in the trait space).
+In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e.
+observations adjacent to each other in the trait space are more likely compared to common ones).
 
@@
 This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
-In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
+In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions, in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
 This approximates our \textbf{filtering mechanism}.
@@
 This is done by selecting $B$ discrete categories to summarise the distribution - i.e.
-bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0()} function in \texttt{R};][]{silverman1986density}.
+bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0} function in \texttt{R};][]{silverman1986density}.
 Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$.
-Where $p$ is a factor increasing the scaling power of the algorithm (here we used $p=3$).
+Where $p$ is a factor increasing the scaling power of the algorithm (here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D).
 In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
@@
 $55$ species are known to have gone extinct prior to $1500$ CE and $26$ after $1500$ CE.
-Note that we used the extinct species data as presented in \cite{matthews2023global}, and thus we do not include de Hawaiian species declared extinct by the IUCN in $2024$.
+Note that we used the extinct species data as presented in \cite{matthews2023global}, and thus we do not include Hawaiian species declared extinct by the IUCN in $2024$.
 For our analyses, we focused on two time periods: (i) the avifauna present in $1500$ CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
@@
 For example, say one is interested in looking at changes in functional diversity using a divergence metric calculated with the probability density method (e.g. \texttt{TPD::REND}).
-They can apply this metric to their trait space and get two relative metric scores of, say, $0.7$ and $0.65$ (pattern) for two levels of extinction (process).
+They can apply this metric to their trait space and get two relative metric scores of $0.7$ and $0.65$ (pattern) for two levels of extinction (process).
 The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.
@@
 
-Our results indicating that the loss of functional diversity (based on several metrics) was not more than expected by chance when focusing on all extinctions could be due to the fact that although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines (approximatively $58$\%), the majority of global known anthropogenic extinctions (approx.
-$75$\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
+Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
+This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
+These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
 However, in Hawaii, where the same proportion of known species were passerines (approx. $58$\%), the opposite is true: the majority of extinctions (approx. $60$\%) have been of passerine species.
@@
 Once the one has clearly defined the pattern they want to capture (\textbf{what}), we believe it gets easier to pick up the type of metric that would be the most adequate.
-For example, we can follow the suggestions from \cite{mammola2021concepts} (Table \ref{Tab:box2}).
+For example, we can follow the suggestions from \cite{mammola2021concepts} highlighting which kind of methodological pipeline best supports which type of metric of interest (Table \ref{Tab:box2}).
 
@@
 \end{table}
-
-%TG: TODO: check the pre-1500 story from email from Tom and Maria
 

### 2025-01-07 Manuscript/ecography_reponse_letter_2.tex
Update ecography_reponse_letter_2.tex

@@
 
-% TODO:  Change the trait definition from single to multiple.
-And say it's because in treats "traits" are treated as processes in n-dimensions but that we agree that this is adds extra complication here.
+We have changed the description of the trait to multiple uni-dimensional traits.
+We used (and maybe insisted on) the previous definition of a single uncorrelated multidimensional trait because the way they can be defined in the \texttt{treats} package (a trait = any number of dimensions as long as the process is the same).
+We changed the description to the following:
+
+\textit{First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
+a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
+We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).} Lines @@@
+
+However, we kept the broader definition of trait.
+\citealt{violle2007let} provided an ``unambiguous definition of plant trait, with a particular emphasis on functional trait'', which as per the authors aims was proposed to clarify and and conceptualise the definition on functional traits (``a surrogate of organismal performance'') specifically in plant ecology.
+Here we build on \citealt{mcgill2006rebuilding}'s more generalised approach, and, more interestingly on the work of \citealt{dawson2021traits} of how ecologists _actually_ operationalise the concept of trait based on researchers understanding vs.
+proposed definitions.
+We propose through the paragraph on lines @@@ to expand that to ``any measurable aspect of an organism'' which we hope allows expansion beyond one dimensional traits and, more importantly, beyond ecology.
+I.e.
+we believe that this definition of trait more accurately also includes how traits are used in microbiology (e.g.
+bacterial colony shape), macroevolution (e.g.
+3D geometric morphometrics) or in palaeontology (e.g.
+``position of a specific foramen in a skull relative to another feature'').
+
+% TODO: also define this in the new in-text glossary table.
 

### 2025-01-07 Manuscript/main_ecography_R2.tex
Update main_ecography_R2.tex

@@
 
-First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
+First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
 a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
-This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
+We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).
 Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive).

### 2025-01-09 Manuscript/ecography_reponse_letter_2.tex
added glossary table

@@
 
-% TODO: add the table to the main text
-% TODO: say we keep our definition of a trait as "measurable characteristic" (but that we use the more common defintion for the simulations)
+We've now added the the glossary table (supplementary table 1) to the main text as suggested with the definition of trait redefined as a characteristic (good catch!).
+We did however keep our broader definition of a trait as opposed to \cite{violle2007let}'s (see response to first comment of Reviewer 3 above).
+We have added this precision in the table:
+
+\textit{Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).} Lines @@@
 

### 2025-01-09 Manuscript/main_ecography_R2.tex
added glossary table

@@
 We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.
+
+\begin{table}[ht]
+\caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
+\centering
+\begin{tabular}{p{2.5cm}p{4cm}p{5cm}p{5cm}}
+Term in this paper & definition & in ecology & in macroevolution \\
+\hline
+Trait space & Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, traitspace, etc.  \\
+\hline
+Observations & Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs etc. \\
+\hline
+Dimensions & Columns ($d$) & Traits, Ordination scores, distances, etc. & Traits, ordination scores, distances, etc. \\
+\hline
+Traits & Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional characteristic*; e.g. geographical location & uni/multidimensional characteristic; e.g. landmark coordinates \\
+\hline
+Group & Rows subset ($m \times d$; $m \leq n$) & Treatments, phylogenetic group (clade), etc. & Clades, geological stratum, etc. \\
+\hline
+Metric & Statistic (i.e. a measure) & Dissimilarity index or metric, hypervolume, functional diversity, etc. & Disparity metric or index \\
+\hline
+Stressors & An algorithm removing a proportion of the observations & Introduction of invasive species, change of landscape use, climate change, pollution, experimental design, etc.
+& Mass extinction, Tectonic change, climate change, etc.
+\\
+\hline
+Stressors' intensity & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction,  etc. \\
+\hline
+\end{tabular}
+* Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).
+\end{table}
 

### 2025-01-09 Manuscript/supplementaries.tex
added glossary table

@@
 %---------------------------------------------
-
-\section{Supplementary materials}
-
-\subsection{Glossary}
-
-\begin{table}[ht]
-\caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
-\centering
-\begin{tabular}{p{2.5cm}p{4cm}p{5cm}p{5cm}}
-Term in this paper & definition & in ecology & in macroevolution \\
-\hline
-Trait space & Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, traitspace, etc.  \\
-\hline
-Observations & Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs etc. \\
-\hline
-Dimensions & Columns ($d$) & Traits, Ordination scores, distances, etc. & Traits, ordination scores, distances, etc. \\
-\hline
-Traits & Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional trait; e.g. geographical location & uni/multidimensional trait; e.g. landmark coordinates \\
-\hline
-Group & Rows subset ($m \times d$; $m \leq n$) & Treatments, phylogenetic group (clade), etc. & Clades, geological stratum, etc. \\
-\hline
-Metric & Statistic (i.e. a measure) & Dissimilarity index or metric, hypervolume, functional diversity, etc. & Disparity metric or index \\
-\hline
-Stressors & An algorithm removing a proportion of the observations & Introduction of invasive species, change of landscape use, climate change, pollution, experimental design, etc.
-& Mass extinction, Tectonic change, climate change, etc.
-\\
-\hline
-Stressors' intensity & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction,  etc. \\
-\end{tabular}
-\end{table}
-
-\newpage
 

### 2025-01-14 Analysis/04_Summarise_results.Rmd
tmp UP
need to finish the plotting bit

--- 2024-11-29 Analysis/04_Summarise_results.Rmd
@@
 ## Extracting all the scaled results
-scaled_results_2d <- extract.table(all_results_2d, scale.method = "between")
-scaled_results_3d <- extract.table(all_results_3d, scale.method = "between")
-scaled_results_4d <- extract.table(all_results_4d, scale.method = "between")
-scaled_results_5d <- extract.table(all_results_5d, scale.method = "between")
-scaled_results_8d <- extract.table(all_results_8d, scale.method = "between")
-scaled_results_emp<- extract.table(empirical_results, scale.method = "between")
+scaled_cent_results_2d <- extract.table(all_results_2d, scale.method = "between")
+scaled_cent_results_3d <- extract.table(all_results_3d, scale.method = "between")
+scaled_cent_results_4d <- extract.table(all_results_4d, scale.method = "between")
+scaled_cent_results_5d <- extract.table(all_results_5d, scale.method = "between")
+scaled_cent_results_8d <- extract.table(all_results_8d, scale.method = "between")
+scaled_cent_results_emp<- extract.table(empirical_results, scale.method = "between")
+
+scaled_cent_results_2d500 <- extract.table(all_results_2d500, scale.method = "between")
+scaled_cent_results_4d500 <- extract.table(all_results_4d500, scale.method = "between")
+scaled_cent_results_2d1000 <- extract.table(all_results_2d1000, scale.method = "between")
+scaled_cent_results_4d1000 <- extract.table(all_results_4d1000, scale.method = "between")
 
 ## Extracting all the results but not centred (still scaled)
-raw_results_2d <- extract.table(all_results_2d, scale.method = "between", centre = FALSE)
-raw_results_3d <- extract.table(all_results_3d, scale.method = "between", centre = FALSE)
-raw_results_4d <- extract.table(all_results_4d, scale.method = "between", centre = FALSE)
-raw_results_5d <- extract.table(all_results_5d, scale.method = "between", centre = FALSE)
-raw_results_8d <- extract.table(all_results_8d, scale.method = "between", centre = FALSE)
-raw_results_emp<- extract.table(empirical_results, scale.method = "between", centre = FALSE)
+scaled_results_2d <- extract.table(all_results_2d, scale.method = "between", centre = FALSE)
+scaled_results_3d <- extract.table(all_results_3d, scale.method = "between", centre = FALSE)
+scaled_results_4d <- extract.table(all_results_4d, scale.method = "between", centre = FALSE)
+scaled_results_5d <- extract.table(all_results_5d, scale.method = "between", centre = FALSE)
+scaled_results_8d <- extract.table(all_results_8d, scale.method = "between", centre = FALSE)
+scaled_results_emp<- extract.table(empirical_results, scale.method = "between", centre = FALSE)
+
+scaled_results_2d500 <- extract.table(all_results_2d500, scale.method = "between", centre = FALSE)
+scaled_results_4d500 <- extract.table(all_results_4d500, scale.method = "between", centre = FALSE)
+scaled_results_2d1000 <- extract.table(all_results_2d1000, scale.method = "between", centre = FALSE)
+scaled_results_4d1000 <- extract.table(all_results_4d1000, scale.method = "between", centre = FALSE)
+
+## Extracting all the results not scaled
+raw_results_2d <- extract.table(all_results_2d, scale = FALSE, centre = FALSE)
+raw_results_3d <- extract.table(all_results_3d, scale = FALSE, centre = FALSE)
+raw_results_4d <- extract.table(all_results_4d, scale = FALSE, centre = FALSE)
+raw_results_5d <- extract.table(all_results_5d, scale = FALSE, centre = FALSE)
+raw_results_8d <- extract.table(all_results_8d, scale = FALSE, centre = FALSE)
+raw_results_emp<- extract.table(empirical_results, scale = FALSE, centre = FALSE)
+
+raw_results_2d500 <- extract.table(all_results_2d500,scale = FALSE, centre = FALSE)
+raw_results_4d500 <- extract.table(all_results_4d500, scale = FALSE, centre = FALSE)
+raw_results_2d1000 <- extract.table(all_results_2d1000, scale = FALSE, centre = FALSE)
+raw_results_4d1000 <- extract.table(all_results_4d1000, scale = FALSE, centre = FALSE)
+
 ```
@@
 ## Slope and R2 for each model
-model_fits_2d  <- lapply(scaled_results_2d, fit.model.rm)
-model_fits_3d  <- lapply(scaled_results_3d, fit.model.rm)
-model_fits_4d  <- lapply(scaled_results_4d, fit.model.rm)
-model_fits_5d  <- lapply(scaled_results_5d, fit.model.rm)
-model_fits_8d  <- lapply(scaled_results_8d, fit.model.rm)
-model_fits_emp <- lapply(scaled_results_emp, fit.model.rm)
+model_fits_2d  <- lapply(scaled_cent_results_2d, fit.model.rm)
+model_fits_3d  <- lapply(scaled_cent_results_3d, fit.model.rm)
+model_fits_4d  <- lapply(scaled_cent_results_4d, fit.model.rm)
+model_fits_5d  <- lapply(scaled_cent_results_5d, fit.model.rm)
+model_fits_8d  <- lapply(scaled_cent_results_8d, fit.model.rm)
+model_fits_emp <- lapply(scaled_cent_results_emp, fit.model.rm)
 
@@
 null_table_emp  <- fancy.table(null_differences_emp, columns = list(c(1,3), 4, c(1,3)+4, 4+4), digits = 2)
+
+model_fits_2d500  <- lapply(scaled_cent_results_2d500, fit.model.rm)
+null_differences_2d500  <- lapply(all_results_2d500, paired.tests)
+model_table_2d500  <- fancy.table(model_fits_2d500, columns = list(c(3,4), 5))
+null_table_2d500   <- fancy.table(null_differences_2d500, columns = columns_diff_ses, digits = 2)
+model_fits_4d500  <- lapply(scaled_cent_results_4d500, fit.model.rm)
+null_differences_4d500  <- lapply(all_results_4d500, paired.tests)
+model_table_4d500  <- fancy.table(model_fits_4d500, columns = list(c(3,4), 5))
+null_table_4d500   <- fancy.table(null_differences_4d500, columns = columns_diff_ses, digits = 2)
+
+model_fits_2d1000  <- lapply(scaled_cent_results_2d1000, fit.model.rm)
+null_differences_2d1000  <- lapply(all_results_2d1000, paired.tests)
+model_table_2d1000  <- fancy.table(model_fits_2d1000, columns = list(c(3,4), 5))
+null_table_2d1000   <- fancy.table(null_differences_2d1000, columns = columns_diff_ses, digits = 2)
+model_fits_4d1000  <- lapply(scaled_cent_results_4d1000, fit.model.rm)
+null_differences_4d1000  <- lapply(all_results_4d1000, paired.tests)
+model_table_4d1000  <- fancy.table(model_fits_4d1000, columns = list(c(3,4), 5))
+null_table_4d1000   <- fancy.table(null_differences_4d1000, columns = columns_diff_ses, digits = 2)
+
 ```
@@
 ## The plot
-fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4), xlab = "scaled metric change")
+fancy.plot(scaled_cent_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4), xlab = "scaled metric change", xlim = c(-1, 1))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_2d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_2d, col = col_vector, lm = model_fits_2d, null = null_differences_2d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_3d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_3d, col = col_vector, lm = model_fits_3d, null = null_differences_3d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_4d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_4d, col = col_vector, lm = model_fits_4d, null = null_differences_4d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_5d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_8d.pdf", height = 9, width = 9)
-fancy.plot(scaled_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4))
+fancy.plot(scaled_cent_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17))
+fancy.plot(scaled_cent_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17), xlim = c(-1, 1))
 
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 11, width = 3.5)
-fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17))
+fancy.plot(scaled_cent_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17), xlim = c(-1, 1))
 dev.off()
@@
 
+```{r}
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_2d500.pdf", height = 9, width = 9)
+fancy.plot(scaled_cent_results_2d500, col = col_vector, lm = model_fits_2d500, null = null_differences_2d500, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_4d500.pdf", height = 9, width = 9)
+fancy.plot(scaled_cent_results_4d500, col = col_vector, lm = model_fits_4d500, null = null_differences_4d500, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
+dev.off()
+
+## Pdf version
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_2d1000.pdf", height = 9, width = 9)
+fancy.plot(scaled_cent_results_2d1000, col = col_vector, lm = model_fits_2d1000, null = null_differences_2d1000, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_4d1000.pdf", height = 9, width = 9)
+fancy.plot(scaled_cent_results_4d1000, col = col_vector, lm = model_fits_4d1000, null = null_differences_4d1000, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
+dev.off()
+```
+
 ## Raw results per metric
@@
 
-pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_2d.pdf", height = 11, width = 8.5)
-fancy.plot(raw_results_2d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
+pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_2d.pdf", height = 11, width = 8.5)
+fancy.plot(scaled_results_2d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
 dev.off()
 
-pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_4d.pdf", height = 11, width = 8.5)
-fancy.plot(raw_results_4d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
+pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_4d.pdf", height = 11, width = 8.5)
+fancy.plot(scaled_results_4d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
 dev.off()
 
-pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_8d.pdf", height = 11, width = 8.5)
-fancy.plot(raw_results_8d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
+pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_8d.pdf", height = 11, width = 8.5)
+fancy.plot(scaled_results_8d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
 dev.off()
 
-pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_emp.pdf", height = 11, width = 5.5)
-fancy.plot(raw_results_emp, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = c(19,15,17), xlab = "scaled metric", xlim = c(0,1))
+pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_emp.pdf", height = 11, width = 5.5)
+fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = c(19,15,17), xlab = "scaled metric", xlim = c(0,1))
 dev.off()
+```
+
+## Raw results per metric
+
+```{r}
+# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_2d.pdf", height = 11, width = 8.5)
+
+# layout(matrix(1:(12*5), 12, 5))
+# for(stressor in 1:5) {
+#     ## For each stressor
+#     plotting_data <- raw_results_2d[[stressor]]
+#     ## Plot the metric
+#     for(one_metric in 1:(ncol(plotting_data)/4)) {
+#         ## Get the metric name
+#         metric_name <- strsplit(colnames(selected_data), split = "_rm")[[1]][1]
+#         ## Get the data
+#         selected_data <- plotting_data[, 1:4]
+
+#         ## Get the color TODO + add transparent
+#         col <- rep("orange", 4)
+
+#         ## Empty plot
+#         plot(NULL, ylim = c(0, 5), xlim = quantile(c(selected_data), prob =  c(0.025, 0.975)), yaxt = "n", ylab = "", xlab = metric_name)
+
+#         ## Add the lines
+#         for(one_line in 1:4) {
+#             ## Get the CIs and median
+#             CIs <- quantile(selected_data[, one_line], prob = c(0.025, 0.25, 0.75, 0.975))
+#             med <- median(selected_data[, one_line])
+#             ## Plot the CIs
+#             lines(y = rep(5-one_line, 2), x = CIs[c(1, 4)], lty = 3, col = col)
+#             lines(y = rep(5-one_line, 2), x = CIs[c(2, 3)], lty = 1, col = col)
+#             ## Plot the median
+#             points(y =5-one_line, x = med, pch = 19, col = col)
+#         }
+#     }
+# }
+
+# fancy.plot(raw_results_2d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "raw metric")
+# dev.off()
+
+# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_4d.pdf", height = 11, width = 8.5)
+# fancy.plot(raw_results_4d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "raw metric")
+# dev.off()
+
+# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_8d.pdf", height = 11, width = 8.5)
+# fancy.plot(raw_results_8d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "raw metric")
+# dev.off()
+
+# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_emp.pdf", height = 11, width = 5.5)
+# fancy.plot(raw_results_emp, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = c(19,15,17), xlab = "raw metric")
+# dev.off()
 ```

### 2025-01-14 Functions/extract.table.R

@@ -14,7 +14,7 @@
 #' @export
 
 ## Extract table
-extract.table <- function(results, centre = TRUE, scale.method) {
+extract.table <- function(results, centre = TRUE, scale = TRUE, scale.method) {
 
     # stop("DEBUG extract.table")
     # results <- all_results
@@ -35,17 +35,19 @@ extract.table <- function(results, centre = TRUE, scale.method) {
     }
 
     ## Scale all the metrics between each stressor
-    if(!missing(scale.method)) {
-        if(scale.method == "between") {
-            ## Merging all the results together to the the max per metric
-            metric_max <- apply(do.call(rbind, results), 2, function(x) return(max(abs(x))))
-            ## Scaling each result by its maximum
-            results <- lapply(results, function(X, max) t(apply(X, 1, function(x, max) return(x/max), max = metric_max)), max = metric_max)
-        }
-        ## Scale all the metrics within each stressor
-        if(scale.method == "within") {
-            ## Merging all the results together to the the max per metric
-             results <- lapply(results, function(one_results) apply(one_results, 2, function(X) return(X/max(abs(X)))))
+    if(scale) {
+        if(!missing(scale.method)) {
+            if(scale.method == "between") {
+                ## Merging all the results together to the the max per metric
+                metric_max <- apply(do.call(rbind, results), 2, function(x) return(max(abs(x))))
+                ## Scaling each result by its maximum
+                results <- lapply(results, function(X, max) t(apply(X, 1, function(x, max) return(x/max), max = metric_max)), max = metric_max)
+            }
+            ## Scale all the metrics within each stressor
+            if(scale.method == "within") {
+                ## Merging all the results together to the the max per metric
+                 results <- lapply(results, function(one_results) apply(one_results, 2, function(X) return(X/max(abs(X)))))
+            }
         }
     }
 

### 2025-01-15 Analysis/04_Summarise_results.Rmd
added raw values

@@
 ```{r}
-# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_2d.pdf", height = 11, width = 8.5)
+plot.raw.data <- function(raw_data, names, levels = 4, col = c(rep("orange", 4), rep("darkgreen", 4), rep("blue", 4))) {
 
-# layout(matrix(1:(12*5), 12, 5))
-# for(stressor in 1:5) {
-#     ## For each stressor
-#     plotting_data <- raw_results_2d[[stressor]]
-#     ## Plot the metric
-#     for(one_metric in 1:(ncol(plotting_data)/4)) {
-#         ## Get the metric name
-#         metric_name <- strsplit(colnames(selected_data), split = "_rm")[[1]][1]
-#         ## Get the data
-#         selected_data <- plotting_data[, 1:4]
+    n_stress <- length(raw_data)
+    n_metric <- ncol(raw_data[[1]])/levels
 
-#         ## Get the color TODO + add transparent
-#         col <- rep("orange", 4)
+    layout(matrix(0:((n_metric*n_stress)+n_stress+n_metric), n_metric+1, n_stress+1))
 
-#         ## Empty plot
-#         plot(NULL, ylim = c(0, 5), xlim = quantile(c(selected_data), prob =  c(0.025, 0.975)), yaxt = "n", ylab = "", xlab = metric_name)
+    ## Plot the rownames
+    for(metric in 1:n_metric) {
+        op <- par(mar = c(0,0,0,0), bty = "n")
+        plot(NULL, xlim = c(0, 1), ylim = c(0,1), xaxt = "n", yaxt = "n", ylab = "", xlab = "")
+        par(op)
+        text(rev(names)[metric], x = 0.5, y = 0.5)  
+    }
 
-#         ## Add the lines
-#         for(one_line in 1:4) {
-#             ## Get the CIs and median
-#             CIs <- quantile(selected_data[, one_line], prob = c(0.025, 0.25, 0.75, 0.975))
-#             med <- median(selected_data[, one_line])
-#             ## Plot the CIs
-#             lines(y = rep(5-one_line, 2), x = CIs[c(1, 4)], lty = 3, col = col)
-#             lines(y = rep(5-one_line, 2), x = CIs[c(2, 3)], lty = 1, col = col)
-#             ## Plot the median
-#             points(y =5-one_line, x = med, pch = 19, col = col)
-#         }
-#     }
-# }
+    for(stressor in 1:n_stress) {
+        ## For each stressor
+        plotting_data <- raw_data[[stressor]]
 
-# fancy.plot(raw_results_2d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "raw metric")
-# dev.off()
+        ## Plot the stressor name
+        op <- par(mar = c(0,0,0,0), bty = "n")
+        plot(NULL, xlim = c(0, 1), ylim = c(0,1), xaxt = "n", yaxt = "n", ylab = "", xlab = "")
+        par(op)
+        text(names(raw_data)[stressor], x = 0.5, y = 0.5)
 
-# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_4d.pdf", height = 11, width = 8.5)
-# fancy.plot(raw_results_4d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "raw metric")
-# dev.off()
+        ## Plot the metric
+        for(one_metric in 1:n_metric) {
+            ## Get the data
+            selected_data <- plotting_data[, 1:levels]
+            ## Remove the current data
+            plotting_data <- plotting_data[, -c(1:levels)]
 
-# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_8d.pdf", height = 11, width = 8.5)
-# fancy.plot(raw_results_8d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "raw metric")
-# dev.off()
+            ## Get the color TODO + add transparent
+            col_selected <- col[one_metric]
+            colours <- rev(sapply(seq(from = 0.2, to = 1, length.out = levels), function(x, col) adjustcolor(col_selected, alpha.f = x), col = col_selected))
 
-# pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_emp.pdf", height = 11, width = 5.5)
-# fancy.plot(raw_results_emp, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = c(19,15,17), xlab = "raw metric")
-# dev.off()
+            ## Empty plot
+            op <- par(mar = c(2,2,2,2), bty = "n")
+            plot(NULL, ylim = c(0, 5), xlim = quantile(c(selected_data), prob =  c(0.025, 0.975)), yaxt = "n", ylab = "", xlab = "")
+
+            ## Add the lines
+            for(one_line in 1:levels) {
+                ## Get the CIs and median
+                CIs <- quantile(selected_data[, one_line], prob = c(0.025, 0.25, 0.75, 0.975))
+                med <- median(selected_data[, one_line])
+                ## Plot the CIs
+                lines(y = rep(5-one_line, 2), x = CIs[c(1, 4)], lty = 3, col = colours[one_line])
+                lines(y = rep(5-one_line, 2), x = CIs[c(2, 3)], lty = 1, col = colours[one_line])
+                ## Plot the median
+                points(y =5-one_line, x = med, pch = 19, col = colours[one_line])
+            }
+            par(op)
+        }
+    }
+}
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_2d.pdf", height = 22.6661, width = 17)
+plot.raw.data(raw_results_2d, names = fancy_names, levels = 4, col = c(rep("orange", 4), rep("darkgreen", 4), rep("blue", 4)))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_4d.pdf", height = 22.6661, width = 17)
+plot.raw.data(raw_results_4d, names = fancy_names, levels = 4, col = c(rep("orange", 4), rep("darkgreen", 4), rep("blue", 4)))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_8d.pdf", height = 22.6661, width = 17)
+plot.raw.data(raw_results_8d, names = fancy_names, levels = 4, col = c(rep("orange", 4), rep("darkgreen", 4), rep("blue", 4)))
+dev.off()
+
+pdf(file = "../Manuscript/Figures/raw_results_per_metric_per_stressor_emp.pdf", height = 22.6661, width = 10.2)
+plot.raw.data(raw_results_emp, names = fancy_names, levels = 2, col = c(rep("orange", 4), rep("darkgreen", 4), rep("blue", 4)))
+dev.off()
+
 ```

### 2025-01-21 Manuscript/ecography_reponse_letter_2.tex
added raw values
Update ecography_reponse_letter_2.tex
up response

@@
 
-% TODO: double check the definition and make sure it aligns with the correct one. Say we fucked up explanation because of the negative effect (i.e. removing X is the same as keeping not X).
+We believe the confusion stems from what probability we are talking here.
+In a way here we approximate both competitive exclusion (i.e. increased ``extinction'' in densely populated regions) the same way as increased ``survival'' in sparsely populated regions.
+The reviewer is true that these two mechanisms are fundamentally different in ecology but we believe that for our approximations, they result in the same patterns: extinction in dense region (due to high competition) and survival in sparse regions (due to low competition).
+
+The formula illustrates how we measure the probability of being sampled which is decreasing in densely populated regions.
+This probability of being sampled reflects the probability of ``surviving'', i.e. of not being removed from the simulation.
+We've now added this precision throughout the explanation:
+
+\textit{(or, inversely, observations are more likely to be ``preserved''/``surviving'' in less dense regions)}. Lines @@@
 
@@
 % TODO: re double checking how the scaling works and what it does to the data!
+% USE SES instead?
 
@@
 
-% TODO: redo the supplementary plots but with different scale bars for every plot (and don't do the scaling)
+We have now updated all these supplementary figures to show the raw results with the scale bars indicating the range of values for each metric.
 

### 2025-01-21 Manuscript/main_ecography_R2.tex
up response

@@
 In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
-This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
+This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space (or, inversely, observations are more likely to be ``preserved''/``surviving'' in less dense regions).
 We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.

### 2025-01-21 Analysis/01_Simulating_spaces_and_stressors.Rmd
updated figure plots

@@
 plot(simulated_data, main = "The simulated BM through time")
-plot(simulated_data, main = "The 2D BM", trait = c(1,2))
+plot(simulated_data, main = "The 2D BM", trait = c(1,2), col = c("nodes" = "grey", "tips" = "blue"))
 
@@
 plot.text.box("A) Tree and trait\nsimulation:", cex = 2)
-plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (1D)", col = c(tips = "blue", "nodes" = "lightgrey"), ylab = "Trait 1")
-plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (2D)", col = c("lightgrey", "blue"), trait = c(1,2), xlab = "Trait 1", ylab = "Trait 2")
+plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (1D)", col = c("tips" = "blue", "nodes" = "lightgrey"), ylab = "Trait 1")
+plot(simulated_data, main = "Tree and trait simulation\nunder a BM process (2D)", col = c("tips" = "blue", "nodes" = "lightgrey"), trait = c(1,2), xlab = "Trait 1", ylab = "Trait 2")
 

### 2025-01-21 Manuscript/ecography_reponse_letter_2.tex
updated figure plots

@@
 
-% TODO: add colour legend to the plot
-% TODO: distinguish between the two scalings in the figure
-% TODO: expand scaling explanation in the caption
+We've now added the colour legend to the plot figure and removed the mention to the scaled metrics in the legend.
+As requested by this reviewer, we now only display the raw metrics scored (in the supplementaries) and the scaled and null centered metrics scores (in the manuscript and in the supplementaries).
 

### 2025-01-21 Functions/extract.table.R

@@ -40,8 +40,10 @@ extract.table <- function(results, centre = TRUE, scale = TRUE, scale.method) {
             if(scale.method == "between") {
                 ## Merging all the results together to the the max per metric
                 metric_max <- apply(do.call(rbind, results), 2, function(x) return(max(abs(x))))
+                metric_min <- apply(do.call(rbind, results), 2, function(x) return(min(abs(x))))
                 ## Scaling each result by its maximum
-                results <- lapply(results, function(X, max) t(apply(X, 1, function(x, max) return(x/max), max = metric_max)), max = metric_max)
+                # results <- lapply(results, function(X, max) t(apply(X, 1, function(x, max) return(x/max), max = metric_max)), max = metric_max)
+                results <- lapply(results, function(X, max, min) t(apply(X, 1, function(x, max, min) return((x-min)/(max-min)), max = metric_max, min = metric_min)), max = metric_max, min = metric_min)
             }
             ## Scale all the metrics within each stressor
             if(scale.method == "within") {

### 2025-01-23 Analysis/04_Summarise_results.Rmd
updated plots with remove.metrics

@@
 
-## Extracting all the results but not centred (still scaled)
-scaled_results_2d <- extract.table(all_results_2d, scale.method = "between", centre = FALSE)
-scaled_results_3d <- extract.table(all_results_3d, scale.method = "between", centre = FALSE)
-scaled_results_4d <- extract.table(all_results_4d, scale.method = "between", centre = FALSE)
-scaled_results_5d <- extract.table(all_results_5d, scale.method = "between", centre = FALSE)
-scaled_results_8d <- extract.table(all_results_8d, scale.method = "between", centre = FALSE)
-scaled_results_emp<- extract.table(empirical_results, scale.method = "between", centre = FALSE)
+# ## Extracting all the results but not centred (still scaled)
+# scaled_results_2d <- extract.table(all_results_2d, scale.method = "between", centre = FALSE)
+# scaled_results_3d <- extract.table(all_results_3d, scale.method = "between", centre = FALSE)
+# scaled_results_4d <- extract.table(all_results_4d, scale.method = "between", centre = FALSE)
+# scaled_results_5d <- extract.table(all_results_5d, scale.method = "between", centre = FALSE)
+# scaled_results_8d <- extract.table(all_results_8d, scale.method = "between", centre = FALSE)
+# scaled_results_emp<- extract.table(empirical_results, scale.method = "between", centre = FALSE)
 
-scaled_results_2d500 <- extract.table(all_results_2d500, scale.method = "between", centre = FALSE)
-scaled_results_4d500 <- extract.table(all_results_4d500, scale.method = "between", centre = FALSE)
-scaled_results_2d1000 <- extract.table(all_results_2d1000, scale.method = "between", centre = FALSE)
-scaled_results_4d1000 <- extract.table(all_results_4d1000, scale.method = "between", centre = FALSE)
+# scaled_results_2d500 <- extract.table(all_results_2d500, scale.method = "between", centre = FALSE)
+# scaled_results_4d500 <- extract.table(all_results_4d500, scale.method = "between", centre = FALSE)
+# scaled_results_2d1000 <- extract.table(all_results_2d1000, scale.method = "between", centre = FALSE)
+# scaled_results_4d1000 <- extract.table(all_results_4d1000, scale.method = "between", centre = FALSE)
 
@@
 ## The plot
-fancy.plot(scaled_cent_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
-
+fancy.plot(scaled_cent_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1), remove.metrics = c(4, 8, 12))
 ## Pdf version
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_5d.pdf", height = 9, width = 9)
-fancy.plot(scaled_cent_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
+fancy.plot(scaled_cent_results_5d, col = col_vector, lm = model_fits_5d, null = null_differences_5d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1), remove.metrics = c(4, 8, 12))
 dev.off()
@@
 pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_8d.pdf", height = 9, width = 9)
-fancy.plot(scaled_cent_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
+fancy.plot(scaled_cent_results_8d, col = col_vector, lm = model_fits_8d, null = null_differences_8d, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1), remove.metrics = c(4, 8, 12))
 dev.off()
@@
 ## The plot
-fancy.plot(scaled_cent_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17), xlim = c(-1, 1))
+fancy.plot(scaled_cent_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17), xlim = c(-1, 1), remove.metrics = c(4, 8, 12))
 
 ## Pdf version
-pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 11, width = 3.5)
-fancy.plot(scaled_cent_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17), xlim = c(-1, 1))
+pdf(file = "../Manuscript/Figures/results_per_metric_per_stressor_emp.pdf", height = 10, width = 3.5)
+fancy.plot(scaled_cent_results_emp, col = col_vector, lm = NULL, null = null_differences_emp, metric.names = fancy_names, pch = c(19,15,17), xlim = c(-1, 1), remove.metrics = c(4, 8, 12))
 dev.off()
@@
 fancy.plot(scaled_cent_results_4d1000, col = col_vector, lm = model_fits_4d1000, null = null_differences_4d1000, metric.names = fancy_names, pch = rep(19,4), xlim = c(-1, 1))
-dev.off()
-```
-
-## Raw results per metric
-
-```{r}
-
-pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_2d.pdf", height = 11, width = 8.5)
-fancy.plot(scaled_results_2d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
-dev.off()
-
-pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_4d.pdf", height = 11, width = 8.5)
-fancy.plot(scaled_results_4d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
-dev.off()
-
-pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_8d.pdf", height = 11, width = 8.5)
-fancy.plot(scaled_results_8d, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = rep(19,5), xlab = "scaled metric", xlim = c(0,1))
-dev.off()
-
-pdf(file = "../Manuscript/Figures/scaled_results_per_metric_per_stressor_emp.pdf", height = 11, width = 5.5)
-fancy.plot(scaled_results_emp, col = col_vector, lm = NULL, null = NULL, metric.names = fancy_names, pch = c(19,15,17), xlab = "scaled metric", xlim = c(0,1))
 dev.off()

### 2025-01-23 Functions/fancy.plot.R

@@ -18,7 +18,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, empirical = FALSE, xlab, xlim = c(-1, 1), ...) {
+fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend = median, col.metrics, lm, null, metric.names = NULL, empirical = FALSE, xlab, xlim = c(-1, 1), remove.metrics = NULL, ...) {
 
     ## Get the plot parameters
     plot_params <- list(...)
@@ -45,7 +45,8 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                            grid  = FALSE,
                            xlab  = xlab,
                            xlim  = xlim,
-                           metric.names = metric.names)
+                           metric.names = metric.names,
+                           remove.metrics = remove.metrics)
             } else {
                 par(mar = c(5, 0, 4, 1))
                 ## Making the empty plot
@@ -54,7 +55,8 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                            add.x = TRUE,
                            add.y = FALSE,
                            xlim  = xlim,
-                           xlab  = xlab)
+                           xlab  = xlab,
+                           remove.metrics = remove.metrics)
 
                 ## Adding the data
                 if(!missing(lm)) {
@@ -67,8 +69,9 @@ fancy.plot <- function(results, probs = c(0.025, 0.25, 0.75, 0.975), cent.tend =
                 } else {
                     null_data <- NULL
                 }
+
                 ## Plot the data
-                add.one.stressor(results[[one_stressor]], plot_params = plot_params, probs = probs, cent.tend = cent.tend, col = col.metrics, lm = model_data, null = null_data)
+                add.one.stressor(results[[one_stressor]], plot_params = plot_params, probs = probs, cent.tend = cent.tend, col = col.metrics, lm = model_data, null = null_data, remove.metrics = remove.metrics)
             }
         }
     } else {
[…]

### 2025-01-23 Manuscript/ecography_reponse_letter_2.tex
Update ecography_reponse_letter_2.tex
updated responses

@@
 
+We have now removed all the curves for the \texttt{TPD} metrics for any space of more than four dimensions (including the empirical space).
+We added this footnote to Table 3:
+
+\textit{-The \texttt{TPD} package currently only allows calculations for traitspaces up to four dimensions. This metric was exclude from traitspaces with more than four dimensions.}
 % TODO: clarify the artificial extinctions in the birds dataset (yes, we always start from 118 points).
-% TODO: Remove TPD metric altogether when it's not calculable?
 % TODO: double check if that's what the reviewer means.

### 2025-01-23 Manuscript/main_ecography_R2.tex
updated responses

@@
 In each pipeline, the dimensions where simulated as uncorrelated.
-We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait} - note that in the supplementary figures, we measured the \texttt{TPD} metrics on only the four first dimensions) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
+We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
 This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.
@@
 \hline
-Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
+Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND}* & \cite{carmona2019trait}\\
  & \cr 
@@
 \hline
-Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
+Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND}* & \cite{carmona2019trait}\\
  & \cr 
@@
 Regularity & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
-the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
+the height of bins in a histogram) & \texttt{TPD::REND}* & \cite{carmona2019trait}\\
  & \cr 
@@
 \label{Tab:metrics}
+* The \texttt{TPD} package currently only allows calculations for traitspaces up to four dimensions. This metric was exclude from traitspaces with more than four dimensions.
 \end{table}
@@
 Plain circles represent the metrics for the species present on Hawaii before 1500 (historic data; 46\% removal) and fainted squares for the species currently present on Hawaii (extant data; 69\% removal).
-\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
 The model results (including $R^{2}$ and t-statistics) are available in supplementary table 4.

### 2025-01-23 Manuscript/ecography_reponse_letter_2.tex
nearly done?

@@
 Standardizing to four dimensions would also ensure consistency with the dimensionality used in the simulated dataset and maintain uniformity throughout the analyses.
-Regarding the 8-dimensional space, since it is not central to the analysis, I suggest removing the TPD approach for this case to avoid introducing unnecessary variability, while clearly justifying this decision as a limit of the package in the text.
-Furthermore, upon focusing on Supplementary Figure 11, I have a concern.
+Regarding the 8-dimensional space, since it is not central to the analysis, I suggest removing the TPD approach for this case to avoid introducing unnecessary variability, while clearly justifying this decision as a limit of the package in the text.}
+
+We have now removed all the curves for the \texttt{TPD} metrics for any space of more than four dimensions (including the empirical space).
+We added this footnote to Table 3:
+
+\textit{-The \texttt{TPD} package currently only allows calculations for traitspaces up to four dimensions. This metric was exclude from traitspaces with more than four dimensions.}
+
+\textcolor{blue}{Furthermore, upon focusing on Supplementary Figure 11, I have a concern.
 Based on the description in the main text, I understand that the analyses are performed starting from a single trait space (comprising 118 points), where species are removed either randomly or based on actual extinction events, and metrics are then recalculated.
@@
 
-We have now removed all the curves for the \texttt{TPD} metrics for any space of more than four dimensions (including the empirical space).
-We added this footnote to Table 3:
-
-\textit{-The \texttt{TPD} package currently only allows calculations for traitspaces up to four dimensions. This metric was exclude from traitspaces with more than four dimensions.}
-% TODO: clarify the artificial extinctions in the birds dataset (yes, we always start from 118 points).
-% TODO: double check if that's what the reviewer means.
+Yes, the empirical trait space is based on 118 observations split between three groups: 37 species present on the island today, 63 species that were present on island in 1500 and 18 species that went extinct before 1500.
+We believe that the confusion here might come from the equivalence between species present (63 at 1500 and 37 in the present) and species absent (55 (46\%) at 1500 and 81 (68\%) in the present).
+We have tried now to clarify this in the caption of the figure 3 and in the one from the supplementary figure concerning the dimensionality selection.
 
@@
 
-% TODO: say we're keeping Box 2 because.
-% TODO: but also develop points they want.
+We believe that this specific point conflicts with reviewers 1 (``The addition of the text box at the end is particularly appreciated, as it effectively highlights the key message of the manuscript'' - reviewer 1 above).
+In this instance we thus prefer keeping the box 2 since we believe it might add value to at least some readers (and hopefully doesn't distract from the reading for other readers).
+Furthermore, it is not entirely clear to us what the reviewer meant with the metrics relationship ``varying across spaces with different dimensionalities''.
+Did they meant that spaces with different dimensionalities affect different metrics differently?
+We did mention the effect of the curse in the methods section:
+
+\textit{This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.} lines @@@
+
+We now also mention it in the discussion:
+
+\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature \citep{bellman1957dynamic}.} Lines @@@
 

### 2025-01-23 Manuscript/main_ecography_R2.tex
nearly done?

@@
 \caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (green = richness, yellow = divergence, blue = regularity).
-Plain circles represent the metrics for the species present on Hawaii before 1500 (historic data; 46\% removal) and fainted squares for the species currently present on Hawaii (extant data; 69\% removal).
+Plain circles represent the metrics for the species present on Hawaii at 1500 (historic data; 46\% removal - 63 species present) and fainted squares for the species currently present on Hawaii (extant data; 69\% removal - 37 species present).
 The model results (including $R^{2}$ and t-statistics) are available in supplementary table 4.
@@
 This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.
+Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature \citep{bellman1957dynamic}.
 

### 2025-01-23 Manuscript/supplementaries.tex
added raw values
updated responses
nearly done?

@@
 Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
-\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
 }}
@@
 \centering
-   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_2d.pdf}
+   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_8d.pdf}
 \caption{\scriptsize{\textbf{Raw metric scores for 8 dimensions:} the y axes represent the different metrics tested (sorted by categories).
@@
    \includegraphics{Figures/empirical_data_dimensions.pdf}
-\caption{\scriptsize{\textbf{Selection of the number of dimensions for the empirical data:} each histograms corresponds to the variance per dimension (dark grey) and the cumulative variance per dimension (light grey) for each group in the empirical data: the 37 currently endemic Hawaiian species (extant species) the 63 species present before 1500 (pre-1500 group).
-The plot shows that although 4 dimensions are enough to describe at least 95\% of the variance for the extant group, 5 dimensions are required for the pre-1500 group (following the method proposed in \cite{guillerme2023innovation}).
+\caption{\scriptsize{\textbf{Selection of the number of dimensions for the empirical data:} each histograms corresponds to the variance per dimension (dark grey) and the cumulative variance per dimension (light grey) for each group in the empirical data: the 37 currently endemic Hawaiian species (extant species) the 63 species present at 1500 (at_1500 group).
+The plot shows that although 4 dimensions are enough to describe at least 95\% of the variance for the extant group, 5 dimensions are required for the at_1500 group (following the method proposed in \cite{guillerme2023innovation}).
+Note that the whole trait space (light grey) contains 118 species.
+The 18 species extinct before 1500 are not contained in either of these groups.
 }}

### 2025-01-23 Analysis/04_Summarise_results.Rmd
updated fecking tables

@@
                       lapply(columns_diff_ses, `+`, 12))
+
 null_table_2d   <- fancy.table(null_differences_2d, columns = columns_diff_ses, digits = 2)
@@
 # kable(model_table_2d, caption = "Results of the model scaled metric ~ removal level per stressor (2D)", align = "c")
-print(xtable(change.names(model_table_2d, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (2D)"), file = "../Manuscript/Tables/results_model2d.tex")
+print(xtable(change.names(model_table_2d, rev(fancy_names)), caption = "Results of the model scaled metric $\\sim$ removal level per stressor (2D)"), file = "../Manuscript/Tables/results_model2d.tex")
 
-print(xtable(change.names(model_table_4d, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (4D)"), file = "../Manuscript/Tables/results_model4d.tex")
+print(xtable(change.names(model_table_4d, rev(fancy_names)), caption = "Results of the model scaled metric $\\sim$ removal level per stressor (4D)"), file = "../Manuscript/Tables/results_model4d.tex")
 
-print(xtable(change.names(model_table_8d, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (8D)"), file = "../Manuscript/Tables/results_model8d.tex")
+print(xtable(change.names(model_table_8d, rev(fancy_names)), caption = "Results of the model scaled metric $\\sim$ removal level per stressor (8D)"), file = "../Manuscript/Tables/results_model8d.tex")
 
-print(xtable(change.names(model_table_emp, fancy_names), caption = "Results of the model scaled metric ~ removal level per stressor (empirical data)"), file = "../Manuscript/Tables/results_empirical.tex")
+print(xtable(change.names(model_table_emp, rev(fancy_names)), caption = "Results of the model scaled metric $\\sim$ removal level per stressor (empirical data)"), file = "../Manuscript/Tables/results_empirical.tex")
 ```
@@
 ```{r, echo = FALSE, eval = FALSE}
-print(xtable(change.names(null_table_2d[, 1:8], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_2d[, 1:8], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences2d_1.tex")
 
-print(xtable(change.names(null_table_2d[, 9:16], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_2d[, 9:16], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences2d_2.tex")
 
-print(xtable(change.names(null_table_2d[, 17:24], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_2d[, 17:24], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences2d_3.tex")
 
-print(xtable(change.names(null_table_2d[, 25:32], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_2d[, 25:32], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences2d_4.tex")
 
-print(xtable(change.names(null_table_4d[, 1:8], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_4d[, 1:8], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences4d_1.tex")
 
-print(xtable(change.names(null_table_4d[, 9:16], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_4d[, 9:16], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences4d_2.tex")
 
-print(xtable(change.names(null_table_4d[, 17:24], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_4d[, 17:24], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences4d_3.tex")
 
-print(xtable(change.names(null_table_4d[, 25:32], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_4d[, 25:32], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences4d_4.tex")
 
-print(xtable(change.names(null_table_8d[, 1:8], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_8d[, 1:8], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences8d_1.tex")
 
-print(xtable(change.names(null_table_8d[, 9:16], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_8d[, 9:16], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences8d_2.tex")
 
-print(xtable(change.names(null_table_8d[, 17:24], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_8d[, 17:24], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences8d_3.tex")
 
-print(xtable(change.names(null_table_8d[, 25:32], fancy_names, colnames = TRUE),
+print(xtable(change.names(null_table_8d[, 25:32], rev(fancy_names)), colnames = TRUE,
   caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\\%, 40\\%, 60\\% and 80\\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differences8d_4.tex")
 
-print(xtable(change.names(null_table_emp, fancy_names, colnames = TRUE),
-  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each
-stressor (empirical); d = difference, s = standardised effect size. Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1."),
+print(xtable(change.names(null_table_emp[, ], rev(fancy_names)), colnames = TRUE,
+  caption = "Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (empirical); d = difference, s = standardised effect size.
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \\textit{g} with a correction for small-sample bias (using \\texttt{effectsize::hedges_g})."),
   file = "../Manuscript/Tables/results_differencesempirical.tex")

### 2025-01-23 Manuscript/Tables/results_differences2d_1.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:55:43 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & EQ.d\_1 & EQ.ses\_1 & EQ.d\_2 & EQ.ses\_2 & EQ.d\_3 & EQ.ses\_3 & EQ.d\_4 & EQ.ses\_4 \\ 
+ & EQ.d\_rm1 & EQ.ses\_rm1 & EQ.d\_rm2 & EQ.ses\_rm2 & EQ.d\_rm3 & EQ.ses\_rm3 & EQ.d\_rm4 & EQ.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 66.11*** & 4.1 & 68.86*** & 3.87 & 60.36*** & 3 & 43.83*** & 2.01 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 43.52*** & 4.02 & 41.95*** & 3.83 & 34.67*** & 2.89 & 23.62*** & 1.85 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 36.44*** & 4.88 & 41.74*** & 5.17 & 39.04*** & 4.21 & 30.16*** & 2.99 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 31.3*** & 5.32 & 37.36*** & 4.98 & 33.84*** & 3.53 & 23.2*** & 2.13 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.11*** & 3.48 & 0.05*** & 1.64 & 0.01*** & 0.44 & -0.01*** & -0.25 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 2.06*** & 4.98 & 1.78*** & 4.39 & 1.38*** & 3.23 & 0.92*** & 2.06 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.24*** & 2.01 & 0.17*** & 1.51 & 0.11*** & 1.03 & 0.06*** & 0.59 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 2.81*** & 20.32 & 2.07*** & 18.96 & 1.43*** & 15.61 & 0.82*** & 11.23 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.09*** & 4.35 & 0.04*** & 2.12 & 0.01*** & 0.7 & -0.01*** & -0.49 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.33*** & -5.46 & -0.24*** & -4.94 & -0.16*** & -3.81 & -0.09*** & -2.12 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0 & -0.01 & 0* & -0.18 & 0*** & -0.27 & 0*** & -0.23 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & -0.02*** & -0.69 & -0.03*** & -1.32 & -0.03*** & -1.58 & -0.03*** & -1.54 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 66.11*** & 4.1 & 68.86*** & 3.87 & 60.36*** & 3 & 43.83*** & 2.01 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 43.52*** & 4.02 & 41.95*** & 3.83 & 34.67*** & 2.89 & 23.62*** & 1.85 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 36.44*** & 4.88 & 41.74*** & 5.17 & 39.04*** & 4.21 & 30.16*** & 2.99 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 31.3*** & 5.32 & 37.36*** & 4.98 & 33.84*** & 3.53 & 23.2*** & 2.13 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.11*** & 3.48 & 0.05*** & 1.64 & 0.01*** & 0.44 & -0.01*** & -0.25 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 2.06*** & 4.98 & 1.78*** & 4.39 & 1.38*** & 3.23 & 0.92*** & 2.06 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.24*** & 2.01 & 0.17*** & 1.51 & 0.11*** & 1.03 & 0.06*** & 0.59 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 2.81*** & 20.32 & 2.07*** & 18.96 & 1.43*** & 15.61 & 0.82*** & 11.23 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.09*** & 4.35 & 0.04*** & 2.12 & 0.01*** & 0.7 & -0.01*** & -0.49 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.33*** & -5.46 & -0.24*** & -4.94 & -0.16*** & -3.81 & -0.09*** & -2.12 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0 & -0.01 & 0* & -0.18 & 0*** & -0.27 & 0*** & -0.23 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & -0.02*** & -0.69 & -0.03*** & -1.32 & -0.03*** & -1.58 & -0.03*** & -1.54 \\ 
    \hline
@@
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences2d_2.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:55:47 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & FA.d\_1 & FA.ses\_1 & FA.d\_2 & FA.ses\_2 & FA.d\_3 & FA.ses\_3 & FA.d\_4 & FA.ses\_4 \\ 
+ & FA.d\_rm1 & FA.ses\_rm1 & FA.d\_rm2 & FA.ses\_rm2 & FA.d\_rm3 & FA.ses\_rm3 & FA.d\_rm4 & FA.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 34.1*** & 1.51 & 32.08*** & 1.19 & 32.28*** & 1.14 & 28.07*** & 0.98 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 21.27*** & 1.29 & 18.72*** & 1.03 & 17.59*** & 0.96 & 14.25*** & 0.77 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 21.81*** & 1.94 & 21.56*** & 1.56 & 22.13*** & 1.53 & 19.58*** & 1.33 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 22.76*** & 2.2 & 26.13*** & 1.63 & 29.21*** & 2.18 & 24.35*** & 1.85 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0 & -0.03 & 0 & 0.1 & -0.01** & -0.16 & -0.01*** & -0.2 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.76*** & 1.12 & 0.6*** & 0.89 & 0.55*** & 0.82 & 0.46*** & 0.69 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.08*** & 0.77 & 0.06*** & 0.59 & 0.05*** & 0.47 & 0.03*** & 0.33 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.88*** & 3.28 & 0.7*** & 3.57 & 0.6*** & 4.16 & 0.44*** & 4.18 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.02*** & 0.61 & 0.01*** & 0.53 & 0.01*** & 0.31 & 0 & -0.09 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.09*** & -1.01 & -0.06*** & -0.84 & -0.05*** & -0.81 & -0.04*** & -0.73 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0** & 0.35 & 0 & 0.09 & 0*** & -0.24 & 0*** & -0.45 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.25*** & 5.31 & 0.15*** & 3.43 & 0.06*** & 2.12 & -0.01*** & -0.45 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 34.1*** & 1.51 & 32.08*** & 1.19 & 32.28*** & 1.14 & 28.07*** & 0.98 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 21.27*** & 1.29 & 18.72*** & 1.03 & 17.59*** & 0.96 & 14.25*** & 0.77 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 21.81*** & 1.94 & 21.56*** & 1.56 & 22.13*** & 1.53 & 19.58*** & 1.33 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 22.76*** & 2.2 & 26.13*** & 1.63 & 29.21*** & 2.18 & 24.35*** & 1.85 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0 & -0.03 & 0 & 0.1 & -0.01** & -0.16 & -0.01*** & -0.2 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.76*** & 1.12 & 0.6*** & 0.89 & 0.55*** & 0.82 & 0.46*** & 0.69 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.08*** & 0.77 & 0.06*** & 0.59 & 0.05*** & 0.47 & 0.03*** & 0.33 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.88*** & 3.28 & 0.7*** & 3.57 & 0.6*** & 4.16 & 0.44*** & 4.18 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.02*** & 0.61 & 0.01*** & 0.53 & 0.01*** & 0.31 & 0 & -0.09 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.09*** & -1.01 & -0.06*** & -0.84 & -0.05*** & -0.81 & -0.04*** & -0.73 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0** & 0.35 & 0 & 0.09 & 0*** & -0.24 & 0*** & -0.45 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.25*** & 5.31 & 0.15*** & 3.43 & 0.06*** & 2.12 & -0.01*** & -0.45 \\ 
    \hline
@@
 \caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
-FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+FA = FA; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences2d_3.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:55:50 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & FI.d\_1 & FI.ses\_1 & FI.d\_2 & FI.ses\_2 & FI.d\_3 & FI.ses\_3 & FI.d\_4 & FI.ses\_4 \\ 
+ & FI.d\_rm1 & FI.ses\_rm1 & FI.d\_rm2 & FI.ses\_rm2 & FI.d\_rm3 & FI.ses\_rm3 & FI.d\_rm4 & FI.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 44.79*** & 2.45 & 42.48*** & 1.98 & 36.94*** & 1.58 & 25.79*** & 1.08 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 31.28*** & 2.26 & 27.45*** & 1.85 & 22.78*** & 1.45 & 14.93*** & 0.95 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 28.59*** & 3.14 & 28.96*** & 2.67 & 26.06*** & 2.22 & 18.98*** & 1.58 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 21.29*** & 3.08 & 24.42*** & 2.89 & 22.75*** & 2.08 & 15.6*** & 1.39 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.04*** & 0.95 & 0.01 & 0.14 & -0.01** & -0.26 & -0.02*** & -0.41 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 1.27*** & 2.4 & 1.03*** & 1.88 & 0.81*** & 1.48 & 0.53*** & 0.98 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.14*** & 1.34 & 0.09*** & 0.92 & 0.06*** & 0.62 & 0.03*** & 0.33 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 1.2*** & 4.88 & 0.99*** & 5.02 & 0.78*** & 4.56 & 0.49*** & 4.4 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.06*** & 2.38 & 0.04*** & 1.71 & 0.03*** & 1.1 & 0.01*** & 0.46 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.05*** & -0.71 & -0.02*** & -0.4 & -0.01*** & -0.23 & 0 & -0.07 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0* & 0.25 & 0. & 0.17 & 0 & 0.11 & 0 & -0.03 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.02*** & 0.46 & 0.01*** & 0.34 & 0 & -0.03 & 0*** & -0.2 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 44.79*** & 2.45 & 42.48*** & 1.98 & 36.94*** & 1.58 & 25.79*** & 1.08 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 31.28*** & 2.26 & 27.45*** & 1.85 & 22.78*** & 1.45 & 14.93*** & 0.95 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 28.59*** & 3.14 & 28.96*** & 2.67 & 26.06*** & 2.22 & 18.98*** & 1.58 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 21.29*** & 3.08 & 24.42*** & 2.89 & 22.75*** & 2.08 & 15.6*** & 1.39 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.04*** & 0.95 & 0.01 & 0.14 & -0.01** & -0.26 & -0.02*** & -0.41 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 1.27*** & 2.4 & 1.03*** & 1.88 & 0.81*** & 1.48 & 0.53*** & 0.98 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.14*** & 1.34 & 0.09*** & 0.92 & 0.06*** & 0.62 & 0.03*** & 0.33 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 1.2*** & 4.88 & 0.99*** & 5.02 & 0.78*** & 4.56 & 0.49*** & 4.4 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.06*** & 2.38 & 0.04*** & 1.71 & 0.03*** & 1.1 & 0.01*** & 0.46 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.05*** & -0.71 & -0.02*** & -0.4 & -0.01*** & -0.23 & 0 & -0.07 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0* & 0.25 & 0. & 0.17 & 0 & 0.11 & 0 & -0.03 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.02*** & 0.46 & 0.01*** & 0.34 & 0 & -0.03 & 0*** & -0.2 \\ 
    \hline
@@
 \caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).
-FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+FI = FI; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences2d_4.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:55:52 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & CO.d\_1 & CO.ses\_1 & CO.d\_2 & CO.ses\_2 & CO.d\_3 & CO.ses\_3 & CO.d\_4 & CO.ses\_4 \\ 
+ & CO.d\_rm1 & CO.ses\_rm1 & CO.d\_rm2 & CO.ses\_rm2 & CO.d\_rm3 & CO.ses\_rm3 & CO.d\_rm4 & CO.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 17.17*** & 0.79 & 18.21*** & 0.76 & 16.88*** & 0.68 & 12.58*** & 0.48 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 13.24*** & 0.89 & 14.64*** & 0.96 & 12.41*** & 0.81 & 9.06*** & 0.57 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 11.24*** & 1.1 & 12.26*** & 1.05 & 11.73*** & 0.95 & 8.88*** & 0.67 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 7.71*** & 1.17 & 11*** & 1.24 & 11.32*** & 1.14 & 8.68*** & 0.74 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.04*** & 0.87 & 0.04*** & 1.01 & 0.03*** & 0.72 & 0.02*** & 0.49 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.51*** & 0.9 & 0.51*** & 0.93 & 0.41*** & 0.77 & 0.31*** & 0.56 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.06*** & 0.58 & 0.05*** & 0.52 & 0.04*** & 0.4 & 0.03*** & 0.26 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.71*** & 3.18 & 0.67*** & 4.05 & 0.53*** & 4.31 & 0.36*** & 3.85 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.04*** & 1.3 & 0.04*** & 1.73 & 0.03*** & 1.25 & 0.02*** & 0.83 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.02** & -0.29 & -0.01. & -0.13 & 0 & -0.08 & 0 & -0.01 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0** & 0.32 & 0*** & 0.35 & 0*** & 0.3 & 0*** & 0.19 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.03*** & 0.76 & 0.03*** & 0.99 & 0.01*** & 0.62 & 0.01*** & 0.32 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 17.17*** & 0.79 & 18.21*** & 0.76 & 16.88*** & 0.68 & 12.58*** & 0.48 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 13.24*** & 0.89 & 14.64*** & 0.96 & 12.41*** & 0.81 & 9.06*** & 0.57 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 11.24*** & 1.1 & 12.26*** & 1.05 & 11.73*** & 0.95 & 8.88*** & 0.67 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 7.71*** & 1.17 & 11*** & 1.24 & 11.32*** & 1.14 & 8.68*** & 0.74 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.04*** & 0.87 & 0.04*** & 1.01 & 0.03*** & 0.72 & 0.02*** & 0.49 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.51*** & 0.9 & 0.51*** & 0.93 & 0.41*** & 0.77 & 0.31*** & 0.56 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.06*** & 0.58 & 0.05*** & 0.52 & 0.04*** & 0.4 & 0.03*** & 0.26 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.71*** & 3.18 & 0.67*** & 4.05 & 0.53*** & 4.31 & 0.36*** & 3.85 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.04*** & 1.3 & 0.04*** & 1.73 & 0.03*** & 1.25 & 0.02*** & 0.83 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.02** & -0.29 & -0.01. & -0.13 & 0 & -0.08 & 0 & -0.01 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0** & 0.32 & 0*** & 0.35 & 0*** & 0.3 & 0*** & 0.19 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.03*** & 0.76 & 0.03*** & 0.99 & 0.01*** & 0.62 & 0.01*** & 0.32 \\ 
    \hline
 \end{tabular}
-\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (2D).CO = CO; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences4d_1.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:55:55 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & EQ.d\_1 & EQ.ses\_1 & EQ.d\_2 & EQ.ses\_2 & EQ.d\_3 & EQ.ses\_3 & EQ.d\_4 & EQ.ses\_4 \\ 
+ & EQ.d\_rm1 & EQ.ses\_rm1 & EQ.d\_rm2 & EQ.ses\_rm2 & EQ.d\_rm3 & EQ.ses\_rm3 & EQ.d\_rm4 & EQ.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 1024.68*** & 2.29 & 1305.5*** & 1.94 & 1221.81*** & 1.37 & 794.58*** & 0.74 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 2277.38*** & 2.33 & 2433.1*** & 2.04 & 2101.05*** & 1.48 & 1349.76*** & 0.84 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 449.97*** & 3.43 & 720.41*** & 3.62 & 780.96*** & 2.77 & 628.64*** & 1.8 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 42.72*** & 4.14 & 51.54*** & 3.7 & 46.87*** & 2.64 & 30.12*** & 1.5 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.13*** & 3.66 & 0.07*** & 1.91 & 0.03*** & 0.92 & 0.01*** & 0.16 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 2*** & 3.53 & 1.65*** & 2.84 & 1.25*** & 2.06 & 0.78*** & 1.23 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.16*** & 1.76 & 0.11*** & 1.25 & 0.08*** & 0.88 & 0.04*** & 0.52 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 4.19*** & 15.4 & 2.92*** & 15.04 & 1.97*** & 11.6 & 1.1*** & 8.48 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.01*** & 1.2 & 0*** & 0.37 & 0*** & -0.21 & 0*** & -0.55 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.23*** & -5.95 & -0.14*** & -5.16 & -0.08*** & -4.15 & -0.04*** & -2.33 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.56 & 0*** & -0.4 & 0*** & -0.49 & 0** & -0.12 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & -0.01** & -0.33 & -0.01*** & -0.68 & -0.01*** & -0.62 & -0.01*** & -0.54 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 1024.68*** & 2.29 & 1305.5*** & 1.94 & 1221.81*** & 1.37 & 794.58*** & 0.74 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 2277.38*** & 2.33 & 2433.1*** & 2.04 & 2101.05*** & 1.48 & 1349.76*** & 0.84 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 449.97*** & 3.43 & 720.41*** & 3.62 & 780.96*** & 2.77 & 628.64*** & 1.8 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 42.72*** & 4.14 & 51.54*** & 3.7 & 46.87*** & 2.64 & 30.12*** & 1.5 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.13*** & 3.66 & 0.07*** & 1.91 & 0.03*** & 0.92 & 0.01*** & 0.16 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 2*** & 3.53 & 1.65*** & 2.84 & 1.25*** & 2.06 & 0.78*** & 1.23 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.16*** & 1.76 & 0.11*** & 1.25 & 0.08*** & 0.88 & 0.04*** & 0.52 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 4.19*** & 15.4 & 2.92*** & 15.04 & 1.97*** & 11.6 & 1.1*** & 8.48 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.01*** & 1.2 & 0*** & 0.37 & 0*** & -0.21 & 0*** & -0.55 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.23*** & -5.95 & -0.14*** & -5.16 & -0.08*** & -4.15 & -0.04*** & -2.33 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0*** & -0.56 & 0*** & -0.4 & 0*** & -0.49 & 0** & -0.12 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & -0.01** & -0.33 & -0.01*** & -0.68 & -0.01*** & -0.62 & -0.01*** & -0.54 \\ 
    \hline

### 2025-01-23 Manuscript/Tables/results_differences4d_2.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:55:57 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & FA.d\_1 & FA.ses\_1 & FA.d\_2 & FA.ses\_2 & FA.d\_3 & FA.ses\_3 & FA.d\_4 & FA.ses\_4 \\ 
+ & FA.d\_rm1 & FA.ses\_rm1 & FA.d\_rm2 & FA.ses\_rm2 & FA.d\_rm3 & FA.ses\_rm3 & FA.d\_rm4 & FA.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 517.78*** & 0.82 & 783.51*** & 0.96 & 877.91*** & 0.9 & 724.23*** & 0.67 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 940.19*** & 0.71 & 1134.97*** & 0.72 & 1169.29*** & 0.68 & 945.03*** & 0.53 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 235.02*** & 1.11 & 401.13*** & 1.49 & 477.06*** & 1.45 & 439.74*** & 1.23 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 29.27*** & 0.74 & 51.57*** & 1.85 & 57.78*** & 2.4 & 43.32*** & 2.14 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0 & -0.03 & 0 & 0.05 & 0.01*** & 0.21 & 0*** & 0.1 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.36*** & 0.47 & 0.4*** & 0.53 & 0.39*** & 0.52 & 0.31*** & 0.43 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.03*** & 0.4 & 0.03*** & 0.37 & 0.03*** & 0.33 & 0.02*** & 0.24 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.53*** & 1.38 & 0.62*** & 2.38 & 0.6*** & 3.1 & 0.46*** & 3.63 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.03*** & 2.88 & 0.02*** & 2.69 & 0.02*** & 2.31 & 0.01*** & 0.93 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.03*** & -0.75 & -0.03*** & -1 & -0.03*** & -1.14 & -0.02*** & -0.92 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0. & -0.21 & 0*** & -0.75 & 0*** & -0.52 & 0*** & -0.4 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.22*** & 4.28 & 0.11*** & 3.5 & 0.03*** & 1.38 & -0.01*** & -0.55 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 517.78*** & 0.82 & 783.51*** & 0.96 & 877.91*** & 0.9 & 724.23*** & 0.67 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 940.19*** & 0.71 & 1134.97*** & 0.72 & 1169.29*** & 0.68 & 945.03*** & 0.53 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 235.02*** & 1.11 & 401.13*** & 1.49 & 477.06*** & 1.45 & 439.74*** & 1.23 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 29.27*** & 0.74 & 51.57*** & 1.85 & 57.78*** & 2.4 & 43.32*** & 2.14 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0 & -0.03 & 0 & 0.05 & 0.01*** & 0.21 & 0*** & 0.1 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.36*** & 0.47 & 0.4*** & 0.53 & 0.39*** & 0.52 & 0.31*** & 0.43 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.03*** & 0.4 & 0.03*** & 0.37 & 0.03*** & 0.33 & 0.02*** & 0.24 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.53*** & 1.38 & 0.62*** & 2.38 & 0.6*** & 3.1 & 0.46*** & 3.63 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.03*** & 2.88 & 0.02*** & 2.69 & 0.02*** & 2.31 & 0.01*** & 0.93 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.03*** & -0.75 & -0.03*** & -1 & -0.03*** & -1.14 & -0.02*** & -0.92 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0. & -0.21 & 0*** & -0.75 & 0*** & -0.52 & 0*** & -0.4 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.22*** & 4.28 & 0.11*** & 3.5 & 0.03*** & 1.38 & -0.01*** & -0.55 \\ 
    \hline
@@
 \caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
-FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+FA = FA; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences4d_3.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:56:00 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & FI.d\_1 & FI.ses\_1 & FI.d\_2 & FI.ses\_2 & FI.d\_3 & FI.ses\_3 & FI.d\_4 & FI.ses\_4 \\ 
+ & FI.d\_rm1 & FI.ses\_rm1 & FI.d\_rm2 & FI.ses\_rm2 & FI.d\_rm3 & FI.ses\_rm3 & FI.d\_rm4 & FI.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 667.99*** & 1.31 & 871.01*** & 1.14 & 808.97*** & 0.83 & 549.73*** & 0.48 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 1625.13*** & 1.34 & 1706.46*** & 1.14 & 1401.89*** & 0.83 & 908.74*** & 0.49 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 398.17*** & 2.59 & 587.5*** & 2.44 & 586.06*** & 1.89 & 437.87*** & 1.12 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 33.97*** & 3.24 & 40.35*** & 3.04 & 36.06*** & 1.83 & 23.54*** & 1.03 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.05*** & 1.29 & 0.04*** & 1.08 & 0.02*** & 0.67 & 0.01*** & 0.31 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 1.16*** & 1.84 & 0.91*** & 1.4 & 0.67*** & 1.01 & 0.42*** & 0.62 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.1*** & 1.15 & 0.07*** & 0.83 & 0.04*** & 0.55 & 0.02*** & 0.32 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 1.83*** & 4.43 & 1.42*** & 4.4 & 1*** & 3.98 & 0.59*** & 3.23 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.01*** & 1 & 0.01*** & 1.18 & 0.01*** & 0.85 & 0*** & 0.34 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.04*** & -1.03 & -0.02*** & -0.87 & -0.01*** & -0.49 & -0.01*** & -0.25 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0* & -0.26 & 0 & -0.14 & 0 & 0.1 & 0 & 0.07 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.02*** & 0.88 & 0.01*** & 0.54 & 0* & 0.18 & 0 & 0.09 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 667.99*** & 1.31 & 871.01*** & 1.14 & 808.97*** & 0.83 & 549.73*** & 0.48 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 1625.13*** & 1.34 & 1706.46*** & 1.14 & 1401.89*** & 0.83 & 908.74*** & 0.49 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 398.17*** & 2.59 & 587.5*** & 2.44 & 586.06*** & 1.89 & 437.87*** & 1.12 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 33.97*** & 3.24 & 40.35*** & 3.04 & 36.06*** & 1.83 & 23.54*** & 1.03 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.05*** & 1.29 & 0.04*** & 1.08 & 0.02*** & 0.67 & 0.01*** & 0.31 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 1.16*** & 1.84 & 0.91*** & 1.4 & 0.67*** & 1.01 & 0.42*** & 0.62 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.1*** & 1.15 & 0.07*** & 0.83 & 0.04*** & 0.55 & 0.02*** & 0.32 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 1.83*** & 4.43 & 1.42*** & 4.4 & 1*** & 3.98 & 0.59*** & 3.23 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.01*** & 1 & 0.01*** & 1.18 & 0.01*** & 0.85 & 0*** & 0.34 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.04*** & -1.03 & -0.02*** & -0.87 & -0.01*** & -0.49 & -0.01*** & -0.25 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0* & -0.26 & 0 & -0.14 & 0 & 0.1 & 0 & 0.07 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.02*** & 0.88 & 0.01*** & 0.54 & 0* & 0.18 & 0 & 0.09 \\ 
    \hline
@@
 \caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).
-FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+FI = FI; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences4d_4.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:56:03 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & CO.d\_1 & CO.ses\_1 & CO.d\_2 & CO.ses\_2 & CO.d\_3 & CO.ses\_3 & CO.d\_4 & CO.ses\_4 \\ 
+ & CO.d\_rm1 & CO.ses\_rm1 & CO.d\_rm2 & CO.ses\_rm2 & CO.d\_rm3 & CO.ses\_rm3 & CO.d\_rm4 & CO.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 160.54*** & 0.28 & 239.24*** & 0.28 & 239.86*** & 0.23 & 181.09*** & 0.15 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 436.95*** & 0.3 & 491.69*** & 0.28 & 445.45*** & 0.24 & 307.31*** & 0.15 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 115.67*** & 0.61 & 153.13*** & 0.48 & 159.12*** & 0.42 & 123.17*** & 0.27 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 6.44*** & 0.69 & 8.26*** & 0.61 & 8.72*** & 0.52 & 6.84*** & 0.36 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.04*** & 0.93 & 0.03*** & 0.91 & 0.02*** & 0.7 & 0.01*** & 0.44 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.28*** & 0.41 & 0.24*** & 0.34 & 0.19*** & 0.28 & 0.13*** & 0.18 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.03*** & 0.35 & 0.02*** & 0.26 & 0.02*** & 0.23 & 0.01*** & 0.14 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.68*** & 2.23 & 0.59*** & 2.72 & 0.44*** & 2.8 & 0.28*** & 2.85 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0*** & 0.39 & 0.01*** & 0.61 & 0.01*** & 0.61 & 0*** & 0.39 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.01*** & -0.43 & -0.01*** & -0.38 & -0.01*** & -0.3 & 0*** & -0.22 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0 & -0.08 & 0 & -0.02 & 0 & -0.02 & 0 & -0.02 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.01*** & 0.48 & 0.01*** & 0.5 & 0.01*** & 0.48 & 0*** & 0.21 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 160.54*** & 0.28 & 239.24*** & 0.28 & 239.86*** & 0.23 & 181.09*** & 0.15 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 436.95*** & 0.3 & 491.69*** & 0.28 & 445.45*** & 0.24 & 307.31*** & 0.15 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 115.67*** & 0.61 & 153.13*** & 0.48 & 159.12*** & 0.42 & 123.17*** & 0.27 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 6.44*** & 0.69 & 8.26*** & 0.61 & 8.72*** & 0.52 & 6.84*** & 0.36 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.04*** & 0.93 & 0.03*** & 0.91 & 0.02*** & 0.7 & 0.01*** & 0.44 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.28*** & 0.41 & 0.24*** & 0.34 & 0.19*** & 0.28 & 0.13*** & 0.18 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.03*** & 0.35 & 0.02*** & 0.26 & 0.02*** & 0.23 & 0.01*** & 0.14 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.68*** & 2.23 & 0.59*** & 2.72 & 0.44*** & 2.8 & 0.28*** & 2.85 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0*** & 0.39 & 0.01*** & 0.61 & 0.01*** & 0.61 & 0*** & 0.39 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.01*** & -0.43 & -0.01*** & -0.38 & -0.01*** & -0.3 & 0*** & -0.22 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0 & -0.08 & 0 & -0.02 & 0 & -0.02 & 0 & -0.02 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.01*** & 0.48 & 0.01*** & 0.5 & 0.01*** & 0.48 & 0*** & 0.21 \\ 
    \hline
 \end{tabular}
-\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (4D).CO = CO; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences8d_1.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:56:13 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & EQ.d\_1 & EQ.ses\_1 & EQ.d\_2 & EQ.ses\_2 & EQ.d\_3 & EQ.ses\_3 & EQ.d\_4 & EQ.ses\_4 \\ 
+ & EQ.d\_rm1 & EQ.ses\_rm1 & EQ.d\_rm2 & EQ.ses\_rm2 & EQ.d\_rm3 & EQ.ses\_rm3 & EQ.d\_rm4 & EQ.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 761.11*** & 1.34 & 962.64*** & 1.11 & 873.68*** & 0.81 & 544.78*** & 0.43 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 1031482.51*** & 0.39 & 1415980.4*** & 0.33 & 1316657.33*** & 0.23 & 798612.09*** & 0.12 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 7348.88*** & 2.06 & 29098.27*** & 1.93 & 48535.88*** & 1.7 & 47957.63*** & 1.11 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 52.07*** & 3.25 & 62.28*** & 3 & 55*** & 2.41 & 33.6*** & 1.35 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.08*** & 2.07 & 0.05*** & 1.37 & 0.03*** & 0.72 & 0.01*** & 0.29 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 2*** & 2.46 & 1.49*** & 1.83 & 1.07*** & 1.27 & 0.62*** & 0.73 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.12*** & 1.6 & 0.08*** & 1.16 & 0.06*** & 0.85 & 0.03*** & 0.5 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 6.09*** & 12.85 & 4.25*** & 12.1 & 2.83*** & 10.04 & 1.56*** & 7.82 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0*** & 0.47 & 0*** & 0.43 & 0*** & 0.22 & 0 & -0.01 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.07*** & -4.87 & -0.03*** & -3.81 & -0.02*** & -2.76 & -0.01*** & -1.71 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.72 & 0*** & -0.42 & 0*** & -0.29 & 0*** & -0.25 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.02*** & 0.86 & 0.01*** & 0.47 & 0*** & 0.25 & 0 & 0.05 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 1031482.51*** & 0.39 & 1415980.4*** & 0.33 & 1316657.33*** & 0.23 & 798612.09*** & 0.12 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 7348.88*** & 2.06 & 29098.27*** & 1.93 & 48535.88*** & 1.7 & 47957.63*** & 1.11 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 52.07*** & 3.25 & 62.28*** & 3 & 55*** & 2.41 & 33.6*** & 1.35 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 2*** & 2.46 & 1.49*** & 1.83 & 1.07*** & 1.27 & 0.62*** & 0.73 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.12*** & 1.6 & 0.08*** & 1.16 & 0.06*** & 0.85 & 0.03*** & 0.5 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 6.09*** & 12.85 & 4.25*** & 12.1 & 2.83*** & 10.04 & 1.56*** & 7.82 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.07*** & -4.87 & -0.03*** & -3.81 & -0.02*** & -2.76 & -0.01*** & -1.71 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0*** & -0.72 & 0*** & -0.42 & 0*** & -0.29 & 0*** & -0.25 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.02*** & 0.86 & 0.01*** & 0.47 & 0*** & 0.25 & 0 & 0.05 \\ 
    \hline

### 2025-01-23 Manuscript/Tables/results_differences8d_2.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:56:18 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & FA.d\_1 & FA.ses\_1 & FA.d\_2 & FA.ses\_2 & FA.d\_3 & FA.ses\_3 & FA.d\_4 & FA.ses\_4 \\ 
+ & FA.d\_rm1 & FA.ses\_rm1 & FA.d\_rm2 & FA.ses\_rm2 & FA.d\_rm3 & FA.ses\_rm3 & FA.d\_rm4 & FA.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 514.85*** & 1.18 & 555.45*** & 0.78 & 600.74*** & 0.78 & 443.13*** & 0.49 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 877212.48*** & 0.53 & 1040106.12*** & 0.36 & 1253555.6*** & 0.35 & 961196.13*** & 0.22 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 5448.5*** & 0.89 & 18489.32*** & 0.91 & 35772.76*** & 1.48 & 39487.59*** & 1.06 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 63.57*** & 1.5 & 83.62*** & 1.59 & 97.82*** & 4.02 & 75.03*** & 2.69 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & -0.02*** & -0.54 & -0.01*** & -0.4 & -0.01*** & -0.22 & 0*** & -0.11 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.13*** & 0.17 & 0.14*** & 0.17 & 0.14*** & 0.18 & 0.13*** & 0.16 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.02*** & 0.39 & 0.01*** & 0.24 & 0.01*** & 0.22 & 0.01*** & 0.16 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.49*** & 0.97 & 0.44*** & 1.4 & 0.43*** & 1.91 & 0.38*** & 2.35 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.02*** & 3.95 & 0.02*** & 2.45 & 0.01*** & 1.65 & 0.01*** & 0.72 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & 0* & -0.25 & -0.01*** & -0.72 & 0*** & -0.77 & 0*** & -0.69 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -1.09 & 0*** & -1.26 & 0*** & -1.19 & 0*** & -0.67 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.25*** & 7.86 & 0.13*** & 4.81 & 0.05*** & 2.64 & 0.01*** & 0.64 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 877212.48*** & 0.53 & 1040106.12*** & 0.36 & 1253555.6*** & 0.35 & 961196.13*** & 0.22 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 5448.5*** & 0.89 & 18489.32*** & 0.91 & 35772.76*** & 1.48 & 39487.59*** & 1.06 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 63.57*** & 1.5 & 83.62*** & 1.59 & 97.82*** & 4.02 & 75.03*** & 2.69 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.13*** & 0.17 & 0.14*** & 0.17 & 0.14*** & 0.18 & 0.13*** & 0.16 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.02*** & 0.39 & 0.01*** & 0.24 & 0.01*** & 0.22 & 0.01*** & 0.16 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.49*** & 0.97 & 0.44*** & 1.4 & 0.43*** & 1.91 & 0.38*** & 2.35 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & 0* & -0.25 & -0.01*** & -0.72 & 0*** & -0.77 & 0*** & -0.69 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0*** & -1.09 & 0*** & -1.26 & 0*** & -1.19 & 0*** & -0.67 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.25*** & 7.86 & 0.13*** & 4.81 & 0.05*** & 2.64 & 0.01*** & 0.64 \\ 
    \hline
@@
 \caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
-FA = facilitation; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+FA = FA; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences8d_3.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:56:25 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & FI.d\_1 & FI.ses\_1 & FI.d\_2 & FI.ses\_2 & FI.d\_3 & FI.ses\_3 & FI.d\_4 & FI.ses\_4 \\ 
+ & FI.d\_rm1 & FI.ses\_rm1 & FI.d\_rm2 & FI.ses\_rm2 & FI.d\_rm3 & FI.ses\_rm3 & FI.d\_rm4 & FI.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 512.45*** & 0.98 & 601.87*** & 0.74 & 538.56*** & 0.54 & 373.2*** & 0.32 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 759454.01*** & 0.51 & 1036742.44*** & 0.41 & 984207.67*** & 0.3 & 724993.83*** & 0.18 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 6282.8*** & 1.77 & 22928.75*** & 1.72 & 33761.59*** & 1.38 & 35235.64*** & 0.95 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 54.91*** & 3.53 & 71.72*** & 3.02 & 67.56*** & 2.39 & 46.86*** & 1.62 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.02*** & 0.45 & 0.01* & 0.15 & 0.01* & 0.15 & 0** & 0.12 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.66*** & 0.77 & 0.45*** & 0.56 & 0.33*** & 0.43 & 0.22*** & 0.29 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.06*** & 0.8 & 0.03*** & 0.5 & 0.02*** & 0.32 & 0.01*** & 0.19 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 1.73*** & 1.99 & 1.19*** & 1.89 & 0.86*** & 2.12 & 0.57*** & 2.4 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.02*** & 1.8 & 0.01*** & 1.45 & 0.01*** & 1.29 & 0*** & 0.52 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.01*** & -0.79 & 0*** & -0.57 & 0*** & -0.49 & 0*** & -0.32 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.96 & 0*** & -0.95 & 0*** & -0.73 & 0*** & -0.5 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.13*** & 1.73 & 0.07*** & 1.59 & 0.03*** & 1.41 & 0.01*** & 0.52 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 759454.01*** & 0.51 & 1036742.44*** & 0.41 & 984207.67*** & 0.3 & 724993.83*** & 0.18 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 6282.8*** & 1.77 & 22928.75*** & 1.72 & 33761.59*** & 1.38 & 35235.64*** & 0.95 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 54.91*** & 3.53 & 71.72*** & 3.02 & 67.56*** & 2.39 & 46.86*** & 1.62 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.66*** & 0.77 & 0.45*** & 0.56 & 0.33*** & 0.43 & 0.22*** & 0.29 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.06*** & 0.8 & 0.03*** & 0.5 & 0.02*** & 0.32 & 0.01*** & 0.19 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 1.73*** & 1.99 & 1.19*** & 1.89 & 0.86*** & 2.12 & 0.57*** & 2.4 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.01*** & -0.79 & 0*** & -0.57 & 0*** & -0.49 & 0*** & -0.32 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0*** & -0.96 & 0*** & -0.95 & 0*** & -0.73 & 0*** & -0.5 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.13*** & 1.73 & 0.07*** & 1.59 & 0.03*** & 1.41 & 0.01*** & 0.52 \\ 
    \hline
@@
 \caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).
-FI = filtering; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+FI = FI; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differences8d_4.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:56:30 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
   \hline
- & CO.d\_1 & CO.ses\_1 & CO.d\_2 & CO.ses\_2 & CO.d\_3 & CO.ses\_3 & CO.d\_4 & CO.ses\_4 \\ 
+ & CO.d\_rm1 & CO.ses\_rm1 & CO.d\_rm2 & CO.ses\_rm2 & CO.d\_rm3 & CO.ses\_rm3 & CO.d\_rm4 & CO.ses\_rm4 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 79.64*** & 0.14 & 130.26*** & 0.16 & 105.45*** & 0.1 & 92.81*** & 0.08 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 67526.58*** & 0.04 & 173367.22*** & 0.06 & 216655.44*** & 0.06 & 150636.07*** & 0.04 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 1312.02*** & 0.3 & 6855.69*** & 0.47 & 8664.07*** & 0.32 & 8093.46*** & 0.2 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 3.79*** & 0.29 & 7.28*** & 0.44 & 6.82*** & 0.32 & 5.38*** & 0.23 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.02*** & 0.54 & 0.02*** & 0.5 & 0.01*** & 0.4 & 0.01*** & 0.3 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.21*** & 0.27 & 0.21*** & 0.28 & 0.16*** & 0.21 & 0.1*** & 0.13 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.01*** & 0.18 & 0.01*** & 0.2 & 0.01*** & 0.14 & 0.01*** & 0.11 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.6*** & 1.4 & 0.61*** & 2.07 & 0.47*** & 2.15 & 0.32*** & 2.17 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0 & 0.11 & 0** & 0.23 & 0*** & 0.37 & 0*** & 0.27 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & 0** & -0.29 & 0*** & -0.54 & 0*** & -0.34 & 0*** & -0.21 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0 & -0.03 & 0 & -0.1 & 0 & 0.05 & 0 & -0.1 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.01* & 0.24 & 0.01*** & 0.31 & 0*** & 0.25 & 0 & 0.07 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 67526.58*** & 0.04 & 173367.22*** & 0.06 & 216655.44*** & 0.06 & 150636.07*** & 0.04 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 1312.02*** & 0.3 & 6855.69*** & 0.47 & 8664.07*** & 0.32 & 8093.46*** & 0.2 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 3.79*** & 0.29 & 7.28*** & 0.44 & 6.82*** & 0.32 & 5.38*** & 0.23 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.21*** & 0.27 & 0.21*** & 0.28 & 0.16*** & 0.21 & 0.1*** & 0.13 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.01*** & 0.18 & 0.01*** & 0.2 & 0.01*** & 0.14 & 0.01*** & 0.11 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.6*** & 1.4 & 0.61*** & 2.07 & 0.47*** & 2.15 & 0.32*** & 2.17 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & 0** & -0.29 & 0*** & -0.54 & 0*** & -0.34 & 0*** & -0.21 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0 & -0.03 & 0 & -0.1 & 0 & 0.05 & 0 & -0.1 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.01* & 0.24 & 0.01*** & 0.31 & 0*** & 0.25 & 0 & 0.07 \\ 
    \hline
 \end{tabular}
-\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).CO = competition; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
+\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (8D).CO = CO; d = difference, s = standardised effect size; 1 to 4 correspond to the levels of data removal (20\%, 40\%, 60\% and 80\%).
 Signif.
-codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.}
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_differencesempirical.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:22:09 2024
+% Thu Jan 23 18:56:35 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllll}
   \hline
- & empirical.d\_1 & empirical.ses\_1 & empirical.d\_2 & empirical.ses\_2 \\ 
+ & empirical.d\_rm1 & empirical.ses\_rm1 & empirical.d\_rm2 & empirical.ses\_rm2 \\ 
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & -7.28*** & -1.53 & -4.8*** & -1.08 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & -44.19*** & -1.74 & -9.72*** & -0.37 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & -17.07*** & -2.29 & -3.07*** & -0.55 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & -6.04*** & -2.38 & -4.32*** & -1.78 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.02*** & 1.45 & -0.03*** & -1.63 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.11*** & 1.28 & -0.05*** & -0.44 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.03*** & 1.64 & -0.03*** & -1.19 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & -0.41*** & -1.91 & -0.16*** & -0.52 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0* & 0.24 & 0.01*** & 2.35 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & 0.01*** & 0.97 & 0.02*** & 1.06 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0*** & -0.72 & 0*** & 0.69 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & -0.03*** & -2.23 & -0.02*** & -0.7 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & -44.19*** & -1.74 & -9.72*** & -0.37 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & -17.07*** & -2.29 & -3.07*** & -0.55 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & -6.04*** & -2.38 & -4.32*** & -1.78 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.11*** & 1.28 & -0.05*** & -0.44 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.03*** & 1.64 & -0.03*** & -1.19 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & -0.41*** & -1.91 & -0.16*** & -0.52 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & 0.01*** & 0.97 & 0.02*** & 1.06 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0*** & -0.72 & 0*** & 0.69 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & -0.03*** & -2.23 & -0.02*** & -0.7 \\ 
    \hline
 \end{tabular}
-\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each
-stressor (empirical); d = difference, s = standardised effect size. Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.} 
+\caption{Average difference between the null metric and the stressed metric (raw) for each level of removal and each stressor (empirical); d = difference, s = standardised effect size.
+Signif.
+codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1.
+Standardised effect size was calculated using the Hedges' \textit{g} with a correction for small-sample bias.}
 \end{table}

### 2025-01-23 Manuscript/Tables/results_model2d.tex
updated fecking tables

@@
 % latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Nov 28 15:16:05 2024
+% Thu Jan 23 18:48:58 2025
 \begin{table}[ht]
+\centering
 \scriptsize
-\centering
 \begin{tabular}{rllllllll}
@@
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 0.017*** & 0.02 & -0.015** & 0.012 & 0.026*** & 0.057 & -0.001 & -0.001 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 0.014** & 0.011 & -0.016** & 0.014 & 0.024*** & 0.038 & -0.009* & 0.007 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 0.007 & 0.002 & -0.006 & 0 & 0.03*** & 0.069 & 0.004 & 0 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 0.002 & -0.001 & -0.055*** & 0.07 & 0.001 & -0.001 & -0.02*** & 0.048 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.185*** & 0.578 & 0.022** & 0.009 & 0.102*** & 0.117 & 0.022** & 0.013 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.08*** & 0.396 & 0.009. & 0.003 & 0.054*** & 0.195 & 0.008* & 0.005 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.018** & 0.009 & -0.023*** & 0.04 & 0.014* & 0.004 & -0.022*** & 0.048 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.066*** & 0.436 & -0.028*** & 0.046 & -0.009. & 0.004 & -0.022*** & 0.038 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.227*** & 0.572 & 0.044*** & 0.037 & 0.091*** & 0.155 & 0.002 & -0.001 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.09*** & 0.293 & -0.002 & -0.001 & -0.027*** & 0.043 & -0.011* & 0.008 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & 0.018*** & 0.049 & 0.046*** & 0.179 & 0.004 & 0 & -0.012*** & 0.02 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.09*** & 0.356 & 0.14*** & 0.28 & 0.022*** & 0.037 & -0.009* & 0.005 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 0.02*** & 0.02 & -0.02** & 0.01 & 0.03*** & 0.06 & 0 & 0 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 0.01* & 0 & -0.02*** & 0.02 & 0.02*** & 0.03 & -0.01*** & 0.01 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 0.01 & 0 & -0.01 & 0 & 0.03*** & 0.06 & 0 & 0 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 0 & 0 & -0.06*** & 0.07 & 0 & 0 & -0.02*** & 0.05 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.19*** & 0.58 & 0.02** & 0.01 & 0.1*** & 0.12 & 0.02** & 0.01 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.08*** & 0.39 & 0.01. & 0 & 0.05*** & 0.19 & 0.01. & 0 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.02* & 0.01 & -0.02*** & 0.04 & 0.01. & 0 & -0.02*** & 0.05 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.06*** & 0.39 & -0.03*** & 0.05 & -0.01* & 0.01 & -0.02*** & 0.04 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.23*** & 0.57 & 0.04*** & 0.04 & 0.09*** & 0.16 & 0 & 0 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.09*** & 0.29 & 0 & 0 & -0.03*** & 0.04 & -0.01* & 0.01 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & 0.02*** & 0.05 & 0.05*** & 0.18 & 0 & 0 & -0.01*** & 0.02 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.09*** & 0.36 & 0.14*** & 0.28 & 0.02*** & 0.04 & -0.01* & 0 \\ 
    \hline
 \end{tabular}
-\caption{Results of the model scaled metric ~ removal level per stressor (2D)} 
+\caption{Results of the model scaled metric $\sim$ removal level per stressor (2D)} 
 \end{table}

### 2025-01-23 Manuscript/Tables/results_model4d.tex
updated fecking tables

@@
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 0.054*** & 0.143 & -0.009 & 0.002 & 0.032*** & 0.076 & 0.002 & 0 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & 0.013* & 0.007 & -0.032*** & 0.092 & 0.015** & 0.012 & -0.002 & 0 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & 0.049*** & 0.122 & 0.008 & 0.002 & 0.061*** & 0.194 & 0.018*** & 0.038 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & 0.03*** & 0.135 & -0.01 & 0.002 & 0.024*** & 0.09 & 0.001 & -0.001 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.164*** & 0.539 & -0.02** & 0.012 & 0.039*** & 0.023 & -0.002 & -0.001 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.108*** & 0.535 & -0.009** & 0.009 & 0.067*** & 0.258 & 0.012*** & 0.025 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.064*** & 0.145 & -0.004 & 0.002 & 0.045*** & 0.05 & 0.005* & 0.004 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.11*** & 0.52 & -0.03*** & 0.071 & 0.034*** & 0.065 & 0.004 & 0.001 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.153*** & 0.393 & 0.059*** & 0.083 & 0.014* & 0.005 & -0.035*** & 0.048 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.118*** & 0.517 & 0.013*** & 0.018 & -0.025*** & 0.06 & -0.002 & 0 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & -0.008** & 0.009 & -0.015*** & 0.016 & -0.006 & 0.002 & -0.001 & -0.001 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.016*** & 0.037 & 0.196*** & 0.535 & 0.014*** & 0.02 & -0.001 & -0.001 \\ 
+Functional richness\\ probability density\\ \texttt{TPD::REND} & 0.05*** & 0.14 & -0.01 & 0 & 0.03*** & 0.08 & 0 & 0 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & 0.01* & 0.01 & -0.03*** & 0.09 & 0.02** & 0.01 & 0 & 0 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & 0.05*** & 0.12 & 0.01 & 0 & 0.06*** & 0.19 & 0.02*** & 0.04 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & 0.03*** & 0.13 & -0.01 & 0 & 0.02*** & 0.09 & 0 & 0 \\ 
+  Divergence\\ probability density\\ \texttt{TPD::REND} & 0.16*** & 0.54 & -0.02** & 0.01 & 0.04*** & 0.02 & 0 & 0 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.11*** & 0.53 & -0.01** & 0.01 & 0.07*** & 0.26 & 0.01*** & 0.03 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.06*** & 0.15 & 0 & 0 & 0.04*** & 0.05 & 0.01* & 0 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.11*** & 0.52 & -0.03*** & 0.07 & 0.03*** & 0.06 & 0 & 0 \\ 
+  Functional evenness\\ probability density\\ \texttt{TPD::REND} & 0.15*** & 0.39 & 0.06*** & 0.08 & 0.01* & 0 & -0.03*** & 0.05 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.12*** & 0.52 & 0.01*** & 0.02 & -0.02*** & 0.06 & 0 & 0 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & -0.01** & 0.01 & -0.01*** & 0.02 & -0.01 & 0 & 0 & 0 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.02*** & 0.04 & 0.2*** & 0.53 & 0.01*** & 0.02 & 0 & 0 \\ 
    \hline
 \end{tabular}
-\caption{Results of the model scaled metric ~ removal level per stressor (4D)} 
+\caption{Results of the model scaled metric $\sim$ removal level per stressor (4D)} 
 \end{table}

### 2025-01-23 Manuscript/Tables/results_model8d.tex
updated fecking tables

@@
   \hline
-Functional evenness\\distances\\\texttt{FD::dbFD()\$FEve} & 0.039*** & 0.071 & 0.013** & 0.009 & 0.026*** & 0.051 & 0 & -0.001 \\ 
-  Functional evenness\\dendrogram\\\texttt{BAT::evenness} & -0.023*** & 0.041 & -0.038*** & 0.122 & -0.024*** & 0.102 & -0.008*** & 0.056 \\ 
-  Functional evenness\\hypervolume\\\texttt{BAT::kernel.evenness} & -0.055*** & 0.159 & -0.051*** & 0.121 & -0.034*** & 0.093 & -0.007*** & 0.015 \\ 
-  Functional evenness\\probability density\\\texttt{TPD::REND} & -0.032*** & 0.134 & -0.118*** & 0.345 & -0.059*** & 0.185 & -0.008*** & 0.03 \\ 
-  Rao's quadratic entropy\\distances\\\texttt{FD::dbFD()\$RaoQ} & 0.088*** & 0.157 & -0.02** & 0.01 & 0.013 & 0.002 & -0.012. & 0.004 \\ 
-  Functional dispersion\\dendrogram\\\texttt{BAT::dispersion} & 0.05*** & 0.166 & -0.022*** & 0.055 & 0.01. & 0.003 & -0.005 & 0.002 \\ 
-  Functional dispersion\\hypervolume\\\texttt{BAT::kernel.dispersion} & 0.035*** & 0.035 & -0.004 & 0.001 & 0.028*** & 0.029 & -0.007** & 0.009 \\ 
-  Divergence\\probability density\\\texttt{TPD::REND} & 0.037*** & 0.121 & -0.029*** & 0.101 & -0.005 & 0 & -0.017*** & 0.041 \\ 
-  Alpha diversity\\dendrogram\\\texttt{BAT::alpha} & 0.03*** & 0.019 & 0.08*** & 0.131 & 0.041*** & 0.028 & -0.041*** & 0.05 \\ 
-  Functional richness\\convex hull\\\texttt{BAT::hull.alpha} & -0.077*** & 0.265 & 0.031*** & 0.09 & -0.009* & 0.006 & 0.002 & -0.001 \\ 
-  Functional richness\\hypervolume\\\texttt{BAT::kernel.alpha} & -0.019** & 0.013 & -0.023*** & 0.017 & -0.02*** & 0.018 & 0.002 & -0.001 \\ 
-  Functional richness\\probability density\\\texttt{TPD::REND} & 0.014** & 0.012 & 0.175*** & 0.574 & 0.08*** & 0.112 & -0.002 & -0.001 \\ 
+  Functional richness\\ hypervolume\\ \texttt{BAT::kernel.alpha} & -0.02*** & 0.04 & -0.04*** & 0.12 & -0.02*** & 0.1 & -0.01*** & 0.06 \\ 
+  Functional richness\\ convex hull\\ \texttt{BAT::hull.alpha} & -0.06*** & 0.16 & -0.05*** & 0.12 & -0.03*** & 0.09 & -0.01*** & 0.02 \\ 
+  Alpha diversity\\ dendrogram\\ \texttt{BAT::alpha} & -0.03*** & 0.13 & -0.12*** & 0.35 & -0.06*** & 0.18 & -0.01*** & 0.03 \\ 
+  Functional dispersion\\ hypervolume\\ \texttt{BAT::kernel.dispersion} & 0.05*** & 0.17 & -0.02*** & 0.06 & 0.01. & 0 & 0 & 0 \\ 
+  Functional dispersion\\ dendrogram\\ \texttt{BAT::dispersion} & 0.03*** & 0.03 & 0 & 0 & 0.03*** & 0.03 & -0.01** & 0.01 \\ 
+  Rao's quadratic entropy\\ distances\\ \texttt{FD::dbFD()\$RaoQ} & 0.04*** & 0.12 & -0.03*** & 0.1 & -0.01 & 0 & -0.02*** & 0.04 \\ 
+  Functional evenness\\ hypervolume\\ \texttt{BAT::kernel.evenness} & -0.08*** & 0.26 & 0.03*** & 0.09 & -0.01* & 0.01 & 0 & 0 \\ 
+  Functional evenness\\ dendrogram\\ \texttt{BAT::evenness} & -0.02** & 0.01 & -0.02*** & 0.02 & -0.02*** & 0.02 & 0 & 0 \\ 
+  Functional evenness\\ distances\\ \texttt{FD::dbFD()\$FEve} & 0.01** & 0.01 & 0.17*** & 0.57 & 0.08*** & 0.11 & 0 & 0 \\ 
    \hline
 \end{tabular}
-\caption{Results of the model scaled metric ~ removal level per stressor (8D)} 
+\caption{Results of the model scaled metric $\sim$ removal level per stressor (8D)} 
 \end{table}

### 2025-01-23 Functions/fancy.table.R

@@ -16,6 +16,10 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
+
+# data[[1]][c(8,9),  c(1, 4)] #OK
+# extracted_data[[1]][c(8,9), , drop = FALSE] #OK
+
 fancy.table <- function(data, columns, symbols = c("", ".", "*", "**", "***"), digits = 2) {
     ## Making the fancy table
     return(do.call(cbind, lapply(data, get.columns, columns, symbols, digits)))

### 2025-01-23 Manuscript/supplementaries.tex
Update supplementaries.tex

--- 2025-01-09 Manuscript/supplementaries.tex
@@
    \includegraphics{Figures/empirical_data_dimensions.pdf}
-\caption{\scriptsize{\textbf{Selection of the number of dimensions for the empirical data:} each histograms corresponds to the variance per dimension (dark grey) and the cumulative variance per dimension (light grey) for each group in the empirical data: the 37 currently endemic Hawaiian species (extant species) the 63 species present before 1500 (pre-1500 group).
-The plot shows that although 4 dimensions are enough to describe at least 95\% of the variance for the extant group, 5 dimensions are required for the pre-1500 group (following the method proposed in \cite{guillerme2023innovation}).
-}}
+\caption{\scriptsize{\textbf{Selection of the number of dimensions for the empirical data:} each histograms corresponds to the variance per dimension (dark grey) and the cumulative variance per dimension (light grey) for each group in the empirical data: the 37 currently endemic Hawaiian species (extant species) the 63 species present at 1500 (``at 1500'' group).
+The plot shows that although 4 dimensions are enough to describe at least 95\% of the variance for the extant group, 5 dimensions are required for the ``at 1500'' group (following the method proposed in \cite{guillerme2023innovation}).Note that the whole trait space (light grey) contains 118 species.
+The 18 species extinct before 1500 are not contained in either of these groups.}}
 \label{Fig:empirical_dimensions}

### 2025-01-23 Manuscript/ecography_reponse_letter_2.tex
One more to do!

@@
 If the data are only divided by the maximum value, the results would generally have a maximum of 1 in cases of consistently negative or positive metrics across all levels of species exclusion (e.g., -2/-2 = 1 and +2/+2 = 1).
-Similarly, when some metrics have negative values and others are positive (across different levels of removal), the range would not necessarily be constrained between -1 and 1 unless all values fall within the range [0, 1] (e.g., functional richness might not range between -1 and 1).
-Furthermore, I am concerned that this second scaling could reduce or bias the patterns of the metrics, potentially affecting the conclusions of the manuscript.
+Similarly, when some metrics have negative values and others are positive (across different levels of removal), the range would not necessarily be constrained between -1 and 1 unless all values fall within the range [0, 1] (e.g., functional richness might not range between -1 and 1).}
+
+We did not use any metrics that output negative values.
+This means that the scaled and null centered values are always ranging between -1 and 1 in theory (even some values might not be achieved in practice).
+Hopefully the raw results in the supplementary materials now highlight this.
+
+\textcolor{blue}{Furthermore, I am concerned that this second scaling could reduce or bias the patterns of the metrics, potentially affecting the conclusions of the manuscript.
 For instance, in Supplementary Figure 8 (simulated data for 4 dimensions), taking FD:FEve for facilitation as a reference, subtracting the MetricStressor from the MetricNull results in a relatively linear pattern, with a general decrease in evenness as species removal increases (if I am seeing well from the figure), which aligns with expectations under this mechanism.
@@
 
-% TODO: re double checking how the scaling works and what it does to the data!
-% USE SES instead?
+We are not entirely sure how to handle this comment.
+If we understood correctly, the reviewer suggests we use a different null scaling method to see how that changes the results for specific metrics.
+Doing so will certainly result in different results.
+Although we believe that there is no need to test a range of scaling methods since the null centering one was chosen specifically to make the results more readable in this specific paper: does metric A captures a ``directional'' change in a specific mechanism induced pattern and is this change distinguishable from a null mechanism induced pattern?
+
+We are definitely sure that there are more possibilities of metric combinations, pattern simulations, results summarising that would be of interest to readers but with our limited time resources, we decide to chose only the specific one presented in this manuscript.
+Thankfully though, we made this whole project entirely reproducible so hopefully some worker wishing to pursue our work could easily add and modify what we did here.
 
@@
 For example, if the reported values are raw, the differences for functional divergence with TPD should be constrained between 0 and 1, as the metric itself is bounded within this range.
-However, some values in the tables exceed this limit.
-Finally, could you clarify how the SES is calculated in these cases?}
+However, some values in the tables exceed this limit.}
 
-% TODO: double check what's in the tables
-% TODO: in caption, explain how SES is calculated (reference the R package)
+Good spotting, the row names where mismatched (reversed) with the rows actual content.
+We have now fixed that in the supplementaries.
+
+\textcolor{blue}{Finally, could you clarify how the SES is calculated in these cases?}
+
+We used the Hedge's \textit{g} method to measure the standardised effect size (from the \texttt{effectsize} package) corrected for small sample sizes:
+
+\begin{equation}
+g = \frac{
+        median_{1} - median_{2}
+      }{
+    \sqrt{
+      \frac{
+           (n_{1}-1)\sigma_{1} + (n_{2}-1)\sigma_{2}
+          }{
+            n_{1}+n_{2}-2
+          }
+        }
+      }
+\end{equation}
 
@@
 
+We have double checked the manuscript again.
+Hopefully there are no more typos.
+
 \end{document}

### 2025-01-23 Manuscript/ecography_reponse_letter_2.tex
Update ecography_reponse_letter_2.tex

@@
 
-\textcolor{blue}{Recommendation by the Subject Editor (Dr. Barry Brook):}
+% \textcolor{blue}{Recommendation by the Subject Editor (Dr. Barry Brook):}
 
@@
 % - please focus on resolving these methodological and conceptual concerns to improve transparency and strengthen the presentation of your core findings.
+
 % - You will also need to address the referees’ points regarding:
@@
 \textit{Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g.
-\cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines @@@.
+Novack-Gottshall (2016a,b))), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines 89-92.
+%@@@
 
@@
 \textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e.
-observations adjacent to each other in the trait space are more likely compared to common ones).} Lines @@@
+observations adjacent to each other in the trait space are more likely compared to common ones).} Lines 175-178.
+%@@@
 
@@
 
-\textit{here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D} Lines @@@.
+\textit{here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D} Lines 202-203.% @@@.
 
@@
 This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
-These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines @@@
+These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines 431-436. %@@@
 
@@
 
-\textit{For example, we can follow the suggestions from \cite{mammola2021concepts} highlighting which kind of methodological pipeline best supports which type of metric of interest (Table 3).} Lines @@@
+\textit{For example, we can follow the suggestions from Mammola et al.
+(2021) highlighting which kind of methodological pipeline best supports which type of metric of interest (Table 3).} Lines 545-547.
+%@@@
 
@@
 \textit{First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
-a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
-We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).} Lines @@@
+a pure birth speciation model) until reaching 200 observations in \texttt{R} (R Core Team, 2024) using \texttt{treats} (Guillerme, 2024).
+We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).} Lines 124-124. %@@@
 
 However, we kept the broader definition of trait.
-\citealt{violle2007let} provided an ``unambiguous definition of plant trait, with a particular emphasis on functional trait'', which as per the authors aims was proposed to clarify and and conceptualise the definition on functional traits (``a surrogate of organismal performance'') specifically in plant ecology.
-Here we build on \citealt{mcgill2006rebuilding}'s more generalised approach, and, more interestingly on the work of \citealt{dawson2021traits} of how ecologists _actually_ operationalise the concept of trait based on researchers understanding vs.
+Violle et al.
+(2007) provided an ``unambiguous definition of plant trait, with a particular emphasis on functional trait'', which as per the authors aims was proposed to clarify and and conceptualise the definition on functional traits (``a surrogate of organismal performance'') specifically in plant ecology.
+Here we build on McGill et al.
+(2006)'s more generalised approach, and, more interestingly on the work of Dawson et al.
+(2021) of how ecologists \textit{actually} operationalise the concept of trait based on researchers understanding vs.
 proposed definitions.
-We propose through the paragraph on lines @@@ to expand that to ``any measurable aspect of an organism'' which we hope allows expansion beyond one dimensional traits and, more importantly, beyond ecology.
+We propose through the paragraph on lines 124-139 to expand that to ``any measurable aspect of an organism'' which we hope allows expansion beyond one dimensional traits and, more importantly, beyond ecology.
 I.e.
@@
 ``position of a specific foramen in a skull relative to another feature'').
-
-% TODO: also define this in the new in-text glossary table.
 
@@
 
-\textit{This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.} lines @@@
-
+\textit{This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.} lines 251-253.
 We now also mention it in the discussion:
 
-\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature \citep{bellman1957dynamic}.} Lines @@@
+\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature (Bellman et al., 1957).} Lines388-390.
+%@@@
 
@@
 
-\textit{Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).} Lines @@@
+\textit{Note that our definition of a trait is a generalisation ofViolle et al. (2007); McGill et al.
+(2006); Dawson et al. (2021)’s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).} At the end of Table 1.
 

### 2025-01-24 Manuscript/ecography_reponse_letter_2.tex
Update ecography_reponse_letter_2.tex

@@
 
-\textit{(or, inversely, observations are more likely to be ``preserved''/``surviving'' in less dense regions)}. Lines @@@
+\textit{(or, inversely, observations are more likely to be ``preserved''/``surviving'' in less dense regions)}. Lines 208-209. %@@@
 
@@
 
-\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature (Bellman et al., 1957).} Lines388-390.
+\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature (Bellman et al., 1957).} Lines 388-390.
 %@@@
@@
 We've now added the the glossary table (supplementary table 1) to the main text as suggested with the definition of trait redefined as a characteristic (good catch!).
-We did however keep our broader definition of a trait as opposed to \cite{violle2007let}'s (see response to first comment of Reviewer 3 above).
+We did however keep our broader definition of a trait as opposed to Violle et al (2007)'s (see response to first comment of Reviewer 3 above).
 We have added this precision in the table:

### 2025-01-27 Manuscript/main_ecography_R2.tex
Update main_ecography_R2.tex

@@
 
-\begin{table}[ht]
+\begin{table}[htbp]
 \caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
@@
 \hline
-Trait space & Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, traitspace, etc.  \\
+Trait space & Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, trait space, etc.  \\
 \hline
-Observations & Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs etc. \\
+Observations & Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs, etc. \\
 \hline
@@
 \hline
-Stressors' intensity & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction,  etc. \\
+Stressors' intensity & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc.
+& Severity of mass extinction, amount of co-occurring species, etc.
+\\
 \hline
@@
 
-\begin{table}
+\begin{table}[htbp]
 \center
@@
 \label{Tab:metrics}
-* The \texttt{TPD} package currently only allows calculations for traitspaces up to four dimensions. This metric was exclude from traitspaces with more than four dimensions.
+* The \texttt{TPD} package currently only allows calculations for trait spaces up to four dimensions. This metric was exclude from trait spaces with more than four dimensions.
 \end{table}
@@
 This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.
-Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature \citep{bellman1957dynamic}.
+Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the trait space and the metric nature \citep{bellman1957dynamic}.
 
@@
    \includegraphics[width=0.5\textwidth]{Figures/empirical_example.pdf}
-\caption{Example of a traitspace based on the empirical data described above.}
+\caption{Example of a trait space based on the empirical data described above.}
 \label{Fig:trait_space_example}

### 2025-01-29 Manuscript/ecography_reponse_letter_2.tex
updates until table 2

@@
 We are grateful for the detailed and useful comments from all two reviewers
-We have taken all their suggestions on board, integrated changes associated with the suggestion into the manuscript and responded to them in detail below
+We have integrated changes associated with the suggestion into the manuscript and responded to them in detail below
 For clarity we have displayed the \textcolor{blue}{reviewers' comments in blue} and our own responses in black
@@
 \textcolor{blue}{Line 89-90: I think that some references are necessary here.}
+
 We changed this sentence and merged it with the next one after discovering some previously missed literature:
 
-\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g.
+\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and processes (e.g.
 Novack-Gottshall (2016a,b))), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines 89-92.
@@
 \textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e.
-observations adjacent to each other in the trait space are more likely compared to common ones).} Lines 175-178.
+observations adjacent to each other in the trait space are more common).} Lines 175-178.
 %@@@
@@
 
-We remove the word ``say''.
+We removed the word ``say''.
 
@@
 
-\textit{First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
+\textit{First, we simulated multiple independent random time dependent traits (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
 a pure birth speciation model) until reaching 200 observations in \texttt{R} (R Core Team, 2024) using \texttt{treats} (Guillerme, 2024).
@@
 
+% TODO: Do you think we should briefly explain why we decided to stick with 5 axes and removed TPD, rather than drop to 4 like they suggest?
 We have now removed all the curves for the \texttt{TPD} metrics for any space of more than four dimensions (including the empirical space).
@@
 
-We believe that this specific point conflicts with reviewers 1 (``The addition of the text box at the end is particularly appreciated, as it effectively highlights the key message of the manuscript'' - reviewer 1 above).
+We believe that this specific point conflicts with reviewer 1 (``The addition of the text box at the end is particularly appreciated, as it effectively highlights the key message of the manuscript'' - reviewer 1 above).
 In this instance we thus prefer keeping the box 2 since we believe it might add value to at least some readers (and hopefully doesn't distract from the reading for other readers).

### 2025-01-29 Manuscript/main_ecography_R2.tex
updates until table 2

@@
 Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
-We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).
-For example competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
+We show that different metrics perform differently when trying to capture signal from different approximations of common mechanisms relative to no mechanism at all (null).
+For example, competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
 This emphasises the importance of not using a one-size-fits-all metric.
@@
 \textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
-Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space.
+Usually this consists of: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space.
 For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
@@
 
-Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g.
+Although some previous work has been done in understanding patterns in the context of mechanisms and processes (e.g.
 \cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.
@@
 Our goal is to test the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
-We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).
+We expect that our ability to capture a pattern does not only depend on the choice of metric (what) but also on the process and mechanism at hand (how and why).
 To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data.
-We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.
+We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data, with the exact same data sometimes being interpreted in opposite ways leading to sometimes opposite potential interpretations.
 
 \begin{table}[htbp]
-\caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
+\caption{Glossary and equivalence of terms used in this paper; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
 \centering
@@
 \hline
-Dimensions & Columns ($d$) & Traits, Ordination scores, distances, etc. & Traits, ordination scores, distances, etc. \\
+Dimensions & Columns ($d$) & Traits, ordination scores, distances, etc. & Traits, ordination scores, distances, etc. \\
 \hline
-Traits & Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional characteristic*; e.g. geographical location & uni/multidimensional characteristic; e.g. landmark coordinates \\
+Traits & Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional characteristic*; e.g. body mass, geographical location & uni/multidimensional characteristic; e.g. landmark coordinates \\
 \hline
@@
 \hline
-Stressors' intensity & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc.
+Stressors' intensity & The amount of observations removed (percentages) & Proportion of species affected, proportion of temperature change, etc.
 & Severity of mass extinction, amount of co-occurring species, etc.
@@
 \end{tabular}
-* Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).
+\label{Tab:glossary}
+* Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions expanded to include traits beyond the ones common in functional ecology (e.g.
+also in macroevolution, palaeontology, microbiology, etc.).
 \end{table}
@@
 
-First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
+First, we simulated multiple independent random time dependent traits (Brownian Motion) under a model where lineages only speciate (no extinction; i.e.
 a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
 We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).
-Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive).
+Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive as traits).
 These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
-For example, for trait defined as ``leaf insertion angle'', this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).
+For example, for a trait defined as ``leaf insertion angle'', this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).
 These observations on which the traits are measured can represent any biological focal entities, e.g.
 tips in a phylogenetic tree, individuals, species, populations, OTUs, etc.
-(see Glossary in the supplementary materials, table 1).
+(see Table \ref{Tab:glossary}).
 This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
@@
 In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e.
-observations adjacent to each other in the trait space are more likely compared to common ones).
+observations adjacent to each other in the trait space are more common).
 
@@
 This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space (or, inversely, observations are more likely to be ``preserved''/``surviving'' in less dense regions).
-We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
+We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share trait combinations.
 This could approximate the fact that observations that have similar trait combinations are more likely to go extinct due to competition (e.g. using the same resources). 
@@
 The algorithms used can lead to similar stressor effects depending on the simulated data distribution.
-For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors).
+For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just an average increase in density for ``increasing density'' and just an overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors).
 These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 intensities of data removal).
 We used 171 replicates because that was empirically the smallest number of replicates required to reach a variance between replicates lower than 1\% across all metrics (i.e.
-any additional replicates beyond 171 added less that 1\% extra variance).
+any additional replicates beyond 171 added less than 1\% extra variance).
 See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.
@@
 Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
-However, this is not tackled here for both simplicity and to understand how they work in isolation. 
-Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \ref{Tab:mechanisms}) should be taken as a cautious guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.
+However, this is not tackled here for both simplicity and to understand how the stressors work in isolation. 
+Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \ref{Tab:mechanisms}) should be taken as a cautionary guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.
 
 We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}), and for 2 and 8 dimensional (see supplementary materials figures 1 and 2).
-In each pipeline, the dimensions where simulated as uncorrelated.
+In each pipeline, the dimensions were simulated as uncorrelated.
 We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
@@
 observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient \citep{chesson2000mechanisms,barot2004mechanisms}.
-& Metrics mainly capturing changes in richness (but less in divergence and not in regularity) \\
+& Metrics mainly capturing changes in richness* (but less in divergence* and not in regularity*) \\
 
@@
 a community of species) with shared traits rather than the opposite.
-& Metrics capturing changes in regularity (but not richness and divergence) \\
+& Metrics capturing changes in regularity* (but not richness* and divergence*) \\
 
@@
 some regions of the trait space become unsuitable).
-& Metrics capturing changes in position (and richness) \\\\
+& Metrics capturing changes in position (and richness*) \\\\
 
@@
 At the extreme, it implies that only dissimilar observations will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}).
-& Metrics capturing changes in divergence (but not richness and regularity) \\\\
+& Metrics capturing changes in divergence* (but not richness* and regularity*) \\\\
 
@@
 Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution), thus, the type of metric expected to recover the mechanism can be variable.
-These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
+These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.
+    * Richness, divergence and regularity are dimensions of functional metrics that have been proposed to respectively reflect the sum of differences, the average differences and how regular the differences are among observations \citep{pavoine2011measuring,tucker2017guide,mammola2021concepts}.
+}
 }

### 2025-01-29 Manuscript/ecography_reponse_letter_2.tex
highligthed todos

@@
 \textit{Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
-This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
-These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines 431-436. %@@@
+This could be due to the taxonomic distribution of the extinct birds: these Hawaiian data comprise approximatively $58$\% passerines while the majority of anthropogenic extinctions have involved non-passerines (approximatively $75$\%).
+Non-passerines tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines 431-436. %@@@
 

### 2025-01-29 Manuscript/main_ecography_R2.tex
highligthed todos

@@
 F) \textbf{approximating competition mechanism}: proportionally removing observations from the centre of the distribution.
-The expected changes indicate the main type of metric affect by the stressor (following the aspects in \citealt{mammola2021concepts}).
+The expected changes indicate the main type of metric affected by the stressor (following the aspects in \citealt{mammola2021concepts}).
 G) we presented the results of the simulations as the scaled metric scores for the approximated mechanisms (C to F) relative to the null mechanism.
@@
 \hline
-Regularity & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g.
+Regularity & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution compared to an even spread of the distribution (e.g.
 the height of bins in a histogram) & \texttt{TPD::REND}* & \cite{carmona2019trait}\\
@@
  & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
+% TODO: comment from pedro: I think I mantioned this before, but this uses a tree, even if it is not an histogram
 \hline
@@
 This allowed us to see if there was any predictable linearity in the metric scores.
-In other words testing whether the metric scores relative to null removals where linearly scaling with the amount of data removed.
+In other words testing whether the metric scores relative to null removals were linearly scaling with the amount of data removed.
 
@@
 A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at $1500$ CE) and extant species list of the native community for each island were taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
-We also added two marine species to the prehistoric species list (\textit{Haliaeetus albicilla} and \textit{Pterodroma hypoleuca}); we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
 We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}.
@@
    \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d.pdf}
-\caption{\scriptsize{\textbf{Simulation results:} the y axes represent the different metrics tested (sorted by categories).
+\caption{\scriptsize{\textbf{Simulation results:} the y-axes represent the different metrics tested (sorted by categories).
 The different columns represent the different stressors.
@@
 Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted in a score increase (e.g.
-the Functional evenness based on a hypervolume for the equalising mechanism) or non-linear responses (e.g.
-Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores; see supplementary tables 2 to 5 and 6 to 17).
-Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
+the Functional evenness based on a hypervolume for the equalising mechanism; positive values in Figure \ref{Fig:simulation_results}) or non-linear responses (e.g.
+Alpha diversity based on a dendrogram for the equalising represent distributions of metric scores; see supplementary tables 2 to 5 and 6 to 17).
 Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulations show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).
@@
 randomly removing elements has more effects on the metric than removing them following the mechanism).
-However, some metrics (e.g. the Functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
-This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.
+However, some metrics (e.g. Functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
+This means that changes in metric values can be counter intuitive and should always be interpreted in the context of the metric, data and question at hand.
 Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the trait space and the metric nature \citep{bellman1957dynamic}.
@@
 Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
-This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
-These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
+This could be due to the taxonomic distribution of the extinct birds: these Hawaiian data comprise approximatively $58$\% passerines while the majority of anthropogenic extinctions have involved non-passerines (approximatively $75$\%).
+Non-passerines tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
 However, in Hawaii, where the same proportion of known species were passerines (approx. $58$\%), the opposite is true: the majority of extinctions (approx. $60$\%) have been of passerine species.
@@
 we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g.
-PCA or PCO), the dimensions have by definition a decreasing variance.
+PCA, PCO or PCoA), the dimensions have by definition a decreasing variance.
 
@@
 more than 200 dimensions in \citealt{van2023should}).
-In a higher number of dimensions (usually $>$10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
-This sometimes makes metric interpretation more chaotic \cite{bellman1957dynamic} in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
+In a higher number of dimensions (usually $>$10 but this is highly variable depending on the system under study), metrics results are more difficult to scale linearly.
+This sometimes makes metric interpretation more chaotic \citep{bellman1957dynamic} in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
 In fact, the relationship between observations has an important effect on how our results scale with higher dimensionality (e.g.
@@
 For example, if removing 80\% of the edges of a distribution was distinguishable or not from removing 80\% of a distribution randomly (e.g.
-facilitation mechanism measured as divergence using the probability density method - not distinguishable from null when removing 80\% or 60\% of the data; Figure \ref{Fig:simulation_results}).
+facilitation mechanism measured as divergence using the probability density method - not distinguishable from the null when removing 80\% or 60\% of the data; Figure \ref{Fig:simulation_results}).
 Some metrics (e.g. Functional evenness using the dendrogram method) were not able to distinguish between the null mechanism and the mechanism of interest (here the mechanism simulating competition).
 In other words, our statistical question was ``Does metric X distinguish between removing N\% of data in a biassed way (the mechanism) and removing N\% of data randomly (the null)?''.
-This definition of null hypothesis is also appropriate for the empirical data if the question is the same.
+This definition of the null hypothesis is also appropriate for the empirical data if the question is the same.
 However, it is very likely that researchers will ask a more exciting or intriguing question based on these data.
@@
 This legitimate question would thus require first a trait space (the pattern - what) and maybe some contrasting groups of interest like extinctions through time (the process - how) to answer the question (the mechanism - why).
-Furthermore, the scaling of the metrics paired with the levels of removal (80\%, 60\%, 40\% and 20\%) creates an expected artifact similarity when comparing the random removals against the other ones: when less data is available overall, the metrics are more likely to be similar (e.g.
-if removing 99\% of the data, we expect the random and removal metric score to be nearly identical to the non-random ones).
+Furthermore, the scaling of the metrics paired with the levels of removal (80\%, 60\%, 40\% and 20\%) creates an expected artefactual similarity when comparing the random removals against the removals due to stressors: when less data are available overall, the metrics are more likely to be similar (e.g.
+if removing 99\% of the data, we expect the random removal metric score to be nearly identical to the non-random ones).
 
 Importantly, it should always be borne in mind that mechanisms ought of course to be more complex in real world scenarios.
-For example, evolutionary mechanisms can vary through time or across clades (and so patterns are often the results of multiple processes); and ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
-facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
+For example, evolutionary mechanisms can vary through time or across clades (and so patterns are often the result of multiple processes); and ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
+facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional}).
 
@@
 
-\textit{We have data of bird traits for native species in Hawaii through time (e.g. before and after human colonisation of the archipelago).
-We want to know how anthropogenic stresses (e.g. hunting, invasive species, land use change) have affected the trait diversity of native Hawaiian birds.}
+\textit{We have data of bird traits for native species in Hawaii through time (e.g. the birds known to be present in 1500 CE and those still extant today).
+We want to know how anthropogenic stressors (e.g. hunting, invasive species, land use change) have affected the trait diversity of native Hawaiian birds.}
 
@@
    \includegraphics[width=0.5\textwidth]{Figures/empirical_example.pdf}
-\caption{Example of a trait space based on the empirical data described above.}
+\caption{Example of a trait space based on the empirical data described above. Species in 1500' refers to extinct species that were believed to have been extant in 1500 CE.}
 \label{Fig:trait_space_example}
@@
     \item \textbf{What?} Choose the appropriate pattern that would capture changes through time in relation to a filtering mechanism.
-In our example, this could be a continuous metric that measures either changes in the trait space position or/and changes in the trait space side.
-The idea being that filtering would remove either the edges or a specific portion of the trait space without affecting to many other parts of the trait space.
+In our example, this could be a continuous metric that measures either changes in the trait space position or/and changes in the trait space size.
+The idea being that filtering would remove either the edges or a specific portion of the trait space without affecting many other parts of the trait space.
 \end{itemize}
@@
 
-Once the one has clearly defined the pattern they want to capture (\textbf{what}), we believe it gets easier to pick up the type of metric that would be the most adequate.
+Once one has clearly defined the pattern they want to capture (\textbf{what}), we believe it becomes easier to select an appropriate metric.
 For example, we can follow the suggestions from \cite{mammola2021concepts} highlighting which kind of methodological pipeline best supports which type of metric of interest (Table \ref{Tab:box2}).
@@
 Regularity & Is one group more evenly disperse in space? & Not really \\
-Something else  & Does the position of a group changes in space? & Yes \\
+Something else  & Does the position of a group change in space? & Yes \\
+% TG: TODO: something about Beta Diversity (Pedro's comment?)
 \end{tabular}
@@
 
-\subsubsection{Get the metric most appropriate to your data:}
+\subsubsection{Select the metric most appropriate to your data:}
 
-So to describe the pattern for our specific example question, we will be interested in size and position of the trait space.
+To describe the pattern for our specific example question, we will be interested in size and position of the trait space.
 There is a lot of different metrics that capture changes in size and in position of the trait space and their choice should depend on the following questions:
@@
 Are you familiar with the metric (is there a software implementation you can use)?
-Of course, these are example questions and are not necessary the only ones you should ask.
+Of course, these are example questions and are not necessarily the only ones you should ask.
 \cite{mammola2021concepts} proposes a very handy flow chart.
@@
 \section{Conclusion}
-Different ecological and evolutionary processes or mechanisms do not always results in different patterns.
+Different ecological and evolutionary processes or mechanisms do not always result in different patterns.
 Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (\textit{aka} disparity, dissimilarity or functional diversity metrics).

### 2025-02-02 Manuscript/ecography_reponse_letter_2.tex
response updates

@@
 
-% TODO: Do you think we should briefly explain why we decided to stick with 5 axes and removed TPD, rather than drop to 4 like they suggest?
+The aim of including the empirical data was not to compare the metrics directly \textit{per se} but rather to provide a second example that readers preferring real world data can hopefully relate to.
+In line with the aim, we have thus decided to keep the empirical dataset to 5 dimensions, namely to hopefully link it to a paper that is currently in preparation by some of the authors (with different objectives but a same subset of the data).
+Furthermore, reducing the dataset to 4 dimensions would require rerunning the full analyses which will require CPU time that we currently don't have available because of other projects using our allocated computational time.
+
+We do agree with the reviewers point that including the \texttt{TPD} metrics that are limited to 4 dimensions can be misleading.
 We have now removed all the curves for the \texttt{TPD} metrics for any space of more than four dimensions (including the empirical space).

### 2025-02-02 Manuscript/main_ecography_R2.tex
response updates

@@
  & \cr 
- & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
-% TODO: comment from pedro: I think I mantioned this before, but this uses a tree, even if it is not an histogram
+ & Dendrogram & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
 \hline
@@
 Something else & Does the position of a group change in space? & Yes \\
-% TG: TODO: something about Beta Diversity (Pedro's comment?)
 \end{tabular}
